### 2024-08-15 backmatter/cv.tex
Add introduction and methods

% Structure

\newenvironment{cvlist}{
	\begin{description}[font=\normalfont]\setlength\itemsep{0em}}
		{\end{description}}

\newcommand{\cvitem}[2]{\item[#1,] #2}

%Preamble

%\begin{minipage}[b]{.7\linewidth}
%	\Large \textbf{Arne Van Den Kerchove}\\
%	\normalsize Doctoral Researcher Computational Neuroscience\\
%	KU Leuven, University of Lille
%  \bigskip
%
%  \texttt{arne.vandenkerchove@kuleuven.be \\
%  arne.vandenkerchove@univ-lille.fr \\
%  arne@vandenkerchove.com \\
%  +32 473 32 78 71 \\
%	\url{https://arne.vandenkerchove.com}\\
%	\url{https://orcid.org/0000-0002-9367-2986} \\
%  \url{https://linkedin.com/in/arnevdk} \\}
%
%%	\begin{tabular}{@{}c l}
%%		\faAt        & arne.vandenkerchove@kuleuven.be             \\
%%		\faAt        & arne.vandenkerchove@univ-lille.fr           \\
%%		\faAt        & arne@vandenkerchove.com                     \\
%%		\faPhone     & +32 473 32 78 71                            \\
%%		\faMapMarker & Leuven, Belgium                             \\
%%		\faMapMarker & Laboratory for Neuro- and Psychophysiology  \\
%%		             & KU Leuven-Campus Gasthuisberg             \\
%%		             & ON II-Herestraat 49-box 1021            \\
%%		             & BE-3000 Leuven                              \\
%%		\faMapMarker & UMR CRIStAL                                 \\
%%		             & Université de Lille-Campus scientifique   \\
%%		             & Bâtiment ESPRIT-Avenue Henri Poincaré     \\
%%		             & FR-59655 Villeneuve d'Ascq                  \\
%%		\faGlobe     & \url{https://arne.vandenkerchove.com}       \\
%%		\aiOrcid     & \url{https://orcid.org/0000-0002-9367-2986} \\
%%		\faLinkedin  & \url{https://linkedin.com/in/arnevdk}
%%	\end{tabular}
%\end{minipage}%
%\begin{minipage}[b]{.29\linewidth}
%	\includegraphics[width=\linewidth]{photo.jpg}
%\end{minipage}

% Curriculum vitae =============================================================
\chapter{Curriculum vitae}
\section*{About me}

I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

\paragraph{Keywords:} brain-computer interfaces, event-related
potentials, multilinear classification, covert visual attention

\subsection*{Education}

\begin{cvlist}
	\cvitem{Ph.D.\ in Biomedical Sciences}{KU Leuven, 2024}
	\cvitem{Ph.D.\ in Control Science and Signal Processing}{University of
		Lille, 2024}
	\cvitem{M.Sc.\ in Engineering Science: Computer Science}{KU Leuven, 2024}
	\cvitem{B.Sc.\ in Informatics}{KU Leuven, 2017}
\end{cvlist}

\subsection*{Publications}
\nocite{%
	VanDenKerchove2020,
	Libert2022,
	VanDenKerchove2022,
	VanDenKerchove2024,
}
\printbibliography[heading=none, resetnumbers=true]

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University, 2021--2024}
\end{cvlist}

\subsection*{Conferences and presentations}
\begin{cvlist}
	\cvitem{CORTICO Days}{CORTICO, Nancy, 2024 \\
		Poster presentation on covert visual attention BCIs for patients with
		oculomotor impairment.}
	\cvitem{10th International BCI Meeting}{BCI Society, Brussels, 2023\\
		Poster presentation on classifier-based latency estimation for
		gaze-independent BCIs.}
	\cvitem{Leuven.AI Scientific Workshop 2022}{Leuven.AI, Leuven, 2022\\
		Poster presentation on (multi-)Kronecker-structured linear discriminant
		analysis for low sample size event-related potential classification.}
	\cvitem{CORTICO Days}{CORTICO, Grenoble, 2022\\
		Presentation on Kronecker-structured LCMV-beamforming for event-related potential
		classification.}
	\cvitem{CORTICO Days}{CORTICO, 2021\\
		Presentation on a multi-component approach
		to spatiotemporal beamforming decoding of event-related potentials.}
	\cvitem{BCI \& Neurotechnology Spring School 2024}{g.tec, 2024}
	\cvitem{Closed Loop Neurotechnologies Autumn School}{NeurotechEU, Lille, 2023}
	\cvitem{CORTICO Days}{CORTICO, Paris, 2023}
	\cvitem{BCI \& Neurotechnology Spring School 2023}{g.tec, 2023}
	\cvitem{BCI \& Neurotechnology Masterclass Belgium}{g.tec, 2022}
	\cvitem{BCI \& Neurotechnology Spring School 2022}{g.tec, 2022}
\end{cvlist}

\subsection*{Teaching Experience}
\subsubsection*{Classes}
\begin{cvlist}
	\cvitem{Teaching Assistant Brain-Computer Interfaces}{KU Leuven, 2022--2024\\
		Teaching exercise sessions in BCI design and signal processing to students in
		the Master of Bioengineering and Advanced Master of Artificial
		Intelligence.}
	\cvitem{Teaching Assistant Fundamentals of Computer Science}{KU Leuven,
		2022 \\
		Teaching exercise sessions in Python programming and algorithmic reasoning to
		students in the Bachelor of	Engineering Sciences.}
\end{cvlist}

\subsubsection{Master theses and internships}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain\\ for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

\subsubsection*{Extracurricular}
\begin{cvlist}
	\cvitem{Project supervisor}{NeuroTech Leuven, 2022--ongoing\\
		Project supervisor and technical advisor for extracurricular
		neurotechnology student projects and a student team competing in the
		annual NeuroTechX BCI competition.}
	\cvitem{PAL tutor Principles of Computer Programming}{KU Leuven, 2015--2016\\
		Organizing and teaching peer-assisted learning sessions in Python programming for first year students.}
\end{cvlist}

\subsection*{Professional and volunteering Experience}
\begin{cvlist}
	\cvitem{Ambulance EMT}{Red Cross, FAST	vzw, 2022-ongoing}
	\cvitem{Python developer}{Mindspeller, 2019
		Python Flask developer in a KU Leuven spin-off that provides neuromarketing services based on
		scientifically validated neuroscience and AI research.}
	\cvitem{Freelance full-stack web developer}{Self-employed,
		2014--2020}
\end{cvlist}

\subsection*{Licenses and certifications}
\begin{cvlist}
	\cvitem{ICH Good Clinical Practice}{E6 TransCelerate BioPharma Inc.}
	\cvitem{Ambulance EMT}{FOD Volksgezondheid}
\end{cvlist}

\subsection*{Languages}
\begin{cvlist}
	\cvitem{Dutch}{Native}
	\cvitem{English}{Fully proficient}
	\cvitem{French}{Advanced}
	\cvitem{German}{Intermediate}
\end{cvlist}

### 2024-08-15 part1/kronlda.tex
Add introduction and methods

\chapter{Kronecker-structured discriminant analysis}

Part of the work in this chapter was published in~\cite{VanDenKerchove2022}.
\todo{Rename to LDA}

\section{Introduction}

Brain-computer interfaces (BCIs) establish a direct communication pathway
between the brain and an external device~\cite{Wolpaw2002}.
Severely disabled patients with impaired or absent communication capabilities
can benefit from BCIs to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
BCIs can be implemented in multiple ways, using non-invasive recording
techniques such as electroencephalography (EEG)~\cite{Abiri2019},
magnetoencephalography (MEG)~\cite{Mellinger2007},
functional Near-Infrared Spectroscopy (fNIRS)~\cite{Hong2015}, and Optically Pumped
Magnetometers (OPM MEG)~\cite{Paek2020}, or semi-invasive and invasive methods such as
electrocorticography (ECoG)~\cite{Schalk2011} or microelectrode
arrays~\cite{Maynard1997} which require surgery to implant a recording device.
While invasive BCIs yield the highest information transfer
rate~\cite{Willett2021}, non-invasive BCIs are preferable for short-term use
since they are not susceptible to the risks that come with surgery.
Of the non-invasive options, EEG is the most cost-effective and practical as it
is not limited to the same controlled settings as MEG and OPM MEG.

Besides the recording method, BCIs differ in the communication paradigms used
for communication~\cite{Abiri2019}.
A popular class of BCI paradigms relies on the evocation
of Event-Related Potentials (ERPs) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
calibration time before usage~\cite{Gao2014, Kapgate2015}.
The study we report on focuses on the visual P300 oddball ERP in response to a
rare but attended visual stimulus.
The decoder detects whether this ERP is present to determine which stimulus
the user attended.
The P300 paradigm has been used extensively in BCI development and is easy to
set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.

There are multiple state-of-the-art P300 classification methods, like Support
Vector Machines (SVMs)~\cite{Tayeb2014}, deep
learning models~\cite{vavreka2020evaluation,Borra2020}, and Riemannian Geometry
classifiers~\cite{Barachant2014}.
While these models often return a high classification accuracy, there is a need
for lightweight models -- lightweight models lead to fast off-line analyses and can be transferred to consumer-grade hardware.
When moving towards plug-and-play solutions, BCI calibration sessions should be short and model training times low.
The spatiotemporal beamformer~\cite{VanVliet2015, Wittevrongel2016} belongs to
this class of ERP decoding models as it achieves state-of-the-art performance and is fast to train.
Earlier work has shown that it is
possible to apply the spatiotemporal beamformer to multiple time-locked visual
BCI paradigms, including the P300 oddball paradigm,
Steady-State Visually Evoked Potentials (SSVEP), code-modulated Visually Evoked Potentials (cVEP)~\cite{Wittevrongel2017a} and
motion-onset Visually Evoked Potentials (mVEP)~\cite{Libert2021}.

This work shows that the original spatiotemporal
beamformer~\cite{Wittevrongel2016} can fall short in performance when BCI
calibration data are restricted.
We also show that the spatiotemporal beamformer does not scale well for
higher spatial and temporal resolution cases.
As a response to these issues, we introduce a regularization method that
exploits prior knowledge about the spatiotemporal nature of the EEG signal to
improve the accuracy for low data availability settings and speed up the
classifier training time, thereby considerably reducing memory usage.
Similar structured regularization approaches have been applied to other linear
ERP classifiers~\cite{gonzalez2017spatio, Vliet2020} and have shown
significant increases in performance.
Additionally, we show that regularization results in an interpretable
classification model, which can aid in analyzing and developing spatiotemporal beamformer-based classifiers.

\section{Materials \& methods}
\subsection{Notation}
We represent matrices with cursive capital letters, vectors with bold
lowercase letters, and scalars with cursive lowercase letters.
The epoched EEG data with $n$ epochs, $c$ channels, and $s$ samples are
represented in epoch format as $\{X_i\in\mathbb{R}^{c\times s}\}^n_{i=1}$ or flattened vector format by concatenating all channels for each epoch.
Flattening results in $\{\mathbf{x}_i\in\mathbb{R}^{cs}\}^n_{i=1}$ such that $\mathbf{x}_i = \text{vec}(X_i)$.
The real covariance matrix of the epochs in vector format is
denoted by $C$, estimators thereof as $\hat{C}$.

\subsection{Spatiotemporal beamforming}
LCMV-beamforming was initially introduced to EEG analysis as a filter for
source localization~\cite{VanVeen1997} to enhance the signal-to-noise ratio
(SNR).
Van Vliet et al.~\cite{VanVliet2015} first applied the spatiotemporal
LCMV-beamformer as a method for the analysis of ERPs.
The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
data-driven approaches proposed by Treder et al.~\cite{TREDER2016279} and
Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.

For the following analyses, we assume that all EEG channels are normalized with zero mean and unit variance without loss of generality.
Solving \autoref{eq:minimum_variance} under the linear constraint given by
\autoref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal LCMV-beamformer.
	\begin{equation}
		\underset{\mathbf{w}}{\arg\min}\mathbf{w}^\intercal C
		\mathbf{w}^\intercal
		\label{eq:minimum_variance}
	\end{equation}
	\begin{equation}
		\mathbf{a}^\intercal\mathbf{w} = 1
		\label{eq:linear_constraint}
	\end{equation}

These weights minimize the variance of the output of the filter while enhancing
the signal characterized by the constraint.
$\mathbf{a} = \text{vec}(A)$ is the data-driven activation pattern, a template
of the signal of interest maximizing the difference between two classes of
epochs, determined as follows:

	\begin{equation}
		\mathbf{a} =
		\frac{1}{|\text{class 1}|}\sum_\text{class 1}\mathbf{x_i} -
		\frac{1}{|\text{class 2}|}\sum_\text{class 2}\mathbf{x_i}
		\label{eq:activation_pattern}
	\end{equation}

The method of Lagrange multipliers then gives the closed-form solution to the minimization problem posed by
\autoref{eq:minimum_variance} and \autoref{eq:linear_constraint} as:

	\begin{equation}
		\mathbf{w} =
		\frac{C^{-1}\mathbf{a}^\intercal}
		{\mathbf{a}C^{-1}\mathbf{a}^\intercal}
		\label{eq:closed_form}
	\end{equation}

The beamformer can be applied to epochs (unseen or not) as:

	\begin{equation}
		y_i = \mathbf{w}\mathbf{x}_i
		\label{eq:apply_beamformer}
	\end{equation}

resulting in a scalar output per epoch.
The linear constraint in \autoref{eq:linear_constraint} ensures that the
beamformer maps epochs containing a target response to a score close to one
and, conversely, epochs not containing a target response to a score close to
zero.

\subsection{Covariance matrix regularization}
While the spatiotemporal beamformer, in theory, achieves optimal separation
between target and non-target classes, in analogy to linear discriminant analysis~\cite{TREDER2016279}, it does not always perform well on unseen data.
The main challenge is to find a good estimator for the inverse covariance matrix $C^{-1}$ since the real underlying covariance matrix generating the data is, in principle, unknown.

\subsubsection{Empirical covariance estimation}
\label{sec:empirical_covariance}
Earlier spatiotemporal beamformer studies~\cite{Wittevrongel2016,
Wittevrongel2016a, Wittevrongel2017, Wittevrongel2017a} use the empirical
covariance and inverse covariance calculated as follows:

	\begin{equation}
		\hat{C}_\text{emp} =
		\frac{1}{n-1}\sum^{n}_{i=1}\mathbf{x}_i\mathbf{x}_i^\intercal
		\label{eq:empirical_covariance}
	\end{equation}

	\begin{equation}
		\widehat{C^{-1}}_\text{emp} = \hat{C}_\text{emp}^+
		\label{eq:empirical_inverse_covariance}
	\end{equation}

The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
\autoref{fig:covs}a and \autoref{fig:covs}b respectively show examples of the
empirical estimators of the covariance and the inverse covariance matrices.
The empirical estimator suffers from performance and
stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.

\subsubsection{Shrunk covariance estimation}
\label{sec:shrunk_covariance}
The shrinkage covariance estimator creates a better-conditioned inversion matrix problem and generally performs better when applied to unseen data.
The estimators for the covariance and inverse covariance are given by:

	\begin{equation}
		\hat{C}_\alpha =
		(1-\alpha) \hat{C}_\text{emp}
		+ \alpha\frac{\text{Tr}(\hat{C}_\text{emp})}{cs}\mathbb{I}
		\label{eq:shrinkage}
	\end{equation}

	\begin{equation}
		\widehat{C^{-1}}_\alpha =
		\hat{C}^+_\alpha
		\label{eq:shrinkage_inverse}
	\end{equation}

with $0<\alpha<1$.
Analogous to L2 regularization of the beamforming problem,
shrinkage reduces the ratio between the smallest and largest eigenvalues
of the covariance matrix by strengthening the diagonal.
\autoref{fig:covs}c and \autoref{fig:covs}d respectively show examples of the
shrunk estimator of the covariance and the inverse covariance matrices.

Earlier work~\cite{Libert2021} applied shrinkage regularization to ERP
decoding with the spatiotemporal beamformer and showed competitive performance
compared to other state-of-the-art decoding techniques like stepwise LDA or SVM.
The abovementioned work chooses the shrinkage coefficient $\alpha$ as a fixed hyperparameter.
However, its optimal value depends on the number of training epochs, the
covariance matrix's dimensionality, and the independence and variance of the
data, which can vary across evaluation settings and per session.
The optimal value for $\alpha$ can be found with a line search using cross-validation, but this can be a costly procedure.
Methods exist to estimate an optimal shrinkage value from the data directly.
Most notable among these are the Ledoit-Wolf procedure~\cite{LEDOIT2004365},
Rao-Blackwell Ledoit-Wolf~\cite{chen2010shrinkage}, and Oracle Approximating Shrinkage~\cite{chen2010shrinkage}.
A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
cross-validation (LOOCV) scheme expressed by the data-driven closed-form
estimate:

	\begin{equation}
		\alpha =
		1-\frac{
			\frac{n}{n-1}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{1}{n(n-1)}\sum_{i=1}^n||\mathbf{x}_i||_2^4
		}
		{
			\frac{n^2 -2n}{(n-1)^2}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
			+ \frac{1}{n(n-1)^2}\sum_{i=1}^n||\mathbf{x}_i||_2^4
		}
		\label{eq:loocv}
	\end{equation}

We opt for the LOOCV shrinkage estimator because it avoids some of the
assumptions made by~\cite{LEDOIT2004365} and~\cite{chen2010shrinkage} and
because it generalizes to structured covariance estimation as described in
\autoref{seq:structured_estimation}.

\subsubsection{Spatiotemporal beamforming with Kronecker-Toeplitz structured covariance}
\label{seq:structured_estimation}
Exploiting prior knowledge about the spatiotemporal structure of the EEG signal leads to a more regularized estimator of the covariance.
When viewing the example of empirical spatiotemporal EEG covariance in
\autoref{fig:covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal EEG epochs is a Kronecker
product of two smaller
matrices~\cite{Munck1992,DeMunck1999,huizenga2002spatiotemporal}, as expressed
by:

	\begin{equation}
		\hat{C}_\text{struct} = \hat{S} \otimes \hat{T}
		\label{eq:kronecker}
	\end{equation}

with $\otimes$ the Kronecker product operator.
$\hat{S} \in \mathbb{R}^{c\times c}$ and $\hat{T} \in \mathbb{R}^{s\times s}$ respectively correspond to estimators of the spatial and temporal covariance of the data.
Furthermore, because the temporal covariance of the EEG-signal is
stationary (i.e., it is only dependent on interval length between covarying
time samples)~\cite{Bijma2003}, it is assumed to have a Toeplitz-matrix structure:

	\begin{equation}
		\hat{T}_{i,j} = \hat{T}_{i+1,j+1}
		\label{eq:toeplitz}
	\end{equation}

\autoref{prop:inverse_kronecker} then leads to
\autoref{eq:cov_inverse_kronecker} to estimate the inverse
covariance.
\begin{property}
	$(U \otimes V)^+ = U^+ \otimes V^+$ for any non-singular
	matrices $U$ and $V$~\cite{Langville2004}.
	\label{prop:inverse_kronecker}
\end{property}

	\begin{equation}
		\widehat{C^{-1}}_\text{struct} = \hat{S}^+ \otimes \hat{T}^+
	\label{eq:cov_inverse_kronecker}
	\end{equation}

Finally, based on \autoref{prop:kronecker_multiplication},
\autoref{eq:closed_form} can be reformulated more efficiently as
\autoref{eq:closed_form_kronecker}.
\begin{property}
	$(U\otimes V)\cdot\text{vec}(W) = \text{vec}(VWU^\intercal)$
	for any matrices $U\in\mathbb{R}^{p\times p}$,
	$V\in\mathbb{R}^{q\times q}$ and $W\in\mathbb{R}^{p\times q}$~\cite{Loan2000}.
	\label{prop:kronecker_multiplication}
\end{property}

	\begin{equation}
		\hat{\mathbf{w}}_\text{struct} =
		\frac{\hat{S}^+A^T\hat{T}^+}
		{\mathbf{a}\cdot\text{vec}(\hat{S}^+A^T\hat{T}^+)}
		\label{eq:closed_form_kronecker}
	\end{equation}

Using \autoref{eq:closed_form_kronecker} removes the need to calculate the full, high dimensional Kronecker product $\hat{S}^+\otimes
\hat{T}^+$.
\autoref{fig:covs}e and \autoref{fig:covs}f respectively show examples of the
structured covariance and inverse covariance estimators,
consisting of a spatial Kronecker factor (\autoref{fig:covs}g and
\autoref{fig:covs}h) and a temporal component (\autoref{fig:covs}i and
\autoref{fig:covs}j).

\begin{figure}
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
	\includegraphics[width=\linewidth]{figures/covs.eps}
\end{adjustwidth}
	\caption{Different estimators of the covariance and inverse covariance
		of 100 epochs of data from \textit{Subject 01} for channels
\textit{Fz}, \textit{Cz}, \textit{Pz}, and \textit{Oz} and time samples between 0.1s and 0.6s.
Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
determined by the closed-form LOOCV method. (\textbf{e,f}) Kronecker-Toeplitz
structured covariance and inverse covariance matrices.
(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
	\label{fig:covs}
\end{figure}
\unskip

The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
applications~\cite{de2002estimating,huizenga2002spatiotemporal,6408231,gonzalez2016kronecker,gonzalez2017spatio}.
Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
Our work goes further by embedding the Kronecker structure in the
spatiotemporal beamformer training process, using a data-adaptive shrinkage
method, and regularizing the covariance further by imposing a Toeplitz
structure on the temporal covariance.

\subsubsection{Kronecker-Toeplitz structured covariance estimation}
\label{sec:structured_covariance}
The question remains how to estimate $\hat{S}$ and $\hat{T}$.
While the Flip-Flop and Non-iterative Flip-Flop
algorithms~\cite{Lu2005, werner2008estimation, wirfalt2010toeplitz} can estimate Kronecker or Kronecker-Toeplitz structured covariances, new results show that a fixed point iteration is more efficient~\cite{Wiesel2012a,Wiesel2012}.
After each iteration, the spatial and temporal covariances matrices are scaled to unit
variance to ensure the fixed point iteration converges.
Finally, shrinkage can also be introduced in the Fixed Point Iteration to
improve stability and achieve more robust
regularization~\cite{Wiesel2012,Greenewald2014,6408231, Breloy2016}.
The spatial and temporal covariance matrices are shrunk at every fixed-point
iteration with shrinkage factors $\beta_k$ and $\gamma_k$ before matrix
inversion in the
next iteration.
Combined, this leads to the iterative estimation algorithm described by the
following equations:

	\begin{subequations}
	\begin{equation}
		\tilde{S}_{k+1} =
		\frac{1}{n}
		\sum^n_{i=1}X_i^\intercal\hat{T}_k^+X_i
		\label{eq:fpi_spatial}
	\end{equation}
	\begin{equation}
		\tilde{T}_{k+1} =
		\frac{1}{n}
		\sum^n_{i=1}X_i\hat{S}_k^+X_i^\intercal
		\label{eq:fpi_temporal}
	\end{equation}
\end{subequations}

	\begin{subequations}
	\begin{equation}
	\tilde{S}_{k+1}^{(\beta)} =
	(1-\beta_{k+1})\tilde{S}_{k+1}
	+\beta_{k+1}\frac{\text{Tr}(\tilde{S}_{k+1})}{c}\mathbb{I}
	\label{eq:fpi_spatial_shrunk}
	\end{equation}
	\begin{equation}
	\tilde{T}_{k+1}^{(\gamma)} =
	(1-\gamma_{k+1})\tilde{T}_{k+1}
	+\gamma_{k+1}\frac{\text{Tr}(\tilde{T}_{k+1})}{s}\mathbb{I}
	\label{eq:fpi_temporal_shrunk}
	\end{equation}
	\end{subequations}

	\begin{subequations}
	\begin{equation}
		\hat{S}_{k+1} =
		\frac{c}{\text{Tr}\left[\tilde{S}_{k+1}^{(\beta)}\right]}
		\tilde{S}_{k+1}^{(\beta)}
		\label{eq:fpi_spatial_norm}
	\end{equation}
	\begin{equation}
		\hat{T}_{k+1} =
		\frac{s}{\text{Tr}\left[\tilde{T}_{k+1}^{(\gamma)}\right]}
		\tilde{T}_{k+1}^{(\gamma)}
		\label{eq:fpi_temporal_norm}
	\end{equation}
\end{subequations}

$\hat{S}_0$ and $\hat{T}_0$ can be initialized to any positive definite matrix.
We choose to use the identity matrices $\mathbb{I}^{c\times c}$ and $\mathbb{I}^{s\times s}$.
After each iteration, all diagonals of $\hat{R}_{k+1}$ are set to their mean
values to ensure that $\hat{R}_{k+1}$ and $\hat{T}_{k+1}$ are Toeplitz structured.

Xie et al.~\cite{xie2021regularized} show that the LOOCV estimates for the
optimal values of $\beta_{k+1}$ and $\gamma_{k+1}$ also yield a closed-form
solution for the Kronecker fixed-point-iteration algorithm:

	\begin{subequations}
	\begin{equation}
		\beta_{k+1} =
		1-
		\frac{
			\frac{n}{n-1}\text{Tr}(\tilde{S}_{k+1}^2)
			- \frac{2}{c}\left[\text{Tr}(\tilde{S}_{k+1})\right]^2
			+ \frac{1}{c}\text{Tr}(\tilde{S}_{k+1}^2)
			- \frac{1}{n(n-1)}\sum_{i=1}^n
			\left[\text{Tr}(X_i\hat{T}_k^+X_i^\intercal)^2\right]
		}{
			\frac{n^2-2n}{(n-1)^2}\text{Tr}(\tilde{S}_{k+1}^2)
			- \frac{2}{c}\left[\text{Tr}(\tilde{S}_{k+1})\right]^2
			+ \frac{1}{c}\text{Tr}(\tilde{S}_{k+1}^2)
			+ \frac{1}{n(n-1)^2}\sum_{i=1}^n
			\left[\text{Tr}(X_i\hat{T}_k^+X_i^\intercal)^2\right]
		}
		\label{eq:spatial_shrinkage}
	\end{equation}
	\begin{equation}
		\gamma_{k+1} =
		1-
		\frac{
			\frac{n}{n-1}\text{Tr}(\tilde{T}_{k+1}^2)
			- \frac{2}{s}\left[\text{Tr}(\tilde{T}_{k+1})\right]^2
			+ \frac{1}{s}\text{Tr}(\tilde{T}_{k+1}^2)
			- \frac{1}{n(n-1)}\sum_{i=1}^n
			\left[\text{Tr}(X_i^\intercal\hat{S}_k^+X_i)^2\right]
		}{
			\frac{n^2-2n}{(n-1)^2}\text{Tr}(\tilde{T}_{k+1}^2)
			- \frac{2}{s}\left[\text{Tr}(\tilde{T}_{k+1})\right]^2
			+ \frac{1}{s}\text{Tr}(\tilde{T}_{k+1}^2)
			+ \frac{1}{n(n-1)^2}\sum_{i=1}^n
			\left[\text{Tr}(X_i^\intercal\hat{S}_k^+X_i)^2\right]
		}
		\label{eq:temporal_shrinkage}
	\end{equation}
\end{subequations}

The shrinkage parameters $0<\beta_{k+1}<1$ and $0<\gamma_{k+1}<1$ should be
re-determined after each iteration.
The Oracle Approximation Shrinkage method can also be used to determine
$\beta_{k+1}$ and $\gamma_{k+1}$~\cite{chen2010,xie2021regularized} but performs worse for spatiotemporal EEG data since not all assumptions are met.

\subsection{Dataset}
We use the dataset from~\cite{Wittevrongel2016}, containing P300 oddball EEG
recordings of 21 healthy subjects since it is a high-quality dataset with a high
number (32) of electrodes and concurrently recorded EOG responses for ocular artifact rejection.
Nine targets were arranged on a monitor before the subject during an
experimental session.
The subject was asked to pay attention to a cued target for a block
of stimulations.
The stimulations in a block are organized in 15 separate subsequent trials.
A trial is defined as 9 stimulations in which each target is flashed
precisely once per trial.
Each target was cued four times, resulting in a dataset consisting of 36 blocks
(4860 stimulations) per subject.
Each stimulation will correspond to a single epoch in the preprocessed dataset.
See~\cite{Wittevrongel2016} for a complete description of the dataset and the recording procedure.

\subsection{Software and preprocessing}
Data processing and classifier analysis were performed in Python using
Scikit-Learn (version 1.0.1)~\cite{pedregosa2011scikit} and SciPy (version
1.7.1)~\cite{2020SciPy-NMeth}.
The preprocessing pipeline was implemented using the MNE-Python toolbox
(version 0.24.0)~\cite{GramfortEtAl2013a}.
The dataset was converted to BIDS-EEG format~\cite{Pernet2019} and managed and
loaded with MNE-BIDS (version 0.9)~\cite{Appelhoff2019}.
The Riemannian classifier from \autoref{sec:riemannian} was implemented using
pyRiemann (version 0.2.7).
Statistical tests were performed in R (version 4.1.2).

The EEG recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
The reference electrodes were dropped from the analysis.
Data were subsequently filtered between 0.5Hz and 16Hz using forward-backward
filtering with a fourth-order Butterworth IIR filter.
The EEG signal was corrected for ocular artifacts using Independent Component
Analysis (ICA) by rejecting components that correlated with the bipolar EOG channels vEOG and hEOG, according to adaptive Z-score thresholding.
Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
No baseline correction was performed since this affects the temporal covariance
of the data, violating the Toeplitz structure assumption~\cite{Bijma2003}.

\subsection{Classification}
\subsubsection{Cross-validation scheme per subject}
We use a variation of grouped fold cross-validation per subject to evaluate the classifiers.
We apply 4-fold cross-validation by splitting the blocks of each subject into
four continuous folds.
Unlike regular cross-validation, we only use a single fold to train the
classifiers while using the other three folds for validation.
This scheme results in a training set of 9 blocks of 135 epochs each.
We chose this approach since we are primarily interested in the performance of the classifiers in the case of low data availability.
The classification task is to determine the cued target for each block.
The fraction of correctly predicted cues provides the accuracy of a classifier.
Data from all trials are used in the training fold, while classifier validation
is performed multiple times per fold, each time using an increasing amount of
trials (i.e., using the first trial, using the first two trials, etc. until all 15 trials
are used).
For each of the 9 stimulated targets, the averages over the corresponding epochs across
the utilized trials are used to predict the cued target in that block.
The target with the maximal classifier score was then chosen as the predicted
cued target.
Before training the classifiers, a Z-score normalization transformation was
developed on the training data to scale all EEG channels to unit variance.
This transformation was then applied to the validation data.

\subsubsection{Spatiotemporal beamformer classifier}
Before calculating the spatiotemporal beamformer (STBF), the signal was downsampled to
32Hz or twice the low-pass frequency 16Hz, resulting in 17 time samples
between 0.1 and 0.6s. According to the Nyquist Theorem, more samples would not
contain more information hence the minimum temporal resolution is chosen to reduce
the dimensionality of the covariance and improve its covariance.
The activation pattern is the difference between the averages of epochs in response to cued targets and the averages of those in response to non-cued targets.
We constructed three variations of the spatiotemporal beamformer:
STBF with empirical covariance estimation (\textsc{stbf-emp}) as in
\autoref{sec:empirical_covariance}, STBF with
LOOCV-shrunk covariance estimation (\textsc{stbf-shrunk}) as in
\autoref{sec:shrunk_covariance}, and STBF with
Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct} )with LOOCV shrinkage for
the Kronecker factors as in \autoref{sec:structured_covariance}.

\subsubsection{Riemannian geometry classifier}
\label{sec:riemannian}
We opted for a Riemannian
geometry-based classifier to compare our results.
The Riemannian model (x\textsc{dawn+rg}) uses the xDAWN spatial filter combined
with Riemannian geometry in tangent space as implemented by Barachant et
al.~\cite{barachant2014meg}.
This classifier uses four xDAWN spatial filters and each epoch's empirical spatial covariance matrix.
The target with the maximum score is the prediction of the cued target.
x\textsc{dawn+rg} was trained and validated without downsampling using epochs
at the original sample rate of 2048Hz.

\todo{check chapters/structure paper}
\todo{check structure Leuven.ai poster}
\section{The spatiotemporal EEG covariance}
\section{The single-KP covariance model}
\section{The sum-of-KP covariance model}
\section{Decoding performance}
\section{Discussion}
\subsection{Extension into the space-time-frequency domain}

### 2024-08-21 part2/stbf_struct.tex
Remove stbf-struct interpretation section

--- 2024-08-15 part1/kronlda.tex
@@
+\newcommand{\pv}[1]{
+	%\ifdim#1pt<0.05pt
+	%	\cellcolor{black!20}
+	%\fi
+	%\ifdim#1pt<0.001pt
+	%	$< 0.001$
+	%\else
+	%	\ifdim#1pt=1.000pt
+	%		\multicolumn{1}{c}{--}
+	%	\else
+	%		$#1$
+	%	\fi
+	%\fi
+	#1
+}
+% Kronecker-structured LDA ====================================================
+\begin{refsection}
 	\chapter{Kronecker-structured discriminant analysis}
@@
 	Part of the work in this chapter was published in~\cite{VanDenKerchove2022}.
+
 	\todo{Rename to LDA}
+	\todo{Work Kronecker sum in}
+	\todo{Add results over time vs toeplitzlda}
+	\todo{Introductory part on LCMV beamforming to intro}
 
@@
 
-Brain-computer interfaces (BCIs) establish a direct communication pathway
-between the brain and an external device~\cite{Wolpaw2002}.
-Severely disabled patients with impaired or absent communication capabilities
-can benefit from BCIs to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
-BCIs can be implemented in multiple ways, using non-invasive recording
-techniques such as electroencephalography (EEG)~\cite{Abiri2019},
-magnetoencephalography (MEG)~\cite{Mellinger2007},
-functional Near-Infrared Spectroscopy (fNIRS)~\cite{Hong2015}, and Optically Pumped
-Magnetometers (OPM MEG)~\cite{Paek2020}, or semi-invasive and invasive methods such as
-electrocorticography (ECoG)~\cite{Schalk2011} or microelectrode
-arrays~\cite{Maynard1997} which require surgery to implant a recording device.
-While invasive BCIs yield the highest information transfer
-rate~\cite{Willett2021}, non-invasive BCIs are preferable for short-term use
-since they are not susceptible to the risks that come with surgery.
-Of the non-invasive options, EEG is the most cost-effective and practical as it
-is not limited to the same controlled settings as MEG and OPM MEG.
+	%Brain-computer interfaces (BCIs) establish a direct communication pathway
+	%between the brain and an external device~\cite{Wolpaw2002}.
+	%Severely disabled patients with impaired or absent communication capabilities
+	%can benefit from BCIs to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
+	%BCIs can be implemented in multiple ways, using non-invasive recording
+	%techniques such as electroencephalography (EEG)~\cite{Abiri2019},
+	%magnetoencephalography (MEG)~\cite{Mellinger2007},
+	%functional Near-Infrared Spectroscopy (fNIRS)~\cite{Hong2015}, and Optically Pumped
+	%Magnetometers (OPM MEG)~\cite{Paek2020}, or semi-invasive and invasive methods such as
+	%electrocorticography (ECoG)~\cite{Schalk2011} or microelectrode
+	%arrays~\cite{Maynard1997} which require surgery to implant a recording device.
+	%While invasive BCIs yield the highest information transfer
+	%rate~\cite{Willett2021}, non-invasive BCIs are preferable for short-term use
+	%since they are not susceptible to the risks that come with surgery.
+	%Of the non-invasive options, EEG is the most cost-effective and practical as it
+	%is not limited to the same controlled settings as MEG and OPM MEG.
 
-Besides the recording method, BCIs differ in the communication paradigms used
-for communication~\cite{Abiri2019}.
-A popular class of BCI paradigms relies on the evocation
-of Event-Related Potentials (ERPs) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
-calibration time before usage~\cite{Gao2014, Kapgate2015}.
-The study we report on focuses on the visual P300 oddball ERP in response to a
-rare but attended visual stimulus.
-The decoder detects whether this ERP is present to determine which stimulus
-the user attended.
-The P300 paradigm has been used extensively in BCI development and is easy to
-set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.
+	%Besides the recording method, BCIs differ in the communication paradigms used
+	%for communication~\cite{Abiri2019}.
+	%A popular class of BCI paradigms relies on the evocation
+	%of Event-Related Potentials (ERPs) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
+	%calibration time before usage~\cite{Gao2014, Kapgate2015}.
+	%The study we report on focuses on the visual P300 oddball ERP in response to a
+	%rare but attended visual stimulus.
+	%The decoder detects whether this ERP is present to determine which stimulus
+	%the user attended.
+	%The P300 paradigm has been used extensively in BCI development and is easy to
+	%set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.
 
@@
 	Vector Machines (SVMs)~\cite{Tayeb2014}, deep
-learning models~\cite{vavreka2020evaluation,Borra2020}, and Riemannian Geometry
+	learning models~\cite{Vareka2020,Borra2020}, and Riemannian Geometry
 	classifiers~\cite{Barachant2014}.
@@
 	Similar structured regularization approaches have been applied to other linear
-ERP classifiers~\cite{gonzalez2017spatio, Vliet2020} and have shown
+	ERP classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
 	significant increases in performance.
@@
 	The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
-data-driven approaches proposed by Treder et al.~\cite{TREDER2016279} and
+	data-driven approaches proposed by Treder et al.~\cite{Treder2016} and
 	Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.
@@
 	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
-\autoref{fig:covs}a and \autoref{fig:covs}b respectively show examples of the
+	\autoref{fig:kronlda-covs}a and \autoref{fig:covs}b respectively show examples of the
 	empirical estimators of the covariance and the inverse covariance matrices.
@@
 	of the covariance matrix by strengthening the diagonal.
-\autoref{fig:covs}c and \autoref{fig:covs}d respectively show examples of the
+	\autoref{fig:kronlda-covs}c and \autoref{fig:covs}d respectively show examples of the
 	shrunk estimator of the covariance and the inverse covariance matrices.
@@
 	Methods exist to estimate an optimal shrinkage value from the data directly.
-Most notable among these are the Ledoit-Wolf procedure~\cite{LEDOIT2004365},
-Rao-Blackwell Ledoit-Wolf~\cite{chen2010shrinkage}, and Oracle Approximating Shrinkage~\cite{chen2010shrinkage}.
+	Most notable among these are the Ledoit-Wolf procedure~\cite{Ledoit2004},
+	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010shrinkage}, and Oracle Approximating Shrinkage~\cite{Chen2010shrinkage}.
 	A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
@@
 	We opt for the LOOCV shrinkage estimator because it avoids some of the
-assumptions made by~\cite{LEDOIT2004365} and~\cite{chen2010shrinkage} and
+	assumptions made by~\cite{Ledoit2004} and~\cite{Chen2010} and
 	because it generalizes to structured covariance estimation as described in
@@
 	When viewing the example of empirical spatiotemporal EEG covariance in
-\autoref{fig:covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
+	\autoref{fig:kronlda-covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
 	Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal EEG epochs is a Kronecker
 	product of two smaller
-matrices~\cite{Munck1992,DeMunck1999,huizenga2002spatiotemporal}, as expressed
+	matrices~\cite{Munck1992,DeMunck1999,Huizenga2002}, as expressed
 	by:
@@
 		\hat{T}^+$.
-\autoref{fig:covs}e and \autoref{fig:covs}f respectively show examples of the
+	\autoref{fig:kronlda-covs}e and \autoref{fig:covs}f respectively show examples of the
 	structured covariance and inverse covariance estimators,
-consisting of a spatial Kronecker factor (\autoref{fig:covs}g and
-\autoref{fig:covs}h) and a temporal component (\autoref{fig:covs}i and
-\autoref{fig:covs}j).
+	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
+	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:covs}i and
+	\autoref{fig:kronlda-covs}j).
 
-\begin{figure}
-	\begin{adjustwidth}{-\extralength}{0cm}
-		\centering
-	\includegraphics[width=\linewidth]{figures/covs.eps}
-\end{adjustwidth}
+	\begin{figure}[p]
+		\includegraphics[width=\linewidth]{part2/figures/stbf_struct/covs.eps}
 		\caption{Different estimators of the covariance and inverse covariance
@@
 			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
-	\label{fig:covs}
+		\label{fig:kronlda-covs}
 	\end{figure}
-\unskip
+	\todo{visualize kronecker sum}
+	\todo{visualize kronpca decomp pvl}
+	\todo{method pvl}
 
 	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
-applications~\cite{de2002estimating,huizenga2002spatiotemporal,6408231,gonzalez2016kronecker,gonzalez2017spatio}.
+	applications~\cite{DeMunck2002,Huizenga2002,6408231,GonzalezNavarro2016,GonzalezNavarro2017}.
 	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
@@
 	While the Flip-Flop and Non-iterative Flip-Flop
-algorithms~\cite{Lu2005, werner2008estimation, wirfalt2010toeplitz} can estimate Kronecker or Kronecker-Toeplitz structured covariances, new results show that a fixed point iteration is more efficient~\cite{Wiesel2012a,Wiesel2012}.
+	algorithms~\cite{Lu2005, Werner2008, Wirfaelt2010} can estimate Kronecker or Kronecker-Toeplitz structured covariances, new results show that a fixed point iteration is more efficient~\cite{Wiesel2012a,Wiesel2012}.
 	After each iteration, the spatial and temporal covariances matrices are scaled to unit
@@
 
-Xie et al.~\cite{xie2021regularized} show that the LOOCV estimates for the
+	Xie et al.~\cite{Xie2021} show that the LOOCV estimates for the
 	optimal values of $\beta_{k+1}$ and $\gamma_{k+1}$ also yield a closed-form
@@
 	The Oracle Approximation Shrinkage method can also be used to determine
-$\beta_{k+1}$ and $\gamma_{k+1}$~\cite{chen2010,xie2021regularized} but performs worse for spatiotemporal EEG data since not all assumptions are met.
+	$\beta_{k+1}$ and $\gamma_{k+1}$~\cite{Chen2010,Xie2021} but performs worse for spatiotemporal EEG data since not all assumptions are met.
 
@@
 	Data processing and classifier analysis were performed in Python using
-Scikit-Learn (version 1.0.1)~\cite{pedregosa2011scikit} and SciPy (version
-1.7.1)~\cite{2020SciPy-NMeth}.
+	Scikit-Learn (version 1.0.1)~\cite{Pedregosa2011} and SciPy (version
+	1.7.1)~\cite{Virtanen2020}.
 	The preprocessing pipeline was implemented using the MNE-Python toolbox
-(version 0.24.0)~\cite{GramfortEtAl2013a}.
+	(version 0.24.0)~\cite{Gramfort2013}.
 	The dataset was converted to BIDS-EEG format~\cite{Pernet2019} and managed and
@@
 	with Riemannian geometry in tangent space as implemented by Barachant et
-al.~\cite{barachant2014meg}.
+	al.~\cite{Barachant2014a}.
 	This classifier uses four xDAWN spatial filters and each epoch's empirical spatial covariance matrix.
@@
 	at the original sample rate of 2048Hz.
+
+	\section{Results}
+	\subsection{Minimum required fixed-point iterations}
+	The fixed point iteration algorithm described
+	in Equations \ref{eq:fpi_spatial}-\ref{eq:fpi_temporal_shrunk} estimates the Kronecker-Toeplitz structured covariance for the
+	\textsc{stbf-struct} classifier.
+	Fixed-point iteration is an iterative procedure starting from (in our case)
+	non-informed initial guesses for the spatial and temporal covariance matrices.
+	As a stopping criterion, one could impose a threshold on the difference in
+	outcome of successive steps, e.g., based on the covariance norm or the
+	classifier accuracy.
+	However, few iterations or even just one~\cite{castaneda2014estimation} suffice to achieve satisfactory performance in practice.
+
+	\autoref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
+	Using more than one fixed-point iteration does not significantly improve the
+	accuracy across the amounts of training data and the number of trials
+	used for evaluation.
+	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
+	the training process.
+
+	\begin{figure}
+		%\includegraphics[width=\linewidth]{figures/fpi.eps}
+		\caption{Average cross-validated \textsc{stbf-struct} Accuracy using
+			one trial per block over all 21 subjects
+			relative to the number of iterations used to estimate the Kronecker-Toeplitz structured shrunk
+			covariance. Error bars represent the first and third quartiles. The
+			accuracy does not improve when using more than one iteration.
+			(\textbf{a}) Results for 1, 2, and 5 trials using only the first block in each
+			training fold for training. (\textbf{b}) Results
+			for 1, 2, and 5 trials using
+			all nine training blocks in the training folds.
+		}
+		\label{fig:iterations}
+	\end{figure}
+
+	\subsection{Classifier accuracy for limited training data}
+	It is of interest to keep the calibration time before BCI
+	operation as short as possible.
+	We mimic this problem by training the classifier with as few training epochs as possible.
+	We evaluate the performance of all classifiers for different levels of
+	available training data and apply the cross-validation procedure nine times (the number of blocks in the training fold) for all subjects, keeping the
+	corresponding number of blocks in the training folds and dropping the rest.
+	\autoref{fig:accuracy} and \autoref{ap:accuracy_ap} show each classifier's
+	accuracy relative to the data availability.
+	We statistically compare the two newly proposed classifiers,
+	\textsc{stbf-struct} and \textsc{stbf-shrunk}, for different levels of training
+	data availability using a one-sided paired Wilcoxon rank-sum test with Holm correction for the multiple pairwise comparisons between classifiers.
+	We performed this analysis three times: by only using the first trial of a
+	block, by averaging epochs across the first two trials of a block and across
+	the first five trials of a block.
+	Results validated on one trial are reported in \autoref{tab:p_values_1}, 2-trial results
+	in \autoref{tab:p_values_2}, and 5-trial results in \autoref{tab:p_values_5}.
+
+	\begin{figure}
+		%\includegraphics[width=\linewidth]{figures/accuracy.eps}
+		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
+			number of blocks available for training. One block consists of 135
+			epochs and corresponds to 27 seconds of stimulation. Accuracies
+			are shown for the evaluation settings averaging over 1, 2, and
+			3 trials of testing stimuli.
+			\autoref{ap:accuracy_ap} contains results for all numbers of trials.
+			While \textsc{stbf-emp} is unstable when little training data
+			are available,
+			regularization of the covariance matrix (\textsc{stbf-shrunk} and
+			\textsc{stbf-struct}) drastically improves performance.}
+		\label{fig:accuracy}
+	\end{figure}
+	\begin{table}[h]
+		\centering
+		\input{part2/stbf_struct_tests_1.tex}
+		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
+			Holm correction using one testing trial for different classifiers
+			and levels of data availability. $p$-values $<0.05$ are considered significant
+			and marked in grey.}
+		\label{tab:p_values_1}
+
+		\centering
+		\input{part2/stbf_struct_tests_2.tex}
+		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
+		\label{tab:p_values_2}
+
+		\centering
+		\input{part2/stbf_struct_tests_3.tex}
+		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
+		\label{tab:p_values_5}
+	\end{table}
+	\unskip
+
+	The tables show that \textsc{stbf-struct} has a significant advantage over
+	\textsc{stbf-shrunk} when the number of training blocks is low.
+	This effect is present for 1-, 2- and 5-trial evaluation.
+	This advantage decreases (the $p$-value increases) when
+	adding more training blocks.
+	Both \textsc{stbf-struct} and \textsc{stbf-shrunk} perform significantly better
+	than \textsc{stbf-emp} for all evaluated settings.
+	Compared to x\textsc{dawn+rg}, \textsc{stbf-struct} also has significantly
+	higher accuracy in almost all evaluated settings, except when using only one training block.
+	\textsc{stbf-shrunk} does not outperform x\textsc{dawn+rg} when training data is low but gains a significant advantage
+	when using more training data.
+
+	\subsection{Classifier training time}
+	In order to evaluate the training time of the investigated classifiers, the
+	cross-validation scheme is run four times for each subject, each time with an
+	increasing number of EEG channels retained in the analysis, to explore the scalability of each classifier for analyses with higher spatial resolutions.
+	The temporal resolutions were not varied, but we expect that increasing the
+	temporal resolution has a similar effect on training time since the
+	training times for the \textsc{stbf}-based as evidenced by the complexity
+	calculations in \autoref{sec:parameters_complexity}.
+	\autoref{fig:training_time} shows the measured training times.
+	These results were obtained using a laptop with an Intel® Core™ i7-8750H CPU and 16GB of RAM.
+
+	\begin{figure}
+		%\includegraphics[width=\linewidth]{figures/training_time.eps}
+		\caption{Median fold training time for different classifiers at different spatial
+			resolution levels evaluated over all training folds for all subjects.
+			Error bars represent standard deviation. The training time of \textsc{stbf-shrunk} increases more
+			steeply with resolution compared to \textsc{stbf-struct}.
+			All \textsc{stbf} classifiers can be trained significantly faster than
+			x\textsc{dawn+rg}.}
+		\label{fig:training_time}
+	\end{figure}
+
+	\autoref{fig:training_time} shows that the training time of
+	\textsc{stbf-struct} increases less steeply than that of \textsc{stbf-shrunk}
+	and \textsc{stbf-emp}. The training time of all three STBF-based classifiers is
+	much lower than that of x\textsc{dawn+rg}, which appears nearly constant when using
+	4, 8, 16, or 32 channels.
+
+	\section{Discussion}
+
+	\subsection{Classification accuracy}
+	As evidenced by \autoref{fig:accuracy} and Tables \ref{tab:p_values_1}-\ref{tab:p_values_5}, the regularized classifiers \textsc{stbf-shrunk}
+	and \textsc{stbf-struct} significantly improve the classification accuracy
+	compared to the original \textsc{stbf-emp} for all numbers of training blocks
+	indicated.
+	We believe there are three reasons for this.
+	First and foremost, the empirical covariance matrix in \textsc{stbf-emp} becomes
+	ill-conditioned when the number of available training epochs is smaller than
+	the number of features ($n<cs$), rendering its inversion with the
+	Moore-Penrose pseudoinverse unstable.
+	This is the case \textsc{stbf-emp} when $n=cs=32*17=544$, after which the
+	accuracy of \textsc{stbf-emp} starts to increase.
+	This effect is visible in \autoref{fig:accuracy}, where the accuracy starts
+	increasing when using more than four training blocks, amounting to 540 epochs.
+	The noticeable dip in accuracy when using around 540 epochs can be explained by
+	numerical effects in the pseudoinverse for very small
+	eigenvalues~\cite{Blankertz2011, Raudys1998, Schaefer2004,
+		Kraemer2009}.
+	Regularization of the covariance matrix with shrinkage ensures that the
+	covariance matrix is non-singular and better conditioned so it can stably be inverted.
+	Second, covariance regularization introduces a trade-off between variance and bias of the model~\cite{Ledoit2004}.
+	Better performance on unseen data can be achieved when some model variance is
+	traded for extra bias.
+	Regularization reduces extreme values present, as shown in
+	\autoref{fig:kronlda-covs}, resulting in a classifier with
+	better generalization.
+	Third, the true spatiotemporal covariance matrix may vary throughout BCI
+	sessions, e.g., due to movement of the EEG-cap, changing impedances of
+	electrodes, subject fatigue, the introduction of new spatiotemporal noise
+	sources, and other possible confounds.
+	A regularized covariance matrix should better account for changes in true covariance.
+	Note that the LOOCV method in principle assumes that the covariances of
+	the training data and unseen data are the same.
+	Because the covariance might have changed for unseen data, the shrinkage
+	estimate obtained with LOOCV is probably still an
+	underestimation of the optimal -- but unknown -- shrinkage coefficient that
+	would yield the best classification accuracy for the unseen data.
+
+	Another observation is the significantly better accuracy score of
+	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available training data is small.
+	This property is an attractive advantage in a BCI setting since it is desirable to keep the calibration (training) phase as short as possible without losing accuracy.
+	The accuracy advantage of the structured estimator is a consequence of the
+	Kronecker-Toeplitz covariance structure, which is informative for the
+	underlying process generating the epochs, if it is assumed that the EEG signal
+	is a linear combination of stationary activity generated by random dipoles in
+	the brain with added noise~\cite{Munck1992, DeMunck2002, GonzalezNavarro2017}.
+	Hence, \textsc{stbf-struct} can utilize this prior information to better estimate the inverse
+	covariance.
+	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \autoref{sec:parameters_complexity}), increasing the stability of matrix inversions.
+
+	When compared to the state-of-the-art x\textsc{dawn+rg} classifier, we conclude
+	that \textsc{stbf-struct} reaches similar accuracy when using only one block of
+	training data.
+	The authors suspect this is due to both
+	classifiers having insufficient training information to reach
+	satisfactory classification accuracy.
+	When more data are available, \textsc{stbf-struct} reaches significantly better accuracies.
+	Combined with the benefits laid out in  \autoref{sec:parameters_complexity} and
+	\autoref{sec:interpretability}, this makes it an attractive option for ERP classification.
+	\textsc{stbf-shrunk} does not show decisive accuracy improvements over
+	x\textsc{dawn+rg} using a few training blocks, but this improves as the training data increases.
+
+	\subsection{Time and memory complexity}
+	\label{sec:parameters_complexity}
+	As mentioned above, inverting the full $cs \times cs$ dimensional covariance
+	matrix to construct \textsc{stbf-emp} and \textsc{stbf-shrunk} can be costly
+	and unstable, in particular in high-resolution settings with many EEG channels or time samples.
+	Constructing the full covariance and inverse covariance matrices also requires a considerable amount of memory.
+	The structured covariance estimator of \textsc{stbf-struct} has two advantages here.
+
+	First, because of \autoref{prop:inverse_kronecker} and
+	\autoref{prop:kronecker_multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
+	symmetric covariance and inverse covariance matrices for \textsc{stbf-struct}; they can instead be replaced by two smaller symmetric matrices respectively of dimensions $c\times c$ and $s\times s$.
+	Furthermore, since the temporal component of the Kronecker product is
+	Toeplitz-structured, it only requires $s$ parameters to
+	estimate.
+	While the inverse covariance of \textsc{stbf-emp} and \textsc{stbf-shrunk} is
+	defined by $\frac{cs(cs+1)}{2}=\frac{32\cdot17(32\cdot17+1)}{2}=\num{122128}$
+	parameters accounting for the symmetric nature of
+	covariance, the structured estimator only requires $\frac{c(c+1)}{2} + s =
+		\frac{32(32+1)}{2} + 17=545$ unique parameters.
+	This reduction in parameters to estimate reduces memory usage and contributes to the regularization effect for low data availability settings.
+	The inverse covariances of \textsc{stbf-emp} and
+	\textsc{stbf-struct}, represented as $32*17\times 32*17$ symmetric matrices of
+	single-precision real floating point numbers for weight calculation,
+	use 9.03MiB of memory.
+	The $32\times 32$ and $17\times 17$ matrices of \textsc{stbf-struct} only
+	require 5.12KiB.
+
+	Second, structured estimation has better time complexity.
+	Covariance estimation and inversion occupy the largest part of the STBF training time.
+	For \textsc{stbf-emp} and \textsc{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(nc^2s^2+c^3s^3)$.
+	Thanks to Property~\ref{prop:inverse_kronecker}, the complexity can be reduced to
+	$\mathcal{O}(nc^2s^2+c^3+s^3)$ for the structured estimator of \textsc{stbf-struct}.
+	The results presented in \autoref{fig:training_time} confirm these
+	calculations.
+	It can be observed that the training time of \textsc{stbf-struct} stays low compared to \textsc{stbf-emp} and \textsc{stbf-shrunk} when dimensionality increases.
+
+	The results shown in \autoref{fig:training_time} also confirm that the
+	STBF-based estimators are very fast to train compared to the
+	state-of-the-art estimator x\textsc{dawn+rg}, which confirms the results in~\cite{Wittevrongel2016}.
+	Since the training times of all STBF-based classifiers are already in
+	the order of tenths of seconds, the question arises whether the
+	improvements achieved by using the structured estimator would be relevant in
+	practice.
+	However, the authors believe that these results could significantly impact some
+	use cases of the spatiotemporal beamformer, like high spatial or temporal resolution ERP analyses.
+	One example is single-trial ERP analysis with a high-temporal
+	resolution to extract ERP time features.
+	Such higher-resolution analyses can later be incorporated into an ERP
+	classification framework.
+	In addition, the speed-up provided by structured estimation yields a faster
+	off-line evaluation of the STBF ERP classifier, where often multiple cross-validation folds, subjects, and hyperparameter settings need to be explored, which can quickly increase runtime.
+	Improvements in computation speed and memory usage can remove the need for dedicated computation hardware and enable running group analyses on a personal computer.
+
+	%	\subsection{Interpreting the weights}
+	%	\label{sec:interpretability}
+	%	The weight matrix of the STBF determines how each
+	%	spatiotemporal feature of a given epoch should contribute to enhancing the SNR
+	%	of the discriminating signal in the classification
+	%	task.
+	%	Alternatively, the activation pattern can be regarded as a forward EEG model of
+	%	the activity generating the discriminating signal and the weights as a
+	%	backward model~\cite{Blankertz2011,Haufe2014}.
+	%	Regularization enables a researcher to interpret better the distribution of the weight over space and time after reshaping the weight vector $\mathbf{w}$ to its spatiotemporal matrix equivalent $W$ such that $\text{vec}(W) = \mathbf{w}$.
+	%	\autoref{fig:interpret_weights} compares the weights calculated in
+	%	\textsc{stbf-emp} and \textsc{stbf-shrunk} with the weights from
+	%	\textsc{stbf-struct}.
+	%
+	%	\begin{figure}
+	%		%\includegraphics[width=\linewidth]{figures/weights.eps}
+	%		\caption{Spatiotemporal beamformer weights calculated using four
+	%			blocks of data (of 1215 epochs) from \textit{Subject 01} from
+	%			0.2s before to 1.0s after stimulus onset.
+	%			Regularized weights show an interpretable sparse pattern,
+	%			while the empirical weights appear noisier.
+	%			(\textbf{a}) Spatiotemporal activation pattern with spatial and temporal global field
+	%			power.
+	%			(\textbf{b}) \textsc{stbf-struct} weights with spatial and temporal average of
+	%			absolute values.
+	%			(\textbf{c}) \textsc{stbf-shrunk} weights. The shrinkage factor $\alpha=0.05$ was
+	%			determined with the closed-form LOOCV-method.
+	%			(\textbf{d}) \textsc{stbf-emp} weights.}
+	%		\label{fig:interpret_weights}
+	%	\end{figure}
+	%
+	%	Since the linear filter's noise suppression and signal amplification functions are deeply entangled, it is not necessarily true that features with a
+	%	high filter weight directly correlate to features containing discriminatory
+	%	information~\cite{Haufe2014}.
+	%	However, it still is possible to interpret the weights in terms of which
+	%	features contribute most to the classification process, be it through noise
+	%	suppression, signal amplification, or -- most likely -- a combination of both.
+	%	The weights obtained by \textsc{stbf-emp} look randomly distributed over space
+	%	and time; the regularized estimator used by \textsc{stbf-shrunk} and
+	%	\textsc{stbf-struct} reveal a more interpretable weight distribution.
+	%	The \textsc{stbf-shrunk} weights show a sparse spatial distribution while
+	%	the \textsc{stbf-struct} weights show a sparse distribution in both space and in
+	%	time.
+	%
+	%	As expected, \autoref{fig:interpret_weights}b and d exhibit weight around
+	%	the central and parietal regions, where the P300 ERP component is present.
+	%	Especially the spatial weights of \textsc{stbf-shrunk} in
+	%	\autoref{fig:interpret_weights}d correspond to the spatial activation pattern
+	%	in \autoref{fig:interpret_weights}a.
+	%	This is not unsurprising, since shrinkage transforms the covariance matrix closer to the
+	%	identity matrix and assuming identity covariance in \autoref{eq:closed_form} yields weights
+	%	identical to the activation pattern (up to a scaling factor).
+	%	Additionaly, \autoref{fig:interpret_weights}b shows that weights in the baseline interval and after 0.6s, which should
+	%	contain no response information, are close to zero for the structured estimator.
+	%	Meanwhile, these weights are high in the occipital region between 0.1s and 0.2s,
+	%	containing early visual components with relatively low SNR.
+	%	This high weight for the early visual components confirms the results from Treder \& Blankertz~\cite{Treder2010}
+	%	that state that, in addition to the P300, the early N1 and P2 ERP components
+	%	are also modulated by oddball attention and contain discriminatory information between
+	%	attended and non-attended stimuli.
+	%
+	%	Using an interpretable classification model has many advantages.
+	%	For instance, one can use the weight matrix determine relevant time samples
+	%	and EEG channels for per-subject feature selection to refine the model further .
+	%	The number of channels is also an important cost factor in practical BCI
+	%	applications.
+	%	Determining which channels do not contribute to the classification accuracy
+	%	helps reduce the number of required electrodes.
+	%	Spatially clustered weights indicate that some electrodes are not used by the
+	%	classifier and can be discarded accordingly with no substantial accuracy
+	%	reduction.
+	%	As another example, information about the timing and spatial distribution of
+	%	the discriminatory information in the response can be extracted from the
+	%	weights
+	%	and linked neurophysiological hypotheses.
+
+	\todo{Remove mentions of interpretability}
+
+	\section{Conclusion}
+	While it is possible to regularize the spatiotemporal LCMV beamformer
+	classifier for ERP detection with other methods such as by employing feature selection,
+	by adding regularizing penalties to the cost
+	function beamforming problem, or by crafting a cleaner activation pattern, our work focused on
+	estimation methods for the spatiotemporal covariance.
+	We introduced a covariance estimator using adaptive shrinkage and an
+	estimator exploiting prior knowledge about the spatiotemporal nature of the EEG
+	signal.
+	We compared these estimators with the original spatiotemporal
+	beamformer and a state-of-the-art method in an off-line P300 detection task.
+	Our results show that the structured estimator performs better when training data are sparsely available and that it can be computed faster and with substantially less memory usage.
+	Since these algorithms are not paradigm-specific,  the conclusions can be generalized to
+	other ERP-based BCI settings.
+
+	Future work should focus on introducing more robust regularization strategies using prior knowledge, such as shrinking the spatial covariance to a population mean or a priorly known matrix based on sensor geometry or characterizing the temporal covariance as a wavelet or autoregressive model.
+	More accurate results could be obtained by expressing the covariance as the sum of multiple Kronecker products to account for spatial variation in temporal
+	covariance.
+	It could also be interesting to explore the impact of covariance regularization on transfer learning of the STBF between subjects to alleviate calibration entirely.
+	Finally, it could be insightful to evaluate the proposed algorithms in a
+	real-world on-line BCI setting.
 
@@
 	\section{Decoding performance}
-\section{Discussion}
 	\subsection{Extension into the space-time-frequency domain}
+	\printbibliography
+\end{refsection}

### 2024-08-24 chapters/stbf_struct.tex
Add figures

--- 2024-08-21 part2/stbf_struct.tex
@@
 
-	\begin{figure}[p]
-		\includegraphics[width=\linewidth]{part2/figures/stbf_struct/covs.eps}
+	\begin{figure}
+		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
 		\caption{Different estimators of the covariance and inverse covariance
@@
   \begin{figure}
-		%\includegraphics[width=\linewidth]{figures/fpi.eps}
+		\includegraphics[width=\linewidth]{figures/stbf_struct/fpi.eps}
 		\caption{Average cross-validated \textsc{stbf-struct} Accuracy using
@@
   \begin{figure}
-		%\includegraphics[width=\linewidth]{figures/accuracy.eps}
+		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
 		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
@@
 		\centering
-		\input{part2/stbf_struct_tests_1.tex}
+		\input{chapters/stbf_struct_tests_1.tex}
 		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
@@
 		\centering
-		\input{part2/stbf_struct_tests_2.tex}
+		\input{chapters/stbf_struct_tests_2.tex}
 		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
@@
 		\centering
-		\input{part2/stbf_struct_tests_3.tex}
+		\input{chapters/stbf_struct_tests_3.tex}
 		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
@@
 	\begin{figure}
-		%\includegraphics[width=\linewidth]{figures/training_time.eps}
+		\includegraphics[width=\linewidth]{figures/stbf_struct/training_time.eps}
 		\caption{Median fold training time for different classifiers at different spatial
@@
 
+  \todo{Include part kronlda, check notes in journal}
+\section{Introduction}
+
+LCMV struct might be to restrictive
+Toeplitzlda
+MEG paper
+
+\section{Methods}
+\subsection{Spatiotemporal Linear Discriminant Analysis}
+\subsection{The single Kronecker product covariance model}
+\begin{equation}
+	\mat{C} = \mat{S}\otimes\mat{T}
+\end{equation}
+Tapering, Toeplitz and shrinkage
+\subsection{The Kronecker sum covariance model}
+\begin{equation}
+	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
+\end{equation}
+pvl, kronpca
+
+\subsection{Kronecker sum covariance estimation}
+\subsection{Classification}
+\section{Results}
+\section{Discussion}
+
+\todo{check chapters/structure paper}
+\todo{check structure Leuven.ai poster}
+\section{The spatiotemporal EEG covariance}
+\section{The single-KP covariance model}
+\section{The sum-of-KP covariance model}
+\section{Decoding performance}
+\section{Extension into the space-time-frequency domain}
+
 	\todo{check chapters/structure paper}

### 2024-08-24 layout/format_booklet.tex
Add figures

\usepackage[paperheight=240mm,paperwidth=160mm]{geometry}

\usepackage{float}
\floatplacement{figure}{p}
\floatplacement{table}{p}

### 2024-08-24 chapters/bci.tex
Add wcble paper

\todo{chapters don't nicely follow imrad structure. Is this  problem?}
\todo{concern: is a bit in between formula 1 and formula 2}
\section{Rationale}
\section{Overview \& state-of-the-art}
different recording technologies and level of invasiveness + infographic
active/reactive/passive + infographic
\section{Visual Event-Related Potential based BCIs}

### 2024-08-24 chapters/conclusion.tex
Add wcble paper

\chapter{Conclusions \& recommendations}
\section{General discussion}
\section{Limitations}
\section{Future directions}
\section{Working with patients}
\section{Conclusion}

\section{Obtained results}

\subsection{Regularized spatiotemporal beamforming}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (xDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

We have tested our proposed methods and similar methods in gaze-independent and
covert VSA settings, but we did not observe a relative increase in performance
in these settings.

%* Preliminary results: BTTDA better than HODA, on par with other state of the
%art methods and in some situations better (data availability, noise structure)

\subsection{Healthy control study}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect observed for covert VSA also holds for split VSA.

We have evaluated our proposed wCBLE algorithm, the baseline CBLE algorithm and
a state-of-the-art algorithm (tLDA)~\cite{Sosulski2022} that uses similar techniques as our
proposed STBF-struct method on this dataset as well as a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the VSA condition and the dataset.
While CBLE performed overall on par with the state-of-the art method for both datasets, wCBLE was
outperformed by tLDA in overt VSA decoding but outperformed
tLDA for covert VSA.
For the split attention conditions in the CVSA-ERP
dataset, wCBLE yielded a significant improvement over CBLE and the
state-of-the-art method only for d = 1 and d = 3.
These results were corroborated by analyzing selection accuracy and information
transfer rate, which showed similar behavior for both datasets, except in overt
VSA, where accuracy and information transfer rate were not harmed by the lower
single-trial selection performance of wCBLE.

To further study the gaze-independent performance of these algorithms, transfer
learning between VSA conditions is studied to simulate conditions where a patient
can end up in different VSA settings within a BCI operation session due to their
lack of eye motor control.
When trained and evaluated on overt VSA data, our proposed wCBLE algorithm results in a significant decrease in performance
over the state-of-the art tLDA decoder, consistent with the within VSA
condition results.
For all other pairs of training and evaluation VSA condition, however, wCBLE
never yields a significant decrease in performance and often significantly
outperforms the other methods.

These results show that there is an interest in developing a new class of ERP-BCI
interfaces for patients that prefer to rest their eyes on a chosen point on the
screen, so as to avoid the effort of redirecting their eye gaze at different spatial locations on
the stimulation screen. Furthermore, the performance gain in the split VSA conditions
and in the between-VSA condition transfer settings are promising for patients with even
less eye motor control, such as patients experiencing involuntary saccades or
fatigue. An accurate decoder would allow them to comfortably operate a BCI while
resting their eyes on whichever portion of the screen they prefer, even when there is
another target present at that location or this location varies during the course of the
experiment. Our results show that whenever the eye gaze is directed at a different
spatial location other than that of the mentally attended target, the wCBLE decoder can
more accurately discern the locus of mental attention, promoting in this way accurate
gaze-independent decoding. This comes, however, at the cost of a decreased performance
compared to overt VSA operation.

### 2024-08-24 chapters/problem_statement.tex
Add wcble paper

\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}

\todo{Check provisionary and final doctoral plan for motivation}
\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities, by leveraging a non-invasive high-ITR
visuospatial ERP-based BCI and improving ERP decoding performance (in general
and specifically in gaze-independent settings).}
\todo{check WIP seminar}

\todo{update text}

\emph{Novelty:}

Brain-Computer Interfaces (BCIs) decode brain activity with the aim to establish a direct communication
pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
devoid of these abilities, for whom BCIs can provide a means to communicate or to
control devices.
The quest for performant and affordable solutions is most evident in
visual BCIs based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
and paradigms tailored to the needs of specific stimulation paradigms and
use contexts.
An effective and proven method is the visual event-related potential (ERP)-based
oddball interface.
Here, targets are shown in short pulses on a computer screen.
Each time the user observes a target, an ERP is evoked, the presence of which can be
detected in the EEG-signal.
The ERP consists of multiple components, of which some are modulated by the
attention of the user, specifically the N2 and the P3 component.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci}.
However, visual BCIs cannot operate efficiently when the user does not direct
their gaze onto the
desired target~\cite{Brunner2010}.

This leads to an often unadressed paradox: the BCI target population consists
of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
(ALS), Multiple Sclerosis (MS), stroke (specifically
brain-stem stroke) or traumatic brain injury, which can lead to several degrees
of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While BCIs are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking, studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from involuntary eye movements to control
issues and even to partial or complete ophtalmoplegia.
\begin{table}
	\centering
	\begin{tabular}{@{}|l|rrrrr|r|@{}}
		\hline
		         & ALS  & MS   & Stroke  &
		DMD      & SMA  & LiS                           \\ \hline
		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		Complete & 17\% & -    & +       & - & - & 2\%  \\
		\hline
	\end{tabular}
	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome).}\label{tab:incidence}
\end{table}
\todo{document methodology}
\todo{find more references, fill out}
Different degrees of eye motor impairment can render visual BCIs
uncomfortable or outright impossible to use.
When operating a visual BCI, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the BCI.
This points to the need for gaze-independent BCIs.

\section{Objective}

The general goal of this PhD is to tackle the gaze-dependency in visual BCIs.
Traditionally, gaze-independency~\cite{Riccio2012, Aloise2012} can be realized in three ways:
Firstly, by avoiding visual
stimulation entirely and opting for auditory or somatosensory stimulation instead.
However, these alternatives often result in lower information transfer rates,
increased mental effort and variable outcomes for different users.
Secondly, gaze-independence can
be achieved by adapting the interface to display stimuli such that they
are always in the center of the field of view, by exploiting visual
non-spatial attention (feature attention), or a combination thereof.
These interfaces suffer equally from reduced information transfer rates.
Moreover, they still rely to some extent on eye motor control, necessitating central gaze fixation.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like Rapid Serial Visual Presentation,
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
Third, stimuli can be presented in a standard BCI paradigm, but \emph{mental} attention
can be decoded separately from \emph{visual} attention.
\todo{table with paradigms and itrs from WIP seminar}

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball ERP-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\begin{quote}
	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
		can achieve similar or higher information transfer rate in patient
		populations with eye motor deficits than other
		gaze-independent alternatives described in literature.}
\end{quote}

To achieve our goal, we innovate on decoder development and interface design,
and we collect data from healthy control subjects, which allow us to test our decoders.
Finally, the findings of these experiments will be leveraged to develop a BCI
for patient use, which will be tested in a case-based patient study in which
both performance and user comfort will be evaluated.

\section{Methods}
\todo{update with new decoders and experiments}

\subsection{Interface design}
We designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
decoders and on our own proposed decoders.
In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent ERP decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention ERP dataset
(CVSA-ERP)
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).

Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG and electro-oculogram (EOG) are
recorded, as well as eye gaze using eye tracking.
To simulate the dissociation between the eye gaze (visual attention) and the
intended target (mental attention) which could occur in patients with hampered eye
motor control, healthy participants are cued to operate the BCI in specific
visuospatial attention (VSA) conditions.
Each subject performs 5 different VSA conditions as illustrated in
Figure~\ref{fig:vsa-conditions}.
\begin{figure}
	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
		asked to dissociate their visual and mental attention by gazing at a
		crosshair while attending a target, depending on the prompted condition.}%
	\label{fig:vsa-conditions}
\end{figure}
\todo{new picture that shows more clearly a user operating freely where they
can gaze}
In the overt case, subjects are instructed to gaze at the cued
target they are also mentally attending; in the covert case, subjects are instructed
to gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent BCI development: split VSA.
In split VSA, the participant mentally focuses on one cued target while gazing at
another (the distractor).
Depending on the inter-target distance along the hull of the hexagon between the
attended target and the distractor we discern three split VSA subconditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

After data collection from healthy participants and decoder development, we
will carry out a patient study using a similar interface.
The goal of the patient study is to evaluate the influence of eye motor
deficits in patient populations on existing state-of-the-art decoders and on
our proposed decoding methods.
\todo{update with current results}

\subsection{Decoders}

During this PhD, we explored different lines in decoding strategies,
trying to tackle several problems that arise from gaze-independence, such as
the lack or decrease in amplitude of specific ERP components, and the increased
non-stationarity of the signal.
As mentioned, state-of-the-art decoders have poor performance in covert attention
settings.
The general goal is thus to design a machine learning classifier that represents
the ERP signal in a specific way, such that it becomes more robust to the problems
occurring in covert attention conditions.

\subsubsection{Regularized spatiotemporal beamforming}
Due to the lack of the N2 component and decrease of amplitude of the
P3 component in covert attention
settings~\cite{Treder2010}, the signal-to-noise ratio (SNR) of
the ERP is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance, might be by increasing overall ERP
decoding performance. A more accurate classifier could yield relative
improvements in those settings where performance is not yet near the achievable
maximum.

Therefore, we have improved upon an in-house developed, state-of-the art ERP
decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(STBF-struct).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, which is of use in our next efforts to cope with the
non-stationarity of the P3 signal component.

\subsubsection{Classifier-based Latency Estimation with Woody iterations}
Literature shows that better covert attention performance can be achieved by
counteracting the non-stationarity of the P3 ERP component in covert
attention settings~\cite{Arico2014}.
This can be done by estimating the latency of each single-trial ERP and aligning
the P3 peaks to all fall at the same moment.
The resulting aligned data and the set of latencies can then be used to train a
classifier that is more robust to jitter.

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low SNR of the ERP.
Classifier-based Latency Estimation (CBLE)~\cite{Mowla2017} is a technique that can leverage
ERP latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic, iterative
method named Classifier-based Latency Estimation with Woody iterations (wCBLE).

\todo{update with current results and all chapters}

### 2024-08-24 chapters/covert_align.tex
Add wcble paper

\section{Data collection \& preprocessing}
\section{Results}
\section{Discussion}
\todo{check chapters paper}

\subsubsection{Riemannian Geometry}
We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L2$-regularized logistic regression (ERPCov-TS-LR).

\subsection{Experiments}
We evaluate our approach on a
publicly available dataset and one specifically recorded for this study, designed to
probe different modalities of covert VSA.

\subsubsection{CVSA-ERP dataset}
We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
\ref{app:dataset} details the stimulation and recording procedure.
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:erps}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
	\end{subfigure}%

	\caption{
		\subref{fig:interface} Interfaces and visuospatial attention (VSA)
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball BCI interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split VSA to the cued target.
		In the BNCI2014-009 overt VSA interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per VSA condition and dataset.
    Overt VSA yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split VSA.
    In split VSA, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

In contrast to the protocol proposed by \cite{Frenzel2011}, split VSA was
performed by instructing the participant to attend the intensifications of the
cued target, and ignore the intensifications of the distractor target.
Since we assume there will be an effect depending on the distance between
attended target and the distractor, we discern three split VSA sub-conditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite to the intended target
($d=3$).

\subsubsection{BNCI2014-009 dataset}
The
BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
dataset~\cite{Aloise2012a} was used in the analysis performed
in \cite{Arico2014}.
It contains data from 10 subjects (median age $24.5\pm1.9$
years)  that performed two spelling tasks illustrated in the second row of
Figure~\ref{fig:interface}: using the P3 Matrix speller
interface to exploit overt VSA, and the GeoSpell covert VSA interface.
To use the GeoSpell interface, the participant gazes at the fixation point at the
center of the screen, while groups of characters flash simultaneously in a
circular layout around the fixation point.
The user directs their visuospatial attention to the location where the intended letter is expected
to appear, and when it does, a P3 ERP component is expected to be evoked.
This results in a specific setting where both visuospatial attention and
feature attention (the attended letter) are exploited.
For a detailed description of the paradigm and dataset, we refer
to \cite{Aloise2012a}.

\subsubsection{Data preprocessing and analysis}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

To evaluate performance, 6-fold cross-validation without shuffling was performed for both
datasets.
At each fold, classifiers were trained on five target selection blocks (300
epochs) and tested on one block (60 epochs) without overlap for CVSA-ERP.
For each subject and run in the BNCI2014-009 dataset, classifiers
were trained on five symbol selections (480 epochs) and tested on one symbol
selection (96 epochs) without overlap at each fold.
A window ranging from 0ms to 600ms after stimulus onset was used for CBLE and WCBLE.
With epochs ranging from -100ms to 700ms relative to stimulus onset, this
allows for extracting latencies ranging from -100ms to +100ms.
\section{Results}

\subsection{BCI decoding performance}
\label{sec:block_accuracy}

We evaluated the BCI decoding performance in a single-trial classification
experiment, as well as in a target selection experiment reflecting BCI
operation.

Figure~\ref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
Receiver Operating Characteristic curve (ROC-AUC) for all pairs of tLDA, CBLE
and WCBLE for single-trial classification to investigate the contributions of
CBLE and WCBLE relative to their first-stage classifier tLDA.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$ and, for the CVSA-ERP dataset, if the user's gaze differed more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:single_trial_roc_auc_diff}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

	\bigskip

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:block_evaluation}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
	\caption{
		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
    single-trial classification receiver operating characteristic curve ($\Delta$ROC-AUC)
    between Classifier-based Latency Estimation (CBLE), Woody
    CBLE (WCBLE),	and their first-stage classifier (tLDA).
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed WCBLE decoder outperforms tLDA and CBLE for covert and split
		($d=3$)	visuospatial attention (VSA). CBLE scores on par with tLDA.
		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
    all decoders plotted n function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves highest
    covert VSA target selection accuracy.}
\end{figure}
Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
and Hochberg's False Discovery Rate procedure (FDR) revealed that
for the BNCI2014-009 dataset, WCBLE significantly outperformed tLDA
($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and CBLE
($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert VSA
but was significantly outperformed by tLDA in overt VSA decoding
($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
For the CVSA-ERP dataset, WCBLE
also achieved significantly better covert VSA performance than tLDA ($\Delta\mathrm{ROC-AUC}
	= 0.023$, $p=0.041$) and CBLE ($\Delta\mathrm{ROC-AUC}
	= 0.036$, $p=0.024$) .
We found no significant difference in WCBLE performance over tLDA in the split
VSA conditions in the CVSA-ERP dataset, but results show a clear trend of
increase in WCBLE performance over tLDA and CBLE as $d$ increases
CBLE failed to significantly outperform its first-stage classifier tLDA in all
evaluated VSA conditions.
Table~\ref{tab:single_trial_full} reports all single-trial classification scores for all considered models, datasets
and conditions.

Figure~\ref{fig:block_evaluation} shows the cross-validated BCI selection
accuracy on the BNCI2014-009 and the CVSA-ERP dataset for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure of BCI
operation intact.
For all datasets and VSA conditions, CBLE scores approximately on par with tLDA.
Yet, WCBLE yields an improved decoding accuracy for covert VSA in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert VSA accuracy increase over tLDA is significant in the BNCI2014-009 dataset
for 1 and 3 repetitions and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial ROC-AUC for
WCBLE in overt VSA, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of WCBLE over other methods was found in split VSA.
While Riemannian methods are significantly outperformed by tLDA, CBLE and WCBLE
in the BNCI2014-009 dataset, they perform approximately on par with tLDA and
WCBLE in CVSA-ERP.

Overall, we observed a 5.10\%pt. accuracy increase with WCBLE over
tLDA for covert VSA in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
These results compare to the performance gain in \cite{Zisk2022}.
They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
step-wise Linear Discriminant Analysis  (SWLDA) for 6 ALS patients, whose SWLDA
performance also suffered from jitter.
Note that interpretation of this comparison may be challenging due to differences in
interface design (number of targets, inter-stimulus interval), subject population,
EEG recording procedure and available training data.

\subsection{Gaze-independence through cross-condition transfer}
\label{sec:cross_results}
To further back our claim of gaze-independence in the case where eye motor
control cannot be assumed, we evaluate our proposed decoder in a transfer
learning setting between VSA conditions.
While performing covert attention requires gaze redirection for each target
selection, performing covert or split VSA continuously still requires
sustained gaze fixation, which might not be possible for some patients that
could benefit from such an application.
Studying the transfer between conditions simulates what happens
when the user performs different VSA conditions throughout the
experimental session.
Furthermore, if our decoder performs well in transfer-learning settings,
it must capture some information about the ERP responses that is independent of
the VSA condition, and hence does not depend on gaze redirection to perform
these conditions.
We introduce an additional setting of interest here, namely on a combination of VSA
conditions, which represents those cases where patients cannot
redirect their gaze and hence can be in any one of the VSA conditions depending on the
target they attend.
For BNCI2014-009, this is implemented as an equal mix of overt and
covert VSA, for CVSA-ERP the combined condition represents an equal mix of overt,
covert and split VSA, disregarding parameter $d$.

Figure~\ref{fig:covert_cross_eval} and Figure~\ref{fig:aloise2012_cross_eval} show
the pair-wise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in Section~\ref{sec:block_accuracy}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert_cross_eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip

	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:aloise2012_cross_eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
	\end{subfigure}\hfill%
	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:jitter}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
	\end{subfigure}

	\caption{
		\subref{fig:covert_cross_eval},\subref{fig:aloise2012_cross_eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt VSA settings, indicating it is more invariant to eye gaze than
    CBLE and tLDA.
		\subref{fig:jitter} Jitter characterized as the inter-quartile range (IQR)
		of target epochs for different VSA conditions. Overt VSA exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing CBLE and tLDA, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.
On the contrary, when considering the comparisons between WCBLE and tLDA, we
see that performance is on par or greater using WCBLE for most conditions,
except for within-overt VSA decoding.

\section{Discussion}

Figure~\ref{fig:single_trial_roc_auc_diff} shows that WCBLE significantly improves
covert VSA decoding.
This is advantageous for the development of a class of ERP-BCI interfaces for
patients who prefer to rest their gaze on a fixation cross on the screen,
avoiding the effort of redirecting their eye gaze for every selection.
Furthermore, the performance gain over the first-stage classifier in the split
VSA ($d=3$) and between-VSA condition transfer settings are promising for
patients with even less eye motor control that can experience involuntary eye
movements or fixation fatigue hence cannot keep their gaze fixed throughout an
entire BCI operation session.
WCBLE would allow them to operate a BCI comfortably while directing
their gaze to whichever portion of the screen they prefer, even when there is
another target present or this location varies during the
course of operation.
Although WCBLE did not significantly improve overt VSA single-trial decoding,
Figure~\ref{fig:block_evaluation} shows that this does not negatively impact target
selection accuracy.
While target selection accuracy also did not improve for split VSA, the
increase in single-trial performance in split ($d=3$) shows that an iterative
alignment procedure has the potential to improve over CBLE and its first-stage
classifier for this case as well.

%, thereby increasing estimated jitter.
We believe the relative increase in performance of our proposed decoder in covert and
split VSA, and the lack thereof in overt VSA, could stem from the following:
\begin{enumerate*}[label=(\arabic*)]
	\item Covert and split VSA exhibit higher P3 jitter than overt VSA.
        In covert and split VSA, participants have to execute a dual task by
        dissociating their visuospatial attention and gaze fixation.
        Evidence shows that ERP latency variability is higher when attention is
        divided~\cite{Polich2007,Arico2014}.
        \cite{Arico2014} also partly attribute higher latency jitter to the covert VSA
        task performed in the BNCI2014-009 dataset since the GeoSpell interface
        requires both spatial and feature attention.
  \item In overt VSA, the first-stage classifier can rely mostly on the modulation
        of early visually evoked potentials (VEPs) like N1 than on the
        P3~\cite{Treder2010}.
        These VEPs are closely related to visual processing hence exhibit
        lower jitter, contrary to P3 which is more prone to the effects of
        attention and workload~\cite{Hu2010}, reducing the contribution of
        alignment.
  \item This property can also result in the estimation of VEP latencies
        instead of the P3 latency and WCBLE would in this case fail to
        increase the P3 SNR, which still could be somewhat jittered in overt
        VSA.
  \item Aligning to the P3 will lower the SNR of early VEPs while aligning to
    VEPs will lower P3 SNR, since they are not time-locked to each other.
	\item Covert and split VSA ERPs may exhibit lower SNR than overt
        VSA due to lower P3 amplitudes or even due to the presence of higher P3
        jitter itself.
        Higher SNR in overt VSA results in higher decoding performance of
        state-of-the-art classifiers, leaving less room for relative improvement
        in this case.
\end{enumerate*}
%When the time window includes multiple ERP components, it is hard to make
%strong claims on component latency distributions using latencies estimated by
%CBLE or WCBLE.
%What is actually measured is the latency of the most discriminatory time shift
%of an epoch.

Although it is not immediately clear if WCBLE actively corrects for higher P3 jitter
present in covert and split VSA compared to overt VSA,
we justify our approach similar to \cite{hardiansyah2020single} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points towards higher P3 jitter as the main
contributing factor.
%Here, the subject using the GeoSpell interface must discriminate between target and
%non-target trials partly based on the stimulus characteristics
%(i.e. the selection of characters in their spelling application), compared to
%discrimination based on spatial location such as is the case in the CVSA-ERP
While Figure~\ref{fig:erps} shows no visible smearing effect in
the shape of the ERPs, a more quantitative analysis on latencies in
Figure~\ref{fig:jitter} indicates the opposite.
To obtain comparable results across VSA conditions, WCBLE was
evaluated per session and trained on all combined VSA conditions as in
Section~\ref{sec:cross_results}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the inter-quartile
range (IQR) of the estimated latencies of target trials, revealed that overt
VSA exhibited significantly lower jitter than all other conditions for both datasets.
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally \cite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert VSA by comparing performance between
overt and covert VSA, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert VSA decoding performance.
However, the large increase in performance for covert VSA in \cite{Arico2014} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.

%If most of the jittered discriminatory information stems from the P3 component,
%WCBLE as a decoder is less sensitive to VEPs and only enhances the P3
%component.
%This property can be exploited in gaze-independent decoding to determine
%visuospatial attention separately from the gaze direction.

%In fact, it is not obvious how the alignment employed by WCBLE
%should result in better performance for overt attention.
%At each iteration, aligning to the P3 latency reduces the contribution from
%early visual potentials such as the N1 component not, which are more time-locked to
%the stimulus presentation than to the P3.
%These early visual potentials, however, provide a large part of the
%discriminatory information in overt VSA, while this is less
%the case in covert attentionConversely, since the N1 component is more closely related to visual processing,
%and therefore exhibits a lower jitter, aligning to the N1 component might not augment the
%Another explanation for the lack of overt VSA improvement could be the choice
%of tLDA as first-stage classifier and comparison method.
%As a classifier, tLDA can outperform Riemannian Geometry
%methods~\cite{Sosulski2022}, which is
%most likely the cause of improvements of WCBLE over \cite{Aydarkhanov2020}.
%It might be harder to improve upon tLDA in overt VSA due to its high
%performance.

Finally, we argue that the increased performance of WCBLE is partly due to the
fact that
tLDA is a suitable first-stage classifier.
Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
Secondly, this method has a synergy with CBLE since both make the same
assumption about the short-time stationary of the EEG background noise
within an epoch.
CBLE does not retrain the first-stage classifier for each time shift, but
rather trains it once within the given window.
After training, the classifier parameters represent some information about the
expected ERP waveform and background noise.
By applying the trained classifier to different time shifts, it assumes this
ERP waveform can be shifted in time, but since the classifier's information about
the background noise was only obtained from the initial window, CBLE assumes
its properties do not vary throughout the epoch.
The block-Toeplitz covariance structure of tLDA also assumes that the
background noise represented by this covariance after subtracting the class
averages is stationary within the epoch~\cite{Sosulski2022}.

We found that CBLE did not improve gaze-independent decoding performance
significantly and also did not increase performance over its first-stage
classifier in overt VSA contrary to what was reported
by \cite{Mowla2017}.
While they report that CBLE is relatively independent to the
first-stage classifiers evaluated in their work, it is evident here that
applying any given classifier in the CBLE scheme does not necessarily increase
its performance.
In our case, this could be due to characteristics of the tested dataset, e.g.
the presence of jitter, or the generally higher performance of tLDA as compared to the
first-stage classifiers tested by \cite{Mowla2017}, leaving less
performance to be gained.

\cite{Thompson2012} already attempted applying CBLE in an iterative scheme
but did not report any results due to convergence issues.
%Just like the Woody scheme, this algorithm can get stuck in a local optimum or,
%in unfavorable SNR conditions, even amplify noise.
We mitigated this by combining the robust latency estimation presented in
~Section~\ref{sec:robust-latency} with the alignment of both target and non-target
epochs.
Aligning only the target epochs containing the jittered P3 component is prone to
overfitting by aligning non-discriminate noise that is present in both
classes, such as environment noise, oscillatory background rhythms or
non-modulated VEPs.
If SNR is low, residual noise varying slightly between classes could dominate
the expected response of the first-stage classifier and subsequently dominate
WCBLE from the start, preventing convergence to a meaningful solution.
Our procedure circumvents this problem by aligning both classes to the time
points where the expected separation between classes is greatest.
This way, noise of which the latencies are estimated in a given iteration will be
perfectly time-locked in all classes in the next
iteration.
The first-stage classifier can then more easily suppress this noise since it is
now clear it is present in both classes and non-discriminative.
This aids the method in converging to a more robust classifier by iteratively
increasing SNR for both classes and class separation over time.

\cite{Zisk2022} addressed P3 jitter in ALS patients by augmenting the training data once
with time-shifted copies based on CBLE-estimated jitter.
While we aim to train the first-stage classifier without the effects of jitter
in its parameters, they do the opposite by intentionally jittering the
training data.
As their focus was on ALS patients and inter-session stability, they
did not assess how their method interacts with visuospatial attention.
We achieved a similar performance gain with our jitter compensation method, but
argue that our method can cope with more granular latency differences, as
\cite{Zisk2022} augment the data with just one positive and negative time
shift.

While \cite{hardiansyah2020single} decoded covert VSA more effectively by
contributing single-trial latency and amplitude features to decoder.
Contrary to our approach, they did not then correct these amplitude
features for the jitter in their latencies by e.g. aligning trials to achieve
better separability.
Hence, their approach would not in principle render the classifier more robust
to jitter.
Furthermore we incorporated estimated latency features in both CBLE and WCBLE,
yet only WCBLE improved covert VSA performance.
This shows that the incorporation of latency features is not the only driver of
covert VSA decoding performance increase.

Despite encouraging results, our study faces some limitations that we plan to
tackle in the future.
Firstly, multiple ERP components can be time-locked to different neural
processes, each with their own jitter, hampering the performance of
single-trial latency estimation and their interpretability.
Adaptations could be made to incorporate prior time windows or probability
distributions on the latency of specific ERP components or to simultaneously
estimate set of multiple component (clusters) latencies per ERP such as in Residue Iteration
Decomposition~\cite{Ouyang2017}.
Future efforts should investigate how strong
spatiotemporal filtering can be combined with methods that allow for a more
flexible non-stationarity of the ERP like Dynamic Time Warping (DTW) or other
techniques lent from time series classification, or methods that explicitly model multiple time
displacements present in one ERP.
Secondly, performance might be improved by venturing beyond the classical
target/non-target binary classification problem.
Due to e.g. the perifoveal stimulus cruciform model~\cite{Vanegas2013}, covert and split VSA
responses might differ based on the relative position in the field of view of
their related stimulus which could be exploited in a multi-class classification
problem.
Similarly, explicitly taking into account the characteristic of the distractor ERP
response might have a beneficial effect.
Thirdly, results were obtained in an off-line and within-session evaluation,
which does not reflect true BCI operation.
Using multiple sessions with on-line feedback, the user could optimize their
performance over time by controlling attention or gaze.
Finally, since this work was conducted with patient applications in mind, we
should highlight that gaze of participants in the conducted experiments was
cued and fixed, which is per definition impossible for the patients we consider.
We are currently conducting a patient study to further investigate whether the
studied VSA conditions are appropriate and to what extent they occur in
natural patient BCI operation.

\section{Conclusion}
Our aim was to improve gaze-independent BCI performance for spatially
organized visual event-related potential (ERP) paradigms by using a suited
decoder.
Earlier results on BCI performance in covert visuospatial attention
(VSA) performance prediction have shown that accounting for single-trial
latency jitter could improve gaze-independent decoding performance.
We applied Classifier-based Latency Estimation (CBLE) as a decoder robust to latency
jitter but found no increase in gaze-independent decoding performance.
To remedy this, we improved CBLE by adapting it into CBLE with Woody
iterations (WCBLE), an iterative scheme using probabilistic latency estimation.
Results for WCBLE within and across VSA condition decoding show that
gaze-independent BCI performance can be improved at the decoding stage.
Overt decoding performance was not improved, but our proposed method can
provide added value for patients who are unable to operate a visual BCI in overt
attention mode.
Later studies should confirm whether our findings hold in patient
populations suffering from a variety of eye-motor impairments, and develop a
solution that is capable of properly handling multiple non-time locked ERP
components.

### 2024-08-24 chapters/gaze_independence.tex
Add wcble paper

\chapter{Gaze-independent BCIs}
\todo{mention partially published}
\chapter{Gaze-independent BCIs}
\section{Oculomotor deficits in Locked-in Syndrome}
\section{Gaze-independent visual BCIs}
\section{Benefits \& drawbacks of covert attention in BCI operation}
\section{Electrophysiological correlates of covert attention}
\section{Compensating for covert attention}
\todo{spatially vs temporally organized interfaces}

Gaze-independent ERP-based BCIs~\cite{Riccio2012} can be realized in three
ways.
Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
do not rely on gaze redirection but often result in lower information transfer
rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
Secondly, visual stimulation can be optimized, e.g. such that the stimuli are always
present in the field of view~\cite{Treder2010,Pires2011,Lees2018}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific ERP components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
%These methods, However, can suffer from reduced information
%transfer rates~\cite{Chennu2013}.
However, they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

Firstly, by avoiding visual stimulation entirely and opting for auditory or
somatosensory stimulation instead.
However, these alternatives often result in lower information transfer
rates, increased mental effort and
user-dependent variability~\cite{Reichert2020b}.
Secondly, gaze-independence can be achieved through tailored stimulus presentation.
This can involve enforcing overt VSA by presenting stimuli that fill the field
of view of the user,
exploiting visual non-spatial attention, or a combination thereof~\cite{Treder2011}.
The most common example of non-spatial attention is feature attention, where
the user directs their attention to an expected property of some of the
presented stimuli, such as its color, its shape or the symbol it represents~\cite{Zhang2010,Hwang2015,Reichert2020a}.
In covert VSA scenarios, stimulus paradigms are often fine-tuned to modulate
specific ERP responses~\cite{Schaeff2012,Xu2022}, increasing the
signal-to-noise ratio (SNR).
Some noteworthy examples include the GIBS Block Speller~\cite{Pires2011},
the GeoSpell interface~\cite{Aloise2012a}, and the RSVP
speller~\cite{Acqualagna2011,Lees2018}.
These systems typically require users to focus on a central fixation point while
selecting peripheral targets.
However, they often suffer from reduced information
transfer rates~\cite{Chennu2013}.
Moreover, they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial attention
can be decoded separately from gaze direction.
Earlier work has made advances in decoding lateralized covert
VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
ERP BCI paradigm, gaze-independent decoding in fast-paced ERP paradigm stimulation
remains underexplored.
\cite{Aloise2012b} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear ERP classifiers on a covert
attention P3 ERP component dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
\cite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{hardiansyah2020single} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference yet did not attempt to correct the amplitude
features for these latencies.
\cite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \cite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.

P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
value is heavily dependent on the subject and the task, and can vary from trial
to trial~\cite{Ouyang2017}.
The work of \cite{Arico2014}
illustrates that the variation in single-trial P3 latencies is important in
gaze-independent decoding and has been hampering covert VSA decoding performance.
In this work we aim to reprise their hypothesis stating that jitter compensation improves
covert VSA performance and extend it by developing a decoder and evaluating it
in a broader range of gaze-independent settings, including split VSA.
\todo{Check connection}

### 2024-08-24 chapters/interface.tex
Add wcble paper

\chapter{Interface design \& experimental setup}
\section{The Hex-o-Spell interface}
\section{Visuospatial attention conditions}
frenzel
\todo{part about split attention etc from chapter on covert alignment to here}

EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed the informed consent form and were seated at a distance of
60 cm before a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \citeasnoun{Treder2010} (\fref{fig:layout_targets}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, \fref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (\fref{fig:layout_cross_overt} and \fref{fig:layout_cross_covert}).
The position of this crosshair determined the VSA condition for this trial:
overt VSA when the crosshair was at the same location as the cued target,
covert VSA when the crosshair appeared in the center of the screen, and split
VSA when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
starts.
All targets were intensified for a duration of 100 ms, in pseudorandom order.
The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
with 200ms uniform jitter added, resulting in an ISI
between 200 and 400 ms.
Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
autocorrelation for a higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
crosshair locations.
The experiment started with five non-recorded practice stimulation blocks, one for
each of the 5 VSA conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30
seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

### 2024-08-24 chapters/patients.tex
Add wcble paper

\section{Patient presentation}
\section{Data collection \& preprocessing}
\todo{Define structure patient results/discussion}
\section{Outcomes}
\section{Discussion}

### 2024-08-24 chapters/wcble.tex
Add wcble paper

\chapter{Woody Classifier-based Latency Estimation}
\section{Latency estimation and correction}
\todo{check earlier draft alignment paper}
\todo{check deleted paragraphs in jitter paper}
\section{Classifier-based Latency Estimation}
\section{Classifier-based Latency Estimation with Woody iterations}
\section{Simulation study}

\section{Introduction}
Single-trial ERP latency jitter introduces a problem in ERP analysis:
the \textit{smearing effect} occurs when aggregating over multiple identical
signals with different latencies.
A familiar example of this happens when averaging over jittered epochs for ERP
analysis.
In the presence of jitter, the shape and amplitude of the average do not
reflect the properties of the original signal.
As the smearing effect reduces the amplitude of the average,
it negatively impacts the signal-to-noise ratio (SNR).
The smearing effect equally impacts the SNR of information captured by a
classifier's parameters when training with a procedure that does not account for
this jitter, and thus also affects its performance~\cite{Thompson2012}.
Hence, effective latency estimation and jitter compensation can contribute to higher
decoding performance.
%In this work, we propose a latency estimation and alignment strategy based on
%the method proposed by \cite{Thompson2012}.

The most straightforward latency estimation method is \textit{peak picking}.
Peak picking determines the latency of an ERP as the time point of its maximum
(or minimum for negative components) amplitude relative to stimulus
onset.
While easy to implement, this method does not perform well in low or negative
SNR conditions because of the risk of picking a noise-related peak, unless
combined with strong filtering in the space, time or frequency
domains~\cite{Avanzo2011,Arico2014,TREDER2016279,Ouyang2017}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Due to the risk of picking noise peaks, template matching is generally
preferred.
Woody's algorithm~\cite{woody1967characterization} is a simple and performant
latency estimation method that is considered to be among
the most performant ones~\cite{Ouyang2017}.
Woody's scheme starts with a template that approximates the expected response,
e.g. the average target ERP, and iteratively enhances the SNR of this template by determining at
each iteration the time point of maximum cross-correlation of the template
with all single-trial ERPs, and aligning them based on these latencies to form
the template for the next iteration.
The final, enhanced template can then be cross-correlated again to retrieve an
accurate estimation of single-trial latencies.
Similar to peak-picking, cross-correlation template matching can be combined
with a spatial filter such as XDAWN~\cite{souloumiac2013improved} or a
spatiotemporal filter \cite{Iturrate2014}.
More recent work achieves good performance in latency estimation using residue
iteration~\cite{Ouyang2017}, genetic
algorithms~\cite{Pelo2018}, a hidden process model~\cite{Kim2020} or
graph-based network analysis~\cite{Dimitriadis2018}.

When multiple clusters of ERP components are not time-locked to
each other, aligning to the latency of one cluster might introduce more jitter
in other component clusters, further decreasing the SNR~\cite{Ouyang2020}.

Finally, algorithms that use time
warping-based methods instead of explicit ERP component latency estimation can
also overcome the smearing problem.
Examples that have been applied to ERPs are Dynamic Time
Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
Correlation-Optimized Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}. Since these methods are fundamentally
different from latency estimation methods,
we will not further consider them in this work.

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target ERP response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for BCI performance
prediction, but not for ERP classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\cite{hardiansyah2020single} incorporated single-trial latencies in
classification by peak-picking within a given
ERP time window, unaware of the class of the epoch under investigation.
The Classifier-Based Latency Estimation (CBLE) algorithm introduced
by \cite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\cite{Thompson2012} initially formulated CBLE as an off-line performance prediction method.
Later, its output was successfully adapted to compensate for jitter to improve decoder performance%
~\cite{Mowla2017,Zisk2022}.

Time series classification algorithms~\cite{Abanda2019}
that are robust to jitter can be used in a decoding setting,
but, in general, have scarcely been applied to ERP decoding.
Data augmentation involving jittering the training
data~\cite{Krell2018,Zisk2022} and Riemannian Geometry methods using spatial
covariances as features~\cite{Aydarkhanov2020} have both been shown
to perform well in the presence of ERP jitter.
In this work, we opted to apply CBLE because it has successfully been applied to
classify jittered ERPs.
We adapt the CBLE to an iterative method akin to Woody's scheme that can better
compensate for jitter, to improve covert VSA decoding performance.

\section{Materials and Methods}
\subsection{Decoders}
\subsubsection{Classifier-based Latency Estimation}
\label{sec:cble}

Consider a training set of $N$ EEG epochs with $C$ channels of $S$
samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
with corresponding training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
	\textrm{non-target}\}_{n=1}^N$, and a
similar testing set of $M$ epochs $\{\mat{X}_m^\mathrm{test} \in
	\mathbb{R}^{C\times S}\}_{m=1}^M$.
We assume that the sampling period is $T$, i.e. that the sample with index $s$ is sampled at time $sT$.
In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
For instance, $\mat{X}[:,s_1:s_2]$ denotes all columns of $\mat{X}$ with indices between $s_1$ (included) and $s_2$ (excluded).

CBLE, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage CBLE for ERP classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the CBLE-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for ERP latency jitter.

When using a simple spatiotemporal linear classifier as first-stage classifier,
CBLE is equivalent to the first iteration of Woody's algorithm with the
spatiotemporal classifier weights as template.
\cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
that CBLE is relatively independent of the first-stage classifier for BCI
accuracy prediction and for ERP classification.
Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
regularized covariance matrix (tLDA) proposed by \cite{Sosulski2022}, as the first-stage
classifier and logistic regression as second stage.

\subsubsection{Robust CBLE latency features}
\label{sec:robust-latency}
The ERP decoding method based on CBLE introduced by \cite{Mowla2017} does
not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
to the second-stage classifier.
%Furthermore, previous CBLE works~\cite{Thompson2012,Mowla2020} only used these
%latencies to correlate them with neurophysiological processes or to predict
%decoder performance.
We noticed that, while CBLE performance was unaffected, the classification
performance of our proposed method can be improved if the estimated
latencies are also made available as features to the second-stage classifier,
after a square transform for linear separability~\cite{Thompson2012}.

These latency features give rise to the following issue when
classifying unseen data.
The CBLE latency estimate is defined only for target epochs as
$s_\mathrm{target}=\argmax_s{y_s}$.
This is the point in time where the target class reaches largest separation
from the background noise and non-target class, indicating the target ERP is
most likely to occur here.
However, in a classifier test phase, it is not known a priori whether an
unseen epoch is a target or a non-target epoch.
This problem is solved by defining an estimated latency per class
$s_\mathrm{target}$ and $s_\mathrm{non-target}$ for every epoch,
regardless of its actual class.
The estimated class latencies can then be used as features  for training and
testing the second-stage classifier.
This way, the latencies of the testing data can be presented to the second-stage
classifier without knowledge of the testing data class labels, making them
useful in decoding.

In a similar manner to the target latency, the non-target latency could be defined
as~$s_\mathrm{non-target}=\argmin_s{y_s}$.
However, this is problematic since it is not evident how to estimate
latency of e.g. a P3 ERP component for a non-target epoch, since the
non-target class is characterized by the absence of this component.
%\footnote{Depending on the stimulation and experimental design, ERP components can be present in one experimental
%	condition or class and missing in another, or appear in multiple classes
%	at different amplitudes. In the latter case, it could be possible
%	to estimate the latency of an ERP in the conditions where it appears, but if
%	its amplitude is lower or negligible in some conditions, latency estimates
%	will be less accurate, hence this case is still problematic.}.
In fact, $y$ can have multiple local minima or entirely lack distinct peaks for
non-targets, rendering the minimum estimate meaningless.

Instead, we opt for a more robust, probabilistic definition of class latencies.
This robust estimation method yields latencies that
\begin{enumerate*}[label=(\arabic*)]
  \item are more meaningful as input for the second-stage
    classifier, and
  \item lead to smoother convergence in our proposed iterative alignment scheme
    for WCBLE, which heavily relies on exact latency estimation.
\end{enumerate*}
Assume classifier $\mathcal{C}(\theta,f,\Pr)$ now can also output a probability
per class $\Pr(l|
\mat{X}[:,s_1:s_2:)], \theta)$ for a given epoch $\mat{X}$, a feature of many
common classifiers.
Analogous to equation~\ref{eq:score}, we can now write
\begin{equation}
\Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
  \quad \forall \quad s\in 1,\ldots,R
	\label{eq:prob}
\end{equation}
The latency features assuming the epoch belongs to a given class
$l\in\{\textrm{target},\textrm{non-target}\}$ are then defined as the median of
the corresponding  distributions
\begin{equation}
  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right] \\
  \label{eq:prob-latency}
\end{equation}
Note that $  \Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
  \Pr(s|\mat{X},\theta,\mathrm{target})$.
The median of the probability distribution over time is more robust to
outliers and noise than the maximum or minimum score.
For the non-target case, the median approach tends towards the center of a
near-uniform distribution, resulting in a more consistent latency estimate over
trials as compared to the minimum approach.

\subsubsection{Classifier-based Latency Estimation with Woody
	Iterations}
To improve performance over CBLE, we propose a new algorithm inspired both by
CBLE and the aforementioned Woody iteration scheme (WCBLE).
Instead of using CBLE to estimate the features of a second-stage classifier
directly, CBLE latency estimation is used as a step in a Woody iteration scheme.
While the Woody algorithm iteratively enhances the SNR of an ERP template to
cross-correlate with the data, WCBLE iteratively re-estimates the parameters of
the first-stage classifier.
To improve convergence and perform well in a classification setting, WCBLE
aligns both targets and non-targets to their corresponding estimated latencies.

The WCBLE algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in CBLE:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\begin{figure*}
	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
	\caption{Schematic representation of the Woody Classifier Based Latency
    Estimation training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is	then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities	over time.
    (3) The medians of these probability distributions are assumed the class
    latencies.
    (4) The epochs are aligned to their corresponding	class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for a	next iteration.
    (6) After the iterative process halts, the scores and	latencies obtained
    from the last iteration are used to train the second-stage
		classifier.}
	\label{fig:diagram-train}
\end{figure*}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a CBLE
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\begin{figure*}
	\includegraphics[width=\linewidth]{figures/wcble/figure2.pdf}
	\caption{Schematic representation of the Woody Classifier Based Latency
    Estimation test phase. (1) The
		first-stage	spatiotemporal binary classifier obtained from the training phase is
		applied to time-shifted copies of the epochs to obtain scores and class probabilities
		over time. (2) The medians of these probability distributions are assumed
		as the new class latencies. (3) The scores and class latencies are input to the
		trained second-stage classifier, which predicts the label of the epochs.}
	\label{fig:diagram-eval}
\end{figure*}
\section{Algorithms}

\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mbfX_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mbfX_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
      \algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mbfX_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mbfY^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mbfX_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mbfX_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mbfY^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$

	\caption{Classifier-Based Latency Estimation}
	\label{alg:cble}
\end{algorithm}

\newpage
\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mbfX_n^\mathrm{train}\}_{n=1}^N, \mathbf{l},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mbfX'_n \gets \mbfX^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mbfX'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[Pr(s|\mbfX_n',\theta^*,l_n)\right]$
		\State $\mbfX'_n \gets \mathrm{align}(\mbfX^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mbfX_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

	\end{algorithmic}
	\textbf{Output:} $\theta^*,
  \mbfY^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$

	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mbfX_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets
      f(\mbfX_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mbfY^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
	\caption{Classifier-Based Latency Estimation with Woody Iterations}
	\label{alg:wcble}
\end{algorithm}

### 2024-08-26 chapters/interface.tex
Add wcble paper

@@
 with a radius of 12.28° of visual angle centered on the midpoint of the screen,
-conforming to the interface proposed by \citeasnoun{Treder2010} (\fref{fig:layout_targets}).
+conforming to the interface proposed by \cite{Treder2010} (Figure~\ref{fig:layout_targets}).
 A hexagonal layout interface with an empty center and a low number of targets
@@
 Targets are full-contrast white and were intensified by scaling them to a
-larger size (5.60° of visual angle, \fref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
+larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
 experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
@@
 was instructed to fixate their gaze on the blue crosshair for the duration of
-the stimulation block (\fref{fig:layout_cross_overt} and \fref{fig:layout_cross_covert}).
+the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
 The position of this crosshair determined the VSA condition for this trial:

### 2024-08-26 chapters/bci.tex
Add wcble paper

@@
 \section{Rationale}
+\todo{people helped and societal impact medical vs recreational}
+\todo{figure BCI loop}
+\todo{definition bci society}
 \section{Overview \& state-of-the-art}
+\subsection{Recording technologies}
+\subsection{BCI paradigms}
+\subsection{Decoders}
+~\cite{Review of classification techniques}
 different recording technologies and level of invasiveness + infographic

### 2024-08-26 chapters/wcble.tex
Add wcble paper

@@
 \section{Algorithms}
-
+\todo{insert comments}
 \begin{algorithm}[H]
@@
 	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mbfX_n^\mathrm{train}\}_{n=1}^N,
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
   \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
@@
 		\State $\theta \gets
-			\mathrm{train}_\mathcal{C}(\{\mbfX_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
-      \algorithmiccomment{Train stage 1}
+			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
+      %\algorithmiccomment{Train stage 1}
 		\For{$n=1\ldots N$}
-      \algorithmiccomment{Feature extraction for stage 2}
+      %\algorithmiccomment{Feature extraction for stage 2}
 		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets f(\mbfX_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
+      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
         \EndFor
         \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta,\mathrm{target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
 		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta,\textrm{non-target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
     \EndFor
@@
   \textbf{Output:} $\theta,
-  \mbfY^\mathrm{train},
+  \mat{Y}^\mathrm{train},
   \mathbf{s}_\mathrm{target}^\mathrm{train},
@@
 	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mbfX_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
   \begin{algorithmic}[1]
 		\For{$m=1\ldots M$}
-      \algorithmiccomment{Feature extraction for stage 2}
+      %\algorithmiccomment{Feature extraction for stage 2}
 		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets f(\mbfX_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
+      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
         \EndFor
         \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta,\mathrm{target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
 		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta,\textrm{non-target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
     \EndFor
 	\end{algorithmic}
-  \textbf{Output:} $\mbfY^\mathrm{test},
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
   \mathbf{s}_\mathrm{target}^\mathrm{test},
@@
 	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mbfX_n^\mathrm{train}\}_{n=1}^N, \mathbf{l},
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l},
 		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
 	\begin{algorithmic}[1]
-		\State $\mbfX'_n \gets \mbfX^\mathrm{train}_n \quad \forall \quad n=1\ldots
+		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
     N$
-      \algorithmiccomment{Train stage 1}
+      %\algorithmiccomment{Train stage 1}
 		\Repeat
-		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mbfX'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
+		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
 		\For{$n=1\ldots N$}
     \State  $s_n \gets
-			    \mathrm{median}\left[Pr(s|\mbfX_n',\theta^*,l_n)\right]$
-		\State $\mbfX'_n \gets \mathrm{align}(\mbfX^\mathrm{train}_n, s^*_n)$
+			    \mathrm{median}\left[Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
+		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
 		\EndFor
     \Until{convergence or maximum iterations reached}
-    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
+    \For{$n=1\ldots N$}%\algorithmiccomment{Feature extraction for stage 2}
 		  \For{$s=1\ldots R$}
       \State $y_{n,s}^\mathrm{train} \gets
-      f(\mbfX_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
+      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
         \EndFor
         \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
 		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[Pr(s|\mbfX_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
     \EndFor
@@
 	\textbf{Output:} $\theta^*,
-  \mbfY^\mathrm{train},
+  \mat{Y}^\mathrm{train},
   \mathbf{s}_\mathrm{target}^\mathrm{train},
@@
 	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mbfX_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
   \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
+		\For{$m=1\ldots M$}%\algorithmiccomment{Feature extraction for stage 2}
 		  \For{$s=1\ldots R$}
       \State $y_{m,s}^\mathrm{test} \gets
-      f(\mbfX_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
+      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
         \EndFor
         \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
 		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[Pr(s|\mbfX_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
+			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
     \EndFor
 	\end{algorithmic}
-  \textbf{Output:} $\mbfY^\mathrm{test},
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
   \mathbf{s}_\mathrm{target}^\mathrm{test},

### 2024-08-26 chapters/gaze_independence.tex
Add dropped gaze-independence literature

@@
 Gaze-independent ERP-based BCIs~\cite{Riccio2012} can be realized in three
-ways.
-Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
+ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
 do not rely on gaze redirection but often result in lower information transfer
 rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
+
 Secondly, visual stimulation can be optimized, e.g. such that the stimuli are always
-present in the field of view~\cite{Treder2010,Pires2011,Lees2018}.
+present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
+Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
+the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
+speller~\parencite{Acqualagna2011}.
 Non-spatial visual attention (feature attention) can also be exploited, such as
@@
 that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
-%These methods, However, can suffer from reduced information
-%transfer rates~\cite{Chennu2013}.
+These systems typically require users to focus on a central fixation point while
+selecting peripheral targets.
+These methods, However, can suffer too from reduced information
+transfer rates~\cite{Chennu2013}.
 However, they still rely to some extent on
 eye motor control, often necessitating central gaze fixation.
-
-Firstly, by avoiding visual stimulation entirely and opting for auditory or
-somatosensory stimulation instead.
-However, these alternatives often result in lower information transfer
-rates, increased mental effort and
-user-dependent variability~\cite{Reichert2020b}.
-Secondly, gaze-independence can be achieved through tailored stimulus presentation.
-This can involve enforcing overt VSA by presenting stimuli that fill the field
-of view of the user,
-exploiting visual non-spatial attention, or a combination thereof~\cite{Treder2011}.
-The most common example of non-spatial attention is feature attention, where
-the user directs their attention to an expected property of some of the
-presented stimuli, such as its color, its shape or the symbol it represents~\cite{Zhang2010,Hwang2015,Reichert2020a}.
-In covert VSA scenarios, stimulus paradigms are often fine-tuned to modulate
-specific ERP responses~\cite{Schaeff2012,Xu2022}, increasing the
-signal-to-noise ratio (SNR).
-Some noteworthy examples include the GIBS Block Speller~\cite{Pires2011},
-the GeoSpell interface~\cite{Aloise2012a}, and the RSVP
-speller~\cite{Acqualagna2011,Lees2018}.
-These systems typically require users to focus on a central fixation point while
-selecting peripheral targets.
-However, they often suffer from reduced information
-transfer rates~\cite{Chennu2013}.
-Moreover, they still rely to some extent on
-eye motor control, often necessitating central gaze fixation.
+\todo{advantages
+disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
 

### 2024-09-16 chapters/bci.tex
BCI chapter

--- 2024-08-24 chapters/bci.tex
@@
-\todo{chapters don't nicely follow imrad structure. Is this  problem?}
-\todo{concern: is a bit in between formula 1 and formula 2}
-\section{Rationale}
-\section{Overview \& state-of-the-art}
+\todo{Shortly describe challenges from the patient point of view, related to
+this thesis}
+
+\section{A direct interface to the brain}
+
+Brain-computer interfaces (BCIs) read your mind.
+BCIs are an entertaining subject with many futuristic and more established
+applications.
+Devices processing direct inputs from the central nervous system can be
+useful for rehabilitation, medical diagnosis or treatment.
+They have always been especially promising as assistive technology for
+paralyzed patients limited in their communication ability. \todo{expand more on this last
+sentence, put patients first in introduction}
+Furthermore, they bring a new paradigm to human-computer interaction,
+especially when paired with virtual or augmented reality.
+
+Simply put, a BCI records the user's brain activity, then extracts some
+relevant output from this brain activity and couples this output to a function
+of a device.
+Optionally, the user can then observe the action of the device and adapt their
+behaviour or brain activity accordingly, closing the loop
+\footnote{
+  \textcite{BCISociety2024} has recently formalized this into the
+  following definition of a BCI:
+  \it``A brain-computer interface is a system that measures brain activity and
+  converts it in (nearly) real-time into functionally useful outputs to replace,
+  restore, enhance, supplement, and/or improve the natural outputs of the brain,
+  thereby changing the ongoing interactions between the brain and its external or
+  internal environments. It may additionally modify brain activity using targeted
+  delivery of stimuli to create functionally useful inputs to the
+  brain.''
+}.
+
+\fullpagefig{figures/bci/bci_loop.pdf}{The Brain-Computer Interfacing loop.}{fig:bci-loop}
+
+Let's break this definition down and focus on the separatate representing what is happening in the brain
+is the neuronal \textit{action potential}.
+The brain consists 86 billion neurons, which are highly interconnected cells that are
+the smallest units of information processing.
+The action potential is an electrical pulse occurring when a neuron receives
+input from another neuron and is activated.\todo{kort door de bocht}
+A neuron's action potential, or the combined action potentials of groups of
+synchronized neurons, thus generate an electrical field in and near the brain,
+which can be measured using electrodes.
+Fluctuations in the brain's electrical potential are a much faster and more
+informative signal for our purposes.
+\todo{lower part to section}
+While a myriad of other techniques can be used to measure functional brain
+activity, such as functional ultrasound imaging (fUS) or magnetoencephalograpy
+(MEG), we'll limit ourselves in this work to measuring the electrical field.
+Section~\ref{sec:bci-recording} gives an overview of current recording
+technologies.\todo{make shorter, move part to section}
+
+\todo{paradigm}
+The brain waves form only one part of the interaction scheme presented
+in Figure~\ref{fig:bci-loop}.
+Most importantly, wee need to know what exactly to look for in this signal.
+For this, the user needs to perform a task which evokes specific activity of
+which it is known that it can be retrieved from the brain signal.
+For example, the imagined movement of an arm is detectable f
+
+Next, this brain signal needs to be interpreted \emph{in function of} the
+task we want our BCI to perform.\todo{feature extraction}
+The electrical activity of the brain is relatively weak compared to that of the
+environment around us, containing electronic devices and other sources of
+interference.
+Furthermore, the brain is continuously processing information and carrying out
+`background tasks', all of which generate their own electrical signals.
+On the other hand, the brain activity we are interested in exploiting often
+originates from a specific functional system within the brain, and its activity
+might not be easily discerned from all other neural and environmental activity,
+akin to trying to listen to a single speaker in a noise environment where
+everyone is shouting over each other.
+Some obvious sources of interference can be easily filtered out, but to obtain
+a good measure reflecting only the activity that is relevant for a given BCI
+problem.
+This problem is called \emph{decoding} and is often solved through supervised
+machine learning.
+Section~\label{sec:bci-decoding} takes a closer look at some common and
+state-of-the art techniques.
+A machine learning algorithm can use some training data for which the performed
+tasks are known, and learn to recognize these tasks in new data.
+The training data can be either obtained from the BCI user themselves or from
+other users.
+In the first case, the user would perform a short calibration session before
+using the BCI, where they are instructed to perform a known task.
+This calibration session can be eliminated by training the decoder on
+preexisting labeled data from other sessions and users, but this is harder due
+to the variability between subjects and sessions.\todo{make shorter, move part
+to section}
+
+Finally, the loop can be closed by coupling this output can be coupled to a
+device or actuator.
+There are two aspects to this.
+On the one hand, the BCI gains its function by allowing the user to interact
+with their environment.
+On the other hand, the actions performed by the BCI can themselves influence
+the user's brain activity, creating an adaptive system.
+A direct form of closed-loop BCI is the use of a neurostimulator.
+Examples of this are deep brain stimulation to mitigate the symptoms of
+Parkinson's disease, or the cochlear implant as an auditive prosthesis.
+More indirectly modulation can be achieved by sensory stimulation.
+This can involve specific, paradigm-related sensory input, like haptic feedback when a
+movement BCI detects an intended action, or simply presenting selected actions
+back to the user.
+The user's brain will then adapt through reinforcement learning, causing
+changes in behavior or strategy.
+Due to neuroplasticity, the brain's ability to adapt, this can have a positive
+impact on BCI performance and opens the pathway to rehabilitation applications.
+
+\todo{check excercise session presentation}
+
+\section{Recording technologies}
+\label{sec:bci-recording}
 different recording technologies and level of invasiveness + infographic
+cost, ease of use
+\section{BCI Paradigms}
+\label{sec:bci-paradigms}
+\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
+
+\section{Decoders}
+\label{sec:bci-decoders}
+evaluated on
+information transfer rate,
+roc auc,
+accuracy
+
+Calibration, transfer learning, riemannian geometry, (multi)-linear techniques,
+paper fabien lotte
+
 active/reactive/passive + infographic
-\section{Visual Event-Related Potential based BCIs}
+
+\section{A case study: the visual oddball ERP paradigm}
+\subsection{stimulation}
+\subsection{The event-related potential}
+\todo{figure of P3 matrix}
+P300 and other components
+\todo{figure of ERP components}
+\subsection{Decoding ERPs}
+tLDA, beamforming
+Linear decoding techniques

### 2024-09-16 chapters/covert_align.tex
BCI chapter

@@
-\section{Data collection \& preprocessing}
-\section{Results}
-\section{Discussion}
-\todo{check chapters paper}
+\section{Introduction}
 
-\subsubsection{Riemannian Geometry}
-We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
-approaches that rely on spatial covariance as features.
-Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
-performance~\cite{Lotte2018}.
-We implemented two Riemannian Geometry pipelines.
-The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
-filters, projects these covariances to a tangent space and classifies the
-result using $L2$-regularized logistic regression
-(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
-Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
-favorable performance in the presence of single-trial ERP latency jitter.
-Shrunk spatial covariance matrices are estimated from epochs that are
-augmented by concatenating the average target and average non-target ERP as
-extra channels, projected to tangent space and classified using
-$L2$-regularized logistic regression (ERPCov-TS-LR).
+frenzel
+\todo{part about split attention etc from chapter on covert alignment to here}
 
-\subsection{Experiments}
+\section{Materials \& methods}
+
+\subsection{Data collection}
 We evaluate our approach on a
@@
 
-\subsubsection{CVSA-ERP dataset}
 We recorded a dataset to validate our approach.
@@
 
+\subsubsection{CVSA-ERP dataset}
+EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
+(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
+international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
+Germany), with electrodes located at AFz and FCz as ground and reference respectively.
+Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
+Electrodes TP9 and TP10, used for off-line re-referencing, were directly
+attached to the skin using stickers for better contact.
+The power line frequency in Belgium is 50 Hz.
+Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
+Canada) in non-fixation mode.
+
+Participants signed the informed consent form and were seated at a distance of
+60 cm before a CRT-emulating monitor (VPixx
+Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
+circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
+with a radius of 12.28° of visual angle centered on the midpoint of the screen,
+conforming to the interface proposed by \cite{Treder2010} (Figure~\ref{fig:layout_targets}).
+A hexagonal layout interface with an empty center and a low number of targets
+counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
+targets, no other target can be between the subject’s gaze and a covertly
+attended target.
+Targets are full-contrast white and were intensified by scaling them to a
+larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
+experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
+peripheral visual field.
+Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.
+
+The participant was instructed to press the space bar when ready for a block
+of stimulations.
+Then, one target was indicated as the cue and the participant
+was instructed to count the number of intensifications of the cued target
+during the following block of stimulations.
+After pressing the space bar again, a blue crosshair appeared, and the subject
+was instructed to fixate their gaze on the blue crosshair for the duration of
+the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
+The position of this crosshair determined the VSA condition for this trial:
+overt VSA when the crosshair was at the same location as the cued target,
+covert VSA when the crosshair appeared in the center of the screen, and split
+VSA when the crosshair appeared on a different target than the cued one.
+After pressing the space bar again and a delay of 5 seconds, the stimulation block
+starts.
+All targets were intensified for a duration of 100 ms, in pseudorandom order.
+The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
+was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
+with 200ms uniform jitter added, resulting in an ISI
+between 200 and 400 ms.
+Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
+longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
+autocorrelation for a higher statistical test precision.
+In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
+This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
+input prompt appeared to enter the mentally counted number of intensifications.
+After inputting this number, the subject was allowed to pause until pressing the space bar again.
+In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
+split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
+($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
+crosshair locations.
+The experiment started with five non-recorded practice stimulation blocks, one for
+each of the 5 VSA conditions.
+During these practice blocks, the participant received feedback about their gaze
+position and counting accuracy.
+Counting the instructions and the participant's response to the
+input prompts, a block lasted about 30
+seconds. In sum, the experiment featured approximately 45 minutes of
+stimulation time.
+After blocks 14 and 28, the participant was allowed to take a longer break.
+Including these longer breaks, the experiment lasted approximately one hour.
+\subsubsection{Riemannian Geometry}
+We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
+approaches that rely on spatial covariance as features.
+Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
+performance~\cite{Lotte2018}.
+We implemented two Riemannian Geometry pipelines.
+The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
+filters, projects these covariances to a tangent space and classifies the
+result using $L2$-regularized logistic regression
+(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
+Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
+favorable performance in the presence of single-trial ERP latency jitter.
+Shrunk spatial covariance matrices are estimated from epochs that are
+augmented by concatenating the average target and average non-target ERP as
+extra channels, projected to tangent space and classified using
+$L2$-regularized logistic regression (ERPCov-TS-LR).
+
 \subsubsection{BNCI2014-009 dataset}
@@
 
-\subsubsection{Data preprocessing and analysis}
+\subsection{Data preprocessing and analysis}
 Analysis was performed using Python and the MNE software package (version

### 2024-09-16 chapters/cv.tex
Add figures
BCI chapter

--- 2024-08-15 backmatter/cv.tex
@@
 \newenvironment{cvlist}{
-	\begin{description}[font=\normalfont]\setlength\itemsep{0em}}
-		{\end{description}}
+	\begin{description}[font=\normalfont]%\setlength\itemsep{0em}
+}{
+  \end{description}
+}
 
@@
 % Curriculum vitae =============================================================
-\chapter{Curriculum vitae}
-\section*{About me}
+\chapter{Curriculum vit\ae}
 
+\subsection*{About me}
 I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
@@
 
-\paragraph{Keywords:} brain-computer interfaces, event-related
-potentials, multilinear classification, covert visual attention
+\subsection*{Keywords}
+brain-computer interfaces, event-related
+potentials, (multi)linear classification, covert visual attention
 
@@
 	}
-	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University, 2021--2024}
+	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
 \end{cvlist}
@@
 
-\subsubsection{Master theses and internships}
+\subsubsection{Master theses and internships supervised}
 \begin{cvlist}
@@
 
-\subsection*{Professional and volunteering Experience}
+\subsection*{Professional experience}
 \begin{cvlist}
@@
 		scientifically validated neuroscience and AI research.}
-	\cvitem{Freelance full-stack web developer}{Self-employed,
+	\cvitem{Freelance full-stack web developer}{self-employed,
 		2014--2020}
@@
 \begin{cvlist}
-	\cvitem{Dutch}{Native}
-	\cvitem{English}{Fully proficient}
-	\cvitem{French}{Advanced}
-	\cvitem{German}{Intermediate}
+	\cvitem{Dutch}{native}
+	\cvitem{English}{fully proficient}
+	\cvitem{French}{advanced}
+	\cvitem{German}{intermediate}
 \end{cvlist}

### 2024-09-16 chapters/gaze_independence.tex
BCI chapter

@@
-\chapter{Gaze-independent BCIs}
 \todo{mention partially published}
-\chapter{Gaze-independent BCIs}
-\section{Oculomotor deficits in Locked-in Syndrome}
-\section{Gaze-independent visual BCIs}
-\section{Benefits \& drawbacks of covert attention in BCI operation}
-\section{Electrophysiological correlates of covert attention}
-\section{Compensating for covert attention}
 \todo{spatially vs temporally organized interfaces}
+\todo{talk about covert shifts in spatial attention alpha}
 
-Gaze-independent ERP-based BCIs~\cite{Riccio2012} can be realized in three
+\section{The gaze-dependence problem}
+
+One of the goals of Brain-computer interfacing is establishing a communication
+channel that does not rely on speech or
+muscular activity~\cite{Naci2012,Chaudhary2016}, which in turn can provide
+solutions to paralyzed individuals.
+%The quest for performant and affordable solutions is most evident in
+%visual BCIs using the electroencephalogram (EEG), where it has led to a gamut
+%of increasingly sophisticated decoders, tailored to the needs
+%of specific stimulation paradigms and usage contexts.
+
+Traditional visual BCI scenarios require the user to overtly direct their
+visuospatial attention (VSA) and gaze toward the screen target they intend to select.
+In most settings, screen targets are overlaid with non-overlapping, transient
+stimuli that evoke event-related EEG potentials (ERPs).
+The selected target can be decoded from these ERPs, as is the case for the
+oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
+component.
+However, a critical challenge arises when users rely solely or in part on covert
+VSA, which involves directing visuospatial attention without corresponding eye gaze.
+In these cases, classical solutions often fall short of the widely accepted
+80\% target selection accuracy threshold deemed necessary for a comfortable user
+experience~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019,Neeling2019},
+calling for alternative, gaze-independent solutions.
+
+In this work, we will use the term \emph{gaze-independent} meaning
+\emph{‘dealing explicitly with the fact that a user cannot control their
+gaze.'}
+In the context of a visual BCI, this means that the user's
+visuospatial attention and their gaze do not necessarily coincide.
+Gaze-independent paradigms are particularly
+promising for individuals with impaired eye motor control, such as those
+suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
+Multiple Sclerosis (MS), stroke, or brain stem
+stroke.
+For these patients, gazing directly at a screen target may be uncomfortable, impractical,
+or even impossible.
+Hence, assistive devices that rely on eye tracking are often inefficient for
+them.
+Consequently, while BCIs hold great promise for these individuals, conventional
+gaze-dependent BCI solutions do not meet their needs due to the absence of gaze
+control. Therefore, the development of decoding strategies that account for covert
+VSA becomes crucial in the pursuit of high-performance gaze independent
+BCIs.
+
+\section{Incidence of eye motor impairment}
+Table~\ref{tab:incidence} reports the relatively high frequencies of
+eye motor problems, which can range from involuntary eye movements to control
+issues and even to partial or complete ophtalmoplegia.
+\begin{table}
+	\centering
+	\begin{tabular}{@{}|l|rrrrr|r|@{}}
+		\hline
+		         & ALS  & MS   & Stroke  &
+		DMD      & SMA  & LiS                           \\ \hline
+		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
+		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
+		Complete & 17\% & -    & +       & - & - & 2\%  \\
+		\hline
+	\end{tabular}
+	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
+		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
+		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
+		Locked-In-Syndrome).}\label{tab:incidence}
+\end{table}
+\todo{document methodology}
+\todo{find more references, fill out}
+\todo{
+Patients with
+Totally locked in/locked in with eye motor issues due to
+Brain stem stroke
+ALS
+MS
+Motor and eye motor impairments due to neurotrauma (tetraplegia)
+...
+}
+
+\todo{define eye motor issues}
+\todo{systematic review of prevalences and prevalences of eye motor issues per disorder}
+\todo{
+cite{Guo2022}
+cite{Prasad2010}
+cite{Serra2018}
+cite{Bogousslavsky1987}
+cite{Castelnovo2016}
+cite{Polet2020}
+}
+
+\section{Gaze-independent BCI paradigms}
+\label{sec:gaze-independence-paradigms}
+Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
 ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
@@
 rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
+\todo{insert table from riccio in text as numbers, not as table}
 
@@
 that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
-These systems typically require users to focus on a central fixation point while
+Nevertheless, these methods, can suffer too from reduced information
+transfer rates~\cite{Chennu2013}.
+Furthermore, these systems typically require users to focus on a central fixation point while
 selecting peripheral targets.
-These methods, However, can suffer too from reduced information
-transfer rates~\cite{Chennu2013}.
-However, they still rely to some extent on
+This entails that they still rely to some extent on
 eye motor control, often necessitating central gaze fixation.
-\todo{advantages
-disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
+\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
+\todo{
+Firstly, by avoiding visual
+stimulation entirely and opting for auditory or somatosensory stimulation instead.
+However, these alternatives often result in lower information transfer rates,
+increased mental effort and variable outcomes for different users.
+Secondly, gaze-independence can
+be achieved by adapting the interface to display stimuli such that they
+are always in the center of the field of view, by exploiting visual
+non-spatial attention (feature attention), or a combination thereof.
+These interfaces suffer equally from reduced information transfer rates.
+}
 
@@
 features for these latencies.
+\todo{Should we already mention Frenzel here?}
 \cite{Frenzel2011} introduced a similar protocol to our split VSA

### 2024-09-16 chapters/patients.tex
BCI chapter

@@
+\section{Introduction}
+
+Communication BCIs have a target population consisting of patients in various
+stages of paralysis or Locked-in Syndrome, often suffering from oculomotor im-
+pairments. This reduces their performance operating visual event-related poten-
+tial (ERP) BCIs, since they are unable to comfortably redirect their gaze at the de-
+sired target, i.e., overt visuospatial attention (VSA). Instead, they are forced to op-
+erate in covert VSA, where the gaze and VSA do not coincide. Several studies show
+that performance drops in covert VSA [2, 1], necessitating gaze-independent so-
+lutions. Our previous study [3] indicates gaze-independent performance can be
+improved in healthy subjects using a suited decoding strategy. Currently, we aim
+to verify this in patients.
+
+\section{Materials \& methods}
+\subsection{Patient recruitment}
+\subsection{Eye motor examination}
+\todo{include eye test document juliette}
+\subsection{Data collection \& preprocessing}
+\subsection{BCI Decoding}
 \section{Patient presentation}
-\section{Data collection \& preprocessing}
 \todo{Define structure patient results/discussion}

### 2024-09-16 chapters/preface.tex
BCI chapter

\chapter{Preface}
The mind works in mysterious ways\ldots\
Let's try to hack it!

I'd like to acknowledge

### 2024-09-16 chapters/problem_statement.tex
BCI chapter

@@
-\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}
-
-\todo{Check provisionary and final doctoral plan for motivation}
-\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
-their residual gaze capabilities, by leveraging a non-invasive high-ITR
-visuospatial ERP-based BCI and improving ERP decoding performance (in general
-and specifically in gaze-independent settings).}
+\todo{provisionary doctoral plan}
 \todo{check WIP seminar}
-
 \todo{update text}
 
-\emph{Novelty:}
+\section{Objective}
 
@@
 However, visual BCIs cannot operate efficiently when the user does not direct
-their gaze onto the
-desired target~\cite{Brunner2010}.
+their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
 
@@
 control, or patients with gaze impairments are excluded.
-Table~\ref{tab:incidence} reports the relatively high frequencies of
-eye motor problems, which can range from involuntary eye movements to control
-issues and even to partial or complete ophtalmoplegia.
-\begin{table}
-	\centering
-	\begin{tabular}{@{}|l|rrrrr|r|@{}}
-		\hline
-		         & ALS  & MS   & Stroke  &
-		DMD      & SMA  & LiS                           \\ \hline
-		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
-		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
-		Complete & 17\% & -    & +       & - & - & 2\%  \\
-		\hline
-	\end{tabular}
-	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
-		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
-		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
-		Locked-In-Syndrome).}\label{tab:incidence}
-\end{table}
-\todo{document methodology}
-\todo{find more references, fill out}
 Different degrees of eye motor impairment can render visual BCIs
@@
 
-\section{Objective}
-
-The general goal of this PhD is to tackle the gaze-dependency in visual BCIs.
-Traditionally, gaze-independency~\cite{Riccio2012, Aloise2012} can be realized in three ways:
-Firstly, by avoiding visual
-stimulation entirely and opting for auditory or somatosensory stimulation instead.
-However, these alternatives often result in lower information transfer rates,
-increased mental effort and variable outcomes for different users.
-Secondly, gaze-independence can
-be achieved by adapting the interface to display stimuli such that they
-are always in the center of the field of view, by exploiting visual
-non-spatial attention (feature attention), or a combination thereof.
-These interfaces suffer equally from reduced information transfer rates.
-Moreover, they still rely to some extent on eye motor control, necessitating central gaze fixation.
+The general goal of this PhD is to tackle the gaze-dependence in visual BCIs.
+\todo{refer to ITRs from previous chapter}
+Section{sec:gaze-independence-paradigms} showed that visual BCI paradigms form
+a good candidate for a performant gaze-independent BCI.
+While the central gaze fixation of most visual gaze-independent paradigms still
+relies to some extent on eye motor control, we aim to circumvent this.
 We make a distinction here between spatially organized interfaces, where
@@
 rate, but are also more gaze-dependent than temporally organized interfaces.
-Third, stimuli can be presented in a standard BCI paradigm, but \emph{mental} attention
-can be decoded separately from \emph{visual} attention.
-\todo{table with paradigms and itrs from WIP seminar}
+In such an interface, stimuli can be presented in a standard BCI paradigm, but
+visuospatial attention can be decoded separately from the position of gaze,
+which do not necessarily need to coincide.
 
@@
 \end{quote}
+\todo{tone this down a bit to what can be proven}
 
@@
 
+\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}
+\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
+their residual gaze capabilities, by leveraging a non-invasive high-ITR
+visuospatial ERP-based BCI and improving ERP decoding performance (in general
+and specifically in gaze-independent settings).}
+
 \section{Methods}
-\todo{update with new decoders and experiments}
-
-\subsection{Interface design}
-We designed a visual oddball interface that helps us study and adapt
-to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
-decoders and on our own proposed decoders.
-In a first series of experiments, we recorded data with this interface from healthy participants.
-The goal of these experiments is to benchmark
-gaze-independent ERP decoding algorithms.
-We denote this dataset as the Covert Visuospatial Attention ERP dataset
-(CVSA-ERP)
-This study was approved by the Ethics Commission of University Hospital Leuven
-(S62547).
-
-Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
-by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
-symbols) to the participant while the EEG and electro-oculogram (EOG) are
-recorded, as well as eye gaze using eye tracking.
-To simulate the dissociation between the eye gaze (visual attention) and the
-intended target (mental attention) which could occur in patients with hampered eye
-motor control, healthy participants are cued to operate the BCI in specific
-visuospatial attention (VSA) conditions.
-Each subject performs 5 different VSA conditions as illustrated in
-Figure~\ref{fig:vsa-conditions}.
-\begin{figure}
-	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
-	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
-		asked to dissociate their visual and mental attention by gazing at a
-		crosshair while attending a target, depending on the prompted condition.}%
-	\label{fig:vsa-conditions}
-\end{figure}
-\todo{new picture that shows more clearly a user operating freely where they
-can gaze}
-In the overt case, subjects are instructed to gaze at the cued
-target they are also mentally attending; in the covert case, subjects are instructed
-to gaze at the center of the screen while mentally attending to the cued target.
-Finally, we introduce a VSA condition that is understudied in the context of
-gaze-independent BCI development: split VSA.
-In split VSA, the participant mentally focuses on one cued target while gazing at
-another (the distractor).
-Depending on the inter-target distance along the hull of the hexagon between the
-attended target and the distractor we discern three split VSA subconditions:
-the distractor is either clockwise or counterclockwise directly next to the
-attended target ($d=1$), there is one other target between the attended target and
-the distractor ($d=2$), or the distractor is opposite the intended target
-($d=3$).
-
-After data collection from healthy participants and decoder development, we
-will carry out a patient study using a similar interface.
-The goal of the patient study is to evaluate the influence of eye motor
-deficits in patient populations on existing state-of-the-art decoders and on
-our proposed decoding methods.
-\todo{update with current results}
 
@@
 
-\todo{update with current results and all chapters}
+\todo{Tensor discriminant analysis}
+
+\subsection{Interface design}
+
+\todo{Overt, covert, split, free, illustration from poster}
+\todo{clear definitions}
+
+\subsection{Experiments}
+\subsubsection{Healthy participant study}
+We designed a visual oddball interface that helps us study and adapt
+to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
+decoders and on our own proposed decoders.
+In a first series of experiments, we recorded data with this interface from healthy participants.
+The goal of these experiments is to benchmark
+gaze-independent ERP decoding algorithms.
+We denote this dataset as the Covert Visuospatial Attention ERP dataset
+(CVSA-ERP)
+This study was approved by the Ethics Commission of University Hospital Leuven
+(S62547).
+
+Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
+by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
+symbols) to the participant while the EEG and electro-oculogram (EOG) are
+recorded, as well as eye gaze using eye tracking.
+To simulate the dissociation between the eye gaze (visual attention) and the
+intended target (mental attention) which could occur in patients with hampered eye
+motor control, healthy participants are cued to operate the BCI in specific
+visuospatial attention (VSA) conditions.
+Each subject performs 5 different VSA conditions as illustrated in
+Figure~\ref{fig:vsa-conditions}.
+\todo{do we already want this here?}
+\begin{figure}
+	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
+	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
+		asked to dissociate their visual and mental attention by gazing at a
+		crosshair while attending a target, depending on the prompted condition.}%
+	\label{fig:vsa-conditions}
+\end{figure}
+\todo{new picture that shows more clearly a user operating freely where they
+can gaze}
+In the overt case, subjects are instructed to gaze at the cued
+target they are also mentally attending; in the covert case, subjects are instructed
+to gaze at the center of the screen while mentally attending to the cued target.
+Finally, we introduce a VSA condition that is understudied in the context of
+gaze-independent BCI development: split VSA.
+In split VSA, the participant mentally focuses on one cued target while gazing at
+another (the distractor).
+Depending on the inter-target distance along the hull of the hexagon between the
+attended target and the distractor we discern three split VSA subconditions:
+the distractor is either clockwise or counterclockwise directly next to the
+attended target ($d=1$), there is one other target between the attended target and
+the distractor ($d=2$), or the distractor is opposite the intended target
+($d=3$).
+
+After data collection from healthy participants and decoder development, we
+will carry out a patient study using a similar interface.
+The goal of the patient study is to evaluate the influence of eye motor
+deficits in patient populations on existing state-of-the-art decoders and on
+our proposed decoding methods.
+\todo{update with current results}
+
+\subsubsection{Patient study}
+\todo{}

### 2024-09-16 chapters/stbf_struct.tex
Add wcble paper
BCI chapter

@@
 % Kronecker-structured LDA ====================================================
-\begin{refsection}
-	\chapter{Kronecker-structured discriminant analysis}
 
@@
   \todo{Include part kronlda, check notes in journal}
-\section{Introduction}
-
+\todo{
 LCMV struct might be to restrictive
 Toeplitzlda
-MEG paper
-
-\section{Methods}
-\subsection{Spatiotemporal Linear Discriminant Analysis}
-\subsection{The single Kronecker product covariance model}
-\begin{equation}
-	\mat{C} = \mat{S}\otimes\mat{T}
-\end{equation}
-Tapering, Toeplitz and shrinkage
-\subsection{The Kronecker sum covariance model}
-\begin{equation}
-	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
-\end{equation}
-pvl, kronpca
-
-\subsection{Kronecker sum covariance estimation}
-\subsection{Classification}
-\section{Results}
-\section{Discussion}
+MEG paper}
+%\begin{equation}
+%	\mat{C} = \mat{S}\otimes\mat{T}
+%\end{equation}
+%Tapering, Toeplitz and shrinkage
+%\subsection{The Kronecker sum covariance model}
+%\begin{equation}
+%	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
+%\end{equation}
+\todo{pvl, kronpca}
 
@@
 \todo{check structure Leuven.ai poster}
-\section{The spatiotemporal EEG covariance}
-\section{The single-KP covariance model}
-\section{The sum-of-KP covariance model}
-\section{Decoding performance}
-\section{Extension into the space-time-frequency domain}
-
-	\todo{check chapters/structure paper}
-	\todo{check structure Leuven.ai poster}
-	\section{The spatiotemporal EEG covariance}
-	\section{The single-KP covariance model}
-	\section{The sum-of-KP covariance model}
-	\section{Decoding performance}
-	\subsection{Extension into the space-time-frequency domain}
-	\printbibliography
-\end{refsection}
+\todo{extension to the TFR domain}

### 2024-09-16 chapters/wcble.tex
BCI chapter

@@
-\chapter{Woody Classifier-based Latency Estimation}
-\section{Latency estimation and correction}
-\todo{check earlier draft alignment paper}
-\todo{check deleted paragraphs in jitter paper}
-\section{Classifier-based Latency Estimation}
-\section{Classifier-based Latency Estimation with Woody iterations}
-\section{Simulation study}
+\section{Introduction}
+
+\todo{General introduction of why we're doing this}
+
+\subsection{ERP latency jitter}
+
+\subsubsection{Neural origins of latency jitter in the time domain}
+While the classic ERP model assumes a constant ERP response over trials,
+research has shown that cortical evoked potentials show significant
+trial-to-trial jitter~\cite{truccolo2002trial}.
+Experiments show that latency jitter in ERPs is the result of the interacion of
+the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
+	kisley1999trial, Curto2009, Arieli1996}.
+
+The latency of an evoked potential is a combination of the modulation of oscillating activity that is
+time-locked, but not necessarily phase-locked to stimulus onset, and a
+delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
+\todo{Check this, \cite{Stokes2016,Mouraux2008}}
+\todo{Read references and check if neural origins and
+	generation model justify approach}
+
+\subsubsection{Implications for ERP analysis}
+The presence of latency jitter introduces multiple issues to ERP analysis, the
+most prominent of which is the smearing effect as illustrated
+in~\autoref{fig:smearing}. In ERP analysis, single-trial ERPs are usually
+averaged per condition to cope with the unfavorable signal-to-noise ratio  When
+averaging over multiple identical signals with jittered latencies, the shape and amplitude of the average do not reflect the properties
+of the original signal. Because the amplitude of the average is lower, the
+signal-to-noise ratio will also be negatively impacted.
+More generally, the smearing effect leads to difficulties interpreting ERP
+results and making inferences about amplitude or latency effects, because these
+are entangled when using averaging~\cite{woody1967characterization, Pfefferbaum1980, McDowell2003, Verleger2005,
+	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
+	these references}.
+
+\todo{illustrate smearing effect}
+%\begin{figure}
+%  \todo{smearing effect}
+%	\caption{The smearing effect: averaging over jittered waveforms leads to a spread out
+%		average with a lower amplitude.}
+%	\label{fig:smearing}
+%\end{figure}
+
+\subsubsection{Benefits of latency estimation}
+A common approach to counteract the smearing effect is correcting single-trial
+ERPs for jitter before averaging.\todo{reference} This can be done by estimating the latency of each
+single-trial ERP, and aligning them before averaging.
+Latency estimation has multiple advantages:
+\begin{enumerate*}[label={(\arabic*)}]
+	\item Latency-based ERP alignment can help reveal the true shape and
+	properties like peak amplitude and latency of ERPs across
+	experimental conditions. These are variables of interest for
+	testing electrophysiological an behavioral hypotheses.
+	Additionally, latency estimation allows for the extraction of
+	other descriptors from a set of single-trial ERPs, like
+	amplidute or latency variance~\cite{Hultsch2004}, which can
+	also be of interest as studied variables. For instance, in
+	\cite{Saville2014}, the inter-trial latency variability itself
+	is determined as a correlate of working memory performance.
+	\item Counteracting the smearing effect will boost the signal-to-noise
+	ratio of averaged ERPs by increasing the amplitude of the
+	ERP signal. This in turn can reveal smaller or more jittered ERP components
+	which are otherwise obscured or blurred due to desynchronization.
+	These deblurred templates can then be used for analysis of the ERP
+	waveform, or as templates of the signal of interest in
+	BCIs~\cite{arico2014influence}.
+	%this enables decoders to take into account the relatively low amplitude
+	%earlier components, leading to applications in settings where
+	%information from other components than the P300 need to be exploited,
+	%like covert visual attention paradigms~\cite{arico2014influence,
+	%hardiansyah2020single}.\todo[inline]{rework this to latency features
+	%because smaller visual components might not really be revealed
+	%since they're already time-locked}
+	\item Finally, the introduction of latency estimation and alignment generaly
+	improves BCI classification performance. Single-trial latencies can be
+	used as features to improve classification~\cite{hardiansyah2020single}.
+	Furthermore, compensating for inter-trial differences, especially when
+	extended to inter-subject differences, can aid in transfer learning or
+	generalization across subjects and protocols\cite{Iturrate2014}.
+\end{enumerate*}
+\todo{Read, summarize and cite BCI	references from bibliography}
+
+\subsection{Existing methods for latency estimation and alignment}
+
+\subsubsection{Single-component approaches}
+The most straight-forward method to estimate single-trial latency
+peak-picking. Here, the latency of an ERP is determined as the
+time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
+onset. While straight-forward, this method
+does not perform well in low or negative SNR conditions, unless combined with
+filtering. Filtering can suppress the noise contaminating EEG trials, lowering
+the risk of picking a noise peak instead of a true peak of the ERP. Several filtering approaches have been developed to be used in
+conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
+	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
+	D’Avanzo2011} and in the time-frequency
+domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
+also be done in the spatial domain. The spatial filter can be constructed
+using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
+	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.
+
+A generally more preferred approach than
+peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
+estimation method that is still being used and is considered amongst the more
+performant techniques~\cite{Ouyang2017}. It starts with the average ERP as a
+rough estimation of the aligned template ERP, and iteratively refines this
+template by determining at each step the time point of maximum
+cross-correlation of this template with all single-trial ERPs and aligning them
+on these time points to form the next template.
+Similarily, cross-correlation template matching can be combined with filtering.
+\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
+template matching the spatially filtered signal. \cite{Iturrate2014} uses a
+spatiotemporal filter. Instead of aligning the
+latencies to obtain the template for the next iteration, weighted averaging can
+also be used~\cite{gasser1983selavco}.
+
+Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
+optimum. To avoid this, a fitness function can be defined to jointly optimize
+the set of single-trial ERP latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
+
+Methods adapted to handle multiple conditions, useful in BCI setting
+\begin{itemize}
+	\item \cite{Kim2020}
+\end{itemize}
+
+\subsubsection{Multi-component approaches}
+The abovementioned methods do not take into account the fact that  ERPs are multi-component signals.
+Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
+time-locked to each other. They each can have different latency
+variabilities. The presence of more than one component can hamper the
+performance of some algorithms, and it often only possible to extract only the
+latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
+to one component and averaging, this will introduce blurring in the other
+components~\cite{Ouyang2020}.
+
+Most of the algorithms above can be adapted to work on multiple ERP components
+by carefully selecting peaks or pre-determined regions of interest.
+\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
+time-window for each ERP component to extract a latency per component.
+Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
+to utilize multiple of their components as filters, resulting in multiple time
+series where peack picking or Woody's algorithm can be applied. In general,
+however, these algorithms lack an integrated approach to deal with multiple
+ERP components. A clear evaluation is lacking of the performance of these
+methods in
+multi-component settings is lacking.
+
+Some algorithms separate components into stimulus and response locked component clusters based on their
+latency distributions~\cite{jung2001analysis, Takeda2008,
+	Zhang1998, Yin2009}. These algorithms do, however, require a response
+button trigger signal as input to determine the response time, which is
+not always present or applicable in an ERP experiment, especially in
+the case of BCIs.
+The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
+	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
+the response time, and can separate stimulus-locked, response-locked,
+and multiple central component clusters. Others, like spatial filtering methods that use
+decomposition methods to determine the spatial filter can be be modified to separate
+components.~\cite{Ouyang2017}.
+
+To categorize:\todo{categorize and analyze these references}
+\begin{itemize}
+	\item \cite{Dimitriadis2018}
+	\item \cite{Thornton2008}
+	\item Maximum likelihood~\cite{tuan1987variable, Moecks1988,Puce1994} Maximum likelihood
+	\item \cite{Gratton1989} (no access) Maximum likelihood?
+\end{itemize}
+
+\todo{Explain beamforming}
+
+Finally, the smearing problem can also be overcome by algorithms that are not
+based on ERP component latency estimation, but which use time warping-based methods
+to enhance the shape of the averaged template ERP, like Dynamic Time
+Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
+Warping~\cite{Skov2006}, or Fast Variational
+Alignment~\cite{flotho2021fast}. Since these methods work
+fundamentally different from the latency estimation methods, and cannot
+directly be used to extract latency features, we will not further consider them
+in this work.
+
+\subsubsection{Contribution}
+In this work, we introduce a new latency method that leverages data-driven,
+joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
+integrates the spatiotemporal filtering in Woody's template matching iteration
+scheme and can easily be extended to multi-component settings.
+Our method attempts to model the spatial and temporal activation of
+each ERP component, as well as it's latency distribution, and takes into acount the
+spatial and temporal correlation of the EEG background noise.
 
@@
 positive, by the difference between the latency and the window onset.
-\begin{figure*}
-	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
-	\caption{Schematic representation of the Woody Classifier Based Latency
+%\begin{figure}
+%	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
+%	\caption{Schematic representation of the Woody Classifier Based Latency
+%    Estimation training phase.
+%    (1) The first-stage spatiotemporal binary classifier is trained on a set of
+%    epochs.
+%    (2) It is	then applied to time-shifted copies of the epochs to obtain scores
+%    and class probabilities	over time.
+%    (3) The medians of these probability distributions are assumed the class
+%    latencies.
+%    (4) The epochs are aligned to their corresponding	class latencies by
+%    shifting in time such that all latencies fall at the same moment.
+%    (5) The spatiotemporal classifier is then retrained on the aligned epochs
+%    for a	next iteration.
+%    (6) After the iterative process halts, the scores and	latencies obtained
+%    from the last iteration are used to train the second-stage
+%		classifier.}
+%	\label{fig:diagram-train}
+%\end{figure}
+\fullpagefig{figures/wcble/figure1.pdf}{Schematic representation of the Woody Classifier Based Latency
     Estimation training phase.
@@
     from the last iteration are used to train the second-stage
-		classifier.}
-	\label{fig:diagram-train}
-\end{figure*}
+    classifier.}{fig:diagram-train}
 The process halts after a fixed amount of iterations or when the estimated set
@@
 in~Section~\ref{sec:robust-latency}.
-\begin{figure*}
-	\includegraphics[width=\linewidth]{figures/wcble/figure2.pdf}
-	\caption{Schematic representation of the Woody Classifier Based Latency
+\fullpagefig{figures/wcble/figure2.pdf}{Schematic representation of the Woody Classifier Based Latency
     Estimation test phase. (1) The
@@
 		as the new class latencies. (3) The scores and class latencies are input to the
-		trained second-stage classifier, which predicts the label of the epochs.}
-	\label{fig:diagram-eval}
-\end{figure*}
+    trained second-stage classifier, which predicts the label of the epochs.}{fig:diagram-eval}
 \section{Algorithms}
-\todo{insert comments}
+\todo{Pr not italic}
+\todo{re-insert comments}
 \begin{algorithm}[H]

### 2024-09-16 main.tex
BCI chapter

% ATTENTION: Don't build this, either build main_booklet.tex or
% main_manuscript.tex
%
% Preamble ====================================================================
\pdfminorversion=7
\usepackage{todonotes}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage[inline]{enumitem}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}

% Commands
\input{math.tex}
\newcommand{\includechapter}[2]{
  \chapter{#1}
  \begin{refsection}
  \input{#2}
    \printbibliography[heading=subbibliography]
  \end{refsection}
  \clearpage
}

% Bibliography
\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}
\addbibresource{publications.bib}

% Title
\title{\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication}
\author{ir. Arne Van Den Kerchove}

\begin{document}

% Front matter ================================================================
\pagestyle{front}
\frontmatter

\include{chapters/preface.tex}
\chapter{Abstract}
\chapter{Samenvatting}
\tableofcontents
\listoffigures
\listoftables

% Main matter =================================================================
\mainmatter%
\pagestyle{main}

\part{Introduction \& Literature}
\includechapter{Brain-Computer Interfaces}{chapters/bci.tex}
\includechapter{Gaze-independent BCIs}{chapters/gaze_independence.tex}
\todo{The role of jitter in gaze-independent ERP decoding}
\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}

\part{Algorithms \& Decoders}
\includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}
\includechapter{Classfier-based Latency Estimation with Woody iterations}{chapters/wcble.tex}

\chapter{Tensor discriminant analysis}
\section{The tensor structure of EEG data}
\section{Higher-order Discriminant Analysis}
\section{Block-term Tensor Discriminant Analysis}
\section{Decoding performance}
\section{Tensorizations coping with latency jitter}

\part{Experiments \& Applications}
\includechapter{Compensating jitter for gaze-independent decoding}{chapters/covert_align.tex}
\todo{Integrating gaze-tracking data}
\includechapter{Patient cases}{chapters/patients.tex}
\includechapter{Conclusion \& recommendations}{chapters/conclusion.tex}

\backmatter%
\pagestyle{front}
\chapter{Funding \& acknowledgments}
\chapter{Personal contribution}
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}
\end{document}

### 2024-09-16 chapters/bci.tex
Haemodynamic signal

@@
 
-Let's break this definition down and focus on the separatate representing what is happening in the brain
-is the neuronal \textit{action potential}.
+Let's break this definition down and focus on its separate parts.
+First of all, we need to identify a signal that is a direct representation of
+what's going on in the brain.
+One might for instance measure the fluctuations in blood flow to specific brain
+regions that are more or less active at a given time, i.e. the
+\textit{haemodynamic signal} with a scanner.
+However, this signal reacts to slowly to reflect the real-time activity that is
+of interest for our purposes, and it carries little information other than the
+brain regions where it is originating.
+A better candidate is the neuronal \textit{action potential}.
 The brain consists 86 billion neurons, which are highly interconnected cells that are

### 2024-09-16 chapters/abstract.tex
Add table, references for eye motor disabilities

Main contribution: improve gaze-independent decoding by isolating component
related to visuospatial attention from visual components

### 2024-09-16 chapters/bci.tex
Add table, references for eye motor disabilities

@@
 this thesis}
+
+tetraplegic, locked-in syndrome
+Regaining capabilities and improving QoL
+Start from wolpaw examples, for spelling
+Cite State of the art examples (walk again, speech avatar Metzger, handwriting
+willet, neuralink)
 
@@
   \textcite{BCISociety2024} has recently formalized this into the
-  following definition of a BCI:
+  following definition:
   \it``A brain-computer interface is a system that measures brain activity and
@@
 regions that are more or less active at a given time, i.e. the
-\textit{haemodynamic signal} with a scanner.
+\emph{hemodynamic signal} with a scanner.
 However, this signal reacts to slowly to reflect the real-time activity that is
@@
 brain regions where it is originating.
-A better candidate is the neuronal \textit{action potential}.
+A better candidate is the neuronal \emph{action potential}.
 The brain consists 86 billion neurons, which are highly interconnected cells that are
@@
 which can be measured using electrodes.
-Fluctuations in the brain's electrical potential are a much faster and more
-informative signal for our purposes.
-\todo{lower part to section}
-While a myriad of other techniques can be used to measure functional brain
-activity, such as functional ultrasound imaging (fUS) or magnetoencephalograpy
-(MEG), we'll limit ourselves in this work to measuring the electrical field.
-Section~\ref{sec:bci-recording} gives an overview of current recording
-technologies.\todo{make shorter, move part to section}
+Fluctuations in the brain's electrical potential can be faster and more
+informative.\todo{ref section}
 
-\todo{paradigm}
-The brain waves form only one part of the interaction scheme presented
+The recorded brain waves form only one part of the interaction scheme presented
 in Figure~\ref{fig:bci-loop}.
-Most importantly, wee need to know what exactly to look for in this signal.
-For this, the user needs to perform a task which evokes specific activity of
-which it is known that it can be retrieved from the brain signal.
-For example, the imagined movement of an arm is detectable f
+It would make little sense to start digging around at random in all the brain
+activity to try to extract the desired output.
+It would help greatly if we knew what we are looking for.
+For this reason, a BCI operates using a defined \emph{paradigm}, the manner in
+which interaction is performed.
+This usually means that the user is instructed to perform a specific task, like
+attending a specific flickering stimulus.
+We can now exploit background knowledge from prior neuroscientific research
+that the brain response evoked by attended stimuli is different of those
+unattended,
+and determine which one of a set of targets was attended.
+In turn, this can be coupled to an action, like typing the letter shown on the
+stimulus.
+The BCI paradigm comprises, if applicable, the manner of stimulation and the
+exact task performed by the user, and is often closely linked to the manner of
+feedback in the case of a closed-loop system.
+As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of BCI
+paradigms based on different systems in the brain.
 
-Next, this brain signal needs to be interpreted \emph{in function of} the
-task we want our BCI to perform.\todo{feature extraction}
+The previous points to another component of a successful BCI: the specific
+brain signals related to our paradigm need to be identified within the recorded
+activity.
+This is commonly referred to as \emph{feature extraction}
+In other words, the signal needs to be interpreted \emph{in function of} the
+task we want our BCI to perform.
 The electrical activity of the brain is relatively weak compared to that of the
@@
 
+\todo{wrap-up}
+
 \todo{check excercise session presentation}
@@
 \label{sec:bci-recording}
+The brain's activity can be recorded using various neuroimaging
+technologies.
+These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
+(fMRI) to more portable technologies like accoustic signals obtained by functional ultrasound imaging
+(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
+(fNIRS)~\cite{Borgheai2020}.
+These technologies all measure the hemodynamic signal, which only indirectly
+reflects brain activity.
+Because of the desirable high temporal resolution of
+electrophysiology~\cite{Easttom2021} and
+because of their practicality, we will focus only on methods to record
+electromagnetical activity originating from neuronal action potentials.
+The magnetical field of the brain can be measured using magnetoencephalography
+(MEG)~\cite{Mellinger2007}.
+While MEG is non-invasive and results in high-quality signals, it is relatively
+impractical due to the size of the equipment needed.
+Recent advances have been made using optically pumped magnetometers
+(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
+lab conditions.
+
+As a consequence, the BCI field relies heavily on technologies measuring the
+electrical activity of the brain.
+As mentioned earlier with MEG, an important concept in determining which technology within this group is
+suited for a BCI application is the invasiveness of the technology.
+single-unit recordings
+microelectrode arrays (utah-array, neuralink)
+ecog/µecog
+eeg
+
+EEG cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
+conduction hence low spatial resolution
+what does EEG measure?
+
 different recording technologies and level of invasiveness + infographic

### 2024-09-16 chapters/gaze_independence.tex
Add table, references for eye motor disabilities

@@
 \todo{talk about covert shifts in spatial attention alpha}
+
+\todo{define eye motor issues}
+\todo{systematic review of prevalences and prevalences of eye motor issues per disorder}
+\todo{
+cite{Guo2022}
+cite{Prasad2010}
+cite{Serra2018}
+cite{Bogousslavsky1987}
+cite{Castelnovo2016}
+cite{Polet2020}
+}
 
@@
 solutions to paralyzed individuals.
-%The quest for performant and affordable solutions is most evident in
-%visual BCIs using the electroencephalogram (EEG), where it has led to a gamut
-%of increasingly sophisticated decoders, tailored to the needs
-%of specific stimulation paradigms and usage contexts.
+In the strictest interpretation, this also means that an interface should not
+rely on the control of eye muscles which allow the subject to redirect their
+gaze or stare at a target.
+
+In fact, many patients in the BCI target population can suffer to some extent
+from eye motor disability, requiring BCIs adapted to their condition.
+Table~\ref{tab:incidence} reports the relatively high frequencies of
+eye motor problems, which can range from minor (nystagmus, other tremors, gaze
+fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia, involuntary
+movements, impaired pursuit, \ldots) and even complete ophtalmoplegia.
+issues and even to partial or complete ophtalmoplegia.
+\begin{table}[h]
+  \sffamily
+  \footnotesize
+	\centering
+	\begin{tabular}{@{}l|rrrrr|r@{}}
+              & \bfseries ALS  & \bfseries MS   & \bfseries Stroke  &\bfseries DMD &\bfseries SMA  &\bfseries LiS  \\ \hline
+		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
+		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
+		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
+	\end{tabular}
+	\caption{Incidence of eye motor impairment in selected BCI target
+		patient populations. (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
+		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
+		Locked-In-Syndrome).}\label{tab:incidence}
+\end{table}
+Among the most affected patients are ALS patients, especially those suffering
+from bulbar onset ALS~\cite{Guo2022}.
+Stroke, brain-stem and cerebellar stroke.
+Furthermore, patients with Locked-in Syndrome
+\footnote{Multiple definitions of Locked-in Syndrome are encountered in
+BCI and neurological literature. Some definitions include only
+tetraplegic patients without eye movements used for communications. Others distinguish Complete
+Locked-in Syndrome (CLIS) with full body paralysis, including no eye motor
+control at all, from a LiS state with some preserved eye movements or minor
+motor output.
+While some definitions only include patients who suffered stroke or traumatic
+brain injury to the brain stem, it can also generally refer to the state of full body paralysis
+or loss of muscle tone, combined with the inability to speak, such as occurs in
+late stage ALS.
+In clinical reality, every tetraplegic patient has their
+unique set of preserved capabilities.}
+\todo{fill out which definition we use, wolpaw criteria}
+\todo{cite studies per pathology}
+Gaze-independent paradigms are particularly
+promising for individuals with impaired eye motor control, such as those
+suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
+Multiple Sclerosis (MS), (brain stem) stroke or Friedreich's Ataxia (FA)
 
@@
 visuospatial attention and their gaze do not necessarily coincide.
-Gaze-independent paradigms are particularly
-promising for individuals with impaired eye motor control, such as those
-suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
-Multiple Sclerosis (MS), stroke, or brain stem
-stroke.
-For these patients, gazing directly at a screen target may be uncomfortable, impractical,
-or even impossible.
+For patients with eye motor impairment, gazing directly at a screen target may
+be uncomfortable, impractical, or even impossible.
 Hence, assistive devices that rely on eye tracking are often inefficient for
@@
 BCIs.
-
-\section{Incidence of eye motor impairment}
-Table~\ref{tab:incidence} reports the relatively high frequencies of
-eye motor problems, which can range from involuntary eye movements to control
-issues and even to partial or complete ophtalmoplegia.
-\begin{table}
-	\centering
-	\begin{tabular}{@{}|l|rrrrr|r|@{}}
-		\hline
-		         & ALS  & MS   & Stroke  &
-		DMD      & SMA  & LiS                           \\ \hline
-		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
-		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
-		Complete & 17\% & -    & +       & - & - & 2\%  \\
-		\hline
-	\end{tabular}
-	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
-		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
-		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
-		Locked-In-Syndrome).}\label{tab:incidence}
-\end{table}
-\todo{document methodology}
-\todo{find more references, fill out}
-\todo{
-Patients with
-Totally locked in/locked in with eye motor issues due to
-Brain stem stroke
-ALS
-MS
-Motor and eye motor impairments due to neurotrauma (tetraplegia)
-...
-}
-
-\todo{define eye motor issues}
-\todo{systematic review of prevalences and prevalences of eye motor issues per disorder}
-\todo{
-cite{Guo2022}
-cite{Prasad2010}
-cite{Serra2018}
-cite{Bogousslavsky1987}
-cite{Castelnovo2016}
-cite{Polet2020}
-}
 
@@
 \todo{Check connection}
+
+\todo{provisionary doctoral plan}
+\todo{check WIP seminar}
+\todo{update text}
+
+\section{Problem statement}
+
+\subsection{Objectives}
+
+Brain-Computer Interfaces (BCIs) decode brain activity with the aim to establish a direct communication
+pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
+devoid of these abilities, for whom BCIs can provide a means to communicate or to
+control devices.
+The quest for performant and affordable solutions is most evident in
+visual BCIs based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
+and paradigms tailored to the needs of specific stimulation paradigms and
+use contexts.
+An effective and proven method is the visual event-related potential (ERP)-based
+oddball interface.
+Here, targets are shown in short pulses on a computer screen.
+Each time the user observes a target, an ERP is evoked, the presence of which can be
+detected in the EEG-signal.
+The ERP consists of multiple components, of which some are modulated by the
+attention of the user, specifically the N2 and the P3 component.
+Decoding this modulation allows for information transfer controlled by the
+user's brain activity, as sketched in Figure~\ref{fig:bci}.
+However, visual BCIs cannot operate efficiently when the user does not direct
+their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
+
+This leads to an often unadressed paradox: the BCI target population consists
+of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
+(ALS), Multiple Sclerosis (MS), stroke (specifically
+brain-stem stroke) or traumatic brain injury, which can lead to several degrees
+of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
+control with preserved consciousness.
+While BCIs are most attractive as a solution for these patients
+compared to other assistive technologies like eye-tracking, studies often
+report bad performance in patient populations precisely due to the lack of adequate eye motor
+control, or patients with gaze impairments are excluded.
+Different degrees of eye motor impairment can render visual BCIs
+uncomfortable or outright impossible to use.
+When operating a visual BCI, the usual rapid series of forced saccades followed
+by fixation is tiring over time, even for healthy control subjects.
+A suitable alternative would allow the user to keep their eyes in an at-rest
+position of their choice while operating the BCI.
+This points to the need for gaze-independent BCIs.
+
+The general goal of this PhD is to tackle the gaze-dependence in visual BCIs.
+\todo{refer to ITRs from previous chapter}
+Section{sec:gaze-independence-paradigms} showed that visual BCI paradigms form
+a good candidate for a performant gaze-independent BCI.
+While the central gaze fixation of most visual gaze-independent paradigms still
+relies to some extent on eye motor control, we aim to circumvent this.
+We make a distinction here between spatially organized interfaces, where
+multiple targets are displayed at the same time at different spatial locations,
+and temporally organized interfaces, like Rapid Serial Visual Presentation,
+where targets or small sets of targets are shown consecutively.
+Traditionally, spatially organized interfaces have a higher information transfer
+rate, but are also more gaze-dependent than temporally organized interfaces.
+In such an interface, stimuli can be presented in a standard BCI paradigm, but
+visuospatial attention can be decoded separately from the position of gaze,
+which do not necessarily need to coincide.
+
+This leads us to adopt a specific approach: improve the performance of a
+spatially organized visual oddball ERP-based brain
+computer interface by using a suited decoding strategy.
+Hence, our working hypothesis is as follows:
+\begin{quote}
+	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
+		can achieve similar or higher information transfer rate in patient
+		populations with eye motor deficits than other
+		gaze-independent alternatives described in literature.}
+\end{quote}
+\todo{tone this down a bit to what can be proven}
+
+To achieve our goal, we innovate on decoder development and interface design,
+and we collect data from healthy control subjects, which allow us to test our decoders.
+Finally, the findings of these experiments will be leveraged to develop a BCI
+for patient use, which will be tested in a case-based patient study in which
+both performance and user comfort will be evaluated.
+
+\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}
+\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
+their residual gaze capabilities, by leveraging a non-invasive high-ITR
+visuospatial ERP-based BCI and improving ERP decoding performance (in general
+and specifically in gaze-independent settings).}
+
+\subsection{Approach}
+
+\subsubsection{Interface design}
+
+\todo{Overt, covert, split, free, illustration from poster}
+\todo{clear definitions}
+
+\subsubsection{Decoders}
+
+During this PhD, we explored different lines in decoding strategies,
+trying to tackle several problems that arise from gaze-independence, such as
+the lack or decrease in amplitude of specific ERP components, and the increased
+non-stationarity of the signal.
+As mentioned, state-of-the-art decoders have poor performance in covert attention
+settings.
+The general goal is thus to design a machine learning classifier that represents
+the ERP signal in a specific way, such that it becomes more robust to the problems
+occurring in covert attention conditions.
+
+\todo{Regularized spatiotemporal beamforming}
+Due to the lack of the N2 component and decrease of amplitude of the
+P3 component in covert attention
+settings~\cite{Treder2010}, the signal-to-noise ratio (SNR) of
+the ERP is lower than in overt attention settings.
+Therefore, a straightforward way to reach satisfactory
+gaze-independent decoding performance, might be by increasing overall ERP
+decoding performance. A more accurate classifier could yield relative
+improvements in those settings where performance is not yet near the achievable
+maximum.
+
+Therefore, we have improved upon an in-house developed, state-of-the art ERP
+decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
+this classifier as a linear discrimination problem and
+imposing regularizing constraints by structuring the noise covariance matrix
+(STBF-struct).
+Furthermore, these regularizing constraints impose temporal stationarity on
+the background noise, which is of use in our next efforts to cope with the
+non-stationarity of the P3 signal component.
+
+\todo{Classifier-based Latency Estimation with Woody iterations}
+Literature shows that better covert attention performance can be achieved by
+counteracting the non-stationarity of the P3 ERP component in covert
+attention settings~\cite{Arico2014}.
+This can be done by estimating the latency of each single-trial ERP and aligning
+the P3 peaks to all fall at the same moment.
+The resulting aligned data and the set of latencies can then be used to train a
+classifier that is more robust to jitter.
+
+Existing latency estimation methods are either not applicable to the
+classification problem of labeling unseen data, or are not robust enough to
+deal with the low SNR of the ERP.
+Classifier-based Latency Estimation (CBLE)~\cite{Mowla2017} is a technique that can leverage
+ERP latency estimation in a decoding setting, but our results show that it
+yields no improvement in gaze-independent settings.
+We improved upon this technique and extended it to a probabilistic, iterative
+method named Classifier-based Latency Estimation with Woody iterations (wCBLE).
+
+\todo{Tensor discriminant analysis}
+
+\subsubsection{Experiments}
+\todo{Healthy participant study}
+We designed a visual oddball interface that helps us study and adapt
+to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
+decoders and on our own proposed decoders.
+In a first series of experiments, we recorded data with this interface from healthy participants.
+The goal of these experiments is to benchmark
+gaze-independent ERP decoding algorithms.
+We denote this dataset as the Covert Visuospatial Attention ERP dataset
+(CVSA-ERP)
+This study was approved by the Ethics Commission of University Hospital Leuven
+(S62547).
+
+Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
+by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
+symbols) to the participant while the EEG and electro-oculogram (EOG) are
+recorded, as well as eye gaze using eye tracking.
+To simulate the dissociation between the eye gaze (visual attention) and the
+intended target (mental attention) which could occur in patients with hampered eye
+motor control, healthy participants are cued to operate the BCI in specific
+visuospatial attention (VSA) conditions.
+Each subject performs 5 different VSA conditions as illustrated in
+Figure~\ref{fig:vsa-conditions}.
+\todo{do we already want this here?}
+\begin{figure}
+	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
+	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
+		asked to dissociate their visual and mental attention by gazing at a
+		crosshair while attending a target, depending on the prompted condition.}%
+	\label{fig:vsa-conditions}
+\end{figure}
+\todo{new picture that shows more clearly a user operating freely where they
+can gaze}
+In the overt case, subjects are instructed to gaze at the cued
+target they are also mentally attending; in the covert case, subjects are instructed
+to gaze at the center of the screen while mentally attending to the cued target.
+Finally, we introduce a VSA condition that is understudied in the context of
+gaze-independent BCI development: split VSA.
+In split VSA, the participant mentally focuses on one cued target while gazing at
+another (the distractor).
+Depending on the inter-target distance along the hull of the hexagon between the
+attended target and the distractor we discern three split VSA subconditions:
+the distractor is either clockwise or counterclockwise directly next to the
+attended target ($d=1$), there is one other target between the attended target and
+the distractor ($d=2$), or the distractor is opposite the intended target
+($d=3$).
+
+After data collection from healthy participants and decoder development, we
+will carry out a patient study using a similar interface.
+The goal of the patient study is to evaluate the influence of eye motor
+deficits in patient populations on existing state-of-the-art decoders and on
+our proposed decoding methods.
+\todo{update with current results}
+
+\todo{Patient study}
+\todo{refer to chapters}

### 2024-09-20 chapters/bci.tex
Recording technologies

@@
 
+Locked-in -\> communicate
+Paraplegic -\> walk again
+Imagine that
 tetraplegic, locked-in syndrome
-Regaining capabilities and improving QoL
+Regaining capabilities and improving quality of life
 Start from wolpaw examples, for spelling
@@
 \section{A direct interface to the brain}
-
-Brain-computer interfaces (BCIs) read your mind.
-BCIs are an entertaining subject with many futuristic and more established
+\comment{This section describes a bci in layman's terms, without too much
+scientific background, jargon and literature, in the following sections of this
+chapter a more scientific style is used}
+A \ac{bci} reads your mind.
+\ac{bci}s are an entertaining subject with many futuristic and more established
 applications.
@@
 paralyzed patients limited in their communication ability. \todo{expand more on this last
-sentence, put patients first in introduction}
+sentence, put patients first in introduction, see above}
 Furthermore, they bring a new paradigm to human-computer interaction,
@@
 
-Simply put, a BCI records the user's brain activity, then extracts some
+Simply put, a \ac{bci} records the user's brain activity, then extracts some
 relevant output from this brain activity and couples this output to a function
@@
 Optionally, the user can then observe the action of the device and adapt their
-behaviour or brain activity accordingly, closing the loop
+behavior or brain activity accordingly, closing the loop
 \footnote{
@@
 
-\fullpagefig{figures/bci/bci_loop.pdf}{The Brain-Computer Interfacing loop.}{fig:bci-loop}
+\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci}loop.
+The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
+visual stimulation. Neural signals are recorded, and neurophysiological features
+related to the paradigm are extracted. Using machine learning, a decision can
+then be made based on these features, which can be presented back to the user.
+In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
+}{fig:bci-loop}
 
@@
 what's going on in the brain.
-One might for instance measure the fluctuations in blood flow to specific brain
-regions that are more or less active at a given time, i.e. the
-\emph{hemodynamic signal} with a scanner.
+Section~\ref{sec:bci-recording} gives an overview of the available options.
+One might for instance measure the fluctuations in bloodflow to specific brain
+regions that are more or less active at a given time with a scanner.
 However, this signal reacts to slowly to reflect the real-time activity that is
@@
 The action potential is an electrical pulse occurring when a neuron receives
-input from another neuron and is activated.\todo{kort door de bocht}
+input from another neuron and is activated.\todo{gross oversimplification,not
+entirely true}
 A neuron's action potential, or the combined action potentials of groups of
@@
 Fluctuations in the brain's electrical potential can be faster and more
-informative.\todo{ref section}
+informative.
 
@@
 It would help greatly if we knew what we are looking for.
-For this reason, a BCI operates using a defined \emph{paradigm}, the manner in
+For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
 which interaction is performed.
@@
 stimulus.
-The BCI paradigm comprises, if applicable, the manner of stimulation and the
+The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
 exact task performed by the user, and is often closely linked to the manner of
 feedback in the case of a closed-loop system.
-As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of BCI
+As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
 paradigms based on different systems in the brain.
 
-The previous points to another component of a successful BCI: the specific
+The previous points to another component of a successful \ac{bci}: the specific
 brain signals related to our paradigm need to be identified within the recorded
@@
 In other words, the signal needs to be interpreted \emph{in function of} the
-task we want our BCI to perform.
+task we want our \ac{bci} to perform.
 The electrical activity of the brain is relatively weak compared to that of the
@@
 Some obvious sources of interference can be easily filtered out, but to obtain
-a good measure reflecting only the activity that is relevant for a given BCI
+a good measure reflecting only the activity that is relevant for a given \ac{bci}
 problem.
@@
 tasks are known, and learn to recognize these tasks in new data.
-The training data can be either obtained from the BCI user themselves or from
+The training data can be either obtained from the \ac{bci} user themselves or from
 other users.
 In the first case, the user would perform a short calibration session before
-using the BCI, where they are instructed to perform a known task.
+using the \ac{bci}, where they are instructed to perform a known task.
 This calibration session can be eliminated by training the decoder on
@@
 There are two aspects to this.
-On the one hand, the BCI gains its function by allowing the user to interact
+On the one hand, the \ac{bci} gains its function by allowing the user to interact
 with their environment.
-On the other hand, the actions performed by the BCI can themselves influence
+On the other hand, the actions performed by the \ac{bci} can themselves influence
 the user's brain activity, creating an adaptive system.
-A direct form of closed-loop BCI is the use of a neurostimulator.
+A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
 Examples of this are deep brain stimulation to mitigate the symptoms of
@@
 This can involve specific, paradigm-related sensory input, like haptic feedback when a
-movement BCI detects an intended action, or simply presenting selected actions
+movement \ac{bci} detects an intended action, or simply presenting selected actions
 back to the user.
@@
 Due to neuroplasticity, the brain's ability to adapt, this can have a positive
-impact on BCI performance and opens the pathway to rehabilitation applications.
+impact on \ac{bci} performance and opens the pathway to rehabilitation applications.
 
@@
 \label{sec:bci-recording}
+
+\todo{why is spatial resolution good}
+
 The brain's activity can be recorded using various neuroimaging
@@
 
-As a consequence, the BCI field relies heavily on technologies measuring the
+As a consequence, the \ac{bci} field relies heavily on the
 electrical activity of the brain.
-As mentioned earlier with MEG, an important concept in determining which technology within this group is
-suited for a BCI application is the invasiveness of the technology.
-single-unit recordings
-microelectrode arrays (utah-array, neuralink)
-ecog/µecog
-eeg
+As mentioned earlier with MEG, an important concept in determining which
+technology is suited for a \ac{bci} application is the \emph{invasiveness} of the technology.
+As illustrated in Figure~\ref{fig:recording-modalities}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
+involved with brain surgery.
+\fullpagefigct{figures/bci/recording_modalities.pdf}{
+	Different electrophysiology recording modalities and \\
+	their respective position in a cross-section of the skull. \\
+	There is a trade-off between invasiveness and  \\
+	spatial resolution.
+}{fig:recording-modalities}
+\todo{find better figure with permission to reuse and without unnecessary
+modalities}
 
-EEG cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
+Specifically, invasive methods come with an increased \emph{spatial resolution},
+meaning they can extract a signal from a specific brain region or even a set of
+neurons of interest.
+Generally, penetrating microelectrodes are considered to have the highest
+spatial resolution.
+Microelectrodes can measure the Local Field Potential (LFP), the extracellular
+potential of a small group of neurons, or even intracellular single neuron
+action potential spikes.
+They come in single-electrode form, or, more popularily, in microelectrode
+arrays.
+Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
+NeuraLink implant~\cite{Musk2019}.
+
+Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
+Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
+and closed-loop adaptive control for Deep Brain Stimulation to mitigate
+Parkinson's disease symptoms~\cite{Arlotti2018}.
+Since these still penetrate the cortex, electrodes that sit on top of the
+cortex are more popular, as they are considered less invasive.
+Electrocorticography (ECoG) is a powerful method
+with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
+Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
+higher number of smaller electrodes per surface area allow for more precise
+measurements.
+
+Some impressive recent breakthroughs\todo{cite} have been realized using
+invasive mictroelectrode arrays and µECoG.
+Recent advances in recording technology focus on improving implant durability and
+lifetime\todo{cite flexible}, or on balancing the invasiveness tradeoff by
+finding new, minimally invasive ways to record from closer to the cortex.
+The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
+through the bloodstream via a catheter, removing the need for open brain
+surgery.
+However
+
+ac{eeg} cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
 conduction hence low spatial resolution
-what does EEG measure?
+what does ac{eeg} measure?
 
@@
 cost, ease of use
-\section{BCI Paradigms}
+\section{\ac{bci} Paradigms}
 \label{sec:bci-paradigms}
 \fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
+\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
+paradigms}
+passive:
+also monitoring for diagnostic purposes
+emotion recognition
+neurofeedback
+
+active:
+motor imagery
+imagined speech
+
+reactive:
+
+somatosensory:
+
+auditory:
+ERP/oddball, steady state
+
+visual paradigms:
+SSVEP, cVEP, mVEP, VEP/P300/Oddball
 

### 2024-09-20 layout/format_booklet.tex
BCI chapter
Add table, references for eye motor disabilities
Recording technologies

@@
 \floatplacement{table}{p}
+
+\usepackage{incgraph}
+\usepackage{tikz}
+%\newcommand{\fullpagefig}[3]{
+%  \incgraph[
+%    overlay={
+%    }
+%  ]{#1}
+%}
+\definecolor{lightgray}{HTML}{ECECEC}
+
+\usepackage[usenames,x11names,dvipsnames,svgnames]{xcolor}
+\usetikzlibrary{backgrounds}
+\newcommand{\fullpagefig}[3]{%
+  \igrcenter{\igrboxset{\includegraphics{#1}}%
+  \begin{tikzpicture}[background rectangle/.style={fill=lightgray}, show background rectangle]]%
+  \igrboxtikzcenter%
+  %\draw[help lines] (0,0) grid (\igrBoxWidth,\igrBoxHeight);
+  \node[above right,text width=10cm,align=justify] at ([xshift=1cm,yshift=1cm]page.south west){{\captionof{figure}{#2}}};
+  \end{tikzpicture}}
+  \label{#3}
+}
+\newcommand{\fullpagefigct}[3]{%
+  \igrcenter{\igrboxset{\includegraphics{#1}}%
+  \begin{tikzpicture}[background rectangle/.style={fill=lightgray}, show background rectangle]]%
+  \igrboxtikzcenter%
+  %\draw[help lines] (0,0) grid (\igrBoxWidth,\igrBoxHeight);
+  \node[below right,text width=10cm,align=justify] at ([xshift=1cm,yshift=-1cm]page.north west){{\captionof{figure}{#2}}};
+  \end{tikzpicture}}
+  \label{#3}
+}
+\renewcommand{\arraystretch}{1.5}

### 2024-09-20 main.tex
Add table, references for eye motor disabilities
Recording technologies

@@
 \input{math.tex}
+\input{acronyms.tex}
 \newcommand{\includechapter}[2]{
@@
 
+\newcommand{\comment}[1]{
+  \todo[color=yellow]{#1}
+}
+
 % Bibliography
@@
 % Title
-\title{\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication}
-\author{ir. Arne Van Den Kerchove}
 
@@
 % Front matter ================================================================
-\pagestyle{front}
 \frontmatter
@@
 \chapter{Abstract}
+\include{chapters/abstract.tex}
 \chapter{Samenvatting}
@@
 \listoftables
+\printacronyms[name=List of acronyms]
 
@@
 \mainmatter%
-\pagestyle{main}
 
@@
 \includechapter{Brain-Computer Interfaces}{chapters/bci.tex}
-\includechapter{Gaze-independent BCIs}{chapters/gaze_independence.tex}
+\includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
 \todo{The role of jitter in gaze-independent ERP decoding}
-\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}
+%\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}
 
@@
 \includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}
-\includechapter{Classfier-based Latency Estimation with Woody iterations}{chapters/wcble.tex}
 
 \chapter{Tensor discriminant analysis}
-\section{The tensor structure of EEG data}
-\section{Higher-order Discriminant Analysis}
-\section{Block-term Tensor Discriminant Analysis}
-\section{Decoding performance}
-\section{Tensorizations coping with latency jitter}
+\todo{add Tensorizations coping with latency jitter}
+
+\includechapter{Classfier-based Latency Estimation with Woody iterations}{chapters/wcble.tex}
 
@@
 \backmatter%
-\pagestyle{front}
 \chapter{Funding \& acknowledgments}

### 2024-09-21 chapters/bci.tex
Add gaze disabilities

@@
+Independent home use of a brain-computer interface by people with
+amyotrophic lateral sclerosis.
 \todo{Shortly describe challenges from the patient point of view, related to
 this thesis}
+\todo{do something with Wolpaw (2018) Independent home use of a brain-computer interface by people with
+amyotrophic lateral sclerosis.}
 
@@
 willet, neuralink)
+
+\todo{put focus on communication BCI}
 
@@
 On the one hand, the \ac{bci} gains its function by allowing the user to interact
-with their environment.
+with their environment.\todo{put focus on communication BCI}
 On the other hand, the actions performed by the \ac{bci} can themselves influence
@@
 
-\todo{why is spatial resolution good}
+\todo{why is spatial resolution a good thing}
+\todo{explain with each method what it actually is (e.g. electrodes placed
+at...)}
 
@@
 cortex are more popular, as they are considered less invasive.
-Electrocorticography (ECoG) is a powerful method
+\ac{ecog} is a powerful method
 with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
@@
 
-Some impressive recent breakthroughs\todo{cite} have been realized using
-invasive mictroelectrode arrays and µECoG.
-Recent advances in recording technology focus on improving implant durability and
-lifetime\todo{cite flexible}, or on balancing the invasiveness tradeoff by
-finding new, minimally invasive ways to record from closer to the cortex.
+Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
+\todo{cite} have been realized using
+invasive mictroelectrode arrays~\cite{Willett2021} and high-density
+\Ac{ecog}\cite{Metzger2023}.
+Furthermore, recent advances in recording technology focus on improving implant
+durability and resolution\todo{cite flexible}, and on balancing the invasiveness
+tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
 The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
@@
 surgery.
-However
+However, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
+potential communication \ac{bci} user groups consistently show that non-invasive
+recording methods are preferred over implanted electrodes, unless the added value of invasive BCI is
+sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
+This provides justification for focusing part of the current research effort on
+non-invasive BCI.
 
-ac{eeg} cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
+\Ac{eeg} is a de facto non-invasive electrophysiology standard, in clinical
+neurology practice as well as in \ac{bci} research.
+Developed in 1924, it is cheap, relatively practical and can offers a
+temporal resolution.
+
+outside skull, environmental/electrical noise, volume conduction activity from
+everywhere in the brain.
+
+cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
 conduction hence low spatial resolution
@@
 
-different recording technologies and level of invasiveness + infographic
 cost, ease of use
+
 \section{\ac{bci} Paradigms}

### 2024-09-21 chapters/gaze_independence.tex
Recording technologies
Add gaze disabilities

@@
 \todo{talk about covert shifts in spatial attention alpha}
+
+\section{The gaze-dependence problem}
 
@@
 
-\section{The gaze-dependence problem}
-
 One of the goals of Brain-computer interfacing is establishing a communication
-channel that does not rely on speech or
-muscular activity~\cite{Naci2012,Chaudhary2016}, which in turn can provide
-solutions to paralyzed individuals.
+channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
+which in turn can provide solutions to paralyzed individuals.
 In the strictest interpretation, this also means that an interface should not
-rely on the control of eye muscles which allow the subject to redirect their
-gaze or stare at a target.
+rely on the control of eye muscles used to redirect the gaze or for blinking.
+\todo{patients suffering paralysis due to damage to the CNS (stroke, spinal
+chord injury, tbi) or neurodegenerative diseases}
 
@@
 Table~\ref{tab:incidence} reports the relatively high frequencies of
-eye motor problems, which can range from minor (nystagmus, other tremors, gaze
-fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia, involuntary
-movements, impaired pursuit, \ldots) and even complete ophtalmoplegia.
-issues and even to partial or complete ophtalmoplegia.
+eye motor problems, which can range from minor (nystagmus
+\footnote{
+Involuntary, rhythmic, and repetitive eye movements recognizable by their
+consistent directionality (horizontal, vertical, or rotational).
+}, other eye tremors
+\footnote{
+These can include square-wave jerks, saccadic intrusions, microtremors or
+microsaccades while resting or fixating the gaze.
+}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
+\footnote{
+Weakness or limited paralysis of one or more of the muscles that control eye
+movement, leading to restricted eye motion, but not complete paralysis.
+Often, a specific movement direction (up-down, left-right) is preserved.
+}, involuntary movements, impaired pursuit, \ldots) and even complete
+ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
+This not only has an effect on the patients' vision and coordination, but also
+on their ability to operate a visual \ac{bci}.
+
 \begin{table}[h]
@@
 		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
-		Locked-In-Syndrome).}\label{tab:incidence}
+		Locked-In-Syndrome). $+$: frequent, $-$: infrequent.}\label{tab:incidence}
 \end{table}
-Among the most affected patients are ALS patients, especially those suffering
-from bulbar onset ALS~\cite{Guo2022}.
-Stroke, brain-stem and cerebellar stroke.
-Furthermore, patients with Locked-in Syndrome
-\footnote{Multiple definitions of Locked-in Syndrome are encountered in
-BCI and neurological literature. Some definitions include only
-tetraplegic patients without eye movements used for communications. Others distinguish Complete
-Locked-in Syndrome (CLIS) with full body paralysis, including no eye motor
-control at all, from a LiS state with some preserved eye movements or minor
-motor output.
+\todo{include FA}
+
+Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
+due to brain-stem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
+\todo{summarize and cite Rowe2019}
+These patients can suffer from severe or complete eye motor impairment immediately
+immediately from the onset of their condition.
+In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
+Although eye movement is often cited as one of the longest preserved
+capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
+fairly common,
+especially for patients with bulbar onset \ac{als}.
+Furthermore \textcite{Hayashi1991} show that, as these patients are supported to beyond
+respiratory failure, the disease progression will eventually also involve
+paralysis of the eye muscles.
+Various forms of eye movement abnormalities also occur often and are especially well studied in
+\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
+These abnormalities can be minor or severe, seldom progressing to complete
+paralysis.
+However, \ac{ms} often comes with vision loss, further complicating interaction
+with visual \acp{bci}
+\todo{FA}
+Patients from other neurodegenerative diseases like
+\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
+sometimes also interested in \ac{bci} use, but their eye motor capabilities
+mostly preserved.
+
+\newcommand\fnlis{\footnote{
+Multiple definitions of \ac{lis} are encountered in
+BCI and neurological literature.
+Some definitions include only tetraplegic patients without eye movements used
+for communications.
+Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
+including no eye motor control at all, from a \ac{lis} state with some preserved eye
+movements or minor motor output.
 While some definitions only include patients who suffered stroke or traumatic
-brain injury to the brain stem, it can also generally refer to the state of full body paralysis
-or loss of muscle tone, combined with the inability to speak, such as occurs in
-late stage ALS.
-In clinical reality, every tetraplegic patient has their
-unique set of preserved capabilities.}
+brain injury to specific regions in the brain (midbrain, brainstem,
+cerebellum),
+it can also generally refer to the state of full body paralysis
+or loss of muscle tone incurred in neurodegenerative diseases, combined with
+the inability to speak, such as occurs in late stage ALS.
+In clinical reality, every tetraplegic patient has their own unique set of
+preserved capabilities and from an solution-oriented \ac{bci} engineering point
+of view, the etiology of the symptoms can be abstracted away.
+}}
+
+Ultimately, patients with \ac{lis} from one of the main
+\ac{bci}.
+In this work, we'll define \ac{lis} patients as those a situation of (near)
+complete paralysis and difficulties or the inability to communicate without
+assistive technology, combined with severe or complete eye motor
+impairment\fnlis.
+\todo{from a solution engineering point of view, the etiology does not matter that
+much}.
 \todo{fill out which definition we use, wolpaw criteria}
 \todo{cite studies per pathology}
-Gaze-independent paradigms are particularly
-promising for individuals with impaired eye motor control, such as those
-suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
-Multiple Sclerosis (MS), (brain stem) stroke or Friedreich's Ataxia (FA)
+\todo{check definition in Wolpaw\&Wolpbaw Brain-Computer Interfaces: Principles
+and Practice}
+\Ac{lis} patients are are (nearly) completely paralyzed and have
+effort communicating without assistive technology.
 
@@
 \todo{insert table from riccio in text as numbers, not as table}
+\todo{true for non-invasive}
 
@@
 
-\subsection{Approach}
+\section{Approach}
 
-\subsubsection{Interface design}
+\subsection{Interface design}
 
@@
 
-\subsubsection{Decoders}
+\subsection{Decoders}
 
@@
 
-\subsubsection{Experiments}
+\subsection{Experiments}
 \todo{Healthy participant study}

### 2024-09-22 chapters/bci.tex
Add decoders

@@
 
+\todo{
 Locked-in -\> communicate
 Paraplegic -\> walk again
-Imagine that
 tetraplegic, locked-in syndrome
@@
 Start from wolpaw examples, for spelling
-Cite State of the art examples (walk again, speech avatar Metzger, handwriting
+Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
 willet, neuralink)
-
+}
 \todo{put focus on communication BCI}
@@
 
-\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci}loop.
+\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci} loop.
 The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
@@
 activity.
+\todo{preprocessing}
 This is commonly referred to as \emph{feature extraction}
@@
 machine learning.
-Section~\label{sec:bci-decoding} takes a closer look at some common and
-state-of-the art techniques.
-A machine learning algorithm can use some training data for which the performed
-tasks are known, and learn to recognize these tasks in new data.
-The training data can be either obtained from the \ac{bci} user themselves or from
-other users.
-In the first case, the user would perform a short calibration session before
-using the \ac{bci}, where they are instructed to perform a known task.
-This calibration session can be eliminated by training the decoder on
-preexisting labeled data from other sessions and users, but this is harder due
-to the variability between subjects and sessions.\todo{make shorter, move part
-to section}
+Useful information that is encoded in the brain signal by the paradigm,
+can be retrieved from it by the decoder.
+Section~\label{sec:bci-decoding} takes a closer look at some common decoding
+techniques.
 
-Finally, the loop can be closed by coupling this output can be coupled to a
+Finally, the loop can be closed by coupling this to a
 device or actuator.
@@
 
-\Ac{eeg} is a de facto non-invasive electrophysiology standard, in clinical
+\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
 neurology practice as well as in \ac{bci} research.
-Developed in 1924, it is cheap, relatively practical and can offers a
-temporal resolution.
+Developed in 1924, it is cheap, relatively practical and can offers great
+temporal resolution up to thousands of Hz.
+As it is easily applicable on the outside of the scalp, many practical
+consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
+caps or helmets.
+Current consumer systems can feature dry electrodes, but clinical and
+research systems often use electrolyte gel to reduce impedance for improved signal
+quality, which slightly decreases practicality.
+Nevertheless, it is the most accessible \ac{bci} technology for both patients
+and researchers.
 
-outside skull, environmental/electrical noise, volume conduction activity from
-everywhere in the brain.
+The major drawback of \ac{eeg} is its poor spatial resolution.
+The electrodes are large and sit on the outside of the scalp, relatively far
+away from the cortex.
+Furthermore, the layers of skull and cerebrospinal fluid in between can
+attenuate the signal and cause volume conduction, which results in electrodes
+measuring a mixture of activity from sources elsewhere in the brain.
+In addition to picking up brain activity other than that of the region of
+interest, \ac{eeg} also records other artifacts, from the
+environment, the power supply, nearby electrical equipment or the user's muscle
+movements.
+Combined, this means that \ac{eeg} is inherently noisy.
 
-cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
-conduction hence low spatial resolution
-what does ac{eeg} measure?
-
-cost, ease of use
+Despite its noisy nature \Ac{eeg} is the recording methodology of choice for
+our \ac{bci} because of its wide acceptance by patients.
+To counteract the noise present in our recording, we must on the one hand evoke
+strong, informative brain signals using a suited \ac{bci} \emph{paradigm}, and on th
+other hand pick a suited \emph{decoder} which can isolate the signal of
+interest and filter out the noise.
 
@@
 \label{sec:bci-paradigms}
-\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
+\fullpagefig{figures/bci/paradigms.eps}{A \ac{bci} paradigm defines which brain
+activity will be used for control. They can be categorized in \emph{active},
+\emph{reactive} and \emph{passive} paradigms, depending on their reliance on
+external stimulation and active participation required from the user.
+Adapted from~\textcite{Muehl2014}.
+}{fig:bci-paradigms}
+\todo{check premissions}
 \todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
@@
 neurofeedback
+error potentials
+cannot easily be used for fast communication
+synchronous <-> asynchronous
+
+\cite{Xu2021}
 
@@
 
-somatosensory:
+somatosensory evoked potentials (cite jimmy)
 
@@
 visual paradigms:
+more performant since it allows us to use our visual system, the most adapted
+sensory system in humans
 SSVEP, cVEP, mVEP, VEP/P300/Oddball
+high ssvep itr
+
+hybrid
 
@@
 \label{sec:bci-decoders}
-evaluated on
-information transfer rate,
-roc auc,
+
+After extracting features, machine learning algorithms can perform \ac{bci}
+decoding.
+This is usually based on
+some training data for which the performed tasks are known, by recognize the
+activity evoked by the task in recorded data.
+The training data can be either obtained from the \ac{bci} user themselves or from
+other users.
+In the first case, the user would perform a short calibration session before
+using the \ac{bci}, where they are instructed to perform a known task.
+This calibration session can be eliminated by training the decoder on
+preexisting labeled data from other sessions and users, but this is harder due
+to the variability between subjects and sessions.\todo{make shorter, move part
+to section}
+
+\todo{regression/classification}
+\todo{for communication, we want classification, focus on classification
+further down the line}
+
+If the goal is to eliminate the per-session calibration, one can also use a
+decoder pre-trained on an existing dataset of the same task.
+This is complicated by large variability in measured brain response between subjects
+and even between different sessions within the same
+subject~\cite{Guger2009,Saha2020}.
+Therefore, pre-trained decoders must either be trained on a very large, diverse
+sample of subjects, or some form of transfer learning must be applied.
+Nevertheless, in practice pre-trained models or models using transfer learning often still
+require some per session fine-tuning, which still necessitates some
+calibration.
+Instead, rather opt for keeping the calibration time as short as possible by using
+an algorithm that can learn from very few training samples.
+This can work well, but requires strong regularization~\cite{VanDenKerchove2022}.
+
+\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
+decoder algorithms.
+Classic linear methods, such as \ac{csp} feature extraction for
+motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
+and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
+relatively well, given some regularizing constraints or extensions.
+Multi-linear techniques exploiting the tensor structure of neural signals are
+also promising.
+Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy,
+Furthermore, they lend themselves well to application like adaptive
+learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}
+Riemannian Geometry classifiers are often considered the current
+state-of-the-art.
+Finally, deep learning\todo{Bhuvaneshwari2021} is also sometimes considered, albeit only when large
+training datasets are available.
+If decoders tailored to a specific user that keep the calibration session as
+short as possible are of interest, not enough training data is available to
+properly train a deep learning model.
+
+Usually, a decoder makes a prediction for a given \emph{trial} while operating
+the \ac{bci}.
+A trial is the smallest unit on which a selection decision can be made, for
+instance be one imagined movement, or one repetition of flashing all targets
+in a visual \ac{erp} \ac{bci}.
+Some valuable metrics to asses decoder performance in the classification case
+are accuracy,
+concept of trial
 accuracy
 
-Calibration, transfer learning, riemannian geometry, (multi)-linear techniques,
-paper fabien lotte
+single erp responses: roc auc, higher roc auc usually results in higher
+accuracy
 
-active/reactive/passive + infographic
+Finall, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
+in general, is \ac{itr}.
+\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
+be calculated as
+\begin{equation}
+	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
+\end{equation}
+\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
+different options that can be selected per trial, $P$, the selection accuracy
+of the decoder, and $Q$, the number of trials per second.
+The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
+building blocks a successful, high-\ac{itr} \ac{bci}.
+To improve performance, we can aim to
+\todo{enumerate}
+\begin{enumerate*}[label={\arabic*})]
+\item increase $N$ by selecting a paradigm and interface that offer a
+broad range of informative selections per trial,
+\item increase $P$ by engineering more performant machine learning methods for
+decoding, and
+\item increase $Q$ by selecting a paradigm and allows fast stimulation or
+responses.
+\end{enumerate*}
 
 \section{A case study: the visual oddball ERP paradigm}
-\subsection{stimulation}
+good choice because it scores high on the parameters above
+\subsection{Stimulation}
+\todo{figure of P3 matrix}
+\todo{cite P300 literature, first example and current sota}
 \subsection{The event-related potential}
-\todo{figure of P3 matrix}
 P300 and other components
+does not only rely on P300, but also heavily dependent on visual early
+components, hence we'll refer to it under the more generall oddball ERP
+paradigm
 \todo{figure of ERP components}
+\subsection{Preprocessing and feature extraction}
+referencing, filtering, epoching, baselining
 \subsection{Decoding ERPs}

### 2024-09-22 chapters/covert_align.tex
Add decoders

@@
+\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}}
+
 \section{Introduction}
+
+\cite{Arico2014} observed higher variability in single-trial P3 peak
+latencies relative to stimulus onset during covert VSA compared to overt VSA.
+This latency variability contributes to reduced covert VSA decoding performance.
+While they proposed an analysis and performance prediction method, they did not provide a
+decoding solution.
+They suggested that compensating for latency jitter could enhance covert VSA
+decoding, but did not actually verify this hypothesis directly.
+Additionally, \cite{hardiansyah2020single} developed a classifier for
+covert VSA ERPs, exploiting single-trial latency features in combination with
+amplitude features for classification with a support vector machine.
+They demonstrated the positive influence of single-trial ERP component latency
+features on covert VSA inference yet did not attempt to correct the amplitude
+features for these latencies.
+\todo{Should we already mention Frenzel here?}
+\cite{Frenzel2011} introduced a similar protocol to our split VSA
+setting.
+They showed that it is possible to perform split VSA and that, in this
+case, visuospatial attention and gaze direction can separately be decoded using
+classical ERP techniques.
+To the best of our
+knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
+However, \cite{Frenzel2011} considered their interface only for the case
+where the user actively intends to select both targets determined by the gaze
+and the VSA.
+In the split VSA setting considered in our work, we rather instruct the participant to ignore
+the distractor and only attend the cued target, since we are interested in
+decoding the visuospatial attention only.
 
@@
 
+\subsubsection{CVSA-ERP dataset}
+
 We recorded a dataset to validate our approach.
@@
 (S62547).
-\ref{app:dataset} details the stimulation and recording procedure.
 Each subject performed different VSA conditions (overt, covert and split),
@@
 
-\subsubsection{CVSA-ERP dataset}
 EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier

### 2024-09-22 chapters/gaze_independence.tex
Add decoders

@@
-\todo{mention partially published}
+\emph{Parts oare dissociated in the
+Section~\ref{sec:gaze-independence} have been published in
+\textcite{VanDenKerchove2024}.}
+
 \todo{spatially vs temporally organized interfaces}
@@
 \section{The gaze-dependence problem}
+\label{sec:gaze-dependence}
 
 \todo{define eye motor issues}
-\todo{systematic review of prevalences and prevalences of eye motor issues per disorder}
-\todo{
-cite{Guo2022}
-cite{Prasad2010}
-cite{Serra2018}
-cite{Bogousslavsky1987}
-cite{Castelnovo2016}
-cite{Polet2020}
-}
 
@@
 brain injury to specific regions in the brain (midbrain, brainstem,
-cerebellum),
+cerebellum)~\cite{Smith2005},
 it can also generally refer to the state of full body paralysis
@@
 
-Ultimately, patients with \ac{lis} from one of the main
-\ac{bci}.
+\newcommand\fnwolpawcrit{\footnote{
+\it``The first class consists of people who are truly totally locked-in (e.g.,
+due to end-stage ALS or severe cerebral palsy), who have no remaining
+useful neuromuscular control of any sort, including no eye movement.
+\elide\ This class is very small. \elide\
+The second class of potential BCI users comprises those who retain
+a very limited capacity for neuromuscular control. This group includes
+people who retain some useful eye movement or enough limb muscle
+function to operate a single-switch system. Such control is often slow,
+unreliable, or easily fatigued.
+This group is much larger than the first.
+\elide\
+The third class of potential BCI users, which is the largest of all,
+includes those who still retain (and can be expected to continue to
+retain) substantial neuromuscular control, particularly speech and/or
+hand control, and can, therefore, operate a wide range of assistive com-
+munication and control devices.''
+}}
+
+Together, some of these patients constitute the \ac{lis} patient group, which
+forms one of the main \ac{bci} target populations.
 In this work, we'll define \ac{lis} patients as those a situation of (near)
 complete paralysis and difficulties or the inability to communicate without
-assistive technology, combined with severe or complete eye motor
-impairment\fnlis.
-\todo{from a solution engineering point of view, the etiology does not matter that
-much}.
-\todo{fill out which definition we use, wolpaw criteria}
-\todo{cite studies per pathology}
-\todo{check definition in Wolpaw\&Wolpbaw Brain-Computer Interfaces: Principles
-and Practice}
-\Ac{lis} patients are are (nearly) completely paralyzed and have
-effort communicating without assistive technology.
+assistive technology, corresponding corresponds to classes one and two defined by
+\textcite{Wolpaw2006}\fnwolpawcrit.
+Additionally some degree of severe or complete eye motor impairment is ususally
+necessary to qualify as locked-in\fnlis.
+Since this work is concerned with gaze-impaired patients, we will adopt this
+interpretation.
+Case study reports~\cite{Patterson1986} show that the population of locked-in
+patients with complete eye motor paralysis is very small.
+Hence, it would be interesting to focus efforts in \ac{bci} development on the
+larger population of \ac{lis} patients that  currently slip through the cracks
+of the current assistive technology offer.
+These are the patients that require assistance to communicate independently,
+which is crucial to retaining an acceptable quality of life, but whose severe
+eye motor impairment prevents them from using eye tracking based solutions.
 
-Traditional visual BCI scenarios require the user to overtly direct their
-visuospatial attention (VSA) and gaze toward the screen target they intend to select.
-In most settings, screen targets are overlaid with non-overlapping, transient
-stimuli that evoke event-related EEG potentials (ERPs).
-The selected target can be decoded from these ERPs, as is the case for the
+\todo{order of concepts in next two paragraphs can be tweaked}
+This leads us to what we'll call the \emph{gaze-dependence problem in visual
+BCI}:
+Traditional visual BCI scenarios require the user to overtly direct their both
+\ac{vsa} and gaze toward the screen target they intend to select.
+For our purposes, screen targets are overlaid with non-overlapping, transient
+stimuli that evoke event-related  potentials \acp{erp} in the \ac{eeg}.
+The selected target can be decoded from these \acp{erp}, as is the case for the
 oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
@@
 In these cases, classical solutions often fall short of the widely accepted
-80\% target selection accuracy threshold deemed necessary for a comfortable user
-experience~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019,Neeling2019},
-calling for alternative, gaze-independent solutions.
+80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019} deemed necessary for a comfortable user
+experience~\cite{Neeling2019}, calling for alternative, gaze-independent
+solutions.
 
@@
 \section{Gaze-independent BCI paradigms}
-\label{sec:gaze-independence-paradigms}
+\label{sec:gaze-independence}
+
 Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
-ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
-do not rely on gaze redirection but often result in lower information transfer
-rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
+ways.
+Firstly, active BCI communication paradigms relying on endogenous activation from the user
+do not rely on sensory stimulation.
+Examples of this are imagined movement or imagined speech paradigms.
+These paradigms can yield very high information transfer
+rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
+complexity that can be captured in commands, but often only do so when paired
+with invasive recording.
+Non-visual reactive paradigms that use e.g. auditory and somatosensory
+stimulation do not rely on gaze redirection exist but also result in lower information transfer
+rates compared to visual paradigms.
+Furthermore, they can suffer from increased mental effort in operation and user-dependent variability~\cite{Reichert2020b}.
 \todo{insert table from riccio in text as numbers, not as table}
-\todo{true for non-invasive}
 
-Secondly, visual stimulation can be optimized, e.g. such that the stimuli are always
-present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
+Secondly, current visual stimulation paradigms can be optimized, e.g. such that
+the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
 Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
@@
 \todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
-\todo{
-Firstly, by avoiding visual
-stimulation entirely and opting for auditory or somatosensory stimulation instead.
-However, these alternatives often result in lower information transfer rates,
-increased mental effort and variable outcomes for different users.
-Secondly, gaze-independence can
-be achieved by adapting the interface to display stimuli such that they
-are always in the center of the field of view, by exploiting visual
-non-spatial attention (feature attention), or a combination thereof.
-These interfaces suffer equally from reduced information transfer rates.
-}
 
-Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial attention
-can be decoded separately from gaze direction.
-Earlier work has made advances in decoding lateralized covert
-VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
-decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
-stimulation paradigms~\cite{Egan2017}.
-Since these methods often require a slower stimulation pace than the classical
-ERP BCI paradigm, gaze-independent decoding in fast-paced ERP paradigm stimulation
-remains underexplored.
+Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial
+attention can be decoded separately from gaze direction.
 \cite{Aloise2012b} aimed to bridge the performance gap between covert and
@@
 decoding for any of the investigated decoders.
-\cite{Arico2014} observed higher variability in single-trial P3 peak
-latencies relative to stimulus onset during covert VSA compared to overt VSA.
-This latency variability contributes to reduced covert VSA decoding performance.
-While they proposed an analysis and performance prediction method, they did not provide a
-decoding solution.
-They suggested that compensating for latency jitter could enhance covert VSA
-decoding, but did not actually verify this hypothesis directly.
-Additionally, \cite{hardiansyah2020single} developed a classifier for
-covert VSA ERPs, exploiting single-trial latency features in combination with
-amplitude features for classification with a support vector machine.
-They demonstrated the positive influence of single-trial ERP component latency
-features on covert VSA inference yet did not attempt to correct the amplitude
-features for these latencies.
-\todo{Should we already mention Frenzel here?}
-\cite{Frenzel2011} introduced a similar protocol to our split VSA
-setting.
-They showed that it is possible to perform split VSA and that, in this
-case, visuospatial attention and gaze direction can separately be decoded using
-classical ERP techniques.
-To the best of our
-knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
-However, \cite{Frenzel2011} considered their interface only for the case
-where the user actively intends to select both targets determined by the gaze
-and the VSA.
-In the split VSA setting considered in our work, we rather instruct the participant to ignore
-the distractor and only attend the cued target, since we are interested in
-decoding the visuospatial attention only.
+More recent work has made advances in decoding lateralized covert
+VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
+decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
+stimulation paradigms~\cite{Egan2017}.
+Since these methods often require a slower stimulation pace than the classical
+ERP BCI paradigm, gaze-independent decoding in the fast-paced reactive ERP paradigm
+leveraging spatial attention remains underexplored.
 
-P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
-value is heavily dependent on the subject and the task, and can vary from trial
-to trial~\cite{Ouyang2017}.
-The work of \cite{Arico2014}
-illustrates that the variation in single-trial P3 latencies is important in
-gaze-independent decoding and has been hampering covert VSA decoding performance.
-In this work we aim to reprise their hypothesis stating that jitter compensation improves
-covert VSA performance and extend it by developing a decoder and evaluating it
-in a broader range of gaze-independent settings, including split VSA.
-\todo{Check connection}
-
-\todo{provisionary doctoral plan}
+\todo{check funding proposal}
+\todo{check provisional doctoral plan}
 \todo{check WIP seminar}
-\todo{update text}
 
@@
 \todo{refer to ITRs from previous chapter}
-Section{sec:gaze-independence-paradigms} showed that visual BCI paradigms form
+Section{sec:gaze-independence} showed that visual BCI paradigms form
 a good candidate for a performant gaze-independent BCI.
@@
 multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like Rapid Serial Visual Presentation,
+and temporally organized interfaces, like Rapid Serial Visual
+Presentation\todo{cite},
 where targets or small sets of targets are shown consecutively.
@@
 	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
-		can achieve similar or higher information transfer rate in patient
-		populations with eye motor deficits than other
-		gaze-independent alternatives described in literature.}
+		can achieve a similar transfer rate in patient
+		populations with eye motor deficits than other gaze-independent alternatives
+    described in literature.}
 \end{quote}
-\todo{tone this down a bit to what can be proven}
 
@@
 
-\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}
-\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
-their residual gaze capabilities, by leveraging a non-invasive high-ITR
-visuospatial ERP-based BCI and improving ERP decoding performance (in general
-and specifically in gaze-independent settings).}
+This specific hypothesis is embedded within a broader goal of enabling
+effective communication for eye-motor impaired \ac{lis} patients.
+Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
+their residual gaze capabilities, by leveraging a non-invasive, high-\ac{itr}
+BCI on the one hand, and on the other hand improving \ac{erp} decoding performance
+in general.
 
@@
 
+\todo{mention, free, split, frenzel}
 \todo{Overt, covert, split, free, illustration from poster}
-\todo{clear definitions}
+\todo{clear definitions, table from paper}
+
+First, we designed a visual oddball interface that helps us study and adapt
+to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
+decoders and on our own proposed decoders.
+Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
+\textcite{Treder2010}, we present six flashing circular targets
+(without letters or symbols) to the participant while the EEG and
+electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
+
+The interface can be operated in different VSA conditions as illustrated in
+Figure~\ref{fig:gaze-vsa}.
+\fourpanefig
+  {\figpane{figures/gaze_independence/attention_overt.eps}{
+    \emph{Overt \ac{vsa}}.
+    Gaze and \ac{vsa} coincide on a target.}{fig:gaze-vsa-overt}}
+  {\figpane{figures/gaze_independence/attention_covert.eps}{\emph{Covert
+  \ac{vsa}}.
+  Gaze rests on the center of the screen, while \ac{vsa} is directed towards a
+  target.}{fig:gaze-vsa-covert}}
+  {\figpane{figures/gaze_independence/attention_split.eps}{
+    \emph{Split \ac{vsa}}. \ac{vsa} is directed towards a target, while the
+    gaze rests on another.}{fig:gaze-vsa-split}}
+  {\figpane{figures/gaze_independence/attention_free.eps}{\emph{Free \ac{vsa}}.
+  The user is free to direct their gaze as they deem most comfortable
+  All previous \ac{vsa} conditions are possible.}{fig:gaze-vsa-free}}
+  {\Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.}
+  {fig:gaze-vsa}
+
+In the overt case, users gaze at the cued
+target they are also mentally attending; in the covert case, users
+gaze at the center of the screen while mentally attending to the cued target.
+Finally, we introduce a VSA condition that is understudied in the context of
+gaze-independent BCI development: split VSA.
+In split VSA, the user mentally focuses on one cued target while gazing at
+another (the distractor).
+
+The option to dissociate gaze and visuospatial attention will allow us to
+investigate the effect of gaze control on \ac{bci} performance.
 
@@
 \todo{Classifier-based Latency Estimation with Woody iterations}
+
+P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
+value is heavily dependent on the subject and the task, and can vary from trial
+to trial~\cite{Ouyang2017}.
+The work of \cite{Arico2014} illustrates that the variation in single-trial P3
+latencies is important in gaze-independent decoding and has been hampering covert
+VSA decoding performance.
+In this work we aim to reprise their hypothesis stating that jitter compensation improves
+covert VSA performance and extend it by developing a decoder and evaluating it
+in a broader range of gaze-independent settings, including split VSA.
+
+\todo{check overlap}
+
 Literature shows that better covert attention performance can be achieved by
@@
 \subsection{Experiments}
+
 \todo{Healthy participant study}
-We designed a visual oddball interface that helps us study and adapt
-to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
-decoders and on our own proposed decoders.
 In a first series of experiments, we recorded data with this interface from healthy participants.
@@
 (S62547).
-
-Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
-by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
-symbols) to the participant while the EEG and electro-oculogram (EOG) are
-recorded, as well as eye gaze using eye tracking.
 To simulate the dissociation between the eye gaze (visual attention) and the
@@
 motor control, healthy participants are cued to operate the BCI in specific
-visuospatial attention (VSA) conditions.
-Each subject performs 5 different VSA conditions as illustrated in
-Figure~\ref{fig:vsa-conditions}.
-\todo{do we already want this here?}
-\begin{figure}
-	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
-	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
-		asked to dissociate their visual and mental attention by gazing at a
-		crosshair while attending a target, depending on the prompted condition.}%
-	\label{fig:vsa-conditions}
-\end{figure}
-\todo{new picture that shows more clearly a user operating freely where they
-can gaze}
-In the overt case, subjects are instructed to gaze at the cued
-target they are also mentally attending; in the covert case, subjects are instructed
-to gaze at the center of the screen while mentally attending to the cued target.
-Finally, we introduce a VSA condition that is understudied in the context of
-gaze-independent BCI development: split VSA.
-In split VSA, the participant mentally focuses on one cued target while gazing at
-another (the distractor).
-Depending on the inter-target distance along the hull of the hexagon between the
-attended target and the distractor we discern three split VSA subconditions:
-the distractor is either clockwise or counterclockwise directly next to the
-attended target ($d=1$), there is one other target between the attended target and
-the distractor ($d=2$), or the distractor is opposite the intended target
-($d=3$).
+\ac{vsa} conditions.
 
+\todo{mention multi-center patient study case studies in cooperation
 After data collection from healthy participants and decoder development, we
@@
 our proposed decoding methods.
+}
 \todo{update with current results}
 
-\todo{Patient study}
 \todo{refer to chapters}

### 2024-09-22 chapters/wcble.tex
Add decoders

@@
+\todo{mention published}
 \section{Introduction}
+
+Single-trial ERP latency jitter introduces a problem in ERP analysis:
+the \textit{smearing effect} occurs when aggregating over multiple identical
+signals with different latencies.
+A familiar example of this happens when averaging over jittered epochs for ERP
+analysis.
+In the presence of jitter, the shape and amplitude of the average do not
+reflect the properties of the original signal.
+As the smearing effect reduces the amplitude of the average,
+it negatively impacts the signal-to-noise ratio (SNR).
+The smearing effect equally impacts the SNR of information captured by a
+classifier's parameters when training with a procedure that does not account for
+this jitter, and thus also affects its performance~\cite{Thompson2012}.
+Hence, effective latency estimation and jitter compensation can contribute to higher
+decoding performance.
+%In this work, we propose a latency estimation and alignment strategy based on
+%the method proposed by \cite{Thompson2012}.
 
@@
 
-\subsection{Existing methods for latency estimation and alignment}
+\section{Literature}
 
-\subsubsection{Single-component approaches}
+\subsection{Single-component approaches}
 The most straight-forward method to estimate single-trial latency
@@
 
-\subsubsection{Multi-component approaches}
+\subsection{Multi-component approaches}
 The abovementioned methods do not take into account the fact that  ERPs are multi-component signals.
@@
 
-\subsubsection{Contribution}
-In this work, we introduce a new latency method that leverages data-driven,
-joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
-integrates the spatiotemporal filtering in Woody's template matching iteration
-scheme and can easily be extended to multi-component settings.
-Our method attempts to model the spatial and temporal activation of
-each ERP component, as well as it's latency distribution, and takes into acount the
-spatial and temporal correlation of the EEG background noise.
-
-\section{Introduction}
-Single-trial ERP latency jitter introduces a problem in ERP analysis:
-the \textit{smearing effect} occurs when aggregating over multiple identical
-signals with different latencies.
-A familiar example of this happens when averaging over jittered epochs for ERP
-analysis.
-In the presence of jitter, the shape and amplitude of the average do not
-reflect the properties of the original signal.
-As the smearing effect reduces the amplitude of the average,
-it negatively impacts the signal-to-noise ratio (SNR).
-The smearing effect equally impacts the SNR of information captured by a
-classifier's parameters when training with a procedure that does not account for
-this jitter, and thus also affects its performance~\cite{Thompson2012}.
-Hence, effective latency estimation and jitter compensation can contribute to higher
-decoding performance.
-%In this work, we propose a latency estimation and alignment strategy based on
-%the method proposed by \cite{Thompson2012}.
+\todo{check overlap with:
 
@@
 
+}
+
+\subsection{Contribution}
+
 Most of these methods suffer from a common drawback: they cannot be used in a
@@
 
-\section{Materials and Methods}
-\subsection{Decoders}
-\subsubsection{Classifier-based Latency Estimation}
+\todo{remove lcmv}
+In this work, we introduce a new latency method that leverages data-driven,
+joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
+integrates the spatiotemporal filtering in Woody's template matching iteration
+scheme and can easily be extended to multi-component settings.
+Our method attempts to model the spatial and temporal activation of
+each ERP component, as well as it's latency distribution, and takes into acount the
+spatial and temporal correlation of the EEG background noise.
+
+\section{Methods}
+
+\subsection{Classifier-based Latency Estimation}
 \label{sec:cble}
@@
 
-\subsubsection{Robust CBLE latency features}
+\subsection{Robust CBLE latency features}
 \label{sec:robust-latency}
@@
 
-\subsubsection{Classifier-based Latency Estimation with Woody
+\subsection{Classifier-based Latency Estimation with Woody
 	Iterations}
@@
 positive, by the difference between the latency and the window onset.
-%\begin{figure}
-%	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
-%	\caption{Schematic representation of the Woody Classifier Based Latency
-%    Estimation training phase.
-%    (1) The first-stage spatiotemporal binary classifier is trained on a set of
-%    epochs.
-%    (2) It is	then applied to time-shifted copies of the epochs to obtain scores
-%    and class probabilities	over time.
-%    (3) The medians of these probability distributions are assumed the class
-%    latencies.
-%    (4) The epochs are aligned to their corresponding	class latencies by
-%    shifting in time such that all latencies fall at the same moment.
-%    (5) The spatiotemporal classifier is then retrained on the aligned epochs
-%    for a	next iteration.
-%    (6) After the iterative process halts, the scores and	latencies obtained
-%    from the last iteration are used to train the second-stage
-%		classifier.}
-%	\label{fig:diagram-train}
-%\end{figure}
 \fullpagefig{figures/wcble/figure1.pdf}{Schematic representation of the Woody Classifier Based Latency
@@
 \todo{re-insert comments}
+
 \begin{algorithm}[H]
@@
 
-\newpage
 \begin{algorithm}[H]
@@
 \end{algorithm}
+
+\section{Results}
+
+\section{Discussion}
+
+\section{Conclusion}

### 2024-09-22 chapters/bci.tex
Decoders section

@@
 Some valuable metrics to asses decoder performance in the classification case
-are accuracy,
-concept of trial
-accuracy
+are accuracy and \ac{rocauc}.
+Accuracy can be obtained from the correctness of predicted trial selection as
+\begin{equation}
+  \text{accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
+\end{equation}
+with TP and TN the numbers of true positives and negatives and FP and FP the
+numbers of false positives and negatives.
+\Ac{rocauc} is also a measure of classifier predictive power, but is more
+suited for evaluation of classification of single epochs of data and in the
+presence of imbalanced data, e.g. when
+comparing single \ac{erps}.
+Higer \ac{rocauc} usually translates to a higher target selection accuracy.
 
-single erp responses: roc auc, higher roc auc usually results in higher
-accuracy
-
-Finall, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
+Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
 in general, is \ac{itr}.

### 2024-09-22 chapters/gaze_independence.tex
Decoders section

@@
-\emph{Parts oare dissociated in the
-Section~\ref{sec:gaze-independence} have been published in
-\textcite{VanDenKerchove2024}.}
+\emph{Parts of Section~\ref{sec:gaze-independence} have been published in \textcite{VanDenKerchove2024}.}
 
@@
 
-\todo{Regularized spatiotemporal beamforming}
+\subsubsection{Regularized spatiotemporal beamforming}
 Due to the lack of the N2 component and decrease of amplitude of the
@@
 
-\todo{Classifier-based Latency Estimation with Woody iterations}
+\subsubsection{Classifier-based Latency Estimation with Woody iterations}
 
@@
 
-\todo{Tensor discriminant analysis}
+\subsubsection{Tensor discriminant analysis}
 
@@
 
-\todo{Healthy participant study}
+\subsubsection{Healthy control study}
 In a first series of experiments, we recorded data with this interface from healthy participants.
@@
 
+\subsubsection{Patient case study}
 \todo{mention multi-center patient study case studies in cooperation

### 2024-09-23 chapters/bci.tex
Started paradigms section

@@
 \todo{put focus on communication BCI}
-
 \section{A direct interface to the brain}
@@
 \label{sec:bci-recording}
+\todo{make sure every part ends with a motivation for why we chose our
+approach}
 
@@
 action potential spikes.
-They come in single-electrode form, or, more popularily, in microelectrode
+They come in single-electrode form, or, more popularly, in microelectrode
 arrays.
@@
 neurology practice as well as in \ac{bci} research.
-Developed in 1924, it is cheap, relatively practical and can offers great
-temporal resolution up to thousands of Hz.
+Developed in 1924~\cite{Berger1929}, it is cheap and relatively practical and
+can offer great temporal resolution up to thousands of Hz.
 As it is easily applicable on the outside of the scalp, many practical
@@
 \label{sec:bci-paradigms}
-\fullpagefig{figures/bci/paradigms.eps}{A \ac{bci} paradigm defines which brain
-activity will be used for control. They can be categorized in \emph{active},
-\emph{reactive} and \emph{passive} paradigms, depending on their reliance on
-external stimulation and active participation required from the user.
+
+The \ac{bci} paradigm~\cite{Xu2021} is the key to translating brain activity into useful
+output commands, since it defines which neural responses will be elicited and
+captured.
+A paradigm can be loosely seen as \emph{a way of interacting with the
+\ac{bci}}.
+According to \textcite{Zander2011}, they can be arranged by their reliance on
+external stimulation and the engagement of the user as shown in
+Figure~\ref{fig:bci-paradigms}.
+
+\fullpagefig{figures/bci/paradigms.eps}{
+  A \ac{bci} paradigm defines which brain activity will be used for control.
+  They can be categorized in \emph{active}, \emph{reactive} and \emph{passive}
+  paradigms, depending on their reliance on external stimulation and active
+  participation required from the user.
   Adapted from~\textcite{Muehl2014}.
 }{fig:bci-paradigms}
-\todo{check premissions}
+
+Paradigms that require little active participation of the user like
+redirecting attention or initiating imagined actions are classified as passive
+\ac{bci}.
+From the user's point of view, this concept is probably the most attractive,
+since their intentions or cognitive state state would be inferred from just
+their brain activity.
+However, without active user participation, it can be hard to establish reliable
+communication or control.
+Therefore, passive \ac{bci} are currently more suited to tasks like
+emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
+monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
+detection~\cite{SiMohammed2020}.
+
+Furthermore, active paradigms make it easier to quickly collect training data for
+supervized machine learning since the conditions can be more easily controlled
+by the user or the stimulation.
+
+active
+
+reactive
+works for most people
+high effort, unintuitive and low expressive power compared to movement and
+speech~\cite{Zander2011}, which are more natural and can be emulated with active bci
+
+\todo{check permissions}
 \todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
@@
 synchronous <-> asynchronous
-
-\cite{Xu2021}
+Again, asynchronous is preferred by users because it is more intuitive and
+comfortable to use, but it is far easier to decode synchronous responses since
+it is known when to expect a response and when a decision should be made.
 
@@
 
-somatosensory evoked potentials (cite jimmy)
+somatosensory evoked potentials~\cite{Petit2021} (cite jimmy)
 
@@
 presence of imbalanced data, e.g. when
-comparing single \ac{erps}.
+comparing single \acp{erp}.
 Higer \ac{rocauc} usually translates to a higher target selection accuracy.
@@
 \subsection{Stimulation}
+comfortable, preferred over SSVEP, similar ITR is this true?
 \todo{figure of P3 matrix}

### 2024-09-23 chapters/cv.tex
Recording technologies
Started paradigms section

@@
 processing, and user interface design with insights from neurophysiology, my
-PhD centers around the development of an event-related potential BCI system
+Ph.D. centers around the development of an event-related potential BCI system
 using EEG for real-time communication. I strongly believe in the potential of
@@
 	\cvitem{LDA in	the combined
-		space-time-frequency domain\\ for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
+		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
 		2023--2023}

### 2024-09-23 chapters/gaze_independence.tex
Started paradigms section

@@
 
-One of the goals of Brain-computer interfacing is establishing a communication
+\todo{Algemeen: in begin uitleggen dat de ziektes er niet veel toe doen en dat
+het gaat over verlamde personen, verlies van spierkracht en coördinatie en dus
+gehinded in communicatie, helemaal in begin van chapter 1 duidelijk maken dat
+het om deze patiënten gaat}
+One of the goals of brain-computer interfacing is establishing a communication
 channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
@@
 
-In fact, many patients in the BCI target population can suffer to some extent
+In fact, many patients in the BCI target population suffer to some extent
 from eye motor disability, requiring BCIs adapted to their condition.
@@
 \footnote{
-These can include square-wave jerks, saccadic intrusions, microtremors or
+These can include square-wave jerks, saccadic intrusions, microtremors, or
 microsaccades while resting or fixating the gaze.
@@
 Often, a specific movement direction (up-down, left-right) is preserved.
-}, involuntary movements, impaired pursuit, \ldots) and even complete
+}, involuntary movements, impaired pursuit, \ldots), and even complete
 ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
-This not only has an effect on the patients' vision and coordination, but also
+This not only has an effect on vision and coordination but also
 on their ability to operate a visual \ac{bci}.
@@
 	\begin{tabular}{@{}l|rrrrr|r@{}}
-              & \bfseries ALS  & \bfseries MS   & \bfseries Stroke  &\bfseries DMD &\bfseries SMA  &\bfseries LiS  \\ \hline
+    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
 		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
@@
 	\caption{Incidence of eye motor impairment in selected BCI target
-		patient populations. (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
-		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
-		Locked-In-Syndrome). $+$: frequent, $-$: infrequent.}\label{tab:incidence}
+    patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
+    \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
+    $+$: frequent, $-$: infrequent.}
+    \label{tab:incidence}
 \end{table}
@@
 Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
-due to brain-stem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
+due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
 \todo{summarize and cite Rowe2019}
-These patients can suffer from severe or complete eye motor impairment immediately
+These patients might suffer from severe or complete eye motor impairment
 immediately from the onset of their condition.
@@
 especially for patients with bulbar onset \ac{als}.
-Furthermore \textcite{Hayashi1991} show that, as these patients are supported to beyond
+Furthermore, \textcite{Hayashi1991} show that, as these patients are supported beyond
 respiratory failure, the disease progression will eventually also involve
@@
 However, \ac{ms} often comes with vision loss, further complicating interaction
-with visual \acp{bci}
+with visual \acp{bci}.
 \todo{FA}
-Patients from other neurodegenerative diseases like
+Patients suffering from other neurodegenerative diseases like
 \ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
-sometimes also interested in \ac{bci} use, but their eye motor capabilities
+sometimes also interested in \ac{bci} use, but their eye motor capabilities are
 mostly preserved.
@@
 While some definitions only include patients who suffered stroke or traumatic
-brain injury to specific regions in the brain (midbrain, brainstem,
+brain injury to specific regions in the brain (midbrain, brainstem, or
 cerebellum)~\cite{Smith2005},
@@
 retain) substantial neuromuscular control, particularly speech and/or
-hand control, and can, therefore, operate a wide range of assistive com-
-munication and control devices.''
+hand control, and can, therefore, operate a wide range of assistive
+communication and control devices.''
 }}
@@
 forms one of the main \ac{bci} target populations.
-In this work, we'll define \ac{lis} patients as those a situation of (near)
+In this work, we'll define \ac{lis} patients as those in a situation of (near)
 complete paralysis and difficulties or the inability to communicate without
-assistive technology, corresponding corresponds to classes one and two defined by
+assistive technology
+This corresponds to classes one and two defined by
 \textcite{Wolpaw2006}\fnwolpawcrit.
-Additionally some degree of severe or complete eye motor impairment is ususally
+Additionally, some degree of severe or complete eye motor impairment is usually
 necessary to qualify as locked-in\fnlis.
 Since this work is concerned with gaze-impaired patients, we will adopt this
-interpretation.
+interpretation into our definition.
 Case study reports~\cite{Patterson1986} show that the population of locked-in
 patients with complete eye motor paralysis is very small.
-Hence, it would be interesting to focus efforts in \ac{bci} development on the
-larger population of \ac{lis} patients that  currently slip through the cracks
-of the current assistive technology offer.
-These are the patients that require assistance to communicate independently,
-which is crucial to retaining an acceptable quality of life, but whose severe
-eye motor impairment prevents them from using eye tracking based solutions.
+Hence, it would be interesting to focus \ac{bci} development efforts on the
+larger population of \ac{lis} patients with severe eye motor impairment that
+currently slip through the cracks of the assistive technology offer.
+These are the patients whose severe eye motor impairment prevents them from
+using eye tracking based solutions.
+They currently use their remaining motor control to
+communicate by indicating symbols on a letterboard with great effort,
+or signal with upwards eye movements or blinks to confirm prompted letters.
+They require a caregiver or relative to interpret their signals, and crave
+the ability to communicate independently which is crucial to retaining an
+acceptable quality of life.
+\todo{order of concepts in next two paragraphs can be tweaked}
 
-\todo{order of concepts in next two paragraphs can be tweaked}
 This leads us to what we'll call the \emph{gaze-dependence problem in visual
@@
 
-This leads to an often unadressed paradox: the BCI target population consists
-of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
-(ALS), Multiple Sclerosis (MS), stroke (specifically
-brain-stem stroke) or traumatic brain injury, which can lead to several degrees
-of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
+This leads to the often unadressed gaze-dependence paradox: the BCI target
+population consists of patients who suffer from various degrees
+of paralysis, or even from Locked-in Syndrome (LiS), the complete loss of muscle
 control with preserved consciousness.

### 2024-09-23 chapters/bci.tex
Comments Liesel

@@
 Devices processing direct inputs from the central nervous system can be
-useful for rehabilitation, medical diagnosis or treatment.
+useful for rehabilitation and medical diagnosis or treatment.
 They have always been especially promising as assistive technology for
@@
 regions that are more or less active at a given time with a scanner.
-However, this signal reacts to slowly to reflect the real-time activity that is
+However, this signal reacts too slowly to reflect the real-time activity that is
 of interest for our purposes, and it carries little information other than the
@@
 `background tasks', all of which generate their own electrical signals.
-On the other hand, the brain activity we are interested in exploiting often
-originates from a specific functional system within the brain, and its activity
-might not be easily discerned from all other neural and environmental activity,
-akin to trying to listen to a single speaker in a noise environment where
+On the other hand, the activity of interest in often
+originates from a specific functional system within the brain.
+This activity might not easily be discerned from all other neural and
+environmental activity.
+Compare this to trying to listen to a single speaker at a noisy party where
 everyone is shouting over each other.
-Some obvious sources of interference can be easily filtered out, but to obtain
-a good measure reflecting only the activity that is relevant for a given \ac{bci}
-problem.
-This problem is called \emph{decoding} and is often solved through supervised
-machine learning.
+Some obvious sources of interference can be easily filtered out by
+\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
+relevant signal is often solved through supervised machine learning.
 Useful information that is encoded in the brain signal by the paradigm,
@@
 with their environment.\todo{put focus on communication BCI}
-On the other hand, the actions performed by the \ac{bci} can themselves influence
-the user's brain activity, creating an adaptive system.
+On the other hand, the actions performed by the \ac{bci} can themselves alter
+the user's cognitive state or brain activity, creating an adaptive system.
 A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
 Examples of this are deep brain stimulation to mitigate the symptoms of
-Parkinson's disease, or the cochlear implant as an auditive prosthesis.
-More indirectly modulation can be achieved by sensory stimulation.
-This can involve specific, paradigm-related sensory input, like haptic feedback when a
-movement \ac{bci} detects an intended action, or simply presenting selected actions
+Parkinson's disease, or the cochlear implant as a hearing prosthesis.
+More indirectly, modulation can be achieved by sensory stimulation.
+This involves specific, paradigm-related sensory input, like haptic selection
+feedback for movement \ac{bci}, or simply presenting selected actions
 back to the user.
@@
 changes in behavior or strategy.
-Due to neuroplasticity, the brain's ability to adapt, this can have a positive
-impact on \ac{bci} performance and opens the pathway to rehabilitation applications.
+Neuroplasticity, the brain's ability to adapt and reinforce specific neural
+pathways, can have a positive impact on \ac{bci} performance and gives rise to
+rehabilitation applications.
 
@@
 (fNIRS)~\cite{Borgheai2020}.
-These technologies all measure the hemodynamic signal, which only indirectly
+The technologies above all measure the hemodynamic signal, which only indirectly
 reflects brain activity.
-Because of the desirable high temporal resolution of
-electrophysiology~\cite{Easttom2021} and
-because of their practicality, we will focus only on methods to record
+Because of the desirable high temporal resolution~\cite{Easttom2021} and
+practicality electrophysiology, we will focus only on methods to record
 electromagnetical activity originating from neuronal action potentials.
@@
 cortex are more popular, as they are considered less invasive.
-\ac{ecog} is a powerful method
+\Ac{ecog} is a powerful method
 with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
@@
 
-\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
+\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
 neurology practice as well as in \ac{bci} research.
-Developed in 1924~\cite{Berger1929}, it is cheap and relatively practical and
-can offer great temporal resolution up to thousands of Hz.
+Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
+offer great temporal resolution up to thousands of Hz.
 As it is easily applicable on the outside of the scalp, many practical
-consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
+consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
 caps or helmets.
-Current consumer systems can feature dry electrodes, but clinical and
+Current consumer systems sometimes feature dry electrodes, but clinical and
 research systems often use electrolyte gel to reduce impedance for improved signal
@@
 measuring a mixture of activity from sources elsewhere in the brain.
+The noise problem mentioned earlier is also prominently present in \ac{eeg}
+recordings.
 In addition to picking up brain activity other than that of the region of
-interest, \ac{eeg} also records other artifacts, from the
-environment, the power supply, nearby electrical equipment or the user's muscle
-movements.
-Combined, this means that \ac{eeg} is inherently noisy.
+interest, \ac{eeg} also records other artifacts, from the environment, the power
+supply, nearby electrical equipment, or the user's muscle movements.
+Combined, this means \ac{eeg} is inherently noisy.
 
-Despite its noisy nature \Ac{eeg} is the recording methodology of choice for
+Despite its noisy nature, \ac{eeg} is the recording methodology of choice for
 our \ac{bci} because of its wide acceptance by patients.
 To counteract the noise present in our recording, we must on the one hand evoke
-strong, informative brain signals using a suited \ac{bci} \emph{paradigm}, and on th
-other hand pick a suited \emph{decoder} which can isolate the signal of
-interest and filter out the noise.
+strong, informative brain signals using a suited \ac{bci} paradigm, and on the
+other hand pick a suited decoder which can isolate the signal of interest and
+filter out the noise.
 
-\section{\ac{bci} Paradigms}
+\section{\ac{bci} paradigms}
 \label{sec:bci-paradigms}
@@
 \section{Decoders}
-\label{sec:bci-decoders}
+\label{sec:bci-decoding}
 
-After extracting features, machine learning algorithms can perform \ac{bci}
+After preprocessing and extracting features, machine learning algorithms can perform \ac{bci}
 decoding.
 This is usually based on
-some training data for which the performed tasks are known, by recognize the
+some training data for which the performed tasks are known, by recognizing the
 activity evoked by the task in recorded data.
@@
 also promising.
-Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy,
+Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
 Furthermore, they lend themselves well to application like adaptive
-learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}
+learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}.
 Riemannian Geometry classifiers are often considered the current
 state-of-the-art.
-Finally, deep learning\todo{Bhuvaneshwari2021} is also sometimes considered, albeit only when large
-training datasets are available.
+Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered,
+albeit only when sufficiently large training datasets are available.
 If decoders tailored to a specific user that keep the calibration session as
@@
 are accuracy and \ac{rocauc}.
-Accuracy can be obtained from the correctness of predicted trial selection as
-\begin{equation}
-  \text{accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
-\end{equation}
-with TP and TN the numbers of true positives and negatives and FP and FP the
-numbers of false positives and negatives.
+Accuracy is calculated as the proportion of correct selections to all
+selections made.
+It should be carefully interpreted in the presence of imbalanced data and always compared to the
+random chance accuracy level in the presence of more than two possible
+selections per trial.
 \Ac{rocauc} is also a measure of classifier predictive power, but is more
 suited for evaluation of classification of single epochs of data and in the
-presence of imbalanced data, e.g. when
-comparing single \acp{erp}.
-Higer \ac{rocauc} usually translates to a higher target selection accuracy.
+presence of imbalanced data, e.g. when comparing single \acp{erp}.
+Higher \ac{rocauc} usually translates to a higher target selection accuracy.
 

### 2024-09-23 chapters/bci.tex
started wcble introduction

@@
 measuring a mixture of activity from sources elsewhere in the brain.
+This also has a negative effect on spectral resolution.
 The noise problem mentioned earlier is also prominently present in \ac{eeg}
@@
 Combined, this means \ac{eeg} is inherently noisy.
+\todo{citations from 'Decoding covert speech from eeg-a cromprehensive review
+EEG has lower signal-to-noise ratio (SNR) than the other
+modalities. It is almost always corrupted by artifacts such as
+muscular artifacts (Eberle et al., 2012; Liu, 2019).
+• EEG has limited spectral and spatial resolution (Peled et al.,
+2001; Lakshmi et al., 2014).
+• Recording EEG for longer duration is challenging since the
+conductive gel or the saline solution applied for reducing the
+electrode impedance dries up over time, thus increasing the
+electrode impedance (Guger et al., 2012; Xu et al., 2018a).
+• A trained personnel is required for placing the EEG
+electrode cap}
 
@@
 
-The \ac{bci} paradigm~\cite{Xu2021} is the key to translating brain activity into useful
+The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity into useful
 output commands, since it defines which neural responses will be elicited and
-captured.
+captured as features.
 A paradigm can be loosely seen as \emph{a way of interacting with the
@@
   A \ac{bci} paradigm defines which brain activity will be used for control.
-  They can be categorized in \emph{active}, \emph{reactive} and \emph{passive}
+  They can be categorized in active, reactive and passive
   paradigms, depending on their reliance on external stimulation and active
@@
 }{fig:bci-paradigms}
+\todo{check permissions}
 
 Paradigms that require little active participation of the user like
-redirecting attention or initiating imagined actions are classified as passive
-\ac{bci}.
+redirecting attention or initiating imagined actions are classified as \emph{passive
+\ac{bci}}.
 From the user's point of view, this concept is probably the most attractive,
@@
 monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
-detection~\cite{SiMohammed2020}.
-
+detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
 Furthermore, active paradigms make it easier to quickly collect training data for
@@
 
-active
+Paradigms with high active participation can be split up in \emph{active
+\ac{bci}} and \emph{reactive \ac{bci}}.
+Active \ac{bci} paradigms encode endogenous activity initiated by the user,
+such as imagined movement or imagined speech.
+These tasks can encode complex information, but current non-invasive
+communication methods often limit the considered domain to a few movement
+directions to control a cursor, or a few words~\cite{Panachakel2021}.
+Invasive recordings are necessary for decoding more complex encoded
+like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
+symbols~\cite{Willett2021} or sign language.
+Active paradigms are also subject to large inter-subject variability due to the
+complexity of the performed tasks.
+Imagined speech or movements can be performed in a multitude of ways, which
+can each have different neural representations for separate individuals.
+This can make it hard to adapt the \ac{bci} to specific individuals, causing
+poor performance and giving rise to the concept of \ac{bci}
+illiteracy~\cite{Allison2010}\footnote{
+  \cite{Becker2022,Thompson2019}
+}.
+\todo{footnote on how this just means it needs to be more adapted,
+inclusivity, harmful to progress, paraphrase abstracts of articles}
 
-reactive
-works for most people
-high effort, unintuitive and low expressive power compared to movement and
-speech~\cite{Zander2011}, which are more natural and can be emulated with active bci
+Reactive \ac{bci} take another approach, by giving the user a discrete set of
+sensory stimulations and decoding towards which stimulus the user's attention
+was directed.
+Compared to active \ac{bci}, reactive paradigms usually require higher effort
+and can induce fatigue due to the constant stimulation.
+These paradigms are also somewhat less intuitive compared to e.g. speech, and
+their expressive power is limited by how many different stimuli can effectively
+presented and attended within a reasonable timeframe.
+Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
+importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.
 
-\todo{check permissions}
-\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
-paradigms}
-passive:
-also monitoring for diagnostic purposes
-emotion recognition
-neurofeedback
-error potentials
-cannot easily be used for fast communication
-synchronous <-> asynchronous
+Reactive \ac{bci} can be realized in multiple ways, depending on the
+stimulation used.
+Some examples include the following:
+In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021},
+the user can attend to one of multiple vibrotactile stimulations in different
+limbs, which encodes the information of the attended limb in the brain signal.
+In auditory paradigms, information can be modulated by the volume, tone, pitch,
+spatial origin of presented sounds, to which the user can
+attend~\cite{Kaongoen2017}.
+Most common are the visual paradigms.
+These can be more performant since they allows the user to use their visual
+system, one of the most adapted sensory system in humans.
+
+The major visual paradigms comprise
+\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} type interfaces.
+In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
+among many that each flicker continuously at different
+frequencies~\cite{Chen2021}.
+In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
+be correlated with the brain activity to retrieve the attended stimulus.
+Instead of stimulating all possible selections at once, the oddball paradigm
+stimulates them one by one with single flashing.\todo{reference}
+Since the times of stimulation of each target is known by the interface,
+information is modulated by time of stimulation, and times where an attention
+response can be found correspond to selected targets.
+\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
+movements in specific directions instead of flashing.
+Information is then also carried by movement
+direction~\cite{Libert2021a,Libert2022}.
+
+The \ac{bci} paradigms mentioned above can also be combined to gain performance.
+Straightforward examples are activating or deactivating \ac{ssvep} stimulation
+with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
+along multiple dimensions at once.
+\textcite{Han2023} recently used this strategy, combining frequancy and phase
+coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
+efficient \ac{bci} where one of 200 targets can be accurately selected with
+only 800ms of stimulation.
+Finally, paradigms can be implemented in a \emph{synchronous} or
+\emph{asynchronous} manner.
+Asynchronous \ac{bci} constantly decodes the signal looking for the response of
+interest, while in synchronous \ac{bci} decisions are made (and eventual
+stimulation takes place) at predetermined time points.
 Again, asynchronous is preferred by users because it is more intuitive and
 comfortable to use, but it is far easier to decode synchronous responses since
-it is known when to expect a response and when a decision should be made.
-
-active:
-motor imagery
-imagined speech
-
-reactive:
-
-somatosensory evoked potentials~\cite{Petit2021} (cite jimmy)
-
-auditory:
-ERP/oddball, steady state
-
-visual paradigms:
-more performant since it allows us to use our visual system, the most adapted
-sensory system in humans
-SSVEP, cVEP, mVEP, VEP/P300/Oddball
-high ssvep itr
-
-hybrid
+it is known when to expect a response and if a decision should be made.
 
@@
 
-\section{A case study: the visual oddball ERP paradigm}
+\section{A case study: the visual oddball paradigm}
+
+The visual oddball paradigm is a reactive, stimulus-dependent, synchronous
+\ac{bci} paradigm.
 good choice because it scores high on the parameters above
+well established, good research, first application cite
+Flashing in various configurations, such as row-colum or in groups,
+
 \subsection{Stimulation}
@@
 \todo{cite P300 literature, first example and current sota}
+
 \subsection{The event-related potential}
-P300 and other components
-does not only rely on P300, but also heavily dependent on visual early
-components, hence we'll refer to it under the more generall oddball ERP
-paradigm
+
+Because of this, the oddball paradigm is often referred to as the P300
+paradigm.
+However, due to equal importance of other \ac{erp} components than the P3 in
+decoding as we will show later, this could be considered a bit of a misnomer
+and we will stick to the oddball paradigm naming.
+
+P300 and other components, explain function (Luck)
+
 \todo{figure of ERP components}
+
+\subsubsection{P1}
+\subsubsection{N1}
+\subsubsection{P2}
+\subsubsection{N2}
+\subsubsection{P3}
+\subsubsection{Late negativity}
+
 \subsection{Preprocessing and feature extraction}

### 2024-09-23 chapters/gaze_independence.tex
started wcble introduction

@@
-\emph{Parts of Section~\ref{sec:gaze-independence} have been published in \textcite{VanDenKerchove2024}.}
+\emph{Parts of Section~\ref{sec:gaze-independence} have been published by \textcite{VanDenKerchove2024}.}
 
@@
 
-In fact, many patients in the BCI target population suffer to some extent
-from eye motor disability, requiring BCIs adapted to their condition.
+In fact, many patients in the \ac{bci} target population suffer to some extent
+from eye motor disability, requiring \ac{bci}s adapted to their condition.
 Table~\ref{tab:incidence} reports the relatively high frequencies of
@@
 	\end{tabular}
-	\caption{Incidence of eye motor impairment in selected BCI target
+	\caption{Incidence of eye motor impairment in selected \ac{bci} target
     patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
@@
 Multiple definitions of \ac{lis} are encountered in
-BCI and neurological literature.
+\ac{bci} and neurological literature.
 Some definitions include only tetraplegic patients without eye movements used
@@
 \elide\ This class is very small. \elide\
-The second class of potential BCI users comprises those who retain
+The second class of potential \ac{bci} users comprises those who retain
 a very limited capacity for neuromuscular control. This group includes
@@
 \elide\
-The third class of potential BCI users, which is the largest of all,
+The third class of potential \ac{bci} users, which is the largest of all,
 includes those who still retain (and can be expected to continue to
@@
 This leads us to what we'll call the \emph{gaze-dependence problem in visual
-BCI}:
-Traditional visual BCI scenarios require the user to overtly direct their both
+\ac{bci}}:
+Traditional visual \ac{bci} scenarios require the user to overtly direct their both
 \ac{vsa} and gaze toward the screen target they intend to select.
@@
 The selected target can be decoded from these \acp{erp}, as is the case for the
-oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
+oddball paradigm where observing a rare but attended stimulus evokes a P3 \ac{erp}
 component.
@@
 In these cases, classical solutions often fall short of the widely accepted
-80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019} deemed necessary for a comfortable user
+80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,RonAngevin2019} deemed necessary for a comfortable user
 experience~\cite{Neeling2019}, calling for alternative, gaze-independent
@@
 gaze.'}
-In the context of a visual BCI, this means that the user's
+In the context of a visual \ac{bci}, this means that the user's
 visuospatial attention and their gaze do not necessarily coincide.
@@
 them.
-Consequently, while BCIs hold great promise for these individuals, conventional
-gaze-dependent BCI solutions do not meet their needs due to the absence of gaze
+Consequently, while \ac{bci}s hold great promise for these individuals, conventional
+gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
 control. Therefore, the development of decoding strategies that account for covert
 VSA becomes crucial in the pursuit of high-performance gaze independent
-BCIs.
+\ac{bci}s.
 
-\section{Gaze-independent BCI paradigms}
+\section{Gaze-independent \ac{bci} paradigms}
 \label{sec:gaze-independence}
 
-Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
+Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
 ways.
-Firstly, active BCI communication paradigms relying on endogenous activation from the user
+Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
 do not rely on sensory stimulation.
@@
 attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
-Alternative visual stimulation paradigms can modulate specific ERP components
+Alternative visual stimulation paradigms can modulate specific \ac{erp} components
 that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
@@
 
-Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial
+Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
 attention can be decoded separately from gaze direction.
-\cite{Aloise2012b} aimed to bridge the performance gap between covert and
+\textcite{Aloise2012} aimed to bridge the performance gap between covert and
 overt VSA decoding performance.
-They compared classical linear and non-linear ERP classifiers on a covert
-attention P3 ERP component dataset.
+They compared classical linear and non-linear \ac{erp} classifiers on a covert
+attention P3 \ac{erp} component dataset.
 The results revealed no significant performance improvement in covert VSA
@@
 More recent work has made advances in decoding lateralized covert
-VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
+VSA harnessing the N2pc \ac{erp} component~\cite{Thiery2016,Reichert2020b,Wang2022},
 decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
@@
 Since these methods often require a slower stimulation pace than the classical
-ERP BCI paradigm, gaze-independent decoding in the fast-paced reactive ERP paradigm
+\ac{erp} \ac{bci} paradigm, gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
 leveraging spatial attention remains underexplored.
@@
 
-Brain-Computer Interfaces (BCIs) decode brain activity with the aim to establish a direct communication
+Brain-Computer Interfaces (\ac{bci}s) decode brain activity with the aim to establish a direct communication
 pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
-devoid of these abilities, for whom BCIs can provide a means to communicate or to
+devoid of these abilities, for whom \ac{bci}s can provide a means to communicate or to
 control devices.
 The quest for performant and affordable solutions is most evident in
-visual BCIs based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
+visual \ac{bci}s based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
 and paradigms tailored to the needs of specific stimulation paradigms and
 use contexts.
-An effective and proven method is the visual event-related potential (ERP)-based
+An effective and proven method is the visual event-related potential (\ac{erp})-based
 oddball interface.
 Here, targets are shown in short pulses on a computer screen.
-Each time the user observes a target, an ERP is evoked, the presence of which can be
+Each time the user observes a target, an \ac{erp} is evoked, the presence of which can be
 detected in the EEG-signal.
-The ERP consists of multiple components, of which some are modulated by the
+The \ac{erp} consists of multiple components, of which some are modulated by the
 attention of the user, specifically the N2 and the P3 component.
@@
 user's brain activity, as sketched in Figure~\ref{fig:bci}.
-However, visual BCIs cannot operate efficiently when the user does not direct
+However, visual \ac{bci}s cannot operate efficiently when the user does not direct
 their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
 
-This leads to the often unadressed gaze-dependence paradox: the BCI target
+This leads to the often unadressed gaze-dependence paradox: the \ac{bci} target
 population consists of patients who suffer from various degrees
@@
 control with preserved consciousness.
-While BCIs are most attractive as a solution for these patients
+While \ac{bci}s are most attractive as a solution for these patients
 compared to other assistive technologies like eye-tracking, studies often
@@
 control, or patients with gaze impairments are excluded.
-Different degrees of eye motor impairment can render visual BCIs
+Different degrees of eye motor impairment can render visual \ac{bci}s
 uncomfortable or outright impossible to use.
-When operating a visual BCI, the usual rapid series of forced saccades followed
+When operating a visual \ac{bci}, the usual rapid series of forced saccades followed
 by fixation is tiring over time, even for healthy control subjects.
 A suitable alternative would allow the user to keep their eyes in an at-rest
-position of their choice while operating the BCI.
-This points to the need for gaze-independent BCIs.
+position of their choice while operating the \ac{bci}.
+This points to the need for gaze-independent \ac{bci}s.
 
-The general goal of this PhD is to tackle the gaze-dependence in visual BCIs.
+The general goal of this PhD is to tackle the gaze-dependence in visual \ac{bci}s.
 \todo{refer to ITRs from previous chapter}
-Section{sec:gaze-independence} showed that visual BCI paradigms form
-a good candidate for a performant gaze-independent BCI.
+Section{sec:gaze-independence} showed that visual \ac{bci} paradigms form
+a good candidate for a performant gaze-independent \ac{bci}.
 While the central gaze fixation of most visual gaze-independent paradigms still
@@
 rate, but are also more gaze-dependent than temporally organized interfaces.
-In such an interface, stimuli can be presented in a standard BCI paradigm, but
+In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
 visuospatial attention can be decoded separately from the position of gaze,
@@
 This leads us to adopt a specific approach: improve the performance of a
-spatially organized visual oddball ERP-based brain
+spatially organized visual oddball \ac{erp}-based brain
 computer interface by using a suited decoding strategy.
@@
 \begin{quote}
-	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
+	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
 		can achieve a similar transfer rate in patient
@@
 and we collect data from healthy control subjects, which allow us to test our decoders.
-Finally, the findings of these experiments will be leveraged to develop a BCI
+Finally, the findings of these experiments will be leveraged to develop a \ac{bci}
 for patient use, which will be tested in a case-based patient study in which
@@
 their residual gaze capabilities, by leveraging a non-invasive, high-\ac{itr}
-BCI on the one hand, and on the other hand improving \ac{erp} decoding performance
+\ac{bci} on the one hand, and on the other hand improving \ac{erp} decoding performance
 in general.
@@
 First, we designed a visual oddball interface that helps us study and adapt
-to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
+to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
 decoders and on our own proposed decoders.
@@
 Finally, we introduce a VSA condition that is understudied in the context of
-gaze-independent BCI development: split VSA.
+gaze-independent \ac{bci} development: split VSA.
 In split VSA, the user mentally focuses on one cued target while gazing at
@@
 trying to tackle several problems that arise from gaze-independence, such as
-the lack or decrease in amplitude of specific ERP components, and the increased
+the lack or decrease in amplitude of specific \ac{erp} components, and the increased
 non-stationarity of the signal.
@@
 The general goal is thus to design a machine learning classifier that represents
-the ERP signal in a specific way, such that it becomes more robust to the problems
+the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
 occurring in covert attention conditions.
@@
 P3 component in covert attention
-settings~\cite{Treder2010}, the signal-to-noise ratio (SNR) of
-the ERP is lower than in overt attention settings.
+settings~\cite{Treder2010}, the \ac{snr} of
+the \ac{erp} is lower than in overt attention settings.
 Therefore, a straightforward way to reach satisfactory
-gaze-independent decoding performance, might be by increasing overall ERP
+gaze-independent decoding performance, might be by increasing overall \ac{erp}
 decoding performance. A more accurate classifier could yield relative
@@
 
-Therefore, we have improved upon an in-house developed, state-of-the art ERP
+Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
 decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
@@
 Literature shows that better covert attention performance can be achieved by
-counteracting the non-stationarity of the P3 ERP component in covert
+counteracting the non-stationarity of the P3 \ac{erp} component in covert
 attention settings~\cite{Arico2014}.
-This can be done by estimating the latency of each single-trial ERP and aligning
+This can be done by estimating the latency of each single-trial \ac{erp} and aligning
 the P3 peaks to all fall at the same moment.
@@
 classification problem of labeling unseen data, or are not robust enough to
-deal with the low SNR of the ERP.
+deal with the low \ac{snr} of the \ac{erp}.
 Classifier-based Latency Estimation (CBLE)~\cite{Mowla2017} is a technique that can leverage
-ERP latency estimation in a decoding setting, but our results show that it
+\ac{erp} latency estimation in a decoding setting, but our results show that it
 yields no improvement in gaze-independent settings.
@@
 The goal of these experiments is to benchmark
-gaze-independent ERP decoding algorithms.
-We denote this dataset as the Covert Visuospatial Attention ERP dataset
-(CVSA-ERP)
+gaze-independent \ac{erp} decoding algorithms.
+We denote this dataset as the Covert Visuospatial Attention \ac{erp} dataset
+(CVSA-\ac{erp})
 This study was approved by the Ethics Commission of University Hospital Leuven
@@
 intended target (mental attention) which could occur in patients with hampered eye
-motor control, healthy participants are cued to operate the BCI in specific
+motor control, healthy participants are cued to operate the \ac{bci} in specific
 \ac{vsa} conditions.

### 2024-09-23 chapters/wcble.tex
started wcble introduction

@@
-\todo{mention published}
+This chapter, except Sections~\ref{sec:wcble/intro},
+\ref{sec:wcble/literature/single-comp} and
+\ref{sec:wcble/literature/multi-comp}
+\todo{properly assess which paragraphs were published and which not}
+
 \section{Introduction}
+\label{sec:wcble/intro}
 
-Single-trial ERP latency jitter introduces a problem in ERP analysis:
-the \textit{smearing effect} occurs when aggregating over multiple identical
-signals with different latencies.
-A familiar example of this happens when averaging over jittered epochs for ERP
-analysis.
-In the presence of jitter, the shape and amplitude of the average do not
-reflect the properties of the original signal.
-As the smearing effect reduces the amplitude of the average,
-it negatively impacts the signal-to-noise ratio (SNR).
-The smearing effect equally impacts the SNR of information captured by a
-classifier's parameters when training with a procedure that does not account for
-this jitter, and thus also affects its performance~\cite{Thompson2012}.
-Hence, effective latency estimation and jitter compensation can contribute to higher
-decoding performance.
-%In this work, we propose a latency estimation and alignment strategy based on
-%the method proposed by \cite{Thompson2012}.
+Visual oddball \acp{bci} establish communication with paralyzed
+individuals by decoding \acp{erp} obtained from time-modulated flashing
+stimuli.
+These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
+\ac{erp} waveform.
+These components each have their own distinct amplitude and latency.
+\Ac{erp} component latency can reflect a wide range of cognitive,
+neurophysiological and clinical properties, that can be confounding factors in
+the development of \acp{bci}.\todo{cite}
+Latency is often assessed as the latency of \ac{erp} components in a subject
+average or grand average.
+Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
+relevant information, yet they are very hard to measure due to low \ac{snr} of
+single trial \acp{erp}.
+Within session variability of \ac{erp} latency, or \ac{erp} latency
+\emph{jitter} has been shown to influence \ac{bci}
+accuracy~\cite{Thompson2012}.
+This effect is especially prominent when bringing \ac{bci} development from the
+lab to the clinical setting with patients in the intended target population.
+\textcite{Zisk2021} showed that patients with \ac{als} have significantly
+increased ERP jitter.
+\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
+performance, and disproportionally so when the user is not directly gazing at
+intended targets, such as is the case for for gaze-impaired patients.
 
-\todo{General introduction of why we're doing this}
+\subsection{\Ac{erp} latency jitter}
 
-\subsection{ERP latency jitter}
-
-\subsubsection{Neural origins of latency jitter in the time domain}
-While the classic ERP model assumes a constant ERP response over trials,
+\subsubsection{Neural origins of latency jitter}
+While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
 research has shown that cortical evoked potentials show significant
-trial-to-trial jitter~\cite{truccolo2002trial}.
-Experiments show that latency jitter in ERPs is the result of the interacion of
+trial-to-trial jitter~\cite{Truccolo2002}.
+Experiments show that latency jitter in \ac{erp}s is the result of the interacion of
 the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
-	kisley1999trial, Curto2009, Arieli1996}.
-
+	Kisley1999, Curto2009, Arieli1996}.
 The latency of an evoked potential is a combination of the modulation of oscillating activity that is
@@
 
-\subsubsection{Implications for ERP analysis}
-The presence of latency jitter introduces multiple issues to ERP analysis, the
+\subsubsection{Implications for \ac{erp} analysis}
+
+The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
 most prominent of which is the smearing effect as illustrated
-in~\autoref{fig:smearing}. In ERP analysis, single-trial ERPs are usually
-averaged per condition to cope with the unfavorable signal-to-noise ratio  When
-averaging over multiple identical signals with jittered latencies, the shape and amplitude of the average do not reflect the properties
-of the original signal. Because the amplitude of the average is lower, the
-signal-to-noise ratio will also be negatively impacted.
-More generally, the smearing effect leads to difficulties interpreting ERP
+in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \ac{erp}s are usually
+averaged per condition to cope with the unfavorable \ac{snr}?
+When averaging over multiple identical signals with jittered latencies, the shape
+and amplitude of the average do not reflect the properties
+of the original signal.
+Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
+More generally, the smearing effect leads to difficulties interpreting \ac{erp}
 results and making inferences about amplitude or latency effects, because these
-are entangled when using averaging~\cite{woody1967characterization, Pfefferbaum1980, McDowell2003, Verleger2005,
+are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
 	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
@@
 \todo{illustrate smearing effect}
-%\begin{figure}
-%  \todo{smearing effect}
-%	\caption{The smearing effect: averaging over jittered waveforms leads to a spread out
-%		average with a lower amplitude.}
-%	\label{fig:smearing}
-%\end{figure}
+
+The smearing effect equally impacts the \ac{snr} of information captured by a
+classifier's parameters when training with a procedure that does not account for
+this jitter, and thus also affects its performance~\cite{Thompson2012}.
+As an illustrative example, think of an \ac{lda} \ac{erp} classifier that uses
+class averages over \ac{erp} epochs to construct its between-class scatter
+matrix.
+These per-class averages will be affected by the smearing effect and will be
+less effective templates for \ac{erp} retrieval.
+Similarly, the noise model in \ac{lda}'s  assumes stationarity and subtracts
+the class-wise averages from the signal to calculate
+the within-class scatter matrix, which will then also be affected.
+Hence, effective latency estimation and jitter compensation can contribute to higher
+decoding performance.
 
 \subsubsection{Benefits of latency estimation}
-A common approach to counteract the smearing effect is correcting single-trial
-ERPs for jitter before averaging.\todo{reference} This can be done by estimating the latency of each
-single-trial ERP, and aligning them before averaging.
+A common approach to counteract the smearing effect is to correct single-trial
+\ac{erp}s for jitter before averaging.\todo{reference}
+This can be done by estimating the latency of each
+single-trial \ac{erp}, and aligning them before averaging.
 Latency estimation has multiple advantages:
 \begin{enumerate*}[label={(\arabic*)}]
-	\item Latency-based ERP alignment can help reveal the true shape and
-	properties like peak amplitude and latency of ERPs across
+	\item Latency-based \ac{erp} alignment can help reveal the true shape and
+	properties like peak amplitude and latency of \ac{erp}s across
 	experimental conditions. These are variables of interest for
@@
 	Additionally, latency estimation allows for the extraction of
-	other descriptors from a set of single-trial ERPs, like
+	other descriptors from a set of single-trial \ac{erp}s, like
 	amplidute or latency variance~\cite{Hultsch2004}, which can
-	also be of interest as studied variables. For instance, in
-	\cite{Saville2014}, the inter-trial latency variability itself
-	is determined as a correlate of working memory performance.
+	also be of interest as studied variables.
+  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
+  variability itself is determined as a correlate of working memory performance.
 	\item Counteracting the smearing effect will boost the signal-to-noise
-	ratio of averaged ERPs by increasing the amplitude of the
-	ERP signal. This in turn can reveal smaller or more jittered ERP components
-	which are otherwise obscured or blurred due to desynchronization.
-	These deblurred templates can then be used for analysis of the ERP
+	ratio of averaged \ac{erp}s by increasing the amplitude of the
+	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
+  components which would otherwise obscured or blurred due to desynchronization.
+	These deblurred templates can then be used for analysis of the \ac{erp}
 	waveform, or as templates of the signal of interest in
@@
 \subsection{Single-component approaches}
+\label{sec:wcble/literature/single-comp}
+
 The most straight-forward method to estimate single-trial latency
-peak-picking. Here, the latency of an ERP is determined as the
+peak-picking. Here, the latency of an \ac{erp} is determined as the
 time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
@@
 filtering. Filtering can suppress the noise contaminating EEG trials, lowering
-the risk of picking a noise peak instead of a true peak of the ERP. Several filtering approaches have been developed to be used in
+the risk of picking a noise peak instead of a true peak of the \ac{erp}. Several filtering approaches have been developed to be used in
 conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
@@
 estimation method that is still being used and is considered amongst the more
-performant techniques~\cite{Ouyang2017}. It starts with the average ERP as a
-rough estimation of the aligned template ERP, and iteratively refines this
+performant techniques~\cite{Ouyang2017}. It starts with the average \ac{erp} as a
+rough estimation of the aligned template \ac{erp}, and iteratively refines this
 template by determining at each step the time point of maximum
-cross-correlation of this template with all single-trial ERPs and aligning them
+cross-correlation of this template with all single-trial \ac{erp}s and aligning them
 on these time points to form the next template.
@@
 optimum. To avoid this, a fitness function can be defined to jointly optimize
-the set of single-trial ERP latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
+the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
 
@@
 \subsection{Multi-component approaches}
-The abovementioned methods do not take into account the fact that  ERPs are multi-component signals.
+\label{sec:wcble/literature/multi-comp}
+The abovementioned methods do not take into account the fact that  \ac{erp}s are multi-component signals.
 Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
@@
 
-Most of the algorithms above can be adapted to work on multiple ERP components
+Most of the algorithms above can be adapted to work on multiple \ac{erp} components
 by carefully selecting peaks or pre-determined regions of interest.
 \cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
-time-window for each ERP component to extract a latency per component.
+time-window for each \ac{erp} component to extract a latency per component.
 Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
@@
 however, these algorithms lack an integrated approach to deal with multiple
-ERP components. A clear evaluation is lacking of the performance of these
+\ac{erp} components. A clear evaluation is lacking of the performance of these
 methods in
@@
 button trigger signal as input to determine the response time, which is
-not always present or applicable in an ERP experiment, especially in
+not always present or applicable in an \ac{erp} experiment, especially in
 the case of BCIs.
@@
 Finally, the smearing problem can also be overcome by algorithms that are not
-based on ERP component latency estimation, but which use time warping-based methods
-to enhance the shape of the averaged template ERP, like Dynamic Time
+based on \ac{erp} component latency estimation, but which use time warping-based methods
+to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
 Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
@@
 The most straightforward latency estimation method is \textit{peak picking}.
-Peak picking determines the latency of an ERP as the time point of its maximum
+Peak picking determines the latency of an \ac{erp} as the time point of its maximum
 (or minimum for negative components) amplitude relative to stimulus
@@
 Woody's scheme starts with a template that approximates the expected response,
-e.g. the average target ERP, and iteratively enhances the SNR of this template by determining at
+e.g. the average target \ac{erp}, and iteratively enhances the SNR of this template by determining at
 each iteration the time point of maximum cross-correlation of the template
-with all single-trial ERPs, and aligning them based on these latencies to form
+with all single-trial \ac{erp}s, and aligning them based on these latencies to form
 the template for the next iteration.
@@
 
-When multiple clusters of ERP components are not time-locked to
+When multiple clusters of \ac{erp} components are not time-locked to
 each other, aligning to the latency of one cluster might introduce more jitter
@@
 Finally, algorithms that use time
-warping-based methods instead of explicit ERP component latency estimation can
+warping-based methods instead of explicit \ac{erp} component latency estimation can
 also overcome the smearing problem.
-Examples that have been applied to ERPs are Dynamic Time
+Examples that have been applied to \ac{erp}s are Dynamic Time
 Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
@@
 decoding scheme to improve performance for incoming epochs with unknown
-class labels that contain either a target or non-target ERP response.
+class labels that contain either a target or non-target \ac{erp} response.
 They can be used offline on a set of labeled epochs for testing hypotheses
 concerning latency and jitter, for aligning templates, or for BCI performance
-prediction, but not for ERP classification.
+prediction, but not for \ac{erp} classification.
 While some of the aforementioned methods could be adapted to perform
@@
 for jitter-resistant decoding.
-\cite{hardiansyah2020single} incorporated single-trial latencies in
+\cite{Hardiansyah2020} incorporated single-trial latencies in
 classification by peak-picking within a given
-ERP time window, unaware of the class of the epoch under investigation.
+\ac{erp} time window, unaware of the class of the epoch under investigation.
 The Classifier-Based Latency Estimation (CBLE) algorithm introduced
@@
 that are robust to jitter can be used in a decoding setting,
-but, in general, have scarcely been applied to ERP decoding.
+but, in general, have scarcely been applied to \ac{erp} decoding.
 Data augmentation involving jittering the training
@@
 covariances as features~\cite{Aydarkhanov2020} have both been shown
-to perform well in the presence of ERP jitter.
+to perform well in the presence of \ac{erp} jitter.
 In this work, we opted to apply CBLE because it has successfully been applied to
-classify jittered ERPs.
+classify jittered \ac{erp}s.
 We adapt the CBLE to an iterative method akin to Woody's scheme that can better
@@
 Our method attempts to model the spatial and temporal activation of
-each ERP component, as well as it's latency distribution, and takes into acount the
+each \ac{erp} component, as well as it's latency distribution, and takes into acount the
 spatial and temporal correlation of the EEG background noise.
@@
 with $R = S-(s_2-s_1)$.
-To leverage CBLE for ERP classification, the score vectors $\mathbf{y}$ can be
+To leverage CBLE for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
 arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
@@
 In the CBLE-decoder, it is this wavelet transform that decreases the sensitivity
-to latency differences, actively compensating for ERP latency jitter.
+to latency differences, actively compensating for \ac{erp} latency jitter.
 
@@
 that CBLE is relatively independent of the first-stage classifier for BCI
-accuracy prediction and for ERP classification.
+accuracy prediction and for \ac{erp} classification.
 Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
@@
 \label{sec:robust-latency}
-The ERP decoding method based on CBLE introduced by \cite{Mowla2017} does
+The \ac{erp} decoding method based on CBLE introduced by \cite{Mowla2017} does
 not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
@@
 This is the point in time where the target class reaches largest separation
-from the background noise and non-target class, indicating the target ERP is
+from the background noise and non-target class, indicating the target \ac{erp} is
 most likely to occur here.
@@
 However, this is problematic since it is not evident how to estimate
-latency of e.g. a P3 ERP component for a non-target epoch, since the
+latency of e.g. a P3 \ac{erp} component for a non-target epoch, since the
 non-target class is characterized by the absence of this component.
-%\footnote{Depending on the stimulation and experimental design, ERP components can be present in one experimental
+%\footnote{Depending on the stimulation and experimental design, \ac{erp} components can be present in one experimental
 %	condition or class and missing in another, or appear in multiple classes
 %	at different amplitudes. In the latter case, it could be possible
-%	to estimate the latency of an ERP in the conditions where it appears, but if
+%	to estimate the latency of an \ac{erp} in the conditions where it appears, but if
 %	its amplitude is lower or negligible in some conditions, latency estimates
@@
 directly, CBLE latency estimation is used as a step in a Woody iteration scheme.
-While the Woody algorithm iteratively enhances the SNR of an ERP template to
+While the Woody algorithm iteratively enhances the SNR of an \ac{erp} template to
 cross-correlate with the data, WCBLE iteratively re-estimates the parameters of

### 2024-09-23 chapters/wcble.tex
Literature wcble

@@
 
-Visual oddball \acp{bci} establish communication with paralyzed
+\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
 individuals by decoding \acp{erp} obtained from time-modulated flashing
@@
   components which would otherwise obscured or blurred due to desynchronization.
-	These deblurred templates can then be used for analysis of the \ac{erp}
+    These \emph{deblurred} templates can then be used for analysis of the \ac{erp}
 	waveform, or as templates of the signal of interest in
-	BCIs~\cite{arico2014influence}.
+	BCIs~\cite{Arico2014}.
 	%this enables decoders to take into account the relatively low amplitude
@@
 	%since they're already time-locked}
-	\item Finally, the introduction of latency estimation and alignment generaly
-	improves BCI classification performance. Single-trial latencies can be
-	used as features to improve classification~\cite{hardiansyah2020single}.
+	\item Finally, the introduction of latency estimation and alignment generally
+	improves BCI classification performance.
+  Single-trial latencies can be	used as features to improve
+  classification~\cite{Hardiansyah2020}.
 	Furthermore, compensating for inter-trial differences, especially when
-	extended to inter-subject differences, can aid in transfer learning or
+	extended to inter-session or inter-subject differences, can aid in transfer learning or
 	generalization across subjects and protocols\cite{Iturrate2014}.
 \end{enumerate*}
-\todo{Read, summarize and cite BCI	references from bibliography}
 
@@
 
-The most straight-forward method to estimate single-trial latency
-peak-picking. Here, the latency of an \ac{erp} is determined as the
-time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
-onset. While straight-forward, this method
-does not perform well in low or negative SNR conditions, unless combined with
-filtering. Filtering can suppress the noise contaminating EEG trials, lowering
-the risk of picking a noise peak instead of a true peak of the \ac{erp}. Several filtering approaches have been developed to be used in
-conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
-	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
-	D’Avanzo2011} and in the time-frequency
-domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
+The most straight-forward method to estimate single-trial latency is
+\emph{peak picking}
+Here, the latency of an \ac{erp} is determined as the
+time-point of it's maximum (or minimum) amplitude relative to stimulus
+onset.
+While straight-forward, this method does not perform well in low or negative
+\ac{snr} conditions, unless combined with filtering.
+Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
+the risk of picking a noise peak instead of a true peak of the \ac{erp}.
+Several filtering approaches have been developed to be used in
+conjuction with peak-picking, in the time domain~\cite{
+  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
+  Georgiadis2005,	D’Avanzo2011}
+and in the time-frequency domain~\cite{
+  Quiroga2003, Wang2007,Hu2010}.
+The aforementioned analysis and performance prediction method proposed
+by \cite{Arico2014} estimates the single-trial P3 latencies of attended
+epochs by P3 peak-picking after filtering in the time-frequency domain, and
+corrects for jitter by temporally aligning all target epochs to the average
+latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
+at the same time instance.
+Filtering can
 also be done in the spatial domain. The spatial filter can be constructed
-using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
-	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.
+using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
+	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.
 
-A generally more preferred approach than
-peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
-estimation method that is still being used and is considered amongst the more
-performant techniques~\cite{Ouyang2017}. It starts with the average \ac{erp} as a
-rough estimation of the aligned template \ac{erp}, and iteratively refines this
-template by determining at each step the time point of maximum
-cross-correlation of this template with all single-trial \ac{erp}s and aligning them
+Template matching is generally preferred to peak picking.
+The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
+latency estimation method that is still being used and is even
+considered amongst the more performant techniques for \acp{erp}~\cite{Ouyang2017}.
+It starts with an average \ac{erp} as a rough estimation of the aligned template
+\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
+cross-correlation of this template with all single-trial \acp{erp} and aligning them
 on these time points to form the next template.
 Similarily, cross-correlation template matching can be combined with filtering.
-\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
+\cite{Souloumiac2013} uses XDAWN to construct spatial filter before
 template matching the spatially filtered signal. \cite{Iturrate2014} uses a
@@
 latencies to obtain the template for the next iteration, weighted averaging can
-also be used~\cite{gasser1983selavco}.
+also be used~\cite{Gasser1983}.
+\todo{cite Thornton2008 hierarchical woody?}
 
 Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
-optimum. To avoid this, a fitness function can be defined to jointly optimize
+optimum.
+To avoid this, a fitness function can be defined to jointly optimize
 the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
-
-Methods adapted to handle multiple conditions, useful in BCI setting
-\begin{itemize}
-	\item \cite{Kim2020}
-\end{itemize}
+Alternative methods are based on graph
+optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020} or by fitting an \ac{erp} model with
+maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988,Puce1994}.
 
@@
 \label{sec:wcble/literature/multi-comp}
-The abovementioned methods do not take into account the fact that  \ac{erp}s are multi-component signals.
-Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
-time-locked to each other. They each can have different latency
-variabilities. The presence of more than one component can hamper the
-performance of some algorithms, and it often only possible to extract only the
+In the previous section, we have abstracted away the fact that \ac{erp}s are
+multi-component signals, which is not taking into account by the aforementioned
+methods.
+Multiple components can be time-locked to different events or neurophysiological
+proccess and therefore are not necessarily time-locked to each other. They
+each can have different latency variabilities.
+The presence of more than one component can hamper the
+performance of some latency estimation algorithms, and it often only possible to extract only the
 latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
@@
 by carefully selecting peaks or pre-determined regions of interest.
-\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
+\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
 time-window for each \ac{erp} component to extract a latency per component.
-Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
+Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
 to utilize multiple of their components as filters, resulting in multiple time
@@
 Some algorithms separate components into stimulus and response locked component clusters based on their
-latency distributions~\cite{jung2001analysis, Takeda2008,
-	Zhang1998, Yin2009}. These algorithms do, however, require a response
+latency distributions~\cite{Jung2001, Takeda2008,	Zhang1998, Yin2009}.
+These algorithms do, however, require a response
 button trigger signal as input to determine the response time, which is
@@
 the case of BCIs.
-The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
-	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
-the response time, and can separate stimulus-locked, response-locked,
-and multiple central component clusters. Others, like spatial filtering methods that use
-decomposition methods to determine the spatial filter can be be modified to separate
-components.~\cite{Ouyang2017}.
-
-To categorize:\todo{categorize and analyze these references}
-\begin{itemize}
-	\item \cite{Dimitriadis2018}
-	\item \cite{Thornton2008}
-	\item Maximum likelihood~\cite{tuan1987variable, Moecks1988,Puce1994} Maximum likelihood
-	\item \cite{Gratton1989} (no access) Maximum likelihood?
-\end{itemize}
-
-\todo{Explain beamforming}
+The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,	Wang2015, Ouyang2016, Ouyang2020}
+is able to do this without entirely relying on the response time, and can
+separate stimulus-locked, response-locked, and multiple central component clusters.
+Others, like spatial filtering methods that use decomposition methods to
+determine the spatial filter can be be modified to separate
+components~\cite{Ouyang2017}.
 
@@
 Warping~\cite{Skov2006}, or Fast Variational
-Alignment~\cite{flotho2021fast}. Since these methods work
+Alignment~\cite{flotho2021fast}.
+These methods work
 fundamentally different from the latency estimation methods, and cannot
-directly be used to extract latency features, we will not further consider them
-in this work.
-
-\todo{check overlap with:
-
-The most straightforward latency estimation method is \textit{peak picking}.
-Peak picking determines the latency of an \ac{erp} as the time point of its maximum
-(or minimum for negative components) amplitude relative to stimulus
-onset.
-While easy to implement, this method does not perform well in low or negative
-SNR conditions because of the risk of picking a noise-related peak, unless
-combined with strong filtering in the space, time or frequency
-domains~\cite{Avanzo2011,Arico2014,TREDER2016279,Ouyang2017}.
-The aforementioned analysis and performance prediction method proposed
-by \cite{Arico2014} estimates the single-trial P3 latencies of attended
-epochs by P3 peak-picking after filtering in the time-frequency domain, and
-corrects for jitter by temporally aligning all target epochs to the average
-latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
-at the same time instance.
-Due to the risk of picking noise peaks, template matching is generally
-preferred.
-Woody's algorithm~\cite{woody1967characterization} is a simple and performant
-latency estimation method that is considered to be among
-the most performant ones~\cite{Ouyang2017}.
-Woody's scheme starts with a template that approximates the expected response,
-e.g. the average target \ac{erp}, and iteratively enhances the SNR of this template by determining at
-each iteration the time point of maximum cross-correlation of the template
-with all single-trial \ac{erp}s, and aligning them based on these latencies to form
-the template for the next iteration.
-The final, enhanced template can then be cross-correlated again to retrieve an
-accurate estimation of single-trial latencies.
-Similar to peak-picking, cross-correlation template matching can be combined
-with a spatial filter such as XDAWN~\cite{souloumiac2013improved} or a
-spatiotemporal filter \cite{Iturrate2014}.
-More recent work achieves good performance in latency estimation using residue
-iteration~\cite{Ouyang2017}, genetic
-algorithms~\cite{Pelo2018}, a hidden process model~\cite{Kim2020} or
-graph-based network analysis~\cite{Dimitriadis2018}.
-
-When multiple clusters of \ac{erp} components are not time-locked to
-each other, aligning to the latency of one cluster might introduce more jitter
-in other component clusters, further decreasing the SNR~\cite{Ouyang2020}.
-
-Finally, algorithms that use time
-warping-based methods instead of explicit \ac{erp} component latency estimation can
-also overcome the smearing problem.
-Examples that have been applied to \ac{erp}s are Dynamic Time
-Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
-Correlation-Optimized Warping~\cite{Skov2006}, or Fast Variational
-Alignment~\cite{Flotho2021}. Since these methods are fundamentally
-different from latency estimation methods,
-we will not further consider them in this work.
-
-}
+directly be used to extract latency features.
 
 \subsection{Contribution}
+
+\todo{cite Kim2020 used to analyze the difference between target and
+non-target}
 
@@
 \section{Results}
+simulation study
+refer to later chapter for use on real erp data
 
@@
 \section{Conclusion}
+Current work is focusing on the extension to multi-component algorithm, in a
+similar fashion to ouyang

### 2024-09-23 chapters/wcble.tex
Simulation methods

@@
 trial-to-trial jitter~\cite{Truccolo2002}.
-Experiments show that latency jitter in \ac{erp}s is the result of the interacion of
+Experiments show that latency jitter in \acp{erp} is the result of the interacion of
 the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
@@
 most prominent of which is the smearing effect as illustrated
-in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \ac{erp}s are usually
+in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
 averaged per condition to cope with the unfavorable \ac{snr}?
@@
 A common approach to counteract the smearing effect is to correct single-trial
-\ac{erp}s for jitter before averaging.\todo{reference}
+\acp{erp} for jitter before averaging.\todo{reference}
 This can be done by estimating the latency of each
@@
 	\item Latency-based \ac{erp} alignment can help reveal the true shape and
-	properties like peak amplitude and latency of \ac{erp}s across
+	properties like peak amplitude and latency of \acp{erp} across
 	experimental conditions. These are variables of interest for
@@
 	Additionally, latency estimation allows for the extraction of
-	other descriptors from a set of single-trial \ac{erp}s, like
+	other descriptors from a set of single-trial \acp{erp}, like
 	amplidute or latency variance~\cite{Hultsch2004}, which can
@@
 	\item Counteracting the smearing effect will boost the signal-to-noise
-	ratio of averaged \ac{erp}s by increasing the amplitude of the
+	ratio of averaged \acp{erp} by increasing the amplitude of the
 	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
@@
 	waveform, or as templates of the signal of interest in
-	BCIs~\cite{Arico2014}.
+	\acp{bci}~\cite{Arico2014}.
 	%this enables decoders to take into account the relatively low amplitude
@@
 	\item Finally, the introduction of latency estimation and alignment generally
-	improves BCI classification performance.
+	improves \ac{bci} classification performance.
   Single-trial latencies can be	used as features to improve
@@
 \label{sec:wcble/literature/multi-comp}
-In the previous section, we have abstracted away the fact that \ac{erp}s are
+In the previous section, we have abstracted away the fact that \acp{erp} are
 multi-component signals, which is not taking into account by the aforementioned
@@
 not always present or applicable in an \ac{erp} experiment, especially in
-the case of BCIs.
+the case of \acp{bci}.
 The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,	Wang2015, Ouyang2016, Ouyang2020}
@@
 to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
-Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
+Warping~\cite{Gupta1996, Wang2001, Zoumpoulaki2015}, Correlation-Optimized
 Warping~\cite{Skov2006}, or Fast Variational
-Alignment~\cite{flotho2021fast}.
+Alignment~\cite{Flotho2021}.
 These methods work
@@
 They can be used offline on a set of labeled epochs for testing hypotheses
-concerning latency and jitter, for aligning templates, or for BCI performance
+concerning latency and jitter, for aligning templates, or for \ac{bci} performance
 prediction, but not for \ac{erp} classification.
@@
 \ac{erp} time window, unaware of the class of the epoch under investigation.
-The Classifier-Based Latency Estimation (CBLE) algorithm introduced
+The \ac{cble} algorithm introduced
 by \cite{Thompson2012} also explicitly applies latency estimation in a
 decoding setting.
-\cite{Thompson2012} initially formulated CBLE as an off-line performance prediction method.
-Later, its output was successfully adapted to compensate for jitter to improve decoder performance%
-~\cite{Mowla2017,Zisk2022}.
+\cite{Thompson2012} initially formulated \ac{cble} as an off-line performance
+prediction method.
+Later, its output was successfully adapted to compensate for jitter to improve
+decoder performance~\cite{Mowla2017,Zisk2022}.
 
@@
 to perform well in the presence of \ac{erp} jitter.
-In this work, we opted to apply CBLE because it has successfully been applied to
-classify jittered \ac{erp}s.
-We adapt the CBLE to an iterative method akin to Woody's scheme that can better
+In this work, we opted to apply \ac{cble} because it has successfully been applied to
+classify jittered \acs{erp}.
+We adapt the \ac{cble} to an iterative method akin to Woody's scheme that can better
 compensate for jitter, to improve covert VSA decoding performance.
-
-\todo{remove lcmv}
-In this work, we introduce a new latency method that leverages data-driven,
-joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
-integrates the spatiotemporal filtering in Woody's template matching iteration
-scheme and can easily be extended to multi-component settings.
-Our method attempts to model the spatial and temporal activation of
-each \ac{erp} component, as well as it's latency distribution, and takes into acount the
-spatial and temporal correlation of the EEG background noise.
 
@@
 
-CBLE, summarized in Algorithm~\ref{alg:cble}, works by training a
+\ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
 \textit{first-stage} classifier $\mathcal{C}(\theta,f)$
@@
 with $R = S-(s_2-s_1)$.
-To leverage CBLE for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
+To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
 arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
@@
 dimensionality before classification with the second-stage classifier.
-In the CBLE-decoder, it is this wavelet transform that decreases the sensitivity
+In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
 to latency differences, actively compensating for \ac{erp} latency jitter.
@@
 When using a simple spatiotemporal linear classifier as first-stage classifier,
-CBLE is equivalent to the first iteration of Woody's algorithm with the
+\ac{cble} is equivalent to the first iteration of Woody's algorithm with the
 spatiotemporal classifier weights as template.
 \cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
-that CBLE is relatively independent of the first-stage classifier for BCI
+that \ac{cble} is relatively independent of the first-stage classifier for \ac{bci}
 accuracy prediction and for \ac{erp} classification.
@@
 
-\subsection{Robust CBLE latency features}
+\subsection{Robust \ac{cble} latency features}
 \label{sec:robust-latency}
-The \ac{erp} decoding method based on CBLE introduced by \cite{Mowla2017} does
+The \ac{erp} decoding method based on \ac{cble} introduced by \cite{Mowla2017} does
 not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
 to the second-stage classifier.
-%Furthermore, previous CBLE works~\cite{Thompson2012,Mowla2020} only used these
-%latencies to correlate them with neurophysiological processes or to predict
-%decoder performance.
-We noticed that, while CBLE performance was unaffected, the classification
+Furthermore, previous \ac{cble} works~\cite{Thompson2012,Mowla2020} only used these
+latencies to correlate them with neurophysiological processes or to predict
+decoder performance.
+We noticed that, while \ac{cble} performance was unaffected, the classification
 performance of our proposed method can be improved if the estimated
@@
 
-These latency features give rise to the following issue when
+However, including these latency features give rise to the following issue when
 classifying unseen data.
-The CBLE latency estimate is defined only for target epochs as
+The \ac{cble} latency estimate is defined only for target epochs as
 $s_\mathrm{target}=\argmax_s{y_s}$.
@@
 non-target class is characterized by the absence of this component.
-%\footnote{Depending on the stimulation and experimental design, \ac{erp} components can be present in one experimental
-%	condition or class and missing in another, or appear in multiple classes
-%	at different amplitudes. In the latter case, it could be possible
-%	to estimate the latency of an \ac{erp} in the conditions where it appears, but if
-%	its amplitude is lower or negligible in some conditions, latency estimates
-%	will be less accurate, hence this case is still problematic.}.
+\footnote{Depending on the stimulation and experimental design, \ac{erp} components can be present in one experimental
+	condition or class and missing in another, or appear in multiple classes
+	at different amplitudes. In the latter case, it could be possible
+	to estimate the latency of an \ac{erp} in the conditions where it appears, but if
+	its amplitude is lower or negligible in some conditions, latency estimates
+	will be less accurate, hence this case is still problematic.}.
 In fact, $y$ can have multiple local minima or entirely lack distinct peaks for
@@
   \item lead to smoother convergence in our proposed iterative alignment scheme
-    for WCBLE, which heavily relies on exact latency estimation.
+    for W\ac{cble}, which heavily relies on exact latency estimation.
 \end{enumerate*}
@@
 
-\subsection{Classifier-based Latency Estimation with Woody
-	Iterations}
-To improve performance over CBLE, we propose a new algorithm inspired both by
-CBLE and the aforementioned Woody iteration scheme (WCBLE).
-Instead of using CBLE to estimate the features of a second-stage classifier
-directly, CBLE latency estimation is used as a step in a Woody iteration scheme.
+\subsection{\acl{wcble}}
+To improve performance over \ac{cble}, we propose a new algorithm inspired both by
+\ac{cble} and the aforementioned Woody iteration scheme termed \ac{wcble}.
+Instead of using \ac{cble} to estimate the features of a second-stage classifier
+directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
 While the Woody algorithm iteratively enhances the SNR of an \ac{erp} template to
-cross-correlate with the data, WCBLE iteratively re-estimates the parameters of
+cross-correlate with the data, W\ac{cble} iteratively re-estimates the parameters of
 the first-stage classifier.
-To improve convergence and perform well in a classification setting, WCBLE
+To improve convergence and perform well in a classification setting, W\ac{cble}
 aligns both targets and non-targets to their corresponding estimated latencies.
 
-The WCBLE algorithm is presented in Algorithm~\ref{alg:wcble}.
+The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
 Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
 The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
-At every iteration, classifier $\mathcal{C}$ is trained like in CBLE:
+At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
 \begin{equation}
@@
 smoother yet longer convergence process compared to the maximum/minimum approach.
-We can then apply the classifier with enhanced parameters $\theta^*$ in a CBLE
+We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
 manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
@@
 simulation study
+equatin snr
+\subsection{Data simulation}
+To verify our approach, we simulated 32 channel \ac{eeg} data with standard
+10-20 electrode positions using the \textttt{simulate_evoked} method of the
+MNE (version 1.8)~\cite{Gramfort2013} software package.
+\Ac{erps} were generated by projecting a source time course in a dipole in the
+left hemisphere to the scalp electrodes using a boundary element method forward
+model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
+We refer to~\textcite{Gramfort2014} for implementation details.
+Pink noise generated using an infinite impulse response filter with transfer
+function $\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^-2}$.
+
+figure
+Three kinds of source time courses, gaussian, a
+morlet wavelet and a sine wave, to study the algorithm's behavior for different
+waveforms.
+equations with parameters
+full simulation fig for wavelet
+
+For each source time course,
+pink noise is generated at all sources? and projected to the scalp
+
+\subsection{Latency estimation}
+cble with prob/median
+wcble with prob+median
+cble with max
+wcble with max
+
+CV/LOOCV?
+\subsection{Classification}
 refer to later chapter for use on real erp data
@@
 \section{Discussion}
+\todo{mention cross validation}
+\todo{remove lcmv}
+In this work, we introduce a new latency method that leverages data-driven,
+joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
+integrates the spatiotemporal filtering in Woody's template matching iteration
+scheme and can easily be extended to multi-component settings.
+Our method attempts to model the spatial and temporal activation of
+each \ac{erp} component, as well as it's latency distribution, and takes into acount the
+spatial and temporal correlation of the EEG background noise.
+
+with lda, it is in the single class case equal to shifting a spatiotemporal
+lcmv beamformer over the epochs and can be used to extract latencies.
+multiclass:
 

### 2024-09-23 chapters/wcble.tex
Describe simulation method

@@
 simulation study
-equatin snr
+
 \subsection{Data simulation}
 To verify our approach, we simulated 32 channel \ac{eeg} data with standard
-10-20 electrode positions using the \textttt{simulate_evoked} method of the
+10-20 electrode positions using the \texttt{simulate\_evoked} method of the
 MNE (version 1.8)~\cite{Gramfort2013} software package.
-\Ac{erps} were generated by projecting a source time course in a dipole in the
+\Acp{erp} were generated by projecting a source time course in a dipole in the
 left hemisphere to the scalp electrodes using a boundary element method forward
@@
 We refer to~\textcite{Gramfort2014} for implementation details.
-Pink noise generated using an infinite impulse response filter with transfer
-function $\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^-2}$.
+Pink spatial and  temporal noise are then added.
+Temporal noise is generated by passing gaussian noise through an infinite
+impulse response filter with transfer function
+$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
+For pink spatial noise, a noise covariance matrix is constructed from the
+cosine distances between electrode positions and scaling the resulting spatial
+frequencies by $1/f$.
+
+Evoked potentials are simulated for three different source time course waveforms,
+a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
+defined by the following formulas:
+\begin{equation}
+  f_\text{gaussian}(t,l) = a
+  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
+\end{equation}
+\begin{equation}
+  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
+\end{equation}
+\begin{equation}
+  f_\text{sine}(t,l) =
+  \begin{cases}
+    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
+    0 & \text{otherwise}
+  \end{cases}
+\end{equation}
+with time $t$, latency $l$, and amplitude $a=\num{1e{-7}}$V.
+The standard deviation that determines the width gaussian $\varsigma=0.1s$.
+For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
+For the sine pulse, the frequency $f=4$
+For each of these, 100 target epochs containing the waveform and noise and 100
+non-target epochs containing only noise were simulated.
+Additionally, the target epochs were jittered by setting the latency offset $l$
+of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
+deviations of respectively $\sigma=0.1,0.2,0.3$s.
+The noise was also scaled at different 32  different noise levels, with the \ac{snr} of a
+target epoch ranging from 0 to -31dB, with
+\begin{equation}
+  \text{SNR} = 10\log_{10}\frac{\text{var}f(t,0)}{\text{var}\left(\text{noise}\right)}
+\end{equation}
 
 figure
-Three kinds of source time courses, gaussian, a
-morlet wavelet and a sine wave, to study the algorithm's behavior for different
-waveforms.
-equations with parameters
-full simulation fig for wavelet
-
-For each source time course,
-pink noise is generated at all sources? and projected to the scalp
 

### 2024-09-24 chapters/bci.tex
Intro, P300 stimulation

@@
-Independent home use of a brain-computer interface by people with
-amyotrophic lateral sclerosis.
-\todo{Shortly describe challenges from the patient point of view, related to
-this thesis}
-\todo{do something with Wolpaw (2018) Independent home use of a brain-computer interface by people with
-amyotrophic lateral sclerosis.}
+\todo{comments liesel paradigms}
+People suffering from heavy paralysis or disability due to
+neurodegenenerative diseases, stroke, traumatic brain or spine injury
+become entirely reliant on caregivers or family members in their day to day
+routine.
+Wouldn't it be great if we could help them regain some quality of life by
+making them more independent?
 
-\todo{
-Locked-in -\> communicate
-Paraplegic -\> walk again
-tetraplegic, locked-in syndrome
-Regaining capabilities and improving quality of life
-Start from wolpaw examples, for spelling
-Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
-willet, neuralink)
-}
+The most basic ability any independent human being needs, is interaction with
+their environment.
+In it's core essence, this is the ability to communicate your intentions,
+emotions, frustrations or thoughts to someone or something, which then might
+act upon them.
+For an able-bodied person, physical interaction happens through body movements
+and communication is usually done through speech and body language, which
+both require sending signals through the body's nervous system to control
+muscles.
+Not everyone has this capability.
+
+The key problem for heavily paralyzed patients is that the mind wants to go
+where the body can't follow.
+Couldn't we therefore design a system that directly interacts with the mind?
+This system can then form an interface between the mind and any kind of
+technology, like a robot arm or speech synthesizer, interacting with the world
+on the user's behalf.
+If the mind's seat is in the brain, and we want to interface this to a digital
+control process, such a device becomes the \ac{bci}\footnote{Sometimes also
+termed brain-machine interface (BMI) when coupled to a physical actuator, like
+a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
+assistive technology and communication devices.}.
+A \ac{bci} directly couples actions in the `real world' to the mental state of
+the user.
+A \ac{bci} reads your mind.
+
+Starting in the 1970's from methods to establish minimal brain activity
+(yes/no, left/right, ...)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
+into sophisticated interaction schemes that restore or enhance many capabilities
+by bypassing part of the nervous system and muscle control.
+By now, classic \acp{bci} methods have matured outside the lab setting and can be
+used as assistive communication technology by patients with impaired
+speech~\cite{Wolpaw2018}.
+Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
+interface allowing a paraplegic patient to walk again\cite{Lorach2023},
+a \ac{bci} translating internal speech to a facial avatar speaking the imagined
+text~\cite{Metzger2023}, and fast communication through decoding imagined
+handwritten symbols~\cite{Willett2021}.
+Recently, the idea has also gained popular traction through Elon Musk's
+Neuralink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.
+
 \todo{put focus on communication BCI}
+
 \section{A direct interface to the brain}
@@
 chapter a more scientific style is used}
-A \ac{bci} reads your mind.
 \ac{bci}s are an entertaining subject with many futuristic and more established
@@
 Devices processing direct inputs from the central nervous system can be
-useful for rehabilitation and medical diagnosis or treatment.
-They have always been especially promising as assistive technology for
-paralyzed patients limited in their communication ability. \todo{expand more on this last
-sentence, put patients first in introduction, see above}
+useful for rehabilitation and medical diagnosis or treatment, but
+have always been especially promising as assistive technology for
+paralyzed patients limited in their communication ability.
 Furthermore, they bring a new paradigm to human-computer interaction,
@@
 To improve performance, we can aim to
-\todo{enumerate}
 \begin{enumerate*}[label={\arabic*})]
@@
 
-\section{A case study: the visual oddball paradigm}
+\section{A case study: the visual oddball \ac{bci}}
 
-The visual oddball paradigm is a reactive, stimulus-dependent, synchronous
-\ac{bci} paradigm.
-good choice because it scores high on the parameters above
-well established, good research, first application cite
-Flashing in various configurations, such as row-colum or in groups,
+The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
+\emph{synchronous} \ac{bci} paradigm with all the benefits and drawbacks
+Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
+The brain response of interest is the visual \ac{erp}, which can be accurately
+measured and decoded from non-invasive \ac{eeg} signals with a relatively low
+computational effort and short calibration time.
+First established by~\cite{Farwell1988} in 1988, it well supported by literature
+and has been proven to work for patients in day-to-day home
+use~\cite{Wolpaw2018}.
 
-\subsection{Stimulation}
+\subsection{Stimulation paradigm}
+One by one, visual elements on a computer screen are \emph{intensified} for a
+short period of time by changing in color, brightness or size.
+On each flash observed by the user, specific brain activity is elicited.
+If one of these intensified elements is rare with respect to the others, i.e. it is the
+\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
+A stimulus can be rare either due to its inherent properties like color or
+location
+Crucially, this is also the case if it is `marked' as rare by the user, e.g.
+by paying explicit attention to only a given stimulus and not the others or
+counting how many times it flashes.
+If these visual elements represent letters, and we know the timing of the
+intensifications of each letter, we can now establish communication by detecting
+for which letter an oddball response was present in the brain signal at its
+time of intensification.
+
+\Ac{itr} in the oddball paradigm is optimized by intensifying
+multiple targets at once in a row-collumnn strategy and using a sequence of
+selections, giving rise to the classic
+P300 matrix speller of~\cite{Farwell1988}.
+Other optimizations increase response \ac{snr}, like using flashing face
+images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
+stimuli~\cite{Treder2011};
+
+Sometimes Flashing in various configurations, such as row-column or in groups,
+other flashing patterns
+faces, feature attention
+visual hex-o-spell
+
 comfortable, preferred over SSVEP, similar ITR is this true?
@@
 \subsection{The event-related potential}
+
+later compoentns P3 more attention/cognitive more centrally in the brain
 
@@
 
+mention VEPs more from visual system occipitally
+
 P300 and other components, explain function (Luck)
@@
 \todo{figure of ERP components}
+
+which components, at what amplitude and latency they are evoked varies per
+person and depends on the type of intensification
 
@@
 \subsection{Decoding ERPs}
+
+Usually, snr low and multiple trials necessary
+
 tLDA, beamforming

### 2024-09-24 chapters/wcble.tex
Intro, P300 stimulation

@@
+\todo{comments liesel}
+
 This chapter, except Sections~\ref{sec:wcble/intro},
@@
 \section{Results}
-simulation study
+
+We compare our proposed \ac{wcble} algorithm to \ac{cble} on both a latency
+estimation task and a binary classification task with simulated data.
+Four settings are evaluated to assess both the impact of the iterative Woody
+procedure and the extraction of robust latency features.
+These are CBLE-median and WCBLE-median, where the median of the latency
+probability distribution is used as latency estimate, and CBLE-peak and
+WCBLE-peak, where the maximum of the score output is used as latency estimate.
 
@@
 model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
-We refer to~\textcite{Gramfort2014} for implementation details.
-Pink spatial and  temporal noise are then added.
-Temporal noise is generated by passing gaussian noise through an infinite
+Pink temporal noise is generated by passing gaussian noise through an infinite
 impulse response filter with transfer function
 $\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
-For pink spatial noise, a noise covariance matrix is constructed from the
-cosine distances between electrode positions and scaling the resulting spatial
-frequencies by $1/f$.
+Pink spatial noise is also added, using a noise covariance matrix constructed from the
+cosine distances between electrode positions and scaling it by the inverse of
+the resulting spatial.
+We refer to~\textcite{Gramfort2014} for implementation details.
 
@@
 
-figure
+A sample of the simulated evoked data is displayed in
+Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
+\ref{fig:wcble/sim-sine}.
+\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
+\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
+\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
+\todo{Properly align plots and label axes}
+
+\todo{figure}
 
 \subsection{Latency estimation}
-cble with prob/median
-wcble with prob+median
-cble with max
-wcble with max
 
@@
 \section{Discussion}
+
 \todo{mention cross validation}
 \todo{remove lcmv}
-In this work, we introduce a new latency method that leverages data-driven,
+
+\todo{In this work, we introduce a new latency method that leverages data-driven,
 joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
@@
 each \ac{erp} component, as well as it's latency distribution, and takes into acount the
-spatial and temporal correlation of the EEG background noise.
+spatial and temporal correlation of the EEG background noise.}
 
-with lda, it is in the single class case equal to shifting a spatiotemporal
+\todo{with lda, it is in the single class case equal to shifting a spatiotemporal
 lcmv beamformer over the epochs and can be used to extract latencies.
-multiclass:
+multiclass:}
 

### 2024-09-24 main.tex
Add gaze disabilities
Add decoders
Comments Liesel
Intro, P300 stimulation

@@
 }
+\newcommand*\elide{\textup{[\,\dots]}}
 
 % Bibliography
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber, doi=false, url=false, isbn=false, eprint=false]{biblatex}
 \addbibresource{references.bib}
@@
 
-% Title
+% Set bibliography font size
+\renewcommand*{\bibfont}{\footnotesize}
+
+\usepackage[inline]{enumitem}
 
@@
 \part{Introduction \& Literature}
-\includechapter{Brain-Computer Interfaces}{chapters/bci.tex}
+\includechapter{Brain-computer interfaces}{chapters/bci.tex}
 \includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
@@
 \part{Experiments \& Applications}
-\includechapter{Compensating jitter for gaze-independent decoding}{chapters/covert_align.tex}
+\includechapter{Compensating jitter for gaze-independce}{chapters/covert_align.tex}
 \todo{Integrating gaze-tracking data}
@@
 \chapter{Funding \& acknowledgments}
+\todo{my grant}\todo{marc's projects}
 \chapter{Personal contribution}
+\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
+Hakim, Marie,Etienne, Philip Van Damme  }
 \chapter{Conflict of interest}

### 2024-09-24 code/setup.py

@@ -1,0 +1,2 @@
+mm = 1 / 25.4  # centimeters in inches
+pagesize = (160 * mm, 240 * mm)

### 2024-09-24 code/wcble/fig_simulated.py

@@ -1,0 +1,120 @@
+import ipdb
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from mne import read_epochs
+from setup import pagesize
+
+epochs = read_epochs("data/wcble/simulated_epochs.fif")
+meta = epochs.metadata
+
+epochs.pick(["C3", "STIM"])
+epochs.crop(-2, 2)
+
+stcs = meta["waveform"].unique()
+snrs = meta["snr"].unique()
+# snrs = [0, -16, -31]
+jitters = meta["jitter"].unique()
+
+
+img_kwargs = dict(
+    vmin=-50,
+    vmax=50,
+    picks="C3",
+)
+
+
+def latency(epochs):
+    ground_truth = np.zeros(len(epochs))
+    for i, epoch in enumerate(epochs):
+        stim_chan = epoch[-1, :]
+        ground_truth[i] = np.nonzero(stim_chan)[0]
+    ground_truth = ground_truth / epochs.info["sfreq"] + epochs.tmin
+    return ground_truth
+
+
+def order(epochs, n_epochs=100):
+    latency_ = latency(epochs)
+    order = np.zeros(len(epochs), dtype=int)
+    order[:n_epochs] = np.argsort(latency_[:n_epochs])
+    order[n_epochs:] = np.arange(n_epochs) + n_epochs
+    return order
+
+
+for i, stc in enumerate(stcs):
+    shape = (2 * len(snrs), len(jitters))
+    fig, axs = plt.subplots(
+        *shape,
+        sharex="col",
+        sharey="row",
+        # layout="tight",
+        height_ratios=[3, 1] * len(snrs),
+        figsize=pagesize,
+    )
+    axs = axs.reshape(shape)
+    fig.suptitle(stc)
+
+    for j, snr in enumerate(snrs):
+        for k, jitter in enumerate(jitters):
+            prefix = f"{stc}/{snr}/{jitter}"
+            print(prefix)
+            curr_epochs = epochs[prefix]
+
+            ax_img = axs[j * 2, k]
+            ax_ts = axs[(j * 2) + 1, k]
+
+            order_ = order(curr_epochs)
+            curr_epochs.plot_image(
+                order=order_,
+                colorbar=False,
+                evoked=False,
+                axes=ax_img,
+                show=False,
+                **img_kwargs,
+            )
+            label = f"SNR={snr}dB, $\sigma$={jitter}s"
+            ax_img.text(
+                0.05,
+                0.95,
+                label,
[…]

### 2024-09-24 chapters/bci.tex
Set manuscript draft

@@
 These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
-(fMRI) to more portable technologies like accoustic signals obtained by functional ultrasound imaging
+(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
 (fUS)~\cite{Zheng2023} and near-infrared spectroscopy
@@
 offer great temporal resolution up to thousands of Hz.
+\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
+reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
 As it is easily applicable on the outside of the scalp, many practical
@@
 Current consumer systems sometimes feature dry electrodes, but clinical and
-research systems often use electrolyte gel to reduce impedance for improved signal
-quality, which slightly decreases practicality.
+research systems often use active electrodes with preamplifiers and
+electrolyte gel to reduce impedance for improved signal
+quality, which slightly decreases practicality, since they require a trained
+professional to properly administer.
 Nevertheless, it is the most accessible \ac{bci} technology for both patients
@@
 inclusivity, harmful to progress, paraphrase abstracts of articles}
+\todo{user training more than with reactive, task instruction}
 
@@
 short period of time by changing in color, brightness or size.
+Intensifcations of different targets are usually 100-200ms apart and last for about
+100ms~\cite{Sellers2006a}.
 On each flash observed by the user, specific brain activity is elicited.
@@
 images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
-stimuli~\cite{Treder2011};
+stimuli~\cite{Treder2011}.
+The number of targets in a visual oddball \ac{bci}is limited by the crowding
+effect, which imposes a limit on how close targets can be to each other while
+not distracting the user's attention~\cite{Sellers2006a,Li2010}.
+This may be overcome by making hierarchical selections of stimuli representing
+groups of selections~\cite{Treder2010}.
 
-Sometimes Flashing in various configurations, such as row-column or in groups,
-other flashing patterns
-faces, feature attention
-visual hex-o-spell
+Together with \ac{ssvep}, oddball paradigms are most frequently used in visual
+\ac{bci}.
+There are however indications that users prefer oddball stimulation over
+\ac{ssvep} which can cause eye strain and fatigue due to the continuous
+oscillatory stimulation of all targets at once~\cite{Xu2021}.
+Furthermore, \ac{ssvep} relies more on directing the gaze at the intended
+target than oddball, detracting from the concept of control independent from
+muscle movement.
 
-comfortable, preferred over SSVEP, similar ITR is this true?
-\todo{figure of P3 matrix}
-\todo{cite P300 literature, first example and current sota}
+\todo{four pane figure:
+  photo of someone using a P300 bci
+  P300 matrix interface
+  simplified figure of ERP components
+  full spatiotemporal ERP response
+}
 
@@
 
-later compoentns P3 more attention/cognitive more centrally in the brain
+The time-domain response ellicited immediately after intensification of a
+stimulus is the visual \ac{erp}~\cite{Luck2014}.
+An \ac{erp} is a waveform consisting of multiple components, of wich some are
+modulated by whether the stimulus was an oddball or not.
+Components are positive peaks and negative troughs in the \ac{erp} waveform.
+They following a naming scheme according to their polarity (Negative or
+Positive) and their latency (e.g. 3 or 300ms).
+The component most prominently modulated by attention is called the P300 or P3
+(positive deflection after $\pm$300ms) and oddball paradigms are therefore
+frequently referred to as P300 paradigms.
+However, due to the equally important contribution of other \ac{erp}
+components than the P3 in decoding as we will show later, this could be
+considered a bit of a misnomer and we will stick to the oddball paradigm naming.
+P300 and other components, explain function (Luck)
+Which components, at what amplitude and latency they are evoked varies per
+person and depends on the exact details of the stimulation.
 
-Because of this, the oddball paradigm is often referred to as the P300
-paradigm.
-However, due to equal importance of other \ac{erp} components than the P3 in
-decoding as we will show later, this could be considered a bit of a misnomer
-and we will stick to the oddball paradigm naming.
+In general, earlier components, such as the \acp{vep}, are more tied to sensory
+information processing and less modulated by attention, while later components
+are more related to attention and cognitive processes.
 
-mention VEPs more from visual system occipitally
-
-P300 and other components, explain function (Luck)
-
-\todo{figure of ERP components}
-
-which components, at what amplitude and latency they are evoked varies per
-person and depends on the type of intensification
-
+\textcite{Luck2013}
 \subsubsection{P1}

### 2024-09-24 chapters/conclusion.tex
BCI chapter
Add gaze disabilities
Add decoders
Literature wcble
Set manuscript draft

@@
-\chapter{Conclusions \& recommendations}
+\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
+\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
+\todo{BCI principles and practice chapter 19}
+
 \section{General discussion}
-\section{Limitations}
-\section{Future directions}
-\section{Working with patients}
-\section{Conclusion}
 
-\section{Obtained results}
+\subsection{Obtained results}
 
-\subsection{Regularized spatiotemporal beamforming}
+\subsubsection{Regularized spatiotemporal beamforming}
 We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
@@
 
-\subsection{Healthy control study}
+\subsubsection{Healthy control study}
 The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
@@
 compared to overt VSA operation.
+
+\section{Limitations}
+
+\section{Current \& future work}
+multicomponent alignment approach
+hankel tensors to cope with jitter
+select best classifier using eye tracking
+
+\section{Recommendations}
+\subsection{Working with patients}
+start from patients
+use all available signals
+\subsection{Linear decoding methods}
+
+\section{Conclusion}

### 2024-09-24 chapters/covert_align.tex
Literature wcble
Set manuscript draft

@@
 \section{Introduction}
+\todo{mention cble and wcble in methods}
 
@@
 Including these longer breaks, the experiment lasted approximately one hour.
-\subsubsection{Riemannian Geometry}
-We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
-approaches that rely on spatial covariance as features.
-Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
-performance~\cite{Lotte2018}.
-We implemented two Riemannian Geometry pipelines.
-The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
-filters, projects these covariances to a tangent space and classifies the
-result using $L2$-regularized logistic regression
-(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
-Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
-favorable performance in the presence of single-trial ERP latency jitter.
-Shrunk spatial covariance matrices are estimated from epochs that are
-augmented by concatenating the average target and average non-target ERP as
-extra channels, projected to tangent space and classified using
-$L2$-regularized logistic regression (ERPCov-TS-LR).
 
@@
 to \cite{Aloise2012a}.
+
+\subsubsection{Decoders}
+We will compare \ac{wcble} and \ac{cble} ad described in
+Section~\ref{sec:wcble/methods}, with their base classifier and tLDA and with
+Riemannian Geometry
+approaches that rely on spatial covariance as features.
+Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
+performance~\cite{Lotte2018}.
+We implemented two Riemannian Geometry pipelines.
+The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
+filters, projects these covariances to a tangent space and classifies the
+result using $L2$-regularized logistic regression
+(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
+Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
+favorable performance in the presence of single-trial ERP latency jitter.
+Shrunk spatial covariance matrices are estimated from epochs that are
+augmented by concatenating the average target and average non-target ERP as
+extra channels, projected to tangent space and classified using
+$L2$-regularized logistic regression (ERPCov-TS-LR).
 

### 2024-09-24 chapters/wcble.tex
Set manuscript draft

@@
 \todo{comments liesel}
-
-This chapter, except Sections~\ref{sec:wcble/intro},
-\ref{sec:wcble/literature/single-comp} and
-\ref{sec:wcble/literature/multi-comp}
-\todo{properly assess which paragraphs were published and which not}
+Sections~\ref{sec:wcble/literature/contrib} were published \ref{sec:wcble/methods}
+by~\textcite{VanDenKerchove2024}
 
@@
 \subsection{Contribution}
+\label{sec:wcble/literature/contrib}
 
@@
 \section{Methods}
+\label{sec:wcble/methods}
 
 \subsection{Classifier-based Latency Estimation}
-\label{sec:cble}
+\label{sec:wcble/methods/cble}
 
@@
 
-\ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
+\Ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
 \textit{first-stage} classifier $\mathcal{C}(\theta,f)$
@@
 	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l},
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
 		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
@@
 
-We compare our proposed \ac{wcble} algorithm to \ac{cble} on both a latency
-estimation task and a binary classification task with simulated data.
-Four settings are evaluated to assess both the impact of the iterative Woody
-procedure and the extraction of robust latency features.
-These are CBLE-median and WCBLE-median, where the median of the latency
-probability distribution is used as latency estimate, and CBLE-peak and
-WCBLE-peak, where the maximum of the score output is used as latency estimate.
+%Four settings are evaluated to assess both the impact of the iterative Woody
+%procedure and the extraction of robust latency features.
+%These are CBLE-median and WCBLE-median, where the median of the latency
+%probability distribution is used as latency estimate, and CBLE-peak and
+%WCBLE-peak, where the maximum of the score output is used as latency estimate.
 
+\todo{
+refer to later chapter for use on real erp data
+}
 \subsection{Data simulation}
-To verify our approach, we simulated 32 channel \ac{eeg} data with standard
+To verify our approach, we simulated 16 channel \ac{eeg} data with standard
 10-20 electrode positions using the \texttt{simulate\_evoked} method of the
 MNE (version 1.8)~\cite{Gramfort2013} software package.
-\Acp{erp} were generated by projecting a source time course in a dipole in the
+\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
 left hemisphere to the scalp electrodes using a boundary element method forward
@@
 We refer to~\textcite{Gramfort2014} for implementation details.
-
-Evoked potentials are simulated for three different source time course waveforms,
-a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
-defined by the following formulas:
+%Evoked potentials are simulated for three different source time course waveforms,
+%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
+%defined by the following formulas:
+%\begin{equation}
+%  f_\text{gaussian}(t,l) = a
+%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
+%\end{equation}
+%\begin{equation}
+%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
+%\end{equation}
+The source time course is defined by the following function:
 \begin{equation}
-  f_\text{gaussian}(t,l) = a
-  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
-\end{equation}
-\begin{equation}
-  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
-\end{equation}
-\begin{equation}
-  f_\text{sine}(t,l) =
+  s(t,l) =
   \begin{cases}
@@
 \end{equation}
-with time $t$, latency $l$, and amplitude $a=\num{1e{-7}}$V.
-The standard deviation that determines the width gaussian $\varsigma=0.1s$.
-For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
-For the sine pulse, the frequency $f=4$
-For each of these, 100 target epochs containing the waveform and noise and 100
+over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
+$f=4$.
+%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
+%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
+%For the sine pulse, the frequency $f=4$
+
+%For each of these,
+100 target epochs containing the evoked potential and noise and 100
 non-target epochs containing only noise were simulated.
-Additionally, the target epochs were jittered by setting the latency offset $l$
+The target epochs were jittered by setting the latency offset $l$
 of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
 deviations of respectively $\sigma=0.1,0.2,0.3$s.
-The noise was also scaled at different 32  different noise levels, with the \ac{snr} of a
+The noise was scaled at different 32  different noise levels, with the \ac{snr} of a
 target epoch ranging from 0 to -31dB, with
 \begin{equation}
-  \text{SNR} = 10\log_{10}\frac{\text{var}f(t,0)}{\text{var}\left(\text{noise}\right)}
+  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
 \end{equation}
 
-A sample of the simulated evoked data is displayed in
-Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
-\ref{fig:wcble/sim-sine}.
-\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
-\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
+%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
+A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
+%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
+%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
 \fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
@@
 
-CV/LOOCV?
+We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
+estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
+jitter and \ac{snr} levels.
+For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
+10-fold cross-validation scheme using the robust median latency method after
+fitting the model.
+\Ac{wcble}'s maximum iteration number was set to 64.
+Results are presented in Figure~\ref{fig:wcble/results/latency}.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{figures/wcble/fig_latency.pdf}
+  \caption{align results}
+  \label{fig:wcble/results/latency}
+\end{figure}
+\todo{invert x axis}
+\todo{caption}
+\todo{properly label axes}
+\todo{discuss results}
+
 \subsection{Classification}
-refer to later chapter for use on real erp data
+Similar to the previous experiment, classification accuracy between target and
+non-target epochs is assessed.
+
+\todo{insert results figure and discuss}
 

### 2024-09-24 code/wcble/fig_latency.py

@@ -1,0 +1,10 @@
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+
+df = pd.read_csv("data/wcble/results_align.csv")
+print(df)
+sns.relplot(
+    data=df, x="snr", y="rho", col="jitter", row="stc", hue="method", kind="line"
+)
+plt.savefig('figures/wcble/fig_latency.pdf')

### 2024-09-25 figures/bci/paradigms.tex
Refactor structure

\documentclass[tikz,border=10pt]{standalone}
\usepackage{fontspec}

\begin{document}
\begin{tikzpicture}

    % Define custom colors
    \definecolor{darkgray}{RGB}{38,35,33}
    \definecolor{lightgray}{RGB}{143,130,125}
    \definecolor{accent1}{RGB}{242,144,24} % Formerly orange
    \definecolor{accent2}{RGB}{246,61,30}  % Formerly red
    \definecolor{accent3}{RGB}{120,71,157}  % Formerly purple
    \definecolor{white}{RGB}{255,255,255}

    % Draw the horizontal line
    \draw[darkgray, thick, ->] (36, 120) -- (124, 120);

    % Draw the vertical line
    \draw[darkgray, thick, ->] (80, 76) -- (80, 164);

    % Draw rectangles
    \draw[accent1, thick] (39, 79) rectangle (73.5, 113.5);
    \draw[accent2, thick] (86, 79) rectangle (120.5, 113.5);
    \draw[accent3, thick] (39, 125) rectangle (120, 158);

    % Add text
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{white}] at (56.5, 90) {};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{lightgray}] at (64.7, 70) {active};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{lightgray}] at (74, 172) {passive};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{lightgray}] at (19, 118) {stimulus\\independent};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{lightgray}] at (139, 118) {stimulus\\dependent};

    % Text in rectangles
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (109, 136) {error\\detection};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (97, 93) {P300};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (56, 90) {motor\\imagery};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (56, 101) {imagined\\speech};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (96, 103) {SSVEP};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (110, 93) {cVEP};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (109, 104) {mVEP};

    % More text
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (56, 133) {attention\\monitoring};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (56, 146) {emotion\\monitoring};
    \node[align=center, font=\fontsize{8}{10}\sffamily\color{darkgray}] at (90, 146) {media induced\\emotion recognition};

\end{tikzpicture}
\end{document}

### 2024-09-25 chapters/bci.tex
Refactor structure

@@
 Starting in the 1970's from methods to establish minimal brain activity
-(yes/no, left/right, ...)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
+(yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
 into sophisticated interaction schemes that restore or enhance many capabilities
@@
 behavior or brain activity accordingly, closing the loop
-\footnote{
+\footnote{%
   \textcite{BCISociety2024} has recently formalized this into the
@@
 \todo{why is spatial resolution a good thing}
-\todo{explain with each method what it actually is (e.g. electrodes placed
+\todo{explain with each method what it actually is (e.g.\ electrodes placed
 at...)}
@@
 involved with brain surgery.
-\fullpagefigct{figures/bci/recording_modalities.pdf}{
+\fullpagefigct{figures/bci/recording_modalities.pdf}{%
 	Different electrophysiology recording modalities and \\

### 2024-09-26 chapters/bci.tex
Add ERP section

@@
 
-\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci} loop.
+\fullpagefigct{%
+  \includegraphics[width=\paperwidth]{figures/bci/bci_loop.pdf}
+}{%
+  The \ac{bci} loop.
   The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
@@
   In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
-}{fig:bci-loop}
+}{%
+  fig:bci/loop
+}
 
@@
 The recorded brain waves form only one part of the interaction scheme presented
-in Figure~\ref{fig:bci-loop}.
+in Figure~\ref{fig:bci/loop}.
 It would make little sense to start digging around at random in all the brain
@@
 technology is suited for a \ac{bci} application is the \emph{invasiveness} of the technology.
-As illustrated in Figure~\ref{fig:recording-modalities}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
+As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
 involved with brain surgery.
-\fullpagefigct{figures/bci/recording_modalities.pdf}{%
+
+\fullpagefigct{%
+  \includegraphics[width=\paperwidth]{figures/bci/recording_modalities.pdf}
+}{%
 	Different electrophysiology recording modalities and \\
@@
 	spatial resolution.
-}{fig:recording-modalities}
+}{%
+  fig:bci/recording
+}
 \todo{find better figure with permission to reuse and without unnecessary
@@
 
-\fullpagefig{figures/bci/paradigms.eps}{
+\fullpagefig{%
+  \input{figures/bci/paradigms.tikz.tex}
+}{%
   A \ac{bci} paradigm defines which brain activity will be used for control.
@@
   Adapted from~\textcite{Muehl2014}.
-}{fig:bci-paradigms}
+}{
+  fig:bci-paradigms
+}
 \todo{check permissions}
@@
 
-The time-domain response ellicited immediately after intensification of a
-stimulus is the visual \ac{erp}~\cite{Luck2014}.
-An \ac{erp} is a waveform consisting of multiple components, of wich some are
-modulated by whether the stimulus was an oddball or not.
-Components are positive peaks and negative troughs in the \ac{erp} waveform.
-They following a naming scheme according to their polarity (Negative or
-Positive) and their latency (e.g. 3 or 300ms).
-The component most prominently modulated by attention is called the P300 or P3
-(positive deflection after $\pm$300ms) and oddball paradigms are therefore
-frequently referred to as P300 paradigms.
-However, due to the equally important contribution of other \ac{erp}
-components than the P3 in decoding as we will show later, this could be
-considered a bit of a misnomer and we will stick to the oddball paradigm naming.
-P300 and other components, explain function (Luck)
-Which components, at what amplitude and latency they are evoked varies per
-person and depends on the exact details of the stimulation.
+The time-domain response elicited immediately after the intensification of a
+stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
+An \ac{erp} is a waveform consisting of multiple components.
+Some of these components are modulated by whether the stimulus was an
+oddball or not.
+These components appear as positive peaks and negative troughs in the \ac{erp}
+waveform.
+They follow a naming scheme based on polarity (Negative or Positive) and
+latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
+The component most prominently modulated by attention is called the P3
+(positive deflection after $\pm300$ ms).
+Therefore, oddball paradigms are frequently referred to as P300 paradigms.
+However, due to the equally important contributions of other \ac{erp}
+components in decoding, as we will show later, this could be considered a
+misnomer.
+Thus, we will adhere to the oddball paradigm naming.
 
-In general, earlier components, such as the \acp{vep}, are more tied to sensory
-information processing and less modulated by attention, while later components
-are more related to attention and cognitive processes.
+The visual \ac{erp} components primarily include P1, N1, P2, N2, and P3.
+Each component is characterized by specific latency, amplitude, and neural
+origins.
+These factors influence the perception of visual stimuli and attentional
+mechanisms~\cite{Luck2013}.
 
-\textcite{Luck2013}
-\subsubsection{P1}
-\subsubsection{N1}
-\subsubsection{P2}
-\subsubsection{N2}
-\subsubsection{P3}
-\subsubsection{Late negativity}
+\paragraph{P1}
+The P1 component occurs approximately 100 ms after stimulus presentation.
+This component reflects initial visual processing and is primarily linked to
+activity in the primary visual cortex.
+P1 is particularly sensitive to gaze fixation, where the gaze is directed
+toward a stimulus.
+When a stimulus is fixated on, the amplitude of P1 increases, reflecting the
+enhanced processing of that visual input.
+In contrast, P1 shows less modulation when stimuli are attended to via
+attention without direct gaze.
 
-\subsection{Preprocessing and feature extraction}
-referencing, filtering, epoching, baselining
+\paragraph{N1}
+The N1 component peaks around 150 to 200 ms post-stimulus.
+It represents an extension of sensory processing related to both gaze fixation
+and attention.
+For attended stimuli, the amplitude is greater, indicating prioritization for
+further cognitive processing.
+This highlights a more selective form of attention allocation, demonstrating
+the impact of both types of attention on visual processing.
+
+\paragraph{P2}
+The P2 component occurs between 200 to 300 ms.
+It reflects higher-order processing, particularly in contexts demanding
+stimulus evaluation and feature detection.
+P2 is sensitive to attention.
+When participants actively engage with specific features or categories of
+stimuli, P2 shows increased amplitude.
+Conversely, P2 may be less responsive when gaze is not directed toward the
+stimulus but attention is still maintained.
+
+\paragraph{N2}
+Peaking around 200 to 350 ms, the N2 component is associated with cognitive
+control processes such as conflict monitoring and inhibition.
+It is particularly pronounced in tasks requiring differentiation between
+competing stimuli.
+N2 reflects the allocation of attentional resources required to resolve
+conflicts, irrespective of whether the gaze is directly on the competing
+stimuli or not.
+Higher amplitudes in N2 are seen when cognitive demands increase, indicating
+the influence of attention strategies.
+
+\paragraph{P3}
+The P3 component is elicited in oddball paradigms and peaks around 300 ms
+post-stimulus.
+It reflects attentional engagement and the processing of rare or unexpected
+events.
+The P3 is typically subdivided into two subcomponents: P3a and P3b.
+The P3a is associated with the allocation of attention to novel or unexpected
+stimuli, indicating the initial detection of a change in the environment.
+It is particularly responsive to stimuli that capture attention, whether
+through gaze fixation or attention.
+P3a typically exhibits a frontal distribution and peaks earlier than P3b.
+
+Conversely, the P3b relates to the evaluation of the stimulus.
+It involves the updating of cognitive resources in response to task relevance.
+This subcomponent is typically observed over parietal regions.
+P3b is elicited when participants must process the stimulus meaningfully,
+often requiring a decision or response.
+Both P3a and P3b amplitudes are sensitive to the probability of occurrence
+and the participant's attentional focus, showing how attentional strategies
+affect cognitive processing.
+
+\paragraph{Late Positive Potential (LPP)}
+The LPP is a later component occurring beyond 400 ms.
+Sustained attention and emotional processing of stimuli are reflected in the
+LPP.
+Typically enhanced for emotionally salient or motivationally relevant
+information, LPP serves as an indicator of cognitive appraisal beyond mere
+recognition tasks.
+This component can be influenced by attention when the participant evaluates
+the emotional significance of stimuli, even if gaze is not directed at them.
+
+In summary, ERP components such as P1, N1, P2, N2, P3, and LPP collectively
+elucidate the neural underpinnings of visual processing and attentional
+mechanisms.
+These components not only inform the design of an effectiv oddball BCI system
+but also enhance our understanding of the cognitive processes involved in visual
+attention.
+This includes both gaze fixation and attention.
+Understanding these components is essential for optimizing BCI protocols and
+improving user interaction, particularly in contexts involving visual oddball
+tasks.
+
 \subsection{Decoding ERPs}

### 2024-09-26 chapters/gaze_independence.tex
Add ERP section

@@
 Figure~\ref{fig:gaze-vsa}.
-\fourpanefig
-  {\figpane{figures/gaze_independence/attention_overt.eps}{
+\fourpanefig{%
+  \figpane{%
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
+  }{%
     \emph{Overt \ac{vsa}}.
-    Gaze and \ac{vsa} coincide on a target.}{fig:gaze-vsa-overt}}
-  {\figpane{figures/gaze_independence/attention_covert.eps}{\emph{Covert
-  \ac{vsa}}.
-  Gaze rests on the center of the screen, while \ac{vsa} is directed towards a
-  target.}{fig:gaze-vsa-covert}}
-  {\figpane{figures/gaze_independence/attention_split.eps}{
-    \emph{Split \ac{vsa}}. \ac{vsa} is directed towards a target, while the
-    gaze rests on another.}{fig:gaze-vsa-split}}
-  {\figpane{figures/gaze_independence/attention_free.eps}{\emph{Free \ac{vsa}}.
+    Gaze and \ac{vsa} coincide on a target.
+  }{%
+    fig:gaze/vsa/overt
+  }
+}{%
+  \figpane{%
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
+  }{%
+    \emph{Covert \ac{vsa}}.
+    Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
+  }{%
+    fig:gaze/vsa/covert
+  }
+}{%
+  \figpane{%
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
+  }{%
+    \emph{Split \ac{vsa}}.
+    \Ac{vsa} is directed towards a target, while the gaze rests on another.
+  }{%
+    fig:gaze/vsa/split
+  }
+}{%
+  \figpane{%
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
+  }{%
+    \emph{Free \ac{vsa}}.
     The user is free to direct their gaze as they deem most comfortable
-  All previous \ac{vsa} conditions are possible.}{fig:gaze-vsa-free}}
-  {\Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.}
-  {fig:gaze-vsa}
+    All previous \ac{vsa} conditions are possible.
+  }{%
+    fig:gaze/vsa/free
+  }
+}{%
+  \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
+}{%
+  fig:gaze-vsa
+}
 

### 2024-09-26 chapters/wcble.tex
Add ERP section

@@
 \todo{comments liesel}
-Sections~\ref{sec:wcble/literature/contrib} were published \ref{sec:wcble/methods}
+Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
 by~\textcite{VanDenKerchove2024}
@@
 positive, by the difference between the latency and the window onset.
-\fullpagefig{figures/wcble/figure1.pdf}{Schematic representation of the Woody Classifier Based Latency
+\fullpagefig{%
+  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
+}{%
+  Schematic representation of the Woody Classifier Based Latency
   Estimation training phase.
@@
   from the last iteration are used to train the second-stage
-    classifier.}{fig:diagram-train}
+  classifier.
+}{%
+  fig:diagram-train
+}
 The process halts after a fixed amount of iterations or when the estimated set
@@
 in~Section~\ref{sec:robust-latency}.
-\fullpagefig{figures/wcble/figure2.pdf}{Schematic representation of the Woody Classifier Based Latency
+\fullpagefig{
+  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
+}{%
+  Schematic representation of the Woody Classifier Based Latency
   Estimation test phase. (1) The
@@
 	as the new class latencies. (3) The scores and class latencies are input to the
-    trained second-stage classifier, which predicts the label of the epochs.}{fig:diagram-eval}
+  trained second-stage classifier, which predicts the label of the epochs.
+}{%
+  fig:diagram-eval
+}
 \section{Algorithms}
@@
 %\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
-\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
+\fullpagefig{\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}}{}{fig:wcble/sim-sine}
 \todo{Properly align plots and label axes}

### 2024-09-26 figures/bci/paradigms.tikz.tex
Add ERP section

\begin{tikzpicture}[scale=\textwidth/2cm]
    \pgfmathsetmacro{\margin}{.1}
    \pgfmathsetmacro{\center}{.5+.5*\margin}
    \pgfmathsetmacro{\textspacing}{.15}

    % Draw axes
    \draw[darkgray, very thick, <->] ({-(1+\margin)}, 0) -- ({1+\margin},0);
    \draw[darkgray, very thick, <->] (0,{-(1+\margin)}) -- (0,{1+\margin});

    % Draw rectangles
    \draw[draw=accent1, fill=white, very thick] ({-\margin}, \margin) rectangle (-1,1);
    \draw[draw=accent2, fill=white, very thick] (\margin, \margin) rectangle (1,1);
    \draw[draw=accent3, fill=white, very thick] (-1, {-\margin}) rectangle (1,-1);

    %% Add text
    \node[color=accent1, align=left,font=\bfseries, anchor=north west] at  (-1,1){active};
    \node[color=accent2, align=left,font=\bfseries, anchor=north east] at  (1,1){reactive};
    \node[color=accent3, align=left,font=\bfseries, anchor=south west] at  (-1,-1){passive};

    \node[color=darkgray, align=center,font=\bfseries, anchor=north] at (0,{-(1+\margin)}) {passive};
    \node[color=darkgray, align=center,font=\bfseries, anchor=south] at (0,{1+\margin}) {active};
    \node[color=darkgray, align=center,font=\bfseries, anchor=east] at ({-(1+\margin)}, 0) {stimulus\\independent};
    \node[color=darkgray, align=center,font=\bfseries, anchor=west] at ({1+\margin},0) {stimulus\\dependent};

    %% Text in rectangles
    \node[align=center] at ({-\center}, {\center+\textspacing}) {motor\\imagery};
    \node[align=center] at ({-\center}, {\center-\textspacing}) {imagined\\speech};

    \node[align=center] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
    \node[align=center] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
    \node[align=center] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
    \node[align=center] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};

    %% More text
    \node[align=center] at (\center, {-\center}) {error\\potentials};
    \node[align=center] at ({-\center}, {-\center}) {attention\\monitoring};
    \node[align=center] at (-.25, -.8) {emotion\\monitoring};
    \node[align=center] at (.25, -.8) {media induced\\emotion recognition};
\end{tikzpicture}

### 2024-09-26 chapters/bci.tex
Add ERP section

@@
 \todo{comments liesel paradigms}
-People suffering from heavy paralysis or disability due to
-neurodegenenerative diseases, stroke, traumatic brain or spine injury
+People suffering from severe paralysis or disability due to
+neurodegenenerative disease, stroke, traumatic brain or spine injury
 become entirely reliant on caregivers or family members in their day to day
@@
 on the user's behalf.
-If the mind's seat is in the brain, and we want to interface this to a digital
-control process, such a device becomes the \ac{bci}\footnote{Sometimes also
+Such a device is the \ac{bci}\footnote{Sometimes also
 termed brain-machine interface (BMI) when coupled to a physical actuator, like
@@
 
-Starting in the 1970's from methods to establish minimal brain activity
-(yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
+Starting in the 1970's from methods to establish basic communication from
+minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
 into sophisticated interaction schemes that restore or enhance many capabilities
-by bypassing part of the nervous system and muscle control.
+by bypassing defective parts of the nervous or muscle system.
 By now, classic \acp{bci} methods have matured outside the lab setting and can be
@@
 \section{A direct interface to the brain}
-\comment{This section describes a bci in layman's terms, without too much
-scientific background, jargon and literature, in the following sections of this
-chapter a more scientific style is used}
-\ac{bci}s are an entertaining subject with many futuristic and more established
+
+Brain-computer interfacing is an entertaining subject with many futuristic and more established
 applications.
@@
 
-Simply put, a \ac{bci} records the user's brain activity, then extracts some
+Simply put, a \ac{bci} records the user's brain activity, extracts some
 relevant output from this brain activity and couples this output to a function
@@
 Section~\ref{sec:bci-recording} gives an overview of the available options.
-One might for instance measure the fluctuations in bloodflow to specific brain
-regions that are more or less active at a given time with a scanner.
+One might for instance measure the fluctuations in bloodflow to establish the
+activity of a specific brain region
 However, this signal reacts too slowly to reflect the real-time activity that is
@@
 and determine which one of a set of targets was attended.
-In turn, this can be coupled to an action, like typing the letter shown on the
-stimulus.
+In turn, this can be coupled to an action, like typing the inteded letter A or B
+in Figure~\cite{fig:bci/loop}.
 The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
@@
 The previous points to another component of a successful \ac{bci}: the specific
-brain signals related to our paradigm need to be identified within the recorded
+brain signatures related to our paradigm need to be identified within the recorded
 activity.
@@
 Furthermore, the brain is continuously processing information and carrying out
-`background tasks', all of which generate their own electrical signals.
+`background tasks', all of which generate their own brain activity.
 On the other hand, the activity of interest in often
-originates from a specific functional system within the brain.
-This activity might not easily be discerned from all other neural and
+originates from a specific region or network within the brain.
+This activity might not easily be discerned from all other brain and
 environmental activity.
@@
 \emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
-relevant signal is often solved through supervised machine learning.
+relevant signal is generally hard.
+Nevertheless, we can attempt to solve it through supervised machine learning.
 Useful information that is encoded in the brain signal by the paradigm,
@@
 
-Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
+Larger implanted electrodes are referred to as intracranial ac{eeg} (i\ac{eeg}).
 Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
@@
 quality, which slightly decreases practicality, since they require a trained
-professional to properly administer.
+individual to properly administer.
 Nevertheless, it is the most accessible \ac{bci} technology for both patients
@@
 
-Reactive \ac{bci} take another approach, by giving the user a discrete set of
+Reactive \ac{bci} take another approach, by providing the user a discrete set of
 sensory stimulations and decoding towards which stimulus the user's attention
@@
 
-The major visual paradigms comprise
-\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} type interfaces.
+The major visual paradigms rely on
+\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} brain responses.
 In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
@@
 stimulates them one by one with single flashing.\todo{reference}
-Since the times of stimulation of each target is known by the interface,
-information is modulated by time of stimulation, and times where an attention
-response can be found correspond to selected targets.
+Since the times of stimulation of each target is known, and time-points at
+which an attention signature is detected to selected targets.
 \Ac{mvep} is usually similarly time modulated, but stimuli make sudden
@@
 in a visual \ac{erp} \ac{bci}.
-Some valuable metrics to asses decoder performance in the classification case
-are accuracy and \ac{rocauc}.
+Accuracy is a valuable metrics to asses decoder performance in the classification
+case.
 Accuracy is calculated as the proportion of correct selections to all
@@
 selections per trial.
-\Ac{rocauc} is also a measure of classifier predictive power, but is more
+Alternatively, \ac{rocauc} is also a measure of classifier predictive power, but is more
 suited for evaluation of classification of single epochs of data and in the
@@
 by paying explicit attention to only a given stimulus and not the others or
-counting how many times it flashes.
+counting how many times it occurs.
 If these visual elements represent letters, and we know the timing of the
@@
 
-\subsection{The event-related potential}
+\subsection{The \acf{erp}}
 

### 2024-09-26 chapters/bci.tex
Comments Marc

@@
 \todo{comments liesel paradigms}
+\todo{spellcheck}
+\todo{are all acronyms defined}
+\todo{fix all latex warnings to make sure no refs are missing (or grep on that
+specific warning}
+
 People suffering from severe paralysis or disability due to
@@
 In it's core essence, this is the ability to communicate your intentions,
-emotions, frustrations or thoughts to someone or something, which then might
-act upon them.
+emotions, frustrations or thoughts such that they might be acted on.
 For an able-bodied person, physical interaction happens through body movements
-and communication is usually done through speech and body language, which
-both require sending signals through the body's nervous system to control
+while communication is usually done through speech and body language
+Both require sending signals through the body's nervous system to control
 muscles.
@@
 Recently, the idea has also gained popular traction through Elon Musk's
-Neuralink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.
+NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.
 
@@
 relevant output from this brain activity and couples this output to a function
-of a device.
+of a device.\todo{refer to fig loop}
 Optionally, the user can then observe the action of the device and adapt their
@@
 brain regions where it is originating.
-A better candidate is the neuronal \emph{action potential}.
+A better candidate is the neuronal electrical activity.
 The brain consists 86 billion neurons, which are highly interconnected cells that are
 the smallest units of information processing.
-The action potential is an electrical pulse occurring when a neuron receives
-input from another neuron and is activated.\todo{gross oversimplification,not
-entirely true}
-A neuron's action potential, or the combined action potentials of groups of
+The \emph{action potential} and its related \emph{postsynaptic potential}
+is an electrical pulse occurring when a neuron receives
+input from another neuron and is activated.
+The firing of a neuron, or the combined firing of groups of
 synchronized neurons, thus generate an electrical field in and near the brain,
@@
 feedback in the case of a closed-loop system.
-As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
+As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
 paradigms based on different systems in the brain.
@@
 Useful information that is encoded in the brain signal by the paradigm,
-can be retrieved from it by the decoder.
-Section~\label{sec:bci-decoding} takes a closer look at some common decoding
+can be retrieved from it by the decoder.\todomvh{the decoder needs to take into
+account the control facilities of the application (degrees of freedom, serial
+/parallel nature, reactive/active...).}
+Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
 techniques.
@@
 More indirectly, modulation can be achieved by sensory stimulation.
-This involves specific, paradigm-related sensory input, like haptic selection
-feedback for movement \ac{bci}, or simply presenting selected actions
+This involves specific, paradigm-related sensory input, like haptic
+feedback\todomvh{or tactile feedback:
+S.N. Flesher et al., “A brain-computer
+interface that evokes tactile sensations improves robotic arm control,”
+Science, vol. 372, no. 6544, pp. 831-836, 2021. } for movement \ac{bci}, or simply presenting selected actions
 back to the user.
@@
 (fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
-(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
+(fUS)~\cite{Zheng2023}\todomvh{non-invasive but not necessarily fUS...} and near-infrared spectroscopy
 (fNIRS)~\cite{Borgheai2020}.
@@
 (MEG)~\cite{Mellinger2007}.
-While MEG is non-invasive and results in high-quality signals, it is relatively
-impractical due to the size of the equipment needed.
+While MEG is non-invasive and results in high-quality signals, the necessary
+equipment is rather expensive and impractical.
 Recent advances have been made using optically pumped magnetometers
@@
 electrical activity of the brain.
-As mentioned earlier with MEG, an important concept in determining which
-technology is suited for a \ac{bci} application is the \emph{invasiveness} of the technology.
+As touched upon earlier,the \emph{invasiveness} of the technology forms an
+important concept in determining its suitability for a given application and
+patient.
 As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
@@
 Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
-NeuraLink implant~\cite{Musk2019}.
+NeuraLink implant~\cite{Musk2019}.\todomvh{+ Neuropixels 2}
 
@@
 \todo{cite} have been realized using
-invasive mictroelectrode arrays~\cite{Willett2021} and high-density
-\Ac{ecog}\cite{Metzger2023}.
+invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
 Furthermore, recent advances in recording technology focus on improving implant
@@
 surgery.
-However, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
+Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
 potential communication \ac{bci} user groups consistently show that non-invasive
-recording methods are preferred over implanted electrodes, unless the added value of invasive BCI is
+technology is preferred over implanted electrodes, unless the added value of invasive BCI is
 sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
@@
 \section{\ac{bci} paradigms}
-\label{sec:bci-paradigms}
+\label{sec:bci/paradigms}
 
 The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity into useful
-output commands, since it defines which neural responses will be elicited and
+output actions, since it defines which neural responses will be elicited and
 captured as features.
@@
 external stimulation and the engagement of the user as shown in
-Figure~\ref{fig:bci-paradigms}.
+Figure~\ref{fig:bci/paradigms}.
 
@@
   Adapted from~\textcite{Muehl2014}.
-}{
-  fig:bci-paradigms
+}{%
+  fig:bci/paradigms
 }
 \todo{check permissions}
+\todo{above Zander and Kothe was mentioned.}
 
@@
 detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
-Furthermore, active paradigms make it easier to quickly collect training data for
+Furthermore, active paradigms\todo{"active" is used in a fuzzy way: active BCI is self-paced, reactive is in response to a cued event. Further down that becomes more clear.
+The "active participation": is needed in both cases as shown in Fig. 1.3.
+I suggest to say active participation} make it easier to quickly collect training data for
 supervized machine learning since the conditions can be more easily controlled
@@
 like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
-symbols~\cite{Willett2021} or sign language.
+symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
 Active paradigms are also subject to large inter-subject variability due to the
@@
 poor performance and giving rise to the concept of \ac{bci}
-illiteracy~\cite{Allison2010}\footnote{
+illiteracy~\cite{Allison2010}\footnote{%
   \cite{Becker2022,Thompson2019}
@@
 
-Reactive \ac{bci} take another approach, by providing the user a discrete set of
-sensory stimulations and decoding towards which stimulus the user's attention
-was directed.
+Reactive \ac{bci} take another approach, by providing a discrete set of
+sensory stimulations towards which the user's attention can directed.
 Compared to active \ac{bci}, reactive paradigms usually require higher effort
 and can induce fatigue due to the constant stimulation.
-These paradigms are also somewhat less intuitive compared to e.g. speech, and
+These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
 their expressive power is limited by how many different stimuli can effectively
@@
 These can be more performant since they allows the user to use their visual
-system, one of the most adapted sensory system in humans.
+system, one of the most evolved sensory system in humans.
 
@@
 only 800ms of stimulation.
-Finally, paradigms can be implemented in a \emph{synchronous} or
-\emph{asynchronous} manner.
-Asynchronous \ac{bci} constantly decodes the signal looking for the response of
-interest, while in synchronous \ac{bci} decisions are made (and eventual
-stimulation takes place) at predetermined time points.
-Again, asynchronous is preferred by users because it is more intuitive and
-comfortable to use, but it is far easier to decode synchronous responses since
-it is known when to expect a response and if a decision should be made.
-
 \section{Decoders}
-\label{sec:bci-decoding}
+\label{sec:bci/decoding}
 
@@
 Multi-linear techniques exploiting the tensor structure of neural signals are
-also promising.
+also promising~\cite{Lotte2018}.
 Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
@@
 suited for evaluation of classification of single epochs of data and in the
-presence of imbalanced data, e.g. when comparing single \acp{erp}.
+presence of imbalanced data, e.g.\ when comparing single \acp{erp}.
 Higher \ac{rocauc} usually translates to a higher target selection accuracy.
 
+\todomvh{"trial" could  be confused with intensification rounds e.g.}
+\todo{remove trial alltogether}
 Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
@@
 \Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
-different options that can be selected per trial, $P$, the selection accuracy
+different options that can be selected per trial\footnote{Given that selections
+are independent on each other. This formula requires some adaptations in the
+case of sequential or hierarchical selection.}, $P$, the selection accuracy
 of the decoder, and $Q$, the number of trials per second.
@@
 The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
-\emph{synchronous} \ac{bci} paradigm with all the benefits and drawbacks
+\ac{bci} paradigm with all the benefits and drawbacks
 Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
@@
 multiple targets at once in a row-collumnn strategy and using a sequence of
-selections, giving rise to the classic
-P300 matrix speller of~\cite{Farwell1988}.
+selections, giving rise to the classic matrix speller of~\cite{Farwell1988}.
 Other optimizations increase response \ac{snr}, like using flashing face

### 2024-09-26 chapters/conclusion.tex
Comments Marc

@@
 
-\subsubsection{Regularized spatiotemporal beamforming}
+\subsubsection{\ac{erp} decoders}
 We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
@@
 spatiotemporal (STBF-emp)
-beamformer and a state-of-the-art Riemannian Geometry method (xDAWN+RG) in an off-line P3 detection task on
+beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
 an existing dataset.
@@
 
+\todo{tensor}
+\todo{wcble}
+
 %* Preliminary results: BTTDA better than HODA, on par with other state of the
@@
 
-\subsubsection{Healthy control study}
+\subsubsection{Gaze-independent \ac{bci} healthy controls}
 The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
@@
 significant impact on performance in covert and split VSA, as well as that the
-amplitude effect observed for covert VSA also holds for split VSA.
+amplitude effect\todomvh{not the right word, maybe the effect of covert VSA on
+ERP...} observed for covert VSA also holds for split VSA.
 
@@
 
+\subsection{Patient case studies}
+
 \section{Limitations}

### 2024-09-26 chapters/covert_align.tex
Comments Marc

@@
 
-\cite{Arico2014} observed higher variability in single-trial P3 peak
+\textcite{Arico2014} observed higher variability in single-trial P3 peak
 latencies relative to stimulus onset during covert VSA compared to overt VSA.
@@
 decoding, but did not actually verify this hypothesis directly.
-Additionally, \cite{hardiansyah2020single} developed a classifier for
+Additionally, \cite{Hardiansyah2020} developed a classifier for
 covert VSA ERPs, exploiting single-trial latency features in combination with
@@
 They demonstrated the positive influence of single-trial ERP component latency
-features on covert VSA inference yet did not attempt to correct the amplitude
+features on covert VSA inference, yet did not attempt to correct the amplitude
 features for these latencies.
 \todo{Should we already mention Frenzel here?}
-\cite{Frenzel2011} introduced a similar protocol to our split VSA
+\textcite{Frenzel2011} introduced a similar protocol to our split VSA
 setting.
@@
 knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
-However, \cite{Frenzel2011} considered their interface only for the case
+However, \textcite{Frenzel2011} considered their interface only for the case
 where the user actively intends to select both targets determined by the gaze
@@
 decoding the visuospatial attention only.
+\todomvh{avoid overlap. it seems already explained in previous sentences. The reader can find details in
+  [6].}
 
@@
 \subsection{Data collection}
-We evaluate our approach on a
-publicly available dataset and one specifically recorded for this study, designed to
+We evaluate our approach on the
+publicly available BNCI2014-009 dataset~\cite{Aloise2012a} and one specifically recorded for this study, designed to
 probe different modalities of covert VSA.
+\todomvh{first explain BNCI2014-009 using fig. 6.1 or revert the announcement
+in the previous sentence.}
 
@@
 Each subject performed different VSA conditions (overt, covert and split),
-illustrated in Figure~\ref{fig:interface}.
+illustrated in Figure~\ref{fig:interface} (top row).
 Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
@@
 with a radius of 12.28° of visual angle centered on the midpoint of the screen,
-conforming to the interface proposed by \cite{Treder2010} (Figure~\ref{fig:layout_targets}).
+conforming to the interface proposed by \textcite{Treder2010}
+(Figure~\ref{fig:layout_targets})\todo{figure interface in four panels and
+update all refs below}.
 A hexagonal layout interface with an empty center and a low number of targets
@@
 
+\subsection{Data processing and analysis}
+\subsubsection{Preprocessing}
+Analysis was performed using Python and the MNE software package (version
+1.3.1)~\cite{Gramfort2013}.
+All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
+Bad channels in the data were automatically detected using the RANSAC
+method~\cite{Fischler1981} and rejected.
+The recorded EEG was re-referenced
+offline to the average of the mastoid electrodes TP9 and TP10.
+Next, the EEG signals were corrected for eye movement artifacts using an
+Independent Component Analysis (ICA).
+Since we have access to EOG data for the CVSA-ERP dataset components correlating
+significantly with the EOG were rejected.
+For the BNCI2014-009 dataset, ICA components were manually rejected.
+Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
+In both datasets, only 16 channels were kept for
+analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
+PO8).
+
 \subsubsection{Decoders}
@@
 
-\subsection{Data preprocessing and analysis}
-Analysis was performed using Python and the MNE software package (version
-1.3.1)~\cite{Gramfort2013}.
-All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
-Bad channels in the data were automatically detected using the RANSAC
-method~\cite{Fischler1981} and rejected.
-The recorded EEG was re-referenced
-offline to the average of the mastoid electrodes TP9 and TP10.
-Next, the EEG signals were corrected for eye movement artifacts using an
-Independent Component Analysis (ICA).
-Since we have access to EOG data for the CVSA-ERP dataset components correlating
-significantly with the EOG were rejected.
-For the BNCI2014-009 dataset, ICA components were manually rejected.
-Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
-In both datasets, only 16 channels were kept for
-analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
-PO8).
-
 To evaluate performance, 6-fold cross-validation without shuffling was performed for both
@@
 allows for extracting latencies ranging from -100ms to +100ms.
+
 \section{Results}
@@
 present in covert and split VSA compared to overt VSA,
-we justify our approach similar to \cite{hardiansyah2020single} by
+we justify our approach similar to \cite{Hardiansyah2020} by
 observing that the increased discrimination performance of a machine-learning
@@
 
-While \cite{hardiansyah2020single} decoded covert VSA more effectively by
-contributing single-trial latency and amplitude features to decoder.
+\textcite{Hardiansyah2020} decoded covert \textcite{vsa} more effectively by
+contributing single-trial latency and amplitude features to decode.
 Contrary to our approach, they did not then correct these amplitude

### 2024-09-26 chapters/wcble.tex
Comments Marc

@@
 These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
-\ac{erp} waveform.
+\ac{erp} waveform.\todomvh{refer to image in bci chapter}
 These components each have their own distinct amplitude and latency.
@@
 the development of \acp{bci}.\todo{cite}
-Latency is often assessed as the latency of \ac{erp} components in a subject
-average or grand average.
-Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
+Latency is often assessed as the timing of an \ac{erp} component peak, obtained
+after averaging over trials.
+Latencies of non-averaged, single-trial \acp{erp} should capture even more
 relevant information, yet they are very hard to measure due to low \ac{snr} of
 single trial \acp{erp}.
-Within session variability of \ac{erp} latency, or \ac{erp} latency
-\emph{jitter} has been shown to influence \ac{bci}
+Within-session variability of \ac{erp} latency, called \ac{erp} latency
+\emph{jitter}, has been shown to influence \ac{bci}
 accuracy~\cite{Thompson2012}.
 This effect is especially prominent when bringing \ac{bci} development from the
-lab to the clinical setting with patients in the intended target population.
+lab to the clinical setting with patients in the target population.
 \textcite{Zisk2021} showed that patients with \ac{als} have significantly
@@
 trial-to-trial jitter~\cite{Truccolo2002}.
-Experiments show that latency jitter in \acp{erp} is the result of the interacion of
+Experiments show that latency jitter in \acp{erp} is the result of the interaction of
 the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
 	Kisley1999, Curto2009, Arieli1996}.
-The latency of an evoked potential is a combination of the modulation of oscillating activity that is
-time-locked, but not necessarily phase-locked to stimulus onset, and a
-delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
-\todo{Check this, \cite{Stokes2016,Mouraux2008}}
-\todo{Read references and check if neural origins and
-	generation model justify approach}
+According to the Firefly model of event-related activity~\cite{Burgess2012},
+the latency of an evoked potential results from a combination of the modulation
+of oscillating activity that is time-locked, but not necessarily phase-locked to stimulus onset,
+and a delay caused by ongoing neurophysiological processes that are not necessarily related to the task
+\cite{Stokes2016,Mouraux2008}.
+\todo{substantiate}
 
@@
 averaged per condition to cope with the unfavorable \ac{snr}?
-When averaging over multiple identical signals with jittered latencies, the shape
+When averaging over multiple identical responses with jittered latencies, the shape
 and amplitude of the average do not reflect the properties
@@
 onset.
-While straight-forward, this method does not perform well in low or negative
+While straight-forward, this method does not perform well in low
 \ac{snr} conditions, unless combined with filtering.
@@
 }
-\section{Algorithms}
 \todo{Pr not italic}

### 2024-09-26 figures/bci/paradigms.tikz.tex
Comments Marc

@@
 
-    \node[color=darkgray, align=center,font=\bfseries, anchor=north] at (0,{-(1+\margin)}) {passive};
-    \node[color=darkgray, align=center,font=\bfseries, anchor=south] at (0,{1+\margin}) {active};
+    \node[color=darkgray, align=center,font=\bfseries, anchor=north] at
+    (0,{-(1+\margin)}) {passive\\participation};
+    \node[color=darkgray, align=center,font=\bfseries, anchor=south] at
+    (0,{1+\margin}) {active\\participation};
     \node[color=darkgray, align=center,font=\bfseries, anchor=east] at ({-(1+\margin)}, 0) {stimulus\\independent};

### 2024-09-27 chapters/wcble.tex
Accuracy results

@@
 
-\todo{figure}
-
 \subsection{Latency estimation}
-
+\todo{refer to figure}
 We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/wcble/fig_latency.pdf}
+  \sansmath
+  \input{figures/wcble/fig_latency.pgf} \\
+  \input{figures/wcble/fig_accuracy.pgf}
   \caption{align results}
@@
 
+As the \ac{snr} decreases, both \ac{cble} and \ac{wcble} can accurately
+estimate evoked potential latencies up to a given threshold. After this
+threshold, correlation between estimated and ground-truth latencies will start
+decreasing.
+However, we notice that this threshold occurs later for \ac{wcble} than for
+\ac{cble} for all evaluated jitter levels, indicating that \ac{wcble} is more
+robust in the presence of pink spatiotemporal noise.
+At around -28dB, \ac{wcble} does not significantly outperform \ac{cble}
+anymore and both methods get overtaken by the strong presence of noise.
+
 \subsection{Classification}
+\todo{refer to figure}
 Similar to the previous experiment, classification accuracy between target and
@@
 \section{Discussion}
+
+Reason that accuracy latency estimation plateaus and accuracy drops for cble,
+but that both plateau for wcble, is because of the alignment
 

### 2024-09-27 code/setup.py

@@ -1,2 +1,92 @@
+import matplotlib.pyplot as plt
+from cycler import cycler
+
 mm = 1 / 25.4  # centimeters in inches
 pagesize = (160 * mm, 240 * mm)
+# colors
+darkgray = "#8D807A"
+
+savefig_kws = dict(transparent=True, bbox_inches="tight", pad_inches=0)
+
+
+# Function to configure matplotlib styles
+def configure_matplotlib_style():
+    custom_color_palette = ["#F29018", "#F63D1E", "#78479D"]  # Custom colors
+
+    # Configure global settings
+    font_size = 9
+
+    plt.rcParams.update(
+        {
+            "text.usetex": True,  # use LaTeX to write all text
+            "pgf.texsystem": "pdflatex",
+            "pgf.rcfonts": False,
+            #'figure.figsize': (8, 6),                 # Plot size
+            "axes.grid": False,  # Disable grid lines
+            "axes.edgecolor": "#8D807A",  # Spine color
+            "axes.labelcolor": "#8D807A",  # Axis label color
+            "axes.titlesize": font_size,  # Axis title font size
+            # "axes.titleweight": "bold",  # Axis title bold
+            "axes.titlelocation": "left",  # Axis title alignment left
+            "axes.titlecolor": "#8D807A",  # Axis title color
+            "axes.labelsize": font_size,  # Axis label font size
+            "xtick.color": "#8D807A",  # X-axis tick color
+            "ytick.color": "#8D807A",  # Y-axis tick color
+            "xtick.labelsize": font_size,  # X-axis tick label size
+            "ytick.labelsize": font_size,  # Y-axis tick label size
+            "xtick.direction": "in",  # X-axis ticks inward
+            "ytick.direction": "in",  # Y-axis ticks inward
+            "xtick.major.size": 6,  # Major tick size for x-axis
+            "ytick.major.size": 6,  # Major tick size for y-axis
+            "lines.linewidth": 1.5,  # Default line width
+            "lines.markersize": 8,  # Default marker size
+            "legend.fontsize": font_size,  # Legend font size
+            # "legend.loc": "upper right",  # Legend position
+            "legend.frameon": False,  # No frame around legend
+            "axes.prop_cycle": cycler(
+                color=custom_color_palette
+            ),  # Custom color palette
+            "text.color": "#8D807A",  # Global text color
+            "figure.facecolor": "white",  # Background color for the figure
[…]

### 2024-09-27 code/wcble/fig_latency.py

@@ -1,10 +1,0 @@
-import matplotlib.pyplot as plt
-import pandas as pd
-import seaborn as sns
-
-df = pd.read_csv("data/wcble/results_align.csv")
-print(df)
-sns.relplot(
-    data=df, x="snr", y="rho", col="jitter", row="stc", hue="method", kind="line"
-)
-plt.savefig('figures/wcble/fig_latency.pdf')

### 2024-09-27 code/wcble/fig_latency_classify.py

@@ -1,0 +1,71 @@
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import configure_matplotlib_style, savefig_kws, set_size
+
+configure_matplotlib_style()
+figsize = set_size(subplots=(1, 3))
+
+
+# Latency
+df = pd.read_csv("data/wcble/results_align.csv")
+df = df[df["stc"] == "sine"]
+
+facet_kws = dict(ylim=[0, 1], xlim=[0, -31])
+fg = sns.relplot(
+    data=df,
+    x="snr",
+    y="rho",
+    col="jitter",
+    # row="stc",
+    hue="method",
+    kind="line",
+    facet_kws=facet_kws,
+)
+fg.set_titles(col_template="$\\sigma={col_name}$s")
+fg.set_xlabels("SNR (dB)")
+fg.set_ylabels("$\\rho$")
+fg.figure.set_size_inches(*figsize)
+# fg.tight_layout()
+
+plt.savefig("figures/wcble/fig_latency.pgf", bbox_inches="tight", pad_inches=0)
+
+# Accuracy
+df = pd.read_csv("data/wcble/results_classify.csv")
+df = pd.concat(
+    [df, pd.read_csv("data/wcble/results_classify_tlda.csv")], ignore_index=True
+)
+df = df[df["stc"] == "sine"]
+df["accuracy"] *= 100
+
+facet_kws = dict(ylim=[0, 100], xlim=[0, -31])
+fg = sns.relplot(
+    data=df,
+    x="snr",
+    y="accuracy",
+    col="jitter",
+    # row="stc",
+    hue="method",
+    kind="line",
+    facet_kws=facet_kws,
+)
+fg.set_titles(col_template="$\\sigma={col_name}$s")
+fg.set_xlabels("SNR (dB)")
+fg.set_ylabels("accuracy (%)")
+fg.figure.set_size_inches(*figsize)
+# fg.tight_layout()
+
+
+# Force drawing of the figure to get the correct bounding boxes
+plt.draw()
+fig = fg.figure
+# Get the bounding box of the axis, in figure coordinatesσ = 0.1s
+bbox = fig.axes[0].get_tightbbox(fig.canvas.get_renderer())
+# Convert the bounding box units to figure-relative coordinates
+bbox_in_fig_units = bbox.transformed(fig.transFigure.inverted())
+# Calculate the amount of space required to align the left spine with the figure border
[…]

### 2024-09-27 code/wcble/fig_simulated.py

@@ -1,10 +1,11 @@
-import ipdb
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 import seaborn as sns
 from mne import read_epochs
-from setup import pagesize
+from setup import configure_matplotlib_style, darkgray, pagesize, savefig_kws
+
+configure_matplotlib_style()
 
 epochs = read_epochs("data/wcble/simulated_epochs.fif")
 meta = epochs.metadata
@@ -12,15 +13,14 @@ meta = epochs.metadata
 epochs.pick(["C3", "STIM"])
 epochs.crop(-2, 2)
 
-stcs = meta["waveform"].unique()
 snrs = meta["snr"].unique()
 # snrs = [0, -16, -31]
 jitters = meta["jitter"].unique()
 
 
 img_kwargs = dict(
-    vmin=-50,
-    vmax=50,
+    vmin=-60,
+    vmax=60,
     picks="C3",
 )
 
@@ -42,79 +42,88 @@ def order(epochs, n_epochs=100):
     return order
 
 
-for i, stc in enumerate(stcs):
-    shape = (2 * len(snrs), len(jitters))
-    fig, axs = plt.subplots(
-        *shape,
-        sharex="col",
-        sharey="row",
-        # layout="tight",
-        height_ratios=[3, 1] * len(snrs),
-        figsize=pagesize,
+def plot_img(ax, epochs):
+    order_ = order(epochs)
+    epochs.plot_image(
+        order=order_,
+        colorbar=False,
+        evoked=False,
+        axes=ax,
+        show=False,
+        **img_kwargs,
     )
-    axs = axs.reshape(shape)
-    fig.suptitle(stc)
+    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=darkgray)
 
-    for j, snr in enumerate(snrs):
-        for k, jitter in enumerate(jitters):
-            prefix = f"{stc}/{snr}/{jitter}"
-            print(prefix)
-            curr_epochs = epochs[prefix]
+    ax_img.set_title("")
+    ax_img.set_xticks([])
+    ax_img.set_xlabel("")
+    ax_img.set_yticks([])
 
-            ax_img = axs[j * 2, k]
-            ax_ts = axs[(j * 2) + 1, k]
 
-            order_ = order(curr_epochs)
-            curr_epochs.plot_image(
-                order=order_,
-                colorbar=False,
-                evoked=False,
-                axes=ax_img,
-                show=False,
[…]

### 2024-09-27 chapters/covert_align.tex
package

@@
 
-Finally, we argue that the increased performance of WCBLE is partly due to the
-fact that
-tLDA is a suitable first-stage classifier.
-Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
-problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
-Secondly, this method has a synergy with CBLE since both make the same
-assumption about the short-time stationary of the EEG background noise
-within an epoch.
-CBLE does not retrain the first-stage classifier for each time shift, but
-rather trains it once within the given window.
-After training, the classifier parameters represent some information about the
-expected ERP waveform and background noise.
-By applying the trained classifier to different time shifts, it assumes this
-ERP waveform can be shifted in time, but since the classifier's information about
-the background noise was only obtained from the initial window, CBLE assumes
-its properties do not vary throughout the epoch.
-The block-Toeplitz covariance structure of tLDA also assumes that the
-background noise represented by this covariance after subtracting the class
-averages is stationary within the epoch~\cite{Sosulski2022}.
+%Finally, we argue that the increased performance of WCBLE is partly due to the
+%fact that
+%tLDA is a suitable first-stage classifier.
+%Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
+%problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
+%Secondly, this method has a synergy with CBLE since both make the same
+%assumption about the short-time stationary of the EEG background noise
+%within an epoch.
+%CBLE does not retrain the first-stage classifier for each time shift, but
+%rather trains it once within the given window.
+%After training, the classifier parameters represent some information about the
+%expected ERP waveform and background noise.
+%By applying the trained classifier to different time shifts, it assumes this
+%ERP waveform can be shifted in time, but since the classifier's information about
+%the background noise was only obtained from the initial window, CBLE assumes
+%its properties do not vary throughout the epoch.
+%The block-Toeplitz covariance structure of tLDA also assumes that the
+%background noise represented by this covariance after subtracting the class
+%averages is stationary within the epoch~\cite{Sosulski2022}.
 

### 2024-09-27 chapters/wcble.tex
package

@@
 \section{Results}
-
-%Four settings are evaluated to assess both the impact of the iterative Woody
-%procedure and the extraction of robust latency features.
-%These are CBLE-median and WCBLE-median, where the median of the latency
-%probability distribution is used as latency estimate, and CBLE-peak and
-%WCBLE-peak, where the maximum of the score output is used as latency estimate.
-
-\todo{
-refer to later chapter for use on real erp data
-}
+\label{sec:wcble/results}
+\todo{refer to later chapter for results on real data}
 \subsection{Data simulation}
@@
 %\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
-\fullpagefig{\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}}{}{fig:wcble/sim-sine}
+\fullpagefig{%
+  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
+}{%
+  Simulated target and non-target evoked data. Targets contain a sine wave
+  pulse jittered by $\sigma$. 32 EEG channels with pink
+  spatiotemporal noise were simulated using a forward head model, channel C3 is
+  plotted here.
+}{fig:wcble/sim-sine}
 \todo{Properly align plots and label axes}
@@
 Results are presented in Figure~\ref{fig:wcble/results/latency}.
+A 95\ confidence intervals on each measure was obtained through bootstrapping
+with 1000 permutations.
 
-\begin{figure}
-  \sansmath
+\begin{figure}[!ht]
   \input{figures/wcble/fig_latency.pgf} \\
@@
 \end{figure}
-\todo{invert x axis}
-\todo{caption}
 \todo{properly label axes}
@@
 \todo{refer to figure}
-Similar to the previous experiment, classification accuracy between target and
-non-target epochs is assessed.
+Analogous to the previous experiment, 10-fold cross-validated classification
+accuracy between target and non-target epochs was assessed.
+Here, \ac{wcble} was compared to \ac{cble}, but both also to their base classifier
+\ac{tlda}.
+\Ac{cble} and \ac{wcble} classifiers were implemented as described
+in~\ref{sec:wcble/methods}, with both the wavelet transformed score time series
+and square transformed latencies as input to a second-stage logistic regression
+classifier with $L_2$-norm regularization ($C=0.2$).
+\todo{ref figure}
 
+The classification accuracy of both \ac{tlda} and \ac{cble} decreases with
+$\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
+This difference is particularily large when $\sigma$ is higher ($\sigma=0.2$
+and $\sigma=0.3$) and \ac{snr} is not yet high.
+\Ac{wcble} always scored better than \ac{cble} and \ac{wcble}, except for
+$\sigma=0.1$ and $\text{\ac{snr}} <27\text{db}$, at which point both \ac{cble}
+and \ac{wcble} performed at chance level.
+When considering the accuracy of \ac{wcble}, we see that it shows more of a
+plateau as \ac{snr} increases.
 \todo{insert results figure and discuss}
 
-\section{Discussion}
+\section{Discussion \& conclusion}
+In this work, we introduced a new \ac{erp} latency estimation and decoding
+method that applies a spatiotemporal classification iteratively along the time
+dimension of the data, refining the dataset at each iteration by aligning
+trials.
+The algorithm combines \acf{cble} with Woody's template matching iteration
+scheme and is therefore named \acf{wcble}.
+Our method attempts to model the spatial and temporal activation of
+each \ac{erp} component, as well as it's latency distribution.
+Through the firs-stage classifier, it takes into acount the
+spatial and temporal structure of EEG background noise and the trials from the
+classes to discriminate.
+\section{Conclusion}
 
+The key finding of the simulation study presented
+in~\ref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
+potential jitter and lower \ac{snr} than \ac{cble}.
+It could therefore be a better candidate to apply in \ac{erp} analysis and
+decoding in those cases where jittered \ac{erp} data is of interest.
 Reason that accuracy latency estimation plateaus and accuracy drops for cble,
 but that both plateau for wcble, is because of the alignment
+As  the noise increases beyond a certain threshold, the improvement vanishes.
+At this point, both methods perform poorly in both tasks, so if latencies
+cannot be properly estimated, no increase in classification performance can be
+gained by aligning in \ac{wcble}.
 
-\todo{mention cross validation}
-\todo{remove lcmv}
+As shown, the proposed method is suitable for \ac{bci} decoding setting
+as well as single-trial \ac{erp} latency analyses.
+In the case where there is only one data class present, the first-stage
+classifier can be replaced by a spatiotemporal filter.
+Instead of outputting a classification score, this filter should now output a
+metric representing the presence of a template response in the signal.
+A suiting candidate is the spatiotemporal
+\ac{lcmv}-beamformer~\cite{VanVliet2015}\todo{refer chapter}.
+The activation pattern can be set to the average of the set of trials to
+analyze, and it can be refined at each \ac{wcble} iteration by taking the
+average of aligned trials.
 
-\todo{In this work, we introduce a new latency method that leverages data-driven,
-joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
-integrates the spatiotemporal filtering in Woody's template matching iteration
-scheme and can easily be extended to multi-component settings.
-Our method attempts to model the spatial and temporal activation of
-each \ac{erp} component, as well as it's latency distribution, and takes into acount the
-spatial and temporal correlation of the EEG background noise.}
+Finally, we argue that \ac{tlda} is a suitable first-stage classifier.
+Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
+problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
+Secondly, this method has a synergy with CBLE since both make the same
+assumption about the short-time stationary of the EEG background noise
+within an epoch.
+CBLE does not retrain the first-stage classifier for each time shift, but
+rather trains it once within the given window.
+After training, the classifier parameters represent some information about the
+expected ERP waveform and background noise.
+By applying the trained classifier to different time shifts, it assumes this
+ERP waveform can be shifted in time, but since the classifier's information about
+the background noise was only obtained from the initial window, CBLE assumes
+its properties do not vary throughout the epoch.
+The block-Toeplitz covariance structure of \ac{tlda} also assumes that the
+background noise represented by this covariance after subtracting the class
+averages is stationary within the epoch~\cite{Sosulski2022}.
 
-\todo{with lda, it is in the single class case equal to shifting a spatiotemporal
-lcmv beamformer over the epochs and can be used to extract latencies.
-multiclass:}
+The main limitation of the proposed method is its supposed lack to handle
+multi-component \ac{erp} data.
+In real \ac{erp} analysis settings, multiple components are usually present,
+and each component has a distinct contribution to class discriminatbility and
+latency distribution.
+A single-component method will yield issues in interpretability and convergence
+in the presence of other components, as it can `lock-on' to a given component
+cluster and flatten out the others.
+This can be avoided, however, by the choice of a proper region of interest
+isolating solely the component of interest.
+Current work focuses on extending \ac{wcble} to a multi-component setting, on
+the one hand to performance latency estimation and decoding performance on real
+\ac{erp} data and on the other hand to apply it as an \ac{erp} component
+separation method in the fashion of~\textcite{Oyang2017}.
 
-\section{Conclusion}
-Current work is focusing on the extension to multi-component algorithm, in a
-similar fashion to ouyang
+The proposed implementation in theory also supports processing multi-class
+data, but the convergence and correctness of the \ac{wcble} solution is yet to
+be properly studied for this case.

### 2024-09-27 code/wcble/fig_simulated.py

@@ -79,7 +79,7 @@ stc = "sine"
 height_ratios = []
 for _ in range(len(snrs)):
     height_ratios.extend([3, 1])  # Closer together for image+timeseries
-    height_ratios.append(0.75)
+    height_ratios.append(0.8)
 gridspec_kw = dict(height_ratios=height_ratios, hspace=0)
 shape = (len(height_ratios), len(jitters))
 fig, axs = plt.subplots(
@@ -120,10 +120,11 @@ for j, snr in enumerate(snrs):
         ax_ts.set_ylim([-10, 10])
         ax_ts.set_yticks([])
 
+        ax_ts.set_yticks([-5, 5])
+        if k:
+            ax_ts.set_yticklabels([])
         ax_ts.set_xticks([-2, -1, 0, 1, 2])
         if j == len(snrs) - 1:
-            ax_ts.set_yticklabels([])
-        else:
             ax_ts.set_xlabel("Time (s)")
 
 fig.savefig(f"figures/wcble/simulated-{stc}.pdf", **savefig_kws)

### 2024-09-28 chapters/bci.tex
Comments François, smearing figure

@@
 the user.
-A \ac{bci} reads your mind.
+%A \ac{bci} reads your mind.
 
@@
 \fullpagefigct{%
-  \includegraphics[width=\paperwidth]{figures/bci/bci_loop.pdf}
+  \includegraphics[width=\textwidth]{figures/bci/bci_loop.pdf}
 }{%
   The \ac{bci} loop.
-  The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
-  visual stimulation. Neural signals are recorded, and neurophysiological features
+  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
+  visual stimulation.
+  Electrical neural signals are recorded, and neurophysiological features
   related to the paradigm are extracted. Using machine learning, a decision can
@@
 A better candidate is the neuronal electrical activity.
-The brain consists 86 billion neurons, which are highly interconnected cells that are
+The brain contains 86 billion neurons, which are highly interconnected cells that are
 the smallest units of information processing.
@@
 which can be measured using electrodes.
-Fluctuations in the brain's electrical potential can be faster and more
+Fluctuations in the brain's electrical activity are be faster and more
 informative.
@@
 \todo{preprocessing}
-This is commonly referred to as \emph{feature extraction}
+This is commonly referred to as \emph{feature extraction}.
 In other words, the signal needs to be interpreted \emph{in function of} the
@@
 `background tasks', all of which generate their own brain activity.
-On the other hand, the activity of interest in often
+On the other hand, the activity of interest often
 originates from a specific region or network within the brain.
@@
 rehabilitation applications.
+\todo{is every section referred to?}
 
@@
 
-\fullpagefigct{%
-  \includegraphics[width=\paperwidth]{figures/bci/recording_modalities.pdf}
+\fullpagefig{%
+  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
 }{%
@@
 
-Larger implanted electrodes are referred to as intracranial ac{eeg} (i\ac{eeg}).
+Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
 Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
@@
 Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
-offer great temporal resolution up to thousands of Hz.
+offers great temporal resolution up to thousands of Hz.
 \Ac{eeg} measures the electrical potential over time at a given electrode relative to a
@@
 As it is easily applicable on the outside of the scalp, many practical
-consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
+consumer-grade \ac{eeg} headsets exist in the form of (often wireless) electrode
 caps or helmets.
@@
 Most common are the visual paradigms.
-These can be more performant since they allows the user to use their visual
+These can be more performant since they allow the user to use their visual
 system, one of the most evolved sensory system in humans.
@@
 along multiple dimensions at once.
-\textcite{Han2023} recently used this strategy, combining frequancy and phase
+\textcite{Han2023} recently used this strategy, combining frequency and phase
 coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
@@
 decoding, and
-\item increase $Q$ by selecting a paradigm and allows fast stimulation or
+\item increase $Q$ by selecting a paradigm that allows fast stimulation or
 responses.
@@
 The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
-\ac{bci} paradigm with all the benefits and drawbacks
+\ac{bci} paradigm with all the benefits and drawbacks.
 Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
@@
 measured and decoded from non-invasive \ac{eeg} signals with a relatively low
-computational effort and short calibration time.
-First established by~\cite{Farwell1988} in 1988, it well supported by literature
+computational effort and short calibration timeand .
+First established by~\cite{Farwell1988} in 1988, it is well supported by literature
 and has been proven to work for patients in day-to-day home
@@
 A stimulus can be rare either due to its inherent properties like color or
-location
+location.
 Crucially, this is also the case if it is `marked' as rare by the user, e.g.
@@
 \Ac{itr} in the oddball paradigm is optimized by intensifying
-multiple targets at once in a row-collumnn strategy and using a sequence of
+multiple targets at once in a row-collumn strategy and using a sequence of
 selections, giving rise to the classic matrix speller of~\cite{Farwell1988}.
@@
 stimuli~\cite{Treder2011}.
-The number of targets in a visual oddball \ac{bci}is limited by the crowding
+The number of targets in a visual oddball \ac{bci} is limited by the crowding
 effect, which imposes a limit on how close targets can be to each other while
@@
 The component most prominently modulated by attention is called the P3
-(positive deflection after $\pm300$ ms).
+(positive deflection after $\pm300$ ms).\todo{add footnote about how ms naming
+is maybe not suited, especially since this work focuses on the fact that there
+is a lot of variability in the timings for visual potentials. the short naming
+scheme builds on 'the first negative component', 'the first positive
+component', and for the visual oddball erp this ordening more or less coincides
+with the timings.}
 Therefore, oddball paradigms are frequently referred to as P300 paradigms.
@@
 
-\paragraph{P1}
-The P1 component occurs approximately 100 ms after stimulus presentation.
+\emph{The P1 component} occurs approximately 100 ms after stimulus presentation.
 This component reflects initial visual processing and is primarily linked to
@@
 
-\paragraph{N1}
-The N1 component peaks around 150 to 200 ms post-stimulus.
+\emph{The N1 component} peaks around 150 to 200 ms post-stimulus.
 It represents an extension of sensory processing related to both gaze fixation
@@
 
-\paragraph{P2}
-The P2 component occurs between 200 to 300 ms.
+\emph{The P2 component} occurs between 200 to 300 ms.
 It reflects higher-order processing, particularly in contexts demanding
@@
 
-\paragraph{N2}
-Peaking around 200 to 350 ms, the N2 component is associated with cognitive
+\emph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
 control processes such as conflict monitoring and inhibition.
@@
 
-\paragraph{P3}
-The P3 component is elicited in oddball paradigms and peaks around 300 ms
+\emph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
 post-stimulus.
@@
 
-\paragraph{Late Positive Potential (LPP)}
-The LPP is a later component occurring beyond 400 ms.
+\emph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
 Sustained attention and emotional processing of stimuli are reflected in the
@@
 Typically enhanced for emotionally salient or motivationally relevant
-information, LPP serves as an indicator of cognitive appraisal beyond mere
+information, the LPP serves as an indicator of cognitive appraisal beyond mere
 recognition tasks.
@@
 mechanisms.
-These components not only inform the design of an effectiv oddball BCI system
+These components not only inform the design of an effective oddball BCI system
 but also enhance our understanding of the cognitive processes involved in visual

### 2024-09-28 chapters/gaze_independence.tex
Comments François, smearing figure

@@
-\emph{Parts of Section~\ref{sec:gaze-independence} have been published by \textcite{VanDenKerchove2024}.}
+\emph{Parts of Sections~\ref{sec:gaze-dependence} and~\ref{sec:gaze-independence}
+have been published by \textcite{VanDenKerchove2024}.}
 
@@
 eye motor problems, which can range from minor (nystagmus
-\footnote{
+\footnote{%
 Involuntary, rhythmic, and repetitive eye movements recognizable by their
@@
 }, other eye tremors
-\footnote{
+\footnote{%
 These can include square-wave jerks, saccadic intrusions, microtremors, or
@@
 }, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
-\footnote{
+\footnote{%
 Weakness or limited paralysis of one or more of the muscles that control eye
@@
 These patients might suffer from severe or complete eye motor impairment
-immediately from the onset of their condition.
+from the onset of their condition.
 In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
@@
 especially for patients with bulbar onset \ac{als}.
-Furthermore, \textcite{Hayashi1991} show that, as these patients are supported beyond
-respiratory failure, the disease progression will eventually also involve
-paralysis of the eye muscles.
+Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
+the point of independent breathing, symptoms will eventually also involve
+eye muscle paralysis.
 Various forms of eye movement abnormalities also occur often and are especially well studied in
@@
 In clinical reality, every tetraplegic patient has their own unique set of
-preserved capabilities and from an solution-oriented \ac{bci} engineering point
+preserved capabilities and from a solution-oriented \ac{bci} engineering point
 of view, the etiology of the symptoms can be abstracted away.
@@
 This group is much larger than the first.
-\elide\
-The third class of potential \ac{bci} users, which is the largest of all,
-includes those who still retain (and can be expected to continue to
-retain) substantial neuromuscular control, particularly speech and/or
-hand control, and can, therefore, operate a wide range of assistive
-communication and control devices.''
 }}
@@
 forms one of the main \ac{bci} target populations.
-In this work, we'll define \ac{lis} patients as those in a situation of (near)
+In this work, we define \ac{lis} patients as those in a situation of (near)
 complete paralysis and difficulties or the inability to communicate without
-assistive technology
+assistive technology.
 This corresponds to classes one and two defined by
@@
 
-This leads us to what we'll call the \emph{gaze-dependence problem in visual
+This leads us to what we call the \emph{gaze-dependence problem in visual
 \ac{bci}}:
@@
 For our purposes, screen targets are overlaid with non-overlapping, transient
-stimuli that evoke event-related  potentials \acp{erp} in the \ac{eeg}.
+stimuli that evoke \acp{erp} in the \ac{eeg}.
 The selected target can be decoded from these \acp{erp}, as is the case for the
@@
 that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
-Nevertheless, these methods, can suffer too from reduced information
+Nevertheless, these methods can suffer too from reduced information
 transfer rates~\cite{Chennu2013}.
@@
 They compared classical linear and non-linear \ac{erp} classifiers on a covert
-attention P3 \ac{erp} component dataset.
+attention oddball \ac{erp} paradigm dataset.
 The results revealed no significant performance improvement in covert VSA
@@
 oddball interface.
-Here, targets are shown in short pulses on a computer screen.
+Here, targets are shown in short flashes on a computer screen.
 Each time the user observes a target, an \ac{erp} is evoked, the presence of which can be
@@
 Decoding this modulation allows for information transfer controlled by the
-user's brain activity, as sketched in Figure~\ref{fig:bci}.
+user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
 However, visual \ac{bci}s cannot operate efficiently when the user does not direct
 their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
+\todomvh{I am wondering whether goal statements and patient target groups
+should be addressed in chapter 1.}
 
@@
 While \ac{bci}s are most attractive as a solution for these patients
-compared to other assistive technologies like eye-tracking, studies often
+compared to other assistive technologies like eye-tracking\todomvh{but with eye
+control while the previous sentence points are absence of it}., studies often
 report bad performance in patient populations precisely due to the lack of adequate eye motor
@@
 \todo{refer to ITRs from previous chapter}
-Section{sec:gaze-independence} showed that visual \ac{bci} paradigms form
+Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
 a good candidate for a performant gaze-independent \ac{bci}.
@@
 for patient use, which will be tested in a case-based patient study in which
-both performance and user comfort will be evaluated.
+both performance and user comfort was be evaluated.\todo{was this actually
+done?}
 
@@
 another (the distractor).
+\todomvh{+ "free" condition in Fig. 21.}
 
@@
 \subsubsection{Regularized spatiotemporal beamforming}
-Due to the lack of the N2 component and decrease of amplitude of the
-P3 component in covert attention
+Due to the decreased amplitude of the N1 and P3 components in covert attention
 settings~\cite{Treder2010}, the \ac{snr} of
@@
 improvements in those settings where performance is not yet near the achievable
-maximum.
+maximum.\todomvh{?? how to find out what is the maximum?}
 
@@
 Furthermore, these regularizing constraints impose temporal stationarity on
-the background noise, which is of use in our next efforts to cope with the
-non-stationarity of the P3 signal component.
+the background noise, yielding insights for our next\todomvh{next? more specificaly?}
+efforts to cope with the non-stationarity of the P3 signal component.
+\todofc{conclue with something like "this approach has been published in ... and
+will be described in chapter XXX}
 
@@
 method named Classifier-based Latency Estimation with Woody iterations (wCBLE).
+\todo{conclue with something like "this approach has been published in ... and
+will be described in chapter XXX}
 
 \subsubsection{Tensor discriminant analysis}
+\todo{todo}
 
@@
 \ac{vsa} conditions.
+\todofc{conclue with something like "this approach has been published in ...
+and will be described in chapter XXX}
 
 \subsubsection{Patient case study}
+\todofc{conclue with something like "this approach has been published in ...
+and will be described in chapter XXX}
+
+\todo{todo}
 \todo{mention multi-center patient study case studies in cooperation

### 2024-09-28 chapters/stbf_struct.tex
started wcble introduction
Comments François, smearing figure

@@
 
-	Part of the work in this chapter was published in~\cite{VanDenKerchove2022}.
+  \emph{This chapter, except section~\ref{sec:stbf-struct-kronlda}, was
+  published by~\textcite{VanDenKerchove2022}}.
 
@@
 	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
-	\autoref{fig:kronlda-covs}a and \autoref{fig:covs}b respectively show examples of the
+	\autoref{fig:kronlda-covs}a and \autoref{fig:kronlda-covs}b respectively show examples of the
 	empirical estimators of the covariance and the inverse covariance matrices.
@@
 	of the covariance matrix by strengthening the diagonal.
-	\autoref{fig:kronlda-covs}c and \autoref{fig:covs}d respectively show examples of the
+	\autoref{fig:kronlda-covs}c and \autoref{fig:kronlda-covs}d respectively show examples of the
 	shrunk estimator of the covariance and the inverse covariance matrices.
@@
 	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
-	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:covs}i and
+	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:kronlda-covs}i and
 	\autoref{fig:kronlda-covs}j).
@@
 	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
-	applications~\cite{DeMunck2002,Huizenga2002,6408231,GonzalezNavarro2016,GonzalezNavarro2017}.
+	applications~\cite{DeMunck2002,Huizenga2002,Beltrachini2013,GonzalezNavarro2016,GonzalezNavarro2017}.
 	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
@@
 	improve stability and achieve more robust
-	regularization~\cite{Wiesel2012,Greenewald2014,6408231, Breloy2016}.
+	regularization~\cite{Wiesel2012,Greenewald2014,Beltrachini2013, Breloy2016}.
 	The spatial and temporal covariance matrices are shrunk at every fixed-point
@@
 	classifier accuracy.
-	However, few iterations or even just one~\cite{castaneda2014estimation} suffice to achieve satisfactory performance in practice.
+	However, few iterations or even just one~\cite{Castaneda2014} suffice to achieve satisfactory performance in practice.
 
@@
 	corresponding number of blocks in the training folds and dropping the rest.
-	\autoref{fig:accuracy} and \autoref{ap:accuracy_ap} show each classifier's
+	\autoref{fig:accuracy} and \autoref{fig:accuracy_ap} show each classifier's
 	accuracy relative to the data availability.
@@
 			3 trials of testing stimuli.
-			\autoref{ap:accuracy_ap} contains results for all numbers of trials.
+			\autoref{fig:accuracy_ap} contains results for all numbers of trials.
 			While \textsc{stbf-emp} is unstable when little training data
@@
 
+  \subsection{Extension to \ac{lda}}
+  \label{sec:stbf-struct-kronlda}
+  \todo{cite proof (or reiterate)}
+
 	\section{Conclusion}

### 2024-09-28 chapters/wcble.tex
Comments François, smearing figure

@@
 performance, and disproportionally so when the user is not directly gazing at
-intended targets, such as is the case for for gaze-impaired patients.
+intended targets, such as is the case for gaze-impaired patients.
 
@@
 most prominent of which is the smearing effect as illustrated
-in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
+in~\autoref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
 averaged per condition to cope with the unfavorable \ac{snr}?
@@
 	these references}.
+
+\begin{figure}[t]
+  \centering
+  \input{figures/wcble/smearing.tikz.tex}
+  \caption{The smearing effect. When obtaining an \ac{erp} template through
+  averaging, high jitter causes a decrease in peak amplitude and a deformation
+  of the shape of the waveform.}
+  \label{fig:wcble/smearing}
+\end{figure}
 
@@
 
-The most straight-forward method to estimate single-trial latency is
+The simplest method to estimate single-trial latency is
 \emph{peak picking}
@@
 onset.
-While straight-forward, this method does not perform well in low
+While straightforward, this method does not perform well in low
 \ac{snr} conditions, unless combined with filtering.
@@
 Multiple components can be time-locked to different events or neurophysiological
-proccess and therefore are not necessarily time-locked to each other. They
+processes and therefore are not necessarily time-locked to each other. They
 each can have different latency variabilities.
@@
 however, these algorithms lack an integrated approach to deal with multiple
-\ac{erp} components. A clear evaluation is lacking of the performance of these
-methods in
-multi-component settings is lacking.
+\ac{erp} components. A clear evaluation of the performance of these
+methods in multi-component settings is lacking.
 
@@
 classes to discriminate.
-\section{Conclusion}
 

### 2024-09-28 include/document.tex
Refactor structure
Comments François, smearing figure

--- 2024-09-24 main.tex
@@
-% ATTENTION: Don't build this, either build main_booklet.tex or
-% main_manuscript.tex
-%
-% Preamble ====================================================================
-\pdfminorversion=7
-\usepackage{todonotes}
-\usepackage{hyperref}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage[inline]{enumitem}
-\usepackage{subcaption}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-% Commands
-\input{math.tex}
-\input{acronyms.tex}
-\newcommand{\includechapter}[2]{
-  \chapter{#1}
-  \begin{refsection}
-  \input{#2}
-    \printbibliography[heading=subbibliography]
-  \end{refsection}
-  \clearpage
-}
-
-\newcommand{\comment}[1]{
-  \todo[color=yellow]{#1}
-}
-\newcommand*\elide{\textup{[\,\dots]}}
-
-% Bibliography
-\usepackage[backend=biber, doi=false, url=false, isbn=false, eprint=false]{biblatex}
-\addbibresource{references.bib}
-\addbibresource{publications.bib}
-
-% Set bibliography font size
-\renewcommand*{\bibfont}{\footnotesize}
-
-\usepackage[inline]{enumitem}
-
 \begin{document}
@@
 \chapter{Tensor discriminant analysis}
-\todo{add Tensorizations coping with latency jitter}
+\todo{insert tensor paper}
+\todo{discuss Tensorizations coping with latency jitter to make link}
 
-\includechapter{Classfier-based Latency Estimation with Woody iterations}{chapters/wcble.tex}
+\includechapter{Latency estimation methods}{chapters/wcble.tex}
 
@@
 \include{chapters/cv.tex}
+
 \end{document}

### 2024-09-28 chapters/covert_align.tex
Comments François

@@
 Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.
+\todo{figures}
 
@@
 starts.
+
 All targets were intensified for a duration of 100 ms, in pseudorandom order.
@@
 between 200 and 400 ms.
-Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
-longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
+\Acp{isi} were jittered to counteract steady-state effects and residue in averaging. A
+longer \ac{isi} will increase component amplitude and aid in counteracting temporal
 autocorrelation for a higher statistical test precision.
@@
 crosshair locations.
+
 The experiment started with five non-recorded practice stimulation blocks, one for
@@
 \subsubsection{Decoders}
-We will compare \ac{wcble} and \ac{cble} ad described in
+We will compare \ac{wcble} and \ac{cble} as described in
 Section~\ref{sec:wcble/methods}, with their base classifier and tLDA and with
@@
 filters, projects these covariances to a tangent space and classifies the
-result using $L2$-regularized logistic regression
+result using $L_2$-regularized logistic regression
 (XDAWNCov-TS-LR)~\cite{Cecotti2017}.
@@
 extra channels, projected to tangent space and classified using
-$L2$-regularized logistic regression (ERPCov-TS-LR).
+$L_2$-regularized logistic regression (ERPCov-TS-LR).
 
@@
 and conditions.
+\todo{table?}
 
@@
 Note that interpretation of this comparison may be challenging due to differences in
-interface design (number of targets, inter-stimulus interval), subject population,
+interface design (number of targets, \ac{isi}), subject population,
 EEG recording procedure and available training data.

### 2024-09-29 chapters/conclusion.tex
Started patients paper

@@
 \todo{BCI principles and practice chapter 19}
+\todo{https://www.nature.com/articles/s44222-024-00239-5}
 
@@
 
+\todo{. Coincidentally, this also
+builds towards a solution for the Midas Touch Problem in BCI, where a BCI user sometimes accidentally
+selects a target while not wanting to give any input. Decoding of mental attention independent of eye
+gaze, with the option of having visual attention without mental attention,
+would counteract this.}
+
 \subsection{Patient case studies}
@@
 \subsection{Working with patients}
+immediately move to patients and on-line implementation, to be able to report
+interpretable metrics
+permanent working collaboration as a lab
+
 start from patients

### 2024-09-29 chapters/gaze_independence.tex
Started patients paper

@@
+\epigraph{%
+  ``\elide\ This begs the question: if a patient had reliable control
+  over their eye gaze, why would one aim to facilitate communication with that
+  patient via a BCI? It would be far simpler and more efficient to use an
+  eye-tracking system.
+  And even if a BCI were the preferred or only available option for some reason,
+  designs requiring eye gaze control could not be used
+  by patients with severe motor impairment, as these patients
+  cannot direct their gaze.
+  Because these are the patients who have the greatest need for BCIs, it is
+  important that accurate gaze-independent BCIs exist.''
+}{%
+  \textcite{Egan2017}
+}
+
 \emph{Parts of Sections~\ref{sec:gaze-dependence} and~\ref{sec:gaze-independence}
@@
 
-\section{The gaze-dependence problem}
+\section{The gaze-dependence problem}%
 \label{sec:gaze-dependence}
@@
 \todo{include FA}
+\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
+\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}
 
@@
 In these cases, classical solutions often fall short of the widely accepted
-80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,RonAngevin2019} deemed necessary for a comfortable user
-experience~\cite{Neeling2019}, calling for alternative, gaze-independent
+80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,RonAngevin2019} deemed necessary for a comfortable user experience~\cite{Neeling2019}, calling for alternative, gaze-independent
 solutions.
+
+To summarize: for patients with eye motor impairment, gazing directly at a screen target may
+be uncomfortable, impractical, or even impossible.
+Hence, assistive devices that rely on eye tracking are often inefficient for
+them.
+Consequently, while \ac{bci}s hold great promise for these individuals, conventional
+gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
+control.
+Therefore, the development of decoding strategies that account for covert
+VSA becomes crucial in the pursuit of high-performance gaze-independent
+\ac{bci}s.
 
@@
 visuospatial attention and their gaze do not necessarily coincide.
-For patients with eye motor impairment, gazing directly at a screen target may
-be uncomfortable, impractical, or even impossible.
-Hence, assistive devices that rely on eye tracking are often inefficient for
-them.
-Consequently, while \ac{bci}s hold great promise for these individuals, conventional
-gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
-control. Therefore, the development of decoding strategies that account for covert
-VSA becomes crucial in the pursuit of high-performance gaze independent
-\ac{bci}s.
 
@@
 rates compared to visual paradigms.
-Furthermore, they can suffer from increased mental effort in operation and user-dependent variability~\cite{Reichert2020b}.
+Furthermore, they can suffer from increased mental effort in operation and
+user-dependent variability~\cite{Severens2014,Reichert2020b}.
 \todo{insert table from riccio in text as numbers, not as table}
@@
 electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
+\todo{chosen because gaze-independent optimized and Peters2020}
 

### 2024-09-29 chapters/patients.tex
Add decoders
Intro, P300 stimulation
Started patients paper

--- 2024-08-24 chapters/patients.tex
@@
-\section{Patient presentation}
-\section{Data collection \& preprocessing}
-\todo{Define structure patient results/discussion}
+\section{Introduction}
+
+Communication BCIs have a target population consisting of patients in various
+stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
+This reduces their performance operating visual oddball \ac{bci}s, since they
+are presumed unable to comfortably redirect their gaze at the desired target,
+i.e., perform overt \ac{vsa}.
+They might more naturally operate in covert \ac{vsa}, where the gaze and
+\ac{vsa} do not coincide.
+Several studies show that performance drops in covert VSA \todo{[2, 1], our
+study}, necessitating gaze-independent solutions.
+Our previous study, presented in chapter \todo{ref chapter}, indicates gaze-independent performance can be
+improved in healthy subjects using a suited decoding strategy.
+
+These studies build on the assumption that patients with eye motor impairment
+would feel comfortable operating
+Visual paradigms
+
+Need for verification in patients, cite different characteristics, differences
+in setting, environment and equipment,
+\todo{goal is class 1, but challenging to perform research experiments and to
+recruit, cite wolpaw}
+
+problems eye tracking
+
+Literature for visual \ac{bci} in eye motor impaired patients is sparse.
+\todo{List conclusions of Peters2020 (severely eye motor impaired als patients
+able to use a hex-o-spell)}
+
+Therefore, we aim to apply the concepts from earlier work an literature to
+patients with motor impairment and, in addition, eye motor impairment.
+The objectives of this case study are as follows:
+\begin{enumerate*}
+  \item Explore capabilities and experienced comfort of eye motor impaired
+    patients,
+  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
+  \item verify if this performance can be improved with a suitable decoding
+    strategy.
+\end{enumerate*}
+
+\section{Materials \& methods}
+\subsection{Patient recruitment}
+Patients were recruited from across the Neuromuscular Reference center at
+University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
+(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
+(Lille, France) and a specialized care home (France).
+Experiments were performed under the supervision of their treating physician.
+Patients were recruited based on the following criteria.
+To qualify for inclusion, patients must
+\begin{enumerate}
+	\item be at least 18 years old and no older than 60
+	years,
+  \item belong to class 2 or 3 according the \ac{bci}
+	patient selection criteria presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
+  \todo{specific cutoff based on vision}
+  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
+\end{enumerate}
+Patients were excluded if they
+\begin{enumerate}
+  \item have a diagnosis of a major medical condition, including any major
+    neurological or psychiatric disorder other than those of interest based on
+    inclusion criteria~\ref{item:patients/inclusion/wolpaw}
+    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
+  \item have a predisposition to or have a history of any kind of epileptic seizures,
+    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
+  \item have a severe loss in vision or hearing, that would significantly impair
+        participation in the experiment,\label{item:patients/exclusion/vision}
+  \item are currently using specific psychoactive medications or substances that could affect the out-
+        come.\label{item:patients/exclusion/cognitive}
+  \item be able to understand the experiment instructions and cooperate,
+  \item have any other limitations preventing them from performing the given task.
+\end{enumerate}
+
+In total, 11 patients were contacted, of which 1 \ac{ms} patient was excluded based on
+criterion~\ref{item:patients/exclusion/vision}, 1 \ac{tbi} patient on
+both~\ref{item:patients/exclusion/epilepsy}
+and~\ref{item:patients/exclusion/cognitive}, and one stroke patient based
+on~\ref{item:patients/exclusion/medical}
+One further stroke patient was excluded due to technical
+difficulties during the experimental session.
+Vision was assessed using a LogMAR chart~\cite{Bailey1976}.
+
+Ultimately, 7 patients were retained.
+Of these, one patient was diagnosed from bulbar-onset \ac{als}.
+\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
+progressive loss of motor function.
+This initially results in general weakness and loss of muscle tone, but
+eventually results in full body paralysis.\todo{cite}
+Although speech and especially eye movements are usually preserved until the
+later stages of the disease progression, the bulbar-onset variant is
+characterized by an early loss of speech and an increased involvement of eye
+motor symptoms~\cite{Guo2022}\todo{which?}.
+Furthermore, one of the goals of \ac{bci} is to support \ac{als} patients whose
+life span has been extended with life support with and assistive technology to
+ensure quality of life.
+In these very progressed stages of \ac{als}, eye movement will eventually also
+be affected~\cite{Hayashi1991}.
+
+Three other patients were diagnosed with \ac{fa}, a neurodegenerative
+disease affecting the
+spinal cord, peripheral nervous system and cerebellum.
+\ac{fa} results in an impairing loss of muscle coordination.
+In progressed stages, the disease can present with nystagmus, saccadic
+intrusions and gaze dysmetria~\cite{Cook2017}.
+
+The final three patients were stroke patients.
+Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
+various eye motor disorders are common depending on the exact location
+~\cite{Bogousslavsky1987, Moncayo2009}.
+Stroke can lead to some of the more severe impairments like ophthalmoplegia or
+partial ophthalmoplegia immediately from onset.
+
+Table~\ref{tab:patients/patients} lists the included patients and their
+diagnoses.
+
+\begin{table}[t]
+  \centering
+  \footnotesize
+  \input{chapters/patients_table.tex}
+  \caption{Presentation of included patients including their diagnosis and
+  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
+  }
+  \label{tab:patients/patients}
+\end{table}
+
+\subsection{Eye tracking and eye motor examination}
+
+Self-reported eye motor abnormalities were recorded.
+Patients were asked for, the paralysis or inability to move in a given
+direction of one or both eyes, the inability to pursue an object to specific
+positions, discomfort or fatigue fixating at specific positions or in general
+when performing visual tasks and eye tremors while fixating, resting or
+pursuing.
+
+Additionaly, we implemented and performed the automated NeuroEye eye movement
+test proposed by~\textcite{Hassan2022}.
+This test is based on calibration-free eye tracking.
+\todo{say which part we used and which part not}
+\todo{motivation and text and references Juliette}
+\todo{Limited in what it can tell us (check paper)}
+\todo{include eye test document juliette}
+Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
+included patients.
+Patient PA1 had the mildest impairment, only reporting fatigue when fixating
+for prolonged times.
+Eye motor function of subjects PC2 and PC3 were most severly affected.
+Patient PC2 was only able to look up and down, and patient PC3 only retained
+partial movement of the right eye, while the left eye was permanently closed.
+
+\begin{table}[t]
+  \centering
+  \footnotesize
+  \input{chapters/patients_table_eye.tex}
+  \caption{Vision and eye motor impairment for included patients.
+  Eye motor impairment was assessed with a combination of self-reported issues
+  by the subject and the NeuroEye~\cite{Hasan2022} test.
+  LogMAR: lower is better.}
+  \label{tab:patients/eye}
+\end{table}
+
+Eye tracking throughout the experimental session was performed using the Tobii~\todo{model, make}
+portable eye tracker.
+\todo{describe analysis during experiment}
+
+\subsection{\Ac{bci} stimulation}
+
+The \ac{bci} stimulation procedure was based on the
+Hex-o-Spell~\cite{Treder2010} implementation presented
+by~\textcite{VanDenKerchove2024}.
+Similar to this study, the task consists of counting the flashes of a cued
+target among all flashing targets.
+We refer to section~\ref{sec:covert-align/stimulation} for the exact details.
+
+To adapt the interface for the patient setting, the number of blocks was
+decreased to 6 per \ac{vsa} setting.
+An additional \ac{free \ac{vsa}} condition was introduced.
+Here, the patient was instructed to perform the task as they deemed most
+comfortable.
+The cued \emph{split attention} setting proposed
+by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
+in natural \ac{vsa} operation settings for gaze-impaired patients.
+If the patient was not fully paralyzed, they were instructed not to move the head.
+\Ac{isi} was increased to\tod{number} to decrease task difficulty.
+The experiment alsy started with a training block in each condition, where the
+patient was instructed with feedback on their performance to make sure they
+understood and were able to perform the task.
+
+\fourpanefig{%
+  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}}{A
+  patient with the stimulation and recording setup.}{fig:patients/photos/side}
+}{%
+  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}}{subcaption}{fig:patients/photos/front}
+}{%
+  the interface
+}{%
+  4
+}{%
+  caption
+}{fig:patients/photos}
+
+Mention change in ISI
+Mention less rounds
+
+\subsection{Data collection \& preprocessing}
+
+\todo{check checklist}
+\subsection{BCI Decoding}
+
 \section{Outcomes}
+
+\subsection{Eye tracking analysis}
+
+PC2 reported that he was unable to fixate the lower two targets, and th
+
+\fullpagefig{%
+  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
+}{%
+  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
+  conditions.
+  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
+  near the middle of the screen, and PC4 near the bottom.
+}{%
+  fig:patients/gaze
+}
+
+\subsection{Decoding performance}
+
+\begin{figure}[t]
+  \input{figures/patients/fig_decode.pgf}
+\end{figure}
+\todo{check if percentages or 0-1 accuracy and roc-auc}
+
+For PB2 and PC4 performing covertly, was free decoding higher than covert
+decoding?
+
+\subsection{Cross-condition calibration}
+
+| indication of fatigue, cinsistency within free case
+| closer to overt or to covert by looking at eye tracker
+|
+| will learn
+| split up per target
+|
+| Findings:
+| 1. Some patients prefer to do covert (check performance results for them),
+|    analyze distance to target
+| 2. covert attention can be improved using WCBLE for all subjects, but not to
+| the
+|    level of overt covert attention works best when training in overt
+| attention
+| s this also true for patients preferring covert?
+| 3. free performs on par (in all subjects?), opens debate of necessity of
+|    gaze-independence BCIs
+| 4. covert attention works best when training in overt attention. Is this also
+|    true for patients preferring covert?
+
+2024-09-18 04:39:35 PM can more overt attention epochs closer to
+| mention problem can't know if patient shifted or eye tracker shifted
+| (difficult mobile gaze tracker)
+| overt lower because can't perform
+
 \section{Discussion}
+
+patients 2 are progressing towards 3
+
+Limitations
+
+Limited sample of patients, FA maybe not most interesting group to represent
+real BCI users
+
+interpreting eye tracker results hard because of inaccuracy eye tracker for eye
+motor impaired patients. Redo with more advanced stationary eye tracker and
+head rest.
+
+advanced eye tracking equipment is necessary
+
+user comfort was not objectively measured using established questionnaires
+
+Further work
+Comparing with a sample of non-gaze impaired patients with  neuromuscular
+disability DMD
+On-line session with these patients
+
+In a follow up study, we aim to revisit these patients with an on-line system
+to be able to calculate a realistic \ac{itr} and measure their user experience.
+
+\section{Conclusion}
+
+\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
+\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
+\todo{BCI principles and practice chapter 19}
+\todo{https://www.nature.com/articles/s44222-024-00239-5}

### 2024-09-29 code/patients/fig_decode.py

@@ -1,0 +1,46 @@
+import pdb
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import colors, configure_matplotlib_style, savefig_kws
+
+configure_matplotlib_style()
+
+df = []
+for pipeline in ["tLDA", "WCBLE", "XDAWNCov+TS+LDA"]:
+    df_pipe = pd.read_csv(f"data/patients/decode/decode_results_{pipeline}.csv")
+    df_pipe["pipeline"] = pipeline
+    df.append(df_pipe)
+df = pd.concat(df, ignore_index=True)
+df = df[df["subject"] != "PD01"]
+df = df[df["subject"] != "PB03"]
+
+df_avg = df.copy()
+df_avg["subject"] = "avg."
+df = pd.concat([df, df_avg], ignore_index=True)
+
+
+fg = sns.catplot(
+    data=df,
+    x="subject",
+    y="test_score",
+    hue="pipeline",
+    kind="bar",
+    col="condition",
+    legend="auto",
+    legend_out=False,
+)
+fg.set(ylim=(0, 1))
+plt.legend(
+    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
+)  # Adjust ncol to fit all labels in one line
+fg.axes[0, 0].legend_.remove()
+for ax in fg.axes.flat:
+    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
+    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
+fg.set_axis_labels("patient", "ROC-AUC")
+fg.set_titles("{col_name} VSA")
+fig = fg.figure
+fig.set_size_inches(6.5, 2)
+fig.savefig("figures/patients/fig_decode.pgf", **savefig_kws)

### 2024-09-29 code/patients/fig_gaze.py

@@ -1,0 +1,102 @@
+import matplotlib.colors as mcolors
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from matplotlib.patches import Rectangle
+from mne import read_epochs
+from setup import colors, configure_matplotlib_style, pagesize, savefig_kws
+
+configure_matplotlib_style()
+
+
+subjects = {
+    "PA1": "PA01",
+    "PB1": "PB01",
+    "PB2": "PB02",
+    "PB4": "PB04",
+    "PC4": "PC04",
+}
+conditions = ["overt", "covert", "free"]
+
+
+N_TARGETS = 6
+# INTERFACE_RADIUS = 12.28 / 2  # degrees of FOV
+INTERFACE_RADIUS = 20 / 2  # degrees of FOV
+TARGET_RADIUS = 4.15 / 2  # degrees of FOV
+TARGET_INTENSE_RADIUS = 5.60 / 2  # degrees of FOV
+FONT_SIZE = 0.5  # degrees of FOG
+
+
+def get_target_pos(pos):
+    if pos == "center":
+        return [0, 0]
+    elif isinstance(pos, int) and pos >= 0 and pos < 6:
+        angle = pos * 2 * np.pi / N_TARGETS
+        x = INTERFACE_RADIUS * np.cos(angle)
+        y = INTERFACE_RADIUS * np.sin(angle)
+        return [x, y]
+    else:
+        raise ValueError
+
+
+xedges = np.load("data/patients/gaze/gaze_xedges.npy")
+yedges = np.load("data/patients/gaze/gaze_yedges.npy")
+
+fig, axs = plt.subplots(
+    len(subjects),
+    len(conditions),
+    sharex=True,
+    sharey=True,
+    figsize=pagesize,
+    layout="constrained",
+)
+axs = axs.reshape((len(subjects), len(conditions)))
+for i, (subject, sub) in enumerate(subjects.items()):
+    for j, condition in enumerate(conditions):
+        axs[i, j].set_title(f"{subject}, {condition} VSA")
+        hist = np.load(f"data/patients/gaze/gaze_sub-{sub}_cond-{condition}.npy")
+        axs[i, j].pcolormesh(xedges, yedges, hist.T)
+        axs[i, j].set_aspect("equal")
+        axs[i, j].set_axis_off()
+        for target in ["center", 0, 1, 2, 3, 4, 5]:
+            pos = get_target_pos(target)
[…]

### 2024-09-29 code/setup.py

@@ -4,14 +4,21 @@ from cycler import cycler
 mm = 1 / 25.4  # centimeters in inches
 pagesize = (160 * mm, 240 * mm)
 # colors
-darkgray = "#8D807A"
+colors = dict(
+    darkgray="#8D807A",
+    lightgray="#E0E0E0",
+    accent1="#F29018",
+    accent2="#F63D1E",
+    accent3="#78479D",
+)
+
 
 savefig_kws = dict(transparent=True, bbox_inches="tight", pad_inches=0)
 
 
 # Function to configure matplotlib styles
 def configure_matplotlib_style():
-    custom_color_palette = ["#F29018", "#F63D1E", "#78479D"]  # Custom colors
+    custom_color_palette = [colors[c] for c in ["accent1", "accent2", "accent3"]]
 
     # Configure global settings
     font_size = 9
@@ -23,15 +30,15 @@ def configure_matplotlib_style():
             "pgf.rcfonts": False,
             #'figure.figsize': (8, 6),                 # Plot size
             "axes.grid": False,  # Disable grid lines
-            "axes.edgecolor": "#8D807A",  # Spine color
-            "axes.labelcolor": "#8D807A",  # Axis label color
+            "axes.edgecolor": colors["darkgray"],  # Spine color
+            "axes.labelcolor": colors["darkgray"],  # Axis label color
             "axes.titlesize": font_size,  # Axis title font size
             # "axes.titleweight": "bold",  # Axis title bold
             "axes.titlelocation": "left",  # Axis title alignment left
-            "axes.titlecolor": "#8D807A",  # Axis title color
+            "axes.titlecolor": colors["darkgray"],  # Axis title color
             "axes.labelsize": font_size,  # Axis label font size
-            "xtick.color": "#8D807A",  # X-axis tick color
-            "ytick.color": "#8D807A",  # Y-axis tick color
+            "xtick.color": colors["darkgray"],  # X-axis tick color
+            "ytick.color": colors["darkgray"],  # Y-axis tick color
             "xtick.labelsize": font_size,  # X-axis tick label size
             "ytick.labelsize": font_size,  # Y-axis tick label size
             "xtick.direction": "in",  # X-axis ticks inward
@@ -46,12 +53,13 @@ def configure_matplotlib_style():
             "axes.prop_cycle": cycler(
                 color=custom_color_palette
             ),  # Custom color palette
-            "text.color": "#8D807A",  # Global text color
+            "text.color": colors["darkgray"],  # Global text color
[…]

### 2024-09-29 chapters/conclusion.tex
Writing patient paper

@@
 
+very small group of interest, but available right now, no implantation needed,
+
+even if it is the most suited strategy for a patient, it will probably be
+improved if a combination of eye tracking, residual motor control and other
+biosensors are used.
+
+true power in the future is probably focused on
+invasive, active paradigms if recording technology and surgical methods
+improve lowering the risk. orders of magnitude higher itr and more natural
+communication.
+characterized by publication date, but might indicate it is of interest to
+heavily focus this research now on applied 'vraagstukken' with patients
+
 \section{Current \& future work}

### 2024-09-29 chapters/patients.tex
Writing patient paper

@@
 stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
-This reduces their performance operating visual oddball \ac{bci}s, since they
+This reduces their performance operating visual oddball \ac{bci}s (see
+section\todo{ref section} for an overview), since they
 are presumed unable to comfortably redirect their gaze at the desired target,
-i.e., perform overt \ac{vsa}.
+i.e., perform overt \ac{vsa}.\todo{cite some visual gaze-independent bci}
 They might more naturally operate in covert \ac{vsa}, where the gaze and
 \ac{vsa} do not coincide.
-Several studies show that performance drops in covert VSA \todo{[2, 1], our
+Several studies show that performance drops when not fixating the intended
+target~\cite{Brunner2010, Frenzel2011,
+Treder2010, RonAngevin2019, VanDenKerchove2024} \todo{[2, 1], our
 study}, necessitating gaze-independent solutions.
-Our previous study, presented in chapter \todo{ref chapter}, indicates gaze-independent performance can be
+
+Usually, they aim to improve gaze-independent \ac{bci} by optimizing
+performance in cued covert \ac{vsa}.
+These studies build on the assumption that patients with eye motor impairment
+would feel comfortable operating an interface in pure covert \ac{vsa} with
+central fixation.
+One could argue that a \ac{bci} that is only verified to work when a central
+fixation is maintained, could also be considered gaze-dependent.
+This does not account for the residual eye motor capabilities of most severly
+paralyzed or locked-in patients and the comfort they experience while
+performing the gaze fixation.
+
+Our previous study, presented in chapter \todo{ref chapter}  partially tried to account for this
+and showed gaze-independent performance can be improved
 improved in healthy subjects using a suited decoding strategy.
-
-These studies build on the assumption that patients with eye motor impairment
-would feel comfortable operating
-Visual paradigms
-
-Need for verification in patients, cite different characteristics, differences
+Yet, there is a strong need for verification of results obtained in healthy
+samples.
+It is a striking constatation that studies reporting on
+gaze-independent visual \ac{bci} in patients that are actually eye-motor
+impaired are very few.
+in patients, cite different characteristics, differences
 in setting, environment and equipment,
-\todo{goal is class 1, but challenging to perform research experiments and to
-recruit, cite wolpaw}
-
-problems eye tracking
-
-Literature for visual \ac{bci} in eye motor impaired patients is sparse.
 \todo{List conclusions of Peters2020 (severely eye motor impaired als patients
@@
 
+Eventually, one of the end goals of this research line is to develop
+gaze-independent \ac{bci} for patients that
+are fully locked-in and have no option left than to use a \ac{bci}.
+However, this group is very small and it is often a challenge to recruit them
+into a study and perform experiments with them~\cite{Wolpaw2006}.
+Patients with less severe paralysis or in less progressed disease stages that struggle with
+eye-tracking technology could also benefit from
+solutions tailored to their specific situation.
+
 Therefore, we aim to apply the concepts from earlier work an literature to
-patients with motor impairment and, in addition, eye motor impairment.
+patients with various degrees motor impairment and various
+degrees of eye motor impairment.
 The objectives of this case study are as follows:
@@
 	years,
-  \item belong to class 2 or 3 according the \ac{bci}
-	patient selection criteria presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
-  \todo{specific cutoff based on vision}
+  \item belong to class 2 or 3 according the \ac{bci}	patient selection criteria
+    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
   \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
@@
 
+\todo{check Usability and Workload of Access Technology for People With Severe
+Motor Impairment: A Comparison of Brain-Computer Interfacing and Eye Tracking}
+
 \begin{table}[t]
@@
 \end{table}
+\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
+Brain-computer interfaces and communication in paralysis: extinction of goal
+directed thinking in completely paralysed patients? Clin Neurophysiol.
+2008;119:2658-2666.}
+\todo{include year of diagnosis}
 
@@
 decreased to 6 per \ac{vsa} setting.
-An additional \ac{free \ac{vsa}} condition was introduced.
+An additional \emph{free \ac{vsa}} condition was introduced.
 Here, the patient was instructed to perform the task as they deemed most
@@
 If the patient was not fully paralyzed, they were instructed not to move the head.
-\Ac{isi} was increased to\tod{number} to decrease task difficulty.
+\Ac{isi} was increased to\todo{number} to decrease task difficulty.
 The experiment alsy started with a training block in each condition, where the

### 2024-09-30 chapters/bci.tex
Started patients paper
Compiling

@@
-\todo{comments liesel paradigms}
+\todo{comments liesel paradigms})
 \todo{spellcheck}
@@
 \todo{fix all latex warnings to make sure no refs are missing (or grep on that
-specific warning}
+specific warning)}
+\todo{uniform ref or autoref}
+\todo{don't capitalize figure, table, algorithm,...}
+\todo{define short captions for tables and figures}
+\todo{fix lacheck/chktex notifications}
 
@@
 their environment.
-In it's core essence, this is the ability to communicate your intentions,
+In its core essence, this is the ability to communicate your intentions,
 emotions, frustrations or thoughts such that they might be acted on.
@@
 
-The key problem for heavily paralyzed patients is that the mind wants to go
+The key problem for severely paralyzed patients is that the mind wants to go
 where the body can't follow.
@@
 termed brain-machine interface (BMI) when coupled to a physical actuator, like
-a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
-assistive technology and communication devices.}.
-A \ac{bci} directly couples actions in the `real world' to the mental state of
+a robot arm or an exoskeleton.
+The term \ac{bci} is usually preferred for
+assistive technology and communication devices.
+},
+a system that directly couples actions in the `real world' to the mental state of
 the user.
-%A \ac{bci} reads your mind.
 
@@
 by bypassing defective parts of the nervous or muscle system.
-By now, classic \acp{bci} methods have matured outside the lab setting and can be
-used as assistive communication technology by patients with impaired
-speech~\cite{Wolpaw2018}.
-Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
+By now, classic \acp{bci} methods are starting to mature outside the lab
+setting and can be used as assistive communication technology by patients with
+impaired speech~\cite{Wolpaw2018,Milekovic2018}.
+Impressive cutting edge examples of experimental use include a brain-spine
 interface allowing a paraplegic patient to walk again\cite{Lorach2023},
@@
 
-\todo{put focus on communication BCI}
+In general, devices processing direct inputs from the central nervous system
+are useful for rehabilitation and medical diagnosis or treatment.
+Additionally, they bring a new paradigm to human-computer interaction,
+especially when paired with virtual or augmented reality.\todo{cite}
+Ultimatly, they have always been especially promising as assistive technology for
+paralyzed patients limited in their communication ability.
+It is this \ac{bci} controlled communication is where our focus lies,
+as it offers a transformative means for individuals with severe motor
+impairments to regain functional independence to write and speak through
+assistive technology.
+This allows them to talk to their loved ones and caregivers, exercise hobbies
+and makes many aspects of daily life more accessible.
 
@@
 
-Brain-computer interfacing is an entertaining subject with many futuristic and more established
-applications.
-Devices processing direct inputs from the central nervous system can be
-useful for rehabilitation and medical diagnosis or treatment, but
-have always been especially promising as assistive technology for
-paralyzed patients limited in their communication ability.
-Furthermore, they bring a new paradigm to human-computer interaction,
-especially when paired with virtual or augmented reality.
-
 Simply put, a \ac{bci} records the user's brain activity, extracts some
 relevant output from this brain activity and couples this output to a function
-of a device.\todo{refer to fig loop}
+of a device as illustrated in Figure~\ref{fig:bci/loop}
 Optionally, the user can then observe the action of the device and adapt their
-behavior or brain activity accordingly, closing the loop
-\footnote{%
+behavior or brain activity accordingly, closing the loop\footnote{%
   \textcite{BCISociety2024} has recently formalized this into the
@@
 In turn, this can be coupled to an action, like typing the inteded letter A or B
-in Figure~\cite{fig:bci/loop}.
+in Figure~\ref{fig:bci/loop}.
 The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
@@
 activity.
-\todo{preprocessing}
 This is commonly referred to as \emph{feature extraction}.
@@
 Some obvious sources of interference can be easily filtered out by
-\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
+\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}), but the problem of retrieving or \emph{decoding} only the
 relevant signal is generally hard.
@@
 Useful information that is encoded in the brain signal by the paradigm,
-can be retrieved from it by the decoder.\todomvh{the decoder needs to take into
-account the control facilities of the application (degrees of freedom, serial
-/parallel nature, reactive/active...).}
+can be retrieved from it by the decoder.
 Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
@@
 On the one hand, the \ac{bci} gains its function by allowing the user to interact
-with their environment.\todo{put focus on communication BCI}
+with their environment.
+For a communication system, this means coupling the decoded information to
+e.g.\ a virtual keyboard or speech synthesizer.
 On the other hand, the actions performed by the \ac{bci} can themselves alter
@@
 More indirectly, modulation can be achieved by sensory stimulation.
-This involves specific, paradigm-related sensory input, like haptic
-feedback\todomvh{or tactile feedback:
-S.N. Flesher et al., “A brain-computer
-interface that evokes tactile sensations improves robotic arm control,”
-Science, vol. 372, no. 6544, pp. 831-836, 2021. } for movement \ac{bci}, or simply presenting selected actions
+This involves specific, paradigm-related sensory input, like tactile
+feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
 back to the user.
@@
 \todo{wrap-up}
-
-\todo{check excercise session presentation}
+\todo{make shorter}
 
@@
 \label{sec:bci-recording}
-\todo{make sure every part ends with a motivation for why we chose our
-approach}
-
 \todo{why is spatial resolution a good thing}
@@
 at...)}
+\todo{make sure every part ends with a short summary motivation for why we chose our
+approach}
 
@@
 (fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
-(fUS)~\cite{Zheng2023}\todomvh{non-invasive but not necessarily fUS...} and near-infrared spectroscopy
+(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
 (fNIRS)~\cite{Borgheai2020}.
@@
 }
-\todo{find better figure with permission to reuse and without unnecessary
-modalities}
 
@@
 Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
-NeuraLink implant~\cite{Musk2019}.\todomvh{+ Neuropixels 2}
+NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
+2.0~\cite{Steinmetz2021}.
 
@@
 Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
-\todo{cite} have been realized using
+have been realized using
 invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
 Furthermore, recent advances in recording technology focus on improving implant
-durability and resolution\todo{cite flexible}, and on balancing the invasiveness
+durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
 tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
@@
 
-\section{\ac{bci} paradigms}
+\section{\Ac{bci} paradigms}
 \label{sec:bci/paradigms}
@@
 only 800ms of stimulation.
-\section{Decoders}
+
+\section{Preprocessing}
+\label{sec:bci/preprocessing}
+\todo{preprocessing}
+EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
+significantly improves the quality of the recorded data for subsequent analysis and
+classification. Common preprocessing techniques in BCI systems include
+rereferencing, band-pass and notch filtering, and \ac{ica}.
+
+Rereferencing is an essential first step, where the potential of each EEG channel is
+recalculated relative to a common reference point, such as the average of all
+electrodes (common average reference) or a selected neutral set of electrodes.
+This technique helps reduce noise common across all channels, improving signal clarity
+and enhancing the accuracy of subsequent processing steps.
+
+Band-pass filtering is then usually applied to isolate the frequency ranges of interest by
+removing frequencies that are too high or too low to contain relevant neural signals.
+Band-pass filters eliminate slow drifts and high-frequency noise, such as
+muscle artifacts or environmental electrical interference, improving the
+\ac{snr}.
+Additionally, notch filtering can be  used to remove strong power line
+interference\footnote{Typically at 50Hz in Europe}.
+This interference can heavily contaminate EEG recordings and even leak through
+when outside the passband of a previous filter.
+Notch filtering efficiently attenuates this narrowband noise without affecting
+the surrounding frequencies.
+
+Finally, to address artifacts like eye blinks and saccades that can significantly
+distort the EEG signal, \ac{ica} is
+performed~\cite{Delorme2007}.
+\Ac{ica} is a blind source separation technique that decomposes the EEG data into
+statistically independent components. Components corresponding to eye movements
+, blinks and other bodily or environmental artifacts can be identified, either
+visually tased on their characteristic topographies and time courses, or
+statistically, and then removed.
+This leads to cleaner data for subsequent processing, especially in the case of
+visual \ac{bci} paradigms.
+Since eye movement plays a role in the paradigm, it can strongly be present in
+the signal and correlate with the task, forming a confounding factor for
+studying control based solely on brain activity.
+
+Together, these preprocessing steps ensure that only the relevant neural signals are
+passed on to the feature extraction and classification stages in BCI applications,
+improving the accuracy and reliability of the system.
+
+\section{Decoders}%
 \label{sec:bci/decoding}
@@
 decoding.
-This is usually based on
-some training data for which the performed tasks are known, by recognizing the
-activity evoked by the task in recorded data.
+Machine learning decoders use some training data for which the performed tasks
+are known, and learn to recognize the activity elicited by the task in recorded data.
 The training data can be either obtained from the \ac{bci} user themselves or from
@@
 preexisting labeled data from other sessions and users, but this is harder due
-to the variability between subjects and sessions.\todo{make shorter, move part
-to section}
-
-\todo{regression/classification}
-\todo{for communication, we want classification, focus on classification
-further down the line}
+to the variability between subjects and sessions.
 
@@
 
+Decoder choice is heavily dependent on constraints set by the paradigm and the
+application.
+The number of output degrees of features, the serial or parallel nature of
+performed tasks, whether the paradigm is active or reactive, etc.\ all need to
+be taken into account.
+On of these choices is to solve a \emph{regression} or \emph{classification}
+problem.
+In regression, a continuous output variable is predicted for a new sample,
+while classification attempts to sort it into one of a set of discrete classes.
+Regression can be useful in \acp{bci} in applications like mapping imagined
+movement to an external robot arm.
+For communication \acp{bci}, however, conveying information in a symbolic
+manner is often of interest.
+Many applications resemble virtual keyboards or map a user's actions to
+discrete words, predefined actions or letters for full control.
+Hence, such problems often present themselves as classification tasks.
+
 \textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
-decoder algorithms.
+classification algorithms for decoding.
 Classic linear methods, such as \ac{csp} feature extraction for
@@
 
+\subsection{Preprocessing}
+In addition to (ref) epoching
+baseline correction, but not in decoding (cite)
+
 \subsection{Decoding ERPs}

### 2024-09-30 chapters/bttda.tex
Compiling

\input{chapters/bttda/tensorviz.tex}
\todo{abbreviations}

\section{Introduction}
\todo{immediately arrive to the novelty or to the point}
\todo{reduce length of the introduction}
\todo{less acronyms}
\todo{put the intuition behind the algorithm (sparse, more flexible, what it
does in one sentence ) in the introduction. Additionally "try to get some
information from the data, continue extracting more information as long as
there is some left}
Brain-computer interfaces (BCIs) have the potential to bypass
defective neural pathways by providing an alternative communication channel
between the brain and an external device.
These interfaces find applications in the development of neuroprosthetics and assistive
technologies, among other applications~\cite{Wolpaw2020}.
To achieve their functionality, BCIs record and process neural data obtained through
a neuroimaging technique, with electroencephalography (EEG) being the most popular.
\todo{explicitly state how BTD is a generalization of PARAFAC and TUCKER}

Due to its multichannel time series structure, EEG data, like most neural
signal acquisition modalities used for BCIs, naturally exist as multiway data,
capturing information in both spatial and temporal domains.
Common preprocessing transformations, such as time-frequency transformation,
time-binning, or integrating information across multiple subjects or conditions,
can further expand the data into additional analytic domains.
This can result in high-dimensional datasets which are usually flattened into a
set of sample vectors, stripping the original data from its structure.
Yet, the intrinsic multiway structure of neural data~\cite{Erol2022} is
well-suited for representation as \emph{tensors}, or multiway arrays, where
each domain is represented as a tensor \emph{mode}.
Tensors provide a structured data representation for this highly dimensional
multiway data
This in turn paves the way to the development of tensor methods which can
counteract some of the drawbacks of the dimensionality problem.
Tensor methods are machine learning techniques that  consider each tensor mode
separately, reducing a given problem into partial, per-mode problems.
This approach has given rise to efficient dimensionality reduction techniques,
such as the Higher Order Singular Value Decomposition
(HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}
and Canonical Polyadic Decomposition
(CPD)~\cite{Hitchcock1927,Nazarpour2006}.

HOSVD and CPD are two examples of unsupervised techniques with applications in EEG
processing.
However, given the low signal-to-noise ratio and
specific, task-related output expected in a BCI application,
supervised feature extraction and machine learning techniques are
favored\todo{cite Lotte review}.
%One approach is to incorporate assumptions of the tensor structure of the data
%directly into the estimation of parameters of classic supervised linear -
%machine learning methods, such as linear discriminant analysis (LDA) or beamforming.
%Advances have been made by leveraging the decoupling of the spatial and temporal
%domains in EEG event-related potential (ERP) classification using Spatiotemporal
%Discriminant Analysis~\cite{Li2010,Zhang2013} or methods regularizing covariance
%matrix estimation~\cite{Kerchove2022,Sosulski2022}.

\todo{come in here with François' comment about restructuring the intro.}
A more structured approach to the same problem is to design a supervised
tensor dimensionality reduction method that optimizes discriminability between the
extracted features, as is the case with Higher Order Discriminant
Analysis (HODA)~\cite{Yan2005,Phan2010,Froelich2018}.
Extracted features can subsequently be further classified, most commonly
using LDA or a support vector machine (SVM) to predict class labels.
Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016}\todo{cite spatiotemporal discriminant
analysis here, 'variants and special cases'} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.

The methods above adhere to the Tucker tensor decomposition
structure,
\todo{
To introduce BTD, I would start with a paragraph explaining the decomposition
itself, or the approximation for dimension reduction, just after the
presentation of tensors: on one side, Tucker-N (1 core only) and on the other
side polyadic decomposition (PD, m scalars) and in the middle the BTD. This
paragraph explains that in most cases the decomposition is not unique and
depends on a priori parameters (nb of cores, dimensions of projection matrices,
etc.) Also defines the residual, difference between the original tensor and the
model in the case of approximation.
Then, the techniques for getting an
"optimal" decomposition or approximation could be presented: HOSVD and equiv.
(just a few refs) in the case of Tucker and CPD and others (just a few refs)
for PD. Then, your state of the art could be just focused on specific
optimizations of BTD used in BCIs (in order to justify the JNE journal
target)
In fact, this is the option chosen by Axel Faes and Marc in the JNE
paper of 2022)...}
meaning that they reduce input tensors of size
$(D_1,D_2,\ldots,D_K)$ to a smaller tensor of size $(r_1,r_2,\ldots,r_K)$ with
each $r_k\leq D_k$, similar to HOSVD.
While effective, other approaches such as the PARAFAC structure employed in
CPD, where a tensor is decomposed into a sum of rank-1 tensors,
might also be suitable to represent the neural data of interest.
Discriminant tensor features can also be extracted
in the PARAFAC structure, for instance through manifold
optimization~\cite{Froelich2018}.

Nevertheless, the PARAFAC structure might still not be able to
efficiently represent all relevant information in a compressed format.
The block-term tensor structure is a generalization of the Tucker and
PARAFAC structures, and can be calculated in an unsupervised way using
the Block-term Tensor Decomposition
(BTD)~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
BTD, of which the HOSVD and CPD are special cases,
represents a tensor as sum of Tucker terms.
Research has shown that supervised decoders adopting this more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Parial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaving room for improvement.
Instead of optimizing features directly for class separability, rather a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in HOPLS, BTTR are still more constrained than
what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low rank common subspace
between the input and classification labels.
\todo{remove acronyms if not used later}

%\todo{Describe spectrum-weighted tda~\cite{Huang2020}, structure is not
%	flexible enough, not broadly applicable, not interpretable}

We propose to mitigate these issues by designing a new supervised feature
extraction tensor method based on the abovementioned HODA algorithm
that is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
	\item  We develop a forward model for HODA to reconstruct a
	      given input tensor from the extracted features.
	\item This allows us to introduce a state-of-the-art BCI feature extraction
        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
	      (BTTDA).
  \item We evaluate a BCI decoder based on BTTDA and its special
	      PARAFAC-structured case on decoding tasks for both ERP and MI
	      paradigm BCI datasets.
\end{enumerate*}
\todo{It's always trichiin in the related work section of a ML/Mathematical
paper, but unless you r reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

\section{Methods}

\subsection{Notation}
Tensors are indicated by bold underlined letters $\ten{X}$, matrices by bold
letters $\mat{U}$, fixed scalars by uppercase letters $K$ and variable
scalars as lower case letters $k$.
The $n^\text{th}$ sample of a tensor dataset with $N$ samples is written as
$\ten{X}(n)$.
A tensor $\ten{X}\in \mathbb{R}^{D_1\times D_2 \times \cdots \times D_K}$ can be
unfolded in mode $k$ to a matrix
$\mat{X}_k\in\mathbb{R}^{(D_k\times\prod_{j\neq k}^K D_j)}$, by concatenating
all mode $j\neq$ fibers.
The tensor-matrix product of tensor $\ten{X}$ with matrix $\mat{U}$ along a
given mode $k$ is written as $\ten{X}\mpr{\mat{U}}{k}$. For ease of notation, let
$\ten{X}\mmpr{\mat{U}} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{K}$.
When skipping one of the modes $k$, this is
written as $\ten{X}\mmprs{\mat{U}}{k} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{k-1}\mpr{\mat{U}}{k+1}\ldots\mpr{\mat{U}}{K}$.
%The Kronecker product is noted as $\otimes$, covariance matrices are indicated  with $\mat{\Sigma}$.

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
visualized in Figure~\ref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
	\caption{A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}
Analogous to the HOSVD, HODA is a dimensionality
reduction decomposition that results in a dense latent tensor $\ten{G}$, and
imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
However, while for the HOSVD decomposition the projection matrices
are chosen to minimize the reconstruction error, the projection matrices
$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
$\ten{G}(n)$ belonging to classes with labels $c_n$.
This is a desirable property in a classification setting where samples are
high-dimensional tensors.

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
The goal is now to find the optimal projection matrices\todo{You should say
somewhere that the shape (r1,r2,...rK) is fixed a priori, no?}
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which can be solved by the iterative algorithm in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
	\caption{Higher order Discriminant Analysis (HODA) backward solution}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.

To apply HODA in a classification setting, the projections
are first learned on a training dataset  with known class labels.
Next, these projections are used to extract latent tensors from the
tensors in the training dataset.
These latent training tensors are then reshaped (\emph{vectorized}) into feature vectors
$\vec{g} =  \vect(\ten{G})$ and used to train a decision classifier with the corresponding class labels.
At the evaluation stage, the projections learned from the training dataset are
used to extract latent tensors from an unseen test dataset with unknown class
labels, which can also be vectorized and passed on to the trained decision
classifier.

To avoid overfitting and improve performance in low sample size settings, the
HODA problem can be regularized by shrinking the partial
within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
$\alpha_k$ at each step such that the eigenvalue problem becomes
\todo{make sure this appears after varphi is defined, move reference to
algorithm to later}
\begin{equation}
	\mat{S}_b^{(-k)} -
	\varphi\left[\left(1-\alpha_k\right)\mat{S}_{-k,\text{w}}+\alpha_k\mat{I}\right] =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
also be estimated in a data driven way in HODA~\cite{Jorajuria2022},
e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.

\subsection{A forward model for HODA}
As a prerequisite to the proposed BTTDA model, we must find a
way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
from $\ten{G}$ after dimensionality reduction.
This is usually referred to as finding a corresponding \emph{forward model}.
While a \emph{backward model} extracts latent sources or properties from the observed
data based on some task related criterion or on prior domain knowledge,
a forward model is a generative model that expresses the observed data in
function of these latent properties or sources that are given.
Forward models are useful for a.o.\ interpretability and data compression.
\todo{explicitly label error terms in equation, use $X\prime$ and error with
iteration sup}

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
	\ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$
and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
	\caption{The forward projection for HODA. By calculating activation
		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
		reconstructed from projected latent tensor $\ten{G}$. $A_k$ are chosen such
		that the variability captured in the latent tensor is maximally expressed in
		the reconstructed tensor and not in the error term.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the forward
projection term $\ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
	\caption{Higher order Discriminant Analysis (HODA) forward solution}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
Equation~\ref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
\todo{eps1 = X - G1 x {A1} still contains discriminative info, so
it is further projected onto another core tensor yielding
eps2 = eps1 - G2 x
{A2}
etc. and finally you give equation (15)
}
We thus extend the HODA feature extraction scheme with backward and
forward models defined in respectively Equations~\ref{eq:HODA-backward}
and~\ref{eq:HODA-forward} to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
	\caption{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
\todo{Explicitly mention that if restricted to rank one, it is parafac, and to
one block, it is hoda}

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
	\caption{Block-term Tensor Discriminant Analysis (BTTDA)}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_BTTDA.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation}from a residual error term
$\ten{\mathbfcal{E}}^{(b)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
	\ten{\mathbfcal{E}}^{(b+1)} = \ten{\mathbfcal{E}}^{(b)} - \ten{G}^{(b)}
	\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(1)}=\ten{X}$.

The resulting latent tensors can be vectorized and concatenated into
one single feature vector per input tensor
\begin{equation}
	\vec{g}
	=\left[\vect\left(\ten{G}^{(1)}\right)\
		\vect\left(\ten{G}^{(2)}\right)\
		\cdots\
		\vect\left(\ten{G}^{(B)}\right)\right]
\end{equation}
so that they can be classified in a similar manner to HODA.

\subsection{Model and feature selection}
Similar to the unsupervised BTD, the performance of BTTDA is
heavily dependent on the rank $(r_1^{(b)}, r_2^{(b)}, \ldots,
	r_K^{(b)})$ and on the number of blocks $B$.
If these are not known a priori, i.e. if they cannot be set based on insights in the
data generation process, a model selection step is necessary in order to
determine what the optimal values for $r_k^{(b)}$ and $B$ are.

Furthermore, HODA, and by extension BTTDA can extract a substantial amount
of redundant features, which can be dropped after projection and before proceeding to the classification
step~\cite{Phan2010}.
Specifcally in BTTDA redundant features can stack up over the number of
blocks, hampering performance.
Relevant features can be retained by calculating the
univariate Fisher score for all features.
The Fisher score $\phi{i}$ for feature $i$ is obtained as
\begin{equation}
	\phi(i) = \frac
	{\sum_c^C N_c \left(\bar{g}_i(c)-\bar{\bar{g}}_i\right)^2}
	{\sum_n^N \left(g_i(n)-\bar{g}_i(c_n)\right)^2}
\end{equation}
The features can then either be sorted by $\phi(i)$ and a given amount of
features retained for classification, or the decision to retain a feature can
be made by a threshold on the statistical $p$-values corresponding to the
$\phi(i)$.
We opted for the latter strategy since it does not require us to redetermine the
optimal number of features to retain at each block.

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
Algorithm~\ref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The area under the receiver operating characteristic curve (ROC-AUC) for
classification of extracted feature vectors after feature selection is used as
cross-validation score.
\begin{algorithm}
	\caption{Greedy model selection}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
highest validation score to determine $B$.

Alternatively, a special case of BTTDA can be constructed using only rank-one
blocks such that the resulting forward model adheres to the CPD structure.
The optimal amount of rank-one blocks can be found by truncating as above.
We refer to this strategy as PARAFACDA.

\section{Experiments}
\subsection{Datasets \& decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MIà paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in Table~\ref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operator
characteristic curve (ROC-AUC).
\begin{table*}[t]
	\centering
	\footnotesize
	\input{chapters/bttda/moabb.tex}
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
    classes (\# Classes), the number of trials or trials per class for ERP
    datasets (\# Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
		number of runs (\#Runs).
    Adapted from~\cite{Aristimunha2023}
    and~\cite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

To use HODA as a decoder, it is paired with LDA to classify the
extracted feature (HODA+LDA), with hyperparameters $r_k$.
Similarly, we implemented BTTDA+LDA with the proposed BTTDA feature
extraction with hyperparameters $r_k^{(b)}$ for each block $b$ and the number of blocks
$1\leq B\leq16$.
Additionally, we also introduce PARAFACDA+LDA, which is the special case of
BTTDA+LDA where each $r_k^{(b)}=1$, with $B$ and the feature selection
threshold as only hyperparameters.
Hyperparameters were determined separately for each fold using nested
stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
greedy model selection algorithm in Algorithm~\ref{alg:model-selection}.
For HODA+LDA and the HODA blocks in BTTDA+LDA, we choose
$r=r_1=r_2=\ldots=r_K$ with possible values
$\textstyle{1,2,4,8,\ldots,\max_kD_k}$
to reduce computational cost.
Other HODA and BTTDA hyperparameters were set to
$B_\text{max}=10$, $\varepsilon=\num{1e-8}$ and $I_\text{max}=128$.

Furthermore, as additional comparison methods, we used the methods evaluated in~\cite{Chevallier2024}.
For the ERP datasets, these were the Riemannian Geometry-based methods
using augmented ERP covariance matrices with and without SVD dimensionality
reduction features for a Minimum Riemannian Distance to Mean classifier
(ERPCov+MDM and ERPCovSVD+MDM), augmented ERP covariance matrices after
XDAWN~\cite{Rivet2009}
filtering paired with a Riemannian Minimum Distance to Mean classifier or a
projection to tangent space and a support vector machine (XDAWNCov+MDM and
XDAWNCov+TS+SVM) and LDA applied after XDAWN filtering (XDAWN+LDA).
For the MI datasets, the comparison methods were picked among Riemannian
methods.
These comprise of projection onto the Riemannian tangent space combined with an
ElasticNet classifier (TS+EL), a Fisher Geodesic Minimum Riemannian Distance to
Mean classifier and the classification of augmented covariance matrices
projected onto the Riemannian tangent space with a support vector machine
(ACM+TS+SVM).
Furthermore, we report the performance of the Common Spatial Patterns filter
and a frequency filterbank combined with a support vector machine (CSP+SVM and
FilterBank+SVM respectively).
We refer to~\cite{Chevallier2024} for implementation details of these comparison
methods.

\subsection{Event-Related Potentials}
ERPs are spatiotemporal features, with each sample forming a $2^\text{nd}$
order tensor with $K=2$ modes (a matrix), representing EEG channels and time samples
per epoch.
EEG signals for the evaluated datasets were recorded at the sample rate given
by Table~\ref{tab:moabb} and band-pass filtered between 1Hz
and 24Hz.
The signals were cut into epochs starting from stimulus onset with a
dataset specific length given by Table~\ref{tab:moabb}.
For HODA+LDA, PARAFAC+LDA and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
For the BNCI2014-008 and BNCI2015-003 datasets, this resulted in matrices of
dimensionality $(8,48)$ and $(8,38)$ respectively.

Table~\ref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\footnotesize
	\centering
	\input{chapters/bttda/score_erp.tex}
  \caption[Whithin-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated whithin-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \cite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
Highest performance is reached with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with significance level $\alpha=0.05$ reveal that both PARAFAC+LDA and BTTDA+LDA
significantly outperform HODA+LDA in both the BNCI2014-008 (PARAFAC+LDA:
$p=0.0039$, BTTDA+LDA: $p=0.0039$) and the BNCI2015-003 (PARAFAC+LDA:
$p=0.0001$, BTTDA+LDA: $p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

\subsection{Motor Imagery}
For MI, discriminatory information is represented in the EEG data as
Event-Related Synchronizations/Desynchronizations (ERS/Ds).
Contrary to the time domain analyses performed on ERPs, ERS/Ds are usually well
discerned in the time-frequency domain.
Hence, for the MI task, we will transform the EEG signal into the
time-frequency domain, forming $3^\text{d}$ order tensors, with $K=3$ modes
representing the channels, frequencies and time bins.

The MI datasets listed in Table~\ref{tab:moabb} were first band-pass filtered
between 8 and 30Hz and cut into
epochs with time windows as specified by Table~\ref{tab:moabb}.
Next, time-frequency transformation was performed using a complex Morlet wavelet
convolution, with 16 wavelet frequencies logarithmically spaced between 8 and
32Hz.
The number of wavelet cycles $c$ varied with wavelet frequency $f$ as
$c=0.7*f$.
Features were extracted by taking the log-transformed envelope of the complex
wavelet transformation an daveraging each epoch along the time axis into time bins of
length $1/4$s.
For the BNCI2014-001 and BNCI2014-004 dataset, this resulted in tensors of dimensionality
$(22, 4, 16)$ and  $(3, 4, 18)$
respectively.

Table~\ref{tab:erp-score} lists the cross-validated classification score for
the evaluated motor imagery datasets.
Note that, in line with~\cite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[%
    Whithin-session classification score for 2 motor imagery datasets.
    ]{Classification score for
		cross-validated whithin-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001 and area under the receiver-operator
    characteriscic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach a performance comparable to current
    state-of the art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet here both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA and BTTDA+LDA.

\subsection{Block contribution}
\todo{Discuss number of features}
To analyze the contribution of extra feature blocks extracted by {BTTDA} over
the first one found by HODA we pick n ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation and test data as a function
of the number of blocks increases, Figure~\ref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\ten{X}_\text{rec}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as
\begin{equation}
	\nmse\left(\ten{X}, \ten{X}_\text{rec}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
	\caption{Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

\section{Discussion}
\todo{subtensors in larger tensor drawing, focus more on sparsity in the
discussion}
\todo{focus on tradeof: more flexibility in model, but this implies more
hypermarameters so it relies more on a model selection procedure.}
The results in Table~\ref{tab:erp-score} and Table~\ref{tab:mi-score} show that
in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
exceeding all comparison methods for the two evaluated datasets, but
for MI datasets results fall short of those of comparison methods.
Performance on the MI classification task could however be greatly influenced
by the exact tensorization method used, i.e. the time-frequency transform in
this case, which is not the main focus of this work.

Nevertheless, when considering at the relative improvement over the original
HODA model, our results show BTTDA+LDA performs consistently on par or higher than HODA+LDA.
The classification score for BTTDA+LDA was also slightly higher than that of
restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
results were not significant and further studies with more datasets and
subjects should show whether this holds.
Figure~\ref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice NMSE monotonically decreases, indicating that
Since the training score keeps increasing, this indicates that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the amount of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

We assume the main benefit of BTTDA is that it can more easily discover relevant
features, while
being more parsimonious due to its block-term structure compared to HODA's full
Tucker structure.
Alternatively, the enhanced performance could also stem from the modeled data
covariance.
Since HODA estimates one whithin-class scatter matrix
$S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
model of the data scatter is determined by these per mode scatter matrices as a
Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
drawn from a multilinear normal distribution~\cite{Ohlson2013}.
However, it is known that the EEG covariance cannot fully be expressed as a
single Kronecker product, but rather is more accurately modeled by a sum of
multiple Kronecker products~\cite{Bijma2005, Sosulski2022}.
Since BTTDA iteratively fits HODA models to the residual error, it will be able
to express the full covariance structure given sufficient blocks.

Additionally, the forward modeling step inherent to the BTTDA results
in an interpretable model, since we can use the activation patterns or the
forward projection to inspect the neural patterns corresponding to the
relevant discriminatory information at each block~\cite{Haufe2014}.
Figure~\ref{fig:forward} shows the activation patterns
of two blocks obtained from the BNCI2014-008 dataset as well as the forward
projection of the difference between the averages of the mean latent tensor per
class (\emph{contrasts}) after forward projection.
While the weights of the backward projection are
uninterpretable~\cite{Haufe2014},
the activation patterns and contrasts after forward projection clearly show
that ERP components can be recognized and separated into different
BTTDA blocks.
\begin{figure*}[t]
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-0.png}
	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
	\caption{Spatial (left two columns) and temporal (middle column) activation patterns and
		condition contrasts (right column) obtained after forward projection of the latent
		features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of BTTDA
		fit on the full dataset BNCI2014-008.
		The separate blocks approximately model different ERP
		components.}
	\label{fig:forward}
\end{figure*}
Given informed or correctly tuned hyperparameters, this method could be used to
e.g. separate ERP components or neural processes based on the task-related
information in the class labels.

Despite favorable results in BCI decoding, the applications of the proposed
BTTDA model are limited mainly by the model selection approach used
to determine the individual block ranks.
While our proposed greedy model selection algorithm is a step in the right
direction, the high computational cost of setting hyperparameters through cross
validation can still hinder the portability of decoders relying on
BTTDA.
Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal between- or within-block feature
correlation, decreasing discriminability etc.
Future efforts should focus on automatic parameter setting, e.g. using
information criteria such as the ones used in BTTR~\cite{Faes2022} or other
statistical measures based on the models application.
\todo{comment as limitation that we test r1=r2=r3}

Another limitation is that BTTDA might yield a disproportionate
improvement for datasets with a low number of features relative to sample size,
while being less effective for datasets with more features
This is reflected
in our ERP results (low dimensionality vs. high number of trials) compared to the
MI results (higher dimensionality due to third order tensorization vs. lower
number of trials).
We expect a dimensionality limit beyond which the forward modeling step cannot
accurately regress from the low dimensional latent tensors to the high
dimensional original tensors, introducing
error in the input data for the next block, which can stack up over blocks.
Since the forward multilinear least squares problem is underdetermined, it is prone to
numerical instability, which calls for regularization of the forward modeling
procedure, but this would introduce another hyperparameter.
It should also thoroughly be investigated what the impact is of going beyond
second- and third-order case to higher order tensors, since this could have a large impact
on the model.
Other tensorization methods of the EEG data, like time-lagged Hankel
tensors~\cite{Papy2005}, or tensors across subjects or sliding windows etc., could also be
of interest if they are appropriately chosen based on prior knowledge of the dataset.
\todo{mention in discussion of model selection procedure that sparser blocks
would be better}

\section{Conclusion}
\todo{main restriction for hoda is model selection. this is also a model
selection problem, but it is more flexible}
We have introduced Block-term Tensor Discriminant Anlysis (BTTDA) a novel,
tensor-based, supervised dimensionality reduction technique optimized for class
discriminability which adheres to the block-term tensor structure.
BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
applied as a special sum-of-rank-one tensors PARAFACDA model.
The model is obtained by iteratively fitting HODA in a deflation scheme,
leveraging a novel forward modeling step.
Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
feature extraction can significantly outperform decoders based on HODA and
reach state-of-the art decoding performance on event-related potential
problems (second order tensors) and scores on par with or higher than HODA  in motor
imagery problems (third order tensors).
Moving from the rigid Tucker tensor structure of HODA to the more flexible
block-term structure shifts the problem from finding optimally constrained multilinear
projections to model and feature selection.
This allows performance to be traded
off for model complexity and number of features, to find a setting that is more
effective for decoding.
Because of the flexibility and minimal assumptions BTTDA
can equally be applied to other neuroimaging modalities (MEG, ECoG, fNIRS,
fMRI, EMG, ...) or tensor classification problems in other fields.

\section*{Acknowledgements}
We thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.
We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
work.

### 2024-09-30 chapters/patients.tex
Compiling

@@
 \section{Introduction}
+\todo{rename patient to bci user and participant}
 
@@
 Several studies show that performance drops when not fixating the intended
-target~\cite{Brunner2010, Frenzel2011,
-Treder2010, RonAngevin2019, VanDenKerchove2024} \todo{[2, 1], our
-study}, necessitating gaze-independent solutions.
+target~\cite{Brunner2010, Frenzel2011, Treder2010, RonAngevin2019,
+VanDenKerchove2024}, necessitating gaze-independent solutions.
 
@@
   \input{chapters/patients_table.tex}
-  \caption{Presentation of included patients including their diagnosis and
-  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
+  \caption[Presentation of included patients including their diagnosis and
+  capabilities.]{Presentation of included patients including their diagnosis and
+  capabilities.
+  (Trach.: patient underwent a tracheotomy, Cls.: classification according
+  to~\textcite{Wolpaw2006}).
   }
@@
   \input{chapters/patients_table_eye.tex}
-  \caption{Vision and eye motor impairment for included patients.
+  \caption[Vision and eye motor impairment for included patients.]{%
+  Vision and eye motor impairment for included patients.
   Eye motor impairment was assessed with a combination of self-reported issues
-  by the subject and the NeuroEye~\cite{Hasan2022} test.
+  by the subject and the NeuroEye~\cite{Hassan2022} test.
   LogMAR: lower is better.}

### 2024-09-30 chapters/gaze_independence.tex
New patient table

@@
 \todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}
+
+\todo{
+14] Fazzi E, Signorini SG, Piana RL, et al. Neuro-ophthalmo-
+logical disorders in cerebral palsy: ophthalmological, oculo-
+motor, and visual aspects. Dev Med Child Neurol. 2012;54:
+730–736.
+[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
+visual function of patients with locked-in syndrome: report
+of 13 cases. J Fr Ophtalmol. 2016;39:437–440.
+[16] Moss HE, McCluskey L, Elman L, et al. Cross-sectional evalu-
+ation of clinical neuro-ophthalmic abnormalities in an
+amyotrophic lateral sclerosis population. J Neurol Sci. 2012;
+314:97–101.
+}
 
@@
 
-\todo{check funding proposal}
-\todo{check provisional doctoral plan}
-\todo{check WIP seminar}
+\todo{
+  Complete with refs from
+
+Fast detection of covert visuospatial
+attention using hybrid N2pc and SSVEP
+features
+}
 
@@
 \todo{mention, free, split, frenzel}
-\todo{Overt, covert, split, free, illustration from poster}
 \todo{clear definitions, table from paper}

### 2024-09-30 chapters/patients.tex
New patient table

@@
 Communication BCIs have a target population consisting of patients in various
-stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
+stages of paralysis or \ac{lis},
+severe motor impairment
+\todo{cite and discuss Usability and Workload of Access
+Technology for People With Severe Motor
+Impairment: A Comparison of Brain-
+Computer Interfacing and Eye Trackin}
+
+often suffering from oculomotor impairments.
+can introduce problems in bci ~\cite{Brain-computer interface (BCI) evaluation
+in people with amyotrophic lateral sclerosis}, discuss als example
+
+\todo{. As with eye tracking SGDs, visual
+and oculomotor impairments may limit PALS’s performance
+with visual BCI systems (Fried-Oken et al., 2020). Although
+}
+
+\cite{FriedOken2020}
+
 This reduces their performance operating visual oddball \ac{bci}s (see
@@
 able to use a hex-o-spell)}
+\todo{discuss Lesenfants2014}
+\todo{discuss Marchetti2013}
 
@@
 pursuing.
+
+Self-reported according to cite.
+Additionally, the participants were asked if they experienced any tremors like
+nystagmus or other (larger) involuntary eye movements.
 
@@
 
+The Coma Recovery Scale
+Revised (CRS-R) [73] provides standardized procedures for admin-
+istration and response criteria to more uniformly describe the vis-
+ual fixation or pursuit skills of a user with SSPI. The Northeastern
+State University College of Optometry (NSUCO) Oculomotor Test
+[74] provides specific procedural techniques as well as scoring cri-
+teria, which allows for uniform description of the accuracy and
+ability of pursuit and saccade skills for adults and children.
+
 In a follow up study, we aim to revisit these patients with an on-line system

### 2024-10-02 chapters/bci.tex
New patient table
Comments hakim

@@
 \todo{fix lacheck/chktex notifications}
+\todo{use terminology individuals with severe
+speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
+\todohsm{more subtitles}
+\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
+\todofc{chapters should look more like full papers}
 
-People suffering from severe paralysis or disability due to
+People with severe physical impairment due to
 neurodegenenerative disease, stroke, traumatic brain or spine injury
@@
 The key problem for severely paralyzed patients is that the mind wants to go
-where the body can't follow.
+where the body cannot follow.
 Couldn't we therefore design a system that directly interacts with the mind?
@@
 impaired speech~\cite{Wolpaw2018,Milekovic2018}.
-Impressive cutting edge examples of experimental use include a brain-spine
+Impressive cutting edge experimental examples include a brain-spine
 interface allowing a paraplegic patient to walk again\cite{Lorach2023},
@@
 \section{A direct interface to the brain}
+\todohsm{Start with state of the art before theses.}
+\todohsm{objectives at the end of chapter 1}
+\todohsm{Introduction part: chapter 1 rationale, chapter 2: literature/sota}
+\todohsm{make less verbose}
 
@@
 First of all, we need to identify a signal that is a direct representation of
-what's going on in the brain.
+what is going on in the brain.
 Section~\ref{sec:bci-recording} gives an overview of the available options.
@@
 It would help greatly if we knew what we are looking for.
-For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
-which interaction is performed.
+A \ac{bci} operates according to a specified paradigm, which defines the method by which interaction is conducted.
 This usually means that the user is instructed to perform a specific task, like
@@
 in Figure~\ref{fig:bci/loop}.
-The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
+The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
 exact task performed by the user, and is often closely linked to the manner of
@@
 Because of the desirable high temporal resolution~\cite{Easttom2021} and
-practicality electrophysiology, we will focus only on methods to record
-electromagnetical activity originating from neuronal action potentials.
+practicality of electrophysiology, we will focus only on methods to record
+electromagnetical activity originating from neuronal action potentials and
+postsynaptic potentials.
 The magnetical field of the brain can be measured using magnetoencephalography
@@
 electrical activity of the brain.
-As touched upon earlier,the \emph{invasiveness} of the technology forms an
+The \emph{invasiveness} of the technology forms an
 important concept in determining its suitability for a given application and
@@
 
-\section{\Ac{bci} paradigms}
+\section{Paradigms}
 \label{sec:bci/paradigms}
+\todohsm{bring more structure, more subtitles}
 
@@
 since their intentions or cognitive state state would be inferred from just
-their brain activity.
+their brain activity.\todohsm{in "passive bci", the user is not supposed to express "intentions". ragher start with active "}
 However, without active user participation, it can be hard to establish reliable
@@
 illiteracy~\cite{Allison2010}\footnote{%
-  \cite{Becker2022,Thompson2019}
+Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
+Instead of attributing poor performance to the user's inherent limitations,
+critiques point out that the issue lies more with the design of the
+systems themselves. BCI systems often fail to account for individual
+variability in brain anatomy and cognitive strategies. As a result, what has
+been labeled as "illiteracy" may, in fact, be a reflection of systems that are
+not flexible enough to adapt to different users. This perspective
+shifts the responsibility to the individual, rather than recognizing that
+better, more adaptive systems could overcome many of these performance
+challenges, thus standing in the way of progress.
+Moreover, the term ``illiteracy'' itself carries negative
+connotations, implying a user deficiency that reinforces normative assumptions,
+when in reality, a more inclusive approach to design could enhance BCI
+accessibility for a wider range of individuals. This evolving view calls for a
+reframing of the problem, focusing on improving the adaptability of BCIs rather
+than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
 }.
-\todo{footnote on how this just means it needs to be more adapted,
-inclusivity, harmful to progress, paraphrase abstracts of articles}
 \todo{user training more than with reactive, task instruction}
@@
 sensory stimulations towards which the user's attention can directed.
-Compared to active \ac{bci}, reactive paradigms usually require higher effort
+Compared to active \ac{bci}, reactive paradigms usually require more effort\todohsm{is this the case?}
 and can induce fatigue due to the constant stimulation.
@@
 
-\section{A case study: the visual oddball \ac{bci}}
+\section{A case study: the visual oddball \acs{bci}}
 

### 2024-10-02 chapters/conclusion.tex
Comments hakim

@@
 
-\section{General discussion}
+\section{Contributions}
 
-\subsection{Obtained results}
+\todo{come back on hypothesis}
 
-\subsubsection{\ac{erp} decoders}
+\subsection{Developed \Ac{erp} decoders}
 We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
@@
 
-\subsubsection{Gaze-independent \ac{bci} healthy controls}
+\subsection{Gathered datasets}
 The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
@@
 ERP...} observed for covert VSA also holds for split VSA.
+
+\todo{describe patients dataset}
+
+\subsection{Investigated gaze-independence visual \acsp{bci}}
 
@@
 
-\subsection{Patient case studies}
+\section{Limitations}
 
-\section{Limitations}
+hard to evaluate experience in long term home use
+unify clinical side and bci development side
 
@@
 
+\section{Recommendations}
+\subsection{Working with patients}
+gaze independent visual, can be good current existing solution
+take into accounrecommendations of \textcite{FriedOken2020}
+
+immediately move to patients and on-line implementation, to be able to report
+interpretable metrics
+permanent working collaboration as a lab
+
+In a mature field like visual bci, start from patients.
+avoid situations where a problem is solved that doesn't pose itself in a
+patient population
+use all available capabilities and signals
+
+%Have attention for the emotional impact
+
+\subsection{Linear decoding methods}
+
 \section{Current \& future work}
@@
 select best classifier using eye tracking
-
-\section{Recommendations}
-\subsection{Working with patients}
-immediately move to patients and on-line implementation, to be able to report
-interpretable metrics
-permanent working collaboration as a lab
-
-start from patients
-use all available signals
-\subsection{Linear decoding methods}
+expand on patient experiemnts
 

### 2024-10-02 chapters/gaze_independence.tex
Comments hakim

@@
 }
+\todohsm{move introduction of paper to paper}
 
@@
 \label{sec:gaze-independence}
+\todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}
 
@@
 with invasive recording.
-Non-visual reactive paradigms that use e.g. auditory and somatosensory
+Non-visual reactive paradigms that use e.g. auditory and
+somatosensory\todo{cite some}
 stimulation do not rely on gaze redirection exist but also result in lower information transfer
@@
 user-dependent variability~\cite{Severens2014,Reichert2020b}.
+\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
+outperformed a somatosensory alternative in participants with \ac{als} whose
+eye motor capabilities were effectively impaired.
+
 \todo{insert table from riccio in text as numbers, not as table}
 
-Secondly, current visual stimulation paradigms can be optimized, e.g. such that
+This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
 the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
-Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
+Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
 the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
@@
 
+\todohsm{ move next two paragraphs to introduction, more context than objectives}
 Brain-Computer Interfaces (\ac{bci}s) decode brain activity with the aim to establish a direct communication
@@
 control devices.
-The quest for performant and affordable solutions is most evident in
-visual \ac{bci}s based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
-and paradigms tailored to the needs of specific stimulation paradigms and
-use contexts.
-An effective and proven method is the visual event-related potential (\ac{erp})-based
-oddball interface.
+An effective and proven method is the \ac{eeg}-based visual event-related potential (\ac{erp})-based
+oddball interface~\cite{Wolpaw2018,Severens2020}.
 Here, targets are shown in short flashes on a computer screen.
@@
 
-This leads to the often unadressed gaze-dependence paradox: the \ac{bci} target
+This leads to the often underadressed gaze-dependence paradox: the \ac{bci} target
 population consists of patients who suffer from various degrees
@@
 Hence, our working hypothesis is as follows:
+\todo{rewrite, this is not what we tested in the patient paper after all}
 \begin{quote}
@@
 \subsection{Interface design}
+\todohsm{rename data collection. start with decoder design, second section data collection with experiments}
 

### 2024-10-02 chapters/patients.tex
Comments hakim

--- 2024-09-30 chapters/patients.tex
@@
 
-Communication BCIs have a target population consisting of patients in various
-stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
-This reduces their performance operating visual oddball \ac{bci}s (see
-section\todo{ref section} for an overview), since they
-are presumed unable to comfortably redirect their gaze at the desired target,
-i.e., perform overt \ac{vsa}.\todo{cite some visual gaze-independent bci}
-They might more naturally operate in covert \ac{vsa}, where the gaze and
-\ac{vsa} do not coincide.
-Several studies show that performance drops when not fixating the intended
-target~\cite{Brunner2010, Frenzel2011, Treder2010, RonAngevin2019,
-VanDenKerchove2024}, necessitating gaze-independent solutions.
+BCIs for communication assistive technology~\cite{Millan2010}
+have a target population consisting of patients people with \ac{sspi}.
 
-Usually, they aim to improve gaze-independent \ac{bci} by optimizing
-performance in cued covert \ac{vsa}.
+Yet, there is a large comorbidity between \ac{sspi}, and eye motor impairment~\cite{FriedOken2020}.
+Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
+vision), and ophthalmoplegia (eye paralysis) can significantly hinder
+the ability to use visual BCIs. These impairments make it difficult for
+individuals to focus on or track visual stimuli accurately, reducing their
+performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
+Unfortunately, it is again for this group that eye tracking solutions also
+perform poorly, making them more reliant on potential developments in \ac{bci}
+that do not rely on eye gaze.
+
+Eye motor impairments are presumed to reduce performance in operating visual
+oddball \ac{bci}s (see section\todo{ref section} for an overview), since users
+cannot comfortably redirect their gaze at the desired target,
+i.e., perform overt \ac{vsa}.
+This is usually circumvented by designing gaze-independent~
+\acp{bci}~\cite{Riccio2012}.
+These interfaces either avoid visual stimulation, or exploit some form of
+covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.
+
+Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
+target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
+gaze-independent solutions.
 These studies build on the assumption that patients with eye motor impairment
@@
 fixation is maintained, could also be considered gaze-dependent.
-This does not account for the residual eye motor capabilities of most severly
-paralyzed or locked-in patients and the comfort they experience while
-performing the gaze fixation.
+This does not account for the residual eye motor capabilities of most people
+with \ac{sspi}, the (dis)comfort they experience while
+performing gaze fixation and other confounding factors due to their eye
+motility or vision.
 
-Our previous study, presented in chapter \todo{ref chapter}  partially tried to account for this
-and showed gaze-independent performance can be improved
+It is a striking constatation that studies reporting on
+gaze-independent visual \ac{bci} with people with \ac{sspi} and eye-motor
+impaired are very few.
+Results are usually different than those obtained with healthy control
+participants in the lab, due to difference in capabilities, brain
+response, equipment and environment.
+
+\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
+participants with \ac{lis} yet only exceeded chance level accuracy in two.
+More recently, \textcite{Peters2020} performed a trial with a
+interface with two participants with late-stage \ac{als} and visual impairment.
+Their system showed high accuracy, outperforming an eye tracking alternative.
+It would be of interest to verify if such results can be verified with other
+patient groups and visual oddball \acp{bci}.
+
+\textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
+participants with \ac{als} and showed that this visual oddball interface optimized
+for gaze-independence can outperform a tactile gaze-independent \ac{bci}.
+While this speaks to the power of visual paradigms even in groups that are
+expected to have eye motor impairment, they did not verify the gaze direction
+of participants during the experiment.
+It was suggested that patients were performing overtly.
+Participants with \ac{als} also had a substantially lower accuracy patients lower than healthy
+(58\% vs 88\%).
+
+Our previous study, presented in chapter \todo{ref chapter} also used the
+visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
+This work partially accounted for the idea that gaze impaired patients might
+not full rely on central gaze fixation and evaluates settings that are not
+strictly dependent on this.
+We showed gaze-independent performance can be improved
 improved in healthy subjects using a suited decoding strategy.
-Yet, there is a strong need for verification of results obtained in healthy
-samples.
-It is a striking constatation that studies reporting on
-gaze-independent visual \ac{bci} in patients that are actually eye-motor
-impaired are very few.
-in patients, cite different characteristics, differences
-in setting, environment and equipment,
-\todo{List conclusions of Peters2020 (severely eye motor impaired als patients
-able to use a hex-o-spell)}
+that accounts for latency jitter in covert \ac{vsa} responses improves
+Yet, there is a strong need for verification of these result in an applied
+setting with people with \ac{sspi}.
+
+\todo{discuss \cite{FriedOken2020}}
 
 Eventually, one of the end goals of this research line is to develop
-gaze-independent \ac{bci} for patients that
-are fully locked-in and have no option left than to use a \ac{bci}.
+gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
 However, this group is very small and it is often a challenge to recruit them
@@
 solutions tailored to their specific situation.
-
 Therefore, we aim to apply the concepts from earlier work an literature to
-patients with various degrees motor impairment and various
-degrees of eye motor impairment.
+people with \ac{sspi} and various degrees motor impairment in a visuall oddball
+\ac{bci}.
 The objectives of this case study are as follows:
@@
     neurological or psychiatric disorder other than those of interest based on
-    inclusion criteria~\ref{item:patients/inclusion/wolpaw}
+    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
     and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
@@
 Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
-various eye motor disorders are common depending on the exact location
+various eye motor disorders are common depending on the exact etiology
 ~\cite{Bogousslavsky1987, Moncayo2009}.
@@
   capabilities.
-  (Trach.: patient underwent a tracheotomy, Cls.: classification according
+  (Trach.: patient underwent a tracheostomy, Cls.: classification according
   to~\textcite{Wolpaw2006}).
@@
 
-\subsection{Eye tracking and eye motor examination}
+\subsection{Visual skills and eye tracking and eye motor examination}
 
-Self-reported eye motor abnormalities were recorded.
-Patients were asked for, the paralysis or inability to move in a given
-direction of one or both eyes, the inability to pursue an object to specific
-positions, discomfort or fatigue fixating at specific positions or in general
-when performing visual tasks and eye tremors while fixating, resting or
-pursuing.
+Self-reported eye motor and visual abnormalities were recorded according to the
+relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
+These include visual acuity, visual fixation, eyelid function, ocular motility
+, binocuar vision and field of vision.
+Additionally, participants were asked about eye tremors (nystagmus or other) and
+other involuntary eye movements.
 
-Additionaly, we implemented and performed the automated NeuroEye eye movement
-test proposed by~\textcite{Hassan2022}.
+As an objective metric, we implemented and performed the automated NeuroEye eye movement
+test proposed by~\textcite{Hassan2022} using calibration-free eye tracking to
+check if it revealed any further eye motor abnormalities.
+This was not the case.
 This test is based on calibration-free eye tracking.
@@
 \todo{include eye test document juliette}
+
 Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
 included patients.
+All participants reported some degree of fatigue or discomfort when fixating.
 Patient PA1 had the mildest impairment, only reporting fatigue when fixating
 for prolonged times.
-Eye motor function of subjects PC2 and PC3 were most severly affected.
-Patient PC2 was only able to look up and down, and patient PC3 only retained
-partial movement of the right eye, while the left eye was permanently closed.
+The \ac{fa} patients were mostly affected by eye tremors and impaired pursuit.
+Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
+Participant PC2 was only able to look up and down and had a deviation in the
+left eye causing diplopia, but this was corrected by a prism glass.
+Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
+Participant PC4 had one deviated eye with a corneal abcess affecting the motility
+and vision in the right eye, and suffered reduced motility in the left.
+\todo{go through notes again}
 
@@
   \input{chapters/patients_table_eye.tex}
-  \caption[Vision and eye motor impairment for included patients.]{%
-  Vision and eye motor impairment for included patients.
-  Eye motor impairment was assessed with a combination of self-reported issues
-  by the subject and the NeuroEye~\cite{Hassan2022} test.
-  LogMAR: lower is better.}
+  \caption[Visual skills of the included participants.]{%
+  Visual skills of the included participants.
+  Visual~\ac{bci} skills~\cite{FriedOken2020}
+  was assessed with a combination of self-reported issues
+  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
+  impaired, \BigDiamondshape\ severely impaired. Visual acuity in logMAR, lower
+  is better.}
   \label{tab:patients/eye}
@@
 
-Eye tracking throughout the experimental session was performed using the Tobii~\todo{model, make}
-portable eye tracker.
-\todo{describe analysis during experiment}
+Finally, we also record gaze position throughout the experimental session, to
+register the participant's gaze relative to the stimulated \ac{bci} targets.
 
@@
 Similar to this study, the task consists of counting the flashes of a cued
-target among all flashing targets.
-We refer to section~\ref{sec:covert-align/stimulation} for the exact details.
+target among 6 round, flashing targets laid out in a hexagonal pattern in the
+field of view of the user.
+We refer to section~\ref{sec:covert-align/stimulation} for implementation details.
 
-To adapt the interface for the patient setting, the number of blocks was
-decreased to 6 per \ac{vsa} setting.
+3 different \ac{vsa} settings were explored
+In overt \ac{vsa}, the participant was instructed to fixate the cued target, or
+try to the maximum extent of their visual skill, even if experiencing slight
+discomfort.
+In covert \ac{vsa}, the participant was instructed to fixate the center of the
+screen, also to the extent of their ability.
 An additional \emph{free \ac{vsa}} condition was introduced.
@@
 comfortable.
+This allows us to investigate the user's natural way of operating the \ac{bci}
+given their individual set of visual skills.
+If the participant was not fully paralyzed, they were instructed not to move the head.
 The cued \emph{split attention} setting proposed
 by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
-in natural \ac{vsa} operation settings for gaze-impaired patients.
-If the patient was not fully paralyzed, they were instructed not to move the head.
+in natural \ac{vsa} operation settings for gaze-impaired individuals.
+
+To make the interface suitable for use by individuals with
+\ac{sspi}~\cite{FriedOken2020}, the
+number of blocks was decreased to 6 per \ac{vsa} setting.
 \Ac{isi} was increased to\todo{number} to decrease task difficulty.
-The experiment alsy started with a training block in each condition, where the
-patient was instructed with feedback on their performance to make sure they
+The experiment also started with a training block in each condition, where the
+participant was instructed with feedback on their performance to make sure they
 understood and were able to perform the task.
@@
 
-Mention change in ISI
-Mention less rounds
-
 \subsection{Data collection \& preprocessing}
 
-\todo{check checklist}
-\subsection{BCI Decoding}
+During the recording session, patients were positioned in their wheelchair in front of a table.
+Stimuli were presented on a laptop \todo{model, make} placed at \todo{distance}
+A Cedrus StimTracker \todo{make} ensured synchronization of stimuli with the
+recorded \ac{eeg}.
+Eye tracking throughout was performed using the Tobii~\todo{model, make}
+portable eye tracker placed at the bottom of the laptop screen.
+
+\Ac{eeg} was recorded at 1000Hz using the Neuroscan Neuvo  portable amplifier (Compumedics Neuroscan,
+Australia) connected to a second laptop for registration
+The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
+according to the international 10-20 layout.
+Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$
+Additionally, the \ac{eog} was recorded.
+
+The \ac{eeg} was band-pass filtered between 0.5 and 16Hz.
+Bad channels were rejected using on the RANSAC algorithm~\cite{Fischler1981}
+and visual inspection.
+Next, the \ac{eeg} was rereferenced to the average of mastoid electrodes TP9
+and TP10, and \ac{ica} was performed to reject artifactual components based on
+correlation with the \ac{eog} or by visual inspection.
+Epochs are cut from -0.1 to 0.9s relative to stimulus onse, no baseline
+correction was performed.
+
+Eye tracking data was cleaned by fusing left and right gaze into one channel
+for the horizontal and vertical gaze position.
+If both were present for a given sample, the fused channel was the mean of both
+values.
+If at a given sample either the left or the right eye was not detected for a
+given channel, the value of the other one was adopted.
+If both were missing, the gaze position remains unset at that time point and no
+interpolation is performed.
+
+\subsection{BCI decoding}
+
+\todo{shortly describe WCBLE}
+\todo{shortly describe tLDA}
+\todo{shortly describe xdawncov+ts+LDA}
 
@@
 \subsection{Eye tracking analysis}
+\label{sec:patients/outcomes/gaze}
 
-PC2 reported that he was unable to fixate the lower two targets, and th
+First, we aimed to shed more light on the actual capabilities of gaze impaired
+individuals with respect to performing overt \ac{vsa} and central gaze
+fixation, as well as investigate how relevant these two settings are when the
+gaze is not cued.
+Figure~\ref{fig:patients/gaze} maps gaze position relative to the stimuli
+across conditions.
+These results should be interpreted with care, since the eye tracker to some
+degree relies on functioning eye motility.
+The participant's position
+relative due the eye tracker might have shifted throughout the experimental
+session despite our best efforts, e.g.\ because they needed aspiration of their
+tracheostomy.
 
@@
 }
+\todo{explaine crosses}
 
-\subsection{Decoding performance}
+PA1 had relatively intact gaze control and was able to correctly perform the
+cued overt and covert settings.
+When gaze was uncued he fixated the cued target.
+This was also mostly the case for PB1, although eye tracking revealed that he
+chose not to perform central gaze fixation when cued in at least one of the
+stimulation blocks. We were unable to record his gaze near the bottom left
+stimulus position, either due to eye tracker failure or because the participant
+was not comfortable fixating this position.
+Eye tracker calibration did not succeed for subject PB4, but given
+transformations of gaze positions to the stimulus space, thy
+
+PB2 was able to perform overt \ac{vsa} and central fixation to some extent,
+yet eye tracking, but gaze position distributions shows a larger spread than
+for PA1 and PB1.
+The distribution in free \ac{vsa}, however, showed that she prefered to attend
+stimuli covertly when the gaze was uncued.
+This was confirmed by the participant.
+
+The overt and central gaze fixation settings were also not properly adapted to
+particpant PC4.
+In free \ac{vsa} eye tracker results show that his gaze was usually near the
+bottom two targets, indicating some degree of covert or split \ac{vsa}.
+
+It was technically impossible to register gaze position with the Tobii
+\todo{model} for participants PC2 and PC3 since they both had one eye that was
+occluded respectively by the prism glass and the eyelid.
+Both participants reported they could not fixate some of the
+stimuli.
+
+\subsection{\Ac{bci} decoding performance}
+
+Figure~\ref{fig:patients/decode} shows single-trial
+\ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.
+Scores were obtained using 6-fold cross-validation where folds corresponded to
+stimulation blocks.
 
@@
   \input{figures/patients/fig_decode.pgf}
+  \caption[%
+    Decoding performance in different \ac{vsa} settings.
+  ]{%
+    Decoding performance in different \ac{vsa} settings reported as
+    single-trial \ac{rocauc}.
+    Free \ac{vsa} is generally on par with overt \ac{vsa}.
+    Performance in covert \ac{vsa} with central gaze fixation is lower, but can
+    be improved with the \ac{wcble} decoder.
+    95\% confidence intervals were calculated using 1000 bootstrapping
+    repetitions.
+  }
+  \label{fig:patients/decode}
 \end{figure}
+\todo{rename subjects}
 \todo{check if percentages or 0-1 accuracy and roc-auc}
 
-For PB2 and PC4 performing covertly, was free decoding higher than covert
-decoding?
+In overt \ac{vsa}, evaluated decoders performed about equal on average
+(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, tLDA 75.99\%).
+In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
+but \ac{wcble} can significantly improve performance over its base classifier
+\ac{tlda} in this condition
+(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, tLDA 59.05\%).
+Decoding performance for this task was at chance level for participants PB4 and
+PC3.
+
+However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
+XDAWNCov+TS+LDA performance was slightly lower here (although not
+significantly).
+(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, tLDA 74.27\%).
+More interestingly, we notice that performances of gaze impaired individuals
+with \ac{sspi} in free \ac{vsa} are close to those in overt \ac{vsa}.
+A substantial decrease in performance from overt to free is present for
+subjects PC3 and PC4.
+For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
+setting, there was no decrease in performance.
 
 \subsection{Cross-condition calibration}
+\label{sec:patients/outcomes/cross}
 
-| indication of fatigue, cinsistency within free case
-| closer to overt or to covert by looking at eye tracker
-|
-| will learn
-| split up per target
-|
-| Findings:
-| 1. Some patients prefer to do covert (check performance results for them),
-|    analyze distance to target
-| 2. covert attention can be improved using WCBLE for all subjects, but not to
-| the
-|    level of overt covert attention works best when training in overt
-| attention
-| s this also true for patients preferring covert?
-| 3. free performs on par (in all subjects?), opens debate of necessity of
-|    gaze-independence BCIs
-| 4. covert attention works best when training in overt attention. Is this also
-|    true for patients preferring covert?
-
-2024-09-18 04:39:35 PM can more overt attention epochs closer to
-| mention problem can't know if patient shifted or eye tracker shifted
-| (difficult mobile gaze tracker)
-| overt lower because can't perform
+\todo{input and describe graph}
 
@@
 
-patients 2 are progressing towards 3
+Due to the heterogenous nature of the participant's conditions, it is hard to support
+claims about significance of presented results.
+This study should therefore be seen as a collection of case studies,
+highlighting different obstacles encountered in developing gaze-independent
+visual oddball \acp{bci} for individuals with \ac{sspi} and gaze impairment.
+Nevertheless, we would like to highlight some general aspects that might be
+of interest for the further development of this class of \acp{bci}.
 
-Limitations
+True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
+Hence our analysis centers around the free \ac{vsa} condition.
+Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
+confirm our assumption that voluntary covert \ac{vsa} can
+occur in individuals with \ac{sspi} and gaze impairment.
+We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
+chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
+fixation can be improved by accounting for latency jitter, and showed that this
+holds for individuals with \ac{sspi} and gaze impairment.
 
-Limited sample of patients, FA maybe not most interesting group to represent
-real BCI users
+Contradicting our assumptions, however, we have shown that this does not
+necessarily improve covert \ac{vsa} when gaze fixation is not cued.
+One possible explanation for this is that actively performing central gaze fixation
+increases task load.
+This in turn reduced overall performance, even though the participant might
+otherwise also have performed covert \ac{vsa}, but would not be occupied by
+maintaining strict central gaze fixation.
+These extended task demands are not present in the free \ac{vsa} condition, so
+there is less performance to be gained here.
+Furthermore cued central gaze fixation combined with counting flashing stimuli
+in the visual periphery, is an explicit example of a dual task.
+Dual tasks have been shown to increase P3 latency
+jitter~\cite{Polich2007,Arico2014, VanDenKerchove2024}
+which is what \ac{wcble} accounts for.
+Hence,increased P3 jitter might be more related to maintaining central gaze fixation
+than to the actual covert \ac{vsa} aspect.
 
-interpreting eye tracker results hard because of inaccuracy eye tracker for eye
+The seamingly stable performance across overt and free \ac{vsa} could be
+misinterpreted as an indication that the Hex-o-Spell \ac{bci} already works
+well for patients with \ac{sspi} and gaze impairment, and no optimization is
+needed.
+However, we assume that overt \ac{vsa} performance was also decreased in some
+subjects or for some blocks if the participant was not able to comfortably
+perform the task.
+Nevertheless, the large difference between free \ac{vsa} and covert \ac{vsa}
+with central gaze fixation, is food for thought about the applicability of
+solutions developed with central fixation in mind.
+
+Individuals with all but the most severe gaze impairments will probably retain
+some degree of gaze direction in visual \ac{bci} operation, which can
+drastically boost performance.
+Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with
+This is also supported by our results on cross contidion calibration
+
+\todo{discuss that training in overt always increases covert/free, focus on
+increased P3 amplitude and relying on visual components, also propose this
+strategy if possible}
+
+\section{Limitations}
+
+Despite results that prompt reflection on gaze-independent \ac{bci} approaches,
+there are some limitations to the presented results that need to be
+addressed in current and future work.
+
+First and foremost, this study works with a limited sample of patients which
+does not represent the full spectrum of individuals with \ac{sspi} and their
+oculomotor impairment.
+\Ac{fa} patients met the inclusion criteria, but they are usually not
+considered as one of the typical interest groups of \ac{bci} communication assistive
+technology, partly due to the rarity, and partly due to the disease
+progression.
+It would be most interesting to verify these results with individuals with
+\ac{lis} and no eye movement capability at all.
+
+Another limiting factor are the difficulties in correctly interpreting eye tracker
+results in studies with eye motor impaired patients.
+If eye tracking at all is possible, it is not guaranteed that the user is able
+to perform the calibration procedure succesfully.
+Further experiments should be carried out with a stationary eye tracker with
+more advanced capabilities, although systems using a head fixator or headrest
+should be avoided.
+This is not practical when working with
+wheelchair-bound individuals who might have undergone a tracheostomy.
+hard because of inaccuracy eye tracker for eye
 motor impaired patients. Redo with more advanced stationary eye tracker and
@@
 
-advanced eye tracking equipment is necessary
+In this study, user comfort in the different conditions was not objectively
+measured.
+Rather, it was assumed that patients operated most comfortably in the free
+\ac{vsa} condition.
+In order to properly contextualize performance results, they should be coupled
+with a measure of user comfort and percieved effort, like the NASA Task Load
+Index~\cite{Hart2006}.
+Performance might after all be traded off for user comfort.
+Eye motor disability can also have been assessed more objectively, using \todo{
 
-user comfort was not objectively measured using established questionnaires
+The Coma Recovery Scale
+Revised (CRS-R) [73] provides standardized procedures for admin-
+istration and response criteria to more uniformly describe the vis-
+ual fixation or pursuit skills of a user with SSPI. The Northeastern
+State University College of Optometry (NSUCO) Oculomotor Test
+[74] provides specific procedural techniques as well as scoring cri-
+teria, which allows for uniform description of the accuracy and
+ability of pursuit and saccade skills for adults and children.
+}
 
-Further work
-Comparing with a sample of non-gaze impaired patients with  neuromuscular
-disability DMD
-On-line session with these patients
-
-In a follow up study, we aim to revisit these patients with an on-line system
-to be able to calculate a realistic \ac{itr} and measure their user experience.
+Finally, the parameters stimulation procedure
+from~\textcite{VanDenKerchove2024} were adapted to make the counting task
+accessible to the participants with \ac{sspi} and gaze impairment.
+The number of repetitions or \ac{isi} were however not optimized to achieve
+maximal \ac{itr}.
+An interface that aims to maximize \ac{itr} could necessitate more and faster
+gaze redirections, which might result in different conclusions regarding the
+comfort and the effect of visual skill.
 
 \section{Conclusion}
+
+To summarize, we present our three main findings.
+\begin{enumerate*}
+  \item We have confirmed that some individuals with \ac{sspi} and gaze
+    impairment find it more comfortable to operate a visual gaze-independent
+    \ac{bci} relying on covert \acf{vsa}.
+  \item We have shown that covert \acf{vsa} with central gaze fixation can be
+    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
+    but this does not hold true if central gaze fixation is not imposed.
+  \item We have shown that covert \ac{vsa} \todo{with or
+    without imposed gaze fixation} decoding can alternatively be improved by
+    using a calibration session with attempted overt \ac{vsa}.
+\end{enumerate*}
+
+In general, this study shows that approaches tailored to the individuals visual
+skills are necessary and that a thorough rethinking of the approach to
+designing gaze-independent visual \acp{bci} is an area of concern.
+Future work includes comparing these results with a sample of non-gaze impaired
+participants with \ac{sspi}
+and carrying out repeated on-line sessions with these patients with built-in feedback to
+properly assess \ac{itr} and learning effects, and obtain realistic measures of percieved effort,
+comfort and performance.
 

### 2024-10-02 code/patients/fig_decode.py

@@ -20,6 +20,7 @@ df_avg = df.copy()
 df_avg["subject"] = "avg."
 df = pd.concat([df, df_avg], ignore_index=True)
 
+pdb.set_trace()
 
 fg = sns.catplot(
     data=df,

### 2024-10-02 code/patients/fig_gaze.py

@@ -61,11 +61,17 @@ for i, (subject, sub) in enumerate(subjects.items()):
         axs[i, j].set_axis_off()
         for target in ["center", 0, 1, 2, 3, 4, 5]:
             pos = get_target_pos(target)
+            if condition=='overt':
+                c=colors["lightgray"] if target == "center" else colors["accent1"]
+            elif condition=='covert':
+                c=colors["accent1"] if target == "center" else colors["lightgray"]
+            elif condition == 'free':
+                c=colors["lightgray"]
             axs[i, j].scatter(
                 pos[0],
                 pos[1],
                 s=50,
-                c=colors["accent2"] if target == "center" else colors["accent1"],
+                c=c,
                 marker="+",
                 linewidth=0.5,
             )

### 2024-10-02 chapters/bci.tex
Comments Marc

@@
 \todofc{chapters should look more like full papers}
+\todo{uniform vsa condition vsa setting}
+\todo{acronyms written in full every new chapter and in captions}
 
@@
   \includegraphics[width=\textwidth]{figures/bci/bci_loop.pdf}
+}{%
+  The \ac{bci} loop.
 }{%
@@
 }{%
-	Different electrophysiology recording modalities and \\
+  Various electrophysiology recording technologies.
+}{%
+	Various electrophysiology recording technologies and \\
 	their respective position in a cross-section of the skull. \\
@@
   \input{figures/bci/paradigms.tikz.tex}
+}{%
+  Overview of \ac{bci} paradigms.
 }{%

### 2024-10-02 chapters/bttda.tex
New patient table
Comments Marc

@@
 	\input{figures/bttda/HODA_backward.tikz.tex}
-	\caption{A visualization of the multilinear projection obtained by Higher Order
+  \caption[The \ac{hoda} backward projection.]{%
+    A visualization of the multilinear projection obtained by Higher Order
     Discriminant Analysis (HODA) applied to a third-order tensor
@@
 \begin{algorithm}
-	\caption{Higher order Discriminant Analysis (HODA) backward solution}
+  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
 	\label{alg:HODA}
@@
 	\input{figures/bttda/HODA_forward.tikz.tex}
-	\caption{The forward projection for HODA. By calculating activation
+  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA. By calculating activation
 		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
@@
 \begin{algorithm}
-	\caption{Higher order Discriminant Analysis (HODA) forward solution}
+  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
 	\label{alg:HODA-fw}
@@
 	\input{figures/bttda/BTTDA.tikz.tex}
-	\caption{A forward model for Block-Term Tensor Discriminant Analysis
+  \caption[A forward model for \ac{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
 		(BTTDA). BTTDA can extract more features
@@
 \begin{algorithm}
-	\caption{Block-term Tensor Discriminant Analysis (BTTDA)}
+  \caption{\Ac{bttda}.}
 	\label{alg:BTTDA}
-	\input{chapters/bttda/alg_BTTDA.tex}
+	\input{chapters/bttda/alg_bttda.tex}
 \end{algorithm}
@@
 \begin{algorithm}
-	\caption{Greedy model selection}
+  \caption[The greedy model selection procedure.]{Greedy model selection}
 	\label{alg:model-selection}
@@
 	\input{figures/bttda/blocks.pgf}
-	\caption{Normalized Mean Square Error (NMSE) (left), and difference in
+  \caption[Analysis of NMSE and classification score per block.]{%
+    Normalized Mean Square Error (NMSE) (left), and difference in
     classification score for the training and validation for the ERP datasets
@@
 	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
-	\caption{Spatial (left two columns) and temporal (middle column) activation patterns and
+  \caption[Extracted \ac{bttda} activation patterns.]{%
+    Spatial (left two columns) and temporal (middle column) activation patterns and
 		condition contrasts (right column) obtained after forward projection of the latent

### 2024-10-02 chapters/conclusion.tex
Comments Marc

@@
 \section{Current \& future work}
+kronecker-sum lda
 multicomponent alignment approach

### 2024-10-02 chapters/gaze_independence.tex
Comments Marc

@@
 	\end{tabular}
-	\caption{Incidence of eye motor impairment in selected \ac{bci} target
+  \caption[Incidence of eye motor impairment in selected \ac{bci} target
+    patient populations.]{%
+      Incidence of eye motor impairment in selected \ac{bci} target
       patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
       \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
-    $+$: frequent, $-$: infrequent.}
+      $+$: frequent, $-$: infrequent.
+    }
     \label{tab:incidence}
@@
 \todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}
+\todo{in a paragraph describe each approach and its up and downsides}
+\todo{move form copied introduction}
 
+\subsection{Gaze-independent modalities}
 Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
@@
 
+\subsection{Gaze-independent visual stimulation}
 This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
@@
 Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
-the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
+the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
 speller~\parencite{Acqualagna2011}.
@@
 
+\subsection{Gaze-independent decoding}
 Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
@@
 multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like Rapid Serial Visual
-Presentation\todo{cite},
+and temporally organized interfaces, like \ac{rsvp}\todo{cite},
 where targets or small sets of targets are shown consecutively.
@@
 Figure~\ref{fig:gaze-vsa}.
-\fourpanefig{%
-  \figpane{%
+
+\begin{figure}
+  \begin{subfigure}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
-  }{%
-    \emph{Overt \ac{vsa}}.
+    \caption[Overt \ac{vsa}]{%
+      \emph{overt} \ac{vsa}.
       Gaze and \ac{vsa} coincide on a target.
-  }{%
-    fig:gaze/vsa/overt
     }
-}{%
-  \figpane{%
+    \label{fig:gaze/vsa/overt}
+  \end{subfigure}\hfill%
+  \begin{subfigure}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
-  }{%
-    \emph{Covert \ac{vsa}}.
+    \caption[Covert \ac{vsa}]{%
+      \emph{Covert} \ac{vsa}.
       Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
-  }{%
-    fig:gaze/vsa/covert
     }
-}{%
-  \figpane{%
+    \label{fig:gaze/vsa/covert}
+  \end{subfigure}
+
+  \begin{subfigure}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
-  }{%
-    \emph{Split \ac{vsa}}.
+    \caption[Split \ac{vsa}]{%
+      \emph{Split} \ac{vsa}.
       \Ac{vsa} is directed towards a target, while the gaze rests on another.
-  }{%
-    fig:gaze/vsa/split
     }
-}{%
-  \figpane{%
+    \label{fig:gaze/vsa/overt}
+  \end{subfigure}\hfill%
+   \begin{subfigure}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
-  }{%
-    \emph{Free \ac{vsa}}.
+    \caption[Free \ac{vsa}]{%
+      \emph{Free} \ac{vsa}.
+      Gaze and \ac{vsa} coincide on a target.
       The user is free to direct their gaze as they deem most comfortable
       All previous \ac{vsa} conditions are possible.
-  }{%
-    fig:gaze/vsa/free
     }
-}{%
+    \label{fig:gaze/vsa/overt}
+  \end{subfigure}\hfill%
+  \caption[\Ac{vsa} conditions]{%
     \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
-}{%
-  fig:gaze-vsa
   }
-
+  \label{fig:gaze-vsa}
+\end{figure}
 In the overt case, users gaze at the cued

### 2024-10-02 chapters/patients.tex
Comments Marc

@@
 \todo{rename patient to bci user and participant}
+\todomvh{separate terminology for sspgi}
+\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}
 
-BCIs for communication assistive technology~\cite{Millan2010}
-have a target population consisting of patients people with \ac{sspi}.
+\todo{A systematic review of research on augmentative and alternative communication brain-computer interface systems for individuals with disabilities}
+
+\Acp{bci} for communication assistive technology~\cite{Millan2010}
+target individuals with \ac{sspi}.
+
+visual bci
 
@@
 central fixation.
-One could argue that a \ac{bci} that is only verified to work when a central
+One could argue that a \ac{bci} that is only verified to work when central
 fixation is maintained, could also be considered gaze-dependent.
@@
 with \ac{sspi}, the (dis)comfort they experience while
-performing gaze fixation and other confounding factors due to their eye
-motility or vision.
+performing gaze fixation and other confounding factors resulting from their eye
+motility.
 
@@
 participants with \ac{lis} yet only exceeded chance level accuracy in two.
-More recently, \textcite{Peters2020} performed a trial with a
-interface with two participants with late-stage \ac{als} and visual impairment.
-Their system showed high accuracy, outperforming an eye tracking alternative.
+More recently, \textcite{Peters2020} performed an trial with an
+two participants with late-stage \ac{als} and visual impairment.
+Their \ac{ssvep} interface was not optimized for gaze-independence, but the
+system showed high accuracy, outperforming an eye tracking alternative.
 It would be of interest to verify if such results can be verified with other
@@
 
+\textcite{Orhan2012} and \textcite{Oken2014} tested the \ac{rsvp} speller with
+individuals with \ac{lis}
+
 \textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
 participants with \ac{als} and showed that this visual oddball interface optimized
-for gaze-independence can outperform a tactile gaze-independent \ac{bci}.
+for gaze-independence can outperform a tactile \ac{bci}.
 While this speaks to the power of visual paradigms even in groups that are
@@
 It was suggested that patients were performing overtly.
-Participants with \ac{als} also had a substantially lower accuracy patients lower than healthy
-(58\% vs 88\%).
+Participants with \ac{als} also had a substantially lower accuracy than healthy
+controls (58\% vs 88\%).
 
@@
 This work partially accounted for the idea that gaze impaired patients might
-not full rely on central gaze fixation and evaluates settings that are not
+not fully rely on central gaze fixation and evaluates settings that are not
 strictly dependent on this.
@@
 
-\fourpanefig{%
-  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}}{A
-  patient with the stimulation and recording setup.}{fig:patients/photos/side}
-}{%
-  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}}{subcaption}{fig:patients/photos/front}
-}{%
-  the interface
-}{%
-  4
-}{%
-  caption
-}{fig:patients/photos}
+\begin{figure}
+  \begin{subfigure}{.45\textwidth}
+    \includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}
+    \caption{A patient with the stimulation and recording setup.}
+    \label{fig:patients/photos/side}
+  \end{subfigure}\hfill%
+  \begin{subfigure}{.45\textwidth}
+    \includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}
+    \caption{caption}
+    \label{fig:patients/photos/front}
+  \end{subfigure}
+  \caption{The recording setup and stimulation interface.}
+  \label{fig:patients/photos}
+\end{figure}
+\todo{caption}
+\todo{include stimulation interface}
 
@@
 
-In overt \ac{vsa}, evaluated decoders performed about equal on average
-(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, tLDA 75.99\%).
+In overt \ac{vsa}\todomvh{specify setting when talking about the setting}, evaluated decoders performed about equal on average
+(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
 In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
@@
 \ac{tlda} in this condition
-(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, tLDA 59.05\%).
+(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
 Decoding performance for this task was at chance level for participants PB4 and
@@
 significantly).
-(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, tLDA 74.27\%).
-More interestingly, we notice that performances of gaze impaired individuals
-with \ac{sspi} in free \ac{vsa} are close to those in overt \ac{vsa}.
+(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
+More interestingly, we notice that performances of the decoders in free
+\ac{vsa} are close to those in overt \ac{vsa}.
 A substantial decrease in performance from overt to free is present for
@@
 Due to the heterogenous nature of the participant's conditions, it is hard to support
-claims about significance of presented results.
+claims about significance of presented results.\todomvh{
+weird formulation
+hard to draw general conclusions}
 This study should therefore be seen as a collection of case studies,
@@
 
-True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
+True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.\todomvh{any reference/evidence in support of this statement?}
 Hence our analysis centers around the free \ac{vsa} condition.
@@
 increases task load.
-This in turn reduced overall performance, even though the participant might
+This can in turn reduce overall performance, even though the participant might
 otherwise also have performed covert \ac{vsa}, but would not be occupied by
 maintaining strict central gaze fixation.
-These extended task demands are not present in the free \ac{vsa} condition, so
+These extra task demand is not present in the free \ac{vsa} condition, so
 there is less performance to be gained here.
@@
 drastically boost performance.
-Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with
-This is also supported by our results on cross contidion calibration
+Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with his
+overt \ac{vsa} performance, although eye tracking showed that he relied mostly
+on over \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
+gaze was uncued.
+This is also supported by our results on cross condition calibration\todo{how}
 
@@
 
-Another limiting factor are the difficulties in correctly interpreting eye tracker
-results in studies with eye motor impaired patients.
+Another limiting factor are the difficulties in correctly interpreting eye
+tracker results in studies with eye motor impaired patients.
 If eye tracking at all is possible, it is not guaranteed that the user is able
-to perform the calibration procedure succesfully.
+to perform the calibration procedure successfully.
 Further experiments should be carried out with a stationary eye tracker with
@@
 This is not practical when working with
-wheelchair-bound individuals who might have undergone a tracheostomy.
-hard because of inaccuracy eye tracker for eye
-motor impaired patients. Redo with more advanced stationary eye tracker and
-head rest.
+wheelchair-bound individuals who might have undergone a tracheostomy and might
+suffer from spasticity.
 
@@
     but this does not hold true if central gaze fixation is not imposed.
-  \item We have shown that covert \ac{vsa} \todo{with or
-    without imposed gaze fixation} decoding can alternatively be improved by
+  \item We have shown that covert \ac{vsa}
+    decoding can alternatively be improved by
     using a calibration session with attempted overt \ac{vsa}.
-\end{enumerate*}
+\end{enumerate*} \todo{with or without imposed gaze fixation}
 

### 2024-10-02 chapters/patients_table.tex
Comments Marc

\begin{tabular}{llllllllr}
  \toprule
  \textbf{ID}  & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
  \textbf{Diagnosis}
  & \textbf{Speech}     & \textbf{Trach.} & \textbf{Communication}          &
  \textbf{Cls.} \\ \midrule
  PA1 & 58  & M   & L     & bulbar-onset \acs{als} & anarthric  & no          & tablet                 & 3  \\
  PB1 & 41  & M   & L     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PB2 & 43  & F   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PB4 & 48  & M   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
  PC2 & 43  & M   & R     & brainstem stroke & anarthric  & yes         & \makecell[l]{prompting\\+eye movement} & 2 \\
  PC3 & 43  & F   & R     & brainstem stroke & anarthric  & yes         & letterboard            & 2 \\
  PC4 & 54  & M   & R     & \makecell[l]{left cerebellar stroke \\ (trombosis of the basilar artery)} & anarthric  & yes & letterboard & 2 \\
  \bottomrule
\end{tabular}

### 2024-10-02 chapters/stbf_struct.tex
Comments Marc

@@
   published by~\textcite{VanDenKerchove2022}}.
-
-	\todo{Rename to LDA}
-	\todo{Work Kronecker sum in}
-	\todo{Add results over time vs toeplitzlda}
-	\todo{Introductory part on LCMV beamforming to intro}
 
@@
 		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
-		\caption{Different estimators of the covariance and inverse covariance
+    \caption[Estimated covariance and inverse covariance.]{Different estimators of the covariance and inverse covariance
 			of 100 epochs of data from \textit{Subject 01} for channels
@@
 	\end{figure}
-	\todo{visualize kronecker sum}
-	\todo{visualize kronpca decomp pvl}
-	\todo{method pvl}
 
@@
 		\includegraphics[width=\linewidth]{figures/stbf_struct/fpi.eps}
-		\caption{Average cross-validated \textsc{stbf-struct} Accuracy using
+    \caption[Average cross-validated \textsc{stbf-struct} accuracy]{%
+      Average cross-validated \textsc{stbf-struct} accuracy using
 			one trial per block over all 21 subjects
@@
 		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
-		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
+    \caption[Clasifier accuracy in function of available training data.]{%
+      Accuracy of the different classifiers for all 21 subjects relative to the
 			number of blocks available for training. One block consists of 135
@@
 		\input{chapters/stbf_struct_tests_1.tex}
-		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
+    \caption[Statistical significance of differences in classifier performance (1 repetition).]{%
+      Statistical significance of differences in classifier performance.
+      $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
 			Holm correction using one testing trial for different classifiers
@@
 		\input{chapters/stbf_struct_tests_2.tex}
-		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
+		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
+      $p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
 		\label{tab:p_values_2}
@@
 		\input{chapters/stbf_struct_tests_3.tex}
-		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
+    \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
+      $p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
 		\label{tab:p_values_5}
@@
 		\includegraphics[width=\linewidth]{figures/stbf_struct/training_time.eps}
-		\caption{Median fold training time for different classifiers at different spatial
+    \caption[Classifier training time.]{Median fold training time for different classifiers at different spatial
 			resolution levels evaluated over all training folds for all subjects.
@@
 
-	\todo{Remove mentions of interpretability}
-
   \subsection{Extension to \ac{lda}}
@@
 
-  \todo{Include part kronlda, check notes in journal}
 \todo{
-LCMV struct might be to restrictive
-Toeplitzlda
-MEG paper}
+in conclusion: LCMV struct might be to restrictive
+Toeplitzlda MEG paper}
+
+\todo{in conclusion: extension to the TFR domain}
 %\begin{equation}
@@
 %\end{equation}
-\todo{pvl, kronpca}
-
-\todo{check chapters/structure paper}
-\todo{check structure Leuven.ai poster}
-\todo{extension to the TFR domain}

### 2024-10-02 chapters/wcble.tex
Started patients paper
Comments hakim
Comments Marc

@@
 \label{sec:wcble/intro}
+\todohsm{do'nt repeat in introduction, rather make line clear}
 
@@
 
-\subsection{\Ac{erp} latency jitter}
+\subsection{\Acs{erp} latency jitter}
 
@@
   \input{figures/wcble/smearing.tikz.tex}
-  \caption{The smearing effect. When obtaining an \ac{erp} template through
+  \caption[The smearing effect.]{The smearing effect. When obtaining an \ac{erp} template through
   averaging, high jitter causes a decrease in peak amplitude and a deformation
@@
 }{%
-  Schematic representation of the Woody Classifier Based Latency
-  Estimation training phase.
+  Schematic representation of the \ac{wcble} training phase.
+}{%
+  Schematic representation of the \ac{wcble} training phase.
   (1) The first-stage spatiotemporal binary classifier is trained on a set of
@@
   \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
+
+}{%
+  Schematic representation of the \ac{wcble} test phase.
 }{%
@@
 
-	\caption{Classifier-Based Latency Estimation}
+  \caption{\Ac{cble}}
 	\label{alg:cble}
@@
   \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-	\caption{Classifier-Based Latency Estimation with Woody Iterations}
+  \caption{\ac{wcble}}
 	\label{alg:wcble}
@@
 }{%
+  Simulated target and non-target evoked data.
+}{%
   Simulated target and non-target evoked data. Targets contain a sine wave
@@
 \todo{Properly align plots and label axes}
+\todo{decrease tick size}
 
@@
   \input{figures/wcble/fig_accuracy.pgf}
-  \caption{align results}
+  \caption{caption}
   \label{fig:wcble/results/latency}
@@
 \todo{properly label axes}
-\todo{discuss results}
+\todo{uniform fonts with rest of document}
+\todo{correct width}
+\todo{figure for data generation}
 

### 2024-10-02 figures/wcble/smearing.tikz.tex
Comments Marc

\begin{tikzpicture}
    % First subplot (low jitter) positioned relative to center
    \begin{axis}[
        at={(-3cm,0cm)}, % Adjust position to the left, closer to the center
        anchor=center,
        width=7cm, height=6cm,
        xmin=20, xmax=80,
        ymin=-0.1, ymax=1.1,
        axis lines=none, % Remove axes
        title={Low Jitter}, % Add title
        title style={yshift=-10pt}, % Adjust title position to make it closer to the plot
        ]
        % Plot individual waveforms (low jitter) in darkgray
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-50)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-52)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-48)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-51)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-49)/5)^2)};

        % Plot average waveform (low jitter) in accent1 color, very thick
        \addplot[ultra thick,accent1,domain=20:80,samples=100] {exp(-0.5*((x-50)/5)^2)};
    \end{axis}

    % Second subplot (high jitter) positioned relative to center
    \begin{axis}[
        at={(3cm,0cm)}, % Adjust position to the right, closer to the center
        anchor=center,
        width=7cm, height=6cm,
        xmin=20, xmax=80,
        ymin=-0.1, ymax=1.1,
        axis lines=none, % Remove axes
        title={High Jitter}, % Add title
        title style={yshift=-10pt}, % Adjust title position to make it closer to the plot
        ]
        % Plot individual waveforms (high jitter) in darkgray
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-45)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-55)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-40)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-60)/5)^2)};
        \addplot[darkgray,domain=20:80,samples=100] {exp(-0.5*((x-50)/5)^2)};

        % Plot average waveform (high jitter) in accent1 color, very thick
        \addplot[ultra thick,accent1,domain=20:80,samples=100] {0.2*(exp(-0.5*((x-45)/5)^2) + exp(-0.5*((x-55)/5)^2) + exp(-0.5*((x-40)/5)^2) + exp(-0.5*((x-60)/5)^2) + exp(-0.5*((x-50)/5)^2))};
    \end{axis}

    % Unified legend positioned further below the center
    \node at (0,-5) {\begin{tikzpicture}
        \begin{axis}[
            hide axis,
            xmin=0, xmax=0, ymin=0, ymax=0,
            legend style={
	    	draw=none, legend columns=2, /tikz/every even column/.append
		style={column sep=2cm}
		}
            ]
            \addlegendimage{darkgray,thick}
            \addlegendentry{Waveforms}
            \addlegendimage{accent1,ultra thick}
            \addlegendentry{Average}
        \end{axis}
    \end{tikzpicture}};
\end{tikzpicture}
\vspace{-5cm}

### 2024-10-03 chapters/bci.tex
Resturcture chap 1 and 2

@@
-\todo{comments liesel paradigms})
+\todo{comments liesel paradigms}
 \todo{spellcheck}
@@
 \todo{acronyms written in full every new chapter and in captions}
+\the\textwidth
 
@@
 Additionally, they bring a new paradigm to human-computer interaction,
-especially when paired with virtual or augmented reality.\todo{cite}
+especially when paired with virtual or augmented reality~\cite{SiMohammed2017}.
 Ultimatly, they have always been especially promising as assistive technology for
@@
 \section{A direct interface to the brain}
-\todohsm{Start with state of the art before theses.}
-\todohsm{objectives at the end of chapter 1}
-\todohsm{Introduction part: chapter 1 rationale, chapter 2: literature/sota}
 \todohsm{make less verbose}
@@
 
-\fullpagefigct{%
-  \includegraphics[width=\textwidth]{figures/bci/bci_loop.pdf}
+\fullpagefig[topleft]{%
+  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
 }{%
@@
 what is going on in the brain.
-Section~\ref{sec:bci-recording} gives an overview of the available options.
-One might for instance measure the fluctuations in bloodflow to establish the
-activity of a specific brain region
-However, this signal reacts too slowly to reflect the real-time activity that is
-of interest for our purposes, and it carries little information other than the
-brain regions where it is originating.
-A better candidate is the neuronal electrical activity.
-The brain contains 86 billion neurons, which are highly interconnected cells that are
-the smallest units of information processing.
-The \emph{action potential} and its related \emph{postsynaptic potential}
-is an electrical pulse occurring when a neuron receives
-input from another neuron and is activated.
-The firing of a neuron, or the combined firing of groups of
-synchronized neurons, thus generate an electrical field in and near the brain,
-which can be measured using electrodes.
-Fluctuations in the brain's electrical activity are be faster and more
-informative.
+Multiple signals can be recorded from the brain, including bloodflow,
+magnetical electrical activity.
+For several reasons laid out in~\ref{sec:bci-recording}, the electrical modality
+is especially interesting.
 
-The recorded brain waves form only one part of the interaction scheme presented
+The recorded brain activity forms only one part of the interaction scheme presented
 in Figure~\ref{fig:bci/loop}.
@@
 activity to try to extract the desired output.
-It would help greatly if we knew what we are looking for.
-A \ac{bci} operates according to a specified paradigm, which defines the method by which interaction is conducted.
+Instead, it would help greatly if we knew which specific activity to look for.
+A \ac{bci} operates according to a specified \emph{paradigm}, which defines the
+method by which interaction is conducted.
 This usually means that the user is instructed to perform a specific task, like
@@
 We can now exploit background knowledge from prior neuroscientific research
-that the brain response evoked by attended stimuli is different of those
-unattended,
-and determine which one of a set of targets was attended.
-In turn, this can be coupled to an action, like typing the inteded letter A or B
-in Figure~\ref{fig:bci/loop}.
+to couple specific activity in the brain to specific information or actions we
+want to convey via the \ac{bci}.
+The brain singal now \emph{modulates} this information.
+In turn, the detection of this signal can be coupled to an action, like
+typing the inteded letter A or B in Figure~\ref{fig:bci/loop}.
 The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
@@
 activity.
-This is commonly referred to as \emph{feature extraction}.
-In other words, the signal needs to be interpreted \emph{in function of} the
-task we want our \ac{bci} to perform.
 The electrical activity of the brain is relatively weak compared to that of the
@@
 Some obvious sources of interference can be easily filtered out by
-\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}), but the problem of retrieving or \emph{decoding} only the
-relevant signal is generally hard.
+\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}) and by choosing
+the correct representation of the signal (i.e. \emph{feature extraction}, but the problem
+of retrieving or \emph{decoding} only the relevant signal is generally hard.
 Nevertheless, we can attempt to solve it through supervised machine learning.
 Useful information that is encoded in the brain signal by the paradigm,
-can be retrieved from it by the decoder.
+can be retrieved from it by the decoder and interpreted in function of the
+desired \ac{bci} output.
 Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
@@
 rehabilitation applications.
-\todo{is every section referred to?}
 
-\todo{wrap-up}
-\todo{make shorter}
+To summarize, a \ac{bci} can facilitate direct interaction central nervous system.
+To do this, information is modulated in the brain signal according to the
+\ac{bci} paradigm by stimulating, the user or having them perform a task.
+Their brain signal is subsequently recorded and preprocessed to clean it.
+Relevant features that contain aspects of the modulated signal are extracted,
+and the modulate information is retrieved by the decoder, a machine learning
+model.
+This information can then acted on and the loop is closed if this action
+affects the user.
+Let us take a look at each of these elements separately in the next section.
 
@@
 \label{sec:bci-recording}
-\todo{why is spatial resolution a good thing}
-\todo{explain with each method what it actually is (e.g.\ electrodes placed
-at...)}
-\todo{make sure every part ends with a short summary motivation for why we chose our
-approach}
 
@@
 (fNIRS)~\cite{Borgheai2020}.
-The technologies above all measure the hemodynamic signal, which only indirectly
+
+The technologies above all measure bloodflow (the hemodynamic signal), which only indirectly
 reflects brain activity.
+This signal reacts too slowly to reflect the real-time activity that is
+of interest in \ac{bci} for communication, and it carries little information
+other than the brain region from where it is originating.
+
+A better candidate is the neuronal electrical activity.
+The brain contains 86 billion neurons, which are highly interconnected cells that are
+the smallest units of information processing.
+The \emph{action potential} and its related \emph{postsynaptic potential}
+is an electrical pulse occurring when a neuron receives
+input from another neuron and is activated.
+The firing of a neuron, or the combined firing of groups of
+synchronized neurons, thus generate an electrical field in and near the brain,
+which can be measured using electrodes.
+The related discipline is \emph{electrophysiology}.
+
 Because of the desirable high temporal resolution~\cite{Easttom2021} and
@@
 patient.
-As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
+As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often
+measure a very specific brain region and can result in better signal quality, at the cost of the risks
 involved with brain surgery.
 
-\fullpagefig{%
-  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
+\fullpagefig[topleft]{%
+  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
 }{%
@@
 
-Specifically, invasive methods come with an increased \emph{spatial resolution},
+Specifically, invasive methods are less noisy and come with an increased \emph{spatial resolution},
 meaning they can extract a signal from a specific brain region or even a set of
 neurons of interest.
-Generally, penetrating microelectrodes are considered to have the highest
+This makes it easier to retrieve a signal with known origin from the brain,
+which might be leveraged by the paradigm to modulate information.
+Generally, microelectrodes penetrating the cortex are considered to have the highest
 spatial resolution.
+These are needles with a diameter smaller than 10µm and one or multiple
+electrodes that extend several millimmeters into the cortex.
 Microelectrodes can measure the Local Field Potential (LFP), the extracellular
@@
 Parkinson's disease symptoms~\cite{Arlotti2018}.
+Depth electrodes are longer rods with multiple electrodes than can extend
+deeper into the brain.
 Since these still penetrate the cortex, electrodes that sit on top of the
 cortex are more popular, as they are considered less invasive.
-\Ac{ecog} is a powerful method
-with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
+\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
 Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
@@
 technology is preferred over implanted electrodes, unless the added value of invasive BCI is
-sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
+sufficiently large~\cite{Kageyama2020}.
 This provides justification for focusing part of the current research effort on
@@
 To counteract the noise present in our recording, we must on the one hand evoke
-strong, informative brain signals using a suited \ac{bci} paradigm, and on the
+strong, informative brain activity using a suited \ac{bci} paradigm, and on the
 other hand pick a suited decoder which can isolate the signal of interest and
@@
 
-\section{Preprocessing}
+\section{Preprocessing}%
 \label{sec:bci/preprocessing}
@@
 \subsection{The \acf{erp}}
+\label{sec:bci/oddball/erp}
 
@@
 
-\subsection{Preprocessing}
+\subsection{Preprocessing \& feature extraction}
 In addition to (ref) epoching
@@
 Linear decoding techniques
+
+\todo{wrap-up}
+
+\section{The gaze-dependence problem}
+
+In a nutshell, we can state that \acp{bci} decode brain activity with the aim to
+establish a direct communication pathway bypassing speech and other forms of
+muscular activity~\cite{Naci2012,Chaudhary2016}.
+They have raised great hopes for individuals devoid of these abilities, for whom
+\acp{bci}s can provide a means to communicate or to
+control devices.
+
+Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based
+oddball interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
+Targets are shown in short flashes on a computer screen, that evoke
+and \ac{erp} when observed, which can be detected in the \ac{eeg}-signal
+detected in the EEG-signal.
+The \ac{erp} consists of multiple components, which some are modulated by
+perception and the attention of the user.
+Decoding this modulation allows for information transfer controlled by the
+user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
+
+However, when carefully examining the properties of the \ac{erp} components,
+we notice that while attention plays a role in the modulation of specific components,
+some part is also related to visual perception.
+How a stimulus is perceived, and, as a consequence, which \ac{erp} components
+are modulated, is directly related to whether the user gazes directly at it
+(i.e.\ \emph{fixates} their gaze on the target) or not.
+Indeed, research has shown that visual oddball \ac{bci}s cannot operate
+efficiently when the user does not direct
+their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
+
+This leads to a problem that is not often addressed:
+a visual \acp{bci} is dependent on the gaze of the user, and hence on
+muscle control.
+This is exactly what \acp{bci} for augmented and assistive technology try to
+avoid.
+The \ac{bci} target population consists of individuals who suffer from various
+degrees of paralysis, or even from \ac{lis}, the complete loss of muscle
+control with preserved consciousness.
+While \ac{bci}s are most attractive as a solution for individuals with
+\ac{sspgi}, studies
+paradoxically often report bad performance in patient populations precisely due
+to the lack of adequate eye motor control.
+Equally often, participants with gaze impairments were excluded from the study.
+
+Different degrees of eye motor impairment can render visual \ac{bci}s
+uncomfortable or outright impossible to use.
+When operating a visual \ac{bci}, the usual rapid series of forced saccades
+followed by fixation is tiring over time, even for healthy control subjects.
+A suitable alternative would allow the user to keep their eyes in an at-rest
+position of their choice while operating the \ac{bci}.
+This points to the need for gaze-independent solutions, which we will explore
+further in Chapter~\ref{sec:gaze-independence} and the rest of this thesis.

### 2024-10-03 chapters/bttda.tex
Resturcture chap 1 and 2

@@
 These latent training tensors are then reshaped (\emph{vectorized}) into feature vectors
-$\vec{g} =  \vect(\ten{G})$ and used to train a decision classifier with the corresponding class labels.
+$\mat{g} =  \vect(\ten{G})$ and used to train a decision classifier with the corresponding class labels.
 At the evaluation stage, the projections learned from the training dataset are
@@
 \begin{equation}
-	\vec{g}
+	\mat{g}
 	=\left[\vect\left(\ten{G}^{(1)}\right)\

### 2024-10-03 chapters/gaze_independence.tex
Resturcture chap 1 and 2

@@
 
-\emph{Parts of Sections~\ref{sec:gaze-dependence} and~\ref{sec:gaze-independence}
-have been published by \textcite{VanDenKerchove2024}.}
-
 \todo{spatially vs temporally organized interfaces}
@@
 
-\section{The gaze-dependence problem}%
+\section{Eye motor impairment in individuals with \acs{sspi}}%
 \label{sec:gaze-dependence}
@@
 het om deze patiënten gaat}
+
 One of the goals of brain-computer interfacing is establishing a communication
-channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
-which in turn can provide solutions to paralyzed individuals.
-In the strictest interpretation, this also means that an interface should not
+channel that does not rely on speech or muscular activity which in turn can
+provide solutions to individuals with \ac{sspi}.
+In the strictest interpretation, this then means that an interface should not
 rely on the control of eye muscles used to redirect the gaze or for blinking.
-\todo{patients suffering paralysis due to damage to the CNS (stroke, spinal
-chord injury, tbi) or neurodegenerative diseases}
+It is exactly this potential of \acp{bci} that makes them suitable as assistive
+communication devices.
+\todo{to the CNS (stroke, spinal chord injury, tbi) or neurodegenerative diseases}
 
-In fact, many patients in the \ac{bci} target population suffer to some extent
-from eye motor disability, requiring \ac{bci}s adapted to their condition.
+Unfortunately, many individuals in the \ac{bci} target population suffer to
+some extent from eye motor disability, requiring \ac{bci}s adapted to their condition.
 Table~\ref{tab:incidence} reports the relatively high frequencies of
@@
 movements or minor motor output.
-While some definitions only include patients who suffered stroke or traumatic
-brain injury to specific regions in the brain (midbrain, brainstem, or
+While some definitions only include stroke or \ac{tbi} patients with damage to
+specific regions in the brain (midbrain, brainstem, or
 cerebellum)~\cite{Smith2005},
@@
 
-Together, some of these patients constitute the \ac{lis} patient group, which
-forms one of the main \ac{bci} target populations.
+The most severely impaired of these patients toghether constitute the \ac{lis}
+patient group, which forms one of the main \ac{bci} target populations.
 In this work, we define \ac{lis} patients as those in a situation of (near)
-complete paralysis and difficulties or the inability to communicate without
-assistive technology.
+complete paralysis and difficulties or the inability to communicate independently,
+even with assistive technology.
 This corresponds to classes one and two defined by
@@
 necessary to qualify as locked-in\fnlis.
-Since this work is concerned with gaze-impaired patients, we will adopt this
-interpretation into our definition.
 Case study reports~\cite{Patterson1986} show that the population of locked-in
@@
 Hence, it would be interesting to focus \ac{bci} development efforts on the
-larger population of \ac{lis} patients with severe eye motor impairment that
-currently slip through the cracks of the assistive technology offer.
-These are the patients whose severe eye motor impairment prevents them from
+larger population of individuals with \ac{sspgi} that currently slip through
+the cracks of the assistive technology offer.
+These are individuals whose severe eye motor impairment prevents them from
 using eye tracking based solutions.
@@
 acceptable quality of life.
-\todo{order of concepts in next two paragraphs can be tweaked}
 
-This leads us to what we call the \emph{gaze-dependence problem in visual
-\ac{bci}}:
-Traditional visual \ac{bci} scenarios require the user to overtly direct their both
+\begin{figure}
+  \input{figures/gaze_independence/lis_venn.tikz.tex}
+  \caption[Venn diagram of potential users of \ac{bci} assistive technology.]{%
+    Venn diagram of potential users of assistive technology.
+    Patients with \acf{sspi} are the general target population.
+    If eye gaze-based solutions are not properly suited for an individual,
+    they have \acf{sspgi}.
+    Individuals with \ac{lis} have no muscular control left, or is is very limited,
+    such as the ability communicate through a binary switching system.
+    \Acp{bci} are especially relevant for the last two groups.
+  }
+  \caption{fig:gaze_independence/venn}
+\end{figure}
+
+This brings us back to the gaze-dependence problem in visual \ac{bci}:
+Traditional visual \ac{bci} scenarios require the user to \emph{overtly} direct both their
 \ac{vsa} and gaze toward the screen target they intend to select.
-For our purposes, screen targets are overlaid with non-overlapping, transient
-stimuli that evoke \acp{erp} in the \ac{eeg}.
-The selected target can be decoded from these \acp{erp}, as is the case for the
-oddball paradigm where observing a rare but attended stimulus evokes a P3 \ac{erp}
-component.
-However, a critical challenge arises when users rely solely or in part on covert
-VSA, which involves directing visuospatial attention without corresponding eye gaze.
+However, a critical challenge arises when users rely solely or in part on
+\emph{covert} \ac{vsa}, which involves directing \ac{vsa} without corresponding
+eye gaze.
 In these cases, classical solutions often fall short of the widely accepted
-80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,RonAngevin2019} deemed necessary for a comfortable user experience~\cite{Neeling2019}, calling for alternative, gaze-independent
+80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,RonAngevin2019}
+deemed necessary for a comfortable user experience~\cite{Neeling2019}, calling for alternative, gaze-independent
 solutions.
-
-To summarize: for patients with eye motor impairment, gazing directly at a screen target may
-be uncomfortable, impractical, or even impossible.
-Hence, assistive devices that rely on eye tracking are often inefficient for
-them.
-Consequently, while \ac{bci}s hold great promise for these individuals, conventional
-gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
-control.
-Therefore, the development of decoding strategies that account for covert
-VSA becomes crucial in the pursuit of high-performance gaze-independent
-\ac{bci}s.
 
@@
 visuospatial attention and their gaze do not necessarily coincide.
+There are multiple approaches to implement gaze-independence in a \ac{bci}.
+The next section highlights some noteworhy examples.
 
-\section{Gaze-independent \ac{bci} paradigms}
+\section{State-of-the-art}
 \label{sec:gaze-independence}
-\todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}
+\todohsm{Address more systematically.}
 \todo{in a paragraph describe each approach and its up and downsides}
@@
 
-\section{Problem statement}
+\section{Research objectives}
 
-\subsection{Objectives}
+To summarize: for patients with eye motor impairment, gazing directly at a screen target may
+be uncomfortable, impractical, or even impossible.
+Hence, assistive devices that rely on eye tracking are often inefficient for
+them.
+Consequently, while \ac{bci}s hold great promise for these individuals, conventional
+gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
+control.
+Therefore, the development of decoding strategies that account for covert
+VSA becomes crucial in the pursuit of high-performance gaze-independent
+\ac{bci}s.
 
-\todohsm{ move next two paragraphs to introduction, more context than objectives}
-Brain-Computer Interfaces (\ac{bci}s) decode brain activity with the aim to establish a direct communication
-pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
-devoid of these abilities, for whom \ac{bci}s can provide a means to communicate or to
-control devices.
-An effective and proven method is the \ac{eeg}-based visual event-related potential (\ac{erp})-based
-oddball interface~\cite{Wolpaw2018,Severens2020}.
-Here, targets are shown in short flashes on a computer screen.
-Each time the user observes a target, an \ac{erp} is evoked, the presence of which can be
-detected in the EEG-signal.
-The \ac{erp} consists of multiple components, of which some are modulated by the
-attention of the user, specifically the N2 and the P3 component.
-Decoding this modulation allows for information transfer controlled by the
-user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
-However, visual \ac{bci}s cannot operate efficiently when the user does not direct
-their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
-\todomvh{I am wondering whether goal statements and patient target groups
-should be addressed in chapter 1.}
-
-This leads to the often underadressed gaze-dependence paradox: the \ac{bci} target
-population consists of patients who suffer from various degrees
-of paralysis, or even from Locked-in Syndrome (LiS), the complete loss of muscle
-control with preserved consciousness.
-While \ac{bci}s are most attractive as a solution for these patients
-compared to other assistive technologies like eye-tracking\todomvh{but with eye
-control while the previous sentence points are absence of it}., studies often
-report bad performance in patient populations precisely due to the lack of adequate eye motor
-control, or patients with gaze impairments are excluded.
-Different degrees of eye motor impairment can render visual \ac{bci}s
-uncomfortable or outright impossible to use.
-When operating a visual \ac{bci}, the usual rapid series of forced saccades followed
-by fixation is tiring over time, even for healthy control subjects.
-A suitable alternative would allow the user to keep their eyes in an at-rest
-position of their choice while operating the \ac{bci}.
-This points to the need for gaze-independent \ac{bci}s.
-
-The general goal of this PhD is to tackle the gaze-dependence in visual \ac{bci}s.
+The general goal of this PhD is to tackle the gaze-dependence in problem in the
+visual oddball \ac{bci}.
 \todo{refer to ITRs from previous chapter}
@@
 \begin{figure}
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}[t]{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
@@
   \end{subfigure}\hfill%
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}[t]{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
@@
 
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}[t]{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
@@
   \end{subfigure}\hfill%
-   \begin{subfigure}{.45\textwidth}
+   \begin{subfigure}[t]{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}

### 2024-10-03 figures/gaze_independence/lis_venn.tikz.tex
Resturcture chap 1 and 2

\sffamily
\begin{tikzpicture}

    % Draw the outer lightgray box with ultra thick darkgray line
    \draw[fill=lightgray, draw=darkgray, ultra thick] (-4,-1.5) rectangle (4,5.5);
    % Label for the outer box in the top left corner
    \node[anchor=north west, text=darkgray] at (-4,5.5) {\textbf{\acs{bci} users}};

    % Draw the outermost circle for SSPI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent1] (0,2) circle (3);
    \node[text=accent1] at (0,4) {\textbf{\acs{sspi}}};  % Centered bold text with line color

    % Draw the middle circle for SSPGI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent2] (0,1) circle (2);
    \node[text=accent2] at (0,1.66) {\textbf{\acs{sspgi}}};  % Centered bold text with line color

    % Draw the innermost circle for LIS with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent3] (0,0) circle (1);
    \node[text=accent3] at (0,0) {\textbf{\acs{lis}}};  % Centered bold text with line color

\end{tikzpicture}

### 2024-10-04 chapters/bci.tex
Rewrite sota

@@
 \todo{acronyms written in full every new chapter and in captions}
-\the\textwidth
+\todo{search for ??}
 
-People with severe physical impairment due to
+People with \ac{sspi} due to
 neurodegenenerative disease, stroke, traumatic brain or spine injury
@@
 lab conditions.
-
 As a consequence, the \ac{bci} field relies heavily on the
 electrical activity of the brain.
+
+\subsection{Invasive electrophysiology}
 The \emph{invasiveness} of the technology forms an
@@
 
+\subsection{\Acl{eeg}}
+
 \Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
@@
 
-The major drawback of \ac{eeg} is its poor spatial resolution.
-The electrodes are large and sit on the outside of the scalp, relatively far
-away from the cortex.
+The major drawback of \ac{eeg} is its poor spatial resolution~\cite{Ferree2001}.
+Since they sit on the outside of the scalp electrodes are large and distance
+relative to the neurons in the cortex they ought to measure.
 Furthermore, the layers of skull and cerebrospinal fluid in between can
-attenuate the signal and cause volume conduction, which results in electrodes
+attenuate the signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
 measuring a mixture of activity from sources elsewhere in the brain.
@@
 interest, \ac{eeg} also records other artifacts, from the environment, the power
-supply, nearby electrical equipment, or the user's muscle movements.
+supply, nearby electrical equipment, or the user's muscle
+movements~\cite{Urigueen2015}.
 Combined, this means \ac{eeg} is inherently noisy.
-\todo{citations from 'Decoding covert speech from eeg-a cromprehensive review
-EEG has lower signal-to-noise ratio (SNR) than the other
-modalities. It is almost always corrupted by artifacts such as
-muscular artifacts (Eberle et al., 2012; Liu, 2019).
-• EEG has limited spectral and spatial resolution (Peled et al.,
-2001; Lakshmi et al., 2014).
-• Recording EEG for longer duration is challenging since the
-conductive gel or the saline solution applied for reducing the
-electrode impedance dries up over time, thus increasing the
-electrode impedance (Guger et al., 2012; Xu et al., 2018a).
-• A trained personnel is required for placing the EEG
-electrode cap}
 
@@
 \label{sec:bci/paradigms}
-\todohsm{bring more structure, more subtitles}
+
+\subsection{Active \& passive \ac{bci}}
 
@@
 }
-\todo{check permissions}
-\todo{above Zander and Kothe was mentioned.}
 
@@
 From the user's point of view, this concept is probably the most attractive,
-since their intentions or cognitive state state would be inferred from just
-their brain activity.\todohsm{in "passive bci", the user is not supposed to express "intentions". ragher start with active "}
+since their cognitive state state would be inferred from just
+their brain activity.
 However, without active user participation, it can be hard to establish reliable
@@
 Therefore, passive \ac{bci} are currently more suited to tasks like
-emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
+emotion or affect monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload
 monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
-detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
-Furthermore, active paradigms\todo{"active" is used in a fuzzy way: active BCI is self-paced, reactive is in response to a cued event. Further down that becomes more clear.
-The "active participation": is needed in both cases as shown in Fig. 1.3.
-I suggest to say active participation} make it easier to quickly collect training data for
-supervized machine learning since the conditions can be more easily controlled
+detection~\cite{SiMohammed2020}.
+It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
+Nevertheless, active participation facilitates training data collection for
+supervised machine learning since the conditions can be more easily controlled
 by the user or the stimulation.
 
-Paradigms with high active participation can be split up in \emph{active
+Paradigms requiring high active participation can be split up in \emph{active
 \ac{bci}} and \emph{reactive \ac{bci}}.
@@
 complexity of the performed tasks.
+They might require extensive user training and are sensitive to
+correct task instruction.
 Imagined speech or movements can be performed in a multitude of ways, which
@@
 variability in brain anatomy and cognitive strategies. As a result, what has
-been labeled as "illiteracy" may, in fact, be a reflection of systems that are
+been labeled as `illiteracy' may, in fact, be a reflection of systems that are
 not flexible enough to adapt to different users. This perspective
@@
 challenges, thus standing in the way of progress.
-Moreover, the term ``illiteracy'' itself carries negative
+Moreover, the term `illiteracy' itself carries negative
 connotations, implying a user deficiency that reinforces normative assumptions,
@@
 }.
-\todo{user training more than with reactive, task instruction}
 
+\subsection{Reactive \ac{bci}}
 Reactive \ac{bci} take another approach, by providing a discrete set of
 sensory stimulations towards which the user's attention can directed.
-Compared to active \ac{bci}, reactive paradigms usually require more effort\todohsm{is this the case?}
-and can induce fatigue due to the constant stimulation.
+Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
+due to the constant stimulation.
 These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
@@
 In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
-be correlated with the brain activity to retrieve the attended stimulus.
+be correlated with the brain activity to retrieve the attended
+stimulus~\cite{Sun2022}.
 Instead of stimulating all possible selections at once, the oddball paradigm
-stimulates them one by one with single flashing.\todo{reference}
+stimulates them one by one with single flashes~\cite{Pan2022}.
 Since the times of stimulation of each target is known, and time-points at
@@
 \label{sec:bci/preprocessing}
-\todo{preprocessing}
 EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
@@
 Additionally, notch filtering can be  used to remove strong power line
-interference\footnote{Typically at 50Hz in Europe}.
+interference\footnote{Typically at 50Hz in Europe.}.
 This interference can heavily contaminate EEG recordings and even leak through
@@
 \label{sec:bci/decoding}
+
+\subsection{Design principles}
 
@@
 Hence, such problems often present themselves as classification tasks.
+
+\subsection{Implemetation \& evaluation}
 
@@
 
+\begin{figure}[t]
+  \begin{subfigure}[t]{.45\textwidth}
+  \includegraphics[width=\textwidth, height=\textwidth]{example-image-a}
+  \end{subfigure}\hfill%
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/bci/p300.tikz.tex}
+    \caption{The P300 matrix speller interface. Entire rows or columns flash at once.}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/bci/erp.tikz.tex}
+    \caption{The idealized \ac{erp} evoked by an attended stimulation.}
+  \end{subfigure}\hfill%
+  \begin{subfigure}[t]{.45\textwidth}
+    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
+  \end{subfigure}
+\end{figure}
+
 \todo{four pane figure:
   photo of someone using a P300 bci
-  P300 matrix interface
-  simplified figure of ERP components
   full spatiotemporal ERP response
@@
 
-\subsection{The \acf{erp}}
+\subsection{The \acf{erp}}%
 \label{sec:bci/oddball/erp}
@@
 The component most prominently modulated by attention is called the P3
-(positive deflection after $\pm300$ ms).\todo{add footnote about how ms naming
-is maybe not suited, especially since this work focuses on the fact that there
-is a lot of variability in the timings for visual potentials. the short naming
-scheme builds on 'the first negative component', 'the first positive
-component', and for the visual oddball erp this ordening more or less coincides
-with the timings.}
-Therefore, oddball paradigms are frequently referred to as P300 paradigms.
+(positive deflection after $\pm300$ ms)
+some \ac{bci}\footnote{In \ac{erp} analysis,
+components are sometimes referred to with the timing around which they occur
+(e.g.\ N170, negative after 170ms), or by their rank in order of occurrence
+(N1, the first negative component).
+The timing nomenclature is based on average latencies for neurotypical
+individuals, but is seldom correct for specific \ac{bci} user due to the large
+variability across \ac{erp} stimulation paradigms and subjects and can also
+depend on the user's pathology.
+Furthermore, as we will see later, this work focuses specifically on the
+intra-session variability in \ac{erp} latency, explicitly assuming the latencies are not
+fixed.
+Therefore, we adhere to the ranking nomenclature.
+Fortunately, for the visual oddball paradigm, the rankings do roughly
+correspond to their expected timings, i.e.\ P1=P100, N1=N100, etc.}.
+Therefore, oddball paradigms are frequently referred to as P3(00) paradigms.
 However, due to the equally important contributions of other \ac{erp}
@@
 \subsection{Preprocessing \& feature extraction}
-In addition to (ref) epoching
-baseline correction, but not in decoding (cite)
+For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz
+and 16Hz.
+Re-referencing can be done to average mastoids to highlight the contribution of
+the P3 component.
+In addition to the steps outlined in section~\ref{sec:bci/preprocessing},
+the signal is usually cut into \emph{epochs} for \ac{erp} analysis.
+These are time windows surrounding the stimulus event.
+They usually include some baseline interval before the stimulus, which can be
+used to normalize the response.
 
-\subsection{Decoding ERPs}
+The epochs contain values representing the voltage of a specific channel relative to the
+reference at a specific time relative to the stimulus.
+In \ac{erp} classification, they are usually directly used as features, by
+concatenating all channels in the matrix format epoch $X$ into one feature
+vector s.t.\ $\mat{x} = \mat\vec(X)$.
 
-Usually, snr low and multiple trials necessary
+\subsection{Decoders}
+Several techniques are suited to decode~\ac{erps}.
+Below are some methods that can achieve state-of-the art performance and are
+relevant to this work.
 
-tLDA, beamforming
-Linear decoding techniques
+\subsubsection{\Acs{lcmv} beamforming}
+Linear decoding techniques such as the \ac{lcmv}
+beamformer~\cite{Wittevrongel2016} generally work well for
+classifying \ac{erp} epochs.
+\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can
+be multiplied with the feature vector of a new epoch to obtain a prediction
+score.
+The weights $\mat{w}$ are obtained by minimizing the variance of the output
+\begin{equation}
+  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
+	\mat{w}^\intercal
+\end{equation}
+under the linear constraint
+\begin{equation}
+	\mat{a}^\intercal\mat{w} = 1
+\end{equation}
+with activation pattern $a$ the difference between the average of the attended epochs and the
+average of the non-attended epochs in the training data.
+This is then solved by  method of Lagrange multipliers as
+\begin{equation}
+	\mat{w} =
+  \frac{\mat{C}^{-1}\mat{a}^\intercal}
+  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
+\end{equation}
+with $\mat{C}$ the background covariance.
+
+\subsubsection{\Ac{lda}}
+\Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
+\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
+It assumes normally distributed data with a different mean per class but a
+common covariance.
+Similar to \ac{lcmv}-beamforming, \ac{lda} learns a weight vector $\mat{w}$
+that satisfies
+\begin{equation}
+  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
+\end{equation}
+The between-class scatter matrix $\mat{S}_\text{b}$ models the variability
+between the expected responses, the within-class scatter matrix
+$\mat{S}_\text{w}$ the expected noise.
+The optimization criterion corresponds to finding a projection that maximizes
+the spread between classes (signal) while minimizing the spread within classes
+(noise).
+$mat{w}$ can be obtained as eigenvector of
+$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.
+
+There is one one caveat with these linear methods, however:
+the amount of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
+is usually relatively high compared to the training dataset, since calibration
+time should be kept minimal.
+Therefore, strong regularization should be applied, such as covariance/scatter
+shrinkage.
+Regularization can also be performed by retaining structure in the data, such
+as is the case for the multilinear methods presented by~\textcite{Lotte2018}.
+
+\subsubsection{\Ac{tlda}}
+
+Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
+stationarity on the background noise to regularize the between-class covariance
+matrix of \ac{lda}
+\begin{equation}
+  \mat{S}_\text{w} = \begin{bmatrix}
+s_0 & s_1 & s_2 & \cdots & s_{n-1} \\
+s_1 & s_0 & s_1 & \cdots & s_{n-2} \\
+s_2 & s_1 & s_0 & \cdots & s_{n-3} \\
+\vdots & \vdots & \vdots & \ddots & \vdots \\
+s_{n-1} & s_{n-2} & s_{n-3} & \cdots & s_0
+\end{bmatrix}
+\end{equation}
+This method is simple yet effective reaching state-of-the-art decoding
+performance.
+
+\subsubsection{Riemannian geometry}
+
+Alternative methods that currently work well include
+non-linear methods include Riemannian geometry~\cite{Barachant2014}.
+Methods leveraging Riemannian geometry use a different feature extraction
+method.
+They do not use the raw epoch values, but rather operate on the spatial
+covariance of an epoch.
+These covariance matrices $C_i$ are symmetric and should have only positive
+eigenvalues, hence they lie on the Riemannian manifold of symmetric positive
+definite matrices, which defined distance metric
+\begin{equation}
+  \delta_\text{R}\left(\mat{C}_1,\mat{C}_2\right)
+  =\left\lVert\log\left(\mat{C}_1^{-\frac{1}{2}} \mat{C}_2 \mat{C}_1^{-\frac{1}{2}}\right)\right\rVert
+\end{equation}
+This geodesic distance metric is naturally suited for this type of data and has
+some desirable properties, like invariance to linear transformations, which
+result in a robustness particularly useful for \ac{bci}
+applications~\cite{Barachant2011}.
+
+With this distance defined, the epochs can than be either directly be
+classified on the Riemannian manifold using a minimum distance to mean
+classifier, or they can be projected to the tangent space.
+This is a euclidean space that locally approximates the Riemannian distance,
+which allows us to apply regular machine learning methods like an SVM or
+\ac{lda}.
+The tangent space is obtained
+To enhance classifications of \ac{erps} which rely on time domain information
+which would otherwise be lost in the spatial representation of the extracted
+feature, the class averages can be concatenated as additional channels to the
+epochs, embedding the temporal structure in the feature~\cite{Barachant2014}.
+To improve performance even further, the \acp{erp} can be decomposed by a method
+like XDAWN before calculating the extended covariances~\cite{Li2020}.
+
+These decoders define the last building blocks of our visual oddball \ac{bci}.
+Yet, the overview until now has been rather theoretical.
+Interesting and complex challenges arise when applying these concepts outside
+of the lab in augmented and assistive communication technologies for patients
+with \acp{sspi}.
 
 \todo{wrap-up}
-
-\section{The gaze-dependence problem}
 

### 2024-10-04 chapters/conclusion.tex
Rewrite sota

@@
 \section{Recommendations}
+
+this work focused on classification methods, and solving problem through
+decoding
+
+classification performance is not everything
+
+ITR, comfort and user satisfaction in on-line operation is the ultimate goal
+
+also do advance calculations of what the information transfer rate would be
+given reasonable accuracy, so that you can go in with a reasonable interface
+that has state-of-the-art ITR
+
+classification performance is convenient since you can do a single data
+collection and iterate offline.
+this constricts you to decoder development
+it is a better approach to only go there after optimizing the interface
+experimentally and with patients with an existing
+decoder, and only moving there once the options for progress are depleted
+
+performance and abilities are probably so user dependent that it is not said that
+findings from a population study with healthy controls will hold for a given
+user with \ac{sspi}.
+Incremental gains in performance in decoding for a specific paradigm will be
+outweighed by how well that paradigm is adapted to the user, factoring in,
+skills and abilities.
+
+best approach should be everything working together, such as is advocated
+by~\textcite{Pan2022} and \textcite{Fouad2020}
+
+instead of decoder > data > patients > online
+go other way
+online > patients > data > decoder
+this is unfortunately also the least practical way to do it for a 4-year
+research project.
+
+It is therefore important that research labs keep a working, long-term collaboration with
+patient centers such that new projects can immediately be tested by the target
+population, and that patients with bci experience can compare multiple systems
+and guide development
+The fact that a user has previous bci experience is often seen as a confounding
+factor, while it is a strength that can help you gain insight into what works
+and doesn't work for them, and where research efforts should be focused.
+
+also maintain a working, on-line system that can be iterated upon
+
+in a phase that is heavily dependent on user experimentation
+
+not best approach for all research projects, but
+Visual oddball bci field is advanced enough to do this
+
 \subsection{Working with patients}

### 2024-10-04 chapters/gaze_independence.tex
Rewrite sota

@@
 }
-\todohsm{move introduction of paper to paper}
 
-\todo{spatially vs temporally organized interfaces}
 \todo{talk about covert shifts in spatial attention alpha}
@@
     they have \acf{sspgi}.
-    Individuals with \ac{lis} have no muscular control left, or is is very limited,
+    Individuals with \ac{lis} have no muscular control left, or is it is very limited,
     such as the ability communicate through a binary switching system.
@@
   }
-  \caption{fig:gaze_independence/venn}
+  \label{fig:gaze_independence/venn}
 \end{figure}
@@
 \todo{in a paragraph describe each approach and its up and downsides}
-\todo{move form copied introduction}
+\todo{rewrite}
 
@@
 rates compared to visual paradigms.
-Furthermore, they can suffer from increased mental effort in operation and
-user-dependent variability~\cite{Severens2014,Reichert2020b}.
+While auditory and somatosensory stimulation on its own might yield poor
+\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
+settings when coupled with oddball stimulation in a hybrid paradigm.
+Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
+In general, their approach is interesting since they follow a philosophy that
+relies on establishing as many communication pathways as possible.
+Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
+operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
 \textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
@@
 speller~\parencite{Acqualagna2011}.
-Non-spatial visual attention (feature attention) can also be exploited, such as
+The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
+\ac{bci}.
+\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
+a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
+We make a distinction here between spatially organized interfaces, where
+multiple targets are displayed at the same time at different spatial locations,
+and temporally organized interfaces, like \ac{rsvp}\todo{cite},
+where targets or small sets of targets are shown consecutively.
+The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
+variants of the \ac{rsvp} paradigm are temporally organized.
+\textcite{Lin2018} follow the philosophy that spatially organized interface
+have historically shown not to perform well in the presence of gaze impairment.
+However, we argue that, spatially organized interfaces in general have a much higher
+\ac{itr} in healthy controls.
+For instance, a regular matrix-based speller can achive an \ac{itr} to up to
+30bits/min.
+With the right set of adaptations, spatial attention could potentionally also
+bring added value to individuals with \ac{sspgi}.
+While spatial interfaces are indeed more gaze-dependent than temporal
+interfaces they also provide an extra channel that can be used to transfer
+information, i.e.\ the location of the stimulus.
+
+As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
 attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
-Alternative visual stimulation paradigms can modulate specific \ac{erp} components
-that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
+The \ac{rsvp} speller is already an example of this, relying on the user to
+attend to attend the appearance of the intended character to type.
+Visual stimulation paradigms relying on alternative types of attention
+can modulate specific \ac{erp} components
+that are more sensitive to stimulation in the visual
+periphery~\cite{Schaeff2012,Xu2022}.\todo{discuss Xu2022}
 Nevertheless, these methods can suffer too from reduced information
 transfer rates~\cite{Chennu2013}.
-Furthermore, these systems typically require users to focus on a central fixation point while
+Furthermore, the previously mentioned systems typically were only tested in
+settings where the user was required to focus on a central fixation point while
 selecting peripheral targets.
@@
 eye motor control, often necessitating central gaze fixation.
-\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
 
@@
 decoding for any of the investigated decoders.
+
 More recent work has made advances in decoding lateralized covert
-VSA harnessing the N2pc \ac{erp} component~\cite{Thiery2016,Reichert2020b,Wang2022},
-decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
-stimulation paradigms~\cite{Egan2017}.
-Since these methods often require a slower stimulation pace than the classical
-\ac{erp} \ac{bci} paradigm, gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
+VSA harnessing the N2pc \ac{erp}
+component~\cite{Thiery2016,Reichert2020b,Wang2022}
+The amplitude of the N2pc component in the contralateral hemisphere, is directly modulated
+by the location of covert \ac{vsa}.
+Another approach is directly decoding covert shifts in \ac{vsa} from spectral
+content~\cite{Tonin2013}.
+These methods are promising, but to date often require a slower stimulation pace than or cannot display as
+many targets as the classical \ac{erp} \ac{bci} paradigms.
+As with the alternative modality approach, they are currently only helpful as
+extra communication channels in a hybrid paradigm.
+\textcite{Xu2016} reached an \ac{itr} of 23.56bits/min by detecting the N2pc
+component while performing \ac{ssvep} stimulation.
+\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
+concurrently decoding the covert \ac{vsa} shift from spectral content, but
+their study only used a few stimulation targets.
+In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
 leveraging spatial attention remains underexplored.
-
-\todo{
-  Complete with refs from
-
-Fast detection of covert visuospatial
-attention using hybrid N2pc and SSVEP
-features
-}
 
@@
 relies to some extent on eye motor control, we aim to circumvent this.
-We make a distinction here between spatially organized interfaces, where
-multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like \ac{rsvp}\todo{cite},
-where targets or small sets of targets are shown consecutively.
-Traditionally, spatially organized interfaces have a higher information transfer
-rate, but are also more gaze-dependent than temporally organized interfaces.
 In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but

### 2024-10-04 fig.tikz.tex
Rewrite sota

\documentclass[tikz]{standalone}
\usepackage{pgfplots}
\begin{document}
\sffamily
\footnotesize
\begin{tikzpicture}
    \begin{axis}[
        width=\textwidth,
        height=0.6180\textwidth,
        scale only axis,
        xlabel={Time (ms)},
        ylabel={Potential ($\mu$V)},
        axis x line=middle,
	ylabel near ticks,
	xlabel near ticks,
        ymin=-5, ymax=8,
        enlarge x limits=false,
        enlarge y limits=false,
        smooth,
        every axis plot/.append style={very thick, accent1}, % Set the line color to accent1
        ytick={-5,0,5}, % Y-ticks at -5, 0, 5
        xtick={200,400,600}, % X-ticks at 200, 400, 600, 800
        axis line style={darkgray}, % Set the axis line color to darkgray
        tick style={darkgray}, % Set the tick color to darkgray
        tick label style={darkgray} % Set the tick label color to darkgray
    ]
    % ERP waveform using cubic splines for smooth connection
    \addplot[
        smooth,
        tension=0.5
    ] coordinates {
        (0, 0)     % Starting point
        (50, 0)    % Pre-P1
        (80, 2)    % P1
        (115, -3)  % N1
        (160, 3)   % P2
        (220, 1)   % N2
        (400, 7)   % P3
        (650, 1)   % LPP (Late Positive Potential)
        (800, 1)   % End point
    };

    % Annotations for the components
    \node[anchor=south, color=darkgray] at (axis cs:80,2) {P1};
    \node[anchor=north, color=darkgray] at (axis cs:115,-3) {N1};
    \node[anchor=south, color=darkgray] at (axis cs:160,3) {P2};
    \node[anchor=west, color=darkgray] at (axis cs:220,1) {N2};
    \node[anchor=south, color=darkgray] at (axis cs:400,7) {P3};
    \node[anchor=south west, color=darkgray] at (axis cs:650,1) {LPP};

    \end{axis}
\end{tikzpicture}
\end{document}

### 2024-10-04 figures/bci/erp.tikz.tex
Rewrite sota

\sffamily
\sansmath
\footnotesize

\begin{tikzpicture}[trim axis left]
    \begin{axis}[
        width=\textwidth,
        height=0.6180\textwidth,
        scale only axis,
        xlabel={Time (ms)},
        ylabel={Potential ($\mu$V)},
        axis lines=middle,
	ylabel near ticks,
	xlabel near ticks,
	y label style={darkgray},
	x label style={darkgray},
        ymin=-5, ymax=10,
        xmin=-100, xmax=800,
        enlarge x limits=false,
        enlarge y limits=false,
        smooth,
        every axis plot/.append style={very thick, accent1}, % Set the line color to accent1
        ytick={-5,0,5}, % Y-ticks at -5, 0, 5
        xtick={200,400,600}, % X-ticks at 200, 400, 600, 800
        axis line style={darkgray}, % Set the axis line color to darkgray
        tick style={darkgray}, % Set the tick color to darkgray
        tick label style={darkgray} % Set the tick label color to darkgray
    ]
    % ERP waveform using cubic splines for smooth connection
    \addplot[
        smooth,
        tension=0.5
    ] coordinates {
        (-100, 0)  % Starting point
        (0, 0)     % Baseline
        (50, 0)    % Pre-P1
        (80, 2)    % P1
        (115, -3)  % N1
        (160, 3)   % P2
        (220, 1)   % N2
        (400, 7)   % P3
        (650, 1)   % LPP (Late Positive Potential)
        (800, 1)   % End point
    };

    % Annotations for the components
    \node[anchor=south, color=muteblack] at (axis cs:80,2) {P1};
    \node[anchor=north, color=muteblack] at (axis cs:115,-3) {N1};
    \node[anchor=south, color=muteblack] at (axis cs:160,3) {P2};
    \node[anchor=west, color=muteblack] at (axis cs:220,1) {N2};
    \node[anchor=south, color=muteblack] at (axis cs:400,7) {P3};
    \node[anchor=south west, color=muteblack] at (axis cs:650,1) {LPP};

    \end{axis}
\end{tikzpicture}

### 2024-10-04 chapters/bci.tex
Update bttda formulation

@@
 \todo{search for ??}
+\todo{acro usage in chapters}
 
@@
 
-\subsection{Active \& passive \ac{bci}}
+\subsection{Active \& passive \acsp{bci}}
 
@@
 
-\subsection{Reactive \ac{bci}}
+\subsection{Reactive \acsp{bci}}
 Reactive \ac{bci} take another approach, by providing a discrete set of
@@
 In \ac{erp} classification, they are usually directly used as features, by
-concatenating all channels in the matrix format epoch $X$ into one feature
-vector s.t.\ $\mat{x} = \mat\vec(X)$.
+concatenating all channels in the matrix format epoch $\mat{X}$ into one feature
+vector s.t.\ $\mat{x} = \mat\vec(\mat{X})$.
 
 \subsection{Decoders}
-Several techniques are suited to decode~\ac{erps}.
+\todo{averaging over trials}
+Several techniques are suited to decode~\acp{erp}.
 Below are some methods that can achieve state-of-the art performance and are
@@
 
-\subsubsection{\Acs{lcmv} beamforming}
+\subsubsection{\Acf{lcmv} beamforming}
 Linear decoding techniques such as the \ac{lcmv}
@@
 
-\subsubsection{\Ac{lda}}
+\subsubsection{\Acf{lda}}
 \Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
@@
 
-\subsubsection{\Ac{tlda}}
+\subsubsection{\Acf{tlda}}
 
 Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
-stationarity on the background noise to regularize the between-class covariance
-matrix of \ac{lda}
+stationarity on the background noise to regularize the problem.
+It does this by imposing a block-Toeplitz structure to the between-class covariance
+matrix of \ac{lda}, such that every temporal covariance block corresponding to
+a channel has the structure
 \begin{equation}
-  \mat{S}_\text{w} = \begin{bmatrix}
-s_0 & s_1 & s_2 & \cdots & s_{n-1} \\
-s_1 & s_0 & s_1 & \cdots & s_{n-2} \\
-s_2 & s_1 & s_0 & \cdots & s_{n-3} \\
+  \mat{S}_\text{w}^(c) = \begin{bmatrix}
+s_0 & s_1 & s_2 & \cdots & s_{k-1} \\
+s_1 & s_0 & s_1 & \cdots & s_{k-2} \\
+s_2 & s_1 & s_0 & \cdots & s_{k-3} \\
 \vdots & \vdots & \vdots & \ddots & \vdots \\
-s_{n-1} & s_{n-2} & s_{n-3} & \cdots & s_0
+s_{k-1} & s_{k-2} & s_{k-3} & \cdots & s_0
 \end{bmatrix}
@@
 \ac{lda}.
-The tangent space is obtained
-To enhance classifications of \ac{erps} which rely on time domain information
+The tangent space is obtained\todo{tangent space}.
+To enhance classifications of \acp{erp} which rely on time domain information
 which would otherwise be lost in the spatial representation of the extracted
@@
 
-\todo{wrap-up}
+\section{The gaze-dependence problem}
 

### 2024-10-04 chapters/bttda.tex
Update bttda formulation

@@
 \input{chapters/bttda/tensorviz.tex}
-\todo{abbreviations}
 
 \section{Introduction}
-\todo{immediately arrive to the novelty or to the point}
 \todo{reduce length of the introduction}
-\todo{less acronyms}
-\todo{put the intuition behind the algorithm (sparse, more flexible, what it
-does in one sentence ) in the introduction. Additionally "try to get some
-information from the data, continue extracting more information as long as
-there is some left}
+
 Brain-computer interfaces (BCIs) have the potential to bypass
@@
 a neuroimaging technique, with electroencephalography (EEG) being the most popular.
-\todo{explicitly state how BTD is a generalization of PARAFAC and TUCKER}
 
@@
 separately, reducing a given problem into partial, per-mode problems.
-This approach has given rise to efficient dimensionality reduction techniques,
-such as the Higher Order Singular Value Decomposition
-(HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}
-and Canonical Polyadic Decomposition
-(CPD)~\cite{Hitchcock1927,Nazarpour2006}.
 
-HOSVD and CPD are two examples of unsupervised techniques with applications in EEG
-processing.
-However, given the low signal-to-noise ratio and
-specific, task-related output expected in a BCI application,
-supervised feature extraction and machine learning techniques are
-favored\todo{cite Lotte review}.
-%One approach is to incorporate assumptions of the tensor structure of the data
-%directly into the estimation of parameters of classic supervised linear -
-%machine learning methods, such as linear discriminant analysis (LDA) or beamforming.
-%Advances have been made by leveraging the decoupling of the spatial and temporal
-%domains in EEG event-related potential (ERP) classification using Spatiotemporal
-%Discriminant Analysis~\cite{Li2010,Zhang2013} or methods regularizing covariance
-%matrix estimation~\cite{Kerchove2022,Sosulski2022}.
+These tensor methods decompose a tensor into a lower dimensional structure of a
+core tensor and factor tensors.
+The most common aproaches adhere to either the Tucker structure or the PARAFAC
+structure.
+A Tucker decomposition reduces an input tensor of size $(D_1,D_2,\ldots,D_K)$ to
+a dense  tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k<=D_k$ using a
+set of per mode factor matrices.
+Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
+using the Higher Order Singular Value
+Decomposition (HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}.
+Alternatively, the PARAFAC structure can be used.
+Here, the tensor is decomposed into a sum of rank-1 tensors, each the product
+of a scalar and a vector per mode.
+This is equivalent to a Tucker structured decomposition with all core elements
+off the hyperdiagonal set to 0.
+One way of obtaining an unsupervised PARAFAC decomposition is through the Canonical Polyadic
+Decomposition~\cite{Hitchcock1927,Nazarpour2006}.
 
-\todo{come in here with François' comment about restructuring the intro.}
-A more structured approach to the same problem is to design a supervised
-tensor dimensionality reduction method that optimizes discriminability between the
-extracted features, as is the case with Higher Order Discriminant
-Analysis (HODA)~\cite{Yan2005,Phan2010,Froelich2018}.
+While commonly used, these Tucker or PARAFAC structure might still not be able to
+efficiently represent relevant neural information in a compressed format.
+The block-term tensor structure is a generalization of the Tucker and
+PARAFAC structures.
+It represents the tensor as a sum of Tucker structured terms.
+If the number of terms is equal to 1, it is equivalent to the Tucker structure;
+if the rank of each term is equal to 1, it is equivalent to the PARAFAC
+structure.
+The generalized structure can be calculated in an unsupervised way using
+the Block-Term Decomposition~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
+Performance of methods leveraging either the Tucker and PARAFAC structure are
+heavily dependent on the a priori choice of hyperparameters describing the
+respectively the multilinear rank or the number of rank-1 terms.
+The block-term structure is more flexible than either the Tucker or the PARAFAC
+structures, since it is not constrained to problems that cannot be expressed by
+one of these structures and the chosen hyperparameters.
+However, this increased flexibility also increased the number of
+hyperparameters to both the number of terms and the multilinear rank of each term.
+
+If the decompositions are not full rank, the Tucker, PARAFAC and block-term
+structures are not unique and can be obtained by optimizing different criteria.
+Given the low signal-to-noise ratio and specific, task-related output expected
+in a BCI application, supervised feature extraction and machine learning techniques are
+favored~\cite{Lotte2018} over the unsupervised decomposition methods presented
+above.
+A decomposition that is helpful for classification purposes should ideally optimize
+the discriminability between classes in the resulting core tensors, which can
+be considered as extracted features.
+In this philosophy, the Tucker decomposition can also be obtained
+using Higher Order Discriminant Analysis
+(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fischer sense, analogous to linear
+discriminant analysis.
 Extracted features can subsequently be further classified, most commonly
 using LDA or a support vector machine (SVM) to predict class labels.
+
 Variants of HODA have been applied to BCI problems such as
-ERP~\cite{Onishi2012,Higashi2016}\todo{cite spatiotemporal discriminant
-analysis here, 'variants and special cases'} and motor imagery (MI)~\cite{Liu2015,Cai2021}
-decoding.
+ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
+decoding with positive results~\cite{Lotte2018}.
 Recent work proposes adaptations such as suited objective
 functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
+Discriminant tensor features have also be extracted
+in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
+However, it is not immediately obvious if either the Tucker or PARAFAC
+structure are most suited to represent the neural data of interest for the \ac{bci}
+paradigm and for decoding.
 
-The methods above adhere to the Tucker tensor decomposition
-structure,
-\todo{
-To introduce BTD, I would start with a paragraph explaining the decomposition
-itself, or the approximation for dimension reduction, just after the
-presentation of tensors: on one side, Tucker-N (1 core only) and on the other
-side polyadic decomposition (PD, m scalars) and in the middle the BTD. This
-paragraph explains that in most cases the decomposition is not unique and
-depends on a priori parameters (nb of cores, dimensions of projection matrices,
-etc.) Also defines the residual, difference between the original tensor and the
-model in the case of approximation.
-Then, the techniques for getting an
-"optimal" decomposition or approximation could be presented: HOSVD and equiv.
-(just a few refs) in the case of Tucker and CPD and others (just a few refs)
-for PD. Then, your state of the art could be just focused on specific
-optimizations of BTD used in BCIs (in order to justify the JNE journal
-target)
-In fact, this is the option chosen by Axel Faes and Marc in the JNE
-paper of 2022)...}
-meaning that they reduce input tensors of size
-$(D_1,D_2,\ldots,D_K)$ to a smaller tensor of size $(r_1,r_2,\ldots,r_K)$ with
-each $r_k\leq D_k$, similar to HOSVD.
-While effective, other approaches such as the PARAFAC structure employed in
-CPD, where a tensor is decomposed into a sum of rank-1 tensors,
-might also be suitable to represent the neural data of interest.
-Discriminant tensor features can also be extracted
-in the PARAFAC structure, for instance through manifold
-optimization~\cite{Froelich2018}.
-
-Nevertheless, the PARAFAC structure might still not be able to
-efficiently represent all relevant information in a compressed format.
-The block-term tensor structure is a generalization of the Tucker and
-PARAFAC structures, and can be calculated in an unsupervised way using
-the Block-term Tensor Decomposition
-(BTD)~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
-BTD, of which the HOSVD and CPD are special cases,
-represents a tensor as sum of Tucker terms.
-Research has shown that supervised decoders adopting this more flexible structure
+Recent research has shown that supervised decoders adopting a more flexible structure
 can improve BCI performance.
@@
 cannot be extended to a multi-class setting.
-Furthermore, structures employed in HOPLS, BTTR are still more constrained than
-what could be achieved with a full block-term tensor structured decomposition
+Furthermore, structures employed in these regression are still more constrained
+than what could be achieved with a full block-term tensor structured decomposition
 optimized for discriminability, since they rely on a low rank common subspace
 between the input and classification labels.
-\todo{remove acronyms if not used later}
+\textcite{Huang2020} propose a supervized approach for finding multiple discriminant
+multilinear spectral filter terms and apply it to motor imagery \ac{bci}, but their
+decomposition is also restricted in flexibility.
 
-%\todo{Describe spectrum-weighted tda~\cite{Huang2020}, structure is not
-%	flexible enough, not broadly applicable, not interpretable}
+\todo{It's always trichiin in the related work section of a ML/Mathematical
+paper, but unless you r reviewers are absolute experts of this specific field,
+you should at least try to better describe the previous methods.}
 
-We propose to mitigate these issues by designing a new supervised feature
-extraction tensor method based on the abovementioned HODA algorithm
-that is more suited for the extraction of discriminant
+A block-term decomposition that is directly optimized for discriminability and with a
+good choice of ranks and number of terms might better represent the behaviour
+of generators of neural activity through increased flexibility, or might
+achieve better regularization through increased sparsity.
+An alternative view on the same approach goes as follows:
+If HODA with a well chosen multilinear rank extracts some discriminant features
+from the input tensor, it is likely that it does not yet retrieve all useful
+information due to the restriction following from its Tucker structure.
+Could we therefore not keep using HODA to extract discriminant Tucker structured
+as long as decoding performance increases?
+
+We propose to implement this idea as a new supervised feature
+extraction tensor method that is a generalization of the aforementioned HODA
+algorithm and is more suited for the extraction of discriminant
 features while adhering to a flexible and efficient block-term tensor
@@
 \end{enumerate*}
-\todo{It's always trichiin in the related work section of a ML/Mathematical
-paper, but unless you r reviewers are absolute experts of this specific field,
-you should at least try to better describe the previous methods.}
-
 \section{Methods}
@@
 these class mean latent tensors.
-The goal is now to find the optimal projection matrices\todo{You should say
-somewhere that the shape (r1,r2,...rK) is fixed a priori, no?}
+If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices
 \begin{equation}
@@
 \end{equation}
-which can be solved by the iterative algorithm in Algorithm~\ref{alg:HODA}.
-\begin{algorithm}
-  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
-	\label{alg:HODA}
-	\input{chapters/bttda/alg_HODA.tex}
-\end{algorithm}
+which is solved through the HODA algorithm.
 To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
@@
 update of each $\mat{U}_k$ is lower than a predetermined threshold.
+The full HODA procedure is summarized in Algorithm~\ref{alg:HODA}.
+\begin{algorithm}
+  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
+	\label{alg:HODA}
+	\input{chapters/bttda/alg_HODA.tex}
+\end{algorithm}
 
@@
 $\alpha_k$ at each step such that the eigenvalue problem becomes
-\todo{make sure this appears after varphi is defined, move reference to
-algorithm to later}
 \begin{equation}
@@
 \begin{equation}
-	\ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}}
+  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
+  \hat{\ten{X}} + \ten{\mathbfcal{E}}
 	\label{eq:HODA-forward}
 \end{equation}
-with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$
-and error term $\ten{\mathbfcal{E}}$.
+with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
+reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
 \begin{figure}[t]
@@
 	\input{figures/bttda/HODA_forward.tikz.tex}
-  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA. By calculating activation
-		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
-		reconstructed from projected latent tensor $\ten{G}$. $A_k$ are chosen such
-		that the variability captured in the latent tensor is maximally expressed in
-		the reconstructed tensor and not in the error term.}
+  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
+    By calculating activation	patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
+    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
+    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
+    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
+    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
+    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
+    the error term $\ten{\mathbfcal{E}}$.}
 	\label{fig:HODA-forward}
@@
 A forward model should make sure that reconstruction error is minimized and
-variation captured in the latent tensor is maximally captured by the forward
-projection term $\ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
+variation captured in the latent tensor is maximally captured by the
+reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
 \todo{not clear ("error is minimized" in the same sentence)}
@@
 \begin{equation}
-	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
+	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
+  -\hat{\ten{X}}(n)\right)^2
+  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
 \end{equation}
@@
 For each block $b$, the latent tensor is extracted using the HODA backward
-projection in \todo{ref equation}from a residual error term
-$\ten{\mathbfcal{E}}^{(b)}$
+projection in \todo{ref equation} from the residual error term of the previous
+block $\ten{\mathbfcal{E}}^{(b-1)}$
 \begin{equation}
-	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b)}\mmpr{\mat{U}^{(b)}}
+	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
 \end{equation}
@@
 \begin{equation}
-	\ten{\mathbfcal{E}}^{(b+1)} = \ten{\mathbfcal{E}}^{(b)} - \ten{G}^{(b)}
-	\mmpr{\mat{A}^{\intercal(b)}}
+  \ten{\mathbfcal{E}}^{(b)}
+  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
+  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
 \end{equation}
-with $\ten{\mathbfcal{E}}^{(1)}=\ten{X}$.
+with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.
 
@@
 \subsection{Model and feature selection}
-Similar to the unsupervised BTD, the performance of BTTDA is
+Similar to the unsupervised Block-Term Decomposition, the performance of BTTDA is
 heavily dependent on the rank $(r_1^{(b)}, r_2^{(b)}, \ldots,
@@
 Alternatively, a special case of BTTDA can be constructed using only rank-one
-blocks such that the resulting forward model adheres to the CPD structure.
+blocks such that the resulting forward model adheres to the PARAFAC structure.
 The optimal amount of rank-one blocks can be found by truncating as above.
@@
 We evaluated our proposed model in two offline EEG-based BCI decoding problems:
-the event-related potential (ERP) and motor imagery (MIà paradigms using a
+the event-related potential (ERP) and motor imagery (MI paradigms using a
 selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
@@
 statistical measures based on the models application.
-\todo{comment as limitation that we test r1=r2=r3}
+\todo{comment as limitation that we test r1=r2=r3, this is against the whole
+idea of increased flexibility, find automated rank selection procedure}
 
@@
 selection problem, but it is more flexible}
-We have introduced Block-term Tensor Discriminant Anlysis (BTTDA) a novel,
+We have introduced Block-Term Tensor Discriminant Anlysis (BTTDA) a novel,
 tensor-based, supervised dimensionality reduction technique optimized for class
@@
 \section*{Acknowledgements}
+\todo{move to the end}
 We thank the Flemish Supercomputer Center (VSC) and the High-Performance

### 2024-10-04 chapters/conclusion.tex
Update bttda formulation

@@
 
-\subsection{Investigated gaze-independence visual \acsp{bci}}
+\subsection{Investigated gaze-independence visual \acp{bci}}
 
@@
 
+\subsection{Decoding: keep it linear}
+
+keep it linear
+use linear methods, but impose specific structure on them reflecting properties
+of the signal, either to make them better regularized or more robust to some
+types of variation
+
+\subsection{Working \emph{with} individuals \acs{sspi}}
+
 this work focused on classification methods, and solving problem through
@@
 
-ITR, comfort and user satisfaction in on-line operation is the ultimate goal
-
+ITR in on-line operation is the ultimate goal
+, and, ultimately, user satisfaction, which depends on performance, comfort and perception
 also do advance calculations of what the information transfer rate would be
@@
 
-%Have attention for the emotional impact
-
-\subsection{Linear decoding methods}
+Have attention for the personal aspect
 

### 2024-10-04 chapters/gaze_independence.tex
Update bttda formulation

@@
 
-\todo{talk about covert shifts in spatial attention alpha}
-
-\section{Eye motor impairment in individuals with \acs{sspi}}%
+\section{Eye motor impairment in individuals with \acl{sspi}}%
 \label{sec:gaze-dependence}
-
-\todo{define eye motor issues}
-
-\todo{Algemeen: in begin uitleggen dat de ziektes er niet veel toe doen en dat
-het gaat over verlamde personen, verlies van spierkracht en coördinatie en dus
-gehinded in communicatie, helemaal in begin van chapter 1 duidelijk maken dat
-het om deze patiënten gaat}
 
@@
 communication devices.
-\todo{to the CNS (stroke, spinal chord injury, tbi) or neurodegenerative diseases}
 
-Unfortunately, many individuals in the \ac{bci} target population suffer to
-some extent from eye motor disability, requiring \ac{bci}s adapted to their condition.
+\Ac{sspi} is often caused by damage to the central or peripheral nervous
+system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
+neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
+\ac{tbi}).
+Many individuals in these groups unfortunately also suffer to
+some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
 Table~\ref{tab:incidence} reports the relatively high frequencies of
-eye motor problems, which can range from minor (nystagmus
+eye motor impairment, which can range from minor (nystagmus
 \footnote{%
@@
 This not only has an effect on vision and coordination but also
-on their ability to operate a visual \ac{bci}.
+on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.
 
-\begin{table}[h]
-  \sffamily
-  \footnotesize
-	\centering
-	\begin{tabular}{@{}l|rrrrr|r@{}}
-    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
-		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
-		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
-		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
-	\end{tabular}
-  \caption[Incidence of eye motor impairment in selected \ac{bci} target
+\begin{table}
+  %\sffamily
+  %\footnotesize
+	%\centering
+  \input{chapters/gaze_independence_incidence.tex}
+  \caption[Incidence of eye motor impairment in selected \acs{bci} target
     patient populations.]{%
       Incidence of eye motor impairment in selected \ac{bci} target
-      patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
-      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
+      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
+      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
       $+$: frequent, $-$: infrequent.
@@
 \end{table}
-\todo{include FA}
-\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
-\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}
 
-\todo{
-14] Fazzi E, Signorini SG, Piana RL, et al. Neuro-ophthalmo-
-logical disorders in cerebral palsy: ophthalmological, oculo-
-motor, and visual aspects. Dev Med Child Neurol. 2012;54:
-730–736.
-[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
-visual function of patients with locked-in syndrome: report
-of 13 cases. J Fr Ophtalmol. 2016;39:437–440.
-[16] Moss HE, McCluskey L, Elman L, et al. Cross-sectional evalu-
-ation of clinical neuro-ophthalmic abnormalities in an
-amyotrophic lateral sclerosis population. J Neurol Sci. 2012;
-314:97–101.
-}
-
-Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
+Among the most  severely affected patients are stroke
+patients~\cite{Pollock2011,Rowe2019}, mostly
 due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
-\todo{summarize and cite Rowe2019}
 These patients might suffer from severe or complete eye motor impairment
 from the onset of their condition.
-In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
+Individuals recovering from stroke with severe eye motor impairment are
+Furthermore, those with
+Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
+with \ac{lis} due to stroke, the group with a complete lack of eye motor
+control is very small.
+
+In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
+are also fairly common.
 Although eye movement is often cited as one of the longest preserved
-capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
-fairly common,
-especially for patients with bulbar onset \ac{als}.
+capabilities in \ac{als}, show that minor issues are still
+fairly common, especially for patients with bulbar onset
+\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
 Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
@@
 eye muscle paralysis.
-Various forms of eye movement abnormalities also occur often and are especially well studied in
-\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
+
+Various forms of eye movement abnormalities also occur often in
+\ac{ms}.
+\ac{ms} is a neurodegenerative disease involving demyelination of nerves.
+Eye motor abnormalities are especially well
+studied~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020} and
+often used as diagnostical tool.
 These abnormalities can be minor or severe, seldom progressing to complete
@@
 with visual \acp{bci}.
-\todo{FA}
-Patients suffering from other neurodegenerative diseases like
+
+\ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
+This almost always affects eye
+movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
+involuntary movements and troubles pursuing or fixating targets.
+They also gradually have more troubles speaking, but often retain some muscular
+control.
+
+Some other groups that are heavily affected are \ac{cp}~\cite{Fazzi2012} and
+\ac{fa}~\cite{Fahey2008,Hocking2010,Furman1983}.
+Individuals with some neurodegenerative diseases like
 \ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
@@
 the inability to speak, such as occurs in late stage ALS.
-In clinical reality, every tetraplegic patient has their own unique set of
-preserved capabilities and from a solution-oriented \ac{bci} engineering point
-of view, the etiology of the symptoms can be abstracted away.
 }}
@@
 
-The most severely impaired of these patients toghether constitute the \ac{lis}
-patient group, which forms one of the main \ac{bci} target populations.
-In this work, we define \ac{lis} patients as those in a situation of (near)
-complete paralysis and difficulties or the inability to communicate independently,
-even with assistive technology.
-This corresponds to classes one and two defined by
-\textcite{Wolpaw2006}\fnwolpawcrit.
-Additionally, some degree of severe or complete eye motor impairment is usually
-necessary to qualify as locked-in\fnlis.
-Case study reports~\cite{Patterson1986} show that the population of locked-in
-patients with complete eye motor paralysis is very small.
-Hence, it would be interesting to focus \ac{bci} development efforts on the
-larger population of individuals with \ac{sspgi} that currently slip through
-the cracks of the assistive technology offer.
-These are individuals whose severe eye motor impairment prevents them from
-using eye tracking based solutions.
-They currently use their remaining motor control to
-communicate by indicating symbols on a letterboard with great effort,
-or signal with upwards eye movements or blinks to confirm prompted letters.
-They require a caregiver or relative to interpret their signals, and crave
-the ability to communicate independently which is crucial to retaining an
-acceptable quality of life.
+We believe it is in general not opportune to delve too deep in the underlying
+ophthalmological and neurological pathophysiology for each of the specific
+conditions.
+In clinical reality, every patient different symptoms, which lead to their own
+unique set of preserved capabilities.
+From a solution-oriented \ac{bci} engineering point
+of view, the etiology of the symptoms can be abstracted away.
+Instead of categorizing patients we will refer to these groups as individuals
+with \ac{sspi} and, if eye motor skills are affected,  \ac{sspgi}, in line with
+the terminology used by~\textcite{FrieDOken2020} (see
+Figure~\ref{fig:gaze_independence/venn}).
 
@@
 \end{figure}
+
+The most severely impaired of these patients constitute the \ac{lis}
+patient group, which forms one of the main \ac{bci} interest groups.
+In this work, we use the term \ac{lis} for a situation of (near)
+complete paralysis and difficulties or the inability to communicate independently,
+even with assistive technology.
+This corresponds to classes one and two defined by
+\textcite{Wolpaw2006}\fnwolpawcrit.
+\todo{[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
+visual function of patients with locked-in syndrome: report
+of 13 cases. J Fr Ophtalmol. 2016;39:437–440.}
+Hence, some degree of severe or complete eye motor impairment is usually
+necessary to qualify as locked-in\fnlis.
+
+In general, the group of locked-in individuals with complete loss of eye motor
+function is very small.
+Hence, it would be interesting to focus \ac{bci} development efforts on the
+larger population of individuals with \ac{sspgi}  that currently slip through
+the cracks of the assistive technology offer.
+These are individuals whose severe eye motor impairment prevents them from
+using eye tracking based solutions.
+They currently use their remaining motor control to
+communicate by indicating symbols on a letterboard with great effort,
+or signal with upwards eye movements or blinks to confirm prompted letters.
+They require a caregiver or relative to interpret their signals, and crave
+the ability to communicate independently which is crucial to retaining an
+acceptable quality of life.
 
@@
 \label{sec:gaze-independence}
-\todohsm{Address more systematically.}
-\todo{in a paragraph describe each approach and its up and downsides}
-\todo{rewrite}
 
@@
 with invasive recording.
-Non-visual reactive paradigms that use e.g. auditory and
-somatosensory\todo{cite some}
+Non-visual reactive paradigms that use e.g. auditory~\cite{Halder2010} and
+somatosensory~\cite{Brouwer2010}
 stimulation do not rely on gaze redirection exist but also result in lower information transfer
@@
 
-\todo{insert table from riccio in text as numbers, not as table}
-
 \subsection{Gaze-independent visual stimulation}
@@
 multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like \ac{rsvp}\todo{cite},
+and temporally organized interfaces, like \ac{rsvp},
 where targets or small sets of targets are shown consecutively.
@@
 Visual stimulation paradigms relying on alternative types of attention
-can modulate specific \ac{erp} components
-that are more sensitive to stimulation in the visual
-periphery~\cite{Schaeff2012,Xu2022}.\todo{discuss Xu2022}
-Nevertheless, these methods can suffer too from reduced information
-transfer rates~\cite{Chennu2013}.
+can modulate specific extra \ac{erp} components that either improve performance
+because the embed extra information in the brain signal~\cite{Xu2022}
+or because they are more sensitive to stimulation in the visual
+periphery~\cite{Schaeff2012}.
+Nevertheless, solely relying on alternative types of attenion can suffer too
+from reduced information transfer rates~\cite{Chennu2013}.
 Furthermore, the previously mentioned systems typically were only tested in
@@
 visual oddball \ac{bci}.
-\todo{refer to ITRs from previous chapter}
 Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
@@
 
-This leads us to adopt a specific approach: improve the performance of a
+This leads us to adopt a specific approach: \emph{improve the performance of a
 spatially organized visual oddball \ac{erp}-based brain
-computer interface by using a suited decoding strategy.
-Hence, our working hypothesis is as follows:
-\todo{rewrite, this is not what we tested in the patient paper after all}
-\begin{quote}
-	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
-		can achieve a similar transfer rate in patient
-		populations with eye motor deficits than other gaze-independent alternatives
-    described in literature.}
-\end{quote}
+computer interface by using a suited decoding strategy}.
+%Hence, our working hypothesis is as follows:
+%\todo{rewrite, this is not what we tested in the patient paper after all}
+%\begin{quote}
+%	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
+%		can achieve a similar transfer rate in patient
+%		populations with eye motor deficits than other gaze-independent alternatives
+%    described in literature.}
+%\end{quote}
 
-To achieve our goal, we innovate on decoder development and interface design,
-and we collect data from healthy control subjects, which allow us to test our decoders.
-Finally, the findings of these experiments will be leveraged to develop a \ac{bci}
-for patient use, which will be tested in a case-based patient study in which
-both performance and user comfort was be evaluated.\todo{was this actually
-done?}
+To achieve our goal, we chose a suitable interface and innovate on decoder
+development.
+We collect data from healthy control subjects, which allow us to test our decoders.
+Finally, the findings of these experiments will be verified in experiments with
+individuals with \ac{sspgi}, including individuals with \ac{lis} evaluating
+both performance and comfort.
 
-This specific hypothesis is embedded within a broader goal of enabling
+This specific target is embedded within a broader goal of enabling
 effective communication for eye-motor impaired \ac{lis} patients.
 Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
-their residual gaze capabilities, by leveraging a non-invasive, high-\ac{itr}
-\ac{bci} on the one hand, and on the other hand improving \ac{erp} decoding performance
-in general.
+their residual gaze capabilities.
+We believe this can be done with a non-invasive, spatially
+organized visual \acp{bci} with high-\ac{itr}.
+Finally, or efforts also aim to improve hand improving \ac{erp} decoding performance
+in general, which will also contribute to their effectiveness in gaze-dependent
+or gaze-independent settings.
 
@@
 
-\subsection{Interface design}
-\todohsm{rename data collection. start with decoder design, second section data collection with experiments}
+\subsection{Decoder design}
+
+During this PhD, we explored different lines in decoding strategies,
+trying to tackle several problems, be it general problems in \ac{erp} decoding
+or those that arise specifically from gaze-independence.
+Examples of this are the lack or decrease in amplitude of specific \ac{erp} components, and the increased
+non-stationarity of the signal.
+As mentioned, state-of-the-art decoders have poor performance in covert attention
+settings.
+The general goal is thus to design a machine learning classifier that represents
+the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
+occurring in covert attention conditions.
+
+\subsubsection{Regularized spatiotemporal beamforming}
+Due to the decreased amplitude of the N1 and P3 components in covert attention
+settings~\cite{Treder2010}, the \ac{snr} of
+the \ac{erp} is lower than in overt attention settings.
+Therefore, a straightforward way to reach satisfactory
+gaze-independent decoding performance, might be by increasing overall \ac{erp}
+decoding performance.
+
+Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
+decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
+this classifier as a linear discrimination problem and
+imposing regularizing constraints by structuring the noise covariance matrix
+(STBF-struct).
+Furthermore, these regularizing constraints impose temporal stationarity on
+the background noise, yielding insights for our next
+efforts to cope with the non-stationarity of the P3 signal component.
+This approach has been published by~\textcite{VanDenKerchove2022} and will be
+described in Chapter~\ref{sec:stbf-struct}.
+
+\subsubsection{Tensor discriminant analysis}
+\todo{fill in}
+
+\subsubsection{Classifier-based Latency Estimation with Woody iterations}
+The previous two methods were aimed at improving decoding performance, but
+did not (yet) yield results specifically in gaze-independent decoding.
+The following approach is targeted towards improving decoding performance in
+gaze-independent settings by accounting for a known property of the covert
+attention \ac{erp} response.
+
+P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
+value is heavily dependent on the subject and the task, and can vary from trial
+to trial~\cite{Ouyang2017}.
+The work of \cite{Arico2014} illustrates that the variation in single-trial P3
+latencies is important in gaze-independent decoding and has been hampering covert
+VSA decoding performance.
+We reprised their hypothesis stating that jitter compensation through latency
+estimation and alignment improves
+covert \ac{vsa} performance and extend it by developing a decoder.
+
+Existing latency estimation methods are either not applicable to the
+classification problem of labeling unseen data, or are not robust enough to
+deal with the low \ac{snr} of the \ac{erp}.
+\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
+\ac{erp} latency estimation in a decoding setting, but our results show that it
+yields no improvement in gaze-independent settings.
+We improved upon this technique and extended it to a probabilistic, iterative
+method named \ac{wcble}.
+This approach has been published by~\cite{VanDenKerchove2024} and is presented
+and evaluated on simulated data in Chapter~\ref{sec:wcble}.
+We have also collected datasets to evaluate this approach on real \ac{erp} data
+in gaze-independent settings.
+
+\subsection{Data collection}
+
+\subsubsection{The CVSA-ERP dataset}
+
+In a first series of experiments, we recorded data with this interface from healthy participants.
+The goal of these experiments is to benchmark
+gaze-independent \ac{erp} decoding algorithms.
+We denote this dataset as the Covert Visuospatial Attention \ac{erp} dataset
+(CVSA-\ac{erp})
+This study was approved by the Ethics Commission of University Hospital Leuven
+(S62547).
+To simulate the dissociation between the eye gaze and the
+intended target (\ac{vsa}) which could occur in individuals with gaze
+impairment, healthy participants are cued to operate the \ac{bci} in specific
+\ac{vsa} conditions.
 
 \todo{mention, free, split, frenzel}
-\todo{clear definitions, table from paper}
 
-First, we designed a visual oddball interface that helps us study and adapt
+We chose a visual oddball interface that helps us study and adapt
 to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
@@
 electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
-\todo{chosen because gaze-independent optimized and Peters2020}
+The Hex-o-Spell was chosen since it is optimiazed for gaze-independent
+performance and because it has already been tested in individuals with
+\ac{sspi} by~\textcite{Severens2014}.
 
@@
 \end{figure}
+
 In the overt case, users gaze at the cued
@@
 another (the distractor).
-\todomvh{+ "free" condition in Fig. 21.}
+This last option has only scarcely be studied, but completes
+the option s to dissociate gaze and visuospatial attention, which allow us to
+investigate the effect of (the lack of) gaze control on \ac{bci} performance.
+This dataset and performance of our proposed~\ac{wcble} decoder has also been
+published by~\textcite{VanDenKerchove2024} and will be presented in
+Chapter~\ref{sec:covert-align}.
+Performance is also evaluated on a publicly available dataset.
 
-The option to dissociate gaze and visuospatial attention will allow us to
-investigate the effect of gaze control on \ac{bci} performance.
-
-\subsection{Decoders}
-
-During this PhD, we explored different lines in decoding strategies,
-trying to tackle several problems that arise from gaze-independence, such as
-the lack or decrease in amplitude of specific \ac{erp} components, and the increased
-non-stationarity of the signal.
-As mentioned, state-of-the-art decoders have poor performance in covert attention
-settings.
-The general goal is thus to design a machine learning classifier that represents
-the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
-occurring in covert attention conditions.
-
-\subsubsection{Regularized spatiotemporal beamforming}
-Due to the decreased amplitude of the N1 and P3 components in covert attention
-settings~\cite{Treder2010}, the \ac{snr} of
-the \ac{erp} is lower than in overt attention settings.
-Therefore, a straightforward way to reach satisfactory
-gaze-independent decoding performance, might be by increasing overall \ac{erp}
-decoding performance. A more accurate classifier could yield relative
-improvements in those settings where performance is not yet near the achievable
-maximum.\todomvh{?? how to find out what is the maximum?}
-
-Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
-decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
-this classifier as a linear discrimination problem and
-imposing regularizing constraints by structuring the noise covariance matrix
-(STBF-struct).
-Furthermore, these regularizing constraints impose temporal stationarity on
-the background noise, yielding insights for our next\todomvh{next? more specificaly?}
-efforts to cope with the non-stationarity of the P3 signal component.
-\todofc{conclue with something like "this approach has been published in ... and
-will be described in chapter XXX}
-
-\subsubsection{Classifier-based Latency Estimation with Woody iterations}
-
-P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
-value is heavily dependent on the subject and the task, and can vary from trial
-to trial~\cite{Ouyang2017}.
-The work of \cite{Arico2014} illustrates that the variation in single-trial P3
-latencies is important in gaze-independent decoding and has been hampering covert
-VSA decoding performance.
-In this work we aim to reprise their hypothesis stating that jitter compensation improves
-covert VSA performance and extend it by developing a decoder and evaluating it
-in a broader range of gaze-independent settings, including split VSA.
-
-\todo{check overlap}
-
-Literature shows that better covert attention performance can be achieved by
-counteracting the non-stationarity of the P3 \ac{erp} component in covert
-attention settings~\cite{Arico2014}.
-This can be done by estimating the latency of each single-trial \ac{erp} and aligning
-the P3 peaks to all fall at the same moment.
-The resulting aligned data and the set of latencies can then be used to train a
-classifier that is more robust to jitter.
-
-Existing latency estimation methods are either not applicable to the
-classification problem of labeling unseen data, or are not robust enough to
-deal with the low \ac{snr} of the \ac{erp}.
-Classifier-based Latency Estimation (CBLE)~\cite{Mowla2017} is a technique that can leverage
-\ac{erp} latency estimation in a decoding setting, but our results show that it
-yields no improvement in gaze-independent settings.
-We improved upon this technique and extended it to a probabilistic, iterative
-method named Classifier-based Latency Estimation with Woody iterations (wCBLE).
-\todo{conclue with something like "this approach has been published in ... and
-will be described in chapter XXX}
-
-\subsubsection{Tensor discriminant analysis}
-\todo{todo}
-
-\subsection{Experiments}
-
-\subsubsection{Healthy control study}
-In a first series of experiments, we recorded data with this interface from healthy participants.
-The goal of these experiments is to benchmark
-gaze-independent \ac{erp} decoding algorithms.
-We denote this dataset as the Covert Visuospatial Attention \ac{erp} dataset
-(CVSA-\ac{erp})
-This study was approved by the Ethics Commission of University Hospital Leuven
-(S62547).
-To simulate the dissociation between the eye gaze (visual attention) and the
-intended target (mental attention) which could occur in patients with hampered eye
-motor control, healthy participants are cued to operate the \ac{bci} in specific
-\ac{vsa} conditions.
-\todofc{conclue with something like "this approach has been published in ...
-and will be described in chapter XXX}
-
-\subsubsection{Patient case study}
-\todofc{conclue with something like "this approach has been published in ...
-and will be described in chapter XXX}
-
-\todo{todo}
+\subsubsection{Case studies with gaze-impaired individuals}
+\todo{fill in}
 \todo{mention multi-center patient study case studies in cooperation
 After data collection from healthy participants and decoder development, we
-will carry out a patient study using a similar interface.
+have carry out a patient study using a similar interface.
 The goal of the patient study is to evaluate the influence of eye motor
@@
 }
-\todo{update with current results}
-
-\todo{refer to chapters}
+The results of this study are presented in chapter~\ref{sec:patients}.

### 2024-10-04 fig.tikz.tex
Update bttda formulation

@@
-\documentclass[tikz]{standalone}
-\usepackage{pgfplots}
+\documentclass{standalone}
+\usepackage{dissertation}
+\input{chapters/bttda/tensorviz.tex}
+
 \begin{document}
-\sffamily
-\footnotesize
-\begin{tikzpicture}
-    \begin{axis}[
-        width=\textwidth,
-        height=0.6180\textwidth,
-        scale only axis,
-        xlabel={Time (ms)},
-        ylabel={Potential ($\mu$V)},
-        axis x line=middle,
-	ylabel near ticks,
-	xlabel near ticks,
-        ymin=-5, ymax=8,
-        enlarge x limits=false,
-        enlarge y limits=false,
-        smooth,
-        every axis plot/.append style={very thick, accent1}, % Set the line color to accent1
-        ytick={-5,0,5}, % Y-ticks at -5, 0, 5
-        xtick={200,400,600}, % X-ticks at 200, 400, 600, 800
-        axis line style={darkgray}, % Set the axis line color to darkgray
-        tick style={darkgray}, % Set the tick color to darkgray
-        tick label style={darkgray} % Set the tick label color to darkgray
-    ]
-    % ERP waveform using cubic splines for smooth connection
-    \addplot[
-        smooth,
-        tension=0.5
-    ] coordinates {
-        (0, 0)     % Starting point
-        (50, 0)    % Pre-P1
-        (80, 2)    % P1
-        (115, -3)  % N1
-        (160, 3)   % P2
-        (220, 1)   % N2
-        (400, 7)   % P3
-        (650, 1)   % LPP (Late Positive Potential)
-        (800, 1)   % End point
-    };
+\begin{tikzpicture}[y=-1cm]
+    \draw[help lines] (-3,-3) grid (20,6);
+    \fill[red] (0,0) circle (2pt);
 
-    % Annotations for the components
-    \node[anchor=south, color=darkgray] at (axis cs:80,2) {P1};
-    \node[anchor=north, color=darkgray] at (axis cs:115,-3) {N1};
-    \node[anchor=south, color=darkgray] at (axis cs:160,3) {P2};
-    \node[anchor=west, color=darkgray] at (axis cs:220,1) {N2};
-    \node[anchor=south, color=darkgray] at (axis cs:400,7) {P3};
-    \node[anchor=south west, color=darkgray] at (axis cs:650,1) {LPP};
+	  \begin{scope}[shift={(-0.0707, -0.0707)}]
+      \MatrixSkewed{$\mat{A}_1$}{$r_1$}{$D_1$}{2}{1}
+    \end{scope}
+	  \begin{scope}[shift={(-0.1,0)}]
+      \MatrixLeft{$\mat{A}_2$}{$r_2$}{$D_2$}{2}{1}
+    \end{scope}
+	  \begin{scope}[shift={(0,1.1)}]
+      \MatrixBelow{$\mat{A}_3$}{$r_3$}{$D_3$}{1}{2}
+    \end{scope}
+    \TensorThree{$\ten{G}$}{}{}{}{1}{1}{1}
 
-    \end{axis}
+    \node at (2,.5){$=$};
+
+	  \begin{scope}[shift={(3,-.5)}]
+    \TensorThree{$\ten{X}^\prime$}{$D_1$}{$D_2$}{$D_3$}{2}{2}{2}
+    \end{scope}
+
+    \node at (6.5,.5){$=$};
+
+	  \begin{scope}[shift={(7.5,-.5)}]
+    \TensorThree{$\ten{X}$}{$D_1$}{$D_2$}{$D_3$}{2}{2}{2}
+    \end{scope}
+
+    \node at (11,.5){$-$};
+
+	  \begin{scope}[shift={(12,-.5)}]
+      \TensorThree{$\ten{\mathbfcal{E}}$}{$D_1$}{$D_2$}{$D_3$}{2}{2}{2}
+    \end{scope}
+
 \end{tikzpicture}

### 2024-10-06 chapters/bci.tex
cover page

@@
 \todo{acro usage in chapters}
-
+\chapter{Brain-computer interfaces}
+\label{sec:bci}
 People with \ac{sspi} due to
@@
 \subsubsection{\Acf{lda}}
-\Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
+\Ac{lda} is a more well-known method that is, in fact, nearly equivalent to
 \ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
@@
 (noise).
-$mat{w}$ can be obtained as eigenvector of
+$\mat{w}$ can be obtained as eigenvector of
 $\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.
@@
 \begin{equation}
-  \mat{S}_\text{w}^(c) = \begin{bmatrix}
+  \mat{S}_\text{w}^{(c)} = \begin{bmatrix}
 s_0 & s_1 & s_2 & \cdots & s_{k-1} \\

### 2024-10-06 chapters/bttda.tex
cover page

@@
+\chapter{Block-term tensor discriminant analysis}
+\label{sec:bttda}
 \input{chapters/bttda/tensorviz.tex}
+\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}}.
 
@@
 a neuroimaging technique, with electroencephalography (EEG) being the most popular.
+
+\subsection{Tensor \& tensor methods}
 
@@
 
+\subsection{Supervised tensor decomposition}
+
 If the decompositions are not full rank, the Tucker, PARAFAC and block-term
@@
 using Higher Order Discriminant Analysis
-(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fischer sense, analogous to linear
+(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fisher sense, analogous to linear
 discriminant analysis.
@@
 you should at least try to better describe the previous methods.}
+
+\subsection{A block-term structured model for classification}
 
@@
 Forward models are useful for a.o.\ interpretability and data compression.
-\todo{explicitly label error terms in equation, use $X\prime$ and error with
-iteration sup}
 
-The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
+The HODA projection in Equation~\ref{eq:HODA-backward} is an example
 of a backward model.
 A straightforward and computationally efficient candidate for a corresponding
-forward model is given by
+forward model, visualized in Figure~\ref{fig:HODA-forward} is given by
 \begin{equation}
@@
 \end{figure}
-\todo{not referred to in text}
-\todo{I would prefer a figure in which the Ai and G are on the left and
-$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
-use a standard equal signal. Finally, in the caption, you could clearly define
-the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
-\todo{explicitly state why choosing this versus full model or HOPLS}
 
-A forward model should make sure that reconstruction error is minimized and
-variation captured in the latent tensor is maximally captured by the
+A forward model should make sure that reconstruction error is minimized.
+In other words, variation captured in the latent tensor should maximally captured by the
 reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
-\todo{not clear ("error is minimized" in the same sentence)}
 $\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
@@
 \end{algorithm}
-\todo{Reorder figures and add error term}
+
+The forward model could alternatively be specified in the classical manner by linear
+regression or with the solution of \textcite{Haufe2014}, but this would require
+vectorizing the tensor representation.
+While the reconstruction error might be brought down, we lose the computational
+efficiency and regularizing constraints of the tensor form.
+More efficiently, the forward model could be obtained through other tensor
+regression methods, like Higher Order Partial
+Least Squares.
+This method still requires fitting more parameters than our
+proposed model, on top of additional hyperparameters defining the
+dimensionality of the common subspace.
+An attractive property of our multilinear forward model for HODA is that it
+estimates exactly as many parameters as the backward model, with the intuition that
+it should not disproportionally contribute to overfitting.
 
@@
 performance.
-\todo{eps1 = X - G1 x {A1} still contains discriminative info, so
-it is further projected onto another core tensor yielding
-eps2 = eps1 - G2 x
-{A2}
-etc. and finally you give equation (15)
-}
-We thus extend the HODA feature extraction scheme with backward and
-forward models defined in respectively Equations~\ref{eq:HODA-backward}
-and~\ref{eq:HODA-forward} to Block-Term Tensor Discriminant Analysis
+Useful features can still be extracted from $\ten{\mathbfcal{eps}} = \ten{X} -
+\hat{\ten{X}}$, so it is further projected onto another core tensor
+$\ten{G}^{(2)}$ (assuming $\ten{G}$ as $\ten{G}^{(1)}).$
+
+We thus extend the HODA feature extraction scheme to Block-Term Tensor Discriminant Analysis
 (BTTDA).
@@
 \end{figure*}
-\todo{Explicitly mention that if restricted to rank one, it is parafac, and to
-one block, it is hoda}
+The block-term structure of this model implies that is a generalization of both
+the Tucker-structured HODA and PARAFAC-structured discriminant feature
+extraction.
+If $B$ in Equation~\ref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
+HODA; if at each term $b$ the rank of the core tensor
+$(r_1^{(b)}=r_2^{(b)}=\ldots=r_k^{(b)}=1)$, a PARAFAC structure is assumed.
 
-Since BTTDA is specified as a forward model, a backward modelling
+Since BTTDA is specified above as a forward model, a backward procedure
 procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
@@
 For each block $b$, the latent tensor is extracted using the HODA backward
-projection in \todo{ref equation} from the residual error term of the previous
+projection in Equition~\ref{eq:HODA-backward} from the residual error term of the previous
 block $\ten{\mathbfcal{E}}^{(b-1)}$
@@
 univariate Fisher score for all features.
-The Fisher score $\phi{i}$ for feature $i$ is obtained as
+The Fisher score $\phi(i)$ for feature $i$ is obtained as
 \begin{equation}
@@
 reconstructed from the truncated BTTDA decomposition
-$\textstyle{\ten{X}_\text{rec}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
+$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
 NMSE is calculated as
 \begin{equation}
-	\nmse\left(\ten{X}, \ten{X}_\text{rec}^{(B)}\right) =
+  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
 	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
@@
 \section{Discussion}
-\todo{subtensors in larger tensor drawing, focus more on sparsity in the
-discussion}
-\todo{focus on tradeof: more flexibility in model, but this implies more
-hypermarameters so it relies more on a model selection procedure.}
+\subsection{Contribution}
 The results in Table~\ref{tab:erp-score} and Table~\ref{tab:mi-score} show that
@@
 being more parsimonious due to its block-term structure compared to HODA's full
-Tucker structure.
+Tucker structure, as illustrated by Figure~\ref{fig:bttda/sparse}.
+\begin{figure}
+  \input{figures/bttda/sparse.tikz.tex}
+  \caption[Sparsity of BTTDA and PARAFACDA]{BTTDA and PARAFACDA can find a
+  sparser expression of the discriminant information captured by the dense,
+  Tucker-structured HODA algorithm.}
+  \label{fig:bttda/sparse}
+\end{figure}
+The same discriminative information captured by a relatively large
+Tucker-structured core tensor, could be expressed more sparsely with a small
+number of block-terms, while avoiding redundant features.
+The PARAFAC structure employed in PARAFACDA is even more sparse, which could be
+a benefit or an advantage depending on the amount of regularization required,
+or on the true underlying structure of the data.
+
 Alternatively, the enhanced performance could also stem from the modeled data
@@
 
+\subsection{Model selection and dimensionality}
+
+BTTDA trades in the rigid HODA model for increased model complexity with more
+hyperparameters to tune, which could open up a setting where performance can be
+improved.
+This turns tensor discriminant analysis into a model selection procedure
+instead of a projection optimization procedure.
 Despite favorable results in BCI decoding, the applications of the proposed
-BTTDA model are limited mainly by the model selection approach used
+BTTDA model are limited mainly by this model selection approach used
 to determine the individual block ranks.
@@
 BTTDA.
+
 Due to its heuristic nature, the greedy algorithm does not always result in the
@@
 correlation, decreasing discriminability etc.
-Future efforts should focus on automatic parameter setting, e.g. using
-information criteria such as the ones used in BTTR~\cite{Faes2022} or other
+Finally, the presented results are limitation that by our artificial restriction
+that $(r_1=r_2=\ldots=r_k)$ to reduce computational demand of the presented
+experiments.
+In a sense, this goes against the proposition of increased model flexibility.
+It might well be that BTTDA in some cases offers little added value over the
+Tucker-structured HODA, when the both are given free choice of rank.
+Future efforts should focus on automatic parameter setting, e.g. using sparsity
+or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
 statistical measures based on the models application.
-\todo{comment as limitation that we test r1=r2=r3, this is against the whole
-idea of increased flexibility, find automated rank selection procedure}
 
@@
 of interest if they are appropriately chosen based on prior knowledge of the dataset.
-\todo{mention in discussion of model selection procedure that sparser blocks
-would be better}
 
 \section{Conclusion}
-\todo{main restriction for hoda is model selection. this is also a model
-selection problem, but it is more flexible}
+
 We have introduced Block-Term Tensor Discriminant Anlysis (BTTDA) a novel,
@@
 leveraging a novel forward modeling step.
+
 Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
@@
 projections to model and feature selection.
-This allows performance to be traded
-off for model complexity and number of features, to find a setting that is more
-effective for decoding.
-Because of the flexibility and minimal assumptions BTTDA
-can equally be applied to other neuroimaging modalities (MEG, ECoG, fNIRS,
-fMRI, EMG, ...) or tensor classification problems in other fields.
 
-\section*{Acknowledgements}
-\todo{move to the end}
-We thank the Flemish Supercomputer Center (VSC) and the High-Performance
-Computing (HPC) center of the KU Leuven for allowing us to execute our
-computational experiments on their systems.
-We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
-work.
+Introducing a flexible block-term tensor model as underlying structure
+reformulates tensor discriminant analysis as a model selection
+problem.
+This allows performance to be traded off for model complexity and number of
+features, to find a setting that is more effective for decoding.
+Because of its general implementation and minimal assumptions on data structure,
+BTTDA can equally be applied to classification for other neuroimaging modalities
+(MEG, ECoG, fNIRS, fMRI, EMG, ...) or to tensor classification problems in other
+fields.
+
+%\section*{Acknowledgements}
+%\todo{move to the end}
+%We thank the Flemish Supercomputer Center (VSC) and the High-Performance
+%Computing (HPC) center of the KU Leuven for allowing us to execute our
+%computational experiments on their systems.
+%We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
+%work.

### 2024-10-06 chapters/conflict.tex
cover page

\chapter*{Conflict of interest}
The authors declare no conflict of interest.

### 2024-10-06 chapters/contribution.tex
cover page

--- 2024-09-28 include/document.tex
@@
-\begin{document}
-
-% Front matter ================================================================
-\frontmatter
-
-\include{chapters/preface.tex}
-\chapter{Abstract}
-\include{chapters/abstract.tex}
-\chapter{Samenvatting}
-\tableofcontents
-\listoffigures
-\listoftables
-\printacronyms[name=List of acronyms]
-
-% Main matter =================================================================
-\mainmatter%
-
-\part{Introduction \& Literature}
-\includechapter{Brain-computer interfaces}{chapters/bci.tex}
-\includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
-\todo{The role of jitter in gaze-independent ERP decoding}
-%\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}
-
-\part{Algorithms \& Decoders}
-\includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}
-
-\chapter{Tensor discriminant analysis}
-\todo{insert tensor paper}
-\todo{discuss Tensorizations coping with latency jitter to make link}
-
-\includechapter{Latency estimation methods}{chapters/wcble.tex}
-
-\part{Experiments \& Applications}
-\includechapter{Compensating jitter for gaze-independce}{chapters/covert_align.tex}
-\todo{Integrating gaze-tracking data}
-\includechapter{Patient cases}{chapters/patients.tex}
-\includechapter{Conclusion \& recommendations}{chapters/conclusion.tex}
-
-\backmatter%
-\chapter{Funding \& acknowledgments}
-\todo{my grant}\todo{marc's projects}
-\chapter{Personal contribution}
+\chapter*{Personal contribution}
+\todo{which works conceptualized}
 \todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
 Hakim, Marie,Etienne, Philip Van Damme  }
-\chapter{Conflict of interest}
-The authors declare no conflict of interest.
-\include{chapters/cv.tex}
-
-\end{document}

### 2024-10-06 chapters/funding.tex
cover page

\chapter*{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}

\todo{acknowledgement wcble paper}
We thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.
We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in Chapter~\ref{sec:bttda}.

\todo{people at TRAINM, MAS and Kristl Claeys}
\todo{juliette meunier}

### 2024-10-06 chapters/gaze_independence.tex
cover page

@@
+\chapter{Gaze-independent visual \acsp{bci}}
+\label{sec:gaze-independent}
 \epigraph{%
@@
 
+\subsection{Incidence of eye motor impairment}
 \Ac{sspi} is often caused by damage to the central or peripheral nervous
@@
 
+\subsection{Gaze impairment and \acl{lis}}
+
 \newcommand\fnlis{\footnote{
@@
 We believe it is in general not opportune to delve too deep in the underlying
-ophthalmological and neurological pathophysiology for each of the specific
-conditions.
+ophthalmological and neurological mechanisms for each of the specific
+conditions, or even the exact symptoms.
 In clinical reality, every patient different symptoms, which lead to their own
-unique set of preserved capabilities.
+unique set of preserved capabilities and visual skills.
 From a solution-oriented \ac{bci} engineering point
 of view, the etiology of the symptoms can be abstracted away.
-Instead of categorizing patients we will refer to these groups as individuals
-with \ac{sspi} and, if eye motor skills are affected,  \ac{sspgi}, in line with
-the terminology used by~\textcite{FrieDOken2020} (see
-Figure~\ref{fig:gaze_independence/venn}).
+Instead of categorizing patients we will refer to those that might benefit from
+\ac{bci} assistive communication technology as with \ac{sspi}, in line with
+the terminology used by~\textcite{FrieDOken2020}.
+If they have eye motor impairment that affects their use of visual \ac{bci} or
+eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).
 
@@
 \section{State-of-the-art}
-\label{sec:gaze-independence}
+\label{sec:gaze-independence/sota}
 
@@
 visual oddball \ac{bci}.
-Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
+Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms form
 a good candidate for a performant gaze-independent \ac{bci}.

### 2024-10-06 chapters/patients.tex
Rewrite sota
Update bttda formulation
cover page

@@
+\chapter{Case studies}
+\label{sec:patients}
 \section{Introduction}
@@
 relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
-These include visual acuity, visual fixation, eyelid function, ocular motility
-, binocuar vision and field of vision.
+These include visual acuity, visual fixation, eyelid function, ocular motility,
+binocular vision and field of vision.
 Additionally, participants were asked about eye tremors (nystagmus or other) and
@@
   \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
+}{%
+  Distribution of the recorded gaze position.
 }{%

### 2024-10-06 chapters/wcble.tex
Resturcture chap 1 and 2
Update bttda formulation
cover page

@@
+\label{sec:wcble}
 \todo{comments liesel}
-Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
-by~\textcite{VanDenKerchove2024}
+\emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
+as part of~\textcite{VanDenKerchove2024}}
 
@@
 individuals by decoding \acp{erp} obtained from time-modulated flashing
-stimuli.
+stimuli (see Figure~\ref{fig:bci/loop}).
 These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
-\ac{erp} waveform.\todomvh{refer to image in bci chapter}
+\ac{erp} waveform (see Figure~\ref{fig:bci/erp}.
 These components each have their own distinct amplitude and latency.
@@
 neurophysiological and clinical properties, that can be confounding factors in
-the development of \acp{bci}.\todo{cite}
+the development of \acp{bci}~\cite{Luck2014}.
 Latency is often assessed as the timing of an \ac{erp} component peak, obtained
@@
 performance, and disproportionally so when the user is not directly gazing at
-intended targets, such as is the case for gaze-impaired patients.
+intended targets, such as is the case for individuals with \ac{sspgi}.
+Therefore, properly accounting for \ac{erp} latency jitter can contribute to
+the development of gaze-independent visual \acp{bci}.
 
-\subsection{\Acs{erp} latency jitter}
-
-\subsubsection{Neural origins of latency jitter}
+\subsection{Neural origins of \ac{erp} latency jitter}
 While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
@@
 \cite{Stokes2016,Mouraux2008}.
-\todo{substantiate}
+If these processes are related to the \ac{bci} task at hand, such as attention
+and visual processing, jitter could have a significant impact on \ac{erp}
+decoding.
 
-\subsubsection{Implications for \ac{erp} analysis}
+\subsection{Implications for \ac{erp} analysis}
 
@@
 results and making inferences about amplitude or latency effects, because these
-are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
-	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
-	these references}.
+are entangled when using averaging~\cite{Luck2014}.
 
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 \end{figure}
-
-\todo{illustrate smearing effect}
 
@@
 
-\subsubsection{Benefits of latency estimation}
-A common approach to counteract the smearing effect is to correct single-trial
-\acp{erp} for jitter before averaging.\todo{reference}
+\subsection{Benefits of latency estimation}
+As we will see later, a common approach to counteract the smearing effect is to correct single-trial
+\acp{erp} for jitter before averaging.
 This can be done by estimating the latency of each
@@
 latency estimation method that is still being used and is even
-considered amongst the more performant techniques for \acp{erp}~\cite{Ouyang2017}.
+considered amongst the more performant techniques for
+\acp{erp}~\cite{Thornton2008, Ouyang2017}.
 It starts with an average \ac{erp} as a rough estimation of the aligned template
@@
 also be used~\cite{Gasser1983}.
-\todo{cite Thornton2008 hierarchical woody?}
 
@@
 \label{sec:wcble/literature/contrib}
-
-\todo{cite Kim2020 used to analyze the difference between target and
-non-target}
 
@@
 
-\subsection{Robust \ac{cble} latency features}
+\subsection{Robust latency features}
 \label{sec:robust-latency}
@@
 }
-\todo{Pr not italic}
 \todo{re-insert comments}
@@
         \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
 		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
     \EndFor
@@
         \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
 		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
     \EndFor
@@
     \State  $s_n \gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
 		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
@@
         \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
 		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
     \EndFor
@@
         \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
 		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
     \EndFor
@@
 
-\begin{figure}[!ht]
-  \input{figures/wcble/fig_latency.pgf} \\
+\begin{figure}[H]
   \input{figures/wcble/fig_accuracy.pgf}
@@
 \end{figure}
+
 \todo{properly label axes}
@@
 \subsection{Classification}
-\todo{refer to figure}
+
 Analogous to the previous experiment, 10-fold cross-validated classification
@@
 \todo{ref figure}
+
+\begin{figure}
+  \input{figures/wcble/fig_accuracy.pgf}
+  \caption{caption}
+  \label{fig:wcble/results/accuracy}
+\end{figure}
 

### 2024-10-06 cover_kul.tex
cover page

\newgeometry{margin=1.5cm}
\begin{titlepage}

    \sffamily
    \begin{minipage}[t]{.3\textwidth}
    \vspace{0pt}
    KU Leuven \\
    Biomedical Sciences Group \\
    Faculty of Medical Science \\
    Department of Neurosciences

    \end{minipage}\hfill%
    \begin{minipage}[t]{.3\textwidth}
      \vspace{0pt}
      \includegraphics[width=\textwidth]{figures/kul_docschool.png}
      \bigskip

      \includegraphics[width=\textwidth]{figures/ulille.png}
    \end{minipage}

    \vfill

    \noindent\makebox[\textwidth][c]{%
      \begin{minipage}{.8\textwidth}
        \centering
        {%
          \Huge
          \bfseries
          A visual Brain-Computer Interface for gaze-free communication
        }
        \bigskip

        \MakeUppercase{%
          \Large
          Algorithms \& Applications
        }
        \bigskip

        {%
          \Large
          Arne  Van Den Kerchove
        }

      \end{minipage}%
    }

    \vfill
    \begin{minipage}[b]{.5\textwidth}
      Jury: \\
      Supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
      Supervisor University of Lille: \emph{prof. François Cabestaing} \\
      Chair examining committee: \emph{prof. John Creemers} \\
      Chair public defence: \emph{prof. Koen Poesen} \\
      Jury members: \\
      \emph{prof. Maarten De Vos} \\
      \emph{prof. Adalberto Simeone} \\
      \emph{prof. Andrea K\"ubler} \\
      \emph{prof. Fabien Lotte} \\
      \emph{prof. Reinhold Scherer}
    \end{minipage}\hfill%
    \begin{minipage}[b]{.31\textwidth}\raggedright
        Dissertation presented in partial fulfillment of the requirements for
        the double degree of \emph{Doctor of Biomedical Sciences} (KU Leuven)
        and \emph{Doctor of Control Science and Signal Processing} (University
        of Lille)
    \end{minipage}
    \bigskip

    \centering\today

\end{titlepage}
\restoregeometry

### 2024-10-07 chapters/bci.tex
funding

@@
 \chapter{Brain-computer interfaces}
+\todo{are textcites used when necessary?}
 \label{sec:bci}
@@
 \subsection{Decoders}
-\todo{averaging over trials}
+Due to the unfavorable \ac{snr} of \acp{erp}, it is often not feasible to
+accurately decode the encoded information based on just one example.
+Therefore, classification algorithm are trained on a set of epochs, but are
+often evaluated on an average of multiple epochs.
+These averages are constructed over multiple stimulations of the stimulus
+representing a given selection, enhancing the \ac{snr}.
+The drawback, however, is that performing these multiple stimulations takes
+more time.
+To optimize the \ac{itr}, the trade-off between number of repetitions,
+stimulation speed and decoding performance must be carefully balanced.
+Performant classifiers can shift this balance to the benefit of \ac{itr}.
 Several techniques are suited to decode~\acp{erp}.

### 2024-10-07 chapters/conclusion.tex
cover page
funding

@@
+\chapter{Conclusion \& recommendations}
 \todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
@@
 
-\subsection{Working \emph{with} individuals \acs{sspi}}
+flexible models
+
+kron-sum is more flexible than kron, tlda even more flexible
+
+similarly, bttda is more flexible than hoda
+
+regularization is absolutely necessary, and when faced with few data relative
+to dimensionality, good first instinct to pick a simple, restrictive structure.
+
+neural data is not linear, everything is fuzzy
+when optimizing for performance, it will soon become clear that there are
+settings where these assumptions can be broken to better describe the ongoing
+ineractions.
+if possible, in doing this, embue as much with prior information about the problem
+
+\subsection{Working \emph{with} individuals with \acs{sspi}}
 
@@
 
-\subsection{Working with patients}
 gaze independent visual, can be good current existing solution

### 2024-10-07 chapters/covert_align.tex
Started patients paper
Comments hakim
Comments Marc
Update bttda formulation
cover page
funding

@@
+\chapter{Compensating \acs{erp} jitter for gaze-independence}
+\label{sec:covert-align}
 \emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}}
@@
 \section{Introduction}
-\todo{mention cble and wcble in methods}
+
+The decoding methods described in \autoref{sec:stbf-struct} and
+\autoref{sec:bttda} are rather general attempts at increasing \ac{erp} decoding accuracy.
+Given the problem statement of this work, we are also interested in targeted
+methods that would specifically advance the development of gaze-independent
+visuall \acp{bci} (see \label{sec:gaze-independence/sota}).
+Methods of directly enhancing gaze-independent decoding through optimizing
+decoder design have not been widely studied (see
+\label{sec:gaze-independence/sota/decoding}).
+We assume that advances in gaze-independent decoding can be made if properties
+of covert \ac{vsa} that affect performance can be identified and, in some way,
+accounted for.
+The latency estimation and alignment approach proposed in \autoref{sec:wcble}
+was specifically designed for this purpose.
 
 \textcite{Arico2014} observed higher variability in single-trial P3 peak
-latencies relative to stimulus onset during covert VSA compared to overt VSA.
+latencies relative to stimulus onset during covert \ac{vsa} compared to overt
+\ac{vsa}.
 This latency variability contributes to reduced covert VSA decoding performance.
@@
 features for these latencies.
-\todo{Should we already mention Frenzel here?}
-\textcite{Frenzel2011} introduced a similar protocol to our split VSA
-setting.
+
+Furthermore, as mentioned in~\autoref{sec:gaze-independence/sota/visual} and
+\autoref{sec:gaze-independence/approach/cvsa-erp},
+our goal is also to explore more flexible gaze-independent settings that the
+usual covert \ac{vsa} with central gaze fixation reported in most literature.
+\textcite{Frenzel2011} introduced a similar protocol to the proposed split
+\ac{vsa} setting (see~\autoref{fig:gaze/vsa/free}).
 They showed that it is possible to perform split VSA and that, in this
@@
 decoding the visuospatial attention only.
-\todomvh{avoid overlap. it seems already explained in previous sentences. The reader can find details in
-  [6].}
-
-frenzel
-\todo{part about split attention etc from chapter on covert alignment to here}
 
@@
 \subsection{Data collection}
-We evaluate our approach on the
-publicly available BNCI2014-009 dataset~\cite{Aloise2012a} and one specifically recorded for this study, designed to
-probe different modalities of covert VSA.
-\todomvh{first explain BNCI2014-009 using fig. 6.1 or revert the announcement
-in the previous sentence.}
+We evaluate our approach on a dataset specifically recorded for this study, designed to
+probe different modalities of covert VSA, and on the publicly available
+BNCI2014-009 dataset~\cite{Aloise2012a}.
 
 \subsubsection{CVSA-ERP dataset}
+\label{sec:covert-align/stimulation}
 
@@
 
-	\caption{
+  \caption[Stimulation interfaces and evoked \acp{erp}.]{
 		\subref{fig:interface} Interfaces and visuospatial attention (VSA)
@@
 conforming to the interface proposed by \textcite{Treder2010}
-(Figure~\ref{fig:layout_targets})\todo{figure interface in four panels and
-update all refs below}.
+(\autoref{fig:covert-align/interface/interface})
 A hexagonal layout interface with an empty center and a low number of targets
@@
 Targets are full-contrast white and were intensified by scaling them to a
-larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
+larger size (5.60° of visual angle,
+(\autoref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
 experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
@@
 Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.
-\todo{figures}
+
+\begin{figure}
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/covert_align/interface_1.tikz.tex}
+    \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
+    \acs{bci}~\cite{Treder2010}.}
+    \label{fig:covert-align/interface/interface}
+  \end{subfigure}\hfill%
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/covert_align/interface_2.tikz.tex}
+    \caption[Intensification.]{When a target is intensified, it is enlarged for
+    a very short time.}
+    \label{fig:covert-align/interface/intensification}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/covert_align/interface_3.tikz.tex}
+    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
+    settings, the
+    participant is cued to fixate on one of the targets.}
+    \label{fig:covert-align/interface/overt}
+  \end{subfigure}\hfill%
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/covert_align/interface_4.tikz.tex}
+    \caption[Gaze fixated centrally]{In the covert \ac{vsa} setting, the participant is
+    cued to fixate the center of the interface.}
+    \label{fig:covert-align/interface/covert}
+  \end{subfigure}
+\end{figure}
 
@@
 was instructed to fixate their gaze on the blue crosshair for the duration of
-the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
+the stimulation block (\autoref{fig:covert-align/interface/overt} and
+\autoref{fig:covert-align/interface/overt}).
 The position of this crosshair determined the VSA condition for this trial:
@@
 We will compare \ac{wcble} and \ac{cble} as described in
-Section~\ref{sec:wcble/methods}, with their base classifier and tLDA and with
+\autoref{sec:wcble/methods}, with their base classifier and tLDA and with
 Riemannian Geometry
@@
 	\end{subfigure}
-	\caption{
+  \caption[Decoder performance.]{%
 		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
@@
 evaluated VSA conditions.
-Table~\ref{tab:single_trial_full} reports all single-trial classification scores for all considered models, datasets
+Table~\ref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets
 and conditions.
-\todo{table?}
+
+\begin{table}
+  \centering
+  \makebox[\textwidth][c]{%
+    \input{chapters/covert_align_performance.tex}
+  }
+  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
+curve for all evaluated models, visuospatial attention conditions and
+  datasets.}
+  \label{tab:covert-align/results/single-trial}
+\end{table}
 
@@
 
-	\caption{
+  \caption[Cross-condition classifier performance and estimated jitter.]{
 		\subref{fig:covert_cross_eval},\subref{fig:aloise2012_cross_eval}

### 2024-10-07 chapters/funding.tex
funding

@@
 
-\todo{acknowledgement wcble paper}
+Arne Van Den Kerchove is supported by the special research fund of the KU Leuven
+(GPUDL/20/031).
+prof. Marc Van Hulle is supported by research grants received from
+the European Union’s Horizon Europe Marie Sklodowska-Curie Action program
+(grant agreement No. 101118964), the European Union’s Horizon 2020 research and
+innovation program (grant agreement No. 857375), the special research fund of
+the KU Leuven (C24/18/098), the Belgian Fund for Scientific Research – Flanders
+(G0A4118N, G0A4321N, G0C1522N), and the Hercules
+Foundation (AKUL 043).
+
 We thank the Flemish Supercomputer Center (VSC) and the High-Performance
@@
 computational experiments on their systems.
-We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
+
+We also thank,  Arno Libert,
+Tjaša Mlinarič, Yide Li and Mani Mirsaeedi for their help in data collection,
+and in general the colleagues at the KU Leuven Computational Neurosciences
+group at KU Leuven and the BCI team at the University of Lille.
+Furthermore, we thank the support teams at TRAINM Neuro Rehab Clinics Antwerp,
+the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
+Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
 the work presented in Chapter~\ref{sec:bttda}.

### 2024-10-07 chapters/gaze_independence.tex
funding

@@
 \ac{bci} assistive communication technology as with \ac{sspi}, in line with
-the terminology used by~\textcite{FrieDOken2020}.
+the terminology used by~\textcite{FriedOken2020}.
 If they have eye motor impairment that affects their use of visual \ac{bci} or
@@
 \subsection{Gaze-independent visual stimulation}
+\label{sec:gaze-independence/sota/visual}
 This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
@@
 \subsection{Gaze-independent decoding}
+\label{sec:gaze-independence/sota/decoding}
 Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
@@
 \subsubsection{The CVSA-ERP dataset}
+\label{sec:gaze-independence/approach/cvsa-erp}
 
@@
 electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
-The Hex-o-Spell was chosen since it is optimiazed for gaze-independent
+The Hex-o-Spell was chosen since it is optimized for gaze-independent
 performance and because it has already been tested in individuals with
@@
 The interface can be operated in different VSA conditions as illustrated in
-Figure~\ref{fig:gaze-vsa}.
+Figure~\ref{fig:gaze/vsa}.
 
@@
     }
-    \label{fig:gaze/vsa/overt}
+    \label{fig:gaze/vsa/split}
   \end{subfigure}\hfill%
@@
     }
-    \label{fig:gaze/vsa/overt}
+    \label{fig:gaze/vsa/free}
   \end{subfigure}\hfill%
@@
   }
-  \label{fig:gaze-vsa}
+  \label{fig:gaze/vsa}
 \end{figure}

### 2024-10-07 chapters/patients.tex
funding

@@
 \section{Introduction}
-\todo{rename patient to bci user and participant}
-\todomvh{separate terminology for sspgi}
-\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}
-
-\todo{A systematic review of research on augmentative and alternative communication brain-computer interface systems for individuals with disabilities}
 
 \Acp{bci} for communication assistive technology~\cite{Millan2010}
-target individuals with \ac{sspi}.
+target individuals with \ac{sspi}~\cite{Peters2022}.
 
-visual bci
+\todo{visual bci}
 
@@
 the ability to use visual BCIs. These impairments make it difficult for
-individuals to focus on or track visual stimuli accurately, reducing their
+\ac{bci} users to focus on or track visual stimuli accurately, reducing their
 performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
@@
 Eye motor impairments are presumed to reduce performance in operating visual
-oddball \ac{bci}s (see section\todo{ref section} for an overview), since users
+oddball \ac{bci}s (see \autoref{sec:gaze-independence} for an overview), since users
 cannot comfortably redirect their gaze at the desired target,
@@
 gaze-independent solutions.
-These studies build on the assumption that patients with eye motor impairment
+These studies build on the assumption that \ac{bci} users with \ac{sspgi}
 would feel comfortable operating an interface in pure covert \ac{vsa} with
@@
 of participants during the experiment.
-It was suggested that patients were performing overtly.
+It was suggested that participants were performing overtly.
 Participants with \ac{als} also had a substantially lower accuracy than healthy
@@
 
-Our previous study, presented in chapter \todo{ref chapter} also used the
+Our previous study, presented in chapter \autoref{sec:covert-align} also used the
 visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
-This work partially accounted for the idea that gaze impaired patients might
+This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
 not fully rely on central gaze fixation and evaluates settings that are not
@@
 
-\todo{discuss \cite{FriedOken2020}}
+%\todo{discuss \cite{FriedOken2020}}
 
@@
 \begin{enumerate*}
-  \item Explore capabilities and experienced comfort of eye motor impaired
-    patients,
-  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
+  \item Explore capabilities and experienced comfort of individuals with
+    \ac{sspgi}, when operating a visual \ac{bci},
+  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
+    group,
   \item verify if this performance can be improved with a suitable decoding
@@
 Patients were recruited based on the following criteria.
-To qualify for inclusion, patients must
+To qualify for inclusion, participants must
 \begin{enumerate}
@@
 	years,
-  \item belong to class 2 or 3 according the \ac{bci}	patient selection criteria
+  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
     presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
@@
 
-In total, 11 patients were contacted, of which 1 \ac{ms} patient was excluded based on
-criterion~\ref{item:patients/exclusion/vision}, 1 \ac{tbi} patient on
+In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
+criterion~\ref{item:patients/exclusion/vision}, one person recovering from
+\ac{tbi} on
 both~\ref{item:patients/exclusion/epilepsy}
-and~\ref{item:patients/exclusion/cognitive}, and one stroke patient based
+and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
+stroke based
 on~\ref{item:patients/exclusion/medical}
-One further stroke patient was excluded due to technical
+One further person recovering from a stroke was excluded due to technical
 difficulties during the experimental session.
@@
 
-Ultimately, 7 patients were retained.
-Of these, one patient was diagnosed from bulbar-onset \ac{als}.
+Ultimately, 7 participants were retained.
+Of these, one participant was diagnosed from bulbar-onset \ac{als}.
 \ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
@@
 This initially results in general weakness and loss of muscle tone, but
-eventually results in full body paralysis.\todo{cite}
+eventually results in full body paralysis.
 Although speech and especially eye movements are usually preserved until the
@@
 characterized by an early loss of speech and an increased involvement of eye
-motor symptoms~\cite{Guo2022}\todo{which?}.
-Furthermore, one of the goals of \ac{bci} is to support \ac{als} patients whose
-life span has been extended with life support with and assistive technology to
+motor symptoms~\cite{Guo2022}.
+Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
+whose life span has been extended with life support with and assistive technology to
 ensure quality of life.
@@
 
-Three other patients were diagnosed with \ac{fa}, a neurodegenerative
+Three other participants were diagnosed with \ac{fa}, a neurodegenerative
 disease affecting the
@@
 
-The final three patients were stroke patients.
+The final three participants were recovering from stroke.
 Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
@@
 
-Table~\ref{tab:patients/patients} lists the included patients and their
+Table~\ref{tab:patients/patients} lists the included participants and their
 diagnoses.
 
-\todo{check Usability and Workload of Access Technology for People With Severe
-Motor Impairment: A Comparison of Brain-Computer Interfacing and Eye Tracking}
-
+\newcommand{\fnkuebler}{%
+  \footnote{
+``With minor degree of impairment,
+we refer to patients who had only slightly impaired limb movement
+and normal speech. Under the category moderate impairment, we
+summarized those patients with restricted limb movement (wheelchair-bound) and
+unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
+and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
+and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
+  }
+}
 \begin{table}[t]
   \centering
+  \makebox[\textwidth][c]{%
   \footnotesize
   \input{chapters/patients_table.tex}
-  \caption[Presentation of included patients including their diagnosis and
-  capabilities.]{Presentation of included patients including their diagnosis and
+  }
+  \caption[Included participants with their diagnosis and
+  capabilities.]{Included participants with their diagnosis and
   capabilities.
-  (Trach.: patient underwent a tracheostomy, Cls.: classification according
-  to~\textcite{Wolpaw2006}).
+  Trach.: underwent a tracheostomy, Cls. W: classification according
+  to~\textcite{Wolpaw2006}, Cls. KB: classification according
+  to~\textcite{Kuebler2008}\fnkuebler.
   }
@@
 This was not the case.
-This test is based on calibration-free eye tracking.
-\todo{say which part we used and which part not}
-\todo{motivation and text and references Juliette}
-\todo{Limited in what it can tell us (check paper)}
-\todo{include eye test document juliette}
 
 Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
-included patients.
+included participants.
 All participants reported some degree of fatigue or discomfort when fixating.
@@
 for prolonged times.
-The \ac{fa} patients were mostly affected by eye tremors and impaired pursuit.
+The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
+PB2 suffered from especially severe horizontal oscillating involuntary eye
+movements.
 Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
@@
 and vision in the right eye, and suffered reduced motility in the left.
-\todo{go through notes again}
 
@@
   Visual~\ac{bci} skills~\cite{FriedOken2020}
-  was assessed with a combination of self-reported issues
+  were assessed with a combination of self-reported issues
   by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
-  impaired, \BigDiamondshape\ severely impaired. Visual acuity in logMAR, lower
-  is better.}
+  impaired, \BigDiamondshape\ severely impaired.
+  logMAR: lower is better.}
   \label{tab:patients/eye}
@@
 3 different \ac{vsa} settings were explored
-In overt \ac{vsa}, the participant was instructed to fixate the cued target, or
+In the overt \ac{vsa} setting, the participant was instructed to fixate the cued target, or
 try to the maximum extent of their visual skill, even if experiencing slight
 discomfort.
-In covert \ac{vsa}, the participant was instructed to fixate the center of the
+In the covert \ac{vsa} setting, the participant was instructed to fixate the center of the
 screen, also to the extent of their ability.
-An additional \emph{free \ac{vsa}} condition was introduced.
-Here, the patient was instructed to perform the task as they deemed most
+An additional \emph{free \ac{vsa}} setting was introduced.
+Here, the participant was instructed to perform the task as they deemed most
 comfortable.
@@
 number of blocks was decreased to 6 per \ac{vsa} setting.
-\Ac{isi} was increased to\todo{number} to decrease task difficulty.
+\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
 The experiment also started with a training block in each condition, where the
@@
 \begin{figure}
-  \begin{subfigure}{.45\textwidth}
+  \begin{minipage}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}
-    \caption{A patient with the stimulation and recording setup.}
-    \label{fig:patients/photos/side}
-  \end{subfigure}\hfill%
-  \begin{subfigure}{.45\textwidth}
+  \end{minipage}\hfill%
+  \begin{minipage}{.45\textwidth}
     \includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}
-    \caption{caption}
-    \label{fig:patients/photos/front}
-  \end{subfigure}
-  \caption{The recording setup and stimulation interface.}
-  \label{fig:patients/photos}
+  \end{minipage}
+  \caption{A participant with the stimulation and recording setup.}
 \end{figure}
-\todo{caption}
-\todo{include stimulation interface}
 
@@
 
-During the recording session, patients were positioned in their wheelchair in front of a table.
-Stimuli were presented on a laptop \todo{model, make} placed at \todo{distance}
-A Cedrus StimTracker \todo{make} ensured synchronization of stimuli with the
+During the recording session, participants were positioned in their wheelchair in front of a table.
+Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
+Inc., Taiwan) placed at a 60cm distance.
+A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
 recorded \ac{eeg}.
@@
 
-First, we aimed to shed more light on the actual capabilities of gaze impaired
-individuals with respect to performing overt \ac{vsa} and central gaze
+First, we aimed to shed more light on the actual capabilities of individuals
+with \ac{sspgi} with respect to performing overt \ac{vsa} and central gaze
 fixation, as well as investigate how relevant these two settings are when the
@@
   conditions.
+  Crosshairs represent stimulus positions, with the orange ones cued during
+  the given condition.
   Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
@@
 }
-\todo{explaine crosses}
 
@@
     single-trial \ac{rocauc}.
-    Free \ac{vsa} is generally on par with overt \ac{vsa}.
-    Performance in covert \ac{vsa} with central gaze fixation is lower, but can
+    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
+    settign.
+    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
     be improved with the \ac{wcble} decoder.
@@
 \end{figure}
-\todo{rename subjects}
-\todo{check if percentages or 0-1 accuracy and roc-auc}
+\todo{figure margin}
+\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}
 
-In overt \ac{vsa}\todomvh{specify setting when talking about the setting}, evaluated decoders performed about equal on average
+In the overt \ac{vsa} setting, evaluated decoders performed about equal on average
 (\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
-In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
+In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorates
 but \ac{wcble} can significantly improve performance over its base classifier
@@
 More interestingly, we notice that performances of the decoders in free
-\ac{vsa} are close to those in overt \ac{vsa}.
-A substantial decrease in performance from overt to free is present for
+\ac{vsa} are close to those in the overt \ac{vsa}.
+A substantial decrease in performance from the overt setting to the free
+setting is present for
 subjects PC3 and PC4.
@@
 
-Due to the heterogenous nature of the participant's conditions, it is hard to support
-claims about significance of presented results.\todomvh{
-weird formulation
-hard to draw general conclusions}
+Due to the heterogenous nature of the participant's conditions, it is hard to
+draw general conclusions.
 This study should therefore be seen as a collection of case studies,
 highlighting different obstacles encountered in developing gaze-independent
-visual oddball \acp{bci} for individuals with \ac{sspi} and gaze impairment.
-Nevertheless, we would like to highlight some general aspects that might be
+visual oddball \acp{bci} for individuals with \ac{sspgi}.
+Nevertheless, we would like to highlight some aspects that might be
 of interest for the further development of this class of \acp{bci}.
 
-True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.\todomvh{any reference/evidence in support of this statement?}
+True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
 Hence our analysis centers around the free \ac{vsa} condition.
@@
 confirm our assumption that voluntary covert \ac{vsa} can
-occur in individuals with \ac{sspi} and gaze impairment.
+occur in individuals with \ac{sspgi}.
 We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
@@
 fixation can be improved by accounting for latency jitter, and showed that this
-holds for individuals with \ac{sspi} and gaze impairment.
+holds for individuals with \ac{sspgi}.
 
@@
 misinterpreted as an indication that the Hex-o-Spell \ac{bci} already works
-well for patients with \ac{sspi} and gaze impairment, and no optimization is
+well for individuals with \ac{sspgi}, and no optimization is
 needed.
@@
 perform the task.
-Nevertheless, the large difference between free \ac{vsa} and covert \ac{vsa}
+Nevertheless, the large difference between the free \ac{vsa} setting and the
+covert \ac{vsa} setting
 with central gaze fixation, is food for thought about the applicability of
@@
 
-First and foremost, this study works with a limited sample of patients which
-does not represent the full spectrum of individuals with \ac{sspi} and their
-oculomotor impairment.
-\Ac{fa} patients met the inclusion criteria, but they are usually not
+First and foremost, this study works with a limited sample size which
+does not represent the full spectrum of individuals with \ac{sspi} and
+\ac{sspgi}, and their specific symptoms and skills?
+Individuals with \ac{fa} met the inclusion criteria, but they are usually not
 considered as one of the typical interest groups of \ac{bci} communication assistive
@@
 Another limiting factor are the difficulties in correctly interpreting eye
-tracker results in studies with eye motor impaired patients.
+tracker results in studies with individuals with gaze impairment.
 If eye tracking at all is possible, it is not guaranteed that the user is able
@@
 measured.
-Rather, it was assumed that patients operated most comfortably in the free
+Rather, it was assumed that participants operated most comfortably in the free
 \ac{vsa} condition.
@@
 from~\textcite{VanDenKerchove2024} were adapted to make the counting task
-accessible to the participants with \ac{sspi} and gaze impairment.
+accessible to the \ac{bci} users with \ac{sspgi}.
 The number of repetitions or \ac{isi} were however not optimized to achieve
@@
 Future work includes comparing these results with a sample of non-gaze impaired
-participants with \ac{sspi}
-and carrying out repeated on-line sessions with these patients with built-in feedback to
+\ac{bci} users with \ac{sspi},
+and carrying out repeated on-line sessions with the current participants with built-in feedback to
 properly assess \ac{itr} and learning effects, and obtain realistic measures of percieved effort,

### 2024-10-07 chapters/patients_table.tex
funding

@@
-\begin{tabular}{llllllllr}
+\begin{tabular}{llllllllrr}
   \toprule
   \textbf{ID}  & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
-  \textbf{Diagnosis}
-  & \textbf{Speech}     & \textbf{Trach.} & \textbf{Communication}          &
-  \textbf{Cls.} \\ \midrule
-  PA1 & 58  & M   & L     & bulbar-onset \acs{als} & anarthric  & no          & tablet                 & 3  \\
-  PB1 & 41  & M   & L     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PB2 & 43  & F   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PB4 & 48  & M   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PC2 & 43  & M   & R     & brainstem stroke & anarthric  & yes         & \makecell[l]{prompting\\+eye movement} & 2 \\
-  PC3 & 43  & F   & R     & brainstem stroke & anarthric  & yes         & letterboard            & 2 \\
-  PC4 & 54  & M   & R     & \makecell[l]{left cerebellar stroke \\ (trombosis of the basilar artery)} & anarthric  & yes & letterboard & 2 \\
+  \textbf{Diagnosis} & \textbf{Speech} & \textbf{Trach.} & \textbf{Communication} &
+  \textbf{Cls. W} & \textbf{Cls. KB} \\ \midrule
+  PA1 & 58  & M   & L     & bulbar-onset \acs{als} & anarthric  & no          & tablet                 & 3  & 4 \\
+  PB1 & 41  & M   & L     & \acs{fa} & dysarthric & no          & verbal                 & 3  & 3 \\
+  PB2 & 43  & F   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  & 3 \\
+  PB4 & 48  & M   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  & 3 \\
+  PC2 & 43  & M   & R     & brainstem stroke & anarthric  & yes         &  \makecell[l]{prompting\\+eye movement} & 2 & 4  \\
+  PC3 & 43  & F   & R     & brainstem stroke & anarthric  & yes         & letterboard            & 2 & 3 \\
+  PC4 & 54  & M   & R     & \makecell[l]{left cerebellar stroke \\ (trombosis of the basilar artery)} & anarthric  & yes & letterboard & 2 & 3 \\
   \bottomrule

### 2024-10-07 chapters/stbf_struct.tex
Update bttda formulation
cover page
funding

@@
+\chapter{Kronecker-structured linear decoding}
+\label{sec:stbf-struct}
 \newcommand{\pv}[1]{
@@
   \emph{This chapter, except section~\ref{sec:stbf-struct-kronlda}, was
-  published by~\textcite{VanDenKerchove2022}}.
+  published as~\textcite{VanDenKerchove2022}}.
 
@@
   \begin{figure}
-		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
+		\includegraphics[width=.66\linewidth]{figures/stbf_struct/accuracy.eps}
     \caption[Clasifier accuracy in function of available training data.]{%
@@
 	\end{figure}
-	\begin{table}[h]
+  \begin{table}[p]
 		\centering
+    \makebox[\textwidth][c]{%
 		\input{chapters/stbf_struct_tests_1.tex}
+  }
     \caption[Statistical significance of differences in classifier performance (1 repetition).]{%
@@
 		\label{tab:p_values_1}
+    \bigskip
 
 		\centering
+    \makebox[\textwidth][c]{%
 		\input{chapters/stbf_struct_tests_2.tex}
+    }
 		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
@@
 		\label{tab:p_values_2}
+    \bigskip
 
 		\centering
+    \makebox[\textwidth][c]{%
 		\input{chapters/stbf_struct_tests_3.tex}
+    }
     \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
@@
 	\begin{figure}
-		\includegraphics[width=\linewidth]{figures/stbf_struct/training_time.eps}
+		\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
     \caption[Classifier training time.]{Median fold training time for different classifiers at different spatial
@@
 
-  \subsection{Extension to \ac{lda}}
-  \label{sec:stbf-struct-kronlda}
-  \todo{cite proof (or reiterate)}
-
 	\section{Conclusion}
@@
 
-\todo{
-in conclusion: LCMV struct might be to restrictive
-Toeplitzlda MEG paper}
-
-\todo{in conclusion: extension to the TFR domain}
 %\begin{equation}

### 2024-10-07 chapters/wcble.tex
funding

@@
+\chapter{\Acs{erp} latency estimation and alignment}
 \label{sec:wcble}
-\todo{comments liesel}
 \emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
@@
 
-\Ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
+\Ac{cble}, summarized in \autoref{sec:wcble/app/cble-alg}, works by training a
 \textit{first-stage} classifier $\mathcal{C}(\theta,f)$
@@
 
-The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
+The W\ac{cble} algorithm is presented in \autoref{sec:wcble/app/wcble-alg}.
 Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
@@
 }
-\todo{re-insert comments}
-
-\begin{algorithm}[H]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
-  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\State $\theta \gets
-			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
-      %\algorithmiccomment{Train stage 1}
-		\For{$n=1\ldots N$}
-      %\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\theta,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}
-      %\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-
-  \caption{\Ac{cble}}
-	\label{alg:cble}
-\end{algorithm}
-
-\begin{algorithm}[H]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
-		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
-	\begin{algorithmic}[1]
-		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
-    N$
-      %\algorithmiccomment{Train stage 1}
-		\Repeat
-		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
-		\For{$n=1\ldots N$}
-    \State  $s_n \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
-		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
-		\EndFor
-    \Until{convergence or maximum iterations reached}
-    \For{$n=1\ldots N$}%\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets
-      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-
-	\end{algorithmic}
-	\textbf{Output:} $\theta^*,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}%\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets
-      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-  \caption{\ac{wcble}}
-	\label{alg:wcble}
-\end{algorithm}
-
 \section{Results}
 \label{sec:wcble/results}
-\todo{refer to later chapter for results on real data}
-\subsection{Data simulation}
-To verify our approach, we simulated 16 channel \ac{eeg} data with standard
+Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
+first evaluate our approach with synthetic data.
+In \autoref{sec:covert-align}, the proposed algorithm will be applied to real
+\ac{eeg} data.
+
+\subsection{Synthetic data generation}
+To verify our approach, we simulated synthetic 16 channel \ac{eeg} data with standard
 10-20 electrode positions using the \texttt{simulate\_evoked} method of the
@@
 \fullpagefig{%
-  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
+  %\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
+  \sffamily
+  \sansmath
+  \inputpgf{figures/wcble}{simulated-sine.pgf}
 }{%
@@
 }{%
-  Simulated target and non-target evoked data. Targets contain a sine wave
-  pulse jittered by $\sigma$. 32 EEG channels with pink
+  Simulated target (bottom half of epochs) and non-target (top half of epoch) evoked data
+  and their average (red: target, yellow: noon-target, shaded are: standard
+  deviation).
+  Targets contain a sine wave pulse jittered by $\sigma$. 32 EEG channels with pink
   spatiotemporal noise were simulated using a forward head model, channel C3 is
   plotted here.
+
 }{fig:wcble/sim-sine}
-\todo{Properly align plots and label axes}
-\todo{decrease tick size}
+\todo{label x axis}
 
 \subsection{Latency estimation}
-\todo{refer to figure}
 We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
-estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
+estimation task using the pearson correlation coefficient $\rho$ at different
 jitter and \ac{snr} levels.
@@
 
-\begin{figure}[H]
-  \input{figures/wcble/fig_accuracy.pgf}
-  \caption{caption}
+\begin{figure}
+  \sffamily\sansmath
+  \input{figures/wcble/fig_latency.pgf}
+  \caption[Synthetic latency estimation results]{Correctness of estimated
+  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
+  target data, measured as pearson correlation coefficient $\rho$ between
+  estimated and ground truth latencies.
+  \Ac{wcble} estimates latencies more accurately in the presence of noise and
+  high jitter.}
   \label{fig:wcble/results/latency}
 \end{figure}
-
-\todo{properly label axes}
-\todo{uniform fonts with rest of document}
-\todo{correct width}
-\todo{figure for data generation}
 
@@
 classifier with $L_2$-norm regularization ($C=0.2$).
-\todo{ref figure}
 
 \begin{figure}
+  \sffamily\sansmath
   \input{figures/wcble/fig_accuracy.pgf}
-  \caption{caption}
+  \caption[Synthetic decoding results]{Decoding performance, measured as binary
+  classification accuracy, for \ac{cble}, \ac{wcble} and their first stage
+  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
+  synthetic evoked target and non-target data.
+  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
   \label{fig:wcble/results/accuracy}
@@
 
-The classification accuracy of both \ac{tlda} and \ac{cble} decreases with
+\autoref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble} decreases with
 $\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
@@
 plateau as \ac{snr} increases.
-\todo{insert results figure and discuss}
 
@@
 A suiting candidate is the spatiotemporal
-\ac{lcmv}-beamformer~\cite{VanVliet2015}\todo{refer chapter}.
+\ac{lcmv}-beamformer~\cite{VanVliet2015} described in
+\autoref{sec:stbf-struct}.
 The activation pattern can be set to the average of the set of trials to
@@
 \ac{erp} data and on the other hand to apply it as an \ac{erp} component
-separation method in the fashion of~\textcite{Oyang2017}.
+separation method in the fashion of~\textcite{Ouyang2017}.
 
@@
 be properly studied for this case.
+
+\clearpage
+\begin{subappendices}
+\section{\Acl{cble} algorithm}
+\label{sec:wcble/app/cble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
+  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\State $\theta \gets
+			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
+      \algorithmiccomment{Train stage 1}
+		\For{$n=1\ldots N$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\theta,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+
+  \caption{\Ac{cble}}
+	\label{alg:cble}
+\end{algorithm}
+
+\clearpage
+\section{\Acl{wcble} algorithm}
+\label{sec:wcble/app/wcble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
+		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
+	\begin{algorithmic}[1]
+		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
+    N$
+      \algorithmiccomment{Train stage 1}
+		\Repeat
+		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
+		\For{$n=1\ldots N$}
+    \State  $s_n \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
+		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
+		\EndFor
+    \Until{convergence or maximum iterations reached}
+    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets
+      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+
+	\end{algorithmic}
+	\textbf{Output:} $\theta^*,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets
+      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+  \caption{\ac{wcble}}
+	\label{alg:wcble}
+\end{algorithm}
+
+\end{subappendices}

### 2024-10-07 figures/bci/erp.tikz.tex
funding

@@
         height=0.6180\textwidth,
-        scale only axis,
         xlabel={Time (ms)},
@@
         axis lines=middle,
-	ylabel near ticks,
-	xlabel near ticks,
-	y label style={darkgray},
-	x label style={darkgray},
         ymin=-5, ymax=10,
         xmin=-100, xmax=800,
-        enlarge x limits=false,
-        enlarge y limits=false,
         smooth,
@@
         xtick={200,400,600}, % X-ticks at 200, 400, 600, 800
-        axis line style={darkgray}, % Set the axis line color to darkgray
-        tick style={darkgray}, % Set the tick color to darkgray
-        tick label style={darkgray} % Set the tick label color to darkgray
+      	ylabel near ticks,
+      	xlabel near ticks,
+        label style={darkgray, font=\footnotesize},
     ]
@@
           tension=0.5
-    ] coordinates {
+      ] coordinates {%
           (-100, 0)  % Starting point

### 2024-10-07 figures/wcble/smearing.tikz.tex
funding

@@
-\begin{tikzpicture}
-    % First subplot (low jitter) positioned relative to center
+\sffamily
+\sansmath
+\begin{tikzpicture}   % First subplot (low jitter) positioned relative to center
     \begin{axis}[
-        at={(-3cm,0cm)}, % Adjust position to the left, closer to the center
+        at={(0,0)}, % Adjust position to the left, closer to the center
         anchor=center,
-        width=7cm, height=6cm,
+        width=.45\textwidth, height=.309\textwidth,
         xmin=20, xmax=80,
@@
         axis lines=none, % Remove axes
-        title={Low Jitter}, % Add title
-        title style={yshift=-10pt}, % Adjust title position to make it closer to the plot
+        title={low jitter}, % Add title
+        title style={yshift=-10pt, color=muteblack}, % Adjust title position to make it closer to the plot
         ]
@@
    \begin{axis}[
-        at={(3cm,0cm)}, % Adjust position to the right, closer to the center
+      at={(0.55\textwidth,0)}, % Adjust position to the left, closer to the center
        anchor=center,
-        width=7cm, height=6cm,
+        width=.45\textwidth, height=.309\textwidth,
        xmin=20, xmax=80,
@@
        axis lines=none, % Remove axes
-        title={High Jitter}, % Add title
-        title style={yshift=-10pt}, % Adjust title position to make it closer to the plot
+       title={high jitter}, % Add title
+       title style={yshift=-10pt, color=muteblack}, % Adjust title position to make it closer to the plot
        ]
@@
    % Unified legend positioned further below the center
-    \node at (0,-5) {\begin{tikzpicture}
+   \node[anchor=center] at (.3\textwidth,-.2\textwidth) {\begin{tikzpicture}
        \begin{axis}[
@@
 \end{tikzpicture}
-\vspace{-5cm}
+%\vspace{-5cm}

### 2024-10-07 code/patients/fig_decode.py

@@ -3,7 +3,7 @@ import pdb
 import matplotlib.pyplot as plt
 import pandas as pd
 import seaborn as sns
-from setup import colors, configure_matplotlib_style, savefig_kws
+from setup import colors, configure_matplotlib_style, save_pgf_trim
 
 configure_matplotlib_style()
 
@@ -20,7 +20,16 @@ df_avg = df.copy()
 df_avg["subject"] = "avg."
 df = pd.concat([df, df_avg], ignore_index=True)
 
-pdb.set_trace()
+df = df.replace({
+    'PA01': 'PA1',
+    'PB01': 'PB1',
+    'PB02': 'PB2',
+    'PB04': 'PB4',
+    'PC02': 'PC2',
+    'PC03': 'PC3',
+    'PC04': 'PC4',
+
+})
 
 fg = sns.catplot(
     data=df,
@@ -43,5 +52,4 @@ for ax in fg.axes.flat:
 fg.set_axis_labels("patient", "ROC-AUC")
 fg.set_titles("{col_name} VSA")
 fig = fg.figure
-fig.set_size_inches(6.5, 2)
-fig.savefig("figures/patients/fig_decode.pgf", **savefig_kws)
+save_pgf_trim(fig, fg.axes[0,0], "figures/patients/fig_decode.pgf", height=2)

### 2024-10-07 code/setup.py

@@ -1,4 +1,5 @@
 import matplotlib.pyplot as plt
+import tikzplotlib
 from cycler import cycler
 
 mm = 1 / 25.4  # centimeters in inches
@@ -13,7 +14,6 @@ colors = dict(
 )
 
 
-savefig_kws = dict(transparent=True, bbox_inches="tight", pad_inches=0)
 
 
 # Function to configure matplotlib styles
@@ -27,74 +27,94 @@ def configure_matplotlib_style():
         {
             "text.usetex": True,  # use LaTeX to write all text
             "pgf.texsystem": "pdflatex",
+            #"axes.grid": False,  # Disable grid lines
             "pgf.rcfonts": False,
-            #'figure.figsize': (8, 6),                 # Plot size
-            "axes.grid": False,  # Disable grid lines
+            "axes.titlelocation": "left",  # Axis title alignment left
+            "figure.facecolor": "white",  # Background color for the figure
+            "image.cmap": "inferno",  # Set default colormap to Plasma
+            "axes.prop_cycle": cycler(
+                color=custom_color_palette
+            ),  # Custom color palette
+ 
+            "figure.titlesize": 11,  # Figure title font size
+            "figure.titleweight": "bold",  # Figure title bold
+
+            "xtick.direction": "in",  # X-axis ticks inward
+            "ytick.direction": "in",  # Y-axis ticks inward
+            "xtick.major.size": 3,  # Major tick size for x-axis
+            "ytick.major.size": 3,  # Major tick size for y-axis
+ 
+
+            'text.latex.preamble' : "\\usepackage{sansmath}\\sansmath",
             "axes.edgecolor": colors["darkgray"],  # Spine color
             "axes.labelcolor": colors["darkgray"],  # Axis label color
             "axes.titlesize": font_size,  # Axis title font size
             # "axes.titleweight": "bold",  # Axis title bold
-            "axes.titlelocation": "left",  # Axis title alignment left
             "axes.titlecolor": colors["darkgray"],  # Axis title color
             "axes.labelsize": font_size,  # Axis label font size
             "xtick.color": colors["darkgray"],  # X-axis tick color
             "ytick.color": colors["darkgray"],  # Y-axis tick color
             "xtick.labelsize": font_size,  # X-axis tick label size
             "ytick.labelsize": font_size,  # Y-axis tick label size
-            "xtick.direction": "in",  # X-axis ticks inward
-            "ytick.direction": "in",  # Y-axis ticks inward
[…]

### 2024-10-07 code/wcble/fig_latency_classify.py

@@ -1,17 +1,19 @@
 import matplotlib.pyplot as plt
+import pdb
 import pandas as pd
 import seaborn as sns
-from setup import configure_matplotlib_style, savefig_kws, set_size
+from setup import configure_matplotlib_style,textwidth_in, save_pgf_trim
 
 configure_matplotlib_style()
-figsize = set_size(subplots=(1, 3))
 
 
 # Latency
 df = pd.read_csv("data/wcble/results_align.csv")
+df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
 df = df[df["stc"] == "sine"]
 
-facet_kws = dict(ylim=[0, 1], xlim=[0, -31])
+facet_kws = dict(ylim=[0, 1], xlim=[0, -31], despine=False, legend_out=False,
+                 )
 fg = sns.relplot(
     data=df,
     x="snr",
@@ -25,10 +27,8 @@ fg = sns.relplot(
 fg.set_titles(col_template="$\\sigma={col_name}$s")
 fg.set_xlabels("SNR (dB)")
 fg.set_ylabels("$\\rho$")
-fg.figure.set_size_inches(*figsize)
-# fg.tight_layout()
+save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_latency.pgf', height=2)
 
-plt.savefig("figures/wcble/fig_latency.pgf", bbox_inches="tight", pad_inches=0)
 
 # Accuracy
 df = pd.read_csv("data/wcble/results_classify.csv")
@@ -38,7 +38,9 @@ df = pd.concat(
 df = df[df["stc"] == "sine"]
 df["accuracy"] *= 100
 
-facet_kws = dict(ylim=[0, 100], xlim=[0, -31])
+df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
+
+facet_kws = dict(ylim=[0, 100], xlim=[0, -31], despine=False, legend_out=False)
 fg = sns.relplot(
     data=df,
     x="snr",
@@ -52,20 +54,5 @@ fg = sns.relplot(
 fg.set_titles(col_template="$\\sigma={col_name}$s")
 fg.set_xlabels("SNR (dB)")
 fg.set_ylabels("accuracy (%)")
-fg.figure.set_size_inches(*figsize)
-# fg.tight_layout()
 
-
[…]

### 2024-10-07 code/wcble/fig_simulated.py

@@ -3,7 +3,8 @@ import numpy as np
 import pandas as pd
 import seaborn as sns
 from mne import read_epochs
-from setup import configure_matplotlib_style, darkgray, pagesize, savefig_kws
+from setup import configure_matplotlib_style,colors, save_pgf_trim
+import pdb
 
 configure_matplotlib_style()
 
@@ -42,17 +43,21 @@ def order(epochs, n_epochs=100):
     return order
 
 
-def plot_img(ax, epochs):
+def plot_img(ax, epochs, cbar_ax=None):
     order_ = order(epochs)
+    if cbar_ax is None:
+        axes=ax
+    else:
+        axes=[ax, cbar_ax]
     epochs.plot_image(
         order=order_,
-        colorbar=False,
+        colorbar=cbar_ax is not None,
         evoked=False,
-        axes=ax,
+        axes=axes,
         show=False,
         **img_kwargs,
     )
-    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=darkgray)
+    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=colors['darkgray'])
 
     ax_img.set_title("")
     ax_img.set_xticks([])
@@ -66,12 +71,14 @@ def plot_ts(ax, epochs):
     df = pd.DataFrame(y.T, index=x)
     df.index.name = "time"
     df = df.melt(ignore_index=False, value_name="amplitude")
-    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts)
+    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts,
+    errorbar='sd')
     y = epochs["target"].pick("C3").get_data().squeeze() * 1e6
     df = pd.DataFrame(y.T, index=x)
     df.index.name = "time"
     df = df.melt(ignore_index=False, value_name="amplitude")
-    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts)
+    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts,
+                 errorbar='sd')
 
 
 stc = "sine"
@@ -79,7 +86,7 @@ stc = "sine"
 height_ratios = []
 for _ in range(len(snrs)):
     height_ratios.extend([3, 1])  # Closer together for image+timeseries
[…]

### 2024-10-07 chapters/bci.tex
All text

@@
 \todo{acro usage in chapters}
+
 \chapter{Brain-computer interfaces}
@@
 }{%
-  The \ac{bci} loop.
+  The \acs{bci} loop.
 }{%
@@
 }{%
-  Overview of \ac{bci} paradigms.
+  Overview of \acs{bci} paradigms.
 }{%
@@
   \begin{subfigure}[t]{.45\textwidth}
-  \includegraphics[width=\textwidth, height=\textwidth]{example-image-a}
+  \includegraphics[width=\textwidth,
+    height=\textwidth]{figures/bci/illustration}
+    \caption{Stimuli are flashed while the \ac{eeg} is recorded}
   \end{subfigure}\hfill%
@@
     \input{figures/bci/erp.tikz.tex}
-    \caption{The idealized \ac{erp} evoked by an attended stimulation.}
+    \caption{The idealized \ac{erp} evoked by an attended stimulation in a
+    single channel.}
   \end{subfigure}\hfill%
@@
 
-\subsection{The \acf{erp}}%
+\subsection{The \acl{erp}}%
 \label{sec:bci/oddball/erp}

### 2024-10-07 chapters/bttda.tex
funding
All text

@@
 	\input{figures/bttda/HODA_backward.tikz.tex}
-  \caption[The \ac{hoda} backward projection.]{%
+  \caption[The \acs{hoda} backward projection.]{%
     A visualization of the multilinear projection obtained by Higher Order
@@
 \begin{algorithm}
-  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
+  \caption[The \acs{hoda} backward solution.]{The \ac{hoda} backward solution.}
 	\label{alg:HODA}
@@
 performance.
-Useful features can still be extracted from $\ten{\mathbfcal{eps}} = \ten{X} -
+Useful features can still be extracted from $\ten{\mathbfcal{E}} = \ten{X} -
 \hat{\ten{X}}$, so it is further projected onto another core tensor
@@
 \subsection{Block contribution}
-\todo{Discuss number of features}
 To analyze the contribution of extra feature blocks extracted by {BTTDA} over
@@
 	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
-  \caption[Extracted \ac{bttda} activation patterns.]{%
+  \caption[Extracted \acs{bttda} activation patterns.]{%
     Spatial (left two columns) and temporal (middle column) activation patterns and

### 2024-10-07 chapters/conclusion.tex
All text

--- 2024-10-04 chapters/conclusion.tex
@@
-\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
-\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
-\todo{BCI principles and practice chapter 19}
-\todo{https://www.nature.com/articles/s44222-024-00239-5}
+\chapter{Conclusion \& recommendations}
 
@@
 
-\todo{come back on hypothesis}
+This work aimed to improve the performance of visual gaze-independent \acsp{bci}, in general
+and applied to individuals with eye motor impairment by developing novel decoding
+strategies and evaluating their effectiveness.
+We addressed the limitations of current gaze-dependent \acp{bci} by proposing methods that
+exploit covert \ac{vsa} and reduce the reliance on eye gaze.
+The following key contributions were made:
 
-\subsection{Developed \Ac{erp} decoders}
+\subsection{Developed \acs{erp} decoders}
 We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
@@
 generalized to other ERP-based BCI settings.
-These results have been published in~\cite{VanDenKerchove2022}.
+These results have been published in~\textcite{VanDenKerchove2022}.
 
-We have tested our proposed methods and similar methods in gaze-independent and
-covert VSA settings, but we did not observe a relative increase in performance
-in these settings.
+Next, \acf{bttda} was introduced as a tensor-based decoder that better captures the
+multidimensional nature of \ac{erp} data.
+By preserving the structure of neural data, \ac{bttda} can cope with noise and other challenges
+arising from gaze-independent \acp{bci}, yielding robust results under certain conditions.
+Results show that \ac{bttda} and its special sum-of-rank-one structured case
+improved over \ac{hoda} and can reach state-of-the-art decoding performance for
+\acp{erp}.
+This work is submitted as~\textcite{Vandenkerchove2024b}.
 
-\todo{tensor}
-\todo{wcble}
-
-%* Preliminary results: BTTDA better than HODA, on par with other state of the
-%art methods and in some situations better (data availability, noise structure)
+Finally, we developed the \acf{wcble} decoder, specifically designed to address
+the challenges posed by P3 latency jitter in covert \ac{vsa} settings.
+Latency variability is a well-known issue in \ac{erp} decoding, particularly
+when users engage in covert attention, where the timing of the P3 component
+fluctuates significantly across trials.
+Traditional decoders struggle in these settings due to the inconsistent timing
+of brain responses, which can reduce the robustness and accuracy of
+\ac{erp} decoding.
+The \ac{wcble} decoder mitigates this issue by introducing an iterative process
+to estimate and align \ac{erp} latencies across trials.
+This is performed in such a way that discriminative power of the \ac{erp}
+signal is preserved and enhanced.
+The method has been published in~\textcite{VanDenKerchove2024}.
+We took care to design a method that can be applied to improve decoding, where
+unseen incoming test data is not yet known.
+\Ac{wcble} was first tested on synthetic data to evaluate its
+effectiveness under conditions of controlled latency variability and noise.
+These results showed that is is robust to higher noise and jitter compared to
+a non-iterative method in a latency estimation task.
+It was also shown that decoding accuracy was higher in high jitter and noise
+settings compared to the non-iterative method and a sate-of-the-art decoder.
 
@@
 $26.38\pm3.15$ years.
-This dataset gives us insight in the ERP dynamics in overt, covert and the
+The dataset was presented in~\textcite{VanDenKerchove2024}.
+The experiment in this dataset implemented a visual odball \ac{bci} with six
+circular targets in a hexagonal layout.
+The gaze fixation of participants was carefully controlled to dissociate the
+visuospatial attention of the participant and their eye gaze.
+This allows us to study effects of gaze-independence on \ac{erp} decoding.
+The CVSA-ERP dataset gives us insight in the ERP dynamics in overt, covert and the
 novel split VSA condition, and confirms our hypothesis that P3 jitter has a
-significant impact on performance in covert and split VSA, as well as that the
-amplitude effect\todomvh{not the right word, maybe the effect of covert VSA on
-ERP...} observed for covert VSA also holds for split VSA.
+significant impact on performance in covert and split VSA.
+It also confirmed the effects of covert \ac{vsa} on \ac{erp} component
+amplitude hold for split \ac{vsa}.
 
-\todo{describe patients dataset}
+Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
+amyotrophic lateral sclerosis (\ac{als}), Friedreich's ataxia (\ac{fa}), and stroke.
+These participants exhibited varying degrees of eye motor impairment, such as
+involuntary eye movements, ophthalmoplegia, and gaze fixation fatigue.
+The data provide invaluable real-world evidence on how gaze-independent
+\acp{bci} perform in populations that experience eye motor difficulties.
+This dataset allowed us to evaluate the proposed decoding strategies in clinical settings and
+highlighted the practical challenges of implementing gaze-independent \acp{bci} in these user
+groups.
 
-\subsection{Investigated gaze-independence visual \acsp{bci}}
+\subsection{Investigated gaze-independent visual \acsp{bci}}
 
-We have evaluated our proposed wCBLE algorithm, the baseline CBLE algorithm and
-a state-of-the-art algorithm (tLDA)~\cite{Sosulski2022} that uses similar techniques as our
-proposed STBF-struct method on this dataset as well as a publicly available
+We have evaluated our proposed \ac{wcble} algorithm, its non-iterative
+counterpart and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
 dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
@@
 Performance was significantly different between decoders, but this
-result was significantly dependent on the VSA condition and the dataset.
-While CBLE performed overall on par with the state-of-the art method for both datasets, wCBLE was
-outperformed by tLDA in overt VSA decoding but outperformed
-tLDA for covert VSA.
+result was significantly dependent on the \ac{vsa} condition and the dataset.
+While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
+decoding, but increased covert \ac{vsa} decoding.
+
 For the split attention conditions in the CVSA-ERP
-dataset, wCBLE yielded a significant improvement over CBLE and the
-state-of-the-art method only for d = 1 and d = 3.
-These results were corroborated by analyzing selection accuracy and information
-transfer rate, which showed similar behavior for both datasets, except in overt
-VSA, where accuracy and information transfer rate were not harmed by the lower
-single-trial selection performance of wCBLE.
+dataset, \ac{wcble} yielded a significant improvement over CBLE and the
+state-of-the-art method only in some cases.
+These results were corroborated by analyzing selection accuracy, which showed
+similar behavior for both datasets, except in overt
+VSA, where accuracy was not harmed by the lower
+single-trial selection performance of \ac{wcble}.
 
 To further study the gaze-independent performance of these algorithms, transfer
-learning between VSA conditions is studied to simulate conditions where a patient
-can end up in different VSA settings within a BCI operation session due to their
-lack of eye motor control.
-When trained and evaluated on overt VSA data, our proposed wCBLE algorithm results in a significant decrease in performance
-over the state-of-the art tLDA decoder, consistent with the within VSA
-condition results.
-For all other pairs of training and evaluation VSA condition, however, wCBLE
-never yields a significant decrease in performance and often significantly
-outperforms the other methods.
+learning between \ac{vsa} conditions is studied to simulate conditions where an
+individual with gaze impairment can end up in different \ac{vsa} settings within
+a \ac{bci}operation session due to their lack of proper motor control.
+When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
+results in a significant decrease in performance compared the state-of-the art,
+consistent with the within-condition results.
+For all other pairs of training and evaluation VSA condition, however,
+\ac{wcble} was equal or significantly better.
 
-These results show that there is an interest in developing a new class of ERP-BCI
-interfaces for patients that prefer to rest their eyes on a chosen point on the
-screen, so as to avoid the effort of redirecting their eye gaze at different spatial locations on
-the stimulation screen. Furthermore, the performance gain in the split VSA conditions
-and in the between-VSA condition transfer settings are promising for patients with even
-less eye motor control, such as patients experiencing involuntary saccades or
-fatigue. An accurate decoder would allow them to comfortably operate a BCI while
-resting their eyes on whichever portion of the screen they prefer, even when there is
-another target present at that location or this location varies during the course of the
-experiment. Our results show that whenever the eye gaze is directed at a different
-spatial location other than that of the mentally attended target, the wCBLE decoder can
-more accurately discern the locus of mental attention, promoting in this way accurate
-gaze-independent decoding. This comes, however, at the cost of a decreased performance
-compared to overt VSA operation.
+Later case studies explored gaze-independent
+\ac{bci} performance in individuals with \ac{sspi} and varying eye motor impairments.
+Participants showed different levels of gaze control, with some using overt
+\ac{vsa}, while others preferred covert \ac{vsa}, resting their gaze and
+mentally attending to targets.
 
-\todo{. Coincidentally, this also
-builds towards a solution for the Midas Touch Problem in BCI, where a BCI user sometimes accidentally
-selects a target while not wanting to give any input. Decoding of mental attention independent of eye
-gaze, with the option of having visual attention without mental attention,
-would counteract this.}
+The \ac{wcble} decoder improved decoding accuracy covert
+\ac{vsa} settings with cued central gaze fixation.
+This is of interest for gaze-independent decoding, but not as optimal as if the
+gaze fixation was uncued.
+The latter was tested in the free \ac{vsa} settion.
+Here, participants were generally more comfortable when allowed to use their
+preferred gaze strategy.
+Decoding performance in free \ac{vsa} was often comparable to overt \ac{vsa},
+showing the system's flexibility.
 
-\section{Limitations}
-
-hard to evaluate experience in long term home use
-unify clinical side and bci development side
-
-very small group of interest, but available right now, no implantation needed,
-
-even if it is the most suited strategy for a patient, it will probably be
-improved if a combination of eye tracking, residual motor control and other
-biosensors are used.
-
-true power in the future is probably focused on
-invasive, active paradigms if recording technology and surgical methods
-improve lowering the risk. orders of magnitude higher itr and more natural
-communication.
-characterized by publication date, but might indicate it is of interest to
-heavily focus this research now on applied 'vraagstukken' with patients
-
-\section{Recommendations}
-
-this work focused on classification methods, and solving problem through
-decoding
-
-classification performance is not everything
-
-ITR, comfort and user satisfaction in on-line operation is the ultimate goal
-
-also do advance calculations of what the information transfer rate would be
-given reasonable accuracy, so that you can go in with a reasonable interface
-that has state-of-the-art ITR
-
-classification performance is convenient since you can do a single data
-collection and iterate offline.
-this constricts you to decoder development
-it is a better approach to only go there after optimizing the interface
-experimentally and with patients with an existing
-decoder, and only moving there once the options for progress are depleted
-
-performance and abilities are probably so user dependent that it is not said that
-findings from a population study with healthy controls will hold for a given
-user with \ac{sspi}.
-Incremental gains in performance in decoding for a specific paradigm will be
-outweighed by how well that paradigm is adapted to the user, factoring in,
-skills and abilities.
-
-best approach should be everything working together, such as is advocated
-by~\textcite{Pan2022} and \textcite{Fouad2020}
-
-instead of decoder > data > patients > online
-go other way
-online > patients > data > decoder
-this is unfortunately also the least practical way to do it for a 4-year
-research project.
-
-It is therefore important that research labs keep a working, long-term collaboration with
-patient centers such that new projects can immediately be tested by the target
-population, and that patients with bci experience can compare multiple systems
-and guide development
-The fact that a user has previous bci experience is often seen as a confounding
-factor, while it is a strength that can help you gain insight into what works
-and doesn't work for them, and where research efforts should be focused.
-
-also maintain a working, on-line system that can be iterated upon
-
-in a phase that is heavily dependent on user experimentation
-
-not best approach for all research projects, but
-Visual oddball bci field is advanced enough to do this
-
-\subsection{Working with patients}
-gaze independent visual, can be good current existing solution
-take into accounrecommendations of \textcite{FriedOken2020}
-
-immediately move to patients and on-line implementation, to be able to report
-interpretable metrics
-permanent working collaboration as a lab
-
-In a mature field like visual bci, start from patients.
-avoid situations where a problem is solved that doesn't pose itself in a
-patient population
-use all available capabilities and signals
-
-%Have attention for the emotional impact
-
-\subsection{Linear decoding methods}
+Together, these results show that there is an interest in developing a new class of ERP-BCI
+interfaces for patients that prefer to choose their own gaze strategy, so as to
+avoid the effort of redirecting their eye gaze at different spatial locations
+on the stimulation screen in manners that might be uncomfortable for them.
+Coincidentally, advances in gaze-independent decoding also build towards a
+solution for the Midas Touch Problem in visual \ac{bci}.
+Here a \ac{bci} user sometimes accidentally selects a target while not wanting
+to give any input.
+Decoding of true intention independent of eye
+gaze, with the option of gazing without paying visuospatial attention to a
+stimulus, would counteract this.
 
 \section{Current \& future work}
-kronecker-sum lda
-multicomponent alignment approach
-hankel tensors to cope with jitter
+On the decoder development side, we have partially implemented some interesting
+extensions of the proposed methods.
+The Kronecker-structured beamformer can be generalized to the \ac{lda} case.
+Furthermore, we obtained promising results when extending the single
+Kronecker product covariance model to a sum of Kronecker product terms, since
+the \ac{eeg} covariance is probably better expressed by such a
+structure~\cite{Bijma2005}.
+
+To overcome the limitations introduced by the presence of multiple \ac{erp}
+components to latency estimation, we have also developed a multi-component
+version of the \ac{wcble} algorithm.
+This algorithm should be able to separate mixed \ac{erp} component clusters
+based on their temporal coherence, in a similar maner to~\textcite{Ouyang2017}.
+It could also yield favorable decoding results in a broader range of settings
+than overt \ac{vsa} since it should be able to account for the presence of both
+visual and attentional \ac{erp}components}.
+
+In gaze-independence decoding, our current efforts focus on,
+counterintuitively, integrating eye tracking into the decoding strategy.
+Since we are able to explicitly discern overt, covert and split \ac{vsa} from
+the \ac{erp}, and different decoders perform best in different settings, it
+could be helpful to derive the current setting from the gaze position and the
+\ac{erp} and propose the most suited decoder for a given data point.
 select best classifier using eye tracking
-expand on patient experiemnts
+
+We also wish to apply the tensor approach to gaze-inependent decoding.
+The problems introduced by jitter could be accounted for as in \ac{wcble} by
+alignment, but also by other methods that model the possible time shifts.
+In ordering the data as a Hankel tensor\footnote{Adding an extra mode with
+time-shifted copys of the temporal response per channel.}, we should have a
+data representation that is more robust to jitter, at the cost of increased
+dimensionality.
+This increased dimensionality can than be countered by a tensor method like
+\ac{hoda} or \ac{bttda}.
+
+Finally, we wish to expand on experimentation with individuals with
+\acp{sspgi}.
+We aim to revisit the patients with an on-line experiment and comfort
+assessment, to establish the usability of the proposed interface and to be able
+to compare its \ac{itr}  to the state-of-the-art literature.
+
+\section{Limitations and recommendations}
+
+\subsection{Decoding: keep it linear; structure it}
+
+Throughout this thesis, we've noticed that linear or multilinear decoding
+methods consistentcly outperform non-linear methods, like SVMs or those based
+in Riemannian Geometry.
+Within these linear models, regularization (shrinkage, covariance structure,
+tensor structure/rank) is of paramount importance, since neural data is
+inherently multivariate and \ac{bci} calibration time is minimized.
+Proper regularization should impose some specific structure on linear model,
+ideally reflecting properties of the signal.
+When faced with new problem and few data relative to dimensionality, it is a
+good first instinct to pick a simple, restrictive structure.
+
+Yet, the brain is a complex organ, hence neural data is not linear in origin and
+these regularizing assumptions based on prior knowledge are probably only
+superficially true.
+When optimizing for performance, it will soon become clear that, given a more
+flexible model, there are
+settings where these assumptions can be broken to better describe the ongoing
+interactions.
+For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
+improves performance in low sample sizes, but is outperformed by more flexible
+methods, e.g.\ \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
+if training sample size increases.
+Similarly, \ac{bttda} is more flexible
+
+This calls for the development of more flexible (multi-)linear models that still rely on some
+structure, but which are combined with efficient model
+selection, such as proposed by \ac{bttda}.
+These models try to combine the best of both worlds.
+
+%\subsection{Visual? gaze-independent \ac{bci}}
+%
+%While visual \acp{bci} are often praised for their easy, non-invasive
+%applicability, they are also inherently unreliable, require wearing the
+%\ac{eeg} cap and force continuous stimulation upon the user, all while forming
+%a rather unintuitive method of communication for humans who are used to speech
+%or motor commands.
+%
+%Currently, the group of users that would directly be helped with \emph{only} a
+%gaze-independent visual \ac{bci} is not that large.
+%Even if some eye motor impairment is the most suited strategy for a patient, it will probably be
+%improved if a combination of eye tracking, residual motor control and other
+%biosensors are used.
+%
+%
+%\todo{https://www.nature.com/articles/s44222-024-00239-5}
+%very small group of interest, but available right now, no implantation needed,
+%
+%true power in the future is probably focused on
+%invasive, active paradigms if recording technology and surgical methods
+%improve lowering the risk. orders of magnitude higher itr and more natural
+%communication.
+%will probably become acceptable if risks of implantation and longevity
+%increase.
+%characterized by publication date, but might indicate it is of interest to
+%heavily focus this research now on applied 'vraagstukken' with patients
+%
+%Why limit ourselves to the visual~\ac{bci} decoding problem postulated in this
+%thesis?
+%
+%
+\subsection{Optimize user experience}
+While our work with \ac{bci} users with \ac{sspgi} results in interesting
+preliminary results, it is currently impossible to make claims about the experience in long term
+home use of the proposed systems as assistive technology.
+This work focused heavily on classification methods, and solving problem
+arising from gaze-independence through decoding.
+A natural metric to then evaluate proposed decoders is \ac{erp} classification
+performance.
+
+However, classification performance is only one aspect of what makes a \ac{bci}
+desirable as assistive technology.
+\Ac{itr} in on-line operation is a better metric, and the ultimate goal
+is user satisfaction.
+The latter can depends on performance, comfort and perception.
+Classification performance is a convenient metric since a researcher can perform a single
+data collection and iterate on the results offline.
+However, this can be too constrictive when seeking the most suited interface
+for a specific (group of users).
+
+In the optimal approach, decoder development, paradigm choice and interface
+design should working together, such as is advocated
+by~\textcite{Pan2022} and~\textcite{Fouad2020}, by unifying the engineering and
+the clinical side.
+Performance and abilities are probably depend in such a degree on the user,
+that off-line decoding performance in healthy controls is probably not a great
+predictor of on-line decoding performance and \ac{itr} in individuals with
+\ac{sspi}.
+Incremental gains in performance in decoding for a specific paradigm might
+also be outweighed by how well that paradigm is adapted to the user, factoring in,
+skills and abilities.
+Longitudinal research should follow several patient groups specifically with gaze
+impairment and experiment with specific paradigms, interfaces and decoders for
+a given \ac{bci} user.
+
+\subsection{Work \emph{with} individuals with \acs{sspi}}
+
+A logical point that arising from this is the question of how to setup an
+effective research project that aims to optimize \ac{bci} user experience, in
+the context of he topics presented in this work.
+The patient-research involvement framework postulates that, ideally, individuals with
+\ac{sspi} should be involved at every stage of the research project, from conceptualization
+to evaluation.
+A good rule of thumb is that research should solves problems \emph{with} the
+users, not \emph{for} them, implying that research projects should start from
+the need of the \ac{bci} user population.
+
+Instead of choosing a research target and then moving on to decoder
+development, data collection with healthy controls, patient studies and,
+finally, verification in on-line operation, it could be more effective to start
+the other way around.
+In a mature field like visual oddball \ac{bci}, the research project can, after
+a thorough literature review, immediately start with on-line experimentation
+involving individuals with \ac{sspgi} and an adequately chosen, existing \ac{bci} system.
+It should then become apparent where challenges for the envisioned user
+(population) lie.
+As an example, \textcite{FriedOken2020} share an overview of their experiences from this step,
+specifically concerning eye motor impairment in visual \ac{bci}.
+After that, hypotheses can be formed on how to mitigate these challenges
+through interface optimization or choice taking into account the full picture
+of the capabilities and skills of the users.
+These hypotheses can then verified in experiments
+with the users and/or a population of healthy controls.
+Finally, the decoder can be optimized, only if it is strongly suspected that
+effects will hold in on-line operation.
+
+Unfortunately, this is also probably one of the least practical ways to set up
+a 4-year research project, e.g.\ in the context of doctoral thesis.
+This approach can only work if the proper facilities can be gathered immediately
+from the start, like ethical approval and access to patients and on-line \acp{bci}
+experimentation systems.
+It is therefore important that research labs keep a working, long-term
+collaboration with patient centers such that new projects can immediately be v
+verified with population, and that interested users can compare multiple systems
+and guide development.
+The fact that a user has previous \ac{bci} experience is often seen as a confounding
+factor in off-line studies.
+When the research is in a phase that is heavily dependent on user
+experimentation,
+this is in fact a strength, sinche the user might help the researcher gain
+insight into what works and doesn't work for them, and where research efforts
+should be focused.
+
+Similarly, there is value in maintaining a working, on-line in house \ac{bci}
+system that can be iterated upon and adapted if necessary.
+Experiments can be supported by advance calculations of hypothetical
+\ac{itr}, such that performed on-line or off-line experiments reasonably reflect
+real-world operation and no effort is wasted exploring settings that should not
+reach satisfactory performance.
+For example, why study a specific number of repetitions or \ac{isi}, if they
+would not be practical to achieve a high \ac{itr}, but can still influence
+decoding performance?
+
+This top-down approach is probably not suited for all types of \ac{bci}
+research.
+Yet, in visual oddbal \ac{bci} where the challenge lies no longer in generating
+a proof of concept but in translating the concept into a useful technology, it
+might help avoid prematurely solving problems that might not necessarily be
+encountered further down the road.
+A top-down approach might improve the quality of study outcomes by enabling the
+reporting of realistic, interpretable metrics,
+and take into account the human aspect of working \emph{with} \ac{bci} users.
 

### 2024-10-07 chapters/funding.tex
All text

@@
 We also thank,  Arno Libert,
-Tjaša Mlinarič, Yide Li and Mani Mirsaeedi for their help in data collection,
+Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
+for the works presented in \autoref{sec:covert-align} and
+\autoref{sec:patients}.
 and in general the colleagues at the KU Leuven Computational Neurosciences
@@
 Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
-the work presented in Chapter~\ref{sec:bttda}.
+the work presented in~\autoref{sec:bttda}.
 

### 2024-10-07 chapters/gaze_independence.tex
All text

@@
   \input{figures/gaze_independence/lis_venn.tikz.tex}
-  \caption[Venn diagram of potential users of \ac{bci} assistive technology.]{%
+  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
     Venn diagram of potential users of assistive technology.
@@
 \subsubsection{Tensor discriminant analysis}
-\todo{fill in}
+
+In the context of \ac{bci} decoding, extracting robust and discriminative
+features from multidimensional neural data is critical.
+Tensor decoding methods offers a powerful approach by preserving the multiway structure
+of \ac{eeg} data while optimizing class separability.
+Unlike traditional methods that flatten data, TDA operates directly on tensors,
+making it particularly suited for \ac{bci} applications, where signals are structured
+across multiple channels and time points.
+
+A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
+optimized for class separation, similar to \ac{lda}, but extended to multidimensional data.
+This method has been applied to various \ac{bci} tasks, including \ac{erp} and
+\ac{mi} decoding, enabling the extraction of class-relevant features while reducing
+the data's dimensionality.
+
+While \ac{hoda} is effective, it is limited by its reliance on a rigid core
+structure, which may not fully capture the complexity of neural data in some
+scenarios. To address this, \ac{bttda}
+introduces a more flexible block-term structure.
+\ac{bttda} iteratively extracts multiple blocks of
+discriminative information, which can help in capturing more complex patterns
+in neural data.
+While \ac{bttda} is not specifically tailored to
+gaze-independent \ac{bci} decoding, specific tensorizations of neural data can
+offer advantages in coping with several obstacles arising from
+gaze-independence.
+The proposed \ac{bttda} method will be discussed in~\autoref{sec:bttda}
 
 \subsubsection{Classifier-based Latency Estimation with Woody iterations}
-The previous two methods were aimed at improving decoding performance, but
-did not (yet) yield results specifically in gaze-independent decoding.
+The previous two methods did not yet yield results specifically in gaze-independent
+decoding.
 The following approach is targeted towards improving decoding performance in
@@
 This approach has been published by~\cite{VanDenKerchove2024} and is presented
-and evaluated on simulated data in Chapter~\ref{sec:wcble}.
+and evaluated on simulated data in~\autoref{sec:wcble}.
 We have also collected datasets to evaluate this approach on real \ac{erp} data
@@
 \ac{vsa} conditions.
-
-\todo{mention, free, split, frenzel}
 
@@
   \caption[\Ac{vsa} conditions]{%
-    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
+    \Acl{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
   }
@@
 \subsubsection{Case studies with gaze-impaired individuals}
-\todo{fill in}
-\todo{mention multi-center patient study case studies in cooperation
-After data collection from healthy participants and decoder development, we
-have carry out a patient study using a similar interface.
-The goal of the patient study is to evaluate the influence of eye motor
-deficits in patient populations on existing state-of-the-art decoders and on
-our proposed decoding methods.
-}
+\label{sec:patients/approach/casestudies}
+
+After collecting data from healthy participants and developing initial decoders, we conducted
+a patient study with gaze-impaired individuals to evaluate the impact of eye motor deficits
+on \ac{bci} performance.
+This study focused on individuals with \ac{sspgi}, recruited from neurorehabilitation centers
+and specialized care homes, including participants with conditions such as \ac{als}, \ac{fa},
+and stroke.
+
+The primary aim was to assess whether gaze-independent decoding strategies could improve
+\ac{bci} performance in this group.
+We used the visual Hex-o-Spell interface~\cite{Treder2010}, adapting the task by adding a
+free \ac{vsa} setting, where participants could operate the interface without being required
+to fixate their gaze.
+This setting allowed us to investigate how participants with varying eye motor capabilities
+naturally interact with a \ac{bci}, compared to standard overt and covert \ac{vsa} tasks.
+
+The hypothesis was that gaze-independent decoders, such as \ac{wcble}~\cite{VanDenKerchove2024},
+would improve performance in covert and free \ac{vsa} conditions compared to standard
+decoders like \ac{tlda}.
+We aimed to determine if calibration in an overt \ac{vsa} setting could also enhance
+performance in free and covert conditions, leveraging residual eye motor control during
+calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
 The results of this study are presented in chapter~\ref{sec:patients}.

### 2024-10-07 chapters/patients.tex
All text

@@
 target individuals with \ac{sspi}~\cite{Peters2022}.
-
-\todo{visual bci}
+Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
+offer several advantages in this context.
+They can work with non-invasive recording technology and can use rapid
+stimulation.
+This makes them well-suited for real-time tasks communication tasks.
 
@@
 into a study and perform experiments with them~\cite{Wolpaw2006}.
-Patients with less severe paralysis or in less progressed disease stages that struggle with
+Individuals with less severe paralysis or in less progressed disease stages that struggle with
 eye-tracking technology could also benefit from
@@
 \section{Materials \& methods}
-\subsection{Patient recruitment}
-Patients were recruited from across the Neuromuscular Reference center at
+\subsection{Recruitment}
+Participants were recruited from across the Neuromuscular Reference center at
 University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
@@
 Experiments were performed under the supervision of their treating physician.
-Patients were recruited based on the following criteria.
+Participants were recruited based on the following criteria.
 To qualify for inclusion, participants must
@@
 \end{enumerate}
-Patients were excluded if they
+Participants were excluded if they
 \begin{enumerate}
@@
 \end{table}
-\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
-Brain-computer interfaces and communication in paralysis: extinction of goal
-directed thinking in completely paralysed patients? Clin Neurophysiol.
-2008;119:2658-2666.}
-\todo{include year of diagnosis}
+\todo{include years since diagnosis}
 
@@
 All participants reported some degree of fatigue or discomfort when fixating.
-Patient PA1 had the mildest impairment, only reporting fatigue when fixating
+Participant PA1 had the mildest impairment, only reporting fatigue when fixating
 for prolonged times.
@@
 recorded \ac{eeg}.
-Eye tracking throughout was performed using the Tobii~\todo{model, make}
-portable eye tracker placed at the bottom of the laptop screen.
+Eye tracking throughout was performed using the Tobii X2-30 Compact (Tobii
+Technology AB, Sweden) portable eye tracker placed at the bottom of the laptop screen.
 
@@
 
-\subsection{BCI decoding}
+\subsection{\Acs{bci} decoding}
 
-\todo{shortly describe WCBLE}
-\todo{shortly describe tLDA}
-\todo{shortly describe xdawncov+ts+LDA}
+We evaluate the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
+and \ac{tlda}~\cite{Sosulski2022}.
+classifiers, and the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
+For \ac{wcble}, a region of interest from 0ms to 800ms relative to stimulus
+onset was used while the epoch was cropped to -100ms to 900ms, for the other
+decoders the epoch was cropped between 0ms and 800ms which resulted in maximal
+performance.
+Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
+stimulation blocks.
+
+refer to~\autoref{sec:covert-align/methods/decoders}.
 
@@
 
-It was technically impossible to register gaze position with the Tobii
-\todo{model} for participants PC2 and PC3 since they both had one eye that was
+It was technically impossible to register gaze position with the Tobii X2-30
+Compact for participants PC2 and PC3 since they both had one eye that was
 occluded respectively by the prism glass and the eyelid.
@@
 
-\subsection{\Ac{bci} decoding performance}
+\subsection{\Acs{bci} decoding performance}
 
@@
 \ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.
-Scores were obtained using 6-fold cross-validation where folds corresponded to
-stimulation blocks.
 
@@
   \caption[%
-    Decoding performance in different \ac{vsa} settings.
+    Decoding performance in different \acs{vsa} settings.
   ]{%
@@
 \end{figure}
-\todo{figure margin}
 \todo{check if percentages or 0-1 accuracy and roc-auc everywhere}
@@
 \label{sec:patients/outcomes/cross}
+As an alternative approach to picking the most suitable decoder, we chose
+\ac{tlda} as a given decoder and verified whether performance can be improved
+if \ac{bci} users with gaze impairment perform the calibration session relying
+maximally on their residual gaze control.
 
-\todo{input and describe graph}
+\begin{figure}[t]
+  \input{figures/patients/fig_cross.pgf}
+  \caption[Cross-setting calibration and decoding performance.]{
+    Decoding performance when calibrating the \ac{tlda} decoder in a given \ac{vsa}
+    setting, and evaluating it in another, reported as
+    single-trial \ac{rocauc}.
+    For participant PA1, free \ac{vsa} performance could be improved when
+    calibrating with overt gaze fixation.
+    For participants PA1, PB2 and PC3, this was the case when operating in the
+    covert \ac{vsa} setting with central gaze fixation and calibrating with
+    overt gaze fixation.
+  }
+  \label{fig:patients/cross}
+\end{figure}
+
+\autoref{fig:patients/cross} shows that, on average, covert~\ac{vsa} decoding
+can be improved when training in overt \ac{vsa}.
+This was especially the case for participants PA1, PB2 and PC3.
+Note that, according to eye tracking data, patients PB1, PB4 and PC4 did not
+always perform the cued central gaze fixation in the covert \ac{vsa} setting,
+which might affect the results.
 
@@
 occur in individuals with \ac{sspgi}.
-We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
-chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
+We have also confirmed part of the results from~\textcite{VanDenKerchove2024}
+presented in~\autoref{sec:covert-align} that state that decoding of covert \ac{vsa} with central gaze
 fixation can be improved by accounting for latency jitter, and showed that this
@@
 gaze was uncued.
-This is also supported by our results on cross condition calibration\todo{how}
-
-\todo{discuss that training in overt always increases covert/free, focus on
-increased P3 amplitude and relying on visual components, also propose this
-strategy if possible}
+This is also supported by our results on cross condition calibration presented
+in~\autoref{sec:patients/outcomes/cross}, which show that leveraging residual
+eye motor control to fixate targets during the calibration phase can improve
+performance in some settings.
+This is likely due to the increased P3 component amplitude in overt \ac{vsa},
+which improves the discriminative power of a classifier trained on this data.
+Cueing this overt gaze fixation only during the calibration phase leaves user
+free to operate in the manner that is most comfortable for them in  the
+operation phase.
+Early VEPs in the training data could also contribute in those cases where the participant was not
+able to perform covert \ac{vsa} with central gaze fixation.
 
@@
 
-Despite results that prompt reflection on gaze-independent \ac{bci} approaches,
+Despite results that prompt interesting reflections on gaze-independent \ac{bci} approaches,
 there are some limitations to the presented results that need to be
@@
 
-First and foremost, this study works with a limited sample size which
+First and foremost, this study works with a limited sample sizekkjk which
 does not represent the full spectrum of individuals with \ac{sspi} and
@@
 Performance might after all be traded off for user comfort.
-Eye motor disability can also have been assessed more objectively, using \todo{
-
-The Coma Recovery Scale
-Revised (CRS-R) [73] provides standardized procedures for admin-
-istration and response criteria to more uniformly describe the vis-
-ual fixation or pursuit skills of a user with SSPI. The Northeastern
-State University College of Optometry (NSUCO) Oculomotor Test
-[74] provides specific procedural techniques as well as scoring cri-
-teria, which allows for uniform description of the accuracy and
-ability of pursuit and saccade skills for adults and children.
-}
+Eye motor disability can also have been assessed more
+objectively~\cite{FriedOken2020}, using e.g.\
+the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
+oculomotor~\cite{Maples1992}.
 
@@
     but this does not hold true if central gaze fixation is not imposed.
-  \item We have shown that covert \ac{vsa}
-    decoding can alternatively be improved by
+  \item We have shown that decoding of covert \ac{vsa}
+    can alternatively be improved by
     using a calibration session with attempted overt \ac{vsa}.
-\end{enumerate*} \todo{with or without imposed gaze fixation}
+\end{enumerate*}
 
@@
 comfort and performance.
-
-\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
-\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
-\todo{BCI principles and practice chapter 19}
-\todo{https://www.nature.com/articles/s44222-024-00239-5}

### 2024-10-07 chapters/wcble.tex
All text

@@
 
-\subsection{Neural origins of \ac{erp} latency jitter}
+\subsection{Neural origins of \acs{erp} latency jitter}
 While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
@@
 
-\subsection{Implications for \ac{erp} analysis}
+\subsection{Implications for \acs{erp} analysis}
 
@@
 }{%
-  Schematic representation of the \ac{wcble} training phase.
+  Schematic representation of the \acs{wcble} training phase.
 }{%
@@
 }{%
-  Schematic representation of the \ac{wcble} test phase.
+  Schematic representation of the \acs{wcble} test phase.
 }{%
@@
 \begin{subappendices}
-\section{\Acl{cble} algorithm}
+\section{\Acs{cble} algorithm}
 \label{sec:wcble/app/cble-alg}
@@
 \clearpage
-\section{\Acl{wcble} algorithm}
+\section{\Acs{wcble} algorithm}
 \label{sec:wcble/app/wcble-alg}

### 2024-10-07 code/patients/fig_cross.py

@@ -7,14 +7,10 @@ from setup import colors, configure_matplotlib_style, save_pgf_trim
 
 configure_matplotlib_style()
 
-df = []
-for pipeline in ["tLDA", "WCBLE", "XDAWNCov+TS+LDA"]:
-    df_pipe = pd.read_csv(f"data/patients/decode/decode_results_{pipeline}.csv")
-    df_pipe["pipeline"] = pipeline
-    df.append(df_pipe)
-df = pd.concat(df, ignore_index=True)
+df = pd.read_csv('data/patients/results_cross.csv')
 df = df[df["subject"] != "PD01"]
 df = df[df["subject"] != "PB03"]
+print(df)
 
 df_avg = df.copy()
 df_avg["subject"] = "avg."
@@ -28,34 +24,36 @@ df = df.replace({
     'PC02': 'PC2',
     'PC03': 'PC3',
     'PC04': 'PC4',
-
+    'overt': 'overt VSA',
+    'covert': 'covert VSA',
+    'free': 'free VSA',
 })
+
+df['test setting'] = df['cond_test']
+df['train setting'] = df['cond_train']
 
 fg = sns.catplot(
     data=df,
     x="subject",
     y="test_score",
-    hue="pipeline",
+    hue="train setting",
     kind="bar",
-    col="condition",
+    col="test setting",
     legend="auto",
     legend_out=False,
+    col_order=['overt VSA', 'covert VSA', 'free VSA']
 )
 fg.set(ylim=(0, 1))
-plt.legend(
-    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
-)  # Adjust ncol to fit all labels in one line
+for ax in fg.axes.flat:
+    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
+    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
+fg.set_axis_labels("patient", "ROC-AUC")
+fg.set_titles("{col_name}")
+
 
 fg._legend.set_frame_on(True)
 fg._legend.get_frame().set_facecolor("white")
 fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
 
-
-
-for ax in fg.axes.flat:
-    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
-    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
[…]

### 2024-10-07 code/setup.py

@@ -77,12 +77,20 @@ def save_pgf_trim(fig, ax,path, width=textwidth_in, height=None):
         golden_ratio = (5**0.5 - 1) / 2
         height=width*golden_ratio
     fig.set_size_inches(width,height)
-    plt.draw()
-    left_margin_in = ax.get_position().x0*width
-    print(left_margin_in)
-    fig.set_size_inches(width+left_margin_in,height)
-    plt.savefig(path, format="pgf")
-    prepend = "\\hspace{-"+str(left_margin_in)+"in}\n"
+    # Get the figure DPI (dots per inch)
+    dpi = fig.dpi
+    # Get the position of the "left" spine (y axis) using the ax object
+    left_spine_vertices = ax.spines['left'].get_path().vertices
+    # Transform these vertices to figure coordinates (from pixels)
+    left_spine_transformed = ax.spines['left'].get_transform().transform(left_spine_vertices)
+    # Extract the y-coordinate (in pixels) of the bottom-left corner (from left spine)
+    bottom_left_corner_y_pixels = left_spine_transformed[0][1]
+    # Convert y-coordinate from pixels to inches
+    margin_left = bottom_left_corner_y_pixels / dpi
+    print(margin_left)
+    fig.set_size_inches(width+margin_left,height)
+    plt.savefig(path, format="pgf", bbox_inches='tight', pad_inches=0)
+    prepend = "\\hspace{-"+str(margin_left)+"in}\n"
     with open(path, 'r') as original:
         data = original.read()
     with open(path, 'w') as modified:

### 2024-10-07 chapters/abstract.tex
Intro, P300 stimulation
cover page
Abstract

@@
-Main contribution: improve gaze-independent decoding by isolating component
-related to visuospatial attention from visual components
+\chapter*{Abstract}
+Individuals with severe motor impairments, such as those with \ac{lis}
+or eye-motor dysfunction, face substantial challenges when using traditional
+gaze-dependent brain-computer interfaces \acp{bci}.
+These systems require users to direct their gaze toward specific targets, a
+task that becomes unfeasible for individuals with limited or no eye control.
+This work addresses this limitation by developing gaze-independent
+\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
+their attention to a target without corresponding eye movements.
+A key contribution of this work is the compensation for ERP latency jitter, a
+variability that negatively impacts decoding performance in covert \ac{vsa}.
+By handling this jitter, the proposed method enhances communication accuracy
+without requiring gaze control, making \acp{bci} more usable for individuals with
+motor impairments.
+
+Beyond gaze independence, the work also advances general \ac{erp} decoding
+techniques by refining the structure of linear and multilinear estimators.
+These methods improve classification accuracy across a range of \ac{bci}
+conditions, particularly when training data are limited.
+The introduction of structured regularization in both linear and multilinear
+models enhances the interpretability of the classifiers and reduces training
+time and computational complexity.
+This allows for more efficient training and better performance on unseen data,
+contributing to more reliable systems that are adaptable to various contexts and user needs.
+
+The proposed methods were validated in experiments involving both healthy
+individuals and 7 individuals with severe physical impairment and impaired eye-motor control.
+Data were recorded from a control group, as well as from seven individuals with
+gaze impairments.
+These experiments demonstrated the robustness of the novel decoding methods,
+showing that the system could effectively decode covert attention even when
+direct gaze was impossible.
+
+\paragraph{Keywords:} brain-computer interface, gaze independence, covert
+visuospatial attention, event-related potentials, linear decoding, multilinear
+decoding, motor impairment, gaze impairment

### 2024-10-07 chapters/abstract_nl.tex
Abstract

\chapter*{Samenvatting}

Personen met ernstige motorische beperkingen, zoals degenen met het
Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
aanzienlijke uitdagingen bij het gebruik van traditionele visuele brain-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
onuitvoerbaar wordt voor personen met beperkte of
geen controle over oogbewegingen.
Dit werk richt zich op deze beperking door
BCI-methoden onafhankelijk van oogbewegingen te ontwikkelen, met een focus op
het verbeteren van de bedekte visuospatiële aandacht (VSA), waarbij gebruikers
hun aandacht kunnen richten op een doel zonder overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van de communicatie zonder afhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met motorische beperkingen van
de ogen.

Naast onafhankelijkheid van oogbewegingen, bevordert dit werk ook de algemene
ERP-decodeertechnieken door de structuur van lineaire en multilineaire
decodeertechnieken te verfijnen.
Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
BCI-condities, met name wanneer trainingsgegevens beperkt zijn.
De introductie van gestructureerde regularisatie in zowel lineaire als
multilineaire modellen verhoogt de interpretateerbaarheid van de classificators
en vermindert de trainingstijd en computationele complexiteit.
Dit maakt efficiëntere training mogelijk en zorgt voor betere prestaties bij
ongeziene data, wat bijdraagt aan betrouwbaardere systemen die zich kunnen
aanpassen aan verschillende contexten en gebruikersbehoeften.

De voorgestelde methoden zijn gevalideerd in experimenten met zowel gezonde
individuen als zeven individuen met een ernstige fysieke beperking en verminderde
oog-motorische controle.
Deze experimenten toonden de robuustheid van de nieuwe decodeermethoden aan,
waarbij werd aangetoond dat het systeem bedekte aandacht effectief kon decoderen,
zelfs wanneer directe blik onmogelijk was.

\paragraph{Trefwoorden:} brain-computer interface, blikonafhankelijkheid, verborgen visuospatiële aandacht, event-gerelateerde potentialen, lineaire decodering, multilineaire decodering, motorische beperking, blikstoornis.

### 2024-10-07 chapters/bci.tex
Abstract

@@
-\todo{comments liesel paradigms}
 \todo{spellcheck}
@@
 
-\todomvh{"trial" could  be confused with intensification rounds e.g.}
-\todo{remove trial alltogether}
 Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
@@
 \end{figure}
-
-\todo{four pane figure:
-  photo of someone using a P300 bci
-  full spatiotemporal ERP response
-}
+\todo{full spatiotemporal erp response}
 
@@
 
-\emph{The N1 component} peaks around 150 to 200 ms post-stimulus.
+\paragraph{The N1 component} peaks around 150 to 200 ms post-stimulus.
 It represents an extension of sensory processing related to both gaze fixation
@@
 
-\emph{The P2 component} occurs between 200 to 300 ms.
+\paragraph{The P2 component} occurs between 200 to 300 ms.
 It reflects higher-order processing, particularly in contexts demanding
@@
 
-\emph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
+\paragraph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
 control processes such as conflict monitoring and inhibition.
@@
 
-\emph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
+\paragraph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
 post-stimulus.
@@
 
-\emph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
+\paragraph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
 Sustained attention and emotional processing of stimuli are reflected in the

### 2024-10-07 chapters/conclusion.tex
Abstract

@@
 structure~\cite{Bijma2005}.
+The Kronecker-lda approach has also be extended to the combined
+space-time-frequency domain, where more information can be modeled at the cost
+of increased dimensionality.
+A properly structured covariance model can strongly regularize the problem.
 
@@
 than overt \ac{vsa} since it should be able to account for the presence of both
-visual and attentional \ac{erp}components}.
+visual and attentional \ac{erp}components.
 
@@
 and take into account the human aspect of working \emph{with} \ac{bci} users.
-
-\section{Conclusion}

### 2024-10-07 chapters/bci.tex
checking bttda

@@
+\todo{comments liesel paradigms}
 \todo{spellcheck}
@@
 \todo{acro usage in chapters}
+\todo{are textcites used when necessary?}
 
 \chapter{Brain-computer interfaces}
-\todo{are textcites used when necessary?}
 \label{sec:bci}
-People with \ac{sspi} due to
-neurodegenenerative disease, stroke, traumatic brain or spine injury
-become entirely reliant on caregivers or family members in their day to day
+People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
+injury become entirely reliant on caregivers or family members in their day-to-day
 routine.
-Wouldn't it be great if we could help them regain some quality of life by
-making them more independent?
+Wouldn't it be great if we could help them regain some quality of life by making them
+more independent?
 
-The most basic ability any independent human being needs, is interaction with
-their environment.
-In its core essence, this is the ability to communicate your intentions,
-emotions, frustrations or thoughts such that they might be acted on.
-For an able-bodied person, physical interaction happens through body movements
-while communication is usually done through speech and body language
-Both require sending signals through the body's nervous system to control
-muscles.
+The most basic ability any independent human being needs is interaction with their
+environment.
+In its core essence, this is the ability to communicate intentions, emotions,
+frustrations, or thoughts such that they might be acted on.
+For an able-bodied person, physical interaction happens through body movements, while
+communication is usually done through speech and body language.
+Both require sending signals through the body's nervous system to control muscles.
 Not everyone has this capability.
 
-The key problem for severely paralyzed patients is that the mind wants to go
-where the body cannot follow.
-Couldn't we therefore design a system that directly interacts with the mind?
-This system can then form an interface between the mind and any kind of
-technology, like a robot arm or speech synthesizer, interacting with the world
-on the user's behalf.
-Such a device is the \ac{bci}\footnote{Sometimes also
-termed brain-machine interface (BMI) when coupled to a physical actuator, like
-a robot arm or an exoskeleton.
-The term \ac{bci} is usually preferred for
-assistive technology and communication devices.
-},
-a system that directly couples actions in the `real world' to the mental state of
-the user.
+The key problem for severely paralyzed patients is that the mind wants to go where the
+body cannot follow.
+Couldn't we, therefore, design a system that directly interacts with the mind?
+This system can then form an interface between the mind and any kind of technology, like
+a robot arm or speech synthesizer, interacting with the world on the user's behalf.
+Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
+(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
+The term \ac{bci} is usually preferred for assistive technology and communication
+devices.}, a system that directly couples actions in the `real world' to the mental state
+of the user.
 
-Starting in the 1970's from methods to establish basic communication from
-minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
-into sophisticated interaction schemes that restore or enhance many capabilities
-by bypassing defective parts of the nervous or muscle system.
-By now, classic \acp{bci} methods are starting to mature outside the lab
-setting and can be used as assistive communication technology by patients with
-impaired speech~\cite{Wolpaw2018,Milekovic2018}.
-Impressive cutting edge experimental examples include a brain-spine
-interface allowing a paraplegic patient to walk again\cite{Lorach2023},
-a \ac{bci} translating internal speech to a facial avatar speaking the imagined
-text~\cite{Metzger2023}, and fast communication through decoding imagined
-handwritten symbols~\cite{Willett2021}.
+Starting in the 1970s from methods to establish basic communication from minimal brain
+activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
+into sophisticated interaction schemes that restore or enhance many capabilities by
+bypassing defective parts of the nervous or muscle system.
+By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
+be used as assistive communication technology by patients with impaired
+speech~\cite{Wolpaw2018,Milekovic2018}.
+Impressive cutting-edge experimental examples include a brain-spine interface allowing a
+paraplegic patient to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
+speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
+communication through decoding imagined handwritten symbols~\cite{Willett2021}.
 Recently, the idea has also gained popular traction through Elon Musk's
@@
 
-In general, devices processing direct inputs from the central nervous system
-are useful for rehabilitation and medical diagnosis or treatment.
-Additionally, they bring a new paradigm to human-computer interaction,
-especially when paired with virtual or augmented reality~\cite{SiMohammed2017}.
-Ultimatly, they have always been especially promising as assistive technology for
+In general, devices processing direct inputs from the central nervous system are useful
+for rehabilitation and medical diagnosis or treatment.
+Additionally, they bring a new paradigm to human-computer interaction, especially when
+paired with virtual or augmented reality~\cite{SiMohammed2017}.
+Ultimately, they have always been especially promising as assistive technology for
 paralyzed patients limited in their communication ability.
-It is this \ac{bci} controlled communication is where our focus lies,
-as it offers a transformative means for individuals with severe motor
-impairments to regain functional independence to write and speak through
-assistive technology.
-This allows them to talk to their loved ones and caregivers, exercise hobbies
-and makes many aspects of daily life more accessible.
+It is this \ac{bci}-controlled communication where our focus lies, as it offers a
+transformative means for individuals with severe motor impairments to regain functional
+independence to write and speak through assistive technology.
+This allows them to talk to their loved ones and caregivers, exercise hobbies, and makes
+many aspects of daily life more accessible.
 
@@
 
-Simply put, a \ac{bci} records the user's brain activity, extracts some
-relevant output from this brain activity and couples this output to a function
-of a device as illustrated in Figure~\ref{fig:bci/loop}
-Optionally, the user can then observe the action of the device and adapt their
-behavior or brain activity accordingly, closing the loop\footnote{%
-  \textcite{BCISociety2024} has recently formalized this into the
-  following definition:
-  \it``A brain-computer interface is a system that measures brain activity and
-  converts it in (nearly) real-time into functionally useful outputs to replace,
-  restore, enhance, supplement, and/or improve the natural outputs of the brain,
-  thereby changing the ongoing interactions between the brain and its external or
-  internal environments. It may additionally modify brain activity using targeted
-  delivery of stimuli to create functionally useful inputs to the
-  brain.''
-}.
+Simply put, a \ac{bci} records the user's brain activity, extracts some relevant output
+from this brain activity, and couples this output to a function of a device as
+illustrated in \autoref{fig:bci/loop}.
+Optionally, the user can then observe the action of the device and adapt their behavior
+or brain activity accordingly, closing the loop\footnote{\textcite{BCISociety2024} has
+recently formalized this into the following definition: \it``A brain-computer interface
+is a system that measures brain activity and converts it in (nearly) real-time into
+functionally useful outputs to replace, restore, enhance, supplement, and/or improve the
+natural outputs of the brain, thereby changing the ongoing interactions between the
+brain and its external or internal environments.
+It may additionally modify brain activity using targeted delivery of stimuli to create
+functionally useful inputs to the brain.''}.
 
@@
   visual stimulation.
-  Electrical neural signals are recorded, and neurophysiological features
-  related to the paradigm are extracted. Using machine learning, a decision can
-  then be made based on these features, which can be presented back to the user.
+  Electrical neural signals are recorded, and neurophysiological features related to the
+  paradigm are extracted.
+  Using machine learning, a decision can then be made based on these features, which can
+  be presented back to the user.
   In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
@@
 Let's break this definition down and focus on its separate parts.
-First of all, we need to identify a signal that is a direct representation of
-what is going on in the brain.
-Multiple signals can be recorded from the brain, including bloodflow,
-magnetical electrical activity.
-For several reasons laid out in~\ref{sec:bci-recording}, the electrical modality
-is especially interesting.
+First of all, we need to identify a signal that is a direct representation of what is
+going on in the brain.
+Multiple signals can be recorded from the brain, including blood flow and magnetic or
+electrical activity.
+For several reasons laid out in \autoref{sec:bci-recording}, the electrical modality is
+especially interesting.
 
-The recorded brain activity forms only one part of the interaction scheme presented
-in Figure~\ref{fig:bci/loop}.
-It would make little sense to start digging around at random in all the brain
-activity to try to extract the desired output.
+The recorded brain activity forms only one part of the interaction scheme presented in
+\autoref{fig:bci/loop}.
+It would make little sense to start digging around at random in all the brain activity
+to try to extract the desired output.
 Instead, it would help greatly if we knew which specific activity to look for.
-A \ac{bci} operates according to a specified \emph{paradigm}, which defines the
-method by which interaction is conducted.
+A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
+by which interaction is conducted.
 This usually means that the user is instructed to perform a specific task, like
 attending a specific flickering stimulus.
-We can now exploit background knowledge from prior neuroscientific research
-to couple specific activity in the brain to specific information or actions we
-want to convey via the \ac{bci}.
-The brain singal now \emph{modulates} this information.
-In turn, the detection of this signal can be coupled to an action, like
-typing the inteded letter A or B in Figure~\ref{fig:bci/loop}.
-The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
-exact task performed by the user, and is often closely linked to the manner of
-feedback in the case of a closed-loop system.
-As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
-paradigms based on different systems in the brain.
+We can now exploit background knowledge from prior neuroscientific research to couple
+specific activity in the brain to specific information or actions we want to convey via
+the \ac{bci}.
+The brain signal now \emph{modulates} this information.
+In turn, the detection of this signal can be coupled to an action, like typing the
+intended letter A or B in \autoref{fig:bci/loop}.
+The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
+task performed by the user, and is often closely linked to the manner of feedback in the
+case of a closed-loop system.
+As shown in \autoref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
+based on different systems in the brain.
 
-The previous points to another component of a successful \ac{bci}: the specific
-brain signatures related to our paradigm need to be identified within the recorded
+The previous points to another component of a successful \ac{bci}: the specific brain
+signatures related to our paradigm need to be identified within the recorded activity.
+The electrical activity of the brain is relatively weak compared to that of the
+environment around us, containing electronic devices and other sources of interference.
+Furthermore, the brain is continuously processing information and carrying out
+`background tasks,' all of which generate their own brain activity.
+On the other hand, the activity of interest often originates from a specific region or
+network within the brain.
+This activity might not easily be discerned from all other brain and environmental
 activity.
-The electrical activity of the brain is relatively weak compared to that of the
-environment around us, containing electronic devices and other sources of
-interference.
-Furthermore, the brain is continuously processing information and carrying out
-`background tasks', all of which generate their own brain activity.
-On the other hand, the activity of interest often
-originates from a specific region or network within the brain.
-This activity might not easily be discerned from all other brain and
-environmental activity.
-Compare this to trying to listen to a single speaker at a noisy party where
-everyone is shouting over each other.
-Some obvious sources of interference can be easily filtered out by
-\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}) and by choosing
-the correct representation of the signal (i.e. \emph{feature extraction}, but the problem
-of retrieving or \emph{decoding} only the relevant signal is generally hard.
+Compare this to trying to listen to a single speaker at a noisy party where everyone is
+shouting over each other.
+Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
+(see \autoref{sec:bci/preprocessing}) and by choosing the correct representation of the
+signal (i.e., \emph{feature extraction}), but the problem of retrieving or
+\emph{decoding} only the relevant signal is generally hard.
 Nevertheless, we can attempt to solve it through supervised machine learning.
-Useful information that is encoded in the brain signal by the paradigm,
-can be retrieved from it by the decoder and interpreted in function of the
-desired \ac{bci} output.
-Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
-techniques.
+Useful information that is encoded in the brain signal by the paradigm can be retrieved
+from it by the decoder and interpreted in function of the desired \ac{bci} output.
+\autoref{sec:bci/decoding} takes a closer look at some common decoding techniques.
 
-Finally, the loop can be closed by coupling this to a
-device or actuator.
+Finally, the loop can be closed by coupling this to a device or actuator.
 There are two aspects to this.
-On the one hand, the \ac{bci} gains its function by allowing the user to interact
-with their environment.
-For a communication system, this means coupling the decoded information to
-e.g.\ a virtual keyboard or speech synthesizer.
-On the other hand, the actions performed by the \ac{bci} can themselves alter
-the user's cognitive state or brain activity, creating an adaptive system.
+On the one hand, the \ac{bci} gains its function by allowing the user to interact with
+their environment.
+For a communication system, this means coupling the decoded information to, e.g., a
+virtual keyboard or speech synthesizer.
+On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
+cognitive state or brain activity, creating an adaptive system.
 A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
-Examples of this are deep brain stimulation to mitigate the symptoms of
-Parkinson's disease, or the cochlear implant as a hearing prosthesis.
+Examples of this are deep brain stimulation to mitigate the symptoms of Parkinson's
+disease, or the cochlear implant as a hearing prosthesis.
 More indirectly, modulation can be achieved by sensory stimulation.
-This involves specific, paradigm-related sensory input, like tactile
-feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
-back to the user.
-The user's brain will then adapt through reinforcement learning, causing
-changes in behavior or strategy.
-Neuroplasticity, the brain's ability to adapt and reinforce specific neural
-pathways, can have a positive impact on \ac{bci} performance and gives rise to
-rehabilitation applications.
+This involves specific, paradigm-related sensory input, like tactile feedback for
+movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
+user.
+The user's brain will then adapt through reinforcement learning, causing changes in
+behavior or strategy.
+Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
+can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
+applications.
 
-To summarize, a \ac{bci} can facilitate direct interaction central nervous system.
-To do this, information is modulated in the brain signal according to the
-\ac{bci} paradigm by stimulating, the user or having them perform a task.
+To summarize, a \ac{bci} can facilitate direct interaction with the central nervous
+system.
+To do this, information is modulated in the brain signal according to the \ac{bci}
+paradigm by stimulating the user or having them perform a task.
 Their brain signal is subsequently recorded and preprocessed to clean it.
-Relevant features that contain aspects of the modulated signal are extracted,
-and the modulate information is retrieved by the decoder, a machine learning
-model.
-This information can then acted on and the loop is closed if this action
-affects the user.
+Relevant features that contain aspects of the modulated signal are extracted, and the
+modulated information is retrieved by the decoder, a machine learning model.
+This information can then be acted on, and the loop is closed if this action affects the
+user.
 Let us take a look at each of these elements separately in the next section.
@@
 
-The brain's activity can be recorded using various neuroimaging
-technologies.
-These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
-(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
-(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
+The brain's activity can be recorded using various neuroimaging technologies.
+These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance
+imaging (fMRI) to more portable technologies like acoustic signals obtained by
+functional ultrasound imaging (fUS)~\cite{Zheng2023} and near-infrared spectroscopy
 (fNIRS)~\cite{Borgheai2020}.
 
-The technologies above all measure bloodflow (the hemodynamic signal), which only indirectly
-reflects brain activity.
-This signal reacts too slowly to reflect the real-time activity that is
-of interest in \ac{bci} for communication, and it carries little information
-other than the brain region from where it is originating.
+The technologies above all measure blood flow (the hemodynamic signal), which only
+indirectly reflects brain activity.
+This signal reacts too slowly to reflect the real-time activity that is of interest in
+\ac{bci} for communication, and it carries little information other than the brain
+region from where it is originating.
 
@@
 the smallest units of information processing.
-The \emph{action potential} and its related \emph{postsynaptic potential}
-is an electrical pulse occurring when a neuron receives
-input from another neuron and is activated.
-The firing of a neuron, or the combined firing of groups of
-synchronized neurons, thus generate an electrical field in and near the brain,
-which can be measured using electrodes.
+The \emph{action potential} and its related \emph{postsynaptic potential} are electrical
+pulses occurring when a neuron receives input from another neuron and is activated.
+The firing of a neuron, or the combined firing of groups of synchronized neurons, thus
+generates an electrical field in and near the brain, which can be measured using
+electrodes.
 The related discipline is \emph{electrophysiology}.
 
-Because of the desirable high temporal resolution~\cite{Easttom2021} and
-practicality of electrophysiology, we will focus only on methods to record
-electromagnetical activity originating from neuronal action potentials and
-postsynaptic potentials.
-The magnetical field of the brain can be measured using magnetoencephalography
+Because of the desirable high temporal resolution~\cite{Easttom2021} and practicality of
+electrophysiology, we will focus only on methods to record electromagnetic activity
+originating from neuronal action potentials and postsynaptic potentials.
+The magnetic field of the brain can be measured using magnetoencephalography
 (MEG)~\cite{Mellinger2007}.
-While MEG is non-invasive and results in high-quality signals, the necessary
-equipment is rather expensive and impractical.
+While MEG is non-invasive and results in high-quality signals, the necessary equipment
+is rather expensive and impractical.
 Recent advances have been made using optically pumped magnetometers
-(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
-lab conditions.
-As a consequence, the \ac{bci} field relies heavily on the
-electrical activity of the brain.
+(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of lab
+conditions.
+As a consequence, the \ac{bci} field relies heavily on the electrical activity of the
+brain.
 
 \subsection{Invasive electrophysiology}
-The \emph{invasiveness} of the technology forms an
-important concept in determining its suitability for a given application and
-patient.
-As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often
-measure a very specific brain region and can result in better signal quality, at the cost of the risks
-involved with brain surgery.
+
+The \emph{invasiveness} of the technology forms an important concept in determining its
+suitability for a given application and patient.
+As illustrated in \autoref{fig:bci/recording}, invasive electrodes can often measure a
+very specific brain region and can result in better signal quality, at the cost of the
+risks involved with brain surgery.
 
@@
 }{%
-	Various electrophysiology recording technologies and \\
-	their respective position in a cross-section of the skull. \\
-	There is a trade-off between invasiveness and  \\
-	spatial resolution.
+	Various electrophysiology recording technologies and their respective position in a
+	cross-section of the skull.
+	There is a trade-off between invasiveness and spatial resolution.
 }{%
@@
 
-Specifically, invasive methods are less noisy and come with an increased \emph{spatial resolution},
-meaning they can extract a signal from a specific brain region or even a set of
-neurons of interest.
-This makes it easier to retrieve a signal with known origin from the brain,
-which might be leveraged by the paradigm to modulate information.
+Specifically, invasive methods are less noisy and come with an increased \emph{spatial
+resolution}, meaning they can extract a signal from a specific brain region or even a
+set of neurons of interest.
+This makes it easier to retrieve a signal with a known origin from the brain, which
+might be leveraged by the paradigm to modulate information.
 Generally, microelectrodes penetrating the cortex are considered to have the highest
 spatial resolution.
-These are needles with a diameter smaller than 10µm and one or multiple
-electrodes that extend several millimmeters into the cortex.
-Microelectrodes can measure the Local Field Potential (LFP), the extracellular
-potential of a small group of neurons, or even intracellular single neuron
-action potential spikes.
-They come in single-electrode form, or, more popularly, in microelectrode
-arrays.
-Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
-NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
-2.0~\cite{Steinmetz2021}.
+These are needles with a diameter smaller than 10µm and one or multiple electrodes that
+extend several millimeters into the cortex.
+Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
+of a small group of neurons, or even intracellular single neuron action potential
+spikes.
+They come in single-electrode form, or, more popularly, in microelectrode arrays.
+Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
+implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.
 
 Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
-Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
-and closed-loop adaptive control for Deep Brain Stimulation to mitigate
-Parkinson's disease symptoms~\cite{Arlotti2018}.
-Depth electrodes are longer rods with multiple electrodes than can extend
-deeper into the brain.
-Since these still penetrate the cortex, electrodes that sit on top of the
-cortex are more popular, as they are considered less invasive.
-\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
-Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
-higher number of smaller electrodes per surface area allow for more precise
-measurements.
+Depth electrodes such as those used in stereo-\ac{eeg} \ac{bci}~\cite{Wu2024} tasks and
+closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
+symptoms~\cite{Arlotti2018}.
+Depth electrodes are longer rods with multiple electrodes that can extend deeper into
+the brain.
+Since these still penetrate the cortex, electrodes that sit on top of the cortex are
+more popular, as they are considered less invasive.
+\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
+epilepsy diagnosis.
+Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
+number of smaller electrodes per surface area, allow for more precise measurements.
 
-Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
-have been realized using
-invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
+Some impressive recent breakthroughs in speech and motor \ac{bci} for communication have
+been realized using invasive microelectrode arrays~\cite{Willett2021} and
+$\mu$\Ac{ecog}~\cite{Metzger2023}.
 Furthermore, recent advances in recording technology focus on improving implant
@@
 tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
-The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
-through the bloodstream via a catheter, removing the need for open brain
-surgery.
+The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered through the
+bloodstream via a catheter, removing the need for open brain surgery.
 Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
 potential communication \ac{bci} user groups consistently show that non-invasive
-technology is preferred over implanted electrodes, unless the added value of invasive BCI is
-sufficiently large~\cite{Kageyama2020}.
+technology is preferred over implanted electrodes, unless the added value of invasive
+BCI is sufficiently large~\cite{Kageyama2020}.
 This provides justification for focusing part of the current research effort on
@@
 
-\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
-neurology practice as well as in \ac{bci} research.
-Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
-offers great temporal resolution up to thousands of Hz.
+\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
+practice as well as in \ac{bci} research.
+Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
+great temporal resolution up to thousands of Hz.
 \Ac{eeg} measures the electrical potential over time at a given electrode relative to a
-reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
-As it is easily applicable on the outside of the scalp, many practical
-consumer-grade \ac{eeg} headsets exist in the form of (often wireless) electrode
-caps or helmets.
-Current consumer systems sometimes feature dry electrodes, but clinical and
-research systems often use active electrodes with preamplifiers and
-electrolyte gel to reduce impedance for improved signal
-quality, which slightly decreases practicality, since they require a trained
-individual to properly administer.
-Nevertheless, it is the most accessible \ac{bci} technology for both patients
-and researchers.
+reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
+As it is easily applicable on the outside of the scalp, many practical consumer-grade
+\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
+Current consumer systems sometimes feature dry electrodes, but clinical and research
+systems often use active electrodes with preamplifiers and electrolyte gel to reduce
+impedance for improved signal quality, which slightly decreases practicality, since they
+require a trained individual to properly administer.
+Nevertheless, it is the most accessible \ac{bci} technology for both patients and
+researchers.
 
 The major drawback of \ac{eeg} is its poor spatial resolution~\cite{Ferree2001}.
-Since they sit on the outside of the scalp electrodes are large and distance
-relative to the neurons in the cortex they ought to measure.
-Furthermore, the layers of skull and cerebrospinal fluid in between can
-attenuate the signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
+Since they sit on the outside of the scalp, electrodes are large and distant relative to
+the neurons in the cortex they ought to measure.
+Furthermore, the layers of skull and cerebrospinal fluid in between can attenuate the
+signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
 measuring a mixture of activity from sources elsewhere in the brain.
 This also has a negative effect on spectral resolution.
-The noise problem mentioned earlier is also prominently present in \ac{eeg}
-recordings.
-In addition to picking up brain activity other than that of the region of
-interest, \ac{eeg} also records other artifacts, from the environment, the power
-supply, nearby electrical equipment, or the user's muscle
-movements~\cite{Urigueen2015}.
+The noise problem mentioned earlier is also prominently present in \ac{eeg} recordings.
+In addition to picking up brain activity other than that of the region of interest,
+\ac{eeg} also records other artifacts from the environment, the power supply, nearby
+electrical equipment, or the user's muscle movements~\cite{Urigueen2015}.
 Combined, this means \ac{eeg} is inherently noisy.
 
-Despite its noisy nature, \ac{eeg} is the recording methodology of choice for
-our \ac{bci} because of its wide acceptance by patients.
-To counteract the noise present in our recording, we must on the one hand evoke
-strong, informative brain activity using a suited \ac{bci} paradigm, and on the
-other hand pick a suited decoder which can isolate the signal of interest and
-filter out the noise.
+Despite its noisy nature, \ac{eeg} is the recording methodology of choice for our
+\ac{bci} because of its wide acceptance by patients.
+To counteract the noise present in our recording, we must on the one hand evoke strong,
+informative brain activity using a suited \ac{bci} paradigm, and on the other hand pick
+a suited decoder which can isolate the signal of interest and filter out the noise.
 
@@
 
-The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity into useful
-output actions, since it defines which neural responses will be elicited and
+The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity
+into useful output actions, since it defines which neural responses will be elicited and
 captured as features.
-A paradigm can be loosely seen as \emph{a way of interacting with the
-\ac{bci}}.
-According to \textcite{Zander2011}, they can be arranged by their reliance on
-external stimulation and the engagement of the user as shown in
-Figure~\ref{fig:bci/paradigms}.
+A paradigm can be loosely seen as \emph{a way of interacting with the \ac{bci}}.
+According to \textcite{Zander2011}, they can be arranged by their reliance on external
+stimulation and the engagement of the user as shown in \autoref{fig:bci/paradigms}.
 
@@
   A \ac{bci} paradigm defines which brain activity will be used for control.
-  They can be categorized in active, reactive and passive
-  paradigms, depending on their reliance on external stimulation and active
-  participation required from the user.
+  They can be categorized in active, reactive and passive paradigms, depending on their
+  reliance on external stimulation and active participation required from the user.
   Adapted from~\textcite{Muehl2014}.
@@
 
-Paradigms that require little active participation of the user like
-redirecting attention or initiating imagined actions are classified as \emph{passive
-\ac{bci}}.
-From the user's point of view, this concept is probably the most attractive,
-since their cognitive state state would be inferred from just
-their brain activity.
+Paradigms that require little active participation from the user, like redirecting
+attention or initiating imagined actions, are classified as \emph{passive \ac{bci}}.
+From the user's point of view, this concept is probably the most attractive, since their
+cognitive state would be inferred from just their brain activity.
 However, without active user participation, it can be hard to establish reliable
 communication or control.
-Therefore, passive \ac{bci} are currently more suited to tasks like
-emotion or affect monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload
-monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
-detection~\cite{SiMohammed2020}.
+Therefore, passive \ac{bci} are currently more suited to tasks like emotion or affect
+monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload monitoring~\cite{Zanetti2021},
+or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
 It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
-Nevertheless, active participation facilitates training data collection for
-supervised machine learning since the conditions can be more easily controlled
-by the user or the stimulation.
+Nevertheless, active participation facilitates training data collection for supervised
+machine learning, since the conditions can be more easily controlled by the user or the
+stimulation.
 
-Paradigms requiring high active participation can be split up in \emph{active
+Paradigms requiring high active participation can be split up into \emph{active
 \ac{bci}} and \emph{reactive \ac{bci}}.
-Active \ac{bci} paradigms encode endogenous activity initiated by the user,
-such as imagined movement or imagined speech.
-These tasks can encode complex information, but current non-invasive
-communication methods often limit the considered domain to a few movement
-directions to control a cursor, or a few words~\cite{Panachakel2021}.
-Invasive recordings are necessary for decoding more complex encoded
-like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
+Active \ac{bci} paradigms encode endogenous activity initiated by the user, such as
+imagined movement or imagined speech.
+These tasks can encode complex information, but current non-invasive communication
+methods often limit the considered domain to a few movement directions to control a
+cursor, or a few words~\cite{Panachakel2021}.
+Invasive recordings are necessary for decoding more complex encoded activities like
+natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
 symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
@@
 complexity of the performed tasks.
-They might require extensive user training and are sensitive to
-correct task instruction.
-Imagined speech or movements can be performed in a multitude of ways, which
-can each have different neural representations for separate individuals.
-This can make it hard to adapt the \ac{bci} to specific individuals, causing
-poor performance and giving rise to the concept of \ac{bci}
-illiteracy~\cite{Allison2010}\footnote{%
+They might require extensive user training and are sensitive to correct task
+instruction.
+Imagined speech or movements can be performed in a multitude of ways, which can each
+have different neural representations for separate individuals.
+This can make it hard to adapt the \ac{bci} to specific individuals, causing poor
+performance and giving rise to the concept of \ac{bci} illiteracy~\cite{Allison2010}\footnote{%
 Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
-Instead of attributing poor performance to the user's inherent limitations,
-critiques point out that the issue lies more with the design of the
-systems themselves. BCI systems often fail to account for individual
-variability in brain anatomy and cognitive strategies. As a result, what has
-been labeled as `illiteracy' may, in fact, be a reflection of systems that are
-not flexible enough to adapt to different users. This perspective
-shifts the responsibility to the individual, rather than recognizing that
-better, more adaptive systems could overcome many of these performance
-challenges, thus standing in the way of progress.
-Moreover, the term `illiteracy' itself carries negative
-connotations, implying a user deficiency that reinforces normative assumptions,
-when in reality, a more inclusive approach to design could enhance BCI
-accessibility for a wider range of individuals. This evolving view calls for a
-reframing of the problem, focusing on improving the adaptability of BCIs rather
-than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
-}.
+Instead of attributing poor performance to the user's inherent limitations, critiques
+point out that the issue lies more with the design of the systems themselves.
+BCI systems often fail to account for individual variability in brain anatomy and
+cognitive strategies.
+As a result, what has been labeled as `illiteracy' may, in fact, be a reflection of
+systems that are not flexible enough to adapt to different users.
+This perspective shifts the responsibility to the individual, rather than recognizing
+that better, more adaptive systems could overcome many of these performance challenges,
+thus standing in the way of progress.
+Moreover, the term `illiteracy' itself carries negative connotations, implying a user
+deficiency that reinforces normative assumptions, when in reality, a more inclusive
+approach to design could enhance BCI accessibility for a wider range of
+individuals.
+This evolving view calls for a reframing of the problem, focusing on improving the
+adaptability of BCIs rather than labeling users as illiterate~\cite{Becker2022,Thompson2019}.}.
 
 \subsection{Reactive \acsp{bci}}
-Reactive \ac{bci} take another approach, by providing a discrete set of
-sensory stimulations towards which the user's attention can directed.
-Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
-due to the constant stimulation.
-These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
-their expressive power is limited by how many different stimuli can effectively
-presented and attended within a reasonable timeframe.
+Reactive \ac{bci} take another approach by providing a discrete set of sensory
+stimulations towards which the user's attention can be directed.
+Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue due to
+the constant stimulation.
+These paradigms are also somewhat less intuitive compared to, e.g., speech, and their
+expressive power is limited by how many different stimuli can effectively be presented
+and attended to within a reasonable timeframe.
 Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
@@
 
-Reactive \ac{bci} can be realized in multiple ways, depending on the
-stimulation used.
+Reactive \ac{bci} can be realized in multiple ways, depending on the stimulation used.
 Some examples include the following:
-In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021},
-the user can attend to one of multiple vibrotactile stimulations in different
-limbs, which encodes the information of the attended limb in the brain signal.
-In auditory paradigms, information can be modulated by the volume, tone, pitch,
-spatial origin of presented sounds, to which the user can
-attend~\cite{Kaongoen2017}.
+In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021}, the user
+can attend to one of multiple vibrotactile stimulations in different limbs, which
+encodes the information of the attended limb in the brain signal.
+In auditory paradigms, information can be modulated by the volume, tone, pitch, or
+spatial origin of presented sounds, to which the user can attend~\cite{Kaongoen2017}.
 Most common are the visual paradigms.
-These can be more performant since they allow the user to use their visual
-system, one of the most evolved sensory system in humans.
+These can be more performant since they allow the user to use their visual system, one
+of the most evolved sensory systems in humans.
 
-The major visual paradigms rely on
-\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} brain responses.
-In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
-among many that each flicker continuously at different
-frequencies~\cite{Chen2021}.
-In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
-be correlated with the brain activity to retrieve the attended
-stimulus~\cite{Sun2022}.
-Instead of stimulating all possible selections at once, the oddball paradigm
-stimulates them one by one with single flashes~\cite{Pan2022}.
-Since the times of stimulation of each target is known, and time-points at
-which an attention signature is detected to selected targets.
-\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
-movements in specific directions instead of flashing.
-Information is then also carried by movement
-direction~\cite{Libert2021a,Libert2022}.
+The major visual paradigms rely on \ac{ssvep}, \ac{cvep}, oddball, and \ac{mvep} brain
+responses.
+In \ac{ssvep}, information is modulated by the frequency of a stimulus attended among
+many that each flicker continuously at different frequencies~\cite{Chen2021}.
+In \ac{cvep}, the stimuli flicker instead with distinct on-off patterns, which can be
+correlated with the brain activity to retrieve the attended stimulus~\cite{Sun2022}.
+Instead of stimulating all possible selections at once, the oddball paradigm stimulates
+them one by one with single flashes~\cite{Pan2022}.
+Since the times of stimulation of each target are known, and time-points at which an
+attention signature is detected can be correlated to selected targets.
+\Ac{mvep} is usually similarly time-modulated, but stimuli make sudden movements in
+specific directions instead of flashing.
+Information is then also carried by movement direction~\cite{Libert2021a,Libert2022}.
 
 The \ac{bci} paradigms mentioned above can also be combined to gain performance.
-Straightforward examples are activating or deactivating \ac{ssvep} stimulation
-with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
+Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
+motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
 along multiple dimensions at once.
-\textcite{Han2023} recently used this strategy, combining frequency and phase
-coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
-efficient \ac{bci} where one of 200 targets can be accurately selected with
-only 800ms of stimulation.
+\textcite{Han2023} recently used this strategy, combining frequency and phase coding in
+\ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
+where one of 200 targets can be accurately selected with only 800ms of stimulation.
 
-\section{Preprocessing}%
+\section{Preprocessing}
 \label{sec:bci/preprocessing}
-EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
-significantly improves the quality of the recorded data for subsequent analysis and
-classification. Common preprocessing techniques in BCI systems include
-rereferencing, band-pass and notch filtering, and \ac{ica}.
 
-Rereferencing is an essential first step, where the potential of each EEG channel is
-recalculated relative to a common reference point, such as the average of all
+\ac{eeg} preprocessing is a critical step in brain-computer interface (BCI) systems, as
+it significantly improves the quality of the recorded data for subsequent analysis and
+classification.
+Common preprocessing techniques in BCI systems include rereferencing, band-pass and
+notch filtering, and \ac{ica}.
+
+Rereferencing is an essential first step, where the potential of each \ac{eeg} channel
+is recalculated relative to a common reference point, such as the average of all
 electrodes (common average reference) or a selected neutral set of electrodes.
@@
 
-Band-pass filtering is then usually applied to isolate the frequency ranges of interest by
-removing frequencies that are too high or too low to contain relevant neural signals.
-Band-pass filters eliminate slow drifts and high-frequency noise, such as
-muscle artifacts or environmental electrical interference, improving the
-\ac{snr}.
-Additionally, notch filtering can be  used to remove strong power line
-interference\footnote{Typically at 50Hz in Europe.}.
-This interference can heavily contaminate EEG recordings and even leak through
-when outside the passband of a previous filter.
-Notch filtering efficiently attenuates this narrowband noise without affecting
-the surrounding frequencies.
+Band-pass filtering is then usually applied to isolate the frequency ranges of interest
+by removing frequencies that are too high or too low to contain relevant neural signals.
+Band-pass filters eliminate slow drifts and high-frequency noise, such as muscle
+artifacts or environmental electrical interference, improving the \ac{snr}.
+Additionally, notch filtering can be used to remove strong power line interference\footnote{%
+Typically at 50Hz in Europe.}.
+This interference can heavily contaminate \ac{eeg} recordings and even leak through when
+outside the passband of a previous filter.
+Notch filtering efficiently attenuates this narrowband noise without affecting the
+surrounding frequencies.
 
 Finally, to address artifacts like eye blinks and saccades that can significantly
-distort the EEG signal, \ac{ica} is
-performed~\cite{Delorme2007}.
-\Ac{ica} is a blind source separation technique that decomposes the EEG data into
-statistically independent components. Components corresponding to eye movements
-, blinks and other bodily or environmental artifacts can be identified, either
-visually tased on their characteristic topographies and time courses, or
-statistically, and then removed.
-This leads to cleaner data for subsequent processing, especially in the case of
-visual \ac{bci} paradigms.
-Since eye movement plays a role in the paradigm, it can strongly be present in
-the signal and correlate with the task, forming a confounding factor for
-studying control based solely on brain activity.
+distort the \ac{eeg} signal, \ac{ica} is performed~\cite{Delorme2007}.
+\ac{ica} is a blind source separation technique that decomposes the \ac{eeg} data into
+statistically independent components.
+Components corresponding to eye movements, blinks, and other bodily or environmental
+artifacts can be identified, either visually based on their characteristic topographies
+and time courses, or statistically, and then removed.
+This leads to cleaner data for subsequent processing, especially in the case of visual
+\ac{bci} paradigms.
+Since eye movement plays a role in the paradigm, it can strongly be present in the
+signal and correlate with the task, forming a confounding factor for studying control
+based solely on brain activity.
 
@@
 
-\section{Decoders}%
+\section{Decoders}
 \label{sec:bci/decoding}
@@
 
-After preprocessing and extracting features, machine learning algorithms can perform \ac{bci}
-decoding.
-Machine learning decoders use some training data for which the performed tasks
-are known, and learn to recognize the activity elicited by the task in recorded data.
-The training data can be either obtained from the \ac{bci} user themselves or from
-other users.
-In the first case, the user would perform a short calibration session before
-using the \ac{bci}, where they are instructed to perform a known task.
-This calibration session can be eliminated by training the decoder on
-preexisting labeled data from other sessions and users, but this is harder due
-to the variability between subjects and sessions.
+After preprocessing and extracting features, machine learning algorithms can perform
+\ac{bci} decoding.
+Machine learning decoders use some training data for which the performed tasks are
+known, and learn to recognize the activity elicited by the task in recorded data.
+The training data can be either obtained from the \ac{bci} user themselves or from other
+users.
+In the first case, the user would perform a short calibration session before using the
+\ac{bci}, where they are instructed to perform a known task.
+This calibration session can be eliminated by training the decoder on preexisting
+labeled data from other sessions and users, but this is harder due to the variability
+between subjects and sessions.
 
-If the goal is to eliminate the per-session calibration, one can also use a
-decoder pre-trained on an existing dataset of the same task.
-This is complicated by large variability in measured brain response between subjects
-and even between different sessions within the same
-subject~\cite{Guger2009,Saha2020}.
-Therefore, pre-trained decoders must either be trained on a very large, diverse
-sample of subjects, or some form of transfer learning must be applied.
-Nevertheless, in practice pre-trained models or models using transfer learning often still
-require some per session fine-tuning, which still necessitates some
-calibration.
-Instead, rather opt for keeping the calibration time as short as possible by using
-an algorithm that can learn from very few training samples.
-This can work well, but requires strong regularization~\cite{VanDenKerchove2022}.
+If the goal is to eliminate the per-session calibration, one can also use a decoder
+pre-trained on an existing dataset of the same task.
+This is complicated by large variability in measured brain responses between subjects
+and even between different sessions within the same subject~\cite{Guger2009,Saha2020}.
+Therefore, pre-trained decoders must either be trained on a very large, diverse sample
+of subjects, or some form of transfer learning must be applied.
+Nevertheless, in practice, pre-trained models or models using transfer learning often
+still require some per-session fine-tuning, which necessitates some calibration.
+Instead, it is better to opt for keeping the calibration time as short as possible by
+using an algorithm that can learn from very few training samples.
+This can work well but requires strong regularization~\cite{VanDenKerchove2022}.
 
@@
 application.
-The number of output degrees of features, the serial or parallel nature of
-performed tasks, whether the paradigm is active or reactive, etc.\ all need to
-be taken into account.
-On of these choices is to solve a \emph{regression} or \emph{classification}
+The number of output degrees of features, the serial or parallel nature of performed
+tasks, and whether the paradigm is active or reactive, all need to be taken into
+account.
+One of these choices is whether to solve a \emph{regression} or \emph{classification}
 problem.
-In regression, a continuous output variable is predicted for a new sample,
-while classification attempts to sort it into one of a set of discrete classes.
-Regression can be useful in \acp{bci} in applications like mapping imagined
-movement to an external robot arm.
-For communication \acp{bci}, however, conveying information in a symbolic
-manner is often of interest.
-Many applications resemble virtual keyboards or map a user's actions to
-discrete words, predefined actions or letters for full control.
+In regression, a continuous output variable is predicted for a new sample, while
+classification attempts to sort it into one of a set of discrete classes.
+Regression can be useful in \acp{bci} for applications like mapping imagined movement to
+an external robot arm.
+For communication \acp{bci}, however, conveying information in a symbolic manner is
+often of interest.
+Many applications resemble virtual keyboards or map a user's actions to discrete words,
+predefined actions, or letters for full control.
 Hence, such problems often present themselves as classification tasks.
 
-\subsection{Implemetation \& evaluation}
+\subsection{Implementation \& evaluation}
 
-\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
+\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the-art
 classification algorithms for decoding.
-Classic linear methods, such as \ac{csp} feature extraction for
-motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
-and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
-relatively well, given some regularizing constraints or extensions.
-Multi-linear techniques exploiting the tensor structure of neural signals are
-also promising~\cite{Lotte2018}.
+Classic linear methods, such as \ac{csp} feature extraction for motor
+imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017}, and \ac{lda}
+for \ac{erp} classification~\cite{Sosulski2022} still perform relatively well, given
+some regularizing constraints or extensions.
+Multi-linear techniques exploiting the tensor structure of neural signals are also
+promising~\cite{Lotte2018}.
 Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
-Furthermore, they lend themselves well to application like adaptive
+Furthermore, they lend themselves well to applications like adaptive
 learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}.
-Riemannian Geometry classifiers are often considered the current
-state-of-the-art.
-Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered,
-albeit only when sufficiently large training datasets are available.
-If decoders tailored to a specific user that keep the calibration session as
-short as possible are of interest, not enough training data is available to
-properly train a deep learning model.
+Riemannian Geometry classifiers are often considered the current state-of-the-art.
+Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered, albeit
+only when sufficiently large training datasets are available.
+If decoders tailored to a specific user that keep the calibration session as short as
+possible are of interest, not enough training data is available to properly train a deep
+learning model.
 
-Usually, a decoder makes a prediction for a given \emph{trial} while operating
-the \ac{bci}.
-A trial is the smallest unit on which a selection decision can be made, for
-instance be one imagined movement, or one repetition of flashing all targets
-in a visual \ac{erp} \ac{bci}.
-Accuracy is a valuable metrics to asses decoder performance in the classification
-case.
-Accuracy is calculated as the proportion of correct selections to all
-selections made.
-It should be carefully interpreted in the presence of imbalanced data and always compared to the
-random chance accuracy level in the presence of more than two possible
+Usually, a decoder makes a prediction for a given \emph{trial} while operating the
+\ac{bci}.
+A trial is the smallest unit on which a selection decision can be made, for instance,
+one imagined movement or one repetition of flashing all targets in a visual \ac{erp}
+\ac{bci}.
+Accuracy is a valuable metric to assess decoder performance in the classification case.
+Accuracy is calculated as the proportion of correct selections to all selections made.
+It should be carefully interpreted in the presence of imbalanced data and always
+compared to the random chance accuracy level in the presence of more than two possible
 selections per trial.
-Alternatively, \ac{rocauc} is also a measure of classifier predictive power, but is more
-suited for evaluation of classification of single epochs of data and in the
-presence of imbalanced data, e.g.\ when comparing single \acp{erp}.
+Alternatively, \ac{rocauc} is also a measure of classifier predictive power but is more
+suited for evaluation of classification of single epochs of data and in the presence of
+imbalanced data, e.g., when comparing single \acp{erp}.
 Higher \ac{rocauc} usually translates to a higher target selection accuracy.
@@
 in general, is \ac{itr}.
-\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
-be calculated as
+\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can be
+calculated as
 \begin{equation}
@@
 \end{equation}
-\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
-different options that can be selected per trial\footnote{Given that selections
-are independent on each other. This formula requires some adaptations in the
-case of sequential or hierarchical selection.}, $P$, the selection accuracy
-of the decoder, and $Q$, the number of trials per second.
-The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
-building blocks a successful, high-\ac{itr} \ac{bci}.
+\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of different options
+that can be selected per trial\footnote{Given that selections are independent of each
+other. This formula requires some adaptations in the case of sequential or hierarchical
+selection.}, $P$, the selection accuracy of the decoder, and $Q$, the number of trials
+per second.
+The parameters $N$, $P$, and $Q$ of this formula give us some insight into the building
+blocks of a successful, high-\ac{itr} \ac{bci}.
 To improve performance, we can aim to
 \begin{enumerate*}[label={\arabic*})]
-\item increase $N$ by selecting a paradigm and interface that offer a
-broad range of informative selections per trial,
-\item increase $P$ by engineering more performant machine learning methods for
-decoding, and
-\item increase $Q$ by selecting a paradigm that allows fast stimulation or
-responses.
+\item increase $N$ by selecting a paradigm and interface that offer a broad range of
+informative selections per trial,
+\item increase $P$ by engineering more performant machine learning methods for decoding,
+and
+\item increase $Q$ by selecting a paradigm that allows fast stimulation or responses.
 \end{enumerate*}
@@
 Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
-The brain response of interest is the visual \ac{erp}, which can be accurately
-measured and decoded from non-invasive \ac{eeg} signals with a relatively low
-computational effort and short calibration timeand .
-First established by~\cite{Farwell1988} in 1988, it is well supported by literature
-and has been proven to work for patients in day-to-day home
-use~\cite{Wolpaw2018}.
+The brain response of interest is the visual \ac{erp}, which can be accurately measured
+and decoded from non-invasive \ac{eeg} signals with a relatively low computational
+effort and short calibration time.
+First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
+has been proven to work for patients in day-to-day home use~\cite{Wolpaw2018}.
 
 \subsection{Stimulation paradigm}
-One by one, visual elements on a computer screen are \emph{intensified} for a
-short period of time by changing in color, brightness or size.
-Intensifcations of different targets are usually 100-200ms apart and last for about
+
+One by one, visual elements on a computer screen are \emph{intensified} for a short
+period of time by changing in color, brightness, or size.
+Intensifications of different targets are usually 100-200ms apart and last for about
 100ms~\cite{Sellers2006a}.
 On each flash observed by the user, specific brain activity is elicited.
-If one of these intensified elements is rare with respect to the others, i.e. it is the
+If one of these intensified elements is rare with respect to the others, i.e., it is the
 \emph{odd-one out} or \emph{oddball}, this brain activity is altered.
-A stimulus can be rare either due to its inherent properties like color or
-location.
-Crucially, this is also the case if it is `marked' as rare by the user, e.g.
-by paying explicit attention to only a given stimulus and not the others or
-counting how many times it occurs.
+A stimulus can be rare either due to its inherent properties like color or location.
+Crucially, this is also the case if it is `marked' as rare by the user, e.g., by paying
+explicit attention to only a given stimulus and not the others or counting how many
+times it occurs.
 If these visual elements represent letters, and we know the timing of the
-intensifications of each letter, we can now establish communication by detecting
-for which letter an oddball response was present in the brain signal at its
-time of intensification.
+intensifications of each letter, we can now establish communication by detecting for
+which letter an oddball response was present in the brain signal at its time of
+intensification.
 
-\Ac{itr} in the oddball paradigm is optimized by intensifying
-multiple targets at once in a row-collumn strategy and using a sequence of
-selections, giving rise to the classic matrix speller of~\cite{Farwell1988}.
-Other optimizations increase response \ac{snr}, like using flashing face
-images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
+\Ac{itr} in the oddball paradigm is optimized by intensifying multiple targets at once
+in a row-column strategy and using a sequence of selections, giving rise to the classic
+matrix speller of~\cite{Farwell1988}.
+Other optimizations increase response \ac{snr}, like using flashing face images as
+intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
 stimuli~\cite{Treder2011}.
-The number of targets in a visual oddball \ac{bci} is limited by the crowding
-effect, which imposes a limit on how close targets can be to each other while
-not distracting the user's attention~\cite{Sellers2006a,Li2010}.
-This may be overcome by making hierarchical selections of stimuli representing
-groups of selections~\cite{Treder2010}.
+The number of targets in a visual oddball \ac{bci} is limited by the crowding effect,
+which imposes a limit on how close targets can be to each other while not distracting
+the user's attention~\cite{Sellers2006a,Li2010}.
+This may be overcome by making hierarchical selections of stimuli representing groups
+of selections~\cite{Treder2010}.
 
-Together with \ac{ssvep}, oddball paradigms are most frequently used in visual
-\ac{bci}.
-There are however indications that users prefer oddball stimulation over
-\ac{ssvep} which can cause eye strain and fatigue due to the continuous
-oscillatory stimulation of all targets at once~\cite{Xu2021}.
-Furthermore, \ac{ssvep} relies more on directing the gaze at the intended
-target than oddball, detracting from the concept of control independent from
-muscle movement.
+Together with \ac{ssvep}, oddball paradigms are most frequently used in visual \ac{bci}.
+There are, however, indications that users prefer oddball stimulation over \ac{ssvep},
+which can cause eye strain and fatigue due to the continuous oscillatory stimulation of
+all targets at once~\cite{Xu2021}.
+Furthermore, \ac{ssvep} relies more on directing the gaze at the intended target than
+oddball, detracting from the concept of control independent from muscle movement.
 
@@
   \begin{subfigure}[t]{.45\textwidth}
-  \includegraphics[width=\textwidth,
-    height=\textwidth]{figures/bci/illustration}
+  \includegraphics[width=\textwidth, height=\textwidth]{figures/bci/illustration}
     \caption{Stimuli are flashed while the \ac{eeg} is recorded}
@@
     \input{figures/bci/erp.tikz.tex}
-    \caption{The idealized \ac{erp} evoked by an attended stimulation in a
-    single channel.}
+    \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
+    channel.}
   \end{subfigure}\hfill%
@@
 
-\subsection{The \acl{erp}}%
+\subsection{The \acl{erp}}
 \label{sec:bci/oddball/erp}
 
-The time-domain response elicited immediately after the intensification of a
-stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
+The time-domain response elicited immediately after the intensification of a stimulus is
+known as the visual \ac{erp}~\cite{Luck2014}.
 An \ac{erp} is a waveform consisting of multiple components.
-Some of these components are modulated by whether the stimulus was an
-oddball or not.
-These components appear as positive peaks and negative troughs in the \ac{erp}
-waveform.
-They follow a naming scheme based on polarity (Negative or Positive) and
-latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
-The component most prominently modulated by attention is called the P3
-(positive deflection after $\pm300$ ms)
-some \ac{bci}\footnote{In \ac{erp} analysis,
-components are sometimes referred to with the timing around which they occur
-(e.g.\ N170, negative after 170ms), or by their rank in order of occurrence
-(N1, the first negative component).
-The timing nomenclature is based on average latencies for neurotypical
-individuals, but is seldom correct for specific \ac{bci} user due to the large
-variability across \ac{erp} stimulation paradigms and subjects and can also
-depend on the user's pathology.
-Furthermore, as we will see later, this work focuses specifically on the
-intra-session variability in \ac{erp} latency, explicitly assuming the latencies are not
-fixed.
+Some of these components are modulated by whether the stimulus was an oddball or not.
+These components appear as positive peaks and negative troughs in the \ac{erp} waveform.
+They follow a naming scheme based on polarity (Negative or Positive) and latency (e.g.,
+N2 or N200 for a negative component after $\pm200$ ms).
+The component most prominently modulated by attention is called the P3 (positive
+deflection after $\pm300$ ms).
+Some \ac{bci}\footnote{In \ac{erp} analysis, components are sometimes referred to with
+the timing around which they occur (e.g., N170, negative after 170ms), or by their rank
+in order of occurrence (N1, the first negative component).
+The timing nomenclature is based on average latencies for neurotypical individuals, but
+is seldom correct for specific \ac{bci} users due to the large variability across
+\ac{erp} stimulation paradigms and subjects, and can also depend on the user's pathology.
+Furthermore, as we will see later, this work focuses specifically on the intra-session
+variability in \ac{erp} latency, explicitly assuming the latencies are not fixed.
 Therefore, we adhere to the ranking nomenclature.
-Fortunately, for the visual oddball paradigm, the rankings do roughly
-correspond to their expected timings, i.e.\ P1=P100, N1=N100, etc.}.
+Fortunately, for the visual oddball paradigm, the rankings do roughly correspond to
+their expected timings, i.e., P1=P100, N1=N100, etc.}.
 Therefore, oddball paradigms are frequently referred to as P3(00) paradigms.
-However, due to the equally important contributions of other \ac{erp}
-components in decoding, as we will show later, this could be considered a
-misnomer.
+However, due to the equally important contributions of other \ac{erp} components in
+decoding, as we will show later, this could be considered a misnomer.
 Thus, we will adhere to the oddball paradigm naming.
@@
 The visual \ac{erp} components primarily include P1, N1, P2, N2, and P3.
-Each component is characterized by specific latency, amplitude, and neural
-origins.
+Each component is characterized by specific latency, amplitude, and neural origins.
 These factors influence the perception of visual stimuli and attentional
@@
 
-\emph{The P1 component} occurs approximately 100 ms after stimulus presentation.
-This component reflects initial visual processing and is primarily linked to
-activity in the primary visual cortex.
-P1 is particularly sensitive to gaze fixation, where the gaze is directed
-toward a stimulus.
-When a stimulus is fixated on, the amplitude of P1 increases, reflecting the
-enhanced processing of that visual input.
-In contrast, P1 shows less modulation when stimuli are attended to via
-attention without direct gaze.
+\paragraph{The P1 component} occurs approximately 100 ms after stimulus presentation.
+This component reflects initial visual processing and is primarily linked to activity in
+the primary visual cortex.
+P1 is particularly sensitive to gaze fixation, where the gaze is directed toward a
+stimulus.
+When a stimulus is fixated on, the amplitude of P1 increases, reflecting the enhanced
+processing of that visual input.
+In contrast, P1 shows less modulation when stimuli are attended to via attention without
+direct gaze.
 
 \paragraph{The N1 component} peaks around 150 to 200 ms post-stimulus.
-It represents an extension of sensory processing related to both gaze fixation
-and attention.
-For attended stimuli, the amplitude is greater, indicating prioritization for
-further cognitive processing.
-This highlights a more selective form of attention allocation, demonstrating
-the impact of both types of attention on visual processing.
+It represents an extension of sensory processing related to both gaze fixation and
+attention.
+For attended stimuli, the amplitude is greater, indicating prioritization for further
+cognitive processing.
+This highlights a more selective form of attention allocation, demonstrating the impact
+of both types of attention on visual processing.
 
 \paragraph{The P2 component} occurs between 200 to 300 ms.
-It reflects higher-order processing, particularly in contexts demanding
-stimulus evaluation and feature detection.
+It reflects higher-order processing, particularly in contexts demanding stimulus
+evaluation and feature detection.
 P2 is sensitive to attention.
-When participants actively engage with specific features or categories of
-stimuli, P2 shows increased amplitude.
-Conversely, P2 may be less responsive when gaze is not directed toward the
-stimulus but attention is still maintained.
+When participants actively engage with specific features or categories of stimuli, P2
+shows increased amplitude.
+Conversely, P2 may be less responsive when gaze is not directed toward the stimulus but
+attention is still maintained.
 
@@
 control processes such as conflict monitoring and inhibition.
-It is particularly pronounced in tasks requiring differentiation between
-competing stimuli.
-N2 reflects the allocation of attentional resources required to resolve
-conflicts, irrespective of whether the gaze is directly on the competing
-stimuli or not.
-Higher amplitudes in N2 are seen when cognitive demands increase, indicating
-the influence of attention strategies.
+It is particularly pronounced in tasks requiring differentiation between competing
+stimuli.
+N2 reflects the allocation of attentional resources required to resolve conflicts,
+irrespective of whether the gaze is directly on the competing stimuli or not.
+Higher amplitudes in N2 are seen when cognitive demands increase, indicating the
+influence of attention strategies.
 
@@
 post-stimulus.
-It reflects attentional engagement and the processing of rare or unexpected
-events.
+It reflects attentional engagement and the processing of rare or unexpected events.
 The P3 is typically subdivided into two subcomponents: P3a and P3b.
-The P3a is associated with the allocation of attention to novel or unexpected
-stimuli, indicating the initial detection of a change in the environment.
-It is particularly responsive to stimuli that capture attention, whether
-through gaze fixation or attention.
+The P3a is associated with the allocation of attention to novel or unexpected stimuli,
+indicating the initial detection of a change in the environment.
+It is particularly responsive to stimuli that capture attention, whether through gaze
+fixation or attention.
 P3a typically exhibits a frontal distribution and peaks earlier than P3b.
@@
 This subcomponent is typically observed over parietal regions.
-P3b is elicited when participants must process the stimulus meaningfully,
-often requiring a decision or response.
-Both P3a and P3b amplitudes are sensitive to the probability of occurrence
-and the participant's attentional focus, showing how attentional strategies
-affect cognitive processing.
+P3b is elicited when participants must process the stimulus meaningfully, often
+requiring a decision or response.
+Both P3a and P3b amplitudes are sensitive to the probability of occurrence and the
+participant's attentional focus, showing how attentional strategies affect cognitive
+processing.
 
 \paragraph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
-Sustained attention and emotional processing of stimuli are reflected in the
-LPP.
-Typically enhanced for emotionally salient or motivationally relevant
-information, the LPP serves as an indicator of cognitive appraisal beyond mere
-recognition tasks.
-This component can be influenced by attention when the participant evaluates
-the emotional significance of stimuli, even if gaze is not directed at them.
+Sustained attention and emotional processing of stimuli are reflected in the LPP.
+Typically enhanced for emotionally salient or motivationally relevant information, the
+LPP serves as an indicator of cognitive appraisal beyond mere recognition tasks.
+This component can be influenced by attention when the participant evaluates the
+emotional significance of stimuli, even if gaze is not directed at them.
 
-In summary, ERP components such as P1, N1, P2, N2, P3, and LPP collectively
-elucidate the neural underpinnings of visual processing and attentional
-mechanisms.
-These components not only inform the design of an effective oddball BCI system
-but also enhance our understanding of the cognitive processes involved in visual
-attention.
+In summary, \ac{erp} components such as P1, N1, P2, N2, P3, and LPP collectively
+elucidate the neural underpinnings of visual processing and attentional mechanisms.
+These components not only inform the design of an effective oddball \ac{bci} system but
+also enhance our understanding of the cognitive processes involved in visual attention.
 This includes both gaze fixation and attention.
-Understanding these components is essential for optimizing BCI protocols and
-improving user interaction, particularly in contexts involving visual oddball
-tasks.
+Understanding these components is essential for optimizing \ac{bci} protocols and
+improving user interaction, particularly in contexts involving visual oddball tasks.
 
 \subsection{Preprocessing \& feature extraction}
-For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz
-and 16Hz.
-Re-referencing can be done to average mastoids to highlight the contribution of
-the P3 component.
-In addition to the steps outlined in section~\ref{sec:bci/preprocessing},
-the signal is usually cut into \emph{epochs} for \ac{erp} analysis.
+For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz and
+16Hz.
+Re-referencing can be done to average mastoids to highlight the contribution of the P3
+component.
+In addition to the steps outlined in \autoref{sec:bci/preprocessing}, the signal is
+usually cut into \emph{epochs} for \ac{erp} analysis.
 These are time windows surrounding the stimulus event.
-They usually include some baseline interval before the stimulus, which can be
-used to normalize the response.
+They usually include some baseline interval before the stimulus, which can be used to
+normalize the response.
 
@@
 reference at a specific time relative to the stimulus.
-In \ac{erp} classification, they are usually directly used as features, by
-concatenating all channels in the matrix format epoch $\mat{X}$ into one feature
-vector s.t.\ $\mat{x} = \mat\vec(\mat{X})$.
+In \ac{erp} classification, they are usually directly used as features, by concatenating
+all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
+$\mat{x} = \vec(\mat{X})$.
 
 \subsection{Decoders}
-Due to the unfavorable \ac{snr} of \acp{erp}, it is often not feasible to
-accurately decode the encoded information based on just one example.
-Therefore, classification algorithm are trained on a set of epochs, but are
-often evaluated on an average of multiple epochs.
-These averages are constructed over multiple stimulations of the stimulus
-representing a given selection, enhancing the \ac{snr}.
-The drawback, however, is that performing these multiple stimulations takes
-more time.
-To optimize the \ac{itr}, the trade-off between number of repetitions,
-stimulation speed and decoding performance must be carefully balanced.
+
+Due to the unfavorable \ac{snr} of \acp{erp}, it is often not feasible to accurately
+decode the encoded information based on just one example.
+Therefore, classification algorithms are trained on a set of epochs but are often
+evaluated on an average of multiple epochs.
+These averages are constructed over multiple stimulations of the stimulus representing
+a given selection, enhancing the \ac{snr}.
+The drawback, however, is that performing these multiple stimulations takes more time.
+To optimize the \ac{itr}, the trade-off between the number of repetitions, stimulation
+speed, and decoding performance must be carefully balanced.
 Performant classifiers can shift this balance to the benefit of \ac{itr}.
-Several techniques are suited to decode~\acp{erp}.
-Below are some methods that can achieve state-of-the art performance and are
-relevant to this work.
+Several techniques are suited to decode \acp{erp}.
+Below are some methods that can achieve state-of-the-art performance and are relevant to
+this work.
 
 \subsubsection{\Acf{lcmv} beamforming}
-Linear decoding techniques such as the \ac{lcmv}
-beamformer~\cite{Wittevrongel2016} generally work well for
-classifying \ac{erp} epochs.
-\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can
-be multiplied with the feature vector of a new epoch to obtain a prediction
-score.
+
+Linear decoding techniques such as the \ac{lcmv} beamformer~\cite{Wittevrongel2016}
+generally work well for classifying \ac{erp} epochs.
+\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can be
+multiplied with the feature vector of a new epoch to obtain a prediction score.
 The weights $\mat{w}$ are obtained by minimizing the variance of the output
@@
 \end{equation}
-with activation pattern $a$ the difference between the average of the attended epochs and the
-average of the non-attended epochs in the training data.
-This is then solved by  method of Lagrange multipliers as
+with activation pattern $\mat{a}$ the difference between the average of the attended
+epochs and the average of the non-attended epochs in the training data.
+This is then solved by the method of Lagrange multipliers as
 \begin{equation}
@@
 \subsubsection{\Acf{lda}}
+
 \Ac{lda} is a more well-known method that is, in fact, nearly equivalent to
 \ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
-It assumes normally distributed data with a different mean per class but a
-common covariance.
-Similar to \ac{lcmv}-beamforming, \ac{lda} learns a weight vector $\mat{w}$
-that satisfies
+It assumes normally distributed data with a different mean per class but a common
+covariance.
+Similar to \ac{lcmv} beamforming, \ac{lda} learns a weight vector $\mat{w}$ that
+satisfies
 \begin{equation}
@@
 \end{equation}
-The between-class scatter matrix $\mat{S}_\text{b}$ models the variability
-between the expected responses, the within-class scatter matrix
-$\mat{S}_\text{w}$ the expected noise.
-The optimization criterion corresponds to finding a projection that maximizes
-the spread between classes (signal) while minimizing the spread within classes
-(noise).
-$\mat{w}$ can be obtained as eigenvector of
+The between-class scatter matrix $\mat{S}_\text{b}$ models the variability between the
+expected responses, while the within-class scatter matrix $\mat{S}_\text{w}$ models the
+expected noise.
+The optimization criterion corresponds to finding a projection that maximizes the spread
+between classes (signal) while minimizing the spread within classes (noise).
+$\mat{w}$ can be obtained as the eigenvector of
 $\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.
 
-There is one one caveat with these linear methods, however:
-the amount of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
-is usually relatively high compared to the training dataset, since calibration
-time should be kept minimal.
+There is one caveat with these linear methods, however:
+the number of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
+is usually relatively high compared to the training dataset, since calibration time
+should be kept minimal.
 Therefore, strong regularization should be applied, such as covariance/scatter
 shrinkage.
-Regularization can also be performed by retaining structure in the data, such
-as is the case for the multilinear methods presented by~\textcite{Lotte2018}.
+Regularization can also be performed by retaining structure in the data, such as in the
+case for the multilinear methods presented by~\textcite{Lotte2018}.
 
@@
 Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
-stationarity on the background noise to regularize the problem.
+stationarity in the background noise to regularize the problem.
 It does this by imposing a block-Toeplitz structure to the between-class covariance
-matrix of \ac{lda}, such that every temporal covariance block corresponding to
-a channel has the structure
+matrix of \ac{lda}, such that every temporal covariance block corresponding to a channel
+has the structure
 \begin{equation}
@@
 \end{equation}
-This method is simple yet effective reaching state-of-the-art decoding
-performance.
+This method is simple yet effective, reaching state-of-the-art decoding performance.
 
@@
 
-Alternative methods that currently work well include
-non-linear methods include Riemannian geometry~\cite{Barachant2014}.
-Methods leveraging Riemannian geometry use a different feature extraction
-method.
-They do not use the raw epoch values, but rather operate on the spatial
-covariance of an epoch.
-These covariance matrices $C_i$ are symmetric and should have only positive
-eigenvalues, hence they lie on the Riemannian manifold of symmetric positive
-definite matrices, which defined distance metric
+Alternative methods that currently work well include non-linear methods using Riemannian
+geometry~\cite{Barachant2014}.
+Methods leveraging Riemannian geometry use a different feature extraction method.
+They do not use the raw epoch values but rather operate on the spatial covariance of an
+epoch.
+These covariance matrices $C_i$ are symmetric and should have only positive eigenvalues;
+hence they lie on the Riemannian manifold of symmetric positive definite matrices, which
+defines the distance metric
 \begin{equation}
@@
 \end{equation}
-This geodesic distance metric is naturally suited for this type of data and has
-some desirable properties, like invariance to linear transformations, which
-result in a robustness particularly useful for \ac{bci}
-applications~\cite{Barachant2011}.
+This geodesic distance metric is naturally suited for this type of data and has some
+desirable properties, like invariance to linear transformations, which result in a
+robustness particularly useful for \ac{bci} applications~\cite{Barachant2011}.
 
-With this distance defined, the epochs can than be either directly be
-classified on the Riemannian manifold using a minimum distance to mean
-classifier, or they can be projected to the tangent space.
-This is a euclidean space that locally approximates the Riemannian distance,
-which allows us to apply regular machine learning methods like an SVM or
-\ac{lda}.
+With this distance defined, the epochs can either be directly classified on the
+Riemannian manifold using a minimum distance to the mean classifier, or they can be
+projected to the tangent space.
+This is a Euclidean space that locally approximates the Riemannian distance, which
+allows us to apply regular machine learning methods like an SVM or \ac{lda}.
 The tangent space is obtained\todo{tangent space}.
-To enhance classifications of \acp{erp} which rely on time domain information
-which would otherwise be lost in the spatial representation of the extracted
-feature, the class averages can be concatenated as additional channels to the
-epochs, embedding the temporal structure in the feature~\cite{Barachant2014}.
-To improve performance even further, the \acp{erp} can be decomposed by a method
-like XDAWN before calculating the extended covariances~\cite{Li2020}.
+To enhance classifications of \acp{erp} which rely on time-domain information, which
+would otherwise be lost in the spatial representation of the extracted feature, the
+class averages can be concatenated as additional channels to the epochs, embedding the
+temporal structure in the feature~\cite{Barachant2014}.
+To improve performance even further, the \acp{erp} can be decomposed by a method like
+XDAWN before calculating the extended covariances~\cite{Li2020}.
 
@@
 Yet, the overview until now has been rather theoretical.
-Interesting and complex challenges arise when applying these concepts outside
-of the lab in augmented and assistive communication technologies for patients
-with \acp{sspi}.
+Interesting and complex challenges arise when applying these concepts outside of the lab
+in augmented and assistive communication technologies for patients with \acp{sspi}.
 
@@
 In a nutshell, we can state that \acp{bci} decode brain activity with the aim to
-establish a direct communication pathway bypassing speech and other forms of
-muscular activity~\cite{Naci2012,Chaudhary2016}.
+establish a direct communication pathway bypassing speech and other forms of muscular
+activity~\cite{Naci2012,Chaudhary2016}.
 They have raised great hopes for individuals devoid of these abilities, for whom
-\acp{bci}s can provide a means to communicate or to
-control devices.
+\acp{bci} can provide a means to communicate or to control devices.
 
-Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based
-oddball interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
-Targets are shown in short flashes on a computer screen, that evoke
-and \ac{erp} when observed, which can be detected in the \ac{eeg}-signal
-detected in the EEG-signal.
-The \ac{erp} consists of multiple components, which some are modulated by
-perception and the attention of the user.
-Decoding this modulation allows for information transfer controlled by the
-user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
+Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based oddball
+interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
+Targets are shown in short flashes on a computer screen that evoke an \ac{erp} when
+observed, which can be detected in the \ac{eeg} signal.
+The \ac{erp} consists of multiple components, some of which are modulated by perception
+and the attention of the user.
+Decoding this modulation allows for information transfer controlled by the user's brain
+activity, as sketched in \autoref{fig:bci/loop}.
 
-However, when carefully examining the properties of the \ac{erp} components,
-we notice that while attention plays a role in the modulation of specific components,
-some part is also related to visual perception.
-How a stimulus is perceived, and, as a consequence, which \ac{erp} components
-are modulated, is directly related to whether the user gazes directly at it
-(i.e.\ \emph{fixates} their gaze on the target) or not.
-Indeed, research has shown that visual oddball \ac{bci}s cannot operate
-efficiently when the user does not direct
-their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
+However, when carefully examining the properties of the \ac{erp} components, we notice
+that while attention plays a role in the modulation of specific components, some part is
+also related to visual perception.
+How a stimulus is perceived, and, as a consequence, which \ac{erp} components are
+modulated, is directly related to whether the user gazes directly at it (i.e.,
+\emph{fixates} their gaze on the target) or not.
+Indeed, research has shown that visual oddball \ac{bci}s cannot operate efficiently when
+the user does not direct their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
 
 This leads to a problem that is not often addressed:
-a visual \acp{bci} is dependent on the gaze of the user, and hence on
-muscle control.
-This is exactly what \acp{bci} for augmented and assistive technology try to
-avoid.
-The \ac{bci} target population consists of individuals who suffer from various
-degrees of paralysis, or even from \ac{lis}, the complete loss of muscle
-control with preserved consciousness.
-While \ac{bci}s are most attractive as a solution for individuals with
-\ac{sspgi}, studies
-paradoxically often report bad performance in patient populations precisely due
+a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
+This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
+The \ac{bci} target population consists of individuals who suffer from various degrees
+of paralysis, or even from \ac{lis}, the complete loss of muscle control with preserved
+consciousness.
+While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
+studies paradoxically often report poor performance in patient populations precisely due
 to the lack of adequate eye motor control.
@@
 
-Different degrees of eye motor impairment can render visual \ac{bci}s
-uncomfortable or outright impossible to use.
-When operating a visual \ac{bci}, the usual rapid series of forced saccades
-followed by fixation is tiring over time, even for healthy control subjects.
-A suitable alternative would allow the user to keep their eyes in an at-rest
-position of their choice while operating the \ac{bci}.
-This points to the need for gaze-independent solutions, which we will explore
-further in Chapter~\ref{sec:gaze-independence} and the rest of this thesis.
+Different degrees of eye motor impairment can render visual \ac{bci}s uncomfortable or
+outright impossible to use.
+When operating a visual \ac{bci}, the usual rapid series of forced saccades followed by
+fixation is tiring over time, even for healthy control subjects.
+A suitable alternative would allow the user to keep their eyes in an at-rest position of
+their choice while operating the \ac{bci}.
+This points to the need for gaze-independent solutions, which we will explore further in
+\autoref{sec:gaze-independence} and the rest of this thesis.

### 2024-10-07 chapters/bttda.tex
checking bttda

@@
 Tensors provide a structured data representation for this highly dimensional
-multiway data
+multiway data.
 This in turn paves the way to the development of tensor methods which can
@@
 core tensor and factor tensors.
-The most common aproaches adhere to either the Tucker structure or the PARAFAC
+The most common approaches adhere to either the Tucker structure or the PARAFAC
 structure.
 A Tucker decomposition reduces an input tensor of size $(D_1,D_2,\ldots,D_K)$ to
-a dense  tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k<=D_k$ using a
-set of per mode factor matrices.
+a dense tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k \leq D_k$ using a
+set of per-mode factor matrices.
 Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
@@
 
-While commonly used, these Tucker or PARAFAC structure might still not be able to
+While commonly used, these Tucker or PARAFAC structures might still not be able to
 efficiently represent relevant neural information in a compressed format.
@@
 It represents the tensor as a sum of Tucker structured terms.
-If the number of terms is equal to 1, it is equivalent to the Tucker structure;
-if the rank of each term is equal to 1, it is equivalent to the PARAFAC
+If the number of terms is equal to 1, it is equivalent to the Tucker structure; if the rank of each term is equal to 1, it is equivalent to the PARAFAC
 structure.
@@
 the Block-Term Decomposition~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
-Performance of methods leveraging either the Tucker and PARAFAC structure are
-heavily dependent on the a priori choice of hyperparameters describing the
-respectively the multilinear rank or the number of rank-1 terms.
+Performance of methods leveraging either the Tucker and PARAFAC structures are
+heavily dependent on the a priori choice of hyperparameters describing
+the multilinear rank or the number of rank-1 terms.
 The block-term structure is more flexible than either the Tucker or the PARAFAC
@@
 one of these structures and the chosen hyperparameters.
-However, this increased flexibility also increased the number of
+However, this increased flexibility also increases the number of
 hyperparameters to both the number of terms and the multilinear rank of each term.
@@
 functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
-Discriminant tensor features have also be extracted
+Discriminant tensor features have also been extracted
 in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
 However, it is not immediately obvious if either the Tucker or PARAFAC
-structure are most suited to represent the neural data of interest for the \ac{bci}
+structure are most suited to represent the neural data of interest for the BCI
 paradigm and for decoding.
@@
 Promising results have been achieved for regression tasks using
-Higher Order Parial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
+Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
 Regression (BTTR)~\cite{Faes2022,Faes2022b}.
 BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
-but this leaving room for improvement.
-Instead of optimizing features directly for class separability, rather a dummy
+but this leaves room for improvement.
+Instead of optimizing features directly for class separability, a dummy
 independent variable was regressed towards, and the method
 cannot be extended to a multi-class setting.
-Furthermore, structures employed in these regression are still more constrained
+Furthermore, structures employed in these regression approaches are still more constrained
 than what could be achieved with a full block-term tensor structured decomposition
-optimized for discriminability, since they rely on a low rank common subspace
+optimized for discriminability, since they rely on a low-rank common subspace
 between the input and classification labels.
-\textcite{Huang2020} propose a supervized approach for finding multiple discriminant
-multilinear spectral filter terms and apply it to motor imagery \ac{bci}, but their
+\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
+multilinear spectral filter terms and apply it to motor imagery BCI, but their
 decomposition is also restricted in flexibility.
 
-\todo{It's always trichiin in the related work section of a ML/Mathematical
-paper, but unless you r reviewers are absolute experts of this specific field,
+\todo{It's always tricky in the related work section of an ML/Mathematical
+paper, but unless your reviewers are absolute experts of this specific field,
 you should at least try to better describe the previous methods.}
@@
 A block-term decomposition that is directly optimized for discriminability and with a
-good choice of ranks and number of terms might better represent the behaviour
+good choice of ranks and number of terms might better represent the behavior
 of generators of neural activity through increased flexibility, or might
@@
 An alternative view on the same approach goes as follows:
-If HODA with a well chosen multilinear rank extracts some discriminant features
+If HODA with a well-chosen multilinear rank extracts some discriminant features
 from the input tensor, it is likely that it does not yet retrieve all useful
@@
 Could we therefore not keep using HODA to extract discriminant Tucker structured
-as long as decoding performance increases?
+terms as long as decoding performance increases?
 
@@
 \end{enumerate*}
+
 \section{Methods}
@@
 Tensors are indicated by bold underlined letters $\ten{X}$, matrices by bold
-letters $\mat{U}$, fixed scalars by uppercase letters $K$ and variable
-scalars as lower case letters $k$.
+letters $\mat{U}$, fixed scalars by uppercase letters $K$, and variable
+scalars as lowercase letters $k$.
 The $n^\text{th}$ sample of a tensor dataset with $N$ samples is written as
@@
 $\mat{X}_k\in\mathbb{R}^{(D_k\times\prod_{j\neq k}^K D_j)}$, by concatenating
-all mode $j\neq$ fibers.
+all mode $j\neq k$ fibers.
 The tensor-matrix product of tensor $\ten{X}$ with matrix $\mat{U}$ along a
@@
 	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{k-1}\mpr{\mat{U}}{k+1}\ldots\mpr{\mat{U}}{K}$.
-%The Kronecker product is noted as $\otimes$, covariance matrices are indicated  with $\mat{\Sigma}$.
 
@@
 dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
-tensor-matrix mode products
+tensor-matrix mode products:
 \begin{equation}
@@
 \end{equation}
-visualized in Figure~\ref{fig:HODA-backward}.
+as visualized in \autoref{fig:HODA-backward}.
 \begin{figure}[t]
@@
 	\input{figures/bttda/HODA_backward.tikz.tex}
-  \caption[The \acs{hoda} backward projection.]{%
+  \caption[A \acs{hoda} backward projection.]{%
     A visualization of the multilinear projection obtained by Higher Order
@@
 \end{figure}
+
 Analogous to the HOSVD, HODA is a dimensionality
@@
 HODA optimizes discriminability in the Fisher sense, by optimizing the
-Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
+Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
 \begin{equation}
@@
 for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
-latent
-tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
+latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
 these class mean latent tensors.
-If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices
+If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
 \begin{equation}
@@
 To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
-orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
+orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
 or as the partial HOSVD of all stacked tensors in the dataset.
@@
 projections but $\mat{U}_k$ corresponding to mode $k$.
-It then finds a partial latent tensor
+It then finds a partial latent tensor:
 \begin{equation}
@@
 Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
-Discriminant Analysis by constructing the within-class scatter matrix
+Discriminant Analysis by constructing the within-class scatter matrix:
 \begin{equation}
@@
 with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
-and the between-class scatter matrix
+and the between-class scatter matrix:
 \begin{equation}
@@
 with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
-and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
+and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
 \begin{equation}
@@
 are obtained by calculating the
-per-mode total scatter matrices
+per-mode total scatter matrices:
 \begin{equation}
@@
 \end{equation}
-and finding the $r_k$ leading eigenvectors of
+and finding the $r_k$ leading eigenvectors of:
 \begin{equation}
@@
 at each iteration~\cite{Wang2007}.
-%and solving the generalized eigenvalue problem
-%\begin{equation}
-%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
-%\end{equation}
-The iterative process halts after a fixed amount or iterations, or when the
+The iterative process halts after a fixed number of iterations, or when the
 update of each $\mat{U}_k$ is lower than a predetermined threshold.
-The full HODA procedure is summarized in Algorithm~\ref{alg:HODA}.
+The full HODA procedure is summarized in \autoref{alg:HODA}.
 \begin{algorithm}
-  \caption[The \acs{hoda} backward solution.]{The \ac{hoda} backward solution.}
+  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
 	\label{alg:HODA}
@@
 within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
-$\alpha_k$ at each step such that the eigenvalue problem becomes
+$\alpha_k$ at each step such that the eigenvalue problem becomes:
 \begin{equation}
@@
 As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
-also be estimated in a data driven way in HODA~\cite{Jorajuria2022},
+also be estimated in a data-driven way in HODA~\cite{Jorajuria2022},
 e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.
@@
 While a \emph{backward model} extracts latent sources or properties from the observed
-data based on some task related criterion or on prior domain knowledge,
+data based on some task-related criterion or on prior domain knowledge,
 a forward model is a generative model that expresses the observed data in
-function of these latent properties or sources that are given.
-Forward models are useful for a.o.\ interpretability and data compression.
+terms of these latent properties or sources that are given.
+Forward models are useful for, among other things, interpretability and data compression.
 
-The HODA projection in Equation~\ref{eq:HODA-backward} is an example
+The HODA projection in \autoref{eq:HODA-backward} is an example
 of a backward model.
 A straightforward and computationally efficient candidate for a corresponding
-forward model, visualized in Figure~\ref{fig:HODA-forward} is given by
+forward model, visualized in \autoref{fig:HODA-forward}, is given by:
 \begin{equation}
@@
 with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
-reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
+reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
 \begin{figure}[t]
@@
 	\input{figures/bttda/HODA_forward.tikz.tex}
-  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
+  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
     By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
@@
 
-A forward model should make sure that reconstruction error is minimized.
-In other words, variation captured in the latent tensor should maximally captured by the
+A forward model should ensure that reconstruction error is minimized.
+In other words, variation captured in the latent tensor should be maximally captured by the
 reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
@@
 Hence, we aim to minimize the expected value of the cross-covariance between
-the noise term and the extracted latent tensors
+the noise term and the extracted latent tensors:
 \begin{equation}
@@
   -\hat{\ten{X}}(n)\right)^2
-  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
+  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmprs{\mat{A}}\right)^2
 \end{equation}
 This least squares tensor approximation problem can be solved efficiently using the
-alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
-iteratively fixing all but one of the activation patterns such that
+alternating least squares (ALS) algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
 \begin{equation}
@@
 Similar to fitting the backward model, the iterative process for the forward
-model halts after a fixed amount of iterations or when the update of each
+model halts after a fixed number of iterations or when the update of each
 $\mat{A}_k$ is lower than a predetermined threshold.
 The full algorithm to determine the HODA forward projection is listed
-in Algorithm~\ref{alg:HODA-fw}.
+in \autoref{alg:HODA-fw}.
 \begin{algorithm}
-  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
+  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
 	\label{alg:HODA-fw}
@@
 vectorizing the tensor representation.
-While the reconstruction error might be brought down, we lose the computational
+While the reconstruction error might be reduced, we would lose the computational
 efficiency and regularizing constraints of the tensor form.
@@
 class separation, the error term $\ten{\mathbfcal{E}}$ in
-Equation~\ref{eq:HODA-forward} should still contain some discriminative
+\autoref{eq:HODA-forward} should still contain some discriminative
 information, which can be exploited to improve classifier
@@
 $\ten{\mathbfcal{E}}$.
-Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
+\autoref{fig:BTTDA} further illustrates the BTTDA model.
 \begin{figure*}[t]
@@
 	\input{figures/bttda/BTTDA.tikz.tex}
-  \caption[A forward model for \ac{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
+  \caption[A forward model for \acs{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
 		(BTTDA). BTTDA can extract more features
@@
 \end{figure*}
-The block-term structure of this model implies that is a generalization of both
+The block-term structure of this model implies that it is a generalization of both
 the Tucker-structured HODA and PARAFAC-structured discriminant feature
 extraction.
-If $B$ in Equation~\ref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
+If $B$ in \autoref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
 HODA; if at each term $b$ the rank of the core tensor
@@
 Since BTTDA is specified above as a forward model, a backward procedure
-procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
+is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
 BTTDA to be useful as a feature extraction method.
 The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
-computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
+computed through a deflation scheme summarized in \autoref{alg:BTTDA}.
 \begin{algorithm}
-  \caption{\Ac{bttda}.}
+  \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
 	\label{alg:BTTDA}
@@
 For each block $b$, the latent tensor is extracted using the HODA backward
-projection in Equition~\ref{eq:HODA-backward} from the residual error term of the previous
-block $\ten{\mathbfcal{E}}^{(b-1)}$
+projection in \autoref{eq:HODA-backward} from the residual error term of the previous
+block $\ten{\mathbfcal{E}}^{(b-1)}$:
 \begin{equation}
@@
 previous error and its reconstruction after backward and forward HODA
-projection
+projection:
 \begin{equation}
@@
 The resulting latent tensors can be vectorized and concatenated into
-one single feature vector per input tensor
+one single feature vector per input tensor:
 \begin{equation}
@@
 	r_K^{(b)})$ and on the number of blocks $B$.
-If these are not known a priori, i.e. if they cannot be set based on insights in the
+If these are not known a priori, i.e., if they cannot be set based on insights into the
 data generation process, a model selection step is necessary in order to
-determine what the optimal values for $r_k^{(b)}$ and $B$ are.
+determine the optimal values for $r_k^{(b)}$ and $B$.
 
-Furthermore, HODA, and by extension BTTDA can extract a substantial amount
+Furthermore, HODA, and by extension BTTDA, can extract a substantial amount
 of redundant features, which can be dropped after projection and before proceeding to the classification
 step~\cite{Phan2010}.
-Specifcally in BTTDA redundant features can stack up over the number of
+Specifically, in BTTDA redundant features can accumulate over the number of
 blocks, hampering performance.
@@
 univariate Fisher score for all features.
-The Fisher score $\phi(i)$ for feature $i$ is obtained as
+The Fisher score $\phi(i)$ for feature $i$ is obtained as:
 \begin{equation}
@@
 \end{equation}
-The features can then either be sorted by $\phi(i)$ and a given amount of
+The features can then either be sorted by $\phi(i)$, and a given number of
 features retained for classification, or the decision to retain a feature can
@@
 $\phi(i)$.
-We opted for the latter strategy since it does not require us to redetermine the
+We opted for the latter strategy since it does not require redetermining the
 optimal number of features to retain at each block.
@@
 To reduce the computational cost of model selection,
-Algorithm~\ref{alg:model-selection} proposes a heuristic model selection
+\autoref{alg:model-selection} proposes a heuristic model selection
 algorithm that leverages cross-validation in a greedy way per block, to
@@
 classification of extracted feature vectors after feature selection is used as
-cross-validation score.
+a cross-validation score.
 \begin{algorithm}
-  \caption[The greedy model selection procedure.]{Greedy model selection}
+  \caption[Greedy model selection procedure.]{Greedy model selection.}
 	\label{alg:model-selection}
@@
 Finally, the series of blocks can be truncated to the point with
-highest validation score to determine $B$.
+the highest validation score to determine $B$.
 
@@
 blocks such that the resulting forward model adheres to the PARAFAC structure.
-The optimal amount of rank-one blocks can be found by truncating as above.
+The optimal number of rank-one blocks can be found by truncating as above.
 We refer to this strategy as PARAFACDA.
@@
 \section{Experiments}
-\subsection{Datasets \& decoders}
+\subsection{Datasets and decoders}
 We evaluated our proposed model in two offline EEG-based BCI decoding problems:
-the event-related potential (ERP) and motor imagery (MI paradigms using a
+the event-related potential (ERP) and motor imagery (MI) paradigms using a
 selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
@@
 demand.
-Details about these datasets are found in Table~\ref{tab:moabb}.
+Details about these datasets are found in \autoref{tab:moabb}.
 For the ERP datasets, the task is to distinguish target from non-target ERPs,
@@
 Within-session classification performance was assessed using stratified 5-fold
-cross-validation to calculate the area under the receiver operator
+cross-validation to calculate the area under the receiver operating
 characteristic curve (ROC-AUC).
@@
 		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
-		number of runs (\#Runs).
+		number of runs (\# Runs).
     Adapted from~\cite{Aristimunha2023}
@@
 stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
-greedy model selection algorithm in Algorithm~\ref{alg:model-selection}.
-For HODA+LDA and the HODA blocks in BTTDA+LDA, we choose
+greedy model selection algorithm in \autoref{alg:model-selection}.
+For HODA+LDA and the HODA blocks in BTTDA+LDA, we chose
 $r=r_1=r_2=\ldots=r_K$ with possible values
@@
 filtering paired with a Riemannian Minimum Distance to Mean classifier or a
-projection to tangent space and a support vector machine (XDAWNCov+MDM and
-XDAWNCov+TS+SVM) and LDA applied after XDAWN filtering (XDAWN+LDA).
-For the MI datasets, the comparison methods were picked among Riemannian
+projection to tangent space and a support vector machine
+(XDAWNCov+MDM and XDAWNCov+TS+SVM), and LDA applied after XDAWN filtering
+(XDAWN+LDA).
+For the MI datasets, the comparison methods were selected from Riemannian
 methods.
-These comprise of projection onto the Riemannian tangent space combined with an
+These include projection onto the Riemannian tangent space combined with an
 ElasticNet classifier (TS+EL), a Fisher Geodesic Minimum Riemannian Distance to
@@
 (ACM+TS+SVM).
-Furthermore, we report the performance of the Common Spatial Patterns filter
+Additionally, we report the performance of the Common Spatial Patterns filter
 and a frequency filterbank combined with a support vector machine (CSP+SVM and
-FilterBank+SVM respectively).
+FilterBank+SVM, respectively).
 We refer to~\cite{Chevallier2024} for implementation details of these comparison
@@
 EEG signals for the evaluated datasets were recorded at the sample rate given
-by Table~\ref{tab:moabb} and band-pass filtered between 1Hz
+by \autoref{tab:moabb} and band-pass filtered between 1Hz
 and 24Hz.
 The signals were cut into epochs starting from stimulus onset with a
-dataset specific length given by Table~\ref{tab:moabb}.
-For HODA+LDA, PARAFAC+LDA and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
+dataset-specific length given by \autoref{tab:moabb}.
+For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
 For the BNCI2014-008 and BNCI2015-003 datasets, this resulted in matrices of
-dimensionality $(8,48)$ and $(8,38)$ respectively.
+dimensionality $(8,48)$ and $(8,38)$, respectively.
 
-Table~\ref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
+\autoref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
 decoders.
@@
 	\input{chapters/bttda/score_erp.tex}
-  \caption[Whithin-session classification score for 2 event-related potential
+  \caption[Within-session classification score for 2 event-related potential
   datasets.]{Area under the receiver operating characteristic curve for
-		cross-validated whithin-session evaluation for HODA and our proposed decoders
+		cross-validated within-session evaluation for HODA and our proposed decoders
 		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
@@
 \end{table}
-Highest performance is reached with the proposed BTTDA+LDA.
-One-sided Wilcoxon signed-rank tests with significance level $\alpha=0.05$ reveal that both PARAFAC+LDA and BTTDA+LDA
-significantly outperform HODA+LDA in both the BNCI2014-008 (PARAFAC+LDA:
-$p=0.0039$, BTTDA+LDA: $p=0.0039$) and the BNCI2015-003 (PARAFAC+LDA:
-$p=0.0001$, BTTDA+LDA: $p=0.0049$) datasets.
+The highest performance is achieved with the proposed BTTDA+LDA.
+One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
+reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
+HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
+$p=0.0039$) and the BNCI2015-003 (PARAFACDA+LDA: $p=0.0001$, BTTDA+LDA:
+$p=0.0049$) datasets.
 No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.
@@
 Event-Related Synchronizations/Desynchronizations (ERS/Ds).
-Contrary to the time domain analyses performed on ERPs, ERS/Ds are usually well
+Contrary to the time-domain analyses performed on ERPs, ERS/Ds are usually well
 discerned in the time-frequency domain.
-Hence, for the MI task, we will transform the EEG signal into the
-time-frequency domain, forming $3^\text{d}$ order tensors, with $K=3$ modes
-representing the channels, frequencies and time bins.
+Hence, for the MI task, we transform the EEG signal into the
+time-frequency domain, forming $3^\text{rd}$ order tensors, with $K=3$ modes
+representing the channels, frequencies, and time bins.
 
-The MI datasets listed in Table~\ref{tab:moabb} were first band-pass filtered
+The MI datasets listed in \autoref{tab:moabb} were first band-pass filtered
 between 8 and 30Hz and cut into
-epochs with time windows as specified by Table~\ref{tab:moabb}.
+epochs with time windows as specified by \autoref{tab:moabb}.
 Next, time-frequency transformation was performed using a complex Morlet wavelet
@@
 Features were extracted by taking the log-transformed envelope of the complex
-wavelet transformation an daveraging each epoch along the time axis into time bins of
+wavelet transformation and averaging each epoch along the time axis into time bins of
 length $1/4$s.
-For the BNCI2014-001 and BNCI2014-004 dataset, this resulted in tensors of dimensionality
-$(22, 4, 16)$ and  $(3, 4, 18)$
-respectively.
+For the BNCI2014-001 and BNCI2014-004 datasets, this resulted in tensors of dimensionality
+$(22, 4, 16)$ and $(3, 4, 18)$, respectively.
 
-Table~\ref{tab:erp-score} lists the cross-validated classification score for
+\autoref{tab:mi-score} lists the cross-validated classification scores for
 the evaluated motor imagery datasets.
@@
 	\input{chapters/bttda/score_mi.tex}
-  \caption[%
-    Whithin-session classification score for 2 motor imagery datasets.
-    ]{Classification score for
-		cross-validated whithin-session evaluation for HODA+LDA and our proposed decoders
+  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
+		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
 		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
 		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
-    is listed for BNCI2014-001 and area under the receiver-operator
-    characteriscic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
-    for BNCI2014-001 but does not reach a performance comparable to current
-    state-of the art. For BNCI2014-004, PARAFACDA and BTTDA perform
+    is listed for BNCI2014-001, and area under the receiver-operator
+    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
+    for BNCI2014-001 but does not reach performance comparable to the current
+    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
     approximately on par with HODA and the current state-of-the art.}
@@
 For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
-satisfactory performance compared to the comparison methods, yet here both
+satisfactory performance compared to the comparison methods, yet both
 PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
 difference is significant ($p=0.0269$).
-For the BNCI2014-004 dataset the performance gap with comparison methods is
+For the BNCI2014-004 dataset, the performance gap with comparison methods is
 smaller, but no significant differences were found between HODA+LDA,
-PARAFACDA+LDA and BTTDA+LDA.
+PARAFACDA+LDA, and BTTDA+LDA.
 
 \subsection{Block contribution}
-To analyze the contribution of extra feature blocks extracted by {BTTDA} over
-the first one found by HODA we pick n ERP ($K=2$) dataset
+To analyze the contribution of extra feature blocks extracted by BTTDA over
+the first one found by HODA, we pick an ERP ($K=2$) dataset
 (BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
-We report within-session ROC-AUC scores for training, validation and test data as a function
-of the number of blocks increases, Figure~\ref{fig:blocks}.
+We report within-session ROC-AUC scores for training, validation, and test data as a function
+of the number of blocks increases, shown in \autoref{fig:blocks}.
 Training and validation folds were taken from the model selection procedure.
@@
 $\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
-NMSE is calculated as
+NMSE is calculated as:
 \begin{equation}
@@
 \subsection{Contribution}
-The results in Table~\ref{tab:erp-score} and Table~\ref{tab:mi-score} show that
+The results in \autoref{tab:erp-score} and \autoref{tab:mi-score} show that
 in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
 exceeding all comparison methods for the two evaluated datasets, but
-for MI datasets results fall short of those of comparison methods.
-Performance on the MI classification task could however be greatly influenced
-by the exact tensorization method used, i.e. the time-frequency transform in
+for MI datasets, results fall short of those of comparison methods.
+Performance on the MI classification task could, however, be greatly influenced
+by the exact tensorization method used, i.e., the time-frequency transform in
 this case, which is not the main focus of this work.
 
-Nevertheless, when considering at the relative improvement over the original
-HODA model, our results show BTTDA+LDA performs consistently on par or higher than HODA+LDA.
+Nevertheless, when considering the relative improvement over the original
+HODA model, our results show that BTTDA+LDA performs consistently on par or higher than HODA+LDA.
 The classification score for BTTDA+LDA was also slightly higher than that of
 restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
-results were not significant and further studies with more datasets and
+results were not significant, and further studies with more datasets and
 subjects should show whether this holds.
-Figure~\ref{fig:blocks} shows that there is added value in finding extra blocks
+\autoref{fig:blocks} shows that there is added value in finding extra blocks
 over the first HODA block.
-While no proof is given, we notice NMSE monotonically decreases, indicating that
-Since the training score keeps increasing, this indicates that
+While no proof is given, we notice that NMSE monotonically decreases, indicating that
+since the training score keeps increasing, this suggests that
 eventually all the variation in the signal will be explained by the model
 while still extracting features that are maximally discriminant.
-Eventually, the amount of blocks will reach a point of diminishing validation
+Eventually, the number of blocks will reach a point of diminishing validation
 score returns, when adding extra features to the decision classifier increases
@@
 We assume the main benefit of BTTDA is that it can more easily discover relevant
-features, while
+features while
 being more parsimonious due to its block-term structure compared to HODA's full
-Tucker structure, as illustrated by Figure~\ref{fig:bttda/sparse}.
+Tucker structure, as illustrated by \autoref{fig:bttda/sparse}.
 \begin{figure}
@@
 The same discriminative information captured by a relatively large
-Tucker-structured core tensor, could be expressed more sparsely with a small
+Tucker-structured core tensor could be expressed more sparsely with a small
 number of block-terms, while avoiding redundant features.
 The PARAFAC structure employed in PARAFACDA is even more sparse, which could be
-a benefit or an advantage depending on the amount of regularization required,
+a benefit or a drawback depending on the amount of regularization required,
 or on the true underlying structure of the data.
@@
 covariance.
-Since HODA estimates one whithin-class scatter matrix
+Since HODA estimates one within-class scatter matrix
 $S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
-model of the data scatter is determined by these per mode scatter matrices as a
+model of the data scatter is determined by these per-mode scatter matrices as a
 Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
@@
 
-Additionally, the forward modeling step inherent to the BTTDA results
-in an interpretable model, since we can use the activation patterns or the
+Additionally, the forward modeling step inherent to BTTDA results
+in an interpretable model since we can use the activation patterns or the
 forward projection to inspect the neural patterns corresponding to the
 relevant discriminatory information at each block~\cite{Haufe2014}.
-Figure~\ref{fig:forward} shows the activation patterns
+\autoref{fig:forward} shows the activation patterns
 of two blocks obtained from the BNCI2014-008 dataset as well as the forward
@@
 Given informed or correctly tuned hyperparameters, this method could be used to
-e.g. separate ERP components or neural processes based on the task-related
-information in the class labels.
-
-\subsection{Model selection and dimensionality}
-
-BTTDA trades in the rigid HODA model for increased model complexity with more
-hyperparameters to tune, which could open up a setting where performance can be
-improved.
-This turns tensor discriminant analysis into a model selection procedure
-instead of a projection optimization procedure.
-Despite favorable results in BCI decoding, the applications of the proposed
-BTTDA model are limited mainly by this model selection approach used
-to determine the individual block ranks.
-While our proposed greedy model selection algorithm is a step in the right
-direction, the high computational cost of setting hyperparameters through cross
-validation can still hinder the portability of decoders relying on
-BTTDA.
-
-Due to its heuristic nature, the greedy algorithm does not always result in the
-set of ranks with the highest achievable performance.
-In combination with the fact that the feature selection cutoff parameter was
-fixed somewhat arbitrarily, it is clear that thorough
-hyperparameter optimization could improve performance.
-Furthermore, it is clear that our proposed model selection procedure does not
-necessarily result in an optimal set of blocks that group coherent projections
-within the same block, according to some some desirable metric.
-Examples of this are sparsity, pattern interpretability, minimal or maximal between- or within-block feature
-correlation, decreasing discriminability etc.
-Finally, the presented results are limitation that by our artificial restriction
-that $(r_1=r_2=\ldots=r_k)$ to reduce computational demand of the presented
-experiments.
-In a sense, this goes against the proposition of increased model flexibility.
-It might well be that BTTDA in some cases offers little added value over the
-Tucker-structured HODA, when the both are given free choice of rank.
-Future efforts should focus on automatic parameter setting, e.g. using sparsity
-or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
-statistical measures based on the models application.
-
-Another limitation is that BTTDA might yield a disproportionate
-improvement for datasets with a low number of features relative to sample size,
-while being less effective for datasets with more features
-This is reflected
-in our ERP results (low dimensionality vs. high number of trials) compared to the
-MI results (higher dimensionality due to third order tensorization vs. lower
-number of trials).
-We expect a dimensionality limit beyond which the forward modeling step cannot
-accurately regress from the low dimensional latent tensors to the high
-dimensional original tensors, introducing
-error in the input data for the next block, which can stack up over blocks.
-Since the forward multilinear least squares problem is underdetermined, it is prone to
-numerical instability, which calls for regularization of the forward modeling
-procedure, but this would introduce another hyperparameter.
-It should also thoroughly be investigated what the impact is of going beyond
-second- and third-order case to higher order tensors, since this could have a large impact
-on the model.
-Other tensorization methods of the EEG data, like time-lagged Hankel
-tensors~\cite{Papy2005}, or tensors across subjects or sliding windows etc., could also be
-of interest if they are appropriately chosen based on prior knowledge of the dataset.
-
-\section{Conclusion}
-
-We have introduced Block-Term Tensor Discriminant Anlysis (BTTDA) a novel,
-tensor-based, supervised dimensionality reduction technique optimized for class
-discriminability which adheres to the block-term tensor structure.
-BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
-applied as a special sum-of-rank-one tensors PARAFACDA model.
-The model is obtained by iteratively fitting HODA in a deflation scheme,
-leveraging a novel forward modeling step.
-
-Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
-feature extraction can significantly outperform decoders based on HODA and
-reach state-of-the art decoding performance on event-related potential
-problems (second order tensors) and scores on par with or higher than HODA  in motor
-imagery problems (third order tensors).
-Moving from the rigid Tucker tensor structure of HODA to the more flexible
-block-term structure shifts the problem from finding optimally constrained multilinear
-projections to model and feature selection.
-
-Introducing a flexible block-term tensor model as underlying structure
-reformulates tensor discriminant analysis as a model selection
-problem.
-This allows performance to be traded off for model complexity and number of
-features, to find a setting that is more effective for decoding.
-Because of its general implementation and minimal assumptions on data structure,
-BTTDA can equally be applied to classification for other neuroimaging modalities
-(MEG, ECoG, fNIRS, fMRI, EMG, ...) or to tensor classification problems in other
-fields.
-
-%\section*{Acknowledgements}
-%\todo{move to the end}
-%We thank the Flemish Supercomputer Center (VSC) and the High-Performance
-%Computing (HPC) center of the KU Leuven for allowing us to execute our
-%computational experiments on their systems.
-%We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
-%work.
+e.g., separate ERP components or neural processes based on the task-related
+information in the class labels. \todo{spellcheck from here}

### 2024-10-07 chapters/gaze_independence.tex
checking bttda

@@
 One of the goals of brain-computer interfacing is establishing a communication
-channel that does not rely on speech or muscular activity which in turn can
+channel that does not rely on speech or muscular activity, which in turn can
 provide solutions to individuals with \ac{sspi}.
-In the strictest interpretation, this then means that an interface should not
+In the strictest interpretation, this means that an interface should not
 rely on the control of eye muscles used to redirect the gaze or for blinking.
@@
 \ac{tbi}).
-Many individuals in these groups unfortunately also suffer to
-some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
+Many individuals in these groups unfortunately also suffer
+to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
 Table~\ref{tab:incidence} reports the relatively high frequencies of
-eye motor impairment, which can range from minor (nystagmus
-\footnote{%
-Involuntary, rhythmic, and repetitive eye movements recognizable by their
-consistent directionality (horizontal, vertical, or rotational).
-}, other eye tremors
-\footnote{%
-These can include square-wave jerks, saccadic intrusions, microtremors, or
-microsaccades while resting or fixating the gaze.
-}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
-\footnote{%
-Weakness or limited paralysis of one or more of the muscles that control eye
+eye motor impairment, which can range from minor (nystagmus,
+\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
+consistent directionality (horizontal, vertical, or rotational).},
+other eye tremors
+\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
+microsaccades while resting or fixating the gaze.},
+gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
+\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
 movement, leading to restricted eye motion, but not complete paralysis.
-Often, a specific movement direction (up-down, left-right) is preserved.
-}, involuntary movements, impaired pursuit, \ldots), and even complete
-ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
-This not only has an effect on vision and coordination but also
-on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.
+Often, a specific movement direction (up-down, left-right) is preserved.},
+involuntary movements, impaired pursuit, \ldots), and even complete
+ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
+This not only affects vision and coordination but also
+their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.
 
 \begin{table}
-  %\sffamily
-  %\footnotesize
-	%\centering
   \input{chapters/gaze_independence_incidence.tex}
@@
 from the onset of their condition.
-Individuals recovering from stroke with severe eye motor impairment are
-Furthermore, those with
 Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
@@
 Although eye movement is often cited as one of the longest preserved
-capabilities in \ac{als}, show that minor issues are still
+capabilities in \ac{als}, studies show that minor issues are still
 fairly common, especially for patients with bulbar onset
 \ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
-Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
+Furthermore, \textcite{Hayashi1991} show that as these patients progress past
 the point of independent breathing, symptoms will eventually also involve
@@
 \ac{ms}.
-\ac{ms} is a neurodegenerative disease involving demyelination of nerves.
+\Ac{ms} is a neurodegenerative disease involving demyelination of nerves.
 Eye motor abnormalities are especially well
 studied~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020} and
-often used as diagnostical tool.
+are often used as diagnostic tools.
 These abnormalities can be minor or severe, seldom progressing to complete
@@
 
-\ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
+\Ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
 This almost always affects eye
 movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
-involuntary movements and troubles pursuing or fixating targets.
-They also gradually have more troubles speaking, but often retain some muscular
+involuntary movements and trouble pursuing or fixating on targets.
+They also gradually have more trouble speaking but often retain some muscular
 control.
@@
 
-\newcommand\fnlis{\footnote{
-Multiple definitions of \ac{lis} are encountered in
+\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
 \ac{bci} and neurological literature.
@@
 specific regions in the brain (midbrain, brainstem, or
-cerebellum)~\cite{Smith2005},
-it can also generally refer to the state of full body paralysis
+cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
 or loss of muscle tone incurred in neurodegenerative diseases, combined with
-the inability to speak, such as occurs in late stage ALS.
-}}
+the inability to speak, such as occurs in late-stage ALS.}}
 
 \newcommand\fnwolpawcrit{\footnote{
-\it``The first class consists of people who are truly totally locked-in (e.g.,
+``The first class consists of people who are truly totally locked-in (e.g.,
 due to end-stage ALS or severe cerebral palsy), who have no remaining
@@
 unreliable, or easily fatigued.
-This group is much larger than the first.
-}}
+This group is much larger than the first.''~\cite{Wolpaw2008}}}
 
-We believe it is in general not opportune to delve too deep in the underlying
+We believe it is generally not opportune to delve too deeply into the underlying
 ophthalmological and neurological mechanisms for each of the specific
 conditions, or even the exact symptoms.
-In clinical reality, every patient different symptoms, which lead to their own
+In clinical reality, every patient has different symptoms, which lead to their own
 unique set of preserved capabilities and visual skills.
-From a solution-oriented \ac{bci} engineering point
-of view, the etiology of the symptoms can be abstracted away.
-Instead of categorizing patients we will refer to those that might benefit from
-\ac{bci} assistive communication technology as with \ac{sspi}, in line with
+From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
+Instead of categorizing patients, we will refer to those who might benefit from
+\ac{bci} assistive communication technology as having \ac{sspi}, in line with
 the terminology used by~\textcite{FriedOken2020}.
 If they have eye motor impairment that affects their use of visual \ac{bci} or
-eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).
+eye-tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).
 
@@
     they have \acf{sspgi}.
-    Individuals with \ac{lis} have no muscular control left, or is it is very limited,
-    such as the ability communicate through a binary switching system.
+    Individuals with \ac{lis} have no muscular control left, or it is very limited,
+    such as the ability to communicate through a binary switching system.
     \Acp{bci} are especially relevant for the last two groups.
@@
 Hence, it would be interesting to focus \ac{bci} development efforts on the
-larger population of individuals with \ac{sspgi}  that currently slip through
-the cracks of the assistive technology offer.
+larger population of individuals with \ac{sspgi} who currently slip through
+the cracks of the assistive technology offerings.
 These are individuals whose severe eye motor impairment prevents them from
-using eye tracking based solutions.
+using eye-tracking-based solutions.
 They currently use their remaining motor control to
 communicate by indicating symbols on a letterboard with great effort,
-or signal with upwards eye movements or blinks to confirm prompted letters.
-They require a caregiver or relative to interpret their signals, and crave
-the ability to communicate independently which is crucial to retaining an
+or signal with upward eye movements or blinks to confirm prompted letters.
+They require a caregiver or relative to interpret their signals and crave
+the ability to communicate independently, which is crucial to retaining an
 acceptable quality of life.
@@
 
-In this work, we will use the term \emph{gaze-independent} meaning
-\emph{‘dealing explicitly with the fact that a user cannot control their
+In this work, we will use the term \emph{gaze-independent} to mean
+\emph{`dealing explicitly with the fact that a user cannot control their
 gaze.'}
@@
 There are multiple approaches to implement gaze-independence in a \ac{bci}.
-The next section highlights some noteworhy examples.
+The next section highlights some noteworthy examples.
 
@@
 \subsection{Gaze-independent modalities}
-Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
+Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
 ways.
@@
 rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
-complexity that can be captured in commands, but often only do so when paired
+complexity that can be captured in commands, but they often do so only when paired
 with invasive recording.
-Non-visual reactive paradigms that use e.g. auditory~\cite{Halder2010} and
+Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
 somatosensory~\cite{Brouwer2010}
-stimulation do not rely on gaze redirection exist but also result in lower information transfer
+stimulation do not rely on gaze redirection but also result in lower information transfer
 rates compared to visual paradigms.
-While auditory and somatosensory stimulation on its own might yield poor
-\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
+While auditory and somatosensory stimulation on their own might yield poor
+\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
 settings when coupled with oddball stimulation in a hybrid paradigm.
-Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
-In general, their approach is interesting since they follow a philosophy that
+Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
+In general, their approach is interesting as it follows a philosophy that
 relies on establishing as many communication pathways as possible.
@@
 \label{sec:gaze-independence/sota/visual}
-This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
-the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
-Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
+
+This leads us to the second approach: current visual stimulation paradigms can be optimized so that
+the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
+Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
 the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
 speller~\parencite{Acqualagna2011}.
-The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
+The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
 \ac{bci}.
 \textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
-a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
+a paradigm with subsequently flashing sets of three characters filling the user's field of view.
 We make a distinction here between spatially organized interfaces, where
 multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like \ac{rsvp},
-where targets or small sets of targets are shown consecutively.
-The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
+and temporally organized interfaces, like \ac{rsvp}, where targets or small sets of targets are shown consecutively.
+The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
 variants of the \ac{rsvp} paradigm are temporally organized.
-\textcite{Lin2018} follow the philosophy that spatially organized interface
-have historically shown not to perform well in the presence of gaze impairment.
-However, we argue that, spatially organized interfaces in general have a much higher
+\textcite{Lin2018} follow the philosophy that spatially organized interfaces
+have historically not performed well in the presence of gaze impairment.
+However, we argue that spatially organized interfaces generally have a much higher
 \ac{itr} in healthy controls.
-For instance, a regular matrix-based speller can achive an \ac{itr} to up to
-30bits/min.
-With the right set of adaptations, spatial attention could potentionally also
+For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
+30 bits/min.
+With the right set of adaptations, spatial attention could potentially also
 bring added value to individuals with \ac{sspgi}.
 While spatial interfaces are indeed more gaze-dependent than temporal
-interfaces they also provide an extra channel that can be used to transfer
-information, i.e.\ the location of the stimulus.
+interfaces, they also provide an extra channel that can be used to transfer
+information, i.e., the location of the stimulus.
 
 As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
-attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
+attention to stimulus color, shape, or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
 The \ac{rsvp} speller is already an example of this, relying on the user to
-attend to attend the appearance of the intended character to type.
+attend to the appearance of the intended character to type.
 Visual stimulation paradigms relying on alternative types of attention
 can modulate specific extra \ac{erp} components that either improve performance
-because the embed extra information in the brain signal~\cite{Xu2022}
+because they embed extra information in the brain signal~\cite{Xu2022}
 or because they are more sensitive to stimulation in the visual
 periphery~\cite{Schaeff2012}.
-Nevertheless, solely relying on alternative types of attenion can suffer too
-from reduced information transfer rates~\cite{Chennu2013}.
-Furthermore, the previously mentioned systems typically were only tested in
+Nevertheless, solely relying on alternative types of attention can also suffer from
+reduced information transfer rates~\cite{Chennu2013}.
+Furthermore, the previously mentioned systems were typically tested only in
 settings where the user was required to focus on a central fixation point while
@@
 \label{sec:gaze-independence/sota/decoding}
+
 Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
@@
 \textcite{Aloise2012} aimed to bridge the performance gap between covert and
-overt VSA decoding performance.
+overt \ac{vsa} decoding performance.
 They compared classical linear and non-linear \ac{erp} classifiers on a covert
 attention oddball \ac{erp} paradigm dataset.
-The results revealed no significant performance improvement in covert VSA
+The results revealed no significant performance improvement in covert \ac{vsa}
 decoding for any of the investigated decoders.
@@
 More recent work has made advances in decoding lateralized covert
-VSA harnessing the N2pc \ac{erp}
-component~\cite{Thiery2016,Reichert2020b,Wang2022}
-The amplitude of the N2pc component in the contralateral hemisphere, is directly modulated
+\ac{vsa} by harnessing the N2pc \ac{erp}
+component~\cite{Thiery2016,Reichert2020b,Wang2022}.
+The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
 by the location of covert \ac{vsa}.
@@
 content~\cite{Tonin2013}.
-These methods are promising, but to date often require a slower stimulation pace than or cannot display as
+These methods are promising, but to date often require a slower stimulation pace or cannot display as
 many targets as the classical \ac{erp} \ac{bci} paradigms.
@@
 extra communication channels in a hybrid paradigm.
-\textcite{Xu2016} reached an \ac{itr} of 23.56bits/min by detecting the N2pc
+\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
 component while performing \ac{ssvep} stimulation.
@@
 
-To summarize: for patients with eye motor impairment, gazing directly at a screen target may
+To summarize, for patients with eye motor impairment, gazing directly at a screen target may
 be uncomfortable, impractical, or even impossible.
@@
 them.
-Consequently, while \ac{bci}s hold great promise for these individuals, conventional
+Consequently, while \acp{bci} hold great promise for these individuals, conventional
 gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
@@
 Therefore, the development of decoding strategies that account for covert
-VSA becomes crucial in the pursuit of high-performance gaze-independent
-\ac{bci}s.
+\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
+\acp{bci}.
 
-The general goal of this PhD is to tackle the gaze-dependence in problem in the
+The general goal of this PhD is to tackle the gaze-dependence problem in the
 visual oddball \ac{bci}.
-Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms form
-a good candidate for a performant gaze-independent \ac{bci}.
+Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
+good candidate for a performant gaze-independent \ac{bci}.
 While the central gaze fixation of most visual gaze-independent paradigms still
@@
 In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
-visuospatial attention can be decoded separately from the position of gaze,
-which do not necessarily need to coincide.
+visuospatial attention can be decoded separately from gaze position, which do not necessarily need to coincide.
 
@@
 computer interface by using a suited decoding strategy}.
-%Hence, our working hypothesis is as follows:
-%\todo{rewrite, this is not what we tested in the patient paper after all}
-%\begin{quote}
-%	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
-%		can achieve a similar transfer rate in patient
-%		populations with eye motor deficits than other gaze-independent alternatives
-%    described in literature.}
-%\end{quote}
 
@@
 development.
-We collect data from healthy control subjects, which allow us to test our decoders.
+We collect data from healthy control subjects to test our decoders.
 Finally, the findings of these experiments will be verified in experiments with
-individuals with \ac{sspgi}, including individuals with \ac{lis} evaluating
+individuals with \ac{sspgi}, including individuals with \ac{lis}, evaluating
 both performance and comfort.
@@
 We believe this can be done with a non-invasive, spatially
-organized visual \acp{bci} with high-\ac{itr}.
-Finally, or efforts also aim to improve hand improving \ac{erp} decoding performance
-in general, which will also contribute to their effectiveness in gaze-dependent
-or gaze-independent settings.
+organized visual \acp{bci} with high \ac{itr}.
+Finally, our efforts also aim to improve \ac{erp} decoding performance
+in general, which will also contribute to their effectiveness in both gaze-dependent
+and gaze-independent settings.
 
@@
 
-During this PhD, we explored different lines in decoding strategies,
-trying to tackle several problems, be it general problems in \ac{erp} decoding
+During this PhD, we explored different lines of decoding strategies,
+trying to tackle several challenges, be they general problems in \ac{erp} decoding
 or those that arise specifically from gaze-independence.
-Examples of this are the lack or decrease in amplitude of specific \ac{erp} components, and the increased
+Examples of these challenges include the lack or decrease in the amplitude of specific \ac{erp} components, and the increased
 non-stationarity of the signal.
@@
 The general goal is thus to design a machine learning classifier that represents
-the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
+the \ac{erp} signal in a way that makes it more robust to the problems
 occurring in covert attention conditions.
@@
 \subsubsection{Regularized spatiotemporal beamforming}
+
 Due to the decreased amplitude of the N1 and P3 components in covert attention
@@
 Therefore, a straightforward way to reach satisfactory
-gaze-independent decoding performance, might be by increasing overall \ac{erp}
+gaze-independent decoding performance may be to increase overall \ac{erp}
 decoding performance.
 
-Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
+To address this, we improved upon an in-house developed, state-of-the-art \ac{erp}
 decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
@@
 This approach has been published by~\textcite{VanDenKerchove2022} and will be
-described in Chapter~\ref{sec:stbf-struct}.
+described in \autoref{sec:stbf-struct}.
 
@@
 features from multidimensional neural data is critical.
-Tensor decoding methods offers a powerful approach by preserving the multiway structure
+Tensor decoding methods offer a powerful approach by preserving the multiway structure
 of \ac{eeg} data while optimizing class separability.
@@
 A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
-optimized for class separation, similar to \ac{lda}, but extended to multidimensional data.
+optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
 This method has been applied to various \ac{bci} tasks, including \ac{erp} and
@@
 gaze-independence.
-The proposed \ac{bttda} method will be discussed in~\autoref{sec:bttda}
+The proposed \ac{bttda} method will be discussed in \autoref{sec:bttda}.
 
 \subsubsection{Classifier-based Latency Estimation with Woody iterations}
-The previous two methods did not yet yield results specifically in gaze-independent
+
+The previous two methods did not yet yield specific results in gaze-independent
 decoding.
-The following approach is targeted towards improving decoding performance in
+The following approach is targeted at improving decoding performance in
 gaze-independent settings by accounting for a known property of the covert
@@
 P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
-value is heavily dependent on the subject and the task, and can vary from trial
+value is heavily dependent on the subject and the task and can vary from trial
 to trial~\cite{Ouyang2017}.
@@
 latencies is important in gaze-independent decoding and has been hampering covert
-VSA decoding performance.
-We reprised their hypothesis stating that jitter compensation through latency
+\ac{vsa} decoding performance.
+We reprised their hypothesis, stating that jitter compensation through latency
 estimation and alignment improves
-covert \ac{vsa} performance and extend it by developing a decoder.
+covert \ac{vsa} performance and extended it by developing a decoder.
 
@@
 method named \ac{wcble}.
-This approach has been published by~\cite{VanDenKerchove2024} and is presented
-and evaluated on simulated data in~\autoref{sec:wcble}.
+This approach has been published by~\textcite{VanDenKerchove2024} and is presented
+and evaluated on simulated data in \autoref{sec:wcble}.
 We have also collected datasets to evaluate this approach on real \ac{erp} data
@@
 
-In a first series of experiments, we recorded data with this interface from healthy participants.
-The goal of these experiments is to benchmark
+In the first series of experiments, we recorded data using this interface from healthy participants.
+The goal of these experiments was to benchmark
 gaze-independent \ac{erp} decoding algorithms.
 We denote this dataset as the Covert Visuospatial Attention \ac{erp} dataset
-(CVSA-\ac{erp})
+(CVSA-\ac{erp}).
 This study was approved by the Ethics Commission of University Hospital Leuven
@@
 To simulate the dissociation between the eye gaze and the
-intended target (\ac{vsa}) which could occur in individuals with gaze
-impairment, healthy participants are cued to operate the \ac{bci} in specific
+intended target (\ac{vsa}), which could occur in individuals with gaze
+impairment, healthy participants were cued to operate the \ac{bci} in specific
 \ac{vsa} conditions.
 
-We chose a visual oddball interface that helps us study and adapt
-to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
-decoders and on our own proposed decoders.
-Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
-\textcite{Treder2010}, we present six flashing circular targets
+We chose a visual oddball interface to study and adapt
+to the effects of eye motor impairment on the \ac{erp}, using both existing state-of-the-art
+decoders and our own proposed decoders.
+Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed by
+\textcite{Treder2010}, we presented six flashing circular targets
 (without letters or symbols) to the participant while the EEG and
-electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
+electro-oculogram (EOG) were recorded, along with eye gaze using eye tracking.
 The Hex-o-Spell was chosen since it is optimized for gaze-independent
@@
 
-The interface can be operated in different VSA conditions as illustrated in
-Figure~\ref{fig:gaze/vsa}.
+The interface can be operated in different \ac{vsa} conditions, as illustrated in
+\autoref{fig:gaze/vsa}.
 
@@
     \caption[Overt \ac{vsa}]{%
-      \emph{overt} \ac{vsa}.
+      \emph{Overt} \ac{vsa}.
       Gaze and \ac{vsa} coincide on a target.
@@
       Gaze and \ac{vsa} coincide on a target.
-      The user is free to direct their gaze as they deem most comfortable
+      The user is free to direct their gaze as they deem most comfortable.
       All previous \ac{vsa} conditions are possible.
@@
 
-In the overt case, users gaze at the cued
-target they are also mentally attending; in the covert case, users
-gaze at the center of the screen while mentally attending to the cued target.
-Finally, we introduce a VSA condition that is understudied in the context of
-gaze-independent \ac{bci} development: split VSA.
-In split VSA, the user mentally focuses on one cued target while gazing at
+In the overt case, users gazed at the cued
+target they were also mentally attending; in the covert case, users
+gazed at the center of the screen while mentally attending to the cued target.
+We introduced a \ac{vsa} condition that is understudied in the context of
+gaze-independent \ac{bci} development: split \ac{vsa}.
+In split \ac{vsa}, the user mentally focuses on one cued target while gazing at
 another (the distractor).
-This last option has only scarcely be studied, but completes
-the option s to dissociate gaze and visuospatial attention, which allow us to
+This last option has been scarcely studied but completes
+the options to dissociate gaze and visuospatial attention, allowing us to
 investigate the effect of (the lack of) gaze control on \ac{bci} performance.
-This dataset and performance of our proposed~\ac{wcble} decoder has also been
-published by~\textcite{VanDenKerchove2024} and will be presented in
-Chapter~\ref{sec:covert-align}.
+This dataset and the performance of our proposed \ac{wcble} decoder have also been
+published by \textcite{VanDenKerchove2024} and will be presented in
+\autoref{sec:covert-align}.
 Performance is also evaluated on a publicly available dataset.
@@
 This setting allowed us to investigate how participants with varying eye motor capabilities
-naturally interact with a \ac{bci}, compared to standard overt and covert \ac{vsa} tasks.
+naturally interacted with a \ac{bci}, compared to standard overt and covert \ac{vsa} tasks.
 
@@
 calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
-The results of this study are presented in chapter~\ref{sec:patients}.
+The results of this study are presented in \autoref{sec:patients}.

### 2024-10-08 chapters/abstract.tex
Proofread

@@
 \chapter*{Abstract}
+
 Individuals with severe motor impairments, such as those with \ac{lis}

### 2024-10-08 chapters/abstract_fr.tex
Proofread

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de \ac{lis} ou de dysfonctionnement oculomoteur, rencontrent des défis considérables lors de l'utilisation d'interfaces cerveau-ordinateur \acp{bci} traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail aborde cette limitation en développant des méthodes de \ac{bci} indépendantes du regard, en se concentrant sur l'amélioration de la \ac{vsa} cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour le jitter de latence ERP, une variabilité qui impacte négativement la performance de décodage dans la \ac{vsa} cachée.
En gérant ce jitter, la méthode proposée améliore la précision de la communication sans nécessiter de contrôle du regard, rendant les \acp{bci} plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les techniques générales de décodage \ac{erp} en affinant la structure des estimateurs linéaires et multilinaires.
Ces méthodes améliorent la précision de classification dans une gamme de conditions \ac{bci}, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et multilinaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.

Les méthodes proposées ont été validées lors d'expériences impliquant à la fois des individus en bonne santé et 7 individus avec des handicaps physiques sévères et un contrôle oculomoteur altéré.
Les données ont été enregistrées à partir d'un groupe de contrôle, ainsi que de sept individus ayant des impairments du regard.
Ces expériences ont démontré la robustesse des nouvelles méthodes de décodage, montrant que le système pouvait efficacement décoder l'attention cachée même lorsque le regard direct était impossible.

\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéraire, handicap moteur, handicap du regard.

### 2024-10-08 chapters/bci.tex
Proofread

@@
-\todo{comments liesel paradigms}
-\todo{spellcheck}
-\todo{are all acronyms defined}
-\todo{fix all latex warnings to make sure no refs are missing (or grep on that
-specific warning)}
-\todo{uniform ref or autoref}
-\todo{don't capitalize figure, table, algorithm,...}
-\todo{define short captions for tables and figures}
-\todo{fix lacheck/chktex notifications}
-\todo{use terminology individuals with severe
+\chapter{Brain-computer interfaces}
+\label{sec:bci}
+\todo{search for ??}
+\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
+\todo{fix all latex warnings to make sure no refs are missing}
+\todo{uniform ref or cref}
+\todo{consistently terminology individuals with severe
 speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
 \todohsm{more subtitles}
-\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
-\todofc{chapters should look more like full papers}
 \todo{uniform vsa condition vsa setting}
-\todo{acronyms written in full every new chapter and in captions}
-\todo{search for ??}
-\todo{acro usage in chapters}
+\todo{are all acronyms defined/Acro corectly used in chapters?}
+\todo{acronyms written in full every new chapter and in captions?}
 \todo{are textcites used when necessary?}
+\todo{fix lacheck/chktex notifications}
+\todo{footnot skips a page in wcble chap}
 
-\chapter{Brain-computer interfaces}
-\label{sec:bci}
 People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
@@
 from this brain activity, and couples this output to a function of a device as
-illustrated in \autoref{fig:bci/loop}.
+illustrated in \cref{fig:bci/loop}.
 Optionally, the user can then observe the action of the device and adapt their behavior
@@
   In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
-}{%
-  fig:bci/loop
-}
+}{fig:bci/loop}
 
@@
 electrical activity.
-For several reasons laid out in \autoref{sec:bci-recording}, the electrical modality is
+For several reasons laid out in \cref{sec:bci-recording}, the electrical modality is
 especially interesting.
@@
 The recorded brain activity forms only one part of the interaction scheme presented in
-\autoref{fig:bci/loop}.
+\cref{fig:bci/loop}.
 It would make little sense to start digging around at random in all the brain activity
@@
 In turn, the detection of this signal can be coupled to an action, like typing the
-intended letter A or B in \autoref{fig:bci/loop}.
+intended letter A or B in \cref{fig:bci/loop}.
 The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
@@
 case of a closed-loop system.
-As shown in \autoref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
+As shown in \cref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
 based on different systems in the brain.
@@
 Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
-(see \autoref{sec:bci/preprocessing}) and by choosing the correct representation of the
+(see \cref{sec:bci/preprocessing}) and by choosing the correct representation of the
 signal (i.e., \emph{feature extraction}), but the problem of retrieving or
@@
 from it by the decoder and interpreted in function of the desired \ac{bci} output.
-\autoref{sec:bci/decoding} takes a closer look at some common decoding techniques.
+\Cref{sec:bci/decoding} takes a closer look at some common decoding techniques.
 
@@
 suitability for a given application and patient.
-As illustrated in \autoref{fig:bci/recording}, invasive electrodes can often measure a
+As illustrated in \cref{fig:bci/recording}, invasive electrodes can often measure a
 very specific brain region and can result in better signal quality, at the cost of the
@@
 According to \textcite{Zander2011}, they can be arranged by their reliance on external
-stimulation and the engagement of the user as shown in \autoref{fig:bci/paradigms}.
+stimulation and the engagement of the user as shown in \cref{fig:bci/paradigms}.
 
@@
 \begin{equation}
-	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
+  \text{\acs{itr}} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
 \end{equation}
@@
 \begin{figure}[t]
+  \centering
   \begin{subfigure}[t]{.45\textwidth}
   \includegraphics[width=\textwidth, height=\textwidth]{figures/bci/illustration}
-    \caption{Stimuli are flashed while the \ac{eeg} is recorded}
+    \caption{Stimuli are flashed while the \acs{eeg} is recorded.}
   \end{subfigure}\hfill%
@@
     channel.}
-  \end{subfigure}\hfill%
-  \begin{subfigure}[t]{.45\textwidth}
-    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
-  \end{subfigure}
+  \end{subfigure}%
 \end{figure}
-\todo{full spatiotemporal erp response}
 
@@
 component.
-In addition to the steps outlined in \autoref{sec:bci/preprocessing}, the signal is
+In addition to the steps outlined in \cref{sec:bci/preprocessing}, the signal is
 usually cut into \emph{epochs} for \ac{erp} analysis.
@@
 allows us to apply regular machine learning methods like an SVM or \ac{lda}.
-The tangent space is obtained\todo{tangent space}.
+%The tangent space is obtained\todo{tangent space}.
 To enhance classifications of \acp{erp} which rely on time-domain information, which
@@
 Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based oddball
-interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
+interface is an effective and proven method~\cite{Wolpaw2018,Severens2014}.
 Targets are shown in short flashes on a computer screen that evoke an \ac{erp} when
@@
 Decoding this modulation allows for information transfer controlled by the user's brain
-activity, as sketched in \autoref{fig:bci/loop}.
+activity, as sketched in \cref{fig:bci/loop}.
 
@@
 This points to the need for gaze-independent solutions, which we will explore further in
-\autoref{sec:gaze-independence} and the rest of this thesis.
+\cref{sec:gaze-independence} and the rest of this thesis.

### 2024-10-08 chapters/bttda.tex
Proofread

@@
 \section{Introduction}
-\todo{reduce length of the introduction}
 
@@
 Recent work proposes adaptations such as suited objective
-functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
+functions and regularization~\cite{JamshidiIdaji2017,Jorajuria2022,Aghili2023}.
 Discriminant tensor features have also been extracted
@@
 Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
-Regression (BTTR)~\cite{Faes2022,Faes2022b}.
+Regression (BTTR)~\cite{Faes2022,Faes2022a}.
 BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
@@
 decomposition is also restricted in flexibility.
-
-\todo{It's always tricky in the related work section of an ML/Mathematical
-paper, but unless your reviewers are absolute experts of this specific field,
-you should at least try to better describe the previous methods.}
 
@@
 \end{equation}
-as visualized in \autoref{fig:HODA-backward}.
+as visualized in \cref{fig:HODA-backward}.
 \begin{figure}[t]
@@
 update of each $\mat{U}_k$ is lower than a predetermined threshold.
-The full HODA procedure is summarized in \autoref{alg:HODA}.
+The full HODA procedure is summarized in \cref{alg:HODA}.
 \begin{algorithm}
@@
 
-The HODA projection in \autoref{eq:HODA-backward} is an example
+The HODA projection in \cref{eq:HODA-backward} is an example
 of a backward model.
 A straightforward and computationally efficient candidate for a corresponding
-forward model, visualized in \autoref{fig:HODA-forward}, is given by:
+forward model, visualized in \cref{fig:HODA-forward}, is given by:
 \begin{equation}
@@
 \begin{equation}
-	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
+	\left\{\mat{A}^*\right\}
+  = \argmin_{\{\mat{A}\}}\text{E}\left[
+      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
+    \right]_n
 \end{equation}
@@
 \begin{equation}
-	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
-  -\hat{\ten{X}}(n)\right)^2
-  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmprs{\mat{A}}\right)^2
+	\left\{\mat{A}^*\right\}
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \hat{\ten{X}}(n)\right)^2
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
 \end{equation}
 This least squares tensor approximation problem can be solved efficiently using the
-alternating least squares (ALS) algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
+alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
 \begin{equation}
@@
 The full algorithm to determine the HODA forward projection is listed
-in \autoref{alg:HODA-fw}.
+in \cref{alg:HODA-fw}.
 \begin{algorithm}
@@
 class separation, the error term $\ten{\mathbfcal{E}}$ in
-\autoref{eq:HODA-forward} should still contain some discriminative
+\cref{eq:HODA-forward} should still contain some discriminative
 information, which can be exploited to improve classifier
@@
 $\ten{\mathbfcal{E}}$.
-\autoref{fig:BTTDA} further illustrates the BTTDA model.
+The BTTDA model is further illustrated by~\cref{fig:BTTDA}.
 \begin{figure*}[t]
@@
 extraction.
-If $B$ in \autoref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
+If $B$ in \cref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
 HODA; if at each term $b$ the rank of the core tensor
@@
 The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
-computed through a deflation scheme summarized in \autoref{alg:BTTDA}.
+computed through a deflation scheme summarized in \cref{alg:BTTDA}.
 \begin{algorithm}
@@
 For each block $b$, the latent tensor is extracted using the HODA backward
-projection in \autoref{eq:HODA-backward} from the residual error term of the previous
+projection in \cref{eq:HODA-backward} from the residual error term of the previous
 block $\ten{\mathbfcal{E}}^{(b-1)}$:
@@
 To reduce the computational cost of model selection,
-\autoref{alg:model-selection} proposes a heuristic model selection
+\cref{alg:model-selection} proposes a heuristic model selection
 algorithm that leverages cross-validation in a greedy way per block, to
@@
 demand.
-Details about these datasets are found in \autoref{tab:moabb}.
+Details about these datasets are found in \cref{tab:moabb}.
 For the ERP datasets, the task is to distinguish target from non-target ERPs,
@@
 	\footnotesize
+  \makebox[\textwidth][c]{
 	  \input{chapters/bttda/moabb.tex}
+  }
   \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
@@
 stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
-greedy model selection algorithm in \autoref{alg:model-selection}.
+greedy model selection algorithm in \cref{alg:model-selection}.
 For HODA+LDA and the HODA blocks in BTTDA+LDA, we chose
@@
 EEG signals for the evaluated datasets were recorded at the sample rate given
-by \autoref{tab:moabb} and band-pass filtered between 1Hz
+by \cref{tab:moabb} and band-pass filtered between 1Hz
 and 24Hz.
 The signals were cut into epochs starting from stimulus onset with a
-dataset-specific length given by \autoref{tab:moabb}.
+dataset-specific length given by \cref{tab:moabb}.
 For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
@@
 
-\autoref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
+\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
 decoders.
@@
 
-The MI datasets listed in \autoref{tab:moabb} were first band-pass filtered
+The MI datasets listed in \cref{tab:moabb} were first band-pass filtered
 between 8 and 30Hz and cut into
-epochs with time windows as specified by \autoref{tab:moabb}.
+epochs with time windows as specified by \cref{tab:moabb}.
 Next, time-frequency transformation was performed using a complex Morlet wavelet
@@
 
-\autoref{tab:mi-score} lists the cross-validated classification scores for
+\cref{tab:mi-score} lists the cross-validated classification scores for
 the evaluated motor imagery datasets.
@@
 We report within-session ROC-AUC scores for training, validation, and test data as a function
-of the number of blocks increases, shown in \autoref{fig:blocks}.
+of the number of blocks increases, shown in \cref{fig:blocks}.
 Training and validation folds were taken from the model selection procedure.
@@
 \subsection{Contribution}
-The results in \autoref{tab:erp-score} and \autoref{tab:mi-score} show that
+The results in \cref{tab:erp-score} and \autoref{tab:mi-score} show that
 in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
@@
 subjects should show whether this holds.
-\autoref{fig:blocks} shows that there is added value in finding extra blocks
+\Cref{fig:blocks} shows that there is added value in finding extra blocks
 over the first HODA block.
@@
 being more parsimonious due to its block-term structure compared to HODA's full
-Tucker structure, as illustrated by \autoref{fig:bttda/sparse}.
+Tucker structure, as illustrated by \cref{fig:bttda/sparse}.
 \begin{figure}
@@
 relevant discriminatory information at each block~\cite{Haufe2014}.
-\autoref{fig:forward} shows the activation patterns
+\Cref{fig:forward} shows the activation patterns
 of two blocks obtained from the BNCI2014-008 dataset as well as the forward
@@
 e.g., separate ERP components or neural processes based on the task-related
-information in the class labels. \todo{spellcheck from here}
+information in the class labels.
+
+\subsection{Model selection and dimensionality}
+
+BTTDA trades in the rigid HODA model for increased model complexity with more
+hyperparameters to tune, which could open up a setting where performance can be
+improved.
+This turns tensor discriminant analysis into a model selection procedure
+instead of a projection optimization procedure.
+Despite favorable results in BCI decoding, the applications of the proposed
+BTTDA model are limited mainly by this model selection approach used
+to determine the individual block ranks.
+While our proposed greedy model selection algorithm is a step in the right
+direction, the high computational cost of setting hyperparameters through cross
+validation can still hinder the portability of decoders relying on
+BTTDA.
+
+Due to its heuristic nature, the greedy algorithm does not always result in the
+set of ranks with the highest achievable performance.
+In combination with the fact that the feature selection cutoff parameter was
+fixed somewhat arbitrarily, it is clear that thorough
+hyperparameter optimization could improve performance.
+Furthermore, it is clear that our proposed model selection procedure does not
+necessarily result in an optimal set of blocks that group coherent projections
+within the same block, according to some desirable metric.
+Examples of this are sparsity, pattern interpretability, minimal or maximal
+between- or within-block feature correlation, decreasing discriminability, etc.
+Finally, the presented results are limited by our artificial restriction
+that $(r_1=r_2=\ldots=r_k)$ to reduce the computational demand of the presented
+experiments.
+In a sense, this goes against the proposition of increased model flexibility.
+It might well be that BTTDA in some cases offers little added value over the
+Tucker-structured HODA, when both are given free choice of rank.
+Future efforts should focus on automatic parameter setting, e.g., using sparsity
+or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
+statistical measures based on the model's application.
+
+Another limitation is that BTTDA might yield a disproportionate
+improvement for datasets with a low number of features relative to sample size,
+while being less effective for datasets with more features.
+This is reflected in our ERP results (low dimensionality vs. high number of
+trials) compared to the MI results (higher dimensionality due to third-order
+tensorization vs. lower number of trials).
+We expect a dimensionality limit beyond which the forward modeling step cannot
+accurately regress from the low-dimensional latent tensors to the high-
+dimensional original tensors, introducing
+error in the input data for the next block, which can stack up over blocks.
+Since the forward multilinear least squares problem is underdetermined, it is
+prone to numerical instability, which calls for regularization of the forward
+modeling procedure, but this would introduce another hyperparameter.
+It should also be thoroughly investigated what the impact is of going beyond
+second- and third-order cases to higher-order tensors, since this could have a
+large impact on the model.
+Other tensorization methods of the EEG data, like time-lagged Hankel
+tensors~\cite{Papy2005}, or tensors across subjects or sliding windows, etc.,
+could also be of interest if they are appropriately chosen based on prior
+knowledge of the dataset.
+
+\section{Conclusion}
+
+We have introduced Block-Term Tensor Discriminant Analysis (BTTDA), a novel,
+tensor-based, supervised dimensionality reduction technique optimized for class
+discriminability, which adheres to the block-term tensor structure.
+BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
+applied as a special sum-of-rank-one tensors PARAFACDA model.
+The model is obtained by iteratively fitting HODA in a deflation scheme,
+leveraging a novel forward modeling step.
+
+Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
+feature extraction can significantly outperform decoders based on HODA and
+reach state-of-the-art decoding performance on event-related potential
+problems (second-order tensors) and scores on par with or higher than HODA in motor
+imagery problems (third-order tensors).
+Moving from the rigid Tucker tensor structure of HODA to the more flexible
+block-term structure shifts the problem from finding optimally constrained
+multilinear projections to model and feature selection.
+
+Introducing a flexible block-term tensor model as the underlying structure
+reformulates tensor discriminant analysis as a model selection
+problem.
+This allows performance to be traded off for model complexity and the number of
+features, to find a setting that is more effective for decoding.
+Because of its general implementation and minimal assumptions on data structure,
+BTTDA can equally be applied to classification for other neuroimaging modalities
+(MEG, ECoG, fNIRS, fMRI, EMG, ...) or to tensor classification problems in other
+fields.
+
+%\section*{Acknowledgements}
+%\todo{move to the end}
+%We thank the Flemish Supercomputer Center (VSC) and the High-Performance
+%Computing (HPC) center of the KU Leuven for allowing us to execute our
+%computational experiments on their systems.
+%We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
+%work.

### 2024-10-08 chapters/conclusion.tex
Proofread

--- 2024-10-07 chapters/conclusion.tex
@@
 This work aimed to improve the performance of visual gaze-independent \acsp{bci}, in general
-and applied to individuals with eye motor impairment by developing novel decoding
+and when applied to individuals with eye motor impairment, by developing novel decoding
 strategies and evaluating their effectiveness.
@@
 \acp{erp}.
-This work is submitted as~\textcite{Vandenkerchove2024b}.
+This work is submitted as~\textcite{VanDenKerchove2024}.
 
@@
 The method has been published in~\textcite{VanDenKerchove2024}.
-We took care to design a method that can be applied to improve decoding, where
+We designed this method to improve decoding where
 unseen incoming test data is not yet known.
@@
 effectiveness under conditions of controlled latency variability and noise.
-These results showed that is is robust to higher noise and jitter compared to
+These results showed that it is robust to higher noise and jitter compared to
 a non-iterative method in a latency estimation task.
 It was also shown that decoding accuracy was higher in high jitter and noise
-settings compared to the non-iterative method and a sate-of-the-art decoder.
+settings compared to the non-iterative method and a state-of-the-art decoder.
 
@@
 The dataset was presented in~\textcite{VanDenKerchove2024}.
-The experiment in this dataset implemented a visual odball \ac{bci} with six
+The experiment in this dataset implemented a visual oddball \ac{bci} with six
 circular targets in a hexagonal layout.
@@
 visuospatial attention of the participant and their eye gaze.
-This allows us to study effects of gaze-independence on \ac{erp} decoding.
-The CVSA-ERP dataset gives us insight in the ERP dynamics in overt, covert and the
+This allowed us to study the effects of gaze-independence on \ac{erp} decoding.
+The CVSA-ERP dataset gives us insight into the ERP dynamics in overt, covert, and the
 novel split VSA condition, and confirms our hypothesis that P3 jitter has a
 significant impact on performance in covert and split VSA.
-It also confirmed the effects of covert \ac{vsa} on \ac{erp} component
+It also confirmed that the effects of covert \ac{vsa} on \ac{erp} component
 amplitude hold for split \ac{vsa}.
@@
 
-We have evaluated our proposed \ac{wcble} algorithm, its non-iterative
-counterpart and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
+We evaluated our proposed \ac{wcble} algorithm, its non-iterative
+counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
 dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
@@
 While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
-decoding, but increased covert \ac{vsa} decoding.
+decoding, it increased covert \ac{vsa} decoding.
 
@@
 To further study the gaze-independent performance of these algorithms, transfer
-learning between \ac{vsa} conditions is studied to simulate conditions where an
+learning between \ac{vsa} conditions was studied to simulate conditions where an
 individual with gaze impairment can end up in different \ac{vsa} settings within
-a \ac{bci}operation session due to their lack of proper motor control.
+a \ac{bci} operation session due to their lack of proper motor control.
 When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
-results in a significant decrease in performance compared the state-of-the art,
+results in a significant decrease in performance compared to the state-of-the-art,
 consistent with the within-condition results.
-For all other pairs of training and evaluation VSA condition, however,
-\ac{wcble} was equal or significantly better.
+For all other pairs of training and evaluation VSA conditions, however,
+\ac{wcble} was equal to or significantly better.
 
@@
 
-The \ac{wcble} decoder improved decoding accuracy covert
+The \ac{wcble} decoder improved decoding accuracy in covert
 \ac{vsa} settings with cued central gaze fixation.
-This is of interest for gaze-independent decoding, but not as optimal as if the
+This is of interest for gaze-independent decoding, but not as optimal as when the
 gaze fixation was uncued.
-The latter was tested in the free \ac{vsa} settion.
+The latter was tested in the free \ac{vsa} setting.
 Here, participants were generally more comfortable when allowed to use their
@@
 Together, these results show that there is an interest in developing a new class of ERP-BCI
-interfaces for patients that prefer to choose their own gaze strategy, so as to
-avoid the effort of redirecting their eye gaze at different spatial locations
+interfaces for patients that prefer to choose their own gaze strategy, to
+avoid the effort of redirecting their eye gaze to different spatial locations
 on the stimulation screen in manners that might be uncomfortable for them.
@@
 solution for the Midas Touch Problem in visual \ac{bci}.
-Here a \ac{bci} user sometimes accidentally selects a target while not wanting
+Here, a \ac{bci} user sometimes accidentally selects a target while not intending
 to give any input.
@@
 structure~\cite{Bijma2005}.
+The Kronecker-\ac{lda} approach has also been extended to the combined
+space-time-frequency domain, where more information can be modeled at the cost
+of increased dimensionality.
+A properly structured covariance model can strongly regularize the problem.
 
 To overcome the limitations introduced by the presence of multiple \ac{erp}
-components to latency estimation, we have also developed a multi-component
+components in latency estimation, we have also developed a multi-component
 version of the \ac{wcble} algorithm.
 This algorithm should be able to separate mixed \ac{erp} component clusters
-based on their temporal coherence, in a similar maner to~\textcite{Ouyang2017}.
+based on their temporal coherence, in a similar manner to~\textcite{Ouyang2017}.
 It could also yield favorable decoding results in a broader range of settings
-than overt \ac{vsa} since it should be able to account for the presence of both
-visual and attentional \ac{erp}components}.
+than overt \ac{vsa}, since it should be able to account for the presence of both
+visual and attentional \ac{erp} components.
 
@@
 counterintuitively, integrating eye tracking into the decoding strategy.
-Since we are able to explicitly discern overt, covert and split \ac{vsa} from
+Since we are able to explicitly discern overt, covert, and split \ac{vsa} from
 the \ac{erp}, and different decoders perform best in different settings, it
 could be helpful to derive the current setting from the gaze position and the
-\ac{erp} and propose the most suited decoder for a given data point.
-select best classifier using eye tracking
+\ac{erp}, and propose the most suited decoder for a given data point.
+This could allow us to select the best classifier using eye tracking.
 
-We also wish to apply the tensor approach to gaze-inependent decoding.
-The problems introduced by jitter could be accounted for as in \ac{wcble} by
+We also aim to apply the tensor approach to gaze-independent decoding.
+The problems introduced by jitter could be accounted for, as in \ac{wcble}, by
 alignment, but also by other methods that model the possible time shifts.
 In ordering the data as a Hankel tensor\footnote{Adding an extra mode with
-time-shifted copys of the temporal response per channel.}, we should have a
+time-shifted copies of the temporal response per channel.}, we should have a
 data representation that is more robust to jitter, at the cost of increased
 dimensionality.
-This increased dimensionality can than be countered by a tensor method like
+This increased dimensionality can then be countered by a tensor method like
 \ac{hoda} or \ac{bttda}.
 
-Finally, we wish to expand on experimentation with individuals with
+Finally, we wish to expand our experimentation with individuals with
 \acp{sspgi}.
 We aim to revisit the patients with an on-line experiment and comfort
-assessment, to establish the usability of the proposed interface and to be able
-to compare its \ac{itr}  to the state-of-the-art literature.
+assessment to establish the usability of the proposed interface and compare its
+\ac{itr} to the state-of-the-art literature.
 
@@
 Throughout this thesis, we've noticed that linear or multilinear decoding
-methods consistentcly outperform non-linear methods, like SVMs or those based
-in Riemannian Geometry.
+methods consistently outperform non-linear methods, such as SVMs or those based
+on Riemannian Geometry.
 Within these linear models, regularization (shrinkage, covariance structure,
@@
 inherently multivariate and \ac{bci} calibration time is minimized.
-Proper regularization should impose some specific structure on linear model,
+Proper regularization should impose some specific structure on the linear model,
 ideally reflecting properties of the signal.
-When faced with new problem and few data relative to dimensionality, it is a
+When faced with a new problem and few data relative to dimensionality, it is a
 good first instinct to pick a simple, restrictive structure.
 
-Yet, the brain is a complex organ, hence neural data is not linear in origin and
-these regularizing assumptions based on prior knowledge are probably only
+However, the brain is a complex organ, so neural data is not linear in origin, and
+these regularizing assumptions based on prior knowledge are likely only
 superficially true.
@@
 For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
-improves performance in low sample sizes, but is outperformed by more flexible
-methods, e.g.\ \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
-if training sample size increases.
-Similarly, \ac{bttda} is more flexible
+improves performance in low sample sizes but is outperformed by more flexible
+methods, such as \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
+when the training sample size increases.
+Similarly, \ac{bttda} offers more flexibility.
 
 This calls for the development of more flexible (multi-)linear models that still rely on some
-structure, but which are combined with efficient model
-selection, such as proposed by \ac{bttda}.
+structure but are combined with efficient model
+selection, such as those proposed by \ac{bttda}.
 These models try to combine the best of both worlds.
 
-%\subsection{Visual? gaze-independent \ac{bci}}
-%
-%While visual \acp{bci} are often praised for their easy, non-invasive
-%applicability, they are also inherently unreliable, require wearing the
-%\ac{eeg} cap and force continuous stimulation upon the user, all while forming
-%a rather unintuitive method of communication for humans who are used to speech
-%or motor commands.
-%
-%Currently, the group of users that would directly be helped with \emph{only} a
-%gaze-independent visual \ac{bci} is not that large.
-%Even if some eye motor impairment is the most suited strategy for a patient, it will probably be
-%improved if a combination of eye tracking, residual motor control and other
-%biosensors are used.
-%
-%
-%\todo{https://www.nature.com/articles/s44222-024-00239-5}
-%very small group of interest, but available right now, no implantation needed,
-%
-%true power in the future is probably focused on
-%invasive, active paradigms if recording technology and surgical methods
-%improve lowering the risk. orders of magnitude higher itr and more natural
-%communication.
-%will probably become acceptable if risks of implantation and longevity
-%increase.
-%characterized by publication date, but might indicate it is of interest to
-%heavily focus this research now on applied 'vraagstukken' with patients
-%
-%Why limit ourselves to the visual~\ac{bci} decoding problem postulated in this
-%thesis?
-%
-%
 \subsection{Optimize user experience}
-While our work with \ac{bci} users with \ac{sspgi} results in interesting
-preliminary results, it is currently impossible to make claims about the experience in long term
+While our work with \ac{bci} users with \ac{sspgi} resulted in interesting
+preliminary findings, it is currently impossible to make claims about the experience in long-term
 home use of the proposed systems as assistive technology.
-This work focused heavily on classification methods, and solving problem
+This work focused heavily on classification methods and solving problems
 arising from gaze-independence through decoding.
@@
 is user satisfaction.
-The latter can depends on performance, comfort and perception.
-Classification performance is a convenient metric since a researcher can perform a single
+The latter depends on factors such as performance, comfort, and user perception.
+Classification performance is a convenient metric because a researcher can perform a single
 data collection and iterate on the results offline.
-However, this can be too constrictive when seeking the most suited interface
-for a specific (group of users).
+However, this can be too limiting when seeking the most suitable interface
+for a specific group of users.
 
-In the optimal approach, decoder development, paradigm choice and interface
-design should working together, such as is advocated
-by~\textcite{Pan2022} and~\textcite{Fouad2020}, by unifying the engineering and
-the clinical side.
-Performance and abilities are probably depend in such a degree on the user,
-that off-line decoding performance in healthy controls is probably not a great
+The optimal approach should unify decoder development, paradigm choice, and interface
+design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
+engineering and clinical perspectives.
+Performance and abilities likely depend to a great degree on the user,
+so off-line decoding performance in healthy controls is probably not a good
 predictor of on-line decoding performance and \ac{itr} in individuals with
 \ac{sspi}.
-Incremental gains in performance in decoding for a specific paradigm might
-also be outweighed by how well that paradigm is adapted to the user, factoring in,
-skills and abilities.
-Longitudinal research should follow several patient groups specifically with gaze
-impairment and experiment with specific paradigms, interfaces and decoders for
+Incremental gains in decoding performance for a specific paradigm might
+also be outweighed by how well that paradigm is adapted to the user, factoring in
+their skills and abilities.
+Longitudinal research should follow several patient groups, specifically those with gaze
+impairment, and experiment with specific paradigms, interfaces, and decoders for
 a given \ac{bci} user.
@@
 
-A logical point that arising from this is the question of how to setup an
+A logical point arising from this is the question of how to set up an
 effective research project that aims to optimize \ac{bci} user experience, in
-the context of he topics presented in this work.
+the context of the topics presented in this work.
 The patient-research involvement framework postulates that, ideally, individuals with
@@
 to evaluation.
-A good rule of thumb is that research should solves problems \emph{with} the
+A good rule of thumb is that research should solve problems \emph{with} the
 users, not \emph{for} them, implying that research projects should start from
-the need of the \ac{bci} user population.
+the needs of the \ac{bci} user population.
 
 Instead of choosing a research target and then moving on to decoder
-development, data collection with healthy controls, patient studies and,
+development, data collection with healthy controls, patient studies, and
 finally, verification in on-line operation, it could be more effective to start
@@
 It should then become apparent where challenges for the envisioned user
-(population) lie.
+population lie.
 As an example, \textcite{FriedOken2020} share an overview of their experiences from this step,
@@
 After that, hypotheses can be formed on how to mitigate these challenges
-through interface optimization or choice taking into account the full picture
+through interface optimization or choice, taking into account the full picture
 of the capabilities and skills of the users.
-These hypotheses can then verified in experiments
+These hypotheses can then be verified in experiments
 with the users and/or a population of healthy controls.
@@
 Unfortunately, this is also probably one of the least practical ways to set up
-a 4-year research project, e.g.\ in the context of doctoral thesis.
+a 4-year research project, such as in the context of a doctoral thesis.
 This approach can only work if the proper facilities can be gathered immediately
-from the start, like ethical approval and access to patients and on-line \acp{bci}
+from the start, such as ethical approval, access to patients, and on-line \acp{bci}
 experimentation systems.
-It is therefore important that research labs keep a working, long-term
-collaboration with patient centers such that new projects can immediately be v
-verified with population, and that interested users can compare multiple systems
-and guide development.
+It is therefore important that research labs maintain a working, long-term
+collaboration with patient centers, so that new projects can immediately be
+verified with the user population. This also allows interested users to compare
+multiple systems and guide development.
 The fact that a user has previous \ac{bci} experience is often seen as a confounding
 factor in off-line studies.
-When the research is in a phase that is heavily dependent on user
-experimentation,
-this is in fact a strength, sinche the user might help the researcher gain
-insight into what works and doesn't work for them, and where research efforts
+When the research is in a phase that heavily depends on user
+experimentation, this is actually a strength, since the user can help the researcher gain
+insight into what works and doesn’t work for them, and where research efforts
 should be focused.
 
-Similarly, there is value in maintaining a working, on-line in house \ac{bci}
+Similarly, there is value in maintaining a working, on-line in-house \ac{bci}
 system that can be iterated upon and adapted if necessary.
 Experiments can be supported by advance calculations of hypothetical
-\ac{itr}, such that performed on-line or off-line experiments reasonably reflect
-real-world operation and no effort is wasted exploring settings that should not
-reach satisfactory performance.
+\ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
+real-world operation, and no effort is wasted exploring settings that will not
+achieve satisfactory performance.
 For example, why study a specific number of repetitions or \ac{isi}, if they
-would not be practical to achieve a high \ac{itr}, but can still influence
+would not be practical to achieve a high \ac{itr}, but could still influence
 decoding performance?
@@
 research.
-Yet, in visual oddbal \ac{bci} where the challenge lies no longer in generating
-a proof of concept but in translating the concept into a useful technology, it
-might help avoid prematurely solving problems that might not necessarily be
-encountered further down the road.
-A top-down approach might improve the quality of study outcomes by enabling the
+Yet, in visual oddball \ac{bci}, where the challenge no longer lies in generating
+a proof of concept but in translating the concept into useful technology, this
+approach might help avoid prematurely solving problems that may not necessarily
+arise further down the road.
+A top-down approach could improve the quality of study outcomes by enabling the
 reporting of realistic, interpretable metrics,
-and take into account the human aspect of working \emph{with} \ac{bci} users.
-
-\section{Conclusion}
+and by taking into account the human aspect of working \emph{with} \ac{bci} users.

### 2024-10-08 chapters/contribution.tex
Proofread

@@
 \chapter*{Personal contribution}
-\todo{which works conceptualized}
-\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
-Hakim, Marie,Etienne, Philip Van Damme  }
+Arne Van Den Kerchove conceptualized the works and performed the analyses in
+this thesis, under supervision and in consultation with prof. Hakim Si-Mohammed,
+prof. Marc Van Hulle and prof. François Cabestaing.
+Dr. Arno Libert and Dr. Benjamin Wittevrongel consulted on the analyses performed in
+\autoref{sec:stbf-struct}
+Data collection was performed by Arne Van Den Kerchove for the dataset
+presented in \autoref{sec:covert-align} and by Arne Van Den Kerchove and
+Juliette Meunier for the dataset presented in \autoref{sec:patients}.
+Participant recruitment at TRAINM was performed by Alixe Willemssens under
+supervision of dr. Edward Schietekatte, at Fondation Partage et Vie by Marie de
+Moura under supervision of prof. Etienne Allart and at UZ Leuven Neuromuscular reference
+center under supervision of prof. Kristl Claeys and prof. Philip Van Damme.
+Prof. Etienne allart provided clinical consultation.

### 2024-10-08 chapters/covert_align.tex
All text
Proofread

@@
 
-The decoding methods described in \autoref{sec:stbf-struct} and
-\autoref{sec:bttda} are rather general attempts at increasing \ac{erp} decoding accuracy.
+The decoding methods described in \cref{sec:stbf-struct} and
+\cref{sec:bttda} are general attempts at increasing \ac{erp} decoding accuracy.
 Given the problem statement of this work, we are also interested in targeted
 methods that would specifically advance the development of gaze-independent
-visuall \acp{bci} (see \label{sec:gaze-independence/sota}).
+visual \acp{bci} (see \cref{sec:gaze-independence/sota}).
 Methods of directly enhancing gaze-independent decoding through optimizing
 decoder design have not been widely studied (see
-\label{sec:gaze-independence/sota/decoding}).
+\cref{sec:gaze-independence/sota/decoding}).
 We assume that advances in gaze-independent decoding can be made if properties
-of covert \ac{vsa} that affect performance can be identified and, in some way,
-accounted for.
-The latency estimation and alignment approach proposed in \autoref{sec:wcble}
+of covert \ac{vsa} that affect performance can be identified and accounted for.
+The latency estimation and alignment approach proposed in \cref{sec:wcble}
 was specifically designed for this purpose.
@@
 \ac{vsa}.
-This latency variability contributes to reduced covert VSA decoding performance.
+This latency variability contributes to reduced covert \ac{vsa} decoding performance.
 While they proposed an analysis and performance prediction method, they did not provide a
 decoding solution.
-They suggested that compensating for latency jitter could enhance covert VSA
-decoding, but did not actually verify this hypothesis directly.
+They suggested that compensating for latency jitter could enhance covert \ac{vsa}
+decoding, but did not verify this hypothesis directly.
 Additionally, \cite{Hardiansyah2020} developed a classifier for
-covert VSA ERPs, exploiting single-trial latency features in combination with
+covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
 amplitude features for classification with a support vector machine.
-They demonstrated the positive influence of single-trial ERP component latency
-features on covert VSA inference, yet did not attempt to correct the amplitude
+They demonstrated the positive influence of single-trial \ac{erp} component latency
+features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
 features for these latencies.
 
-Furthermore, as mentioned in~\autoref{sec:gaze-independence/sota/visual} and
-\autoref{sec:gaze-independence/approach/cvsa-erp},
-our goal is also to explore more flexible gaze-independent settings that the
+Furthermore, as mentioned in \cref{sec:gaze-independence/sota/visual} and
+\cref{sec:gaze-independence/approach/cvsa-erp},
+our goal is also to explore more flexible gaze-independent settings than the
 usual covert \ac{vsa} with central gaze fixation reported in most literature.
 \textcite{Frenzel2011} introduced a similar protocol to the proposed split
-\ac{vsa} setting (see~\autoref{fig:gaze/vsa/free}).
-They showed that it is possible to perform split VSA and that, in this
-case, visuospatial attention and gaze direction can separately be decoded using
-classical ERP techniques.
+\ac{vsa} setting (see~\cref{fig:gaze/vsa/free}).
+They showed that it is possible to perform split \ac{vsa} and that, in this
+case, visuospatial attention and gaze direction can be decoded separately using
+classical \ac{erp} techniques.
 To the best of our
-knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
+knowledge, this is the sole study that investigated split attention in \ac{erp}-based \acp{bci}.
 However, \textcite{Frenzel2011} considered their interface only for the case
 where the user actively intends to select both targets determined by the gaze
-and the VSA.
-In the split VSA setting considered in our work, we rather instruct the participant to ignore
-the distractor and only attend the cued target, since we are interested in
+and the \ac{vsa}.
+In the split \ac{vsa} setting considered in our work, we instruct the participant to ignore
+the distractor and only attend to the cued target, as we are interested in
 decoding the visuospatial attention only.
@@
 We evaluate our approach on a dataset specifically recorded for this study, designed to
-probe different modalities of covert VSA, and on the publicly available
+probe different modalities of covert \ac{vsa}, and on the publicly available
 BNCI2014-009 dataset~\cite{Aloise2012a}.
@@
 We recorded a dataset to validate our approach.
-The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
+The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
 consists of 15 participants, mean age $26.34\pm3.04$ years.
@@
 (S62547).
-Each subject performed different VSA conditions (overt, covert and split),
+Each subject performed different \ac{vsa} conditions (overt, covert, and split),
 illustrated in Figure~\ref{fig:interface} (top row).
 Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
-by \cite{Treder2010}, we present six flashing targets (without letters or
-symbols) to the participant while the EEG, electrooculogram (EOG), and the
+by \cite{Treder2010}, we presented six flashing targets (without letters or
+symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
 participant's eye gaze using eye tracking were recorded.
-The VSA conditions described in the first row of Figure~\ref{fig:interface} are
+The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
 considered.
@@
 
-  \caption[Stimulation interfaces and evoked \acp{erp}.]{
-		\subref{fig:interface} Interfaces and visuospatial attention (VSA)
+  \caption[Stimulation interfaces and evoked \acsp{erp}.]{
+		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
     conditions in the CVSA-ERP and BNCI2014-009 datasets.
-		In the CVSA-ERP oddball BCI interface, screen targets are intensified one after
+		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
     the other in pseudorandom order while the participant can either pay overt,
-    covert, or split VSA to the cued target.
-		In the BNCI2014-009 overt VSA interface, entire rows and columns are
+    covert, or split \ac{vsa} to the cued target.
+		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
 		intensified at once. In its covert counterpart,
@@
 		\subref{fig:erps} Contrast between target (color) and non-target, and
-    distractor (gray) and non-target grand average event-related potentials per VSA condition and dataset.
-    Overt VSA yields a strong modulation of the N1 component in both datasets;
-    the P3 amplitude decreases with the degree of split VSA.
-    In split VSA, N1 and P2 are more prominently evoked by the distractor,
+    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
+    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
+    the P3 amplitude decreases with the degree of split \ac{vsa}.
+    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
     while the P3 is evoked by the target.
@@
 
-In contrast to the protocol proposed by \cite{Frenzel2011}, split VSA was
-performed by instructing the participant to attend the intensifications of the
-cued target, and ignore the intensifications of the distractor target.
+In contrast to the protocol proposed by \cite{Frenzel2011}, split \ac{vsa} was
+performed by instructing the participant to attend to the intensifications of the
+cued target and ignore the intensifications of the distractor target.
 Since we assume there will be an effect depending on the distance between
-attended target and the distractor, we discern three split VSA sub-conditions:
+the attended target and the distractor, we discern three split \ac{vsa} sub-conditions:
 the distractor is either clockwise or counterclockwise directly next to the
 attended target ($d=1$), there is one other target between the attended target and
-the distractor ($d=2$), or the distractor is opposite to the intended target
+the distractor ($d=2$), or the distractor is opposite the intended target
 ($d=3$).
 
-EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
+\ac{eeg} for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
 (Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
 international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
-Germany), with electrodes located at AFz and FCz as ground and reference respectively.
+Germany), with electrodes located at AFz and FCz as ground and reference, respectively.
 Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
@@
 The power line frequency in Belgium is 50 Hz.
-Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
+The participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
 Canada) in non-fixation mode.
 
-Participants signed the informed consent form and were seated at a distance of
-60 cm before a CRT-emulating monitor (VPixx
+Participants signed an informed consent form and were seated at a distance of
+60 cm from a CRT-emulating monitor (VPixx
 Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
@@
 conforming to the interface proposed by \textcite{Treder2010}
-(\autoref{fig:covert-align/interface/interface})
+(\cref{fig:covert-align/interface/interface}).
 A hexagonal layout interface with an empty center and a low number of targets
-counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
+counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
 targets, no other target can be between the subject’s gaze and a covertly
@@
 larger size (5.60° of visual angle,
-(\autoref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
+(\cref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
 experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
 peripheral visual field.
-Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.
+Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.
 
@@
     \input{figures/covert_align/interface_4.tikz.tex}
-    \caption[Gaze fixated centrally]{In the covert \ac{vsa} setting, the participant is
-    cued to fixate the center of the interface.}
+    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
+    cued to fixate at the center of the interface.}
     \label{fig:covert-align/interface/covert}
@@
 of stimulations.
-Then, one target was indicated as the cue and the participant
+Then, one target was indicated as the cue, and the participant
 was instructed to count the number of intensifications of the cued target
@@
 was instructed to fixate their gaze on the blue crosshair for the duration of
-the stimulation block (\autoref{fig:covert-align/interface/overt} and
-\autoref{fig:covert-align/interface/overt}).
-The position of this crosshair determined the VSA condition for this trial:
-overt VSA when the crosshair was at the same location as the cued target,
-covert VSA when the crosshair appeared in the center of the screen, and split
-VSA when the crosshair appeared on a different target than the cued one.
+the stimulation block (\cref{fig:covert-align/interface/overt} and
+\cref{fig:covert-align/interface/covert}).
+The position of this crosshair determined the \ac{vsa} condition for this trial:
+overt \ac{vsa} when the crosshair was at the same location as the cued target,
+covert \ac{vsa} when the crosshair appeared in the center of the screen, and split
+\ac{vsa} when the crosshair appeared on a different target than the cued one.
 After pressing the space bar again and a delay of 5 seconds, the stimulation block
-starts.
+started.
 
-All targets were intensified for a duration of 100 ms, in pseudorandom order.
-The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
-was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
-with 200ms uniform jitter added, resulting in an ISI
+All targets were intensified for a duration of 100 ms in pseudorandom order.
+The inter-stimulus-interval (\ac{isi}), the time between the onsets of subsequent intensifications,
+was variable and consisted of a fixed 300 ms interval (of which 100 ms with an intensified target onscreen)
+with 200 ms of uniform jitter added, resulting in an \ac{isi}
 between 200 and 400 ms.
@@
 longer \ac{isi} will increase component amplitude and aid in counteracting temporal
-autocorrelation for a higher statistical test precision.
+autocorrelation for higher statistical test precision.
 In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
@@
 After inputting this number, the subject was allowed to pause until pressing the space bar again.
-In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
-split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
-($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
+In total, six blocks were presented for overt \ac{vsa}, six blocks for covert \ac{vsa}, 12 blocks for
+split ($d=1$) \ac{vsa}, 12 blocks for split ($d=2$) \ac{vsa}, and 6 blocks for split
+($d=3$) \ac{vsa}, covering all possible combinations of \ac{vsa} conditions, cued targets, and
 crosshair locations.
@@
 The experiment started with five non-recorded practice stimulation blocks, one for
-each of the 5 VSA conditions.
+each of the five \ac{vsa} conditions.
 During these practice blocks, the participant received feedback about their gaze
@@
 Counting the instructions and the participant's response to the
-input prompts, a block lasted about 30
-seconds. In sum, the experiment featured approximately 45 minutes of
+input prompts, a block lasted about 30 seconds. In sum, the experiment featured approximately 45 minutes of
 stimulation time.
@@
 \subsubsection{BNCI2014-009 dataset}
-The
-BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
+The BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
 dataset~\cite{Aloise2012a} was used in the analysis performed
-in \cite{Arico2014}.
+by \cite{Arico2014}.
 It contains data from 10 subjects (median age $24.5\pm1.9$
-years)  that performed two spelling tasks illustrated in the second row of
-Figure~\ref{fig:interface}: using the P3 Matrix speller
-interface to exploit overt VSA, and the GeoSpell covert VSA interface.
+years) who performed two spelling tasks illustrated in the second row of
+\cref{fig:interface}: using the P3 Matrix speller
+interface to exploit overt \ac{vsa}, and the GeoSpell covert \ac{vsa} interface.
 To use the GeoSpell interface, the participant gazes at the fixation point at the
@@
 The user directs their visuospatial attention to the location where the intended letter is expected
-to appear, and when it does, a P3 ERP component is expected to be evoked.
+to appear, and when it does, a P3 \ac{erp} component is expected to be evoked.
 This results in a specific setting where both visuospatial attention and
@@
 1.3.1)~\cite{Gramfort2013}.
-All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
+All datasets were band-pass filtered between 0.1 Hz and 20 Hz with a 4th-order Butterworth filter.
 Bad channels in the data were automatically detected using the RANSAC
 method~\cite{Fischler1981} and rejected.
-The recorded EEG was re-referenced
+The recorded \ac{eeg} was re-referenced
 offline to the average of the mastoid electrodes TP9 and TP10.
-Next, the EEG signals were corrected for eye movement artifacts using an
+Next, the \ac{eeg} signals were corrected for eye movement artifacts using
 Independent Component Analysis (ICA).
-Since we have access to EOG data for the CVSA-ERP dataset components correlating
-significantly with the EOG were rejected.
+Since we have access to \ac{eog} data for the CVSA-ERP dataset, components correlating
+significantly with the \ac{eog} were rejected.
 For the BNCI2014-009 dataset, ICA components were manually rejected.
-Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
+Finally, the \ac{eeg} signal was divided into epochs ranging from 100 ms before stimulus onset to 700 ms after stimulus onset and down-sampled to 128 Hz.
 In both datasets, only 16 channels were kept for
-analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
+analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7, and
 PO8).
@@
 \subsubsection{Decoders}
-We will compare \ac{wcble} and \ac{cble} as described in
-\autoref{sec:wcble/methods}, with their base classifier and tLDA and with
-Riemannian Geometry
-approaches that rely on spatial covariance as features.
-Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
+\label{sec:covert-align/methods/decoders}
+We compared \ac{wcble} and \ac{cble} as described in
+\cref{sec:wcble/methods}, with their base classifier \ac{tlda} and with
+Riemannian Geometry approaches that rely on spatial covariance as features.
+Together with \ac{tlda}, Riemannian Geometry generally achieves state-of-the-art decoding
 performance~\cite{Lotte2018}.
 We implemented two Riemannian Geometry pipelines.
-The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
-filters, projects these covariances to a tangent space and classifies the
+The first estimates shrunk covariances from the \acp{erp} filtered with 6 XDAWN
+filters, projects these covariances to a tangent space, and classifies the
 result using $L_2$-regularized logistic regression
 (XDAWNCov-TS-LR)~\cite{Cecotti2017}.
-Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
-favorable performance in the presence of single-trial ERP latency jitter.
-Shrunk spatial covariance matrices are estimated from epochs that are
-augmented by concatenating the average target and average non-target ERP as
-extra channels, projected to tangent space and classified using
+Secondly, we adopted the pipeline from \cite{Aydarkhanov2020}, as their work shows
+favorable performance in the presence of single-trial \ac{erp} latency jitter.
+Shrunk spatial covariance matrices were estimated from epochs that were
+augmented by concatenating the average target and average non-target \ac{erp} as
+extra channels, projected to tangent space, and classified using
 $L_2$-regularized logistic regression (ERPCov-TS-LR).
@@
 selection (96 epochs) without overlap at each fold.
-A window ranging from 0ms to 600ms after stimulus onset was used for CBLE and WCBLE.
-With epochs ranging from -100ms to 700ms relative to stimulus onset, this
-allows for extracting latencies ranging from -100ms to +100ms.
+A window ranging from 0 ms to 600 ms after stimulus onset was used for \ac{cble} and \ac{wcble}.
+With epochs ranging from -100 ms to 700 ms relative to stimulus onset, this
+allows for extracting latencies ranging from -100 ms to +100 ms.
 
@@
 
-Figure~\ref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
-Receiver Operating Characteristic curve (ROC-AUC) for all pairs of tLDA, CBLE
-and WCBLE for single-trial classification to investigate the contributions of
-CBLE and WCBLE relative to their first-stage classifier tLDA.
+\Cref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
+Receiver Operating Characteristic curve (ROC-AUC) for all pairs of \ac{tlda}, \ac{cble},
+and \ac{wcble} for single-trial classification to investigate the contributions of
+\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
 For this evaluation, epochs were rejected when the peak-to-peak amplitude
-exceeded 800$\mu V$ and, for the CVSA-ERP dataset, if the user's gaze differed more than
+exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
 10 degrees of visual angle from the fixation crosshair.
@@
 		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
-    all decoders plotted n function of the number of test repetitions in
+    all decoders plotted as a function of the number of test repetitions in
     different VSA conditions. Significance was determined using one-sided
     (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
-    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves highest
+    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves the highest
     covert VSA target selection accuracy.}
@@
 and Hochberg's False Discovery Rate procedure (FDR) revealed that
-for the BNCI2014-009 dataset, WCBLE significantly outperformed tLDA
-($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and CBLE
-($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert VSA
-but was significantly outperformed by tLDA in overt VSA decoding
+for the BNCI2014-009 dataset, \ac{wcble} significantly outperformed \ac{tlda}
+($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and \ac{cble}
+($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert \ac{vsa},
+but was significantly outperformed by \ac{tlda} in overt \ac{vsa} decoding
 ($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
-For the CVSA-ERP dataset, WCBLE
-also achieved significantly better covert VSA performance than tLDA ($\Delta\mathrm{ROC-AUC}
-	= 0.023$, $p=0.041$) and CBLE ($\Delta\mathrm{ROC-AUC}
-	= 0.036$, $p=0.024$) .
-We found no significant difference in WCBLE performance over tLDA in the split
-VSA conditions in the CVSA-ERP dataset, but results show a clear trend of
-increase in WCBLE performance over tLDA and CBLE as $d$ increases
-CBLE failed to significantly outperform its first-stage classifier tLDA in all
-evaluated VSA conditions.
-Table~\ref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets
+For the CVSA-ERP dataset, \ac{wcble}
+also achieved significantly better covert \ac{vsa} performance than \ac{tlda} ($\Delta\mathrm{ROC-AUC}
+	= 0.023$, $p=0.041$) and \ac{cble} ($\Delta\mathrm{ROC-AUC}
+	= 0.036$, $p=0.024$).
+We found no significant difference in \ac{wcble} performance over \ac{tlda} in the split
+\ac{vsa} conditions in the CVSA-ERP dataset, but results show a clear trend of
+increased \ac{wcble} performance over \ac{tlda} and \ac{cble} as $d$ increases.
+\ac{cble} failed to significantly outperform its first-stage classifier \ac{tlda} in all
+evaluated \ac{vsa} conditions.
+\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets,
 and conditions.
@@
   \caption{Cross-validated single-trial classification area under the receiver operating characteristic
-curve for all evaluated models, visuospatial attention conditions and
+curve for all evaluated models, visuospatial attention conditions, and
   datasets.}
@@
 
-Figure~\ref{fig:block_evaluation} shows the cross-validated BCI selection
-accuracy on the BNCI2014-009 and the CVSA-ERP dataset for all investigated
+\Cref{fig:block_evaluation} shows the cross-validated BCI selection
+accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
 decoders.
@@
 operation intact.
-For all datasets and VSA conditions, CBLE scores approximately on par with tLDA.
-Yet, WCBLE yields an improved decoding accuracy for covert VSA in both
+For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
+Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
 datasets, which is greatest for smaller numbers of repetitions and
 decreases as the number of repetitions increases.
-This covert VSA accuracy increase over tLDA is significant in the BNCI2014-009 dataset
-for 1 and 3 repetitions and in CVSA-ERP for 1-5 and 10 repetitions.
+This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
+for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
 Furthermore, while we reported a relative decrease in single-trial ROC-AUC for
-WCBLE in overt VSA, this does not seem to result in a consistent decrease in target
+\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
 selection accuracy.
-No significant increase of WCBLE over other methods was found in split VSA.
-While Riemannian methods are significantly outperformed by tLDA, CBLE and WCBLE
-in the BNCI2014-009 dataset, they perform approximately on par with tLDA and
-WCBLE in CVSA-ERP.
+No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
+While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
+in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
+\ac{wcble} in CVSA-ERP.
 
-Overall, we observed a 5.10\%pt. accuracy increase with WCBLE over
-tLDA for covert VSA in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
+Overall, we observed a 5.10\%pt. accuracy increase with \ac{wcble} over
+\ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
 These results compare to the performance gain in \cite{Zisk2022}.
@@
 interface design (number of targets, \ac{isi}), subject population,
-EEG recording procedure and available training data.
+\ac{eeg} recording procedure, and available training data.
 
@@
 control cannot be assumed, we evaluate our proposed decoder in a transfer
-learning setting between VSA conditions.
-While performing covert attention requires gaze redirection for each target
-selection, performing covert or split VSA continuously still requires
-sustained gaze fixation, which might not be possible for some patients that
+learning setting between \ac{vsa} conditions.
+While performing overt attention requires gaze redirection for each target
+selection, performing covert or split \ac{vsa} continuously still requires
+sustained gaze fixation, which might not be possible for some patients who
 could benefit from such an application.
 Studying the transfer between conditions simulates what happens
-when the user performs different VSA conditions throughout the
+when the user performs different \ac{vsa} conditions throughout the
 experimental session.
 Furthermore, if our decoder performs well in transfer-learning settings,
-it must capture some information about the ERP responses that is independent of
-the VSA condition, and hence does not depend on gaze redirection to perform
+it must capture some information about the \ac{erp} responses that is independent of
+the \ac{vsa} condition, and hence does not depend on gaze redirection to perform
 these conditions.
-We introduce an additional setting of interest here, namely on a combination of VSA
+We introduce an additional setting of interest here, namely a combination of \ac{vsa}
 conditions, which represents those cases where patients cannot
-redirect their gaze and hence can be in any one of the VSA conditions depending on the
-target they attend.
+redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
+target they attend to.
 For BNCI2014-009, this is implemented as an equal mix of overt and
-covert VSA, for CVSA-ERP the combined condition represents an equal mix of overt,
-covert and split VSA, disregarding parameter $d$.
+covert \ac{vsa}; for CVSA-ERP, the combined condition represents an equal mix of overt,
+covert, and split \ac{vsa}, disregarding parameter $d$.
 
 Figure~\ref{fig:covert_cross_eval} and Figure~\ref{fig:aloise2012_cross_eval} show
-the pair-wise differences in area under the ROC curve ($\Delta$ROC-AUC)
+the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
 between the investigated decoders.
-In this evaluation, bad epochs were rejected as in Section~\ref{sec:block_accuracy}.
+In this evaluation, bad epochs were rejected as in \cref{sec:block_accuracy}.
 \begin{figure*}
@@
     (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
-    BNCI2014-009 datasets respectively.
+    BNCI2014-009 datasets, respectively.
     A decoder is each	time trained on a visuospatial attention (VSA) condition
@@
     WCBLE yields an improvement in most
-		non-overt VSA settings, indicating it is more invariant to eye gaze than
-    CBLE and tLDA.
-		\subref{fig:jitter} Jitter characterized as the inter-quartile range (IQR)
-		of target epochs for different VSA conditions. Overt VSA exhibits lower
+		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
+    CBLE and \ac{tlda}.
+		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
+		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
     jitter than other conditions.
@@
 \end{figure*}
-When comparing CBLE and tLDA, we do not observe large differences in any of the
+When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
 evaluated settings, similar to the within-subject conditions.
-On the contrary, when considering the comparisons between WCBLE and tLDA, we
-see that performance is on par or greater using WCBLE for most conditions,
-except for within-overt VSA decoding.
+On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
+see that performance is on par or greater using \ac{wcble} for most conditions,
+except for within-overt \ac{vsa} decoding.
 
@@
 
-Figure~\ref{fig:single_trial_roc_auc_diff} shows that WCBLE significantly improves
-covert VSA decoding.
-This is advantageous for the development of a class of ERP-BCI interfaces for
+\cref{fig:single_trial_roc_auc_diff} shows that \ac{wcble} significantly improves
+covert \ac{vsa} decoding.
+This is advantageous for the development of a class of \ac{erp}-\ac{bci} interfaces for
 patients who prefer to rest their gaze on a fixation cross on the screen,
 avoiding the effort of redirecting their eye gaze for every selection.
-Furthermore, the performance gain over the first-stage classifier in the split
-VSA ($d=3$) and between-VSA condition transfer settings are promising for
-patients with even less eye motor control that can experience involuntary eye
-movements or fixation fatigue hence cannot keep their gaze fixed throughout an
-entire BCI operation session.
-WCBLE would allow them to operate a BCI comfortably while directing
+Furthermore, the performance gain over the first-stage classifier in split
+\ac{vsa} ($d=3$) and between-\ac{vsa} condition transfer settings are promising for
+patients with even less eye motor control who may experience involuntary eye
+movements or fixation fatigue and hence cannot keep their gaze fixed throughout an
+entire \ac{bci} operation session.
+\ac{wcble} would allow them to operate a \ac{bci} comfortably while directing
 their gaze to whichever portion of the screen they prefer, even when there is
@@
 course of operation.
-Although WCBLE did not significantly improve overt VSA single-trial decoding,
-Figure~\ref{fig:block_evaluation} shows that this does not negatively impact target
+Although \ac{wcble} did not significantly improve overt \ac{vsa} single-trial decoding,
+\cref{fig:block_evaluation} shows that this does not negatively impact target
 selection accuracy.
-While target selection accuracy also did not improve for split VSA, the
+While target selection accuracy also did not improve for split \ac{vsa}, the
 increase in single-trial performance in split ($d=3$) shows that an iterative
-alignment procedure has the potential to improve over CBLE and its first-stage
-classifier for this case as well.
+alignment procedure has the potential to improve over \ac{cble} and its first-stage
+classifier in this case as well.
 
-%, thereby increasing estimated jitter.
 We believe the relative increase in performance of our proposed decoder in covert and
-split VSA, and the lack thereof in overt VSA, could stem from the following:
+split \ac{vsa}, and the lack thereof in overt \ac{vsa}, could stem from the following:
 \begin{enumerate*}[label=(\arabic*)]
-	\item Covert and split VSA exhibit higher P3 jitter than overt VSA.
-        In covert and split VSA, participants have to execute a dual task by
+	\item Covert and split \ac{vsa} exhibit higher P3 jitter than overt \ac{vsa}.
+        In covert and split \ac{vsa}, participants have to execute a dual task by
         dissociating their visuospatial attention and gaze fixation.
-        Evidence shows that ERP latency variability is higher when attention is
+        Evidence shows that \ac{erp} latency variability is higher when attention is
         divided~\cite{Polich2007,Arico2014}.
-        \cite{Arico2014} also partly attribute higher latency jitter to the covert VSA
-        task performed in the BNCI2014-009 dataset since the GeoSpell interface
+        \textcite{Arico2014} also partly attribute higher latency jitter to the covert \ac{vsa}
+        task performed in the BNCI2014-009 dataset, since the GeoSpell interface
         requires both spatial and feature attention.
-  \item In overt VSA, the first-stage classifier can rely mostly on the modulation
-        of early visually evoked potentials (VEPs) like N1 than on the
+  \item In overt \ac{vsa}, the first-stage classifier can rely mostly on the modulation
+        of early visually evoked potentials (VEPs) like N1 rather than on the
         P3~\cite{Treder2010}.
-        These VEPs are closely related to visual processing hence exhibit
-        lower jitter, contrary to P3 which is more prone to the effects of
+        These VEPs are closely related to visual processing, hence exhibit
+        lower jitter, contrary to P3, which is more prone to the effects of
         attention and workload~\cite{Hu2010}, reducing the contribution of
@@
   \item This property can also result in the estimation of VEP latencies
-        instead of the P3 latency and WCBLE would in this case fail to
-        increase the P3 SNR, which still could be somewhat jittered in overt
-        VSA.
-  \item Aligning to the P3 will lower the SNR of early VEPs while aligning to
-    VEPs will lower P3 SNR, since they are not time-locked to each other.
-	\item Covert and split VSA ERPs may exhibit lower SNR than overt
-        VSA due to lower P3 amplitudes or even due to the presence of higher P3
+        instead of the P3 latency, and \ac{wcble} would in this case fail to
+        increase the P3 \ac{snr}, which still could be somewhat jittered in overt
+        \ac{vsa}.
+  \item Aligning to the P3 will lower the \ac{snr} of early VEPs, while aligning to
+        VEPs will lower P3 \ac{snr}, since they are not time-locked to each other.
+	\item Covert and split \ac{vsa} \acp{erp} may exhibit lower \ac{snr} than overt
+        \ac{vsa} due to lower P3 amplitudes or even due to the presence of higher P3
         jitter itself.
-        Higher SNR in overt VSA results in higher decoding performance of
+        Higher \ac{snr} in overt \ac{vsa} results in higher decoding performance of
         state-of-the-art classifiers, leaving less room for relative improvement
@@
 \end{enumerate*}
-%When the time window includes multiple ERP components, it is hard to make
-%strong claims on component latency distributions using latencies estimated by
-%CBLE or WCBLE.
-%What is actually measured is the latency of the most discriminatory time shift
-%of an epoch.
 
-Although it is not immediately clear if WCBLE actively corrects for higher P3 jitter
-present in covert and split VSA compared to overt VSA,
-we justify our approach similar to \cite{Hardiansyah2020} by
+Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
+present in covert and split \ac{vsa} compared to overt \ac{vsa},
+we justify our approach similarly to \cite{Hardiansyah2020} by
 observing that the increased discrimination performance of a machine-learning
 model accounting for jitter forgoes the need for characterizing the underlying
-physiological processes while still objectively quantifying the presence of
+physiological processes, while still objectively quantifying the presence of
 jitter between data classes.
-Furthermore, some evidence points towards higher P3 jitter as the main
+Furthermore, some evidence points toward higher P3 jitter as the main
 contributing factor.
-%Here, the subject using the GeoSpell interface must discriminate between target and
-%non-target trials partly based on the stimulus characteristics
-%(i.e. the selection of characters in their spelling application), compared to
-%discrimination based on spatial location such as is the case in the CVSA-ERP
-While Figure~\ref{fig:erps} shows no visible smearing effect in
-the shape of the ERPs, a more quantitative analysis on latencies in
-Figure~\ref{fig:jitter} indicates the opposite.
-To obtain comparable results across VSA conditions, WCBLE was
-evaluated per session and trained on all combined VSA conditions as in
-Section~\ref{sec:cross_results}.
+While \autoref{fig:erps} shows no visible smearing effect in
+the shape of the \acp{erp}, a more quantitative analysis on latencies in
+\autoref{fig:jitter} indicates the opposite.
+To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
+evaluated per session and trained on all combined \ac{vsa} conditions as in
+\cref{sec:cross_results}.
 Since all conditions have the P3 in common, the estimated latencies can be
 interpreted as P3 latencies.
-Two-sided Wilcoxon rank-sum tests on jitter, expressed as the inter-quartile
+Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
 range (IQR) of the estimated latencies of target trials, revealed that overt
-VSA exhibited significantly lower jitter than all other conditions for both datasets.
+\ac{vsa} exhibited significantly lower jitter than all other conditions for both datasets.
 No other significant differences were found.
@@
 correction.
-Additionally \cite{Arico2014} prove that P3 jitter does play a
-non-negligible role in covert VSA by comparing performance between
-overt and covert VSA, while including or excluding early VEPs.
+Additionally, \textcite{Arico2014} prove that P3 jitter does play a
+non-negligible role in covert \ac{vsa} by comparing performance between
+overt and covert \ac{vsa}, while including or excluding early VEPs.
 This analysis showed that the absence of the N1 and other early VEPs is not the
-only factor hampering covert VSA decoding performance.
-However, the large increase in performance for covert VSA in \cite{Arico2014} could also
+only factor hampering covert \ac{vsa} decoding performance.
+However, the large increase in performance for covert \ac{vsa} in \cite{Arico2014} could also
 be explained by overfitting on artifacts amplified by aligning, since their
@@
 
-%If most of the jittered discriminatory information stems from the P3 component,
-%WCBLE as a decoder is less sensitive to VEPs and only enhances the P3
-%component.
-%This property can be exploited in gaze-independent decoding to determine
-%visuospatial attention separately from the gaze direction.
+We also argue that \ac{wcble} does not only enhance the P3 signal, but effectively reduces the
+negative impact of noise that would otherwise mask the \ac{erp} response, thereby increasing the
+discriminative ability of the decoder.
+This process likely results in more reliable feature extraction, especially in conditions
+with high jitter.
 
-%In fact, it is not obvious how the alignment employed by WCBLE
-%should result in better performance for overt attention.
-%At each iteration, aligning to the P3 latency reduces the contribution from
-%early visual potentials such as the N1 component not, which are more time-locked to
-%the stimulus presentation than to the P3.
-%These early visual potentials, however, provide a large part of the
-%discriminatory information in overt VSA, while this is less
-%the case in covert attentionConversely, since the N1 component is more closely related to visual processing,
-%and therefore exhibits a lower jitter, aligning to the N1 component might not augment the
-%Another explanation for the lack of overt VSA improvement could be the choice
-%of tLDA as first-stage classifier and comparison method.
-%As a classifier, tLDA can outperform Riemannian Geometry
-%methods~\cite{Sosulski2022}, which is
-%most likely the cause of improvements of WCBLE over \cite{Aydarkhanov2020}.
-%It might be harder to improve upon tLDA in overt VSA due to its high
-%performance.
-
-%Finally, we argue that the increased performance of WCBLE is partly due to the
-%fact that
-%tLDA is a suitable first-stage classifier.
-%Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
-%problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
-%Secondly, this method has a synergy with CBLE since both make the same
-%assumption about the short-time stationary of the EEG background noise
-%within an epoch.
-%CBLE does not retrain the first-stage classifier for each time shift, but
-%rather trains it once within the given window.
-%After training, the classifier parameters represent some information about the
-%expected ERP waveform and background noise.
-%By applying the trained classifier to different time shifts, it assumes this
-%ERP waveform can be shifted in time, but since the classifier's information about
-%the background noise was only obtained from the initial window, CBLE assumes
-%its properties do not vary throughout the epoch.
-%The block-Toeplitz covariance structure of tLDA also assumes that the
-%background noise represented by this covariance after subtracting the class
-%averages is stationary within the epoch~\cite{Sosulski2022}.
-
-We found that CBLE did not improve gaze-independent decoding performance
+We found that \ac{cble} did not improve gaze-independent decoding performance
 significantly and also did not increase performance over its first-stage
-classifier in overt VSA contrary to what was reported
-by \cite{Mowla2017}.
-While they report that CBLE is relatively independent to the
+classifier in overt \ac{vsa}, contrary to what was reported by \cite{Mowla2017}.
+While they report that \ac{cble} is relatively independent of the
 first-stage classifiers evaluated in their work, it is evident here that
-applying any given classifier in the CBLE scheme does not necessarily increase
+applying any given classifier in the \ac{cble} scheme does not necessarily increase
 its performance.
-In our case, this could be due to characteristics of the tested dataset, e.g.
-the presence of jitter, or the generally higher performance of tLDA as compared to the
+In our case, this could be due to characteristics of the tested dataset, e.g.,
+the presence of jitter or the generally higher performance of \ac{tlda} compared to the
 first-stage classifiers tested by \cite{Mowla2017}, leaving less
@@
 
-\cite{Thompson2012} already attempted applying CBLE in an iterative scheme
+\cite{Thompson2012} already attempted applying \ac{cble} in an iterative scheme
 but did not report any results due to convergence issues.
-%Just like the Woody scheme, this algorithm can get stuck in a local optimum or,
-%in unfavorable SNR conditions, even amplify noise.
 We mitigated this by combining the robust latency estimation presented in
-~Section~\ref{sec:robust-latency} with the alignment of both target and non-target
+\cref{sec:robust-latency} with the alignment of both target and non-target
 epochs.
 Aligning only the target epochs containing the jittered P3 component is prone to
-overfitting by aligning non-discriminate noise that is present in both
-classes, such as environment noise, oscillatory background rhythms or
+overfitting by aligning non-discriminative noise that is present in both
+classes, such as environment noise, oscillatory background rhythms, or
 non-modulated VEPs.
-If SNR is low, residual noise varying slightly between classes could dominate
+If \ac{snr} is low, residual noise varying slightly between classes could dominate
 the expected response of the first-stage classifier and subsequently dominate
-WCBLE from the start, preventing convergence to a meaningful solution.
+\ac{wcble} from the start, preventing convergence to a meaningful solution.
 Our procedure circumvents this problem by aligning both classes to the time
@@
 This aids the method in converging to a more robust classifier by iteratively
-increasing SNR for both classes and class separation over time.
+increasing \ac{snr} for both classes and class separation over time.
 
 \cite{Zisk2022} addressed P3 jitter in ALS patients by augmenting the training data once
-with time-shifted copies based on CBLE-estimated jitter.
+with time-shifted copies based on \ac{cble}-estimated jitter.
 While we aim to train the first-stage classifier without the effects of jitter
@@
 
-\textcite{Hardiansyah2020} decoded covert \textcite{vsa} more effectively by
-contributing single-trial latency and amplitude features to decode.
-Contrary to our approach, they did not then correct these amplitude
-features for the jitter in their latencies by e.g. aligning trials to achieve
+\textcite{Hardiansyah2020} decoded covert \ac{vsa} more effectively by
+contributing single-trial latency and amplitude features for decoding.
+Contrary to our approach, they did not correct these amplitude
+features for the jitter in their latencies by, e.g., aligning trials to achieve
 better separability.
-Hence, their approach would not in principle render the classifier more robust
+Hence, their approach would not, in principle, render the classifier more robust
 to jitter.
-Furthermore we incorporated estimated latency features in both CBLE and WCBLE,
-yet only WCBLE improved covert VSA performance.
+Furthermore, we incorporated estimated latency features in both \ac{cble} and \ac{wcble},
+yet only \ac{wcble} improved covert \ac{vsa} performance.
 This shows that the incorporation of latency features is not the only driver of
-covert VSA decoding performance increase.
+covert \ac{vsa} decoding performance increase.
 
@@
 tackle in the future.
-Firstly, multiple ERP components can be time-locked to different neural
-processes, each with their own jitter, hampering the performance of
+Firstly, multiple \ac{erp} components can be time-locked to different neural
+processes, each with its own jitter, hampering the performance of
 single-trial latency estimation and their interpretability.
 Adaptations could be made to incorporate prior time windows or probability
-distributions on the latency of specific ERP components or to simultaneously
-estimate set of multiple component (clusters) latencies per ERP such as in Residue Iteration
+distributions on the latency of specific \ac{erp} components or to simultaneously
+estimate a set of multiple component (clusters) latencies per \ac{erp}, such as in Residue Iteration
 Decomposition~\cite{Ouyang2017}.
@@
 spatiotemporal filtering can be combined with methods that allow for a more
-flexible non-stationarity of the ERP like Dynamic Time Warping (DTW) or other
-techniques lent from time series classification, or methods that explicitly model multiple time
-displacements present in one ERP.
+flexible non-stationarity of the \ac{erp}, like Dynamic Time Warping (DTW) or other
+techniques borrowed from time series classification, or methods that explicitly model multiple time
+displacements present in one \ac{erp}.
+
 Secondly, performance might be improved by venturing beyond the classical
 target/non-target binary classification problem.
-Due to e.g. the perifoveal stimulus cruciform model~\cite{Vanegas2013}, covert and split VSA
+Due to, for example, the perifoveal stimulus cruciform model~\cite{Vanegas2013}, covert and split \ac{vsa}
 responses might differ based on the relative position in the field of view of
-their related stimulus which could be exploited in a multi-class classification
+their related stimulus, which could be exploited in a multi-class classification
 problem.
-Similarly, explicitly taking into account the characteristic of the distractor ERP
+Similarly, explicitly taking into account the characteristics of the distractor \ac{erp}
 response might have a beneficial effect.
-Thirdly, results were obtained in an off-line and within-session evaluation,
-which does not reflect true BCI operation.
-Using multiple sessions with on-line feedback, the user could optimize their
+Thirdly, results were obtained in an offline and within-session evaluation,
+which does not reflect true \ac{bci} operation.
+Using multiple sessions with online feedback, the user could optimize their
 performance over time by controlling attention or gaze.
 Finally, since this work was conducted with patient applications in mind, we
-should highlight that gaze of participants in the conducted experiments was
+should highlight that the gaze of participants in the conducted experiments was
 cued and fixed, which is per definition impossible for the patients we consider.
 We are currently conducting a patient study to further investigate whether the
-studied VSA conditions are appropriate and to what extent they occur in
-natural patient BCI operation.
+studied \ac{vsa} conditions are appropriate and to what extent they occur in
+natural patient \ac{bci} operation.
 
 \section{Conclusion}
-Our aim was to improve gaze-independent BCI performance for spatially
-organized visual event-related potential (ERP) paradigms by using a suited
+
+Our aim was to improve gaze-independent \ac{bci} performance for spatially
+organized visual event-related potential (\ac{erp}) paradigms by using a suitable
 decoder.
-Earlier results on BCI performance in covert visuospatial attention
-(VSA) performance prediction have shown that accounting for single-trial
+Earlier results on \ac{bci} performance in covert visuospatial attention
+(\ac{vsa}) performance prediction have shown that accounting for single-trial
 latency jitter could improve gaze-independent decoding performance.
-We applied Classifier-based Latency Estimation (CBLE) as a decoder robust to latency
-jitter but found no increase in gaze-independent decoding performance.
-To remedy this, we improved CBLE by adapting it into CBLE with Woody
-iterations (WCBLE), an iterative scheme using probabilistic latency estimation.
-Results for WCBLE within and across VSA condition decoding show that
-gaze-independent BCI performance can be improved at the decoding stage.
+We applied Classifier-based Latency Estimation (\ac{cble}) as a decoder robust to latency
+jitter, but found no increase in gaze-independent decoding performance.
+To remedy this, we improved \ac{cble} by adapting it into \ac{cble} with Woody
+iterations (\ac{wcble}), an iterative scheme using probabilistic latency estimation.
+Results for \ac{wcble} within and across \ac{vsa} condition decoding show that
+gaze-independent \ac{bci} performance can be improved at the decoding stage.
 Overt decoding performance was not improved, but our proposed method can
-provide added value for patients who are unable to operate a visual BCI in overt
+provide added value for patients who are unable to operate a visual \ac{bci} in overt
 attention mode.
 Later studies should confirm whether our findings hold in patient
-populations suffering from a variety of eye-motor impairments, and develop a
-solution that is capable of properly handling multiple non-time locked ERP
+populations suffering from a variety of eye-motor impairments and develop a
+solution that is capable of properly handling multiple non-time-locked \ac{erp}
 components.
+
+\clearpage
+\begin{subappendices}
+\section{\Acs{cble} algorithm}
+\label{sec:wcble/app/cble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
+  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\State $\theta \gets
+			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
+      \algorithmiccomment{Train stage 1}
+		\For{$n=1\ldots N$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\theta,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+
+  \caption{\Acs{cble}}
+	\label{alg:cble}
+\end{algorithm}
+
+\clearpage
+\section{\Acs{wcble} algorithm}
+\label{sec:wcble/app/wcble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
+		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
+	\begin{algorithmic}[1]
+		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
+    N$
+      \algorithmiccomment{Train stage 1}
+		\Repeat
+		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
+		\For{$n=1\ldots N$}
+    \State  $s_n \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
+		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
+		\EndFor
+    \Until{convergence or maximum iterations reached}
+    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets
+      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+
+	\end{algorithmic}
+	\textbf{Output:} $\theta^*,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets
+      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+  \caption{\Acs{wcble}}
+	\label{alg:wcble}
+\end{algorithm}
+\end{subappendices}

### 2024-10-08 chapters/cv.tex
Comments Marc
Comments François, smearing figure
Compiling
Update bttda formulation
cover page
Proofread

--- 2024-09-16 chapters/cv.tex
@@
 
-\newenvironment{cvlist}{
+\newenvironment{cvlist}{%
 	\begin{description}[font=\normalfont]%\setlength\itemsep{0em}
-}{
+}{%
   \end{description}
@@
 % Curriculum vitae =============================================================
-\chapter{Curriculum vit\ae}
+\chapter*{Curriculum vit\ae}
+Arne Van Den Kerchove \\
+\texttt{arne@vandenkerchove.com}
 
 \subsection*{About me}
-I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
+I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
 and the University of Lille Research Center for Informatics, Signal Processing,
@@
 \begin{cvlist}
-	\cvitem{Ph.D.\ in Biomedical Sciences}{KU Leuven, 2024}
-	\cvitem{Ph.D.\ in Control Science and Signal Processing}{University of
+	\cvitem{PhD\ in Biomedical Sciences}{KU Leuven, 2024}
+	\cvitem{PhD\ in Control Science and Signal Processing}{University of
 		Lille, 2024}
@@
 \subsection*{Publications}
+\begin{refsection}
+\setmaxbibnames{9999}
 \nocite{%
@@
 	VanDenKerchove2024,
+	VanDenKerchove2024a,
 }
 \printbibliography[heading=none, resetnumbers=true]
+\end{refsection}
 
@@
 	}
-	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
+	\cvitem{Global PhD Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
 \end{cvlist}

### 2024-10-08 chapters/funding.tex
Proofread

@@
 \chapter*{Funding \& acknowledgments}
-\todo{my grant}\todo{marc's projects}
 
@@
 
-We also thank,  Arno Libert,
+We also thank  Arno Libert,
 Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
@@
 the work presented in~\autoref{sec:bttda}.
-
-\todo{people at TRAINM, MAS and Kristl Claeys}
-\todo{juliette meunier}

### 2024-10-08 chapters/gaze_independence.tex
Proofread

@@
 to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
-Table~\ref{tab:incidence} reports the relatively high frequencies of
+\Cref{tab:incidence} reports the relatively high frequencies of
 eye motor impairment, which can range from minor (nystagmus,
@@
 \begin{table}
+  \centering
+  \makebox[\textwidth][c]{%
   \input{chapters/gaze_independence_incidence.tex}
+  }
   \caption[Incidence of eye motor impairment in selected \acs{bci} target
@@
 If they have eye motor impairment that affects their use of visual \ac{bci} or
-eye-tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).
+eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).
 
@@
     such as the ability to communicate through a binary switching system.
-    \Acp{bci} are especially relevant for the last two groups.
+    \Acp{bci} are especially relevant for the last two groups, as they are not
+    able to use eye tracking solutions.
   }
@@
 \textcite{Wolpaw2006}\fnwolpawcrit.
-\todo{[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
-visual function of patients with locked-in syndrome: report
-of 13 cases. J Fr Ophtalmol. 2016;39:437–440.}
 Hence, some degree of severe or complete eye motor impairment is usually
@@
 visual oddball \ac{bci}.
-Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
+\Cref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
 good candidate for a performant gaze-independent \ac{bci}.
@@
 To address this, we improved upon an in-house developed, state-of-the-art \ac{erp}
-decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
+decoder, the \ac{stbf}~\cite{Wittevrongel2016}, by reformulating
 this classifier as a linear discrimination problem and
 imposing regularizing constraints by structuring the noise covariance matrix
-(STBF-struct).
+(\acs{stbf-struct}).
 Furthermore, these regularizing constraints impose temporal stationarity on
@@
 This approach has been published by~\textcite{VanDenKerchove2022} and will be
-described in \autoref{sec:stbf-struct}.
+described in \cref{sec:stbf-struct}.
 
@@
 gaze-independence.
-The proposed \ac{bttda} method will be discussed in \autoref{sec:bttda}.
+The proposed \ac{bttda} method will be discussed in \cref{sec:bttda}.
 
@@
 This approach has been published by~\textcite{VanDenKerchove2024} and is presented
-and evaluated on simulated data in \autoref{sec:wcble}.
+and evaluated on simulated data in \cref{sec:wcble}.
 We have also collected datasets to evaluate this approach on real \ac{erp} data
@@
 The interface can be operated in different \ac{vsa} conditions, as illustrated in
-\autoref{fig:gaze/vsa}.
+\cref{fig:gaze/vsa}.
 
@@
 published by \textcite{VanDenKerchove2024} and will be presented in
-\autoref{sec:covert-align}.
+\cref{sec:covert-align}.
 Performance is also evaluated on a publicly available dataset.
@@
 calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
-The results of this study are presented in \autoref{sec:patients}.
+The results of this study are presented in \cref{sec:patients}.

### 2024-10-08 chapters/patients.tex
Proofread

@@
 \Acp{bci} for communication assistive technology~\cite{Millan2010}
-target individuals with \ac{sspi}~\cite{Peters2022}.
+target individuals with \acf{sspi}~\cite{Peters2022}.
 Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
@@
 stimulation.
-This makes them well-suited for real-time tasks communication tasks.
+This makes them well-suited for real-time communication tasks.
 
-Yet, there is a large comorbidity between \ac{sspi}, and eye motor impairment~\cite{FriedOken2020}.
+Yet, there is a large comorbidity between \ac{sspi} and eye motor impairment~\cite{FriedOken2020}.
 Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
@@
 Eye motor impairments are presumed to reduce performance in operating visual
-oddball \ac{bci}s (see \autoref{sec:gaze-independence} for an overview), since users
+oddball \ac{bci}s (see \cref{sec:gaze-independence} for an overview), since users
 cannot comfortably redirect their gaze at the desired target,
@@
 \acp{bci}~\cite{Riccio2012}.
-These interfaces either avoid visual stimulation, or exploit some form of
+These interfaces either avoid visual stimulation or exploit some form of
 covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.
@@
 One could argue that a \ac{bci} that is only verified to work when central
-fixation is maintained, could also be considered gaze-dependent.
+fixation is maintained could also be considered gaze-dependent.
 This does not account for the residual eye motor capabilities of most people
@@
 participants with \ac{lis} yet only exceeded chance level accuracy in two.
-More recently, \textcite{Peters2020} performed an trial with an
-two participants with late-stage \ac{als} and visual impairment.
+More recently, \textcite{Peters2020} performed a trial with two
+participants with late-stage \ac{als} and visual impairment.
 Their \ac{ssvep} interface was not optimized for gaze-independence, but the
 system showed high accuracy, outperforming an eye tracking alternative.
-It would be of interest to verify if such results can be verified with other
+It would be of interest to verify if such results can be replicated with other
 patient groups and visual oddball \acp{bci}.
@@
 \textcite{Orhan2012} and \textcite{Oken2014} tested the \ac{rsvp} speller with
-individuals with \ac{lis}
+individuals with \ac{lis}.
 
@@
 
-Our previous study, presented in chapter \autoref{sec:covert-align} also used the
+Our previous study, presented in \cref{sec:covert-align} also used the
 visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
 This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
-not fully rely on central gaze fixation and evaluates settings that are not
+not fully rely on central gaze fixation and evaluated settings that are not
 strictly dependent on this.
-We showed gaze-independent performance can be improved
-improved in healthy subjects using a suited decoding strategy.
-that accounts for latency jitter in covert \ac{vsa} responses improves
-Yet, there is a strong need for verification of these result in an applied
+We showed gaze-independent performance can be improved in healthy subjects
+using a suited decoding strategy.
+That accounts for latency jitter in covert \ac{vsa} responses.
+Yet, there is a strong need for verification of these results in an applied
 setting with people with \ac{sspi}.
 
-%\todo{discuss \cite{FriedOken2020}}
-
 Eventually, one of the end goals of this research line is to develop
-gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
+gaze-independent \ac{bci} for people that are fully locked-in and have no option left than to use a \ac{bci}.
 However, this group is very small and it is often a challenge to recruit them
@@
 solutions tailored to their specific situation.
-Therefore, we aim to apply the concepts from earlier work an literature to
-people with \ac{sspi} and various degrees motor impairment in a visuall oddball
+Therefore, we aim to apply the concepts from earlier work and literature to
+people with \ac{sspi} and various degrees of motor impairment in a visual oddball
 \ac{bci}.
@@
 (Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
-(Lille, France) and a specialized care home (France).
+(Lille, France), and a specialized care home (France).
 Experiments were performed under the supervision of their treating physician.
 Participants were recruited based on the following criteria.
-To qualify for inclusion, participants must
+To qualify for inclusion, participants must:
 \begin{enumerate}
@@
 	years,
-  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
+  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
     presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
@@
 \end{enumerate}
-Participants were excluded if they
+Participants were excluded if they:
 \begin{enumerate}
@@
     including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
-  \item have a severe loss in vision or hearing, that would significantly impair
+  \item have a severe loss in vision or hearing that would significantly impair
         participation in the experiment,\label{item:patients/exclusion/vision}
-  \item are currently using specific psychoactive medications or substances that could affect the out-
-        come.\label{item:patients/exclusion/cognitive}
-  \item be able to understand the experiment instructions and cooperate,
+  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
+  \item be unable to understand the experiment instructions and cooperate,
   \item have any other limitations preventing them from performing the given task.
@@
 stroke based
-on~\ref{item:patients/exclusion/medical}
+on~\ref{item:patients/exclusion/medical}.
 One further person recovering from a stroke was excluded due to technical
@@
 Ultimately, 7 participants were retained.
-Of these, one participant was diagnosed from bulbar-onset \ac{als}.
+Of these, one participant was diagnosed with bulbar-onset \ac{als}.
 \ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
@@
 This initially results in general weakness and loss of muscle tone, but
-eventually results in full body paralysis.
+eventually leads to full body paralysis.
 Although speech and especially eye movements are usually preserved until the
@@
 Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
-whose life span has been extended with life support with and assistive technology to
+whose life span has been extended with life support and assistive technology to
 ensure quality of life.
@@
 disease affecting the
-spinal cord, peripheral nervous system and cerebellum.
+spinal cord, peripheral nervous system, and cerebellum.
 \ac{fa} results in an impairing loss of muscle coordination.
 In progressed stages, the disease can present with nystagmus, saccadic
-intrusions and gaze dysmetria~\cite{Cook2017}.
+intrusions, and gaze dysmetria~\cite{Cook2017}.
 
@@
 unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
-and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
+and swallowing). Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
 and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
@@
 \end{table}
-\todo{include years since diagnosis}
+%\todo{include years since diagnosis}
 
@@
 These include visual acuity, visual fixation, eyelid function, ocular motility,
-binocular vision and field of vision.
+binocular vision, and field of vision.
 Additionally, participants were asked about eye tremors (nystagmus or other) and
@@
 
-Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
-included participants.
+Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
+the included participants.
 All participants reported some degree of fatigue or discomfort when fixating.
@@
 movements.
-Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
+Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
 Participant PC2 was only able to look up and down and had a deviation in the
@@
 Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
-Participant PC4 had one deviated eye with a corneal abcess affecting the motility
+Participant PC4 had one deviated eye with a corneal abscess affecting the motility
 and vision in the right eye, and suffered reduced motility in the left.
@@
 
-Finally, we also record gaze position throughout the experimental session, to
+Finally, we also recorded gaze position throughout the experimental session to
 register the participant's gaze relative to the stimulated \ac{bci} targets.
@@
 field of view of the user.
-We refer to section~\ref{sec:covert-align/stimulation} for implementation details.
+We refer to \cref{sec:covert-align/stimulation} for implementation details.
 
-3 different \ac{vsa} settings were explored
-In the overt \ac{vsa} setting, the participant was instructed to fixate the cued target, or
+Three different \ac{vsa} settings were explored.
+In the overt \ac{vsa} setting, the participant was instructed to fixate on the cued target or
 try to the maximum extent of their visual skill, even if experiencing slight
 discomfort.
-In the covert \ac{vsa} setting, the participant was instructed to fixate the center of the
-screen, also to the extent of their ability.
+In the covert \ac{vsa} setting, the participant was instructed to fixate on the center of the
+screen, to the extent of their ability.
 An additional \emph{free \ac{vsa}} setting was introduced.
@@
 comfortable.
-This allows us to investigate the user's natural way of operating the \ac{bci}
+This allowed us to investigate the user's natural way of operating the \ac{bci}
 given their individual set of visual skills.
-If the participant was not fully paralyzed, they were instructed not to move the head.
+If the participant was not fully paralyzed, they were instructed not to move their head.
 The cued \emph{split attention} setting proposed
-by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
+by~\textcite{VanDenKerchove2024} was not studied here, as we were interested
 in natural \ac{vsa} operation settings for gaze-impaired individuals.
@@
 The experiment also started with a training block in each condition, where the
-participant was instructed with feedback on their performance to make sure they
+participant was instructed with feedback on their performance to ensure they
 understood and were able to perform the task.
@@
 During the recording session, participants were positioned in their wheelchair in front of a table.
-Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
-Inc., Taiwan) placed at a 60cm distance.
+Stimuli were presented on an Acer Predator Helios laptop with an 18" screen (Acer,
+Inc., Taiwan) placed at a 60 cm distance.
 A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
 recorded \ac{eeg}.
-Eye tracking throughout was performed using the Tobii X2-30 Compact (Tobii
+Eye tracking was performed throughout using the Tobii X2-30 Compact (Tobii
 Technology AB, Sweden) portable eye tracker placed at the bottom of the laptop screen.
 
-\Ac{eeg} was recorded at 1000Hz using the Neuroscan Neuvo  portable amplifier (Compumedics Neuroscan,
-Australia) connected to a second laptop for registration
+\Ac{eeg} was recorded at 1000 Hz using the Neuroscan Neuvo portable amplifier (Compumedics Neuroscan,
+Australia) connected to a second laptop for registration.
 The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
 according to the international 10-20 layout.
-Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$
+Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$.
 Additionally, the \ac{eog} was recorded.
 
-The \ac{eeg} was band-pass filtered between 0.5 and 16Hz.
-Bad channels were rejected using on the RANSAC algorithm~\cite{Fischler1981}
+The \ac{eeg} was band-pass filtered between 0.5 and 16 Hz.
+Bad channels were rejected using the RANSAC algorithm~\cite{Fischler1981}
 and visual inspection.
-Next, the \ac{eeg} was rereferenced to the average of mastoid electrodes TP9
+Next, the \ac{eeg} was re-referenced to the average of mastoid electrodes TP9
 and TP10, and \ac{ica} was performed to reject artifactual components based on
 correlation with the \ac{eog} or by visual inspection.
-Epochs are cut from -0.1 to 0.9s relative to stimulus onse, no baseline
+Epochs were cut from -0.1 to 0.9s relative to stimulus onset, and no baseline
 correction was performed.
@@
 given channel, the value of the other one was adopted.
-If both were missing, the gaze position remains unset at that time point and no
-interpolation is performed.
+If both were missing, the gaze position remained unset at that time point, and no
+interpolation was performed.
 
@@
 
-We evaluate the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
-and \ac{tlda}~\cite{Sosulski2022}.
-classifiers, and the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
-For \ac{wcble}, a region of interest from 0ms to 800ms relative to stimulus
-onset was used while the epoch was cropped to -100ms to 900ms, for the other
-decoders the epoch was cropped between 0ms and 800ms which resulted in maximal
+We evaluated the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
+and \ac{tlda}~\cite{Sosulski2022}
+classifiers, as well as the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
+For \ac{wcble}, a region of interest from 0 ms to 800 ms relative to stimulus
+onset was used while the epoch was cropped to -100 ms to 900 ms. For the other
+decoders, the epoch was cropped between 0 ms and 800 ms, which resulted in maximal
 performance.
@@
 stimulation blocks.
-
-refer to~\autoref{sec:covert-align/methods/decoders}.
 
@@
 First, we aimed to shed more light on the actual capabilities of individuals
-with \ac{sspgi} with respect to performing overt \ac{vsa} and central gaze
-fixation, as well as investigate how relevant these two settings are when the
+with \ac{sspgi} regarding performing overt \ac{vsa} and central gaze
+fixation, as well as to investigate how relevant these two settings are when the
 gaze is not cued.
-Figure~\ref{fig:patients/gaze} maps gaze position relative to the stimuli
+\Cref{fig:patients/gaze} maps gaze position relative to the stimuli
 across conditions.
-These results should be interpreted with care, since the eye tracker to some
+These results should be interpreted with care, as the eye tracker to some
 degree relies on functioning eye motility.
-The participant's position
-relative due the eye tracker might have shifted throughout the experimental
-session despite our best efforts, e.g.\ because they needed aspiration of their
+The participant's position relative to the eye tracker might have shifted
+throughout the experimental session despite our best efforts, e.g., because they needed aspiration of their
 tracheostomy.
@@
 cued overt and covert settings.
-When gaze was uncued he fixated the cued target.
+When gaze was uncued, he fixated on the cued target.
 This was also mostly the case for PB1, although eye tracking revealed that he
 chose not to perform central gaze fixation when cued in at least one of the
-stimulation blocks. We were unable to record his gaze near the bottom left
+stimulation blocks. We were unable to record his gaze near the bottom-left
 stimulus position, either due to eye tracker failure or because the participant
-was not comfortable fixating this position.
+was not comfortable fixating on this position.
 Eye tracker calibration did not succeed for subject PB4, but given
-transformations of gaze positions to the stimulus space, thy
+transformations of gaze positions to the stimulus space, they were assumed to be overtly performing the task.
 
 PB2 was able to perform overt \ac{vsa} and central fixation to some extent,
-yet eye tracking, but gaze position distributions shows a larger spread than
-for PA1 and PB1.
-The distribution in free \ac{vsa}, however, showed that she prefered to attend
+yet eye tracking shows a larger spread in gaze position compared to
+PA1 and PB1.
+In the free \ac{vsa} condition, however, she preferred to attend
 stimuli covertly when the gaze was uncued.
@@
 The overt and central gaze fixation settings were also not properly adapted to
-particpant PC4.
-In free \ac{vsa} eye tracker results show that his gaze was usually near the
+participant PC4.
+In the free \ac{vsa} condition, eye tracker results show that his gaze was usually near the
 bottom two targets, indicating some degree of covert or split \ac{vsa}.
@@
 occluded respectively by the prism glass and the eyelid.
-Both participants reported they could not fixate some of the
+Both participants reported they could not fixate on some of the
 stimuli.
@@
 
-Figure~\ref{fig:patients/decode} shows single-trial
-\ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.
+\Cref{fig:patients/decode} shows single-trial
+\ac{rocauc} in the evaluated \ac{vsa} settings for the different decoders.
 
@@
     Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
-    settign.
+    setting.
     Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
@@
 \end{figure}
-\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}
 
-In the overt \ac{vsa} setting, evaluated decoders performed about equal on average
+In the overt \ac{vsa} setting, the evaluated decoders performed similarly on average
 (\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
-In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorates
-but \ac{wcble} can significantly improve performance over its base classifier
+In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorated,
+but \ac{wcble} significantly improved performance over the base classifier
 \ac{tlda} in this condition
@@
 
-However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
-XDAWNCov+TS+LDA performance was slightly lower here (although not
+However, \ac{wcble} did not improve \ac{tlda} performance in the free \ac{vsa} setting, but
+XDAWNCov+TS+LDA performance was slightly lower here (though not
 significantly).
 (\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
-More interestingly, we notice that performances of the decoders in free
-\ac{vsa} are close to those in the overt \ac{vsa}.
+More interestingly, we noticed that performances of the decoders in free
+\ac{vsa} were close to those in the overt \ac{vsa}.
 A substantial decrease in performance from the overt setting to the free
-setting is present for
+setting was observed for
 subjects PC3 and PC4.
@@
 \label{sec:patients/outcomes/cross}
-As an alternative approach to picking the most suitable decoder, we chose
-\ac{tlda} as a given decoder and verified whether performance can be improved
-if \ac{bci} users with gaze impairment perform the calibration session relying
+
+As an alternative approach to selecting the most suitable decoder, we used
+\ac{tlda} as the base decoder and verified whether performance could be improved
+if \ac{bci} users with gaze impairment performed the calibration session relying
 maximally on their residual gaze control.
+
+\Cref{fig:patients/cross} shows that, on average, covert \ac{vsa} decoding
+improved when training with overt \ac{vsa}.
+This was especially true for participants PA1, PB2, and PC3.
+Note that, according to eye tracking data, participants PB1, PB4, and PC4 did not
+always perform cued central gaze fixation in the covert \ac{vsa} setting,
+which might have affected the results.
 
@@
     single-trial \ac{rocauc}.
-    For participant PA1, free \ac{vsa} performance could be improved when
-    calibrating with overt gaze fixation.
-    For participants PA1, PB2 and PC3, this was the case when operating in the
-    covert \ac{vsa} setting with central gaze fixation and calibrating with
-    overt gaze fixation.
+    For participant PA1, free \ac{vsa} performance improved when
+    calibrated with overt gaze fixation.
+    For participants PA1, PB2, and PC3, performance in the covert \ac{vsa} setting with central gaze fixation
+    improved when calibrating with overt gaze fixation.
   }
@@
 
-\autoref{fig:patients/cross} shows that, on average, covert~\ac{vsa} decoding
-can be improved when training in overt \ac{vsa}.
-This was especially the case for participants PA1, PB2 and PC3.
-Note that, according to eye tracking data, patients PB1, PB4 and PC4 did not
-always perform the cued central gaze fixation in the covert \ac{vsa} setting,
-which might affect the results.
-
 \section{Discussion}
 
-Due to the heterogenous nature of the participant's conditions, it is hard to
+Due to the heterogeneous nature of the participants' conditions, it is difficult to
 draw general conclusions.
@@
 
-True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
-Hence our analysis centers around the free \ac{vsa} condition.
-Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
+True gaze-independent visual \acp{bci} should not rely on gaze fixation.
+Hence, our analysis centers around the free \ac{vsa} condition.
+Eye tracking results presented in \cref{sec:patients/outcomes/gaze}
 confirm our assumption that voluntary covert \ac{vsa} can
 occur in individuals with \ac{sspgi}.
-We have also confirmed part of the results from~\textcite{VanDenKerchove2024}
-presented in~\autoref{sec:covert-align} that state that decoding of covert \ac{vsa} with central gaze
-fixation can be improved by accounting for latency jitter, and showed that this
-holds for individuals with \ac{sspgi}.
+We also confirmed part of the results from~\textcite{VanDenKerchove2024}
+presented in \cref{sec:covert-align}, which state that decoding of covert \ac{vsa} with central gaze
+fixation can be improved by accounting for latency jitter. We showed that this
+also holds for individuals with \ac{sspgi}.
 
-Contradicting our assumptions, however, we have shown that this does not
+Contrary to our assumptions, however, we have shown that this does not
 necessarily improve covert \ac{vsa} when gaze fixation is not cued.
-One possible explanation for this is that actively performing central gaze fixation
+One possible explanation is that actively performing central gaze fixation
 increases task load.
-This can in turn reduce overall performance, even though the participant might
-otherwise also have performed covert \ac{vsa}, but would not be occupied by
+This, in turn, can reduce overall performance, even though the participant might
+have otherwise performed covert \ac{vsa}, but would not be occupied with
 maintaining strict central gaze fixation.
-These extra task demand is not present in the free \ac{vsa} condition, so
-there is less performance to be gained here.
-Furthermore cued central gaze fixation combined with counting flashing stimuli
-in the visual periphery, is an explicit example of a dual task.
+This extra task demand is not present in the free \ac{vsa} condition, so
+there is less performance to be gained.
+Furthermore, cued central gaze fixation combined with counting flashing stimuli
+in the visual periphery is an explicit example of a dual task.
 Dual tasks have been shown to increase P3 latency
-jitter~\cite{Polich2007,Arico2014, VanDenKerchove2024}
+jitter~\cite{Polich2007,Arico2014, VanDenKerchove2024},
 which is what \ac{wcble} accounts for.
-Hence,increased P3 jitter might be more related to maintaining central gaze fixation
+Hence, increased P3 jitter might be more related to maintaining central gaze fixation
 than to the actual covert \ac{vsa} aspect.
 
-The seamingly stable performance across overt and free \ac{vsa} could be
+The seemingly stable performance across overt and free \ac{vsa} could be
 misinterpreted as an indication that the Hex-o-Spell \ac{bci} already works
@@
 covert \ac{vsa} setting
-with central gaze fixation, is food for thought about the applicability of
+with central gaze fixation is food for thought about the applicability of
 solutions developed with central fixation in mind.
 
-Individuals with all but the most severe gaze impairments will probably retain
+Individuals with all but the most severe gaze impairments will likely retain
 some degree of gaze direction in visual \ac{bci} operation, which can
@@
 overt \ac{vsa} performance, although eye tracking showed that he relied mostly
-on over \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
+on overt \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
 gaze was uncued.
-This is also supported by our results on cross condition calibration presented
-in~\autoref{sec:patients/outcomes/cross}, which show that leveraging residual
+This is also supported by our results on cross-condition calibration presented
+in \cref{sec:patients/outcomes/cross}, which show that leveraging residual
 eye motor control to fixate targets during the calibration phase can improve
@@
 which improves the discriminative power of a classifier trained on this data.
-Cueing this overt gaze fixation only during the calibration phase leaves user
+Cueing this overt gaze fixation only during the calibration phase leaves the user
 free to operate in the manner that is most comfortable for them in the
@@
 
-First and foremost, this study works with a limited sample sizekkjk which
+First and foremost, this study works with a limited sample size, which
 does not represent the full spectrum of individuals with \ac{sspi} and
-\ac{sspgi}, and their specific symptoms and skills?
+\ac{sspgi}, and their specific symptoms and skills.
 Individuals with \ac{fa} met the inclusion criteria, but they are usually not
-considered as one of the typical interest groups of \ac{bci} communication assistive
-technology, partly due to the rarity, and partly due to the disease
+considered one of the typical interest groups for \ac{bci} communication assistive
+technology, partly due to the rarity of the disease and partly due to its
 progression.
@@
 
-Another limiting factor are the difficulties in correctly interpreting eye
-tracker results in studies with individuals with gaze impairment.
-If eye tracking at all is possible, it is not guaranteed that the user is able
-to perform the calibration procedure successfully.
+Another limiting factor is the difficulty in correctly interpreting eye
+tracker results in studies with individuals with gaze impairments.
+If eye tracking is possible at all, it is not guaranteed that the user is able
+to successfully perform the calibration procedure.
 Further experiments should be carried out with a stationary eye tracker with
@@
 This is not practical when working with
-wheelchair-bound individuals who might have undergone a tracheostomy and might
+wheelchair-bound individuals who might have undergone a tracheostomy and may
 suffer from spasticity.
@@
 measured.
-Rather, it was assumed that participants operated most comfortably in the free
+Instead, it was assumed that participants operated most comfortably in the free
 \ac{vsa} condition.
-In order to properly contextualize performance results, they should be coupled
-with a measure of user comfort and percieved effort, like the NASA Task Load
+To properly contextualize performance results, they should be coupled
+with a measure of user comfort and perceived effort, like the NASA Task Load
 Index~\cite{Hart2006}.
-Performance might after all be traded off for user comfort.
-Eye motor disability can also have been assessed more
-objectively~\cite{FriedOken2020}, using e.g.\
+Performance might, after all, be traded off for user comfort.
+Eye motor disability could also have been assessed more
+objectively~\cite{FriedOken2020}, using, e.g.,
 the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
@@
 
-Finally, the parameters stimulation procedure
+Finally, the stimulation procedure parameters
 from~\textcite{VanDenKerchove2024} were adapted to make the counting task
 accessible to the \ac{bci} users with \ac{sspgi}.
-The number of repetitions or \ac{isi} were however not optimized to achieve
+However, the number of repetitions and \ac{isi} were not optimized to achieve
 maximal \ac{itr}.
@@
 comfort and the effect of visual skill.
-
-\section{Conclusion}
-
-To summarize, we present our three main findings.
-\begin{enumerate*}
-  \item We have confirmed that some individuals with \ac{sspi} and gaze
-    impairment find it more comfortable to operate a visual gaze-independent
-    \ac{bci} relying on covert \acf{vsa}.
-  \item We have shown that covert \acf{vsa} with central gaze fixation can be
-    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
-    but this does not hold true if central gaze fixation is not imposed.
-  \item We have shown that decoding of covert \ac{vsa}
-    can alternatively be improved by
-    using a calibration session with attempted overt \ac{vsa}.
-\end{enumerate*}
-
-In general, this study shows that approaches tailored to the individuals visual
-skills are necessary and that a thorough rethinking of the approach to
-designing gaze-independent visual \acp{bci} is an area of concern.
-Future work includes comparing these results with a sample of non-gaze impaired
-\ac{bci} users with \ac{sspi},
-and carrying out repeated on-line sessions with the current participants with built-in feedback to
-properly assess \ac{itr} and learning effects, and obtain realistic measures of percieved effort,
-comfort and performance.

### 2024-10-08 chapters/stbf_struct.tex
checking bttda
Proofread

@@
 
-  \emph{This chapter, except section~\ref{sec:stbf-struct-kronlda}, was
+  \emph{This chapter was
   published as~\textcite{VanDenKerchove2022}}.
@@
 
-	%Brain-computer interfaces (BCIs) establish a direct communication pathway
+	%Brain-computer interfaces (\acp{bci}) establish a direct communication pathway
 	%between the brain and an external device~\cite{Wolpaw2002}.
 	%Severely disabled patients with impaired or absent communication capabilities
-	%can benefit from BCIs to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
-	%BCIs can be implemented in multiple ways, using non-invasive recording
-	%techniques such as electroencephalography (EEG)~\cite{Abiri2019},
+	%can benefit from \acp{bci} to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
+	%\acp{bci} can be implemented in multiple ways, using non-invasive recording
+	%techniques such as electroencephalography (\ac{eeg})~\cite{Abiri2019},
 	%magnetoencephalography (MEG)~\cite{Mellinger2007},
@@
 	%arrays~\cite{Maynard1997} which require surgery to implant a recording device.
-	%While invasive BCIs yield the highest information transfer
-	%rate~\cite{Willett2021}, non-invasive BCIs are preferable for short-term use
+	%While invasive \acp{bci} yield the highest information transfer
+	%rate~\cite{Willett2021}, non-invasive \acp{bci} are preferable for short-term use
 	%since they are not susceptible to the risks that come with surgery.
-	%Of the non-invasive options, EEG is the most cost-effective and practical as it
+	%Of the non-invasive options, \ac{eeg} is the most cost-effective and practical as it
 	%is not limited to the same controlled settings as MEG and OPM MEG.
 
-	%Besides the recording method, BCIs differ in the communication paradigms used
+	%Besides the recording method, \acp{bci} differ in the communication paradigms used
 	%for communication~\cite{Abiri2019}.
-	%A popular class of BCI paradigms relies on the evocation
-	%of Event-Related Potentials (ERPs) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
+	%A popular class of \ac{bci} paradigms relies on the evocation
+	%of Event-Related Potentials (\acp{erp}) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
 	%calibration time before usage~\cite{Gao2014, Kapgate2015}.
-	%The study we report on focuses on the visual P300 oddball ERP in response to a
+	%The study we report on focuses on the visual P300 oddball \ac{erp} in response to a
 	%rare but attended visual stimulus.
-	%The decoder detects whether this ERP is present to determine which stimulus
+	%The decoder detects whether this \ac{erp} is present to determine which stimulus
 	%the user attended.
-	%The P300 paradigm has been used extensively in BCI development and is easy to
+	%The P300 paradigm has been used extensively in \ac{bci} development and is easy to
 	%set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.
 
-	There are multiple state-of-the-art P300 classification methods, like Support
-	Vector Machines (SVMs)~\cite{Tayeb2014}, deep
+  There are multiple state-of-the-art P300 classification methods, like
+  \acp{svm}~\cite{Tayeb2014}, deep
 	learning models~\cite{Vareka2020,Borra2020}, and Riemannian Geometry
@@
 	for lightweight models -- lightweight models lead to fast off-line analyses and can be transferred to consumer-grade hardware.
-	When moving towards plug-and-play solutions, BCI calibration sessions should be short and model training times low.
-	The spatiotemporal beamformer~\cite{VanVliet2015, Wittevrongel2016} belongs to
-	this class of ERP decoding models as it achieves state-of-the-art performance and is fast to train.
+	When moving towards plug-and-play solutions, \ac{bci} calibration sessions should be short and model training times low.
+  The \ac{stbf}~\cite{VanVliet2015, Wittevrongel2016} belongs to
+	this class of \ac{erp} decoding models as it achieves state-of-the-art performance and is fast to train.
 	Earlier work has shown that it is
 	possible to apply the spatiotemporal beamformer to multiple time-locked visual
-	BCI paradigms, including the P300 oddball paradigm,
-	Steady-State Visually Evoked Potentials (SSVEP), code-modulated Visually Evoked Potentials (cVEP)~\cite{Wittevrongel2017a} and
-	motion-onset Visually Evoked Potentials (mVEP)~\cite{Libert2021}.
+  \ac{bci} paradigms, including the P300 oddball paradigm, \ac{ssvep},
+  \ac{cvep}~\cite{Wittevrongel2017a}, \ac{mvep}~\cite{Libert2021}.
 
 	This work shows that the original spatiotemporal
-	beamformer~\cite{Wittevrongel2016} can fall short in performance when BCI
+	beamformer~\cite{Wittevrongel2016} can fall short in performance when \ac{bci}
 	calibration data are restricted.
@@
 	As a response to these issues, we introduce a regularization method that
-	exploits prior knowledge about the spatiotemporal nature of the EEG signal to
+	exploits prior knowledge about the spatiotemporal nature of the \ac{eeg} signal to
 	improve the accuracy for low data availability settings and speed up the
@@
 	Similar structured regularization approaches have been applied to other linear
-	ERP classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
+	\ac{erp} classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
 	significant increases in performance.
@@
 	\subsection{Notation}
-	We represent matrices with cursive capital letters, vectors with bold
-	lowercase letters, and scalars with cursive lowercase letters.
-	The epoched EEG data with $n$ epochs, $c$ channels, and $s$ samples are
-	represented in epoch format as $\{X_i\in\mathbb{R}^{c\times s}\}^n_{i=1}$ or flattened vector format by concatenating all channels for each epoch.
-	Flattening results in $\{\mathbf{x}_i\in\mathbb{R}^{cs}\}^n_{i=1}$ such that $\mathbf{x}_i = \text{vec}(X_i)$.
+	We represent matrices with bold capital letters, vectors with bold
+	lowercase letters, fixed scalars with uppercase cursive letters and variable
+  scalars with cursive lowercase letters.
+	The epoched \ac{eeg} data with $N$ epochs, $C$ channels, and $S$ samples are
+  represented in epoch format as $\{\mat{X}_i\in\mathbb{R}^{C\times S}\}^N_{i=1}$
+  or flattened vector format by concatenating all channels for each epoch.
+	Flattening results in $\{\mat{x}_i\in\mathbb{R}^{CS}\}^N_{i=1}$ such that
+  $\mat{x}_i = \text{vec}(\mat{X}_i)$.
 	The real covariance matrix of the epochs in vector format is
-	denoted by $C$, estimators thereof as $\hat{C}$.
+  denoted by $\mat{C}$, estimators thereof as $\hat{\mat{C}}$.
 
 	\subsection{Spatiotemporal beamforming}
-	LCMV-beamforming was initially introduced to EEG analysis as a filter for
+  \ac{lcmv}-beamforming was initially introduced to \ac{eeg} analysis as a filter for
 	source localization~\cite{VanVeen1997} to enhance the signal-to-noise ratio
 	(SNR).
-	Van Vliet et al.~\cite{VanVliet2015} first applied the spatiotemporal
-	LCMV-beamformer as a method for the analysis of ERPs.
+	\textcite{VanVliet2015} first applied the spatiotemporal
+	\ac{lcmv}-beamformer as a method for the analysis of \acp{erp}.
 	The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
@@
 
-	For the following analyses, we assume that all EEG channels are normalized with zero mean and unit variance without loss of generality.
-	Solving \autoref{eq:minimum_variance} under the linear constraint given by
-	\autoref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal LCMV-beamformer.
+	For the following analyses, we assume that all \ac{eeg} channels are normalized with zero mean and unit variance without loss of generality.
+	Solving \cref{eq:minimum_variance} under the linear constraint given by
+	\cref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal \ac{lcmv}-beamformer.
 	\begin{equation}
-		\underset{\mathbf{w}}{\arg\min}\mathbf{w}^\intercal C
-		\mathbf{w}^\intercal
+    \underset{\mat{w}}{\arg\min}\mat{w}^\intercal \mat{C}
+		\mat{w}^\intercal
 		\label{eq:minimum_variance}
@@
 	\begin{equation}
-		\mathbf{a}^\intercal\mathbf{w} = 1
+		\mat{a}^\intercal\mat{w} = 1
 		\label{eq:linear_constraint}
@@
 	the signal characterized by the constraint.
-	$\mathbf{a} = \text{vec}(A)$ is the data-driven activation pattern, a template
+  $\mat{a} = \text{vec}(\mat{A})$ is the data-driven activation pattern, a template
 	of the signal of interest maximizing the difference between two classes of
@@
 	The method of Lagrange multipliers then gives the closed-form solution to the minimization problem posed by
-	\autoref{eq:minimum_variance} and \autoref{eq:linear_constraint} as:
+	\cref{eq:minimum_variance} and \cref{eq:linear_constraint} as:
 
 	\begin{equation}
-		\mathbf{w} =
-		\frac{C^{-1}\mathbf{a}^\intercal}
-		{\mathbf{a}C^{-1}\mathbf{a}^\intercal}
+		\mat{w} =
+    \frac{\mat{C}^{-1}\mat{a}^\intercal}
+		{\mat{a}C^{-1}\mat{a}^\intercal}
 		\label{eq:closed_form}
@@
 	resulting in a scalar output per epoch.
-	The linear constraint in \autoref{eq:linear_constraint} ensures that the
+	The linear constraint in \cref{eq:linear_constraint} ensures that the
 	beamformer maps epochs containing a target response to a score close to one
@@
 	While the spatiotemporal beamformer, in theory, achieves optimal separation
-	between target and non-target classes, in analogy to linear discriminant analysis~\cite{TREDER2016279}, it does not always perform well on unseen data.
-	The main challenge is to find a good estimator for the inverse covariance matrix $C^{-1}$ since the real underlying covariance matrix generating the data is, in principle, unknown.
+	between target and non-target classes, in analogy to linear discriminant
+  analysis~\cite{Treder201}, it does not always perform well on unseen data.
+	The main challenge is to find a good estimator for the inverse covariance
+  matrix $\mat{C}^{-1}$ since the real underlying covariance matrix generating the data is, in principle, unknown.
 
@@
 	\begin{equation}
-		\hat{C}_\text{emp} =
-		\frac{1}{n-1}\sum^{n}_{i=1}\mathbf{x}_i\mathbf{x}_i^\intercal
+    \hat{\mat{C}}_\text{emp} =
+		\frac{1}{N-1}\sum^{N}_{i=1}\mat{x}_i\mat{x}_i^\intercal
 		\label{eq:empirical_covariance}
@@
 	\begin{equation}
-		\widehat{C^{-1}}_\text{emp} = \hat{C}_\text{emp}^+
+    \widehat{\mat{C}^{-1}}_\text{emp} = \hat{\mat{C}}_\text{emp}^+
 		\label{eq:empirical_inverse_covariance}
@@
 
-	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
-	\autoref{fig:kronlda-covs}a and \autoref{fig:kronlda-covs}b respectively show examples of the
+	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when
+  $\hat{\mat{C}}_\text{emp}$ is singular.
+	\cref{fig:kronlda-covs}a and \cref{fig:kronlda-covs}b respectively show examples of the
 	empirical estimators of the covariance and the inverse covariance matrices.
@@
 	\begin{equation}
-		\hat{C}_\alpha =
-		(1-\alpha) \hat{C}_\text{emp}
-		+ \alpha\frac{\text{Tr}(\hat{C}_\text{emp})}{cs}\mathbb{I}
+    \hat{\mat{C}}_\alpha =
+    (1-\alpha) \hat{\mat{C}}_\text{emp}
+    + \alpha\frac{\text{Tr}(\hat{\mat{C}}_\text{emp})}{CS}\mathbb{I}
 		\label{eq:shrinkage}
@@
 	\begin{equation}
-		\widehat{C^{-1}}_\alpha =
-		\hat{C}^+_\alpha
+    \widehat{\mat{C}^{-1}}_\alpha =
+    \hat{\mat{C}}^+_\alpha
 		\label{eq:shrinkage_inverse}
@@
 	with $0<\alpha<1$.
-	Analogous to L2 regularization of the beamforming problem,
+	Analogous to $L_2$ regularization of the beamforming problem,
 	shrinkage reduces the ratio between the smallest and largest eigenvalues
 	of the covariance matrix by strengthening the diagonal.
-	\autoref{fig:kronlda-covs}c and \autoref{fig:kronlda-covs}d respectively show examples of the
+	\cref{fig:kronlda-covs}c and \cref{fig:kronlda-covs}d respectively show examples of the
 	shrunk estimator of the covariance and the inverse covariance matrices.
 
-	Earlier work~\cite{Libert2021} applied shrinkage regularization to ERP
+	Earlier work~\cite{Libert2021} applied shrinkage regularization to \ac{erp}
 	decoding with the spatiotemporal beamformer and showed competitive performance
@@
 	Most notable among these are the Ledoit-Wolf procedure~\cite{Ledoit2004},
-	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010shrinkage}, and Oracle Approximating Shrinkage~\cite{Chen2010shrinkage}.
+	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010}, and Oracle Approximating Shrinkage~\cite{Chen2010}.
 	A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
-	cross-validation (LOOCV) scheme expressed by the data-driven closed-form
+	cross-validation (\ac{loocv}) scheme expressed by the data-driven closed-form
 	estimate:
@@
 		1-\frac{
-			\frac{n}{n-1}\text{Tr}(\hat{C}_\text{emp}^2)
-			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
-			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
-			- \frac{1}{n(n-1)}\sum_{i=1}^n||\mathbf{x}_i||_2^4
+      \frac{N}{N-1}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
+      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+			- \frac{1}{N(N-1)}\sum_{i=1}^N||\mat{x}_i||_2^4
 		}
 		{
-			\frac{n^2 -2n}{(n-1)^2}\text{Tr}(\hat{C}_\text{emp}^2)
-			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
-			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
-			+ \frac{1}{n(n-1)^2}\sum_{i=1}^n||\mathbf{x}_i||_2^4
+      \frac{N^2 -2N}{(N-1)^2}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
+      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+			+ \frac{1}{N(N-1)^2}\sum_{i=1}^N||\mat{x}_i||_2^4
 		}
@@
 
-	We opt for the LOOCV shrinkage estimator because it avoids some of the
+	We opt for the \ac{loocv} shrinkage estimator because it avoids some of the
 	assumptions made by~\cite{Ledoit2004} and~\cite{Chen2010} and
 	because it generalizes to structured covariance estimation as described in
-	\autoref{seq:structured_estimation}.
+	\cref{seq:structured_estimation}.
 
@@
 	\label{seq:structured_estimation}
-	Exploiting prior knowledge about the spatiotemporal structure of the EEG signal leads to a more regularized estimator of the covariance.
-	When viewing the example of empirical spatiotemporal EEG covariance in
-	\autoref{fig:kronlda-covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
-	Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal EEG epochs is a Kronecker
+	Exploiting prior knowledge about the spatiotemporal structure of the \ac{eeg} signal leads to a more regularized estimator of the covariance.
+	When viewing the example of empirical spatiotemporal \ac{eeg} covariance in
+	\cref{fig:kronlda-covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
+	Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal \ac{eeg} epochs is a Kronecker
 	product of two smaller
@@
 	\begin{equation}
-		\hat{C}_\text{struct} = \hat{S} \otimes \hat{T}
+    \hat{\mat{C}}_\text{struct} = \hat{\mat{S}} \otimes \hat{\mat{T}}
 		\label{eq:kronecker}
@@
 	with $\otimes$ the Kronecker product operator.
-	$\hat{S} \in \mathbb{R}^{c\times c}$ and $\hat{T} \in \mathbb{R}^{s\times s}$ respectively correspond to estimators of the spatial and temporal covariance of the data.
-	Furthermore, because the temporal covariance of the EEG-signal is
+  $\hat{\mat{S}} \in \mathbb{R}^{C\times C}$ and $\hat{\mat{T}} \in \mathbb{R}^{S\times S}$ respectively correspond to estimators of the spatial and temporal covariance of the data.
+	Furthermore, because the temporal covariance of the \ac{eeg}-signal is
 	stationary (i.e., it is only dependent on interval length between covarying
@@
 	\begin{equation}
-		\hat{T}_{i,j} = \hat{T}_{i+1,j+1}
+    \hat{\mat{T}}_{i,j} = \hat{\mat{T}}_{i+1,j+1}
 		\label{eq:toeplitz}
@@
 
-	\autoref{prop:inverse_kronecker} then leads to
-	\autoref{eq:cov_inverse_kronecker} to estimate the inverse
+	\Cref{prop:inverse_kronecker} then leads to
+	\cref{eq:cov_inverse_kronecker} to estimate the inverse
 	covariance.
 	\begin{property}
-		$(U \otimes V)^+ = U^+ \otimes V^+$ for any non-singular
-		matrices $U$ and $V$~\cite{Langville2004}.
+    $(\mat{U} \otimes \mat{V})^+ = \mat{U}^+ \otimes \mat{V}^+$ for any non-singular
+    matrices $\mat{U}$ and $\mat{V}$~\cite{Langville2004}.
 		\label{prop:inverse_kronecker}
@@
 	\begin{equation}
-		\widehat{C^{-1}}_\text{struct} = \hat{S}^+ \otimes \hat{T}^+
+    \widehat{\mat{C}^{-1}}_\text{struct} = \hat{\mat{S}}^+ \otimes \hat{\mat{T}}^+
 		\label{eq:cov_inverse_kronecker}
@@
 
-	Finally, based on \autoref{prop:kronecker_multiplication},
-	\autoref{eq:closed_form} can be reformulated more efficiently as
-	\autoref{eq:closed_form_kronecker}.
+	Finally, based on \cref{prop:kronecker_multiplication},
+	\cref{eq:closed_form} can be reformulated more efficiently as
+	\cref{eq:closed_form_kronecker}.
 	\begin{property}
-		$(U\otimes V)\cdot\text{vec}(W) = \text{vec}(VWU^\intercal)$
-		for any matrices $U\in\mathbb{R}^{p\times p}$,
-		$V\in\mathbb{R}^{q\times q}$ and $W\in\mathbb{R}^{p\times q}$~\cite{Loan2000}.
+    $(\mat{U}\otimes \mat{V})\cdot\text{vec}(\mat{W}) =
+    \text{vec}(\mat{V}\mat{W}\mat{U}^\intercal)$
+    for any matrices $\mat{U}\in\mathbb{R}^{P\times P}$,
+    $\mat{V}\in\mathbb{R}^{Q\times Q}$ and $\mat{W}\in\mathbb{R}^{P\times Q}$~\cite{Loan2000}.
 		\label{prop:kronecker_multiplication}
@@
 	\begin{equation}
-		\hat{\mathbf{w}}_\text{struct} =
-		\frac{\hat{S}^+A^T\hat{T}^+}
-		{\mathbf{a}\cdot\text{vec}(\hat{S}^+A^T\hat{T}^+)}
+		\hat{\mat{w}}_\text{struct} =
+    \frac{\hat{\mat{S}}^+\mat{A}^\intercal\hat{\mat{T}}^+}
+    {\mat{a}\cdot\text{vec}(\hat{\mat{S}}^+A^\intercal\hat{\mat{T}}^+)}
 		\label{eq:closed_form_kronecker}
@@
 
-	Using \autoref{eq:closed_form_kronecker} removes the need to calculate the full, high dimensional Kronecker product $\hat{S}^+\otimes
-		\hat{T}^+$.
-	\autoref{fig:kronlda-covs}e and \autoref{fig:covs}f respectively show examples of the
+	Using \cref{eq:closed_form_kronecker} removes the need to calculate the
+  full, high dimensional Kronecker product $\hat{\mat{S}}^+\otimes
+    \hat{\mat{T}}^+$.
+	\Cref{fig:kronlda-covs}e and \cref{fig:covs}f respectively show examples of the
 	structured covariance and inverse covariance estimators,
-	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
-	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:kronlda-covs}i and
-	\autoref{fig:kronlda-covs}j).
+	consisting of a spatial Kronecker factor (\cref{fig:kronlda-covs}g and
+	\cref{fig:kronlda-covs}h) and a temporal component (\cref{fig:kronlda-covs}i and
+	\cref{fig:kronlda-covs}j).
 
@@
 			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
-			determined by the closed-form LOOCV method. (\textbf{e,f}) Kronecker-Toeplitz
+			determined by the closed-form \ac{loocv} method. (\textbf{e,f}) Kronecker-Toeplitz
 			structured covariance and inverse covariance matrices.
@@
 
-	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
+	The Kronecker approach has shown significant performance yields in different linear spatiotemporal \ac{eeg} and MEG
 	applications~\cite{DeMunck2002,Huizenga2002,Beltrachini2013,GonzalezNavarro2016,GonzalezNavarro2017}.
-	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
+	\textcite{Vliet2020} have applied a Kronecker-structured covariance estimator to \ac{erp} classification with linear models in a post-hoc fashion.
 	Our work goes further by embedding the Kronecker structure in the
@@
 		\begin{equation}
-			\tilde{S}_{k+1} =
-			\frac{1}{n}
-			\sum^n_{i=1}X_i^\intercal\hat{T}_k^+X_i
+      \tilde{\mat{S}}_{k+1} =
+			\frac{1}{N}
+      \sum^N_{i=1}\mat{X}_i^\intercal\hat{\mat{T}}_k^+\mat{X}_i
 			\label{eq:fpi_spatial}
@@
 		\begin{equation}
-			\tilde{T}_{k+1} =
-			\frac{1}{n}
-			\sum^n_{i=1}X_i\hat{S}_k^+X_i^\intercal
+      \tilde{\mat{T}}_{k+1} =
+			\frac{1}{N}
+      \sum^N_{i=1}\mat{X}_i\hat{\mat{S}}_k^+\mat{X}_i^\intercal
 			\label{eq:fpi_temporal}
@@
 		\begin{equation}
-			\tilde{S}_{k+1}^{(\beta)} =
-			(1-\beta_{k+1})\tilde{S}_{k+1}
-			+\beta_{k+1}\frac{\text{Tr}(\tilde{S}_{k+1})}{c}\mathbb{I}
+      \tilde{\mat{S}}_{k+1}^{(\beta)} =
+      (1-\beta_{k+1})\tilde{\mat{S}}_{k+1}
+      +\beta_{k+1}\frac{\text{Tr}(\tilde{\mat{S}}_{k+1})}{C}\mathbb{I}
 			\label{eq:fpi_spatial_shrunk}
@@
 		\begin{equation}
-			\tilde{T}_{k+1}^{(\gamma)} =
-			(1-\gamma_{k+1})\tilde{T}_{k+1}
-			+\gamma_{k+1}\frac{\text{Tr}(\tilde{T}_{k+1})}{s}\mathbb{I}
+      \tilde{\mat{T}}_{k+1}^{(\gamma)} =
+      (1-\gamma_{k+1})\tilde{\mat{T}}_{k+1}
+      +\gamma_{k+1}\frac{\text{Tr}(\tilde{\mat{T}}_{k+1})}{S}\mathbb{I}
 			\label{eq:fpi_temporal_shrunk}
@@
 		\begin{equation}
-			\hat{S}_{k+1} =
-			\frac{c}{\text{Tr}\left[\tilde{S}_{k+1}^{(\beta)}\right]}
-			\tilde{S}_{k+1}^{(\beta)}
+      \hat{\mat{S}}_{k+1} =
+      \frac{C}{\text{Tr}\left[\tilde{\mat{S}}_{k+1}^{(\beta)}\right]}
+      \tilde{\mat{S}}_{k+1}^{(\beta)}
 			\label{eq:fpi_spatial_norm}
@@
 		\begin{equation}
-			\hat{T}_{k+1} =
-			\frac{s}{\text{Tr}\left[\tilde{T}_{k+1}^{(\gamma)}\right]}
-			\tilde{T}_{k+1}^{(\gamma)}
+      \hat{\mat{T}}_{k+1} =
+      \frac{S}{\text{Tr}\left[\tilde{\mat{T}}_{k+1}^{(\gamma)}\right]}
+      \tilde{\mat{T}}_{k+1}^{(\gamma)}
 			\label{eq:fpi_temporal_norm}
@@
 
-	$\hat{S}_0$ and $\hat{T}_0$ can be initialized to any positive definite matrix.
-	We choose to use the identity matrices $\mathbb{I}^{c\times c}$ and $\mathbb{I}^{s\times s}$.
-	After each iteration, all diagonals of $\hat{R}_{k+1}$ are set to their mean
-	values to ensure that $\hat{R}_{k+1}$ and $\hat{T}_{k+1}$ are Toeplitz structured.
+  $\hat{\mat{S}}_0$ and $\hat{\mat{T}}_0$ can be initialized to any positive definite matrix.
+	We choose to use the identity matrices $\mathbb{I}^{C\times C}$ and $\mathbb{I}^{S\times S}$.
+  After each iteration, all diagonals of $\hat{\mat{R}}_{k+1}$ are set to their mean
+  values to ensure that $\hat{\mat{R}}_{k+1}$ and $\hat{\mat{T}}_{k+1}$ are Toeplitz structured.
 
-	Xie et al.~\cite{Xie2021} show that the LOOCV estimates for the
+	\textcite{Xie2021} show that the \ac{loocv} estimates for the
 	optimal values of $\beta_{k+1}$ and $\gamma_{k+1}$ also yield a closed-form
@@
 			\frac{
-				\frac{n}{n-1}\text{Tr}(\tilde{S}_{k+1}^2)
-				- \frac{2}{c}\left[\text{Tr}(\tilde{S}_{k+1})\right]^2
-				+ \frac{1}{c}\text{Tr}(\tilde{S}_{k+1}^2)
-				- \frac{1}{n(n-1)}\sum_{i=1}^n
-				\left[\text{Tr}(X_i\hat{T}_k^+X_i^\intercal)^2\right]
+        \frac{N}{N-1}\text{Tr}(\tilde{\hat{S}}_{k+1}^2)
+        - \frac{2}{C}\left[\text{Tr}(\tilde{\mat{S}}_{k+1})\right]^2
+        + \frac{1}{C}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
+				- \frac{1}{N(N-1)}\sum_{i=1}^N
+        \left[\text{Tr}(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
 			}{
-				\frac{n^2-2n}{(n-1)^2}\text{Tr}(\tilde{S}_{k+1}^2)
-				- \frac{2}{c}\left[\text{Tr}(\tilde{S}_{k+1})\right]^2
-				+ \frac{1}{c}\text{Tr}(\tilde{S}_{k+1}^2)
-				+ \frac{1}{n(n-1)^2}\sum_{i=1}^n
-				\left[\text{Tr}(X_i\hat{T}_k^+X_i^\intercal)^2\right]
+        \frac{N^2-2N}{(N-1)^2}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
+        - \frac{2}{C}\left[\text{Tr}(\tilde{\mat{S}}_{k+1})\right]^2
+        + \frac{1}{C}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
+				+ \frac{1}{N(N-1)^2}\sum_{i=1}^N
+        \left[\text{Tr}(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
 			}
@@
 			\frac{
-				\frac{n}{n-1}\text{Tr}(\tilde{T}_{k+1}^2)
-				- \frac{2}{s}\left[\text{Tr}(\tilde{T}_{k+1})\right]^2
-				+ \frac{1}{s}\text{Tr}(\tilde{T}_{k+1}^2)
-				- \frac{1}{n(n-1)}\sum_{i=1}^n
-				\left[\text{Tr}(X_i^\intercal\hat{S}_k^+X_i)^2\right]
+        \frac{N}{N-1}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+        - \frac{2}{S}\left[\text{Tr}(\tilde{\mat{T}}_{k+1})\right]^2
+				+ \frac{1}{S}\text{Tr}(\tilde{T}_{k+1}^2)
+				- \frac{1}{N(N-1)}\sum_{i=1}^N
+        \left[\text{Tr}(\mat{X}_i^\intercal\mat{\mat{S}}_k^+\mat{X}_i)^2\right]
 			}{
-				\frac{n^2-2n}{(n-1)^2}\text{Tr}(\tilde{T}_{k+1}^2)
-				- \frac{2}{s}\left[\text{Tr}(\tilde{T}_{k+1})\right]^2
-				+ \frac{1}{s}\text{Tr}(\tilde{T}_{k+1}^2)
-				+ \frac{1}{n(n-1)^2}\sum_{i=1}^n
-				\left[\text{Tr}(X_i^\intercal\hat{S}_k^+X_i)^2\right]
+        \frac{N^2-2N}{(N-1)^2}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+        - \frac{2}{S}\left[\text{Tr}(\tilde{\mat{T}}_{k+1})\right]^2
+        + \frac{1}{S}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+				+ \frac{1}{N(N-1)^2}\sum_{i=1}^N
+        \left[\text{Tr}(\mat{X}_i^\intercal\hat{\mat{S}}_k^+\mat{X}_i)^2\right]
 			}
@@
 	The Oracle Approximation Shrinkage method can also be used to determine
-	$\beta_{k+1}$ and $\gamma_{k+1}$~\cite{Chen2010,Xie2021} but performs worse for spatiotemporal EEG data since not all assumptions are met.
+	$\beta_{k+1}$ and $\gamma_{k+1}$~\cite{Chen2010,Xie2021} but performs worse for spatiotemporal \ac{eeg} data since not all assumptions are met.
 
 	\subsection{Dataset}
-	We use the dataset from~\cite{Wittevrongel2016}, containing P300 oddball EEG
+	We use the dataset from~\textcite{Wittevrongel2016}, containing P300 oddball \ac{eeg}
 	recordings of 21 healthy subjects since it is a high-quality dataset with a high
@@
 	Each stimulation will correspond to a single epoch in the preprocessed dataset.
-	See~\cite{Wittevrongel2016} for a complete description of the dataset and the recording procedure.
+	See~\textcite{Wittevrongel2016} for a complete description of the dataset and the recording procedure.
 
@@
 	(version 0.24.0)~\cite{Gramfort2013}.
-	The dataset was converted to BIDS-EEG format~\cite{Pernet2019} and managed and
+	The dataset was converted to BIDS-\ac{eeg} format~\cite{Pernet2019} and managed and
 	loaded with MNE-BIDS (version 0.9)~\cite{Appelhoff2019}.
-	The Riemannian classifier from \autoref{sec:riemannian} was implemented using
+	The Riemannian classifier from \cref{sec:riemannian} was implemented using
 	pyRiemann (version 0.2.7).
@@
 
-	The EEG recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
+	The \ac{eeg} recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
 	The reference electrodes were dropped from the analysis.
@@
 	filtering with a fourth-order Butterworth IIR filter.
-	The EEG signal was corrected for ocular artifacts using Independent Component
+	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
 	Analysis (ICA) by rejecting components that correlated with the bipolar EOG channels vEOG and hEOG, according to adaptive Z-score thresholding.
@@
 	Before training the classifiers, a Z-score normalization transformation was
-	developed on the training data to scale all EEG channels to unit variance.
+	developed on the training data to scale all \ac{eeg} channels to unit variance.
 	This transformation was then applied to the validation data.
 
-	\subsubsection{Spatiotemporal beamformer classifier}
-	Before calculating the spatiotemporal beamformer (STBF), the signal was downsampled to
+  \subsubsection{\Acl{stbf} classifier}
+	Before calculating the \ac{stbf}, the signal was downsampled to
 	32Hz or twice the low-pass frequency 16Hz, resulting in 17 time samples
@@
 	We constructed three variations of the spatiotemporal beamformer:
-	STBF with empirical covariance estimation (\textsc{stbf-emp}) as in
-	\autoref{sec:empirical_covariance}, STBF with
-	LOOCV-shrunk covariance estimation (\textsc{stbf-shrunk}) as in
-	\autoref{sec:shrunk_covariance}, and STBF with
-	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct} )with LOOCV shrinkage for
-	the Kronecker factors as in \autoref{sec:structured_covariance}.
+	\ac{stbf} with empirical covariance estimation (\textsc{stbf-emp}) as in
+	\cref{sec:empirical_covariance}, \ac{stbf} with
+	\ac{loocv}-shrunk covariance estimation (\textsc{stbf-shrunk}) as in
+	\cref{sec:shrunk_covariance}, and \ac{stbf} with
+	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct} )with \ac{loocv} shrinkage for
+	the Kronecker factors as in \cref{sec:structured_covariance}.
 
@@
 	geometry-based classifier to compare our results.
-	The Riemannian model (x\textsc{dawn+rg}) uses the xDAWN spatial filter combined
+	The Riemannian model (\textsc{xdawn+rg}) uses the XDAWN spatial filter combined
 	with Riemannian geometry in tangent space as implemented by Barachant et
 	al.~\cite{Barachant2014a}.
-	This classifier uses four xDAWN spatial filters and each epoch's empirical spatial covariance matrix.
+	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
 	The target with the maximum score is the prediction of the cued target.
-	x\textsc{dawn+rg} was trained and validated without downsampling using epochs
+	\textsc{xdawn+rg} was trained and validated without downsampling using epochs
 	at the original sample rate of 2048Hz.
@@
 	The fixed point iteration algorithm described
-	in Equations \ref{eq:fpi_spatial}-\ref{eq:fpi_temporal_shrunk} estimates the Kronecker-Toeplitz structured covariance for the
+	in
+  \crefrange{eq:fpi_spatial}{eq:fpi-temporal-norm}
+  estimates the Kronecker-Toeplitz structured covariance for the
 	\textsc{stbf-struct} classifier.
@@
 
-	\autoref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
+	\cref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
 	Using more than one fixed-point iteration does not significantly improve the
@@
 	\subsection{Classifier accuracy for limited training data}
-	It is of interest to keep the calibration time before BCI
+	It is of interest to keep the calibration time before \ac{bci}
 	operation as short as possible.
@@
 	corresponding number of blocks in the training folds and dropping the rest.
-	\autoref{fig:accuracy} and \autoref{fig:accuracy_ap} show each classifier's
+	\cref{fig:accuracy} and \cref{fig:accuracy_ap} show each classifier's
 	accuracy relative to the data availability.
@@
 	the first five trials of a block.
-	Results validated on one trial are reported in \autoref{tab:p_values_1}, 2-trial results
-	in \autoref{tab:p_values_2}, and 5-trial results in \autoref{tab:p_values_5}.
+	Results validated on one trial are reported in \cref{tab:p_values_1}, 2-trial results
+	in \cref{tab:p_values_2}, and 5-trial results in \cref{tab:p_values_5}.
 
@@
 			3 trials of testing stimuli.
-			\autoref{fig:accuracy_ap} contains results for all numbers of trials.
+			\cref{fig:accuracy_ap} contains results for all numbers of trials.
 			While \textsc{stbf-emp} is unstable when little training data
@@
 		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
-      $p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
+      $p$-values as in \cref{tab:p_values_1} averaging over two testing trials.}
 		\label{tab:p_values_2}
@@
     \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
-      $p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
+      $p$-values as in \cref{tab:p_values_1} averaging over five testing trials.}
 		\label{tab:p_values_5}
@@
 	than \textsc{stbf-emp} for all evaluated settings.
-	Compared to x\textsc{dawn+rg}, \textsc{stbf-struct} also has significantly
+	Compared to \textsc{xdawn+rg}, \textsc{stbf-struct} also has significantly
 	higher accuracy in almost all evaluated settings, except when using only one training block.
-	\textsc{stbf-shrunk} does not outperform x\textsc{dawn+rg} when training data is low but gains a significant advantage
+	\textsc{stbf-shrunk} does not outperform \textsc{xdawn+rg} when training data is low but gains a significant advantage
 	when using more training data.
@@
 	cross-validation scheme is run four times for each subject, each time with an
-	increasing number of EEG channels retained in the analysis, to explore the scalability of each classifier for analyses with higher spatial resolutions.
+	increasing number of \ac{eeg} channels retained in the analysis, to explore the scalability of each classifier for analyses with higher spatial resolutions.
 	The temporal resolutions were not varied, but we expect that increasing the
@@
 	training times for the \textsc{stbf}-based as evidenced by the complexity
-	calculations in \autoref{sec:parameters_complexity}.
-	\autoref{fig:training_time} shows the measured training times.
+	calculations in \cref{sec:parameters_complexity}.
+	\cref{fig:training_time} shows the measured training times.
 	These results were obtained using a laptop with an Intel® Core™ i7-8750H CPU and 16GB of RAM.
@@
 			All \textsc{stbf} classifiers can be trained significantly faster than
-			x\textsc{dawn+rg}.}
+			\textsc{xdawn+rg}.}
 		\label{fig:training_time}
@@
 
-	\autoref{fig:training_time} shows that the training time of
+	\cref{fig:training_time} shows that the training time of
 	\textsc{stbf-struct} increases less steeply than that of \textsc{stbf-shrunk}
-	and \textsc{stbf-emp}. The training time of all three STBF-based classifiers is
-	much lower than that of x\textsc{dawn+rg}, which appears nearly constant when using
+	and \textsc{stbf-emp}. The training time of all three \ac{stbf}-based classifiers is
+	much lower than that of \textsc{xdawn+rg}, which appears nearly constant when using
 	4, 8, 16, or 32 channels.
@@
 	\subsection{Classification accuracy}
-	As evidenced by \autoref{fig:accuracy} and Tables \ref{tab:p_values_1}-\ref{tab:p_values_5}, the regularized classifiers \textsc{stbf-shrunk}
+	As evidenced by \cref{fig:accuracy} and Tables \ref{tab:p_values_1}-\ref{tab:p_values_5}, the regularized classifiers \textsc{stbf-shrunk}
 	and \textsc{stbf-struct} significantly improve the classification accuracy
@@
 	ill-conditioned when the number of available training epochs is smaller than
-	the number of features ($n<cs$), rendering its inversion with the
+	the number of features ($N<CS$), rendering its inversion with the
 	Moore-Penrose pseudoinverse unstable.
-	This is the case \textsc{stbf-emp} when $n=cs=32*17=544$, after which the
+	This is the case \textsc{stbf-emp} when $N=CS=32*17=544$, after which the
 	accuracy of \textsc{stbf-emp} starts to increase.
-	This effect is visible in \autoref{fig:accuracy}, where the accuracy starts
+	This effect is visible in \cref{fig:accuracy}, where the accuracy starts
 	increasing when using more than four training blocks, amounting to 540 epochs.
@@
 	Regularization reduces extreme values present, as shown in
-	\autoref{fig:kronlda-covs}, resulting in a classifier with
+	\cref{fig:kronlda-covs}, resulting in a classifier with
 	better generalization.
-	Third, the true spatiotemporal covariance matrix may vary throughout BCI
-	sessions, e.g., due to movement of the EEG-cap, changing impedances of
+	Third, the true spatiotemporal covariance matrix may vary throughout \ac{bci}
+	sessions, e.g., due to movement of the \ac{eeg}-cap, changing impedances of
 	electrodes, subject fatigue, the introduction of new spatiotemporal noise
@@
 	A regularized covariance matrix should better account for changes in true covariance.
-	Note that the LOOCV method in principle assumes that the covariances of
+	Note that the \ac{loocv} method in principle assumes that the covariances of
 	the training data and unseen data are the same.
 	Because the covariance might have changed for unseen data, the shrinkage
-	estimate obtained with LOOCV is probably still an
+	estimate obtained with \ac{loocv} is probably still an
 	underestimation of the optimal -- but unknown -- shrinkage coefficient that
@@
 	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available training data is small.
-	This property is an attractive advantage in a BCI setting since it is desirable to keep the calibration (training) phase as short as possible without losing accuracy.
+	This property is an attractive advantage in a \ac{bci} setting since it is desirable to keep the calibration (training) phase as short as possible without losing accuracy.
 	The accuracy advantage of the structured estimator is a consequence of the
 	Kronecker-Toeplitz covariance structure, which is informative for the
-	underlying process generating the epochs, if it is assumed that the EEG signal
+	underlying process generating the epochs, if it is assumed that the \ac{eeg} signal
 	is a linear combination of stationary activity generated by random dipoles in
@@
 	covariance.
-	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \autoref{sec:parameters_complexity}), increasing the stability of matrix inversions.
+	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:parameters_complexity}), increasing the stability of matrix inversions.
 
-	When compared to the state-of-the-art x\textsc{dawn+rg} classifier, we conclude
+	When compared to the state-of-the-art \textsc{xdawn+rg} classifier, we conclude
 	that \textsc{stbf-struct} reaches similar accuracy when using only one block of
@@
 	When more data are available, \textsc{stbf-struct} reaches significantly better accuracies.
-	Combined with the benefits laid out in  \autoref{sec:parameters_complexity} and
-	\autoref{sec:interpretability}, this makes it an attractive option for ERP classification.
+	Combined with the benefits laid out in  \cref{sec:parameters_complexity} and
+	\cref{sec:interpretability}, this makes it an attractive option for \ac{erp} classification.
 	\textsc{stbf-shrunk} does not show decisive accuracy improvements over
-	x\textsc{dawn+rg} using a few training blocks, but this improves as the training data increases.
+	\textsc{xdawn+rg} using a few training blocks, but this improves as the training data increases.
 
@@
 	\label{sec:parameters_complexity}
-	As mentioned above, inverting the full $cs \times cs$ dimensional covariance
+	As mentioned above, inverting the full $CS \times CS$ dimensional covariance
 	matrix to construct \textsc{stbf-emp} and \textsc{stbf-shrunk} can be costly
-	and unstable, in particular in high-resolution settings with many EEG channels or time samples.
+	and unstable, in particular in high-resolution settings with many \ac{eeg} channels or time samples.
 	Constructing the full covariance and inverse covariance matrices also requires a considerable amount of memory.
@@
 
-	First, because of \autoref{prop:inverse_kronecker} and
-	\autoref{prop:kronecker_multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
+	First, because of \cref{prop:inverse_kronecker} and
+	\cref{prop:kronecker_multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
 	symmetric covariance and inverse covariance matrices for \textsc{stbf-struct}; they can instead be replaced by two smaller symmetric matrices respectively of dimensions $c\times c$ and $s\times s$.
@@
 	While the inverse covariance of \textsc{stbf-emp} and \textsc{stbf-shrunk} is
-	defined by $\frac{cs(cs+1)}{2}=\frac{32\cdot17(32\cdot17+1)}{2}=\num{122128}$
+	defined by $\frac{CS(CS+1)}{2}=\frac{32\cdot17(32\cdot17+1)}{2}=\num{122128}$
 	parameters accounting for the symmetric nature of
-	covariance, the structured estimator only requires $\frac{c(c+1)}{2} + s =
+	covariance, the structured estimator only requires $\frac{C(C+1)}{2} + S =
 		\frac{32(32+1)}{2} + 17=545$ unique parameters.
@@
 	Second, structured estimation has better time complexity.
-	Covariance estimation and inversion occupy the largest part of the STBF training time.
-	For \textsc{stbf-emp} and \textsc{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(nc^2s^2+c^3s^3)$.
+	Covariance estimation and inversion occupy the largest part of the \ac{stbf} training time.
+	For \textsc{stbf-emp} and \textsc{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
 	Thanks to Property~\ref{prop:inverse_kronecker}, the complexity can be reduced to
-	$\mathcal{O}(nc^2s^2+c^3+s^3)$ for the structured estimator of \textsc{stbf-struct}.
-	The results presented in \autoref{fig:training_time} confirm these
+	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \textsc{stbf-struct}.
+	The results presented in \cref{fig:training_time} confirm these
 	calculations.
@@
 
-	The results shown in \autoref{fig:training_time} also confirm that the
-	STBF-based estimators are very fast to train compared to the
-	state-of-the-art estimator x\textsc{dawn+rg}, which confirms the results in~\cite{Wittevrongel2016}.
-	Since the training times of all STBF-based classifiers are already in
+	The results shown in \cref{fig:training_time} also confirm that the
+	\ac{stbf}-based estimators are very fast to train compared to the
+	state-of-the-art estimator \textsc{xdawn+rg}, which confirms the results in~\cite{Wittevrongel2016}.
+	Since the training times of all \ac{stbf}-based classifiers are already in
 	the order of tenths of seconds, the question arises whether the
@@
 	However, the authors believe that these results could significantly impact some
-	use cases of the spatiotemporal beamformer, like high spatial or temporal resolution ERP analyses.
-	One example is single-trial ERP analysis with a high-temporal
-	resolution to extract ERP time features.
-	Such higher-resolution analyses can later be incorporated into an ERP
+	use cases of the spatiotemporal beamformer, like high spatial or temporal resolution \ac{erp} analyses.
+	One example is single-trial \ac{erp} analysis with a high-temporal
+	resolution to extract \ac{erp} time features.
+	Such higher-resolution analyses can later be incorporated into an \ac{erp}
 	classification framework.
 	In addition, the speed-up provided by structured estimation yields a faster
-	off-line evaluation of the STBF ERP classifier, where often multiple cross-validation folds, subjects, and hyperparameter settings need to be explored, which can quickly increase runtime.
+	off-line evaluation of the \ac{stbf} \ac{erp} classifier, where often multiple cross-validation folds, subjects, and hyperparameter settings need to be explored, which can quickly increase runtime.
 	Improvements in computation speed and memory usage can remove the need for dedicated computation hardware and enable running group analyses on a personal computer.
@@
 	%	\label{sec:interpretability}
-	%	The weight matrix of the STBF determines how each
+	%	The weight matrix of the \ac{stbf} determines how each
 	%	spatiotemporal feature of a given epoch should contribute to enhancing the SNR
@@
 	%	task.
-	%	Alternatively, the activation pattern can be regarded as a forward EEG model of
+	%	Alternatively, the activation pattern can be regarded as a forward \ac{eeg} model of
 	%	the activity generating the discriminating signal and the weights as a
@@
 	%	Regularization enables a researcher to interpret better the distribution of the weight over space and time after reshaping the weight vector $\mathbf{w}$ to its spatiotemporal matrix equivalent $W$ such that $\text{vec}(W) = \mathbf{w}$.
-	%	\autoref{fig:interpret_weights} compares the weights calculated in
+	%	\cref{fig:interpret_weights} compares the weights calculated in
 	%	\textsc{stbf-emp} and \textsc{stbf-shrunk} with the weights from
@@
 	%			(\textbf{c}) \textsc{stbf-shrunk} weights. The shrinkage factor $\alpha=0.05$ was
-	%			determined with the closed-form LOOCV-method.
+	%			determined with the closed-form \ac{loocv}-method.
 	%			(\textbf{d}) \textsc{stbf-emp} weights.}
@@
 	%
-	%	As expected, \autoref{fig:interpret_weights}b and d exhibit weight around
-	%	the central and parietal regions, where the P300 ERP component is present.
+	%	As expected, \cref{fig:interpret_weights}b and d exhibit weight around
+	%	the central and parietal regions, where the P300 \ac{erp} component is present.
 	%	Especially the spatial weights of \textsc{stbf-shrunk} in
-	%	\autoref{fig:interpret_weights}d correspond to the spatial activation pattern
-	%	in \autoref{fig:interpret_weights}a.
+	%	\cref{fig:interpret_weights}d correspond to the spatial activation pattern
+	%	in \cref{fig:interpret_weights}a.
 	%	This is not unsurprising, since shrinkage transforms the covariance matrix closer to the
-	%	identity matrix and assuming identity covariance in \autoref{eq:closed_form} yields weights
+	%	identity matrix and assuming identity covariance in \cref{eq:closed_form} yields weights
 	%	identical to the activation pattern (up to a scaling factor).
-	%	Additionaly, \autoref{fig:interpret_weights}b shows that weights in the baseline interval and after 0.6s, which should
+	%	Additionaly, \cref{fig:interpret_weights}b shows that weights in the baseline interval and after 0.6s, which should
 	%	contain no response information, are close to zero for the structured estimator.
@@
 	%	This high weight for the early visual components confirms the results from Treder \& Blankertz~\cite{Treder2010}
-	%	that state that, in addition to the P300, the early N1 and P2 ERP components
+	%	that state that, in addition to the P300, the early N1 and P2 \ac{erp} components
 	%	are also modulated by oddball attention and contain discriminatory information between
@@
 	%	For instance, one can use the weight matrix determine relevant time samples
-	%	and EEG channels for per-subject feature selection to refine the model further .
-	%	The number of channels is also an important cost factor in practical BCI
+	%	and \ac{eeg} channels for per-subject feature selection to refine the model further .
+	%	The number of channels is also an important cost factor in practical \ac{bci}
 	%	applications.
@@
 	\section{Conclusion}
-	While it is possible to regularize the spatiotemporal LCMV beamformer
-	classifier for ERP detection with other methods such as by employing feature selection,
+	While it is possible to regularize the spatiotemporal \ac{lcmv} beamformer
+	classifier for \ac{erp} detection with other methods such as by employing feature selection,
 	by adding regularizing penalties to the cost
@@
 	We introduced a covariance estimator using adaptive shrinkage and an
-	estimator exploiting prior knowledge about the spatiotemporal nature of the EEG
+	estimator exploiting prior knowledge about the spatiotemporal nature of the \ac{eeg}
 	signal.
@@
 	Since these algorithms are not paradigm-specific,  the conclusions can be generalized to
-	other ERP-based BCI settings.
+	other \ac{erp}-based \ac{bci} settings.
 
@@
 	covariance.
-	It could also be interesting to explore the impact of covariance regularization on transfer learning of the STBF between subjects to alleviate calibration entirely.
+	It could also be interesting to explore the impact of covariance regularization on transfer learning of the \ac{stbf} between subjects to alleviate calibration entirely.
 	Finally, it could be insightful to evaluate the proposed algorithms in a
-	real-world on-line BCI setting.
+	real-world on-line \ac{bci} setting.
 

### 2024-10-08 chapters/wcble.tex
Proofread

@@
 \label{sec:wcble/intro}
-\todohsm{do'nt repeat in introduction, rather make line clear}
 
@@
 individuals by decoding \acp{erp} obtained from time-modulated flashing
-stimuli (see Figure~\ref{fig:bci/loop}).
-These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
-\ac{erp} waveform (see Figure~\ref{fig:bci/erp}.
+stimuli (see~\cref{fig:bci/loop}).
+These \acp{erp} consist of multiple \ac{erp}-components, peaks, and troughs in
+\ac{erp} waveform (see \cref{fig:bci/erp}).
 These components each have their own distinct amplitude and latency.
 \Ac{erp} component latency can reflect a wide range of cognitive,
-neurophysiological and clinical properties, that can be confounding factors in
+neurophysiological, and clinical properties, that can be confounding factors in
 the development of \acp{bci}~\cite{Luck2014}.
@@
 Latencies of non-averaged, single-trial \acp{erp} should capture even more
-relevant information, yet they are very hard to measure due to low \ac{snr} of
-single trial \acp{erp}.
+relevant information, yet they are very hard to measure due to the low \ac{snr} of
+single-trial \acp{erp}.
 Within-session variability of \ac{erp} latency, called \ac{erp} latency
@@
 \textcite{Zisk2021} showed that patients with \ac{als} have significantly
-increased ERP jitter.
-\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
-performance, and disproportionally so when the user is not directly gazing at
+increased \ac{erp} jitter.
+\textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
+performance, and disproportionately so when the user is not directly gazing at
 intended targets, such as is the case for individuals with \ac{sspgi}.
@@
 The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
-most prominent of which is the smearing effect as illustrated
-in~\autoref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
-averaged per condition to cope with the unfavorable \ac{snr}?
+most prominent of which is the smearing effect, as illustrated
+in~\cref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
+averaged per condition to cope with the unfavorable \ac{snr}.
 When averaging over multiple identical responses with jittered latencies, the shape
@@
 More generally, the smearing effect leads to difficulties interpreting \ac{erp}
-results and making inferences about amplitude or latency effects, because these
+results and making inferences about amplitude or latency effects because these
 are entangled when using averaging~\cite{Luck2014}.
@@
 less effective templates for \ac{erp} retrieval.
-Similarly, the noise model in \ac{lda}'s  assumes stationarity and subtracts
+Similarly, the noise model in \ac{lda} assumes stationarity and subtracts
 the class-wise averages from the signal to calculate
@@
 This can be done by estimating the latency of each
-single-trial \ac{erp}, and aligning them before averaging.
+single-trial \ac{erp} and aligning them before averaging.
 Latency estimation has multiple advantages:
@@
 	experimental conditions. These are variables of interest for
-	testing electrophysiological an behavioral hypotheses.
+	testing electrophysiological and behavioral hypotheses.
 	Additionally, latency estimation allows for the extraction of
 	other descriptors from a set of single-trial \acp{erp}, like
-	amplidute or latency variance~\cite{Hultsch2004}, which can
+	amplitude or latency variance~\cite{Hultsch2004}, which can
 	also be of interest as studied variables.
@@
 	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
-  components which would otherwise obscured or blurred due to desynchronization.
+  components which would otherwise be obscured or blurred due to desynchronization.
     These \emph{deblurred} templates can then be used for analysis of the \ac{erp}
@@
 	extended to inter-session or inter-subject differences, can aid in transfer learning or
-	generalization across subjects and protocols\cite{Iturrate2014}.
+	generalization across subjects and protocols~\cite{Iturrate2014}.
 \end{enumerate*}
@@
 The simplest method to estimate single-trial latency is
-\emph{peak picking}
+\emph{peak picking}.
 Here, the latency of an \ac{erp} is determined as the
-time-point of it's maximum (or minimum) amplitude relative to stimulus
+time-point of its maximum (or minimum) amplitude relative to stimulus
 onset.
@@
 Several filtering approaches have been developed to be used in
-conjuction with peak-picking, in the time domain~\cite{
+conjunction with peak-picking, in the time domain~\cite{
   Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999, Sparacino2002,
@@
 and in the time-frequency domain~\cite{
-  Quiroga2003, Wang2007,Hu2010}.
+  Quiroga2003, Wang2007, Hu2010}.
 The aforementioned analysis and performance prediction method proposed
-by \cite{Arico2014} estimates the single-trial P3 latencies of attended
-epochs by P3 peak-picking after filtering in the time-frequency domain, and
+by \textcite{Arico2014} estimates the single-trial P3 latencies of attended
+epochs by P3 peak-picking after filtering in the time-frequency domain and
 corrects for jitter by temporally aligning all target epochs to the average
@@
 also be done in the spatial domain. The spatial filter can be constructed
-using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
+using \ac{pca}~\cite{Ouyang2017}, \ac{ica}~\cite{Townsend1999, Jung2001,
 	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.
@@
 latency estimation method that is still being used and is even
-considered amongst the more performant techniques for
+considered among the more performant techniques for
 \acp{erp}~\cite{Thornton2008, Ouyang2017}.
@@
 on these time points to form the next template.
-Similarily, cross-correlation template matching can be combined with filtering.
-\cite{Souloumiac2013} uses XDAWN to construct spatial filter before
-template matching the spatially filtered signal. \cite{Iturrate2014} uses a
+Similarly, cross-correlation template matching can be combined with filtering.
+\textcite{Souloumiac2013} uses XDAWN to construct a spatial filter before
+template matching the spatially filtered signal. \textcite{Iturrate2014} uses a
 spatiotemporal filter. Instead of aligning the
@@
 Alternative methods are based on graph
-optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020} or by fitting an \ac{erp} model with
-maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988,Puce1994}.
+optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020}, or by fitting an \ac{erp} model with
+maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988, Puce1994}.
 
@@
 In the previous section, we have abstracted away the fact that \acp{erp} are
-multi-component signals, which is not taking into account by the aforementioned
+multi-component signals, which is not taken into account by the aforementioned
 methods.
@@
 The presence of more than one component can hamper the
-performance of some latency estimation algorithms, and it often only possible to extract only the
-latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
+performance of some latency estimation algorithms, and it is often only possible to extract the latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
 to one component and averaging, this will introduce blurring in the other
@@
 by carefully selecting peaks or pre-determined regions of interest.
-\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
+\textcite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
 time-window for each \ac{erp} component to extract a latency per component.
-Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
+Methods based on spatial decomposition filters like \ac{pca}, \ac{ica}, or XDAWN can be leveraged
 to utilize multiple of their components as filters, resulting in multiple time
-series where peack picking or Woody's algorithm can be applied. In general,
+series where peak picking or Woody's algorithm can be applied. In general,
 however, these algorithms lack an integrated approach to deal with multiple
@@
 
-Some algorithms separate components into stimulus and response locked component clusters based on their
+Some algorithms separate components into stimulus- and response-locked component clusters based on their
 latency distributions~\cite{Jung2001, Takeda2008, Zhang1998, Yin2009}.
@@
 The RIDE algorithm~\cite{Ouyang2011, Ouyang2015, Wang2015, Ouyang2016, Ouyang2020}
-is able to do this without entirely relying on the response time, and can
+is able to do this without entirely relying on the response time and can
 separate stimulus-locked, response-locked, and multiple central component clusters.
 Others, like spatial filtering methods that use decomposition methods to
-determine the spatial filter can be be modified to separate
+determine the spatial filter, can be modified to separate
 components~\cite{Ouyang2017}.
@@
 Finally, the smearing problem can also be overcome by algorithms that are not
-based on \ac{erp} component latency estimation, but which use time warping-based methods
+based on \ac{erp} component latency estimation but use time-warping-based methods
 to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
@@
 These methods work
-fundamentally different from the latency estimation methods, and cannot
+fundamentally differently from the latency estimation methods, and cannot
 directly be used to extract latency features.
@@
 for jitter-resistant decoding.
-\cite{Hardiansyah2020} incorporated single-trial latencies in
+\textcite{Hardiansyah2020} incorporated single-trial latencies in
 classification by peak-picking within a given
@@
 The \ac{cble} algorithm introduced
-by \cite{Thompson2012} also explicitly applies latency estimation in a
+by \textcite{Thompson2012} also explicitly applies latency estimation in a
 decoding setting.
-\cite{Thompson2012} initially formulated \ac{cble} as an off-line performance
+\textcite{Thompson2012} initially formulated \ac{cble} as an offline performance
 prediction method.
 Later, its output was successfully adapted to compensate for jitter to improve
-decoder performance~\cite{Mowla2017,Zisk2022}.
+decoder performance~\cite{Mowla2017, Zisk2022}.
 
-Time series classification algorithms~\cite{Abanda2019}
+Time-series classification algorithms~\cite{Abanda2019}
 that are robust to jitter can be used in a decoding setting,
@@
 Data augmentation involving jittering the training
-data~\cite{Krell2018,Zisk2022} and Riemannian Geometry methods using spatial
+data~\cite{Krell2018, Zisk2022} and Riemannian Geometry methods using spatial
 covariances as features~\cite{Aydarkhanov2020} have both been shown
@@
 	\mathbb{R}^{C\times S}\}_{m=1}^M$.
-We assume that the sampling period is $T$, i.e. that the sample with index $s$ is sampled at time $sT$.
+We assume that the sampling period is $T$, i.e., that the sample with index $s$ is sampled at time $sT$.
 In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
@@
 
-\Ac{cble}, summarized in \autoref{sec:wcble/app/cble-alg}, works by training a
+\Ac{cble}, summarized in \cref{sec:wcble/app/cble-alg}, works by training a
 \textit{first-stage} classifier $\mathcal{C}(\theta,f)$
@@
 Then,
-%instead of applying the classifier directly to an unseen, cropped
-%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
 $f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
@@
 spatiotemporal classifier weights as template.
-\cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
+\textcite{Thompson2012}, \textcite{Mowla2017}, and \textcite{Mowla2020} show
 that \ac{cble} is relatively independent of the first-stage classifier for \ac{bci}
@@
 Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
-regularized covariance matrix (tLDA) proposed by \cite{Sosulski2022}, as the first-stage
+regularized covariance matrix (tLDA) proposed by \textcite{Sosulski2022}, as the first-stage
 classifier and logistic regression as second stage.
@@
 \label{sec:robust-latency}
-The \ac{erp} decoding method based on \ac{cble} introduced by \cite{Mowla2017} does
+The \ac{erp} decoding method based on \ac{cble} introduced by \textcite{Mowla2017} does
 not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
 to the second-stage classifier.
-Furthermore, previous \ac{cble} works~\cite{Thompson2012,Mowla2020} only used these
+Furthermore, previous \ac{cble} works~\cite{Thompson2012, Mowla2020} only used these
 latencies to correlate them with neurophysiological processes or to predict
@@
 
-However, including these latency features give rise to the following issue when
+However, including these latency features gives rise to the following issue when
 classifying unseen data.
@@
 $s_\mathrm{target}=\argmax_s{y_s}$.
-This is the point in time where the target class reaches largest separation
+This is the point in time where the target class reaches the largest separation
 from the background noise and non-target class, indicating the target \ac{erp} is
@@
 However, this is problematic since it is not evident how to estimate
-latency of e.g. a P3 \ac{erp} component for a non-target epoch, since the
+the latency of, e.g., a P3 \ac{erp} component for a non-target epoch, since the
 non-target class is characterized by the absence of this component.
@@
   \item lead to smoother convergence in our proposed iterative alignment scheme
-    for W\ac{cble}, which heavily relies on exact latency estimation.
+    for \ac{wcble}, which heavily relies on exact latency estimation.
 \end{enumerate*}
@@
 per class $\Pr(l|
-\mat{X}[:,s_1:s_2:)], \theta)$ for a given epoch $\mat{X}$, a feature of many
+\mat{X}[:,s_1:s_2)], \theta)$ for a given epoch $\mat{X}$, a feature of many
 common classifiers.
@@
 \Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
-  \quad \forall \quad s\in 1,\ldots,R
+  \quad \forall s\in 1,\ldots,R
 	\label{eq:prob}
@@
 \begin{equation}
-  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right] \\
+  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right]
   \label{eq:prob-latency}
 \end{equation}
-Note that $  \Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
+Note that $\Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
   \Pr(s|\mat{X},\theta,\mathrm{target})$.
@@
 directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
-While the Woody algorithm iteratively enhances the SNR of an \ac{erp} template to
-cross-correlate with the data, W\ac{cble} iteratively re-estimates the parameters of
+While the Woody algorithm iteratively enhances the \ac{snr} of an \ac{erp} template to
+cross-correlate with the data, \ac{wcble} iteratively re-estimates the parameters of
 the first-stage classifier.
-To improve convergence and perform well in a classification setting, W\ac{cble}
+To improve convergence and perform well in a classification setting, \ac{wcble}
 aligns both targets and non-targets to their corresponding estimated latencies.
 
-The W\ac{cble} algorithm is presented in \autoref{sec:wcble/app/wcble-alg}.
-Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
+The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
+Its training phase is visualized in~\cref{fig:diagram-train}.
 The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
@@
 Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
-to its class label $l_n$ using equation~\ref{eq:prob-latency}.
+to its class label $l_n$ using \cref{eq:prob-latency}.
 Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
@@
 \begin{equation}
-  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
+  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
 \end{equation}
@@
   and class probabilities over time.
-  (3) The medians of these probability distributions are assumed the class
+  (3) The medians of these probability distributions are assumed as the class
   latencies.
@@
   (5) The spatiotemporal classifier is then retrained on the aligned epochs
-  for a	next iteration.
+  for the next iteration.
   (6) After the iterative process halts, the scores and latencies obtained
@@
 }
-The process halts after a fixed amount of iterations or when the estimated set
+The process halts after a fixed number of iterations or when the estimated set
 of latencies has been encountered before, indicating it ended up in a loop.
@@
 We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
-manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
-vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
-in~Section~\ref{sec:robust-latency}.
+manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
+vector of scores over time as in~\cref{sec:cble} and the estimated latencies as
+in~\cref{sec:robust-latency}.
 \fullpagefig{
@@
 }{%
-  Schematic representation of the Woody Classifier Based Latency
+  Schematic representation of the Woody Classifier-Based Latency
   Estimation test phase. (1) The
@@
 first evaluate our approach with synthetic data.
-In \autoref{sec:covert-align}, the proposed algorithm will be applied to real
+In \cref{sec:covert-align}, the proposed algorithm will be applied to real
 \ac{eeg} data.
@@
 left hemisphere to the scalp electrodes using a boundary element method forward
-model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
-Pink temporal noise is generated by passing gaussian noise through an infinite
+model~\cite{Mosher1999} and MNE's \texttt{fsaverage} source space and anatomy.
+Pink temporal noise is generated by passing Gaussian noise through an infinite
 impulse response filter with transfer function
@@
 cosine distances between electrode positions and scaling it by the inverse of
-the resulting spatial.
-We refer to~\textcite{Gramfort2014} for implementation details.
-%Evoked potentials are simulated for three different source time course waveforms,
-%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
-%defined by the following formulas:
-%\begin{equation}
-%  f_\text{gaussian}(t,l) = a
-%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
-%\end{equation}
-%\begin{equation}
-%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
-%\end{equation}
+the resulting spatial covariance.
+We refer to \textcite{Gramfort2014} for implementation details.
+
 The source time course is defined by the following function:
@@
 \end{equation}
-over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
+over time $t$, with latency $l$, amplitude $a=\num{1e{-7}}$V, and frequency
 $f=4$.
-%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
-%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
-%For the sine pulse, the frequency $f=4$
 
-%For each of these,
 100 target epochs containing the evoked potential and noise and 100
@@
 deviations of respectively $\sigma=0.1,0.2,0.3$s.
-The noise was scaled at different 32  different noise levels, with the \ac{snr} of a
+The noise was scaled at 32 different noise levels, with the \ac{snr} of a
 target epoch ranging from 0 to -31dB, with
@@
 
-%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
-A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
-%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
-%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
+A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
 \fullpagefig{%
-  %\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
   \sffamily
@@
 }{%
-  Simulated target (bottom half of epochs) and non-target (top half of epoch) evoked data
-  and their average (red: target, yellow: noon-target, shaded are: standard
+  Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
+  and their average (red: target, yellow: non-target, shaded area: standard
   deviation).
   Targets contain a sine wave pulse jittered by $\sigma$. 32 EEG channels with pink
-  spatiotemporal noise were simulated using a forward head model, channel C3 is
+  spatiotemporal noise were simulated using a forward head model, and channel C3 is
   plotted here.
-
 }{fig:wcble/sim-sine}
-\todo{label x axis}
 
@@
 We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
-estimation task using the pearson correlation coefficient $\rho$ at different
+estimation task using the Pearson correlation coefficient $\rho$ at different
 jitter and \ac{snr} levels.
-For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
+For both \ac{cble} and \ac{wcble}, target latencies were extracted in a
 10-fold cross-validation scheme using the robust median latency method after
@@
 \Ac{wcble}'s maximum iteration number was set to 64.
-Results are presented in Figure~\ref{fig:wcble/results/latency}.
-A 95\ confidence intervals on each measure was obtained through bootstrapping
+Results are presented in \cref{fig:wcble/results/latency}.
+A 95\% confidence interval on each measure was obtained through bootstrapping
 with 1000 permutations.
@@
   latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
-  target data, measured as pearson correlation coefficient $\rho$ between
+  data, measured as Pearson correlation coefficient $\rho$ between
   estimated and ground truth latencies.
@@
 At around -28dB, \ac{wcble} does not significantly outperform \ac{cble}
-anymore and both methods get overtaken by the strong presence of noise.
+anymore, and both methods get overtaken by the strong presence of noise.
 
@@
 \Ac{cble} and \ac{wcble} classifiers were implemented as described
-in~\ref{sec:wcble/methods}, with both the wavelet transformed score time series
-and square transformed latencies as input to a second-stage logistic regression
+in \cref{sec:wcble/methods}, with both the wavelet-transformed score time series
+and square-transformed latencies as input to a second-stage logistic regression
 classifier with $L_2$-norm regularization ($C=0.2$).
@@
   \caption[Synthetic decoding results]{Decoding performance, measured as binary
-  classification accuracy, for \ac{cble}, \ac{wcble} and their first stage
+  classification accuracy, for \ac{cble}, \ac{wcble}, and their first stage
   classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
@@
 
-\autoref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble} decreases with
-$\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
-This difference is particularily large when $\sigma$ is higher ($\sigma=0.2$
+\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble} decreases with
+$\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
+This difference is particularly large when $\sigma$ is higher ($\sigma=0.2$
 and $\sigma=0.3$) and \ac{snr} is not yet high.
-\Ac{wcble} always scored better than \ac{cble} and \ac{wcble}, except for
-$\sigma=0.1$ and $\text{\ac{snr}} <27\text{db}$, at which point both \ac{cble}
+\Ac{wcble} always scored better than \ac{cble} and \ac{tlda}, except for
+$\sigma=0.1$ and $\text{\ac{snr}} < 27 \text{dB}$, at which point both \ac{cble}
 and \ac{wcble} performed at chance level.
@@
 \section{Discussion \& conclusion}
+
 In this work, we introduced a new \ac{erp} latency estimation and decoding
@@
 Our method attempts to model the spatial and temporal activation of
-each \ac{erp} component, as well as it's latency distribution.
-Through the firs-stage classifier, it takes into acount the
-spatial and temporal structure of EEG background noise and the trials from the
+each \ac{erp} component, as well as its latency distribution.
+Through the first-stage classifier, it takes into account the
+spatial and temporal structure of \ac{eeg} background noise and the trials from the
 classes to discriminate.
@@
 The key finding of the simulation study presented
-in~\ref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
+in \cref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
 potential jitter and lower \ac{snr} than \ac{cble}.
@@
 decoding in those cases where jittered \ac{erp} data is of interest.
-Reason that accuracy latency estimation plateaus and accuracy drops for cble,
-but that both plateau for wcble, is because of the alignment
-As  the noise increases beyond a certain threshold, the improvement vanishes.
-At this point, both methods perform poorly in both tasks, so if latencies
-cannot be properly estimated, no increase in classification performance can be
-gained by aligning in \ac{wcble}.
+At a certain noise threshold, both methods begin to fail, but \ac{wcble}
+delivers higher accuracy at intermediate \ac{snr} levels than \ac{cble}, likely due to the iterative alignment process.
 
-As shown, the proposed method is suitable for \ac{bci} decoding setting
+In general, noise beyond a certain level will overwhelm both \ac{cble} and
+\ac{wcble} performance, making it impossible to accurately estimate latencies
+or perform meaningful classification.
+When latencies cannot be properly estimated, aligning epochs using \ac{wcble} does not help to improve classification.
+
+As shown, the proposed method is suitable for \ac{bci} decoding settings
 as well as single-trial \ac{erp} latency analyses.
@@
 metric representing the presence of a template response in the signal.
-A suiting candidate is the spatiotemporal
-\ac{lcmv}-beamformer~\cite{VanVliet2015} described in
-\autoref{sec:stbf-struct}.
+A suitable candidate is the spatiotemporal
+\ac{lcmv}-beamformer~\cite{VanVliet2015}, described in
+\cref{sec:stbf-struct}.
 The activation pattern can be set to the average of the set of trials to
@@
 Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
-problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
-Secondly, this method has a synergy with CBLE since both make the same
-assumption about the short-time stationary of the EEG background noise
+problem~\cite{Sosulski2022, VanDenKerchove2022}, benefiting decoding performance.
+Secondly, this method has a synergy with \ac{cble} since both make the same
+assumption about the short-time stationarity of the \ac{eeg} background noise
 within an epoch.
-CBLE does not retrain the first-stage classifier for each time shift, but
+\Ac{cble} does not retrain the first-stage classifier for each time shift but
 rather trains it once within the given window.
 After training, the classifier parameters represent some information about the
-expected ERP waveform and background noise.
+expected \ac{erp} waveform and background noise.
 By applying the trained classifier to different time shifts, it assumes this
-ERP waveform can be shifted in time, but since the classifier's information about
-the background noise was only obtained from the initial window, CBLE assumes
+\ac{erp} waveform can be shifted in time, but since the classifier's information about
+the background noise was only obtained from the initial window, \ac{cble} assumes
 its properties do not vary throughout the epoch.
@@
 
-The main limitation of the proposed method is its supposed lack to handle
+The main limitation of the proposed method is its lack of capacity to handle
 multi-component \ac{erp} data.
 In real \ac{erp} analysis settings, multiple components are usually present,
-and each component has a distinct contribution to class discriminatbility and
+and each component has a distinct contribution to class discriminability and
 latency distribution.
@@
 Current work focuses on extending \ac{wcble} to a multi-component setting, on
-the one hand to performance latency estimation and decoding performance on real
-\ac{erp} data and on the other hand to apply it as an \ac{erp} component
-separation method in the fashion of~\textcite{Ouyang2017}.
+the one hand to improve latency estimation and decoding performance on real
+\ac{erp} data, and on the other hand to apply it as an \ac{erp} component
+separation method in the fashion of \textcite{Ouyang2017}.
 
-The proposed implementation in theory also supports processing multi-class
+The proposed implementation theoretically supports processing multi-class
 data, but the convergence and correctness of the \ac{wcble} solution is yet to
 be properly studied for this case.
-
-\clearpage
-\begin{subappendices}
-\section{\Acs{cble} algorithm}
-\label{sec:wcble/app/cble-alg}
-\begin{algorithm}[h]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
-  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\State $\theta \gets
-			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
-      \algorithmiccomment{Train stage 1}
-		\For{$n=1\ldots N$}
-      \algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\theta,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}
-      \algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-
-  \caption{\Ac{cble}}
-	\label{alg:cble}
-\end{algorithm}
-
-\clearpage
-\section{\Acs{wcble} algorithm}
-\label{sec:wcble/app/wcble-alg}
-\begin{algorithm}[h]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
-		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
-	\begin{algorithmic}[1]
-		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
-    N$
-      \algorithmiccomment{Train stage 1}
-		\Repeat
-		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
-		\For{$n=1\ldots N$}
-    \State  $s_n \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
-		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
-		\EndFor
-    \Until{convergence or maximum iterations reached}
-    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets
-      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-
-	\end{algorithmic}
-	\textbf{Output:} $\theta^*,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets
-      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-  \caption{\ac{wcble}}
-	\label{alg:wcble}
-\end{algorithm}
-
-\end{subappendices}

### 2024-10-09 chapters/abstract.tex
Comments François patients

@@
 or eye-motor dysfunction, face substantial challenges when using traditional
-gaze-dependent brain-computer interfaces \acp{bci}.
+gaze-dependent \acp{bci}.
 These systems require users to direct their gaze toward specific targets, a
@@
 The proposed methods were validated in experiments involving both healthy
-individuals and 7 individuals with severe physical impairment and impaired eye-motor control.
-Data were recorded from a control group, as well as from seven individuals with
-gaze impairments.
+individuals and seven individuals with severe physical impairment and impaired eye-motor control.
 These experiments demonstrated the robustness of the novel decoding methods,
@@
 
-\paragraph{Keywords:} brain-computer interface, gaze independence, covert
+\paragraph{Keywords:} \acl{bci}, gaze independence, covert
 visuospatial attention, event-related potentials, linear decoding, multilinear

### 2024-10-09 chapters/abstract_fr.tex
Comments François patients

@@
 \chapter*{R\'esum\'e}
-Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de \ac{lis} ou de dysfonctionnement oculomoteur, rencontrent des défis considérables lors de l'utilisation d'interfaces cerveau-ordinateur \acp{bci} traditionnelles.
-Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
-Ce travail aborde cette limitation en développant des méthodes de \ac{bci} indépendantes du regard, en se concentrant sur l'amélioration de la \ac{vsa} cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
-Une contribution clé de ce travail est la compensation pour le jitter de latence ERP, une variabilité qui impacte négativement la performance de décodage dans la \ac{vsa} cachée.
-En gérant ce jitter, la méthode proposée améliore la précision de la communication sans nécessiter de contrôle du regard, rendant les \acp{bci} plus utilisables pour les personnes ayant des handicaps moteurs.
+Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
+syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
+lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
+Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
+cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
+Ce travail vise \`a palier cette limitation en développant des méthodes de BCI
+indépendantes du regard, en se concentrant sur l'amélioration de la attention visuo-spatiale (VSA)
+cachée,
+où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements
+oculaires correspondants.
+Une contribution clé de ce travail est la compensation pour la compensation de
+la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
+la performance de décodage dans la
+VSA cachée.
+En gérant cette variabilite, la méthode proposée améliore la précision de la communication
+sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.
 
-Au-delà de l'indépendance du regard, le travail fait également progresser les techniques générales de décodage \ac{erp} en affinant la structure des estimateurs linéaires et multilinaires.
-Ces méthodes améliorent la précision de classification dans une gamme de conditions \ac{bci}, en particulier lorsque les données d'entraînement sont limitées.
-L'introduction de régularisation structurée dans les modèles linéaires et multilinaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
+Au-delà de l'indépendance du regard, le travail fait également progresser les
+techniques générales de décodage ERP en affinant la structure des estimateurs
+linéaires et multilinéaires.
+Ces méthodes améliorent la précision de classification dans une gamme de
+conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
+L'introduction de régularisation structurée dans les modèles linéaires et
+multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
 Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.
 
-Les méthodes proposées ont été validées lors d'expériences impliquant à la fois des individus en bonne santé et 7 individus avec des handicaps physiques sévères et un contrôle oculomoteur altéré.
-Les données ont été enregistrées à partir d'un groupe de contrôle, ainsi que de sept individus ayant des impairments du regard.
-Ces expériences ont démontré la robustesse des nouvelles méthodes de décodage, montrant que le système pouvait efficacement décoder l'attention cachée même lorsque le regard direct était impossible.
+Les méthodes proposées ont été validées lors d'expériences impliquant à la fois
+des individus en bonne santé et sept individus avec des handicaps physiques sévères et un contrôle oculomoteur altéré.
+Ces expériences ont démontré la robustesse des nouvelles méthodes de décodage, montrant que le système pouvait efficacement décoder l'attention cachée même lorsque l'observation directe de la cible  était impossible.
 
-\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéraire, handicap moteur, handicap du regard.
+\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéaire, handicap moteur, handicap du regard.

### 2024-10-09 chapters/acknowledgment.tex
Comments François patients

\chapter*{Acknowledgments}
We thank Arno Libert,
Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
for the works presented in \cref{sec:covert-align} and
\cref{sec:patients}.
We would also like to thank the other colleagues at the KU Leuven Computational
Neurosciences group at KU Leuven and the BCI team at the University of Lille,
for their general support.
Furthermore, we thank the teams at TRAINM Neuro Rehab Clinics Antwerp,
the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in~\cref{sec:bttda}.
We thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.

### 2024-10-09 chapters/bci.tex
Comments François patients

@@
 \label{sec:bci}
-\todo{search for ??}
-\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
-\todo{fix all latex warnings to make sure no refs are missing}
-\todo{uniform ref or cref}
 \todo{consistently terminology individuals with severe
-speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
-\todohsm{more subtitles}
+speech and physical impairments (SSPI)}
 \todo{uniform vsa condition vsa setting}
 \todo{are all acronyms defined/Acro corectly used in chapters?}
-\todo{acronyms written in full every new chapter and in captions?}
-\todo{are textcites used when necessary?}
-\todo{fix lacheck/chktex notifications}
 \todo{footnot skips a page in wcble chap}
@@
 
-The key problem for severely paralyzed patients is that the mind wants to go where the
+The key problem for severely paralyzed individuals is that the mind wants to go where the
 body cannot follow.
@@
 By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
-be used as assistive communication technology by patients with impaired
+be used as assistive communication technology by individuals with \acl{sspi}.
 speech~\cite{Wolpaw2018,Milekovic2018}.
 Impressive cutting-edge experimental examples include a brain-spine interface allowing a
-paraplegic patient to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
+paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
 speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
@@
 Ultimately, they have always been especially promising as assistive technology for
-paralyzed patients limited in their communication ability.
+people limited in their communication ability.
 It is this \ac{bci}-controlled communication where our focus lies, as it offers a
@@
 imaging (fMRI) to more portable technologies like acoustic signals obtained by
-functional ultrasound imaging (fUS)~\cite{Zheng2023} and near-infrared spectroscopy
+functional ultrasound imaging (fUS)~\cite{Zheng2023} and functional near-infrared spectroscopy
 (fNIRS)~\cite{Borgheai2020}.
@@
 The \emph{invasiveness} of the technology forms an important concept in determining its
-suitability for a given application and patient.
+suitability for a given application and user.
 As illustrated in \cref{fig:bci/recording}, invasive electrodes can often measure a
@@
 	There is a trade-off between invasiveness and spatial resolution.
-}{%
-  fig:bci/recording
-}
+}{fig:bci/recording}
 
@@
 Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
-Depth electrodes such as those used in stereo-\ac{eeg} \ac{bci}~\cite{Wu2024} tasks and
-closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
+Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
+ and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
 symptoms~\cite{Arlotti2018}.
@@
 Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
-great temporal resolution up to thousands of Hz.
+great temporal sampling resolution up to thousands of Hz.
 \Ac{eeg} measures the electrical potential over time at a given electrode relative to a
@@
 require a trained individual to properly administer.
-Nevertheless, it is the most accessible \ac{bci} technology for both patients and
+Nevertheless, it is the most accessible \ac{bci} technology for both users and
 researchers.
@@
 Despite its noisy nature, \ac{eeg} is the recording methodology of choice for our
-\ac{bci} because of its wide acceptance by patients.
+\ac{bci} because of its wide acceptance by users.
 To counteract the noise present in our recording, we must on the one hand evoke strong,
@@
   Adapted from~\textcite{Muehl2014}.
-}{%
-  fig:bci/paradigms
-}
+}{fig:bci/paradigms}
 
@@
 application.
-The number of output degrees of features, the serial or parallel nature of performed
+The number of output degrees of freedom, the serial or parallel nature of performed
 tasks, and whether the paradigm is active or reactive, all need to be taken into
@@
 First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
-has been proven to work for patients in day-to-day home use~\cite{Wolpaw2018}.
+has been proven to work for those with \acl{sspi} in day-to-day home use~\cite{Wolpaw2018}.
 
@@
     channel.}
+    \label{fig:bci/erp}
   \end{subfigure}%
@@
 Interesting and complex challenges arise when applying these concepts outside of the lab
-in augmented and assistive communication technologies for patients with \acp{sspi}.
+in augmented and assistive communication technologies for users with \acl{sspi}.
 
@@
 This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
-The \ac{bci} target population consists of individuals who suffer from various degrees
-of paralysis, or even from \ac{lis}, the complete loss of muscle control with preserved
+The \ac{bci} target population consists of individuals with severe physical
+impairment due to  with various degrees
+of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
 consciousness.
 While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
-studies paradoxically often report poor performance in patient populations precisely due
-to the lack of adequate eye motor control.
+studies paradoxically often report poor performance in exactly this group,
+precisely due to their lack of adequate eye motor control.
 Equally often, participants with gaze impairments were excluded from the study.
@@
 This points to the need for gaze-independent solutions, which we will explore further in
-\cref{sec:gaze-independence} and the rest of this thesis.
+\cref{sec:gaze-independent} and the rest of this thesis.

### 2024-10-09 chapters/bttda.tex
Comments François patients

@@
 \input{chapters/bttda/tensorviz.tex}
-\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}}.
+\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
 
@@
 \begin{figure*}[t]
+  \centering
+  \makebox[\linewidth][c]{
 	  \input{figures/bttda/blocks.pgf}
+  }
   \caption[Analysis of NMSE and classification score per block.]{%
@@
 between- or within-block feature correlation, decreasing discriminability, etc.
-Finally, the presented results are limited by our artificial restriction
-that $(r_1=r_2=\ldots=r_k)$ to reduce the computational demand of the presented
-experiments.
+Finally, the completeness of the presented results is limited by the artificial
+restriction in hyperparameter choice.
+We imposed $(r_1=r_2=\ldots=r_k)$ that to reduce the computational demand of
+the performed experiments.
 In a sense, this goes against the proposition of increased model flexibility.

### 2024-10-09 chapters/conclusion.tex
Comments François patients

@@
 Together, these results show that there is an interest in developing a new class of ERP-BCI
-interfaces for patients that prefer to choose their own gaze strategy, to
+interfaces for users that prefer to choose their own gaze strategy, to
 avoid the effort of redirecting their eye gaze to different spatial locations
@@
 \acp{sspgi}.
-We aim to revisit the patients with an on-line experiment and comfort
+We aim to revisit the participants with an on-line experiment and comfort
 assessment to establish the usability of the proposed interface and compare its
@@
 their skills and abilities.
-Longitudinal research should follow several patient groups, specifically those with gaze
+Longitudinal research should follow several user groups, specifically those with gaze
 impairment, and experiment with specific paradigms, interfaces, and decoders for

### 2024-10-09 chapters/conflict.tex
Comments François patients

@@
-\chapter*{Conflict of interest}
+\chapter*{Conflict of interest \& funding}
 The authors declare no conflict of interest.
+
+Arne Van Den Kerchove is funded by a bilateral cooperation program between KU
+Leuven (special research fund of the KU Leuven GPUDL/20/031) and the university
+of Lille under the Global PhD Partnership program.
+Prof. Marc Van Hulle is supported by research grants received from
+the European Union’s Horizon Europe Marie Sklodowska-Curie Action program
+(grant agreement No. 101118964), the European Union’s Horizon 2020 research and
+innovation program (grant agreement No. 857375), the special research fund of
+the KU Leuven (C24/18/098), the Belgian Fund for Scientific Research – Flanders
+(G0A4118N, G0A4321N, G0C1522N), and the Hercules
+Foundation (AKUL 043).

### 2024-10-09 chapters/contribution.tex
Comments François patients

@@
 Dr. Arno Libert and Dr. Benjamin Wittevrongel consulted on the analyses performed in
-\autoref{sec:stbf-struct}
+\cref{sec:stbf-struct}
 Data collection was performed by Arne Van Den Kerchove for the dataset
-presented in \autoref{sec:covert-align} and by Arne Van Den Kerchove and
-Juliette Meunier for the dataset presented in \autoref{sec:patients}.
+presented in \cref{sec:covert-align} and by Arne Van Den Kerchove and
+Juliette Meunier for the dataset presented in \cref{sec:patients}.
 Participant recruitment at TRAINM was performed by Alixe Willemssens under
 supervision of dr. Edward Schietekatte, at Fondation Partage et Vie by Marie de
-Moura under supervision of prof. Etienne Allart and at UZ Leuven Neuromuscular reference
-center under supervision of prof. Kristl Claeys and prof. Philip Van Damme.
-Prof. Etienne allart provided clinical consultation.
+Moura under supervision of prof. Etienne Allart and at the UZ Leuven Neuromuscular
+Reference Center under supervision of prof. Kristl Claeys and prof. Philip Van Damme.
+Prof. Etienne Allart provided clinical consultation.

### 2024-10-09 chapters/covert_align.tex
Comments François patients

@@
 \label{sec:covert-align}
-\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}}
+\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}.}
 
@@
 They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
-step-wise Linear Discriminant Analysis  (SWLDA) for 6 ALS patients, whose SWLDA
-performance also suffered from jitter.
+step-wise Linear Discriminant Analysis  (SWLDA) for 6 participants with
+\ac{als}, whose SWLDA performance also suffered from jitter.
 Note that interpretation of this comparison may be challenging due to differences in
@@
 selection, performing covert or split \ac{vsa} continuously still requires
-sustained gaze fixation, which might not be possible for some patients who
+sustained gaze fixation, which might be impaired in those who
 could benefit from such an application.
@@
 We introduce an additional setting of interest here, namely a combination of \ac{vsa}
-conditions, which represents those cases where patients cannot
+conditions, which represents those cases where the user cannot
 redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
@@
 This is advantageous for the development of a class of \ac{erp}-\ac{bci} interfaces for
-patients who prefer to rest their gaze on a fixation cross on the screen,
+users who prefer to rest their gaze on a fixation cross on the screen,
 avoiding the effort of redirecting their eye gaze for every selection.
@@
 \ac{vsa} ($d=3$) and between-\ac{vsa} condition transfer settings are promising for
-patients with even less eye motor control who may experience involuntary eye
+users with even less eye motor control who may experience involuntary eye
 movements or fixation fatigue and hence cannot keep their gaze fixed throughout an
@@
 
-\cite{Zisk2022} addressed P3 jitter in ALS patients by augmenting the training data once
-with time-shifted copies based on \ac{cble}-estimated jitter.
+\cite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
+training data once with time-shifted copies based on \ac{cble}-estimated jitter.
 While we aim to train the first-stage classifier without the effects of jitter
@@
 training data.
-As their focus was on ALS patients and inter-session stability, they
+As their focus was on \ac{als} and inter-session stability, they
 did not assess how their method interacts with visuospatial attention.
@@
 performance over time by controlling attention or gaze.
-Finally, since this work was conducted with patient applications in mind, we
+Finally, since this work was conducted with applications for individuals with
+\ac{sspgi} in mind, we
 should highlight that the gaze of participants in the conducted experiments was
-cued and fixed, which is per definition impossible for the patients we consider.
-We are currently conducting a patient study to further investigate whether the
-studied \ac{vsa} conditions are appropriate and to what extent they occur in
-natural patient \ac{bci} operation.
+cued and fixed, which is per definition impossible for the end user group we consider.
+In \cref{sec:patients}, we present a study to further investigate whether the
+studied \ac{vsa} conditions are appropriate and to what extent they occur when
+individuals with \ac{sspgi} operate a \ac{bci}.
 
@@
 Overt decoding performance was not improved, but our proposed method can
-provide added value for patients who are unable to operate a visual \ac{bci} in overt
+provide added value for users who are unable to operate a visual \ac{bci} in overt
 attention mode.
-Later studies should confirm whether our findings hold in patient
-populations suffering from a variety of eye-motor impairments and develop a
+Later studies should confirm whether our findings hold in individuals with
+severe physical impairment and a variety of eye-motor impairments, and develop a
 solution that is capable of properly handling multiple non-time-locked \ac{erp}
 components.
-
-\clearpage
-\begin{subappendices}
-\section{\Acs{cble} algorithm}
-\label{sec:wcble/app/cble-alg}
-\begin{algorithm}[h]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
-  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\State $\theta \gets
-			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
-      \algorithmiccomment{Train stage 1}
-		\For{$n=1\ldots N$}
-      \algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\theta,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}
-      \algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-
-  \caption{\Acs{cble}}
-	\label{alg:cble}
-\end{algorithm}
-
-\clearpage
-\section{\Acs{wcble} algorithm}
-\label{sec:wcble/app/wcble-alg}
-\begin{algorithm}[h]
-	\textsc{Train}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
-		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
-	\begin{algorithmic}[1]
-		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
-    N$
-      \algorithmiccomment{Train stage 1}
-		\Repeat
-		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
-		\For{$n=1\ldots N$}
-    \State  $s_n \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
-		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
-		\EndFor
-    \Until{convergence or maximum iterations reached}
-    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{n,s}^\mathrm{train} \gets
-      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
-		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-
-	\end{algorithmic}
-	\textbf{Output:} $\theta^*,
-  \mat{Y}^\mathrm{train},
-  \mathbf{s}_\mathrm{target}^\mathrm{train},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
-
-	\smallskip \hrule \smallskip
-	\textsc{Evaluate}
-	\smallskip \hrule \smallskip
-	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
-  \begin{algorithmic}[1]
-		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
-		  \For{$s=1\ldots R$}
-      \State $y_{m,s}^\mathrm{test} \gets
-      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
-        \EndFor
-        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
-		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
-			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
-    \EndFor
-	\end{algorithmic}
-  \textbf{Output:} $\mat{Y}^\mathrm{test},
-  \mathbf{s}_\mathrm{target}^\mathrm{test},
-  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
-  \caption{\Acs{wcble}}
-	\label{alg:wcble}
-\end{algorithm}
-\end{subappendices}

### 2024-10-09 chapters/gaze_independence.tex
Comments François patients

@@
 \subsection{Incidence of eye motor impairment}
+\label{sec:gaze-independent/gaze-dependence/incidence}
 \Ac{sspi} is often caused by damage to the central or peripheral nervous
@@
 \ac{tbi}).
-Many individuals in these groups unfortunately also suffer
-to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
+Many individuals in these groups are unfortunately also afflicted by some form
+of eye motor impairment, requiring \acp{bci} adapted to their condition.
 \Cref{tab:incidence} reports the relatively high frequencies of
@@
   }
-  \caption[Incidence of eye motor impairment in selected \acs{bci} target
-    patient populations.]{%
-      Incidence of eye motor impairment in selected \ac{bci} target
-      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
+  \caption[Incidence of eye motor impairment in selected \acs{bci} user target
+    populations.]{%
+      Incidence of eye motor impairment in selected \ac{bci} user target
+      populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
       \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
@@
 
-Among the most severely affected patients are stroke
-patients~\cite{Pollock2011,Rowe2019}, mostly
-due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
-These patients might suffer from severe or complete eye motor impairment
-from the onset of their condition.
-Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
+Among the most affected are those recovering from stroke~\cite{Pollock2011,Rowe2019},
+and most of all those with due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
+Stroke can lead to the most severe severe or even complete eye motor impairment
+from the onset of their condition, resulting in the~\ac{lis}.
+However, case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
 with \ac{lis} due to stroke, the group with a complete lack of eye motor
@@
 
-In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
-are also fairly common.
+\ac{als} is progressive disease affecting motor neurons.
+This initially results in general weakness and loss of muscle tone, but
+eventually leads to full body paralysis.
 Although eye movement is often cited as one of the longest preserved
 capabilities in \ac{als}, studies show that minor issues are still
-fairly common, especially for patients with bulbar onset
-\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
-Furthermore, \textcite{Hayashi1991} show that as these patients progress past
+fairly common~\cite{Kang2018, Guo2022,Moss2012}.
+The bulbar-onset variant is characterized by an early loss of speech
+and increased involvement of eye motor symptoms~\cite{Guo2022}.
+Furthermore, \textcite{Hayashi1991} show that as \ac{als} progresses past
 the point of independent breathing, symptoms will eventually also involve
 eye muscle paralysis.
+One of the goals of \ac{bci} has always been to support these individuals to
+ensure quality of life.
 
@@
 
-\Ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
-This almost always affects eye
-movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
+\Ac{fa} is a neurodegenerative disease affecting the
+spinal cord, peripheral nervous system, and cerebellum, resulting in an
+impairing loss of muscle coordination.
+This almost always heavily affects eye
+movements~\cite{Fahey2008,Hocking2010,Furman1983,Cook2017}, with various forms of
 involuntary movements and trouble pursuing or fixating on targets.
@@
 \ac{bci} and neurological literature.
-Some definitions include only tetraplegic patients without eye movements used
+Some definitions include only those with tetraplegy without eye movements used
 for communications.
@@
 movements or minor motor output.
-While some definitions only include stroke or \ac{tbi} patients with damage to
+While some definitions only include stroke or \ac{tbi} with damage to
 specific regions in the brain (midbrain, brainstem, or
@@
 unreliable, or easily fatigued.
-This group is much larger than the first.''~\cite{Wolpaw2008}}}
+This group is much larger than the first.''~\cite{Wolpaw2006}}}
 
@@
 conditions, or even the exact symptoms.
-In clinical reality, every patient has different symptoms, which lead to their own
+In clinical reality, every \ac{bci} user has different symptoms, which lead to their own
 unique set of preserved capabilities and visual skills.
 From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
-Instead of categorizing patients, we will refer to those who might benefit from
+Instead of categorizing users by their etiology, we will refer to those who might benefit from
 \ac{bci} assistive communication technology as having \ac{sspi}, in line with
@@
 
-The most severely impaired of these patients constitute the \ac{lis}
-patient group, which forms one of the main \ac{bci} interest groups.
+The most severely impaired of the user groups above constitute the \ac{lis}
+group, which forms one of the main \ac{bci} interest groups.
 In this work, we use the term \ac{lis} for a situation of (near)
@@
 
-To summarize, for patients with eye motor impairment, gazing directly at a screen target may
+To summarize, for individuals with \ac{sspgi}, gazing directly at a screen target may
 be uncomfortable, impractical, or even impossible.
@@
 This specific target is embedded within a broader goal of enabling
-effective communication for eye-motor impaired \ac{lis} patients.
+effective communication for those with \ac{sspgi}.
 Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
@@
 After collecting data from healthy participants and developing initial decoders, we conducted
-a patient study with gaze-impaired individuals to evaluate the impact of eye motor deficits
+a study with seven individuals with \ac{sspgi} to evaluate the impact of eye motor deficits
 on \ac{bci} performance.
-This study focused on individuals with \ac{sspgi}, recruited from neurorehabilitation centers
+This study recruited participants from neurorehabilitation centers
 and specialized care homes, including participants with conditions such as \ac{als}, \ac{fa},

### 2024-10-09 chapters/patients.tex
Comments François patients

@@
 Eye motor impairments are presumed to reduce performance in operating visual
-oddball \ac{bci}s (see \cref{sec:gaze-independence} for an overview), since users
+oddball \ac{bci}s (see \cref{sec:gaze-independent} for an overview), since users
 cannot comfortably redirect their gaze at the desired target,
@@
 impaired are very few.
-Results are usually different than those obtained with healthy control
+Results are usually different from those obtained with healthy control
 participants in the lab, due to difference in capabilities, brain
@@
 system showed high accuracy, outperforming an eye tracking alternative.
-It would be of interest to verify if such results can be replicated with other
-patient groups and visual oddball \acp{bci}.
+It would be of interest to verify if such results can be replicated with
+participants with other conditions, and with a visual oddball \acp{bci}.
 
@@
 We showed gaze-independent performance can be improved in healthy subjects
-using a suited decoding strategy.
-That accounts for latency jitter in covert \ac{vsa} responses.
+using a suited decoding strategy that accounts for latency jitter in covert \ac{vsa} responses.
 Yet, there is a strong need for verification of these results in an applied
@@
 Ultimately, 7 participants were retained.
-Of these, one participant was diagnosed with bulbar-onset \ac{als}.
-\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
-progressive loss of motor function.
-This initially results in general weakness and loss of muscle tone, but
-eventually leads to full body paralysis.
-Although speech and especially eye movements are usually preserved until the
-later stages of the disease progression, the bulbar-onset variant is
-characterized by an early loss of speech and an increased involvement of eye
-motor symptoms~\cite{Guo2022}.
-Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
-whose life span has been extended with life support and assistive technology to
-ensure quality of life.
-In these very progressed stages of \ac{als}, eye movement will eventually also
-be affected~\cite{Hayashi1991}.
-
-Three other participants were diagnosed with \ac{fa}, a neurodegenerative
-disease affecting the
-spinal cord, peripheral nervous system, and cerebellum.
-\ac{fa} results in an impairing loss of muscle coordination.
-In progressed stages, the disease can present with nystagmus, saccadic
-intrusions, and gaze dysmetria~\cite{Cook2017}.
-
-The final three participants were recovering from stroke.
-Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
-various eye motor disorders are common depending on the exact etiology
-~\cite{Bogousslavsky1987, Moncayo2009}.
-Stroke can lead to some of the more severe impairments like ophthalmoplegia or
-partial ophthalmoplegia immediately from onset.
-
+Of these, one participant was diagnosed with bulbar-onset \ac{als}, three with
+\ac{fa} an three were recovering from stroke.
+We refer to \cref{sec:gaze-independent/gaze-dependence/incidence} for a short
+description of these conditions.
 Table~\ref{tab:patients/patients} lists the included participants and their
@@
 binocular vision, and field of vision.
-Additionally, participants were asked about eye tremors (nystagmus or other) and
-other involuntary eye movements.
+Additionally, participants and their caregivers were asked about eye tremors
+(nystagmus or other) and other involuntary eye movements.
 
@@
 Participant PC4 had one deviated eye with a corneal abscess affecting the motility
-and vision in the right eye, and suffered reduced motility in the left.
+and vision in the right eye, and reducing motility in the left.
 
@@
 \end{table}
+\todo{where these values corrected for distance?}
 
@@
   near the middle of the screen, and PC4 near the bottom.
-}{%
-  fig:patients/gaze
-}
+}{fig:patients/gaze}
 
@@
 Eye tracker calibration did not succeed for subject PB4, but given
-transformations of gaze positions to the stimulus space, they were assumed to be overtly performing the task.
+transformation of gaze positions to the stimulus space, they were assumed to
+be overtly performing the free task.
 

### 2024-10-09 chapters/stbf_struct.tex
Comments François patients

@@
   \emph{This chapter was
-  published as~\textcite{VanDenKerchove2022}}.
+  published as~\textcite{VanDenKerchove2022}.}
 
@@
 	\end{equation}
-
 	These weights minimize the variance of the output of the filter while enhancing
@@
 	epochs, determined as follows:
-
 	\begin{equation}
@@
 	\cref{eq:minimum_variance} and \cref{eq:linear_constraint} as:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	The beamformer can be applied to epochs (unseen or not) as:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	resulting in a scalar output per epoch.
@@
 	between target and non-target classes, in analogy to linear discriminant
-  analysis~\cite{Treder201}, it does not always perform well on unseen data.
+  analysis~\cite{Treder2016}, it does not always perform well on unseen data.
 	The main challenge is to find a good estimator for the inverse covariance
@@
 	covariance and inverse covariance calculated as follows:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	\begin{equation}
@@
 	\end{equation}
-
 	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when
@@
 	The empirical estimator suffers from performance and
-	stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.
+	stability issues if the number of epochs $N$ used or estimation is not much larger than the number of features $CS$~\cite{Stein1956,Khatri1987}.
 
@@
 	The estimators for the covariance and inverse covariance are given by:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	with $0<\alpha<1$.
@@
 	estimate:
-
 	\begin{equation}
@@
 	by:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	with $\otimes$ the Kronecker product operator.
@@
 	time samples)~\cite{Bijma2003}, it is assumed to have a Toeplitz-matrix structure:
-
 	\begin{equation}
@@
 	\end{equation}
-
 	\Cref{prop:inverse_kronecker} then leads to
@@
 	covariance.
+
 	\begin{property}
@@
     \hat{\mat{T}}^+$.
-	\Cref{fig:kronlda-covs}e and \cref{fig:covs}f respectively show examples of the
+	\Cref{fig:kronlda-covs}e and \cref{fig:kronlda-covs}f respectively show examples of the
 	structured covariance and inverse covariance estimators,
@@
 	\end{subequations}
-
 	The shrinkage parameters $0<\beta_{k+1}<1$ and $0<\gamma_{k+1}<1$ should be
@@
 	\cref{sec:shrunk_covariance}, and \ac{stbf} with
-	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct} )with \ac{loocv} shrinkage for
+	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct}) with \ac{loocv} shrinkage for
 	the Kronecker factors as in \cref{sec:structured_covariance}.
@@
 	in
-  \crefrange{eq:fpi_spatial}{eq:fpi-temporal-norm}
+  \crefrange{eq:fpi_spatial}{eq:fpi_temporal_norm}
   estimates the Kronecker-Toeplitz structured covariance for the
@@
 	corresponding number of blocks in the training folds and dropping the rest.
-	\cref{fig:accuracy} and \cref{fig:accuracy_ap} show each classifier's
+	\Cref{fig:accuracy} and \cref{fig:accuracy_ap} show each classifier's
 	accuracy relative to the data availability.
@@
 	\end{figure}
+
+  \fullpagefig{%
+		\includegraphics[width=\textwidth]{figures/stbf_struct/accuracy_app.eps}
+  }{%
+    Accuracy averaging over different numbers of trials.
+  }{%
+    Accuracy of the different classifiers for all 21 subjects relative to the
+		number of blocks available for training. One block consists of 135
+		epochs and corresponds to 27 seconds of stimulation. Accuracies
+		are shown for the evaluation settings averaging over different
+	        numbers of trials, ranging from 1 to 15.
+
+  }{fig:accuracy_ap}
+
   \begin{table}[p]
@@
 	When more data are available, \textsc{stbf-struct} reaches significantly better accuracies.
-	Combined with the benefits laid out in  \cref{sec:parameters_complexity} and
-	\cref{sec:interpretability}, this makes it an attractive option for \ac{erp} classification.
+	Combined with the benefits laid out in \cref{sec:parameters_complexity}, this
+  makes it an attractive option for \ac{erp} classification.
 	\textsc{stbf-shrunk} does not show decisive accuracy improvements over

### 2024-10-09 chapters/wcble.tex
Comments François patients

@@
 \emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
-as part of~\textcite{VanDenKerchove2024}}
+as part of~\textcite{VanDenKerchove2024}.}
 
@@
 stimuli (see~\cref{fig:bci/loop}).
-These \acp{erp} consist of multiple \ac{erp}-components, peaks, and troughs in
+These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
 \ac{erp} waveform (see \cref{fig:bci/erp}).
@@
 This effect is especially prominent when bringing \ac{bci} development from the
-lab to the clinical setting with patients in the target population.
-\textcite{Zisk2021} showed that patients with \ac{als} have significantly
+lab to the clinical setting.
+For instance, \textcite{Zisk2021} showed that \ac{als} results significantly
 increased \ac{erp} jitter.
@@
   classifier.
-}{%
-  fig:diagram-train
-}
+}{fig:diagram-train}
 The process halts after a fixed number of iterations or when the estimated set
@@
 manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
-vector of scores over time as in~\cref{sec:cble} and the estimated latencies as
+vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
 in~\cref{sec:robust-latency}.
@@
   trained second-stage classifier, which predicts the label of the epochs.
-}{%
-  fig:diagram-eval
-}
+}{fig:diagram-eval}
 \section{Results}
@@
 
-\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble} decreases with
+\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble}
+decreases with
 $\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
@@
 be properly studied for this case.
+
+\clearpage
+\begin{subappendices}
+\section{\Acs{cble} algorithm}
+\label{sec:wcble/app/cble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
+  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\State $\theta \gets
+			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
+      \algorithmiccomment{Train stage 1}
+		\For{$n=1\ldots N$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\theta,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}
+      \algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+
+  \caption{\Acs{cble}}
+	\label{alg:cble}
+\end{algorithm}
+
+\clearpage
+\section{\Acs{wcble} algorithm}
+\label{sec:wcble/app/wcble-alg}
+\begin{algorithm}[h]
+	\textsc{Train}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
+		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
+	\begin{algorithmic}[1]
+		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
+    N$
+      \algorithmiccomment{Train stage 1}
+		\Repeat
+		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
+		\For{$n=1\ldots N$}
+    \State  $s_n \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
+		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
+		\EndFor
+    \Until{convergence or maximum iterations reached}
+    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{n,s}^\mathrm{train} \gets
+      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
+		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+
+	\end{algorithmic}
+	\textbf{Output:} $\theta^*,
+  \mat{Y}^\mathrm{train},
+  \mathbf{s}_\mathrm{target}^\mathrm{train},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
+
+	\smallskip \hrule \smallskip
+	\textsc{Evaluate}
+	\smallskip \hrule \smallskip
+	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
+  \begin{algorithmic}[1]
+		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
+		  \For{$s=1\ldots R$}
+      \State $y_{m,s}^\mathrm{test} \gets
+      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
+        \EndFor
+        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
+		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
+			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
+    \EndFor
+	\end{algorithmic}
+  \textbf{Output:} $\mat{Y}^\mathrm{test},
+  \mathbf{s}_\mathrm{target}^\mathrm{test},
+  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
+  \caption{\Acs{wcble}}
+	\label{alg:wcble}
+\end{algorithm}
+\end{subappendices}

### 2024-10-09 code/setup.py

@@ -76,17 +76,13 @@ def save_pgf_trim(fig, ax,path, width=textwidth_in, height=None):
     if height is None:
         golden_ratio = (5**0.5 - 1) / 2
         height=width*golden_ratio
+    fig.set_constrained_layout(True)
+    print(width)
+    print(height)
     fig.set_size_inches(width,height)
-    # Get the figure DPI (dots per inch)
+    plt.draw()
     dpi = fig.dpi
-    # Get the position of the "left" spine (y axis) using the ax object
-    left_spine_vertices = ax.spines['left'].get_path().vertices
-    # Transform these vertices to figure coordinates (from pixels)
-    left_spine_transformed = ax.spines['left'].get_transform().transform(left_spine_vertices)
-    # Extract the y-coordinate (in pixels) of the bottom-left corner (from left spine)
-    bottom_left_corner_y_pixels = left_spine_transformed[0][1]
-    # Convert y-coordinate from pixels to inches
-    margin_left = bottom_left_corner_y_pixels / dpi
+    margin_left = ax.get_window_extent().x0/dpi
     print(margin_left)
     fig.set_size_inches(width+margin_left,height)
     plt.savefig(path, format="pgf", bbox_inches='tight', pad_inches=0)

### 2024-10-09 code/wcble/fig_simulated.py

@@ -86,7 +86,6 @@ stc = "sine"
 height_ratios = []
 for _ in range(len(snrs)):
     height_ratios.extend([3, 1])  # Closer together for image+timeseries
-    height_ratios.append(0.5)
 gridspec_kw = dict(height_ratios=height_ratios, hspace=0)
 shape = (len(height_ratios), len(jitters))
 fig, axs = plt.subplots(
@@ -96,23 +95,16 @@ fig, axs = plt.subplots(
     gridspec_kw=gridspec_kw,
 )
 axs = axs.reshape(shape)
-# Hide every third row that is used as a spacer
-for i in range(2, len(axs), 3):
-    for j in range(len(jitters)):  # Loop over columns (since axs is 2D)
-        axs[i, j].axis("off")
 
-for ax in axs[-1,:]:
-    ax.remove()
-
-cbar_ax = fig.add_axes([.92, 0.15, 0.02, 0.1])
+cbar_ax = fig.add_axes([1.02, 0.05, 0.02, 0.1])
 for j, snr in enumerate(snrs):
     for k, jitter in enumerate(jitters):
         prefix = f"{stc}/{snr}/{jitter}"
         print(prefix)
         curr_epochs = epochs[prefix]
 
-        ax_img = axs[j * 3, k]
-        ax_ts = axs[(j * 3) + 1, k]
+        ax_img = axs[j * 2, k]
+        ax_ts = axs[(j * 2) + 1, k]
 
         plot_img(ax_img, curr_epochs, cbar_ax=cbar_ax)
         ax_img.lines[0].remove()
@@ -126,9 +118,10 @@ for j, snr in enumerate(snrs):
 
         plot_ts(ax_ts, curr_epochs)
         ax_ts.set_yticks([-10,10])
-        ax_ts.set_ylabel('$\mu V$')
+        ax_ts.set_ylabel('µV')
         ax_ts.set_xticks([-2, -1, 0, 1, 2])
         ax_ts.set_xlabel("time (s)")
 
-save_pgf_trim(plt.gcf(), axs[1,0], f'figures/wcble/simulated-{stc}.pgf',
+
+save_pgf_trim(plt.gcf(), axs[0,0], f'figures/wcble/simulated-{stc}.pgf',
               height=8)

### 2024-10-09 chapters/abstract.tex
Self-plagiarism check

@@
 
-\paragraph{Keywords:} \acl{bci}, gaze independence, covert
-visuospatial attention, event-related potentials, linear decoding, multilinear
-decoding, motor impairment, gaze impairment
+\paragraph{Keywords:} brain-computer interface, electroencephalography,
+event-related potentials, covert visuospatial attention, (multi)linear
+decoding, severe physical impairment

### 2024-10-09 chapters/abstract_fr.tex
Self-plagiarism check

@@
 
-\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéaire, handicap moteur, handicap du regard.
+\paragraph{Mots-clés :} interface cerveau-ordinateur, électroencéphalographie,potentiels évoqués
+attnetion visuo-spatiale cachée, decodage (multi-)linéaire, patients lourdements handicapées

### 2024-10-09 chapters/abstract_nl.tex
Self-plagiarism check

@@
 Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
-aanzienlijke uitdagingen bij het gebruik van traditionele visuele brain-computer interfaces (BCI's).
+aanzienlijke uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
 Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
@@
 
-\paragraph{Trefwoorden:} brain-computer interface, blikonafhankelijkheid, verborgen visuospatiële aandacht, event-gerelateerde potentialen, lineaire decodering, multilineaire decodering, motorische beperking, blikstoornis.
+\paragraph{Trefwoorden:} brein-computer interface, electroencephalografie,
+event-gerelateerde potentialen, bedekte visuospatiale attentie, (multi)lineaire
+decodering, zware fysieke beperkingen

### 2024-10-09 chapters/bci.tex
Self-plagiarism check

@@
 \label{sec:bci}
-\todo{consistently terminology individuals with severe
-speech and physical impairments (SSPI)}
-\todo{uniform vsa condition vsa setting}
-\todo{are all acronyms defined/Acro corectly used in chapters?}
-\todo{footnot skips a page in wcble chap}
 
@@
 \section{A direct interface to the brain}
-\todohsm{make less verbose}
 
@@
 \cref{fig:bci/loop}.
-It would make little sense to start digging around at random in all the brain activity
-to try to extract the desired output.
-Instead, it would help greatly if we knew which specific activity to look for.
-A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
-by which interaction is conducted.
-This usually means that the user is instructed to perform a specific task, like
-attending a specific flickering stimulus.
-We can now exploit background knowledge from prior neuroscientific research to couple
-specific activity in the brain to specific information or actions we want to convey via
-the \ac{bci}.
+It would be inefficient to search all brain activity randomly for the desired output.
+Instead, knowing the specific activity to target would help significantly.
+A \ac{bci} operates under a specified \emph{paradigm}, which defines how interaction
+is conducted.
+This typically means instructing the user to perform a task, like attending to a
+flickering stimulus.
+Background knowledge from neuroscience research allows us to couple specific brain
+activity with the information or actions conveyed via the \ac{bci}.
 The brain signal now \emph{modulates} this information.
-In turn, the detection of this signal can be coupled to an action, like typing the
-intended letter A or B in \cref{fig:bci/loop}.
-The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
-task performed by the user, and is often closely linked to the manner of feedback in the
-case of a closed-loop system.
-As shown in \cref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
-based on different systems in the brain.
+In turn, detecting this signal can trigger an action, like typing the intended letter A
+or B in \cref{fig:bci/loop}.
+The \ac{bci} paradigm includes both the method of stimulation (if applicable) and the
+user's task, and is often closely tied to the feedback mechanism in a closed-loop system.
+As shown in \cref{sec:bci/paradigms}, \ac{bci} paradigms vary widely based on different
+brain systems.
 
-The previous points to another component of a successful \ac{bci}: the specific brain
-signatures related to our paradigm need to be identified within the recorded activity.
-The electrical activity of the brain is relatively weak compared to that of the
-environment around us, containing electronic devices and other sources of interference.
-Furthermore, the brain is continuously processing information and carrying out
-`background tasks,' all of which generate their own brain activity.
-On the other hand, the activity of interest often originates from a specific region or
-network within the brain.
-This activity might not easily be discerned from all other brain and environmental
-activity.
-Compare this to trying to listen to a single speaker at a noisy party where everyone is
-shouting over each other.
-Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
-(see \cref{sec:bci/preprocessing}) and by choosing the correct representation of the
-signal (i.e., \emph{feature extraction}), but the problem of retrieving or
-\emph{decoding} only the relevant signal is generally hard.
-Nevertheless, we can attempt to solve it through supervised machine learning.
-Useful information that is encoded in the brain signal by the paradigm can be retrieved
-from it by the decoder and interpreted in function of the desired \ac{bci} output.
-\Cref{sec:bci/decoding} takes a closer look at some common decoding techniques.
+Another critical component of a successful \ac{bci} is identifying brain signatures
+related to the paradigm within the recorded activity.
+The brain's electrical activity is weak compared to environmental noise from electronic
+devices and interference.
+Additionally, the brain continuously processes information and carries out
+`background tasks', which generate more brain activity.
+The desired activity often originates from specific brain regions or networks.
+This signal can be difficult to isolate from all other brain and environmental activity.
+It is similar to trying to hear a single speaker at a noisy party where everyone is
+shouting.
+Some interference can be filtered out through \emph{preprocessing}
+(see \cref{sec:bci/preprocessing}) and by selecting the correct signal representation
+(i.e., \emph{feature extraction}), but retrieving or \emph{decoding} the relevant signal
+remains challenging.
+Supervised machine learning can help solve this.
+The information encoded in the brain signal can be retrieved by the decoder and used
+to determine the desired \ac{bci} output.
+\Cref{sec:bci/decoding} reviews some common decoding techniques.
 
@@
 cognitive state or brain activity, creating an adaptive system.
-A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
-Examples of this are deep brain stimulation to mitigate the symptoms of Parkinson's
-disease, or the cochlear implant as a hearing prosthesis.
-More indirectly, modulation can be achieved by sensory stimulation.
-This involves specific, paradigm-related sensory input, like tactile feedback for
+This happens either directly through electrical neurostimulation, as is the
+case for adaptive deep brain stimulation in Parkinson's disease or for the
+cochlear implant as a hearing aide, or indirectly through sensory stimulation.
+The latter involves specific, paradigm-related, sensory input, like tactile feedback for
 movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
@@
 all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
-$\mat{x} = \vec(\mat{X})$.
+$\mat{x} = \text{vec}(\mat{X})$.
 

### 2024-10-09 chapters/bttda.tex
Self-plagiarism check

@@
 
-Brain-computer interfaces (BCIs) have the potential to bypass
+\Acp{bci} have the potential to bypass
 defective neural pathways by providing an alternative communication channel
@@
 technologies, among other applications~\cite{Wolpaw2020}.
-To achieve their functionality, BCIs record and process neural data obtained through
-a neuroimaging technique, with electroencephalography (EEG) being the most popular.
+To achieve their functionality, \acp{bci} record and process neural data obtained through
+a neuroimaging technique, with \ac{eeg} being the most popular.
 
@@
 
-Due to its multichannel time series structure, EEG data, like most neural
-signal acquisition modalities used for BCIs, naturally exist as multiway data,
+Due to its multichannel time series structure, \ac{eeg} data, like most neural
+signal acquisition modalities used for \acp{bci}, naturally exist as multiway data,
 capturing information in both spatial and temporal domains.
@@
 Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
-using the Higher Order Singular Value
-Decomposition (HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}.
+using the \ac{hosvd}~\cite{DeLathauwer2000,SoleCasals2018}.
 Alternatively, the PARAFAC structure can be used.
@@
 Performance of methods leveraging either the Tucker and PARAFAC structures are
-heavily dependent on the a priori choice of hyperparameters describing
+heavily dependent on the prior choice of hyperparameters describing
 the multilinear rank or the number of rank-1 terms.
@@
 In this philosophy, the Tucker decomposition can also be obtained
-using Higher Order Discriminant Analysis
-(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fisher sense, analogous to linear
+using \ac{hoda}~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fisher sense, analogous to linear
 discriminant analysis.
 Extracted features can subsequently be further classified, most commonly
-using LDA or a support vector machine (SVM) to predict class labels.
+using \ac{lda} or a \ac{svm} to predict class labels.
 
-Variants of HODA have been applied to BCI problems such as
-ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
+Variants of \ac{hoda} have been applied to \ac{bci} problems such as
+\ac{erp}~\cite{Onishi2012,Higashi2016} and \ac{mi}~\cite{Liu2015,Cai2021}
 decoding with positive results~\cite{Lotte2018}.
@@
 However, it is not immediately obvious if either the Tucker or PARAFAC
-structure are most suited to represent the neural data of interest for the BCI
+structure are most suited to represent the neural data of interest for the
+\ac{bci}
 paradigm and for decoding.
@@
 Recent research has shown that supervised decoders adopting a more flexible structure
-can improve BCI performance.
+can improve \ac{bci} performance.
 Promising results have been achieved for regression tasks using
@@
 We propose to implement this idea as a new supervised feature
-extraction tensor method that is a generalization of the aforementioned HODA
+extraction tensor method that is a generalization of the aforementioned
+\ac{hoda}
 algorithm and is more suited for the extraction of discriminant
@@
 \begin{enumerate*}[label={\arabic*)}]
-	\item  We develop a forward model for HODA to reconstruct a
+  \item  We develop a forward model for \ac{hoda} to reconstruct a
 	      given input tensor from the extracted features.
-	\item This allows us to introduce a state-of-the-art BCI feature extraction
-        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
-	      (BTTDA).
-  \item We evaluate a BCI decoder based on BTTDA and its special
-	      PARAFAC-structured case on decoding tasks for both ERP and MI
-	      paradigm BCI datasets.
+      \item This allows us to introduce a state-of-the-art \ac{bci} feature extraction
+        method based on the block-term tensor structure, named \acf{bttda}.
+      \item We evaluate a \ac{bci} decoder based on \ac{bttda} and its special
+        PARAFAC-structured case on decoding tasks for both \ac{erp} and \ac{mi}
+        paradigm \ac{bci} datasets.
 \end{enumerate*}
@@
 
-\subsection{Higher Order Discriminant Analysis}
-Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
+\subsection{\Acl{hoda}}
+\Acl{hoda}~\cite{Phan2010} is a
 supervised tensor-based feature extraction technique.
@@
   \caption[A \acs{hoda} backward projection.]{%
-    A visualization of the multilinear projection obtained by Higher Order
-    Discriminant Analysis (HODA) applied to a third-order tensor
+    A visualization of the multilinear projection obtained by \acf{hoda} applied to a third-order tensor
     sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
-		HODA finds projection matrices $\mat{U}_k$ such that maximal
+		\Ac{hoda} finds projection matrices $\mat{U}_k$ such that maximal
 		discriminability between classes is achieved in the projected latent tensors
@@
 
-Analogous to the HOSVD, HODA is a dimensionality
+Analogous to the \ac{hosvd}, \ac{hoda} is a dimensionality
 reduction decomposition that results in a dense latent tensor $\ten{G}$, and
 imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
-However, while for the HOSVD decomposition the projection matrices
+However, while for the \ac{hosvd} decomposition the projection matrices
 are chosen to minimize the reconstruction error, the projection matrices
-$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
+$\mat{U}_k$ of \ac{hoda} are optimized for maximal discriminability of tensors
 $\ten{G}(n)$ belonging to classes with labels $c_n$.
@@
 
-HODA optimizes discriminability in the Fisher sense, by optimizing the
+\Ac{hoda} optimizes discriminability in the Fisher sense, by optimizing the
 Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
@@
 orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
-or as the partial HOSVD of all stacked tensors in the dataset.
+or as the partial \ac{hosvd} of all stacked tensors in the dataset.
 At each iteration, the algorithm loops through the modes and fixes all
@@
 update of each $\mat{U}_k$ is lower than a predetermined threshold.
-The full HODA procedure is summarized in \cref{alg:HODA}.
+The full \ac{hoda} procedure is summarized in \cref{alg:HODA}.
 \begin{algorithm}
@@
 
-To apply HODA in a classification setting, the projections
+To apply \ac{hoda} in a classification setting, the projections
 are first learned on a training dataset with known class labels.
@@
 
-\subsection{A forward model for HODA}
-As a prerequisite to the proposed BTTDA model, we must find a
+\subsection{A forward model for \acs{hoda}}
+As a prerequisite to the proposed \ac{bttda} model, we must find a
 way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
@@
 
-The HODA projection in \cref{eq:HODA-backward} is an example
+The \ac{hoda} projection in \cref{eq:HODA-backward} is an example
 of a backward model.
@@
 
-\subsection{Block-Term Tensor Discriminant Analysis}
+\subsection{\Acl{bttda}}
 After defining the forward model, we can construct our proposed block-term
@@
 
-We thus extend the HODA feature extraction scheme to Block-Term Tensor Discriminant Analysis
-(BTTDA).
-BTTDA finds multiple discriminative blocks, such that its forward
+We thus extend the \ac{hoda} feature extraction scheme to \acf{bttda}.
+\Ac{bttda} finds multiple discriminative blocks, such that its forward
 model adheres to the block-term tensor structure:
@@
 	\input{figures/bttda/BTTDA.tikz.tex}
-  \caption[A forward model for \acs{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
-		(BTTDA). BTTDA can extract more features
-		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
+  \caption[A forward model for \acs{bttda}.]{A forward model for \acf{bttda}.
+    \Ac{bttda} can extract more features
+    than \ac{hoda} by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
 		deflation scheme.
-		The HODA backward projection is first applied. Next, the
+    The \ac{hoda} backward projection is first applied. Next, the
 		input data is reconstructed via the HODA forward model and the
@@
 The block-term structure of this model implies that it is a generalization of both
-the Tucker-structured HODA and PARAFAC-structured discriminant feature
+the Tucker-structured \ac{hoda} and PARAFAC-structured discriminant feature
 extraction.
-If $B$ in \cref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
-HODA; if at each term $b$ the rank of the core tensor
+If $B$ in \cref{eq:BTTDA-forward} is set to one, \ac{bttda} is equivalent to
+\ac{hoda}; if at each term $b$ the rank of the core tensor
 $(r_1^{(b)}=r_2^{(b)}=\ldots=r_k^{(b)}=1)$, a PARAFAC structure is assumed.
 
-Since BTTDA is specified above as a forward model, a backward procedure
+Since \ac{bttda} is specified above as a forward model, a backward procedure
 is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
-BTTDA to be useful as a feature extraction method.
+\ac{bttda} to be useful as a feature extraction method.
 The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
@@
 This residual error term is calculated by finding the difference between the
-previous error and its reconstruction after backward and forward HODA
+previous error and its reconstruction after backward and forward \ac{hoda}
 projection:
@@
 \subsection{Model and feature selection}
-Similar to the unsupervised Block-Term Decomposition, the performance of BTTDA is
+Similar to the unsupervised Block-Term Decomposition, the performance of
+\ac{bttda} is
 heavily dependent on the rank $(r_1^{(b)}, r_2^{(b)}, \ldots,
@@
 
-Furthermore, HODA, and by extension BTTDA, can extract a substantial amount
+Furthermore, \ac{hoda}, and by extension \ac{bttda}, can extract a substantial amount
 of redundant features, which can be dropped after projection and before proceeding to the classification
 step~\cite{Phan2010}.
-Specifically, in BTTDA redundant features can accumulate over the number of
+Specifically, in \ac{bttda} redundant features can accumulate over the number of
 blocks, hampering performance.
@@
 previous block.
-The area under the receiver operating characteristic curve (ROC-AUC) for
-classification of extracted feature vectors after feature selection is used as
+The \ac{rocauc}, or the accuracy in the case of a balanced multi-class problem,
+of classification of extracted feature vectors after feature selection is used as
 a cross-validation score.
@@
 
-Alternatively, a special case of BTTDA can be constructed using only rank-one
+Alternatively, a special case of \ac{bttda} can be constructed using only rank-one
 blocks such that the resulting forward model adheres to the PARAFAC structure.
@@
 \end{figure*}
-\todo{format figure}
 
@@
 		condition contrasts (right column) obtained after forward projection of the latent
-		features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of BTTDA
+    features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of \ac{bttda}
     fit on the full dataset BNCI2014-008.
-		The separate blocks approximately model different ERP
+    The separate blocks approximately model different \ac{erp}
 		components.}
@@
 Given informed or correctly tuned hyperparameters, this method could be used to
-e.g., separate ERP components or neural processes based on the task-related
+e.g., separate \ac{erp} components or neural processes based on the task-related
 information in the class labels.
@@
 
-BTTDA trades in the rigid HODA model for increased model complexity with more
+\Ac{bttda} trades in the rigid \ac{hoda} model for increased model complexity with more
 hyperparameters to tune, which could open up a setting where performance can be
@@
 instead of a projection optimization procedure.
-Despite favorable results in BCI decoding, the applications of the proposed
-BTTDA model are limited mainly by this model selection approach used
+Despite favorable results in \ac{bci} decoding, the applications of the proposed
+\ac{bttda} model are limited mainly by this model selection approach used
 to determine the individual block ranks.
@@
 validation can still hinder the portability of decoders relying on
-BTTDA.
+\ac{bttda}.
 
@@
 In a sense, this goes against the proposition of increased model flexibility.
-It might well be that BTTDA in some cases offers little added value over the
-Tucker-structured HODA, when both are given free choice of rank.
+It might well be that \ac{bttda} in some cases offers little added value over the
+Tucker-structured \ac{hoda}, when both are given free choice of rank.
 Future efforts should focus on automatic parameter setting, e.g., using sparsity
-or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
-statistical measures based on the model's application.
+or information criteria such as the ones used in Block-Term Tensor Regression~\cite{Faes2022}
+or other statistical measures based on the model's application.
 
-Another limitation is that BTTDA might yield a disproportionate
+Another limitation is that \ac{bttda} might yield a disproportionate
 improvement for datasets with a low number of features relative to sample size,
 while being less effective for datasets with more features.
-This is reflected in our ERP results (low dimensionality vs. high number of
-trials) compared to the MI results (higher dimensionality due to third-order
+This is reflected in our \ac{erp} results (low dimensionality vs. high number of
+trials) compared to the \ac{mi} results (higher dimensionality due to third-order
 tensorization vs. lower number of trials).
@@
 large impact on the model.
-Other tensorization methods of the EEG data, like time-lagged Hankel
+Other tensorization methods of the \ac{eeg} data, like time-lagged Hankel
 tensors~\cite{Papy2005}, or tensors across subjects or sliding windows, etc.,
@@
 
-We have introduced Block-Term Tensor Discriminant Analysis (BTTDA), a novel,
+We have introduced \acf{bttda}, a novel,
 tensor-based, supervised dimensionality reduction technique optimized for class
 discriminability, which adheres to the block-term tensor structure.
-BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
+\ac{bttda} is a generalization of \acf{hoda} and can also be
 applied as a special sum-of-rank-one tensors PARAFACDA model.
-The model is obtained by iteratively fitting HODA in a deflation scheme,
+The model is obtained by iteratively fitting \ac{hoda} in a deflation scheme,
 leveraging a novel forward modeling step.
 
-Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
-feature extraction can significantly outperform decoders based on HODA and
-reach state-of-the-art decoding performance on event-related potential
-problems (second-order tensors) and scores on par with or higher than HODA in motor
+Via an accompanying heuristic model selection procedure, \ac{bci}decoders using
+\ac{bttda} feature extraction can significantly outperform decoders based on
+\ac{hoda} and reach state-of-the-art decoding performance on event-related potential
+problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
 imagery problems (third-order tensors).
-Moving from the rigid Tucker tensor structure of HODA to the more flexible
+Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible
 block-term structure shifts the problem from finding optimally constrained

### 2024-10-09 chapters/conclusion.tex
Self-plagiarism check

@@
 \chapter{Conclusion \& recommendations}
+\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
+part of \textcite{VanDenKerchove2022}.}
 
@@
 \subsection{Developed \acs{erp} decoders}
+\label{sec:conclusion/contrib/decoders}
 We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
@@
 \acp{erp}.
-This work is submitted as~\textcite{VanDenKerchove2024}.
+This work is submitted as~\textcite{VanDenKerchove2024a}.
 

### 2024-10-09 chapters/contribution.tex
Self-plagiarism check

@@
 \chapter*{Personal contribution}
-Arne Van Den Kerchove conceptualized the works and performed the analyses in
+Arne Van Den Kerchove conceptualized the works and performed the analyses
+presented in
 this thesis, under supervision and in consultation with prof. Hakim Si-Mohammed,

### 2024-10-09 chapters/covert_align.tex
Self-plagiarism check

@@
 \label{sec:covert-align}
-\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}.}
+\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
 
@@
 
-\subsection{BCI decoding performance}
+\subsection{\Acs{bci} decoding performance}
 \label{sec:block_accuracy}
 
-We evaluated the BCI decoding performance in a single-trial classification
-experiment, as well as in a target selection experiment reflecting BCI
+We evaluated the \ac{bci} decoding performance in a single-trial classification
+experiment, as well as in a target selection experiment reflecting \ac{bci}
 operation.
 
-\Cref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
-Receiver Operating Characteristic curve (ROC-AUC) for all pairs of \ac{tlda}, \ac{cble},
+\Cref{fig:single_trial_roc_auc_diff} shows a comparison of \ac{rocauc}
+for all pairs of \ac{tlda}, \ac{cble},
 and \ac{wcble} for single-trial classification to investigate the contributions of
@@
 
-\Cref{fig:block_evaluation} shows the cross-validated BCI selection
+\Cref{fig:block_evaluation} shows the cross-validated \ac{bci} selection
 accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
@@
 corrected for FDR over decoders.
-For this evaluation, no epochs were rejected to keep the trial-based structure of BCI
-operation intact.
+For this evaluation, no epochs were rejected to keep the trial-based structure
+of \ac{bci} operation intact.
 For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
@@
 for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
-Furthermore, while we reported a relative decrease in single-trial ROC-AUC for
+Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
 \ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target

### 2024-10-09 chapters/stbf_struct.tex
Self-plagiarism check

@@
 	%calibration time before usage~\cite{Gao2014, Kapgate2015}.
-	%The study we report on focuses on the visual P300 oddball \ac{erp} in response to a
+	%The study we report on focuses on the visual P3 oddball \ac{erp} in response to a
 	%rare but attended visual stimulus.
@@
 	%the user attended.
-	%The P300 paradigm has been used extensively in \ac{bci} development and is easy to
+	%The P3 paradigm has been used extensively in \ac{bci} development and is easy to
 	%set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.
 
-  There are multiple state-of-the-art P300 classification methods, like
+  There are multiple state-of-the-art P3 classification methods, like
   \acp{svm}~\cite{Tayeb2014}, deep
@@
 	possible to apply the spatiotemporal beamformer to multiple time-locked visual
-  \ac{bci} paradigms, including the P300 oddball paradigm, \ac{ssvep},
+  \ac{bci} paradigms, including the P3 oddball paradigm, \ac{ssvep},
   \ac{cvep}~\cite{Wittevrongel2017a}, \ac{mvep}~\cite{Libert2021}.
@@
 	\subsection{Dataset}
-	We use the dataset from~\textcite{Wittevrongel2016}, containing P300 oddball \ac{eeg}
+	We use the dataset from~\textcite{Wittevrongel2016}, containing P3 oddball \ac{eeg}
 	recordings of 21 healthy subjects since it is a high-quality dataset with a high
@@
 	%	As expected, \cref{fig:interpret_weights}b and d exhibit weight around
-	%	the central and parietal regions, where the P300 \ac{erp} component is present.
+	%	the central and parietal regions, where the P3 \ac{erp} component is present.
 	%	Especially the spatial weights of \textsc{stbf-shrunk} in
@@
 	%	This high weight for the early visual components confirms the results from Treder \& Blankertz~\cite{Treder2010}
-	%	that state that, in addition to the P300, the early N1 and P2 \ac{erp} components
+	%	that state that, in addition to the P3, the early N1 and P2 \ac{erp} components
 	%	are also modulated by oddball attention and contain discriminatory information between
@@
 	We compared these estimators with the original spatiotemporal
-	beamformer and a state-of-the-art method in an off-line P300 detection task.
+	beamformer and a state-of-the-art method in an off-line P3 detection task.
 	Our results show that the structured estimator performs better when training data are sparsely available and that it can be computed faster and with substantially less memory usage.

### 2024-10-09 chapters/wcble.tex
Self-plagiarism check

@@
 \label{sec:wcble}
-\emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
+\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
+\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
+\cref{sec:wcble/conclusion} were published
 as part of~\textcite{VanDenKerchove2024}.}
@@
 }{%
-  Schematic representation of the \ac{wcble} training phase.
+  Schematic representation of the \acf{wcble} training phase.
   (1) The first-stage spatiotemporal binary classifier is trained on a set of
@@
 }{%
-  Schematic representation of the Woody Classifier-Based Latency
-  Estimation test phase. (1) The
+  Schematic representation of the \acf{wcble}. (1) The
 	first-stage spatiotemporal binary classifier obtained from the training phase is
@@
   deviation).
-  Targets contain a sine wave pulse jittered by $\sigma$. 32 EEG channels with pink
+  Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
   spatiotemporal noise were simulated using a forward head model, and channel C3 is
@@
 \section{Discussion \& conclusion}
+\label{sec:wcble/conclusion}
 

### 2024-10-09 code/wcble/fig_simulated.py

@@ -79,6 +79,7 @@ def plot_ts(ax, epochs):
     df = df.melt(ignore_index=False, value_name="amplitude")
     sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts,
                  errorbar='sd')
+    ax.set_xlim([-2,2])
 
 
 stc = "sine"
@@ -90,7 +91,6 @@ gridspec_kw = dict(height_ratios=height_ratios, hspace=0)
 shape = (len(height_ratios), len(jitters))
 fig, axs = plt.subplots(
     *shape,
-    sharex="col",
     sharey="row",
     gridspec_kw=gridspec_kw,
 )
@@ -115,12 +115,18 @@ for j, snr in enumerate(snrs):
             ax_img.set_ylabel("")
         else:
             ax_img.set_ylabel("epochs")
+        ax_img.set_xticks([-2, -1, 0, 1, 2])
+        ax_img.tick_params(axis='x', labelbottom=False)
 
         plot_ts(ax_ts, curr_epochs)
         ax_ts.set_yticks([-10,10])
         ax_ts.set_ylabel('µV')
         ax_ts.set_xticks([-2, -1, 0, 1, 2])
-        ax_ts.set_xlabel("time (s)")
+        if j == len(snrs)-1:
+            ax_ts.set_xlabel("time (s)")
+        else:
+            ax_ts.set_xlabel("")
+            ax_ts.tick_params(axis='x', labelbottom=False)
 
 
 save_pgf_trim(plt.gcf(), axs[0,0], f'figures/wcble/simulated-{stc}.pgf',

### 2024-11-14 chapters/abstract.tex
Started processing comments MDV

@@
 time and computational complexity.
-This allows for more efficient training and better performance on unseen data,
-contributing to more reliable systems that are adaptable to various contexts and user needs.
+This allows for more efficient training and contributes to generalizability
+allowing for more reliable systems that are adaptable to various contexts and
+user needs.
 
@@
 event-related potentials, covert visuospatial attention, (multi)linear
-decoding, severe physical impairment
+decoding, physical impairment

### 2024-11-14 chapters/abstract_fr.tex
Started processing comments MDV

@@
 Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
-cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
-Ce travail vise \`a palier cette limitation en développant des méthodes de BCI
-indépendantes du regard, en se concentrant sur l'amélioration de la attention visuo-spatiale (VSA)
-cachée,
-où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements
-oculaires correspondants.
+cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
+Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
+indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
 Une contribution clé de ce travail est la compensation pour la compensation de
 la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
-la performance de décodage dans la
-VSA cachée.
-En gérant cette variabilite, la méthode proposée améliore la précision de la communication
+la performance de décodage de VSA cachée.
+En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
 sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.
@@
 multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
-Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.
+Cela permet un entraînement plus efficace et contribue à la generalisation, ce qui
+permets le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
+contextes et besoins des utilisateurs.
 
@@
 
-\paragraph{Mots-clés :} interface cerveau-ordinateur, électroencéphalographie,potentiels évoqués
-attnetion visuo-spatiale cachée, decodage (multi-)linéaire, patients lourdements handicapées
+\paragraph{Mots-clés :}
+interfaces neuronales directes,
+\'electroenc\'ephalographie,
+potentiels \'evoqu\'es,
+attention visuo-spatiale cachée,
+decodage (multi-)linéaire,
+handicap\'es moteurs

### 2024-11-14 chapters/abstract_nl.tex
Started processing comments MDV

@@
 
-Personen met ernstige motorische beperkingen, zoals degenen met het
-Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
-aanzienlijke uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
-Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
-onuitvoerbaar wordt voor personen met beperkte of
-geen controle over oogbewegingen.
-Dit werk richt zich op deze beperking door
-BCI-methoden onafhankelijk van oogbewegingen te ontwikkelen, met een focus op
-het verbeteren van de bedekte visuospatiële aandacht (VSA), waarbij gebruikers
-hun aandacht kunnen richten op een doel zonder overeenkomstige oogbewegingen.
+Personen met ernstige motorische beperkingen, bijvoorbeeld het Locked-in
+Syndroom (LIS) of oog-motorische disfunctie, ondervinden aanzienlijke
+uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
+Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen,
+wat onuitvoerbaar wordt voor personen met beperkte of geen controle
+over oogbewegingen.
+Dit werk richt zich op deze beperking door BCI-methoden onafhankelijk van
+oogbewegingen te ontwikkelen, met een focus op het verbeteren van de bedekte
+visuospati\"ele aandacht (VSA), de  aandacht voor een doel zonder
+overeenkomstige oogbewegingen.
 Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
@@
 Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
-nauwkeurigheid van de communicatie zonder afhankelijkheid van oogcontrole,
-waardoor BCI's bruikbaarder worden voor personen met motorische beperkingen van
-de ogen.
+nauwkeurigheid van communicatie onafhankelijkheid van oogcontrole,
+waardoor BCI's bruikbaarder worden voor personen met deze motorische
+beperkingen.
 
-Naast onafhankelijkheid van oogbewegingen, bevordert dit werk ook de algemene
-ERP-decodeertechnieken door de structuur van lineaire en multilineaire
-decodeertechnieken te verfijnen.
+Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
+ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
 Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
@@
 en vermindert de trainingstijd en computationele complexiteit.
-Dit maakt efficiëntere training mogelijk en zorgt voor betere prestaties bij
-ongeziene data, wat bijdraagt aan betrouwbaardere systemen die zich kunnen
+Dit maakt efficiëntere training mogelijk en draagt bij aan generalisatie, wat
+de ontwikkeling van betrouwbaardere systemen toelaat die zich kunnen
 aanpassen aan verschillende contexten en gebruikersbehoeften.
@@
 
-\paragraph{Trefwoorden:} brein-computer interface, electroencephalografie,
-event-gerelateerde potentialen, bedekte visuospatiale attentie, (multi)lineaire
-decodering, zware fysieke beperkingen
+\paragraph{Trefwoorden:}
+brein-computer interface,
+electroencephalografie,
+event-gerelateerde potentialen,
+bedekte visuospatiale aandacht,
+(multi)lineaire decodering,
+fysieke beperkingen

### 2024-11-14 chapters/bci.tex
Started processing comments MDV

@@
 }{%
-  The \acs{bci} loop.
-}{%
+  \caption[The \acs{bci} loop.]{%
     The \ac{bci} loop.
@@
     In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
-}{fig:bci/loop}
+  }
+  \label{fig:bci/loop}
+}
 
@@
 }{%
-  Various electrophysiology recording technologies.
-}{%
+  \caption[Various electrophysiology recording technologies.]{%
 	  Various electrophysiology recording technologies and their respective position in a
@@
 	  There is a trade-off between invasiveness and spatial resolution.
-}{fig:bci/recording}
+  }
+  \label{fig:bci/recording}
+}
 
@@
 }{%
-  Overview of \acs{bci} paradigms.
-}{%
+  \caption[Overview of \acs{bci} paradigms.]{%
     A \ac{bci} paradigm defines which brain activity will be used for control.
@@
     Adapted from~\textcite{Muehl2014}.
-}{fig:bci/paradigms}
+  }
+  \label{fig:bci/paradigms}
+}
 
@@
 In \ac{erp} classification, they are usually directly used as features, by concatenating
-all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
-$\mat{x} = \text{vec}(\mat{X})$.
+all $C$ channels of length $S$ time samples in the matrix format epoch
+$\mat{X}\in\mathbb{R}^{C\times S}$ into one feature vector
+$\mat{x}\in\mathbb{R}^{CS}$ such that $\mat{x} = \text{vec}(\mat{X})$.
 
@@
 generally work well for classifying \ac{erp} epochs.
-\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can be
-multiplied with the feature vector of a new epoch to obtain a prediction score.
-The weights $\mat{w}$ are obtained by minimizing the variance of the output
+\Ac{lcmv} beamforming learns a weight vector $\mat{w}\in\mathbb{R}^{CS}$ from
+a set of epochs that can be multiplied with the feature vector $\mat{x}$ of a
+new epoch to obtain a prediction score.
+The weights are obtained by minimizing the variance of the output
 \begin{equation}
@@
 \end{equation}
-with activation pattern $\mat{a}$ the difference between the average of the attended
-epochs and the average of the non-attended epochs in the training data.
+with \emph{activation pattern} $\mat{a}\in\mathbb{R}^{CS}$ the difference between the
+average of feature vectores of the attended and non-attended epochs in the
+training data,
+and $\mat{C}\in\mathbb{R}^{CS\times CS}$ the background covariance matrix which
+can be estimated empirically from the training data.
 This is then solved by the method of Lagrange multipliers as
@@
 \end{equation}
-with $\mat{C}$ the background covariance.
 
@@
 covariance.
-Similar to \ac{lcmv} beamforming, \ac{lda} learns a weight vector $\mat{w}$ that
-satisfies
+Similar to \ac{lcmv} beamforming, \ac{lda} learns a weight vector
+$\mat{w}\in\mathbb{R}^{CS}$ which satisfies
 \begin{equation}
@@
 \end{equation}
-The between-class scatter matrix $\mat{S}_\text{b}$ models the variability between the
-expected responses, while the within-class scatter matrix $\mat{S}_\text{w}$ models the
+The between-class scatter matrix $\mat{S}_\text{b}\in\mathbb{R}^{CS\times CS}$ models the variability between the
+expected responses, while the within-class scatter matrix $\mat{S}_\text{w}\in\mathbb{R}^{CS\times CS}$ models the
 expected noise.
@@
 stationarity in the background noise to regularize the problem.
-It does this by imposing a block-Toeplitz structure to the between-class covariance
-matrix of \ac{lda}, such that every temporal covariance block corresponding to a channel
-has the structure
+It does this by imposing a block-Toeplitz structure to the between-class
+scatter $\mat{S}_\text{b}$ matrix of \ac{lda}.
+This means
 \begin{equation}
-  \mat{S}_\text{w}^{(c)} = \begin{bmatrix}
-s_0 & s_1 & s_2 & \cdots & s_{k-1} \\
-s_1 & s_0 & s_1 & \cdots & s_{k-2} \\
-s_2 & s_1 & s_0 & \cdots & s_{k-3} \\
-\vdots & \vdots & \vdots & \ddots & \vdots \\
-s_{k-1} & s_{k-2} & s_{k-3} & \cdots & s_0
+  \mat{S}_\text{w} = \begin{bmatrix}
+    \mat{S}_\text{w}^{(1,1)} & \mat{S}_\text{w}^{(1,2)} & \cdots & \mat{S}_\text{w}^{(1,C)}\\
+    \mat{S}_\text{w}^{(2,1)} & \mat{S}_\text{w}^{(2,2)} & \cdots & \mat{S}_\text{w}^{(2,C)}\\
+    \vdots             & \vdots             & \ddots & \vdots            \\
+    \mat{S}_\text{w}^{(C,1)} & \mat{S}_\text{w}^{(C,2)} & \cdots & \mat{S}_\text{w}^{(C,C)}\\
   \end{bmatrix}
 \end{equation}
+consists of $C\times C$ temporal scatter matrices
+$\mat{S}_\text{w}^{(c_1,c_2)}\in\mathbb{R}^{S\times S}$
+corresponding to a pair of channels $(c_1,c_2)$, such
+that every temporal scatter matrix has the Toeplitz structure
+\begin{equation}
+  \mat{S}_\text{w}^{(c_1,c_2)} = \begin{bmatrix}
+    s_1^{(c_1,c_2)}  & s_2^{(c_1,c_2)}  & \cdots & s_S^{(c_1,c_2)}  \\
+    s_2^{(c_1,c_2)}  & s_1^{(c_1,c_2)}  & \cdots & s_{S-1}^{(c_1,c_2)}  \\
+    \vdots & \vdots &  \ddots & \vdots \\
+    s_S^{(c_1,c_2)}  & s_{S-1}^{(c_1,c_2)}  & \cdots & s_1^{(c_1,c_2)}
+\end{bmatrix}
+\end{equation}
+The structuring is performed by setting all elements to the average of their
+corresponding block-subdiagonal and subsequently applying shrinkage
+regularization.
 This method is simple yet effective, reaching state-of-the-art decoding performance.

### 2024-11-14 chapters/bttda.tex
Started processing comments MDV

@@
 
-\subsection{Tensor \& tensor methods}
+\subsection{Tensors \& tensor methods}
 
@@
 restriction in hyperparameter choice.
-We imposed $(r_1=r_2=\ldots=r_k)$ that to reduce the computational demand of
+We imposed $(r_1=r_2=\ldots=r_k)$ to reduce the computational demand of
 the performed experiments.
@@
 
-Via an accompanying heuristic model selection procedure, \ac{bci}decoders using
+Via an accompanying heuristic model selection procedure, \ac{bci} decoders using
 \ac{bttda} feature extraction can significantly outperform decoders based on

### 2024-11-14 chapters/covert_align.tex
Started processing comments MDV

@@
-\chapter{Compensating \acs{erp} jitter for gaze-independence}
+\chapter{Compensating \acs{erp} jitter for gaze-independence}%
 \label{sec:covert-align}
@@
 decoding, but did not verify this hypothesis directly.
-Additionally, \cite{Hardiansyah2020} developed a classifier for
+Additionally, \textcite{Hardiansyah2020} developed a classifier for
 covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
@@
 
-\subsubsection{CVSA-ERP dataset}
+\subsubsection{CVSA-ERP dataset}%
 \label{sec:covert-align/stimulation}
@@
 Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
-by \cite{Treder2010}, we presented six flashing targets (without letters or
+by \textcite{Treder2010}, we presented six flashing targets (without letters or
 symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
@@
 	\begin{subfigure}{\linewidth}
-		\caption{}
+		\caption{}%
 		\label{fig:interface}
@@
 	\begin{subfigure}{\linewidth}
-		\caption{}
+		\caption{}%
 		\label{fig:erps}
@@
 
-  \caption[Stimulation interfaces and evoked \acsp{erp}.]{
+  \caption[Stimulation interfaces and evoked \acsp{erp}.]{%
 		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
@@
 
-In contrast to the protocol proposed by \cite{Frenzel2011}, split \ac{vsa} was
+In contrast to the protocol proposed by \textcite{Frenzel2011}, split \ac{vsa} was
 performed by instructing the participant to attend to the intensifications of the
@@
 Targets are full-contrast white and were intensified by scaling them to a
-larger size (5.60° of visual angle,
-(\cref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
+larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
+instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
 experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
@@
     \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
-    \acs{bci}~\cite{Treder2010}.}
+    \acs{bci}~\cite{Treder2010}.}%
     \label{fig:covert-align/interface/interface}
@@
     \caption[Intensification.]{When a target is intensified, it is enlarged for
-    a very short time.}
+    a very short time.}%
     \label{fig:covert-align/interface/intensification}
@@
     settings, the
-    participant is cued to fixate on one of the targets.}
+    participant is cued to fixate on one of the targets.}%
     \label{fig:covert-align/interface/overt}
@@
     \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
-    cued to fixate at the center of the interface.}
+    cued to fixate at the center of the interface.}%
     \label{fig:covert-align/interface/covert}
@@
 dataset~\cite{Aloise2012a} was used in the analysis performed
-by \cite{Arico2014}.
+by \textcite{Arico2014}.
 It contains data from 10 subjects (median age $24.5\pm1.9$
@@
 For a detailed description of the paradigm and dataset, we refer
-to \cite{Aloise2012a}.
+to \textcite{Aloise2012a}.
 
@@
 
-\subsubsection{Decoders}
+\subsubsection{Decoders}%
 \label{sec:covert-align/methods/decoders}
@@
 (XDAWNCov-TS-LR)~\cite{Cecotti2017}.
-Secondly, we adopted the pipeline from \cite{Aydarkhanov2020}, as their work shows
+Secondly, we adopted the pipeline from \textcite{Aydarkhanov2020}, as their work shows
 favorable performance in the presence of single-trial \ac{erp} latency jitter.
@@
 
-\subsection{\Acs{bci} decoding performance}
-\label{sec:block_accuracy}
+\subsection{\Acs{bci} decoding performance}%
+\label{sec:covert-align/results/bloc-acc}
 
@@
 
-\Cref{fig:single_trial_roc_auc_diff} shows a comparison of \ac{rocauc}
+\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
 for all pairs of \ac{tlda}, \ac{cble},
@@
 		\caption{}
-		\label{fig:single_trial_roc_auc_diff}
+		\label{fig:covert-align/single-trial-roc-auc-diff}%
 		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
@@
 		\caption{}
-		\label{fig:block_evaluation}
+		\label{fig:covert-align/block-eval}%
 		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
@@
   \caption[Decoder performance.]{%
-		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
-    single-trial classification receiver operating characteristic curve ($\Delta$ROC-AUC)
-    between Classifier-based Latency Estimation (CBLE), Woody
-    CBLE (WCBLE),	and their first-stage classifier (tLDA).
+		\subref{fig:covert-align/single-trial-roc-auc-diff} Difference in cross-validated
+    single-trial classification \acl{rocauc} ($\Delta$ROC-AUC)
+    between \ac{cble}, \ac{wcble}, and their first-stage classifier \ac{tlda}.
     95\% confidence intervals were determined using $k=1000$ bootstrapping.
-    Our proposed WCBLE decoder outperforms tLDA and CBLE for covert and split
-		($d=3$)	visuospatial attention (VSA). CBLE scores on par with tLDA.
-		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
+    Our proposed \ac{wcble} decoder outperforms \ac{tlda} and \ac{cble} for covert and split
+    ($d=3$)	\ac{vsa}. \ac{cble} scores on par with \ac{tlda}.
+		\subref{fig:covert-align/block-eval} Cross-validated target selection accuracy for
     all decoders plotted as a function of the number of test repetitions in
     different VSA conditions. Significance was determined using one-sided
-    (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
-    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves the highest
-    covert VSA target selection accuracy.}
+    (\ac{wcble} $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
+    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
+    covert \ac{vsa} target selection accuracy.}
 \end{figure}
@@
 
-\Cref{fig:block_evaluation} shows the cross-validated \ac{bci} selection
+\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
 accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
@@
 \subsection{Gaze-independence through cross-condition transfer}
-\label{sec:cross_results}
+\label{sec:covert-align/results/cross}
 To further back our claim of gaze-independence in the case where eye motor
@@
 
-Figure~\ref{fig:covert_cross_eval} and Figure~\ref{fig:aloise2012_cross_eval} show
+Figure~\ref{fig:covert-align/covert-cross-eval} and Figure~\ref{fig:covert-align/aloise2012-cross-eval} show
 the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
 between the investigated decoders.
-In this evaluation, bad epochs were rejected as in \cref{sec:block_accuracy}.
+In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
 \begin{figure*}
@@
 		\caption{}
-		\label{fig:covert_cross_eval}
+		\label{fig:covert-align/covert-cross-eval}
 		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
@@
 		\caption{}
-		\label{fig:aloise2012_cross_eval}
+		\label{fig:covert-align/aloise2012-cross-eval}
 		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
@@
   \caption[Cross-condition classifier performance and estimated jitter.]{
-		\subref{fig:covert_cross_eval},\subref{fig:aloise2012_cross_eval}
+		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
     Difference in cross-validated area under the receiver
@@
 
-\cref{fig:single_trial_roc_auc_diff} shows that \ac{wcble} significantly improves
+\cref{fig:covert-align/single-trial-roc-auc-diff} shows that \ac{wcble} significantly improves
 covert \ac{vsa} decoding.
@@
 Although \ac{wcble} did not significantly improve overt \ac{vsa} single-trial decoding,
-\cref{fig:block_evaluation} shows that this does not negatively impact target
+\cref{fig:covert-align/block-eval} shows that this does not negatively impact target
 selection accuracy.
@@
 evaluated per session and trained on all combined \ac{vsa} conditions as in
-\cref{sec:cross_results}.
+\cref{sec:covert-align/results/cross}.
 Since all conditions have the P3 in common, the estimated latencies can be

### 2024-11-14 chapters/gaze_independence.tex
Self-plagiarism check
Started processing comments MDV

@@
 }
+
+\noindent\emph{The last paragraph of \cref{sec:gaze-independence/sota},
+\cref{sec:gaze-independence/oculomotor/gaze}, and the first paragraph of
+\cref{sec:gaze-independence/objectives} were adapted from
+\textcite{VanDenKerchove2024}.}
 
@@
 \subsection{Gaze impairment and \acl{lis}}
-
+\label{sec:gaze-independence/oculomotor/gaze}
 \newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
@@
 If they have eye motor impairment that affects their use of visual \ac{bci} or
-eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).
+eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze-independence/venn}).
 
@@
   }
-  \label{fig:gaze_independence/venn}
+  \label{fig:gaze-independence/venn}
 \end{figure}
@@
 \section{Research objectives}
+\label{sec:gaze-independence/objectives}
 

### 2024-11-14 chapters/patients.tex
Self-plagiarism check
Started processing comments MDV

@@
 \end{table}
-\todo{where these values corrected for distance?}
 
@@
 
-\subsection{\Ac{bci} stimulation}
+\subsection{\Acs{bci} stimulation}
 
@@
 }{%
-  Distribution of the recorded gaze position.
-}{%
+  \caption[Distribution of the recorded gaze position.]{%
     Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
@@
     near the middle of the screen, and PC4 near the bottom.
-}{fig:patients/gaze}
+  }
+  \label{fig:patients/gaze}
+}
 
@@
 
-\begin{figure}[t]
-  \input{figures/patients/fig_decode.pgf}
-  \caption[%
-    Decoding performance in different \acs{vsa} settings.
-  ]{%
-    Decoding performance in different \ac{vsa} settings reported as
-    single-trial \ac{rocauc}.
-    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
-    setting.
-    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
-    be improved with the \ac{wcble} decoder.
-    95\% confidence intervals were calculated using 1000 bootstrapping
-    repetitions.
-  }
-  \label{fig:patients/decode}
-\end{figure}
-
 In the overt \ac{vsa} setting, the evaluated decoders performed similarly on average
@@
 
+\begin{figure}
+  \input{figures/patients/fig_decode.pgf}
+  \caption[%
+    Decoding performance in different \acs{vsa} settings.
+  ]{%
+    Decoding performance in different \ac{vsa} settings reported as
+    single-trial \ac{rocauc}.
+    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
+    setting.
+    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
+    be improved with the \ac{wcble} decoder.
+    95\% confidence intervals were calculated using 1000 bootstrapping
+    repetitions.
+  }
+  \label{fig:patients/decode}
+\end{figure}
+
 \subsection{Cross-condition calibration}
@@
 
-\begin{figure}[t]
+\begin{figure}
   \input{figures/patients/fig_cross.pgf}

### 2024-11-14 chapters/stbf_struct.tex
Started processing comments MDV

@@
   \ac{bci} paradigms, including the P3 oddball paradigm, \ac{ssvep},
-  \ac{cvep}~\cite{Wittevrongel2017a}, \ac{mvep}~\cite{Libert2021}.
+  \ac{cvep}~\cite{Wittevrongel2017a}, and the \ac{mvep}~\cite{Libert2021}.
 
@@
 	classifier training time, thereby considerably reducing memory usage.
-	Similar structured regularization approaches have been applied to other linear
+	Similarly structured regularization approaches have been applied to other linear
 	\ac{erp} classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
 	significant increases in performance.
-	Additionally, we show that regularization results in an interpretable
-	classification model, which can aid in analyzing and developing spatiotemporal beamformer-based classifiers.
+%	Additionally, we show that regularization results in an interpretable
+%	classification model, which can aid in analyzing and developing spatiotemporal beamformer-based classifiers.
 
@@
 	The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
-	data-driven approaches proposed by Treder et al.~\cite{Treder2016} and
+	data-driven approaches proposed by \textcite{Treder2016} and
 	Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.
@@
 	For the following analyses, we assume that all \ac{eeg} channels are normalized with zero mean and unit variance without loss of generality.
-	Solving \cref{eq:minimum_variance} under the linear constraint given by
-	\cref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal \ac{lcmv}-beamformer.
+	Solving \cref{eq:stbf-struct/minimum-variance} under the linear constraint given by
+  \cref{eq:stbf-struct/linear-constraint} returns the filter weights $\mat{w}$ defining the spatiotemporal \ac{lcmv}-beamformer.
 	\begin{equation}
-    \underset{\mat{w}}{\arg\min}\mat{w}^\intercal \mat{C}
+    \argmin_\mat{w}\mat{w}^\intercal \mat{C}
 		\mat{w}^\intercal
-		\label{eq:minimum_variance}
+		\label{eq:stbf-struct/minimum-variance}
 	\end{equation}
@@
 		\mat{a}^\intercal\mat{w} = 1
-		\label{eq:linear_constraint}
+		\label{eq:stbf-struct/linear-constraint}
 	\end{equation}
@@
 		\frac{1}{|\text{class 2}|}\sum_\text{class 2}\mathbf{x_i}
-		\label{eq:activation_pattern}
+		\label{eq:stbf-struct/ap}
 	\end{equation}
@@
 	The method of Lagrange multipliers then gives the closed-form solution to the minimization problem posed by
-	\cref{eq:minimum_variance} and \cref{eq:linear_constraint} as:
+	\cref{eq:stbf-struct/minimum-variance} and \cref{eq:stbf-struct/linear-constraint} as:
 	\begin{equation}
@@
     \frac{\mat{C}^{-1}\mat{a}^\intercal}
-		{\mat{a}C^{-1}\mat{a}^\intercal}
-		\label{eq:closed_form}
+    {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
+		\label{eq:stbf-struct/closed-form}
 	\end{equation}
@@
 		y_i = \mathbf{w}\mathbf{x}_i
-		\label{eq:apply_beamformer}
+		\label{eq:stbf-struct/apply-bf}
 	\end{equation}
 	resulting in a scalar output per epoch.
-	The linear constraint in \cref{eq:linear_constraint} ensures that the
+	The linear constraint in \cref{eq:stbf-struct/linear-constraint} ensures that the
 	beamformer maps epochs containing a target response to a score close to one
@@
 	\subsubsection{Empirical covariance estimation}
-	\label{sec:empirical_covariance}
+	\label{sec:stbf-struct/methods/emp-cov}
 	Earlier spatiotemporal beamformer studies~\cite{Wittevrongel2016,
@@
 		\frac{1}{N-1}\sum^{N}_{i=1}\mat{x}_i\mat{x}_i^\intercal
-		\label{eq:empirical_covariance}
+		\label{eq:stbf-struct/emp-cov}
 	\end{equation}
@@
     \widehat{\mat{C}^{-1}}_\text{emp} = \hat{\mat{C}}_\text{emp}^+
-		\label{eq:empirical_inverse_covariance}
+		\label{eq:stbf-struct/emp-inv-cov}
 	\end{equation}
@@
 	\subsubsection{Shrunk covariance estimation}
-	\label{sec:shrunk_covariance}
+	\label{sec:stbf-struct/methods/shrunk-cov}
 	The shrinkage covariance estimator creates a better-conditioned inversion matrix problem and generally performs better when applied to unseen data.
@@
     \hat{\mat{C}}^+_\alpha
-		\label{eq:shrinkage_inverse}
+		\label{eq:stbf-struct/shrinkage-inv}
 	\end{equation}
@@
 	because it generalizes to structured covariance estimation as described in
-	\cref{seq:structured_estimation}.
+	\cref{sec:stbf-struct/methods/structured-estimation}.
 
 	\subsubsection{Spatiotemporal beamforming with Kronecker-Toeplitz structured covariance}
-	\label{seq:structured_estimation}
+	\label{sec:stbf-struct/methods/structured-estimation}
 	Exploiting prior knowledge about the spatiotemporal structure of the \ac{eeg} signal leads to a more regularized estimator of the covariance.
@@
 	\end{equation}
-	\Cref{prop:inverse_kronecker} then leads to
-	\cref{eq:cov_inverse_kronecker} to estimate the inverse
+	\Cref{prop:stbf-struct/inverse-kronecker} then leads to
+	\cref{eq:stbf-struct/cov-inverse-kronecker} to estimate the inverse
 	covariance.
@@
     matrices $\mat{U}$ and $\mat{V}$~\cite{Langville2004}.
-		\label{prop:inverse_kronecker}
+		\label{prop:stbf-struct/inverse-kronecker}
 	\end{property}
@@
     \widehat{\mat{C}^{-1}}_\text{struct} = \hat{\mat{S}}^+ \otimes \hat{\mat{T}}^+
-		\label{eq:cov_inverse_kronecker}
+		\label{eq:stbf-struct/cov-inverse-kronecker}
 	\end{equation}
 
-	Finally, based on \cref{prop:kronecker_multiplication},
-	\cref{eq:closed_form} can be reformulated more efficiently as
-	\cref{eq:closed_form_kronecker}.
+	Finally, based on \cref{propr:stbf-struct/kron-multiplication},
+	\cref{eq:stbf-struct/closed-form} can be reformulated more efficiently as
+	\cref{eq:stbf-struct/closed-form-kron}.
 	\begin{property}
@@
     $\mat{V}\in\mathbb{R}^{Q\times Q}$ and $\mat{W}\in\mathbb{R}^{P\times Q}$~\cite{Loan2000}.
-		\label{prop:kronecker_multiplication}
+		\label{propr:stbf-struct/kron-multiplication}
 	\end{property}
@@
     {\mat{a}\cdot\text{vec}(\hat{\mat{S}}^+A^\intercal\hat{\mat{T}}^+)}
-		\label{eq:closed_form_kronecker}
+		\label{eq:stbf-struct/closed-form-kron}
 	\end{equation}
 
-	Using \cref{eq:closed_form_kronecker} removes the need to calculate the
+	Using \cref{eq:stbf-struct/closed-form-kron} removes the need to calculate the
   full, high dimensional Kronecker product $\hat{\mat{S}}^+\otimes
@@
 	\subsubsection{Kronecker-Toeplitz structured covariance estimation}
-	\label{sec:structured_covariance}
-	The question remains how to estimate $\hat{S}$ and $\hat{T}$.
+	\label{sec:stbf-struct/methods/struct-cov}
+  The question remains how to estimate $\mat{\hat{S}}$ and $\mat{\hat{T}}$.
 	While the Flip-Flop and Non-iterative Flip-Flop
@@
 	next iteration.
+
 	Combined, this leads to the iterative estimation algorithm described by the
 	following equations:
-
 	\begin{subequations}
@@
       \sum^N_{i=1}\mat{X}_i^\intercal\hat{\mat{T}}_k^+\mat{X}_i
-			\label{eq:fpi_spatial}
+			\label{eq:stbf-struct/fpi-spatial}
 		\end{equation}
@@
       \sum^N_{i=1}\mat{X}_i\hat{\mat{S}}_k^+\mat{X}_i^\intercal
-			\label{eq:fpi_temporal}
+			\label{eq:stbf-struct/fpi-temporal}
 		\end{equation}
 	\end{subequations}
-
 	\begin{subequations}
@@
       +\beta_{k+1}\frac{\text{Tr}(\tilde{\mat{S}}_{k+1})}{C}\mathbb{I}
-			\label{eq:fpi_spatial_shrunk}
+			\label{eq:stbf-struct/fpi-spatial-shrunk}
 		\end{equation}
@@
       +\gamma_{k+1}\frac{\text{Tr}(\tilde{\mat{T}}_{k+1})}{S}\mathbb{I}
-			\label{eq:fpi_temporal_shrunk}
+			\label{eq:stbf-struct/fpi-temporal-shrunk}
 		\end{equation}
 	\end{subequations}
-
 	\begin{subequations}
@@
       \tilde{\mat{S}}_{k+1}^{(\beta)}
-			\label{eq:fpi_spatial_norm}
+			\label{eq:stbf-struct/fpi-spatial-norm}
 		\end{equation}
@@
       \tilde{\mat{T}}_{k+1}^{(\gamma)}
-			\label{eq:fpi_temporal_norm}
+			\label{eq:stbf-struct/fpi-temporal-norm}
 		\end{equation}
 	\end{subequations}
-
   $\hat{\mat{S}}_0$ and $\hat{\mat{T}}_0$ can be initialized to any positive definite matrix.
@@
 			}
-			\label{eq:spatial_shrinkage}
+			\label{eq:stbf-struct/spatial-shrinkage}
 		\end{equation}
@@
 			}
-			\label{eq:temporal_shrinkage}
+			\label{eq:stbf-struct/temporal-shrinkage}
 		\end{equation}
@@
 	\ac{stbf} with empirical covariance estimation (\textsc{stbf-emp}) as in
-	\cref{sec:empirical_covariance}, \ac{stbf} with
+	\cref{sec:stbf-struct/methods/emp-cov}, \ac{stbf} with
 	\ac{loocv}-shrunk covariance estimation (\textsc{stbf-shrunk}) as in
-	\cref{sec:shrunk_covariance}, and \ac{stbf} with
+	\cref{sec:stbf-struct/methods/shrunk-cov}, and \ac{stbf} with
 	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct}) with \ac{loocv} shrinkage for
-	the Kronecker factors as in \cref{sec:structured_covariance}.
+	the Kronecker factors as in \cref{sec:stbf-struct/methods/struct-cov}.
 
@@
 	in
-  \crefrange{eq:fpi_spatial}{eq:fpi_temporal_norm}
+  \crefrange{eq:stbf-struct/fpi-spatial}{eq:stbf-struct/fpi-temporal-norm}
   estimates the Kronecker-Toeplitz structured covariance for the
@@
 	corresponding number of blocks in the training folds and dropping the rest.
-	\Cref{fig:accuracy} and \cref{fig:accuracy_ap} show each classifier's
+	\Cref{fig:accuracy} and \cref{fig:stbf-struct/accuracy-ap} show each classifier's
 	accuracy relative to the data availability.
@@
 	the first five trials of a block.
-	Results validated on one trial are reported in \cref{tab:p_values_1}, 2-trial results
-	in \cref{tab:p_values_2}, and 5-trial results in \cref{tab:p_values_5}.
+	Results validated on one trial are reported in \cref{tab:stbf-struct/p-values-1}, 2-trial results
+	in \cref{tab:stbf-struct/p-values-2}, and 5-trial results in \cref{tab:stbf-struct/p-values-5}.
 
@@
 			3 trials of testing stimuli.
-			\cref{fig:accuracy_ap} contains results for all numbers of trials.
+			\cref{fig:stbf-struct/accuracy-ap} contains results for all numbers of trials.
 			While \textsc{stbf-emp} is unstable when little training data
@@
   }{%
-    Accuracy averaging over different numbers of trials.
-  }{%
+    \caption[Accuracy averaging over different numbers of trials.]{%
       Accuracy of the different classifiers for all 21 subjects relative to the
@@
 	    numbers of trials, ranging from 1 to 15.
-
-  }{fig:accuracy_ap}
+    }
+    \label{fig:stbf-struct/accuracy-ap}
+  }
 
@@
 			and marked in grey.}
-		\label{tab:p_values_1}
+		\label{tab:stbf-struct/p-values-1}
     \bigskip
@@
 		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
-      $p$-values as in \cref{tab:p_values_1} averaging over two testing trials.}
-		\label{tab:p_values_2}
+      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over two testing trials.}
+		\label{tab:stbf-struct/p-values-2}
     \bigskip
@@
     \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
-      $p$-values as in \cref{tab:p_values_1} averaging over five testing trials.}
-		\label{tab:p_values_5}
+      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over five testing trials.}
+		\label{tab:stbf-struct/p-values-5}
 	\end{table}
@@
 	training times for the \textsc{stbf}-based as evidenced by the complexity
-	calculations in \cref{sec:parameters_complexity}.
-	\cref{fig:training_time} shows the measured training times.
+	calculations in \cref{sec:stbf-struct/discussion/param-complex}.
+	\cref{fig:stbf-struct/training-time} shows the measured training times.
 	These results were obtained using a laptop with an Intel® Core™ i7-8750H CPU and 16GB of RAM.
@@
 			\textsc{xdawn+rg}.}
-		\label{fig:training_time}
+		\label{fig:stbf-struct/training-time}
 	\end{figure}
 
-	\cref{fig:training_time} shows that the training time of
+	\cref{fig:stbf-struct/training-time} shows that the training time of
 	\textsc{stbf-struct} increases less steeply than that of \textsc{stbf-shrunk}
@@
 	\subsection{Classification accuracy}
-	As evidenced by \cref{fig:accuracy} and Tables \ref{tab:p_values_1}-\ref{tab:p_values_5}, the regularized classifiers \textsc{stbf-shrunk}
+	As evidenced by \cref{fig:accuracy} and Tables \ref{tab:stbf-struct/p-values-1}-\ref{tab:stbf-struct/p-values-5}, the regularized classifiers \textsc{stbf-shrunk}
 	and \textsc{stbf-struct} significantly improve the classification accuracy
@@
 	covariance.
-	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:parameters_complexity}), increasing the stability of matrix inversions.
+	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:stbf-struct/discussion/param-complex}), increasing the stability of matrix inversions.
 
@@
 	When more data are available, \textsc{stbf-struct} reaches significantly better accuracies.
-	Combined with the benefits laid out in \cref{sec:parameters_complexity}, this
+	Combined with the benefits laid out in \cref{sec:stbf-struct/discussion/param-complex}, this
   makes it an attractive option for \ac{erp} classification.
@@
 	\subsection{Time and memory complexity}
-	\label{sec:parameters_complexity}
+	\label{sec:stbf-struct/discussion/param-complex}
 	As mentioned above, inverting the full $CS \times CS$ dimensional covariance
@@
 
-	First, because of \cref{prop:inverse_kronecker} and
-	\cref{prop:kronecker_multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
+	First, because of \cref{prop:stbf-struct/inverse-kronecker} and
+	\cref{propr:stbf-struct/kron-multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
 	symmetric covariance and inverse covariance matrices for \textsc{stbf-struct}; they can instead be replaced by two smaller symmetric matrices respectively of dimensions $c\times c$ and $s\times s$.
@@
 	For \textsc{stbf-emp} and \textsc{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
-	Thanks to Property~\ref{prop:inverse_kronecker}, the complexity can be reduced to
+	Thanks to Property~\ref{prop:stbf-struct/inverse-kronecker}, the complexity can be reduced to
 	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \textsc{stbf-struct}.
-	The results presented in \cref{fig:training_time} confirm these
+	The results presented in \cref{fig:stbf-struct/training-time} confirm these
 	calculations.
@@
 
-	The results shown in \cref{fig:training_time} also confirm that the
+	The results shown in \cref{fig:stbf-struct/training-time} also confirm that the
 	\ac{stbf}-based estimators are very fast to train compared to the
@@
 	%	This is not unsurprising, since shrinkage transforms the covariance matrix closer to the
-	%	identity matrix and assuming identity covariance in \cref{eq:closed_form} yields weights
+	%	identity matrix and assuming identity covariance in \cref{eq:stbf-struct/closed-form} yields weights
 	%	identical to the activation pattern (up to a scaling factor).

### 2024-11-14 chapters/wcble.tex
Started processing comments MDV

@@
 }{%
-  Schematic representation of the \acs{wcble} training phase.
-}{%
+  \caption[Schematic representation of the \acs{wcble} training phase.]{%
     Schematic representation of the \acf{wcble} training phase.
@@
     classifier.
-}{fig:diagram-train}
+  }
+  \label{fig:diagram-train}
+}
 The process halts after a fixed number of iterations or when the estimated set
@@
 in~\cref{sec:robust-latency}.
-\fullpagefig{
+\fullpagefig{%
   \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
-
 }{%
-  Schematic representation of the \acs{wcble} test phase.
-}{%
+  \caption[Schematic representation of the \acs{wcble} test phase.]{%
     Schematic representation of the \acf{wcble}. (1) The
@@
     trained second-stage classifier, which predicts the label of the epochs.
-}{fig:diagram-eval}
+  }
+  \label{fig:diagram-eval}
+}
 \section{Results}
@@
 
+% ERROR BETWEEN HERE ==========================================================
 100 target epochs containing the evoked potential and noise and 100
@@
 \fullpagefig{%
-  \sffamily
-  \sansmath
   \inputpgf{figures/wcble}{simulated-sine.pgf}
 }{%
-  Simulated target and non-target evoked data.
-}{%
+  \caption[Simulated target and non-target evoked data.]{%
     Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
@@
     plotted here.
-}{fig:wcble/sim-sine}
+  }
+  \label{fig:wcble/sim-sine}
+}
 
@@
 
+% AND HERE ====================================================================
 As the \ac{snr} decreases, both \ac{cble} and \ac{wcble} can accurately

### 2024-11-14 cover_kul.tex
Started processing comments MDV

@@
-\newgeometry{margin=1.5cm}
 \begin{titlepage}
-
     \sffamily
-    \begin{minipage}[t]{.3\textwidth}
+    \begin{minipage}[t]{.5\textwidth}
     \vspace{0pt}
@@
     Biomedical Sciences Group \\
+    Doctoral School of Biomedical Sciences \\
     Faculty of Medical Science \\
     Department of Neurosciences
+    \bigskip
+
+    University of Lille \\
+    Doctoral School MADIS \\
+    UMR CRIStAL \\
 
@@
       \includegraphics[width=\textwidth]{figures/kul_docschool.png}
-      \bigskip
+      \smallskip
 
@@
     \begin{minipage}[b]{.5\textwidth}
+      \raggedright
+      \small
       Jury: \\
-      Supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
-      Supervisor University of Lille: \emph{prof. François Cabestaing} \\
+      Co-supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
+      Co-supervisor Univ. Lille: \emph{prof. François Cabestaing} \\
       Chair examining committee: \emph{prof. John Creemers} \\
-      Chair public defence: \emph{prof. Koen Poesen} \\
-      Jury members: \\
-      \emph{prof. Maarten De Vos} \\
-      \emph{prof. Adalberto Simeone} \\
+      Chair public defense: \emph{prof. Koen Poesen} \\
       \emph{prof. Andrea K\"ubler} \\
-      \emph{prof. Fabien Lotte} \\
-      \emph{prof. Reinhold Scherer}
+      \emph{prof.\ Fabien Lotte} \\
+      \emph{prof.\ Reinhold Scherer} \\
+      \emph{prof.\ Maarten De Vos} \\
+      \emph{prof.\ Adalberto Simeone}
     \end{minipage}\hfill%
-    \begin{minipage}[b]{.31\textwidth}\raggedright
+    \begin{minipage}[b]{.31\textwidth}
+        \raggedright
+        \small
         Dissertation presented in partial fulfillment of the requirements for
@@
 
-    \centering\today
+    \centering
+    Defended December 16, 2024
 
 \end{titlepage}
-\restoregeometry

### 2024-11-14 cover_ulille.tex
Started processing comments MDV

\begin{titlepage}
    \sffamily
    \begin{minipage}[t]{.5\textwidth}
    \vspace{0pt}
    Universit\'e de Lille \\
    \'Ecole Doctorale MADIS \\
    UMR CRIStAL
    \bigskip

    KU Leuven \\
    Groupe des Sciences Biom\'edicales \\
    \'Ecole Doctorale des Sciences Biom\'edicales \\
    Facult\'e des Sciences M\'edicales\\
    D\'epartment des Neurosciences

    \end{minipage}\hfill%
    \begin{minipage}[t]{.3\textwidth}
      \vspace{0pt}
      \includegraphics[width=\textwidth]{figures/ulille.png}
      \smallskip

      \includegraphics[width=\textwidth]{figures/edmadis.pdf}
      \smallskip

      \includegraphics[width=\textwidth]{figures/kul.png}
    \end{minipage}

    \vfill

    \noindent\makebox[\textwidth][c]{%
      \begin{minipage}{.85\textwidth}
        \centering
        {%
          \Huge
          \bfseries
          Interface Cerveau-Ordinateur visuelle pour la communication sans le regard
        }
        \bigskip

        \MakeUppercase{%
          \Large
          Algorithmes \& Applications
        }
        \bigskip

        {%
          \Large
          Arne  Van Den Kerchove
        }

      \end{minipage}%
    }

    \vfill
      \small
      \noindent
      Th\`ese soutenue le 16 d\'ecembre 2024 en satisfaction partielle des exigences de double
      dipl\^ome de Docteur en Traitement du Signal et des Images
      (Universit\'e de Lille) et Docteur en Sciences Biom\'edicales (KU Leuven)
      devant le jury compos\'e de: \\
      \smallskip

      \noindent
      Co-directeur de th\`ese: \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
      Co-directeur de th\`ese: \emph{Prof.\ Marc Van Hulle, KU Leuven, D\'epartment des Neurosciences} \\
      Pr\'esident: \emph{Prof.\ Koen Poesen, KU Leuven, D\'epartment des Neurosciences} \\
      Rapporteuse: \emph{Prof.\ Andrea K\"ubler, Univ. W\"urzburg, Institut de Psychologie} \\
      Rapporteur: \emph{Prof.\ Fabien Lotte, Inria Bordeaux, LaBRI} \\
      Rapporteur: \emph{Prof.\ Reinhold Scherer, Univ. Essex, École d'informatique et de G\'enie \'Electronique} \\
      Examinateur: \emph{Prof.\ Maarten De Vos, KU Leuven,  D\'epartment de G\'enie \'Electrique} \\
      Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique}
\end{titlepage}

### 2024-11-14 code/patients/fig_cross.py

@@ -1,59 +1,0 @@
-import pdb
-
-import matplotlib.pyplot as plt
-import pandas as pd
-import seaborn as sns
-from setup import colors, configure_matplotlib_style, save_pgf_trim
-
-configure_matplotlib_style()
-
-df = pd.read_csv('data/patients/results_cross.csv')
-df = df[df["subject"] != "PD01"]
-df = df[df["subject"] != "PB03"]
-print(df)
-
-df_avg = df.copy()
-df_avg["subject"] = "avg."
-df = pd.concat([df, df_avg], ignore_index=True)
-
-df = df.replace({
-    'PA01': 'PA1',
-    'PB01': 'PB1',
-    'PB02': 'PB2',
-    'PB04': 'PB4',
-    'PC02': 'PC2',
-    'PC03': 'PC3',
-    'PC04': 'PC4',
-    'overt': 'overt VSA',
-    'covert': 'covert VSA',
-    'free': 'free VSA',
-})
-
-df['test setting'] = df['cond_test']
-df['train setting'] = df['cond_train']
-
-fg = sns.catplot(
-    data=df,
-    x="subject",
-    y="test_score",
-    hue="train setting",
-    kind="bar",
-    col="test setting",
-    legend="auto",
-    legend_out=False,
-    col_order=['overt VSA', 'covert VSA', 'free VSA']
-)
-fg.set(ylim=(0, 1))
-for ax in fg.axes.flat:
-    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
-    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
-fg.set_axis_labels("patient", "ROC-AUC")
-fg.set_titles("{col_name}")
-
-
-fg._legend.set_frame_on(True)
-fg._legend.get_frame().set_facecolor("white")
-fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
-
-fig = fg.figure
-save_pgf_trim(fig, fg.axes[0,0], "figures/patients/fig_cross.pgf", height=2)

### 2024-11-14 code/patients/fig_decode.py

@@ -1,61 +1,0 @@
-import pdb
-
-import matplotlib.pyplot as plt
-import pandas as pd
-import seaborn as sns
-from setup import colors, configure_matplotlib_style, save_pgf_trim
-
-configure_matplotlib_style()
-
-df = []
-for pipeline in ["tLDA", "WCBLE", "XDAWNCov+TS+LDA"]:
-    df_pipe = pd.read_csv(f"data/patients/decode/decode_results_{pipeline}.csv")
-    df_pipe["pipeline"] = pipeline
-    df.append(df_pipe)
-df = pd.concat(df, ignore_index=True)
-df = df[df["subject"] != "PD01"]
-df = df[df["subject"] != "PB03"]
-
-df_avg = df.copy()
-df_avg["subject"] = "avg."
-df = pd.concat([df, df_avg], ignore_index=True)
-
-df = df.replace({
-    'PA01': 'PA1',
-    'PB01': 'PB1',
-    'PB02': 'PB2',
-    'PB04': 'PB4',
-    'PC02': 'PC2',
-    'PC03': 'PC3',
-    'PC04': 'PC4',
-
-})
-
-fg = sns.catplot(
-    data=df,
-    x="subject",
-    y="test_score",
-    hue="pipeline",
-    kind="bar",
-    col="condition",
-    legend="auto",
-    legend_out=False,
-)
-fg.set(ylim=(0, 1))
-plt.legend(
-    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
-)  # Adjust ncol to fit all labels in one line
-
-fg._legend.set_frame_on(True)
-fg._legend.get_frame().set_facecolor("white")
-fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
-
-
-
-for ax in fg.axes.flat:
-    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
-    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
-fg.set_axis_labels("patient", "ROC-AUC")
-fg.set_titles("{col_name} VSA")
-fig = fg.figure
-save_pgf_trim(fig, fg.axes[0,0], "figures/patients/fig_decode.pgf", height=2)

### 2024-11-14 code/patients/fig_gaze.py

@@ -1,108 +1,0 @@
-import matplotlib.colors as mcolors
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import seaborn as sns
-from matplotlib.patches import Rectangle
-from mne import read_epochs
-from setup import colors, configure_matplotlib_style, pagesize, savefig_kws
-
-configure_matplotlib_style()
-
-
-subjects = {
-    "PA1": "PA01",
-    "PB1": "PB01",
-    "PB2": "PB02",
-    "PB4": "PB04",
-    "PC4": "PC04",
-}
-conditions = ["overt", "covert", "free"]
-
-
-N_TARGETS = 6
-# INTERFACE_RADIUS = 12.28 / 2  # degrees of FOV
-INTERFACE_RADIUS = 20 / 2  # degrees of FOV
-TARGET_RADIUS = 4.15 / 2  # degrees of FOV
-TARGET_INTENSE_RADIUS = 5.60 / 2  # degrees of FOV
-FONT_SIZE = 0.5  # degrees of FOG
-
-
-def get_target_pos(pos):
-    if pos == "center":
-        return [0, 0]
-    elif isinstance(pos, int) and pos >= 0 and pos < 6:
-        angle = pos * 2 * np.pi / N_TARGETS
-        x = INTERFACE_RADIUS * np.cos(angle)
-        y = INTERFACE_RADIUS * np.sin(angle)
-        return [x, y]
-    else:
-        raise ValueError
-
-
-xedges = np.load("data/patients/gaze/gaze_xedges.npy")
-yedges = np.load("data/patients/gaze/gaze_yedges.npy")
-
-fig, axs = plt.subplots(
-    len(subjects),
-    len(conditions),
-    sharex=True,
-    sharey=True,
-    figsize=pagesize,
-    layout="constrained",
-)
-axs = axs.reshape((len(subjects), len(conditions)))
-for i, (subject, sub) in enumerate(subjects.items()):
-    for j, condition in enumerate(conditions):
-        axs[i, j].set_title(f"{subject}, {condition} VSA")
-        hist = np.load(f"data/patients/gaze/gaze_sub-{sub}_cond-{condition}.npy")
-        axs[i, j].pcolormesh(xedges, yedges, hist.T)
-        axs[i, j].set_aspect("equal")
-        axs[i, j].set_axis_off()
-        for target in ["center", 0, 1, 2, 3, 4, 5]:
-            pos = get_target_pos(target)
[…]

### 2024-11-14 code/setup.py

@@ -1,124 +1,0 @@
-import matplotlib.pyplot as plt
-import tikzplotlib
-from cycler import cycler
-
-mm = 1 / 25.4  # centimeters in inches
-pagesize = (160 * mm, 240 * mm)
-# colors
-colors = dict(
-    darkgray="#8D807A",
-    lightgray="#E0E0E0",
-    accent1="#F29018",
-    accent2="#F63D1E",
-    accent3="#78479D",
-)
-
-
-
-
-# Function to configure matplotlib styles
-def configure_matplotlib_style():
-    custom_color_palette = [colors[c] for c in ["accent1", "accent2", "accent3"]]
-
-    # Configure global settings
-    font_size = 9
-
-    plt.rcParams.update(
-        {
-            "text.usetex": True,  # use LaTeX to write all text
-            "pgf.texsystem": "pdflatex",
-            #"axes.grid": False,  # Disable grid lines
-            "pgf.rcfonts": False,
-            "axes.titlelocation": "left",  # Axis title alignment left
-            "figure.facecolor": "white",  # Background color for the figure
-            "image.cmap": "inferno",  # Set default colormap to Plasma
-            "axes.prop_cycle": cycler(
-                color=custom_color_palette
-            ),  # Custom color palette
- 
-            "figure.titlesize": 11,  # Figure title font size
-            "figure.titleweight": "bold",  # Figure title bold
-
-            "xtick.direction": "in",  # X-axis ticks inward
-            "ytick.direction": "in",  # Y-axis ticks inward
-            "xtick.major.size": 3,  # Major tick size for x-axis
-            "ytick.major.size": 3,  # Major tick size for y-axis
- 
-
-            'text.latex.preamble' : "\\usepackage{sansmath}\\sansmath",
-            "axes.edgecolor": colors["darkgray"],  # Spine color
-            "axes.labelcolor": colors["darkgray"],  # Axis label color
-            "axes.titlesize": font_size,  # Axis title font size
-            # "axes.titleweight": "bold",  # Axis title bold
-            "axes.titlecolor": colors["darkgray"],  # Axis title color
-            "axes.labelsize": font_size,  # Axis label font size
-            "xtick.color": colors["darkgray"],  # X-axis tick color
-            "ytick.color": colors["darkgray"],  # Y-axis tick color
-            "xtick.labelsize": font_size,  # X-axis tick label size
-            "ytick.labelsize": font_size,  # Y-axis tick label size
[…]

### 2024-11-14 code/wcble/fig_latency_classify.py

@@ -1,58 +1,0 @@
-import matplotlib.pyplot as plt
-import pdb
-import pandas as pd
-import seaborn as sns
-from setup import configure_matplotlib_style,textwidth_in, save_pgf_trim
-
-configure_matplotlib_style()
-
-
-# Latency
-df = pd.read_csv("data/wcble/results_align.csv")
-df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
-df = df[df["stc"] == "sine"]
-
-facet_kws = dict(ylim=[0, 1], xlim=[0, -31], despine=False, legend_out=False,
-                 )
-fg = sns.relplot(
-    data=df,
-    x="snr",
-    y="rho",
-    col="jitter",
-    # row="stc",
-    hue="method",
-    kind="line",
-    facet_kws=facet_kws,
-)
-fg.set_titles(col_template="$\\sigma={col_name}$s")
-fg.set_xlabels("SNR (dB)")
-fg.set_ylabels("$\\rho$")
-save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_latency.pgf', height=2)
-
-
-# Accuracy
-df = pd.read_csv("data/wcble/results_classify.csv")
-df = pd.concat(
-    [df, pd.read_csv("data/wcble/results_classify_tlda.csv")], ignore_index=True
-)
-df = df[df["stc"] == "sine"]
-df["accuracy"] *= 100
-
-df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
-
-facet_kws = dict(ylim=[0, 100], xlim=[0, -31], despine=False, legend_out=False)
-fg = sns.relplot(
-    data=df,
-    x="snr",
-    y="accuracy",
-    col="jitter",
-    # row="stc",
-    hue="method",
-    kind="line",
-    facet_kws=facet_kws,
-)
-fg.set_titles(col_template="$\\sigma={col_name}$s")
-fg.set_xlabels("SNR (dB)")
-fg.set_ylabels("accuracy (%)")
-
-save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_accuracy.pgf', height=2)

### 2024-11-14 scripts/patients/fig_cross.py

@@ -1,0 +1,59 @@
+import pdb
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import colors, configure_matplotlib_style, save_pgf_trim
+
+configure_matplotlib_style()
+
+df = pd.read_csv('data/patients/results_cross.csv')
+df = df[df["subject"] != "PD01"]
+df = df[df["subject"] != "PB03"]
+print(df)
+
+df_avg = df.copy()
+df_avg["subject"] = "avg."
+df = pd.concat([df, df_avg], ignore_index=True)
+
+df = df.replace({
+    'PA01': 'PA1',
+    'PB01': 'PB1',
+    'PB02': 'PB2',
+    'PB04': 'PB4',
+    'PC02': 'PC2',
+    'PC03': 'PC3',
+    'PC04': 'PC4',
+    'overt': 'overt VSA',
+    'covert': 'covert VSA',
+    'free': 'free VSA',
+})
+
+df['test setting'] = df['cond_test']
+df['train setting'] = df['cond_train']
+
+fg = sns.catplot(
+    data=df,
+    x="subject",
+    y="test_score",
+    hue="train setting",
+    kind="bar",
+    col="test setting",
+    legend="auto",
+    legend_out=False,
+    col_order=['overt VSA', 'covert VSA', 'free VSA']
+)
+fg.set(ylim=(0, 1))
+for ax in fg.axes.flat:
+    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
+    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
+fg.set_axis_labels("patient", "ROC-AUC")
+fg.set_titles("{col_name}")
+
+
+fg._legend.set_frame_on(True)
+fg._legend.get_frame().set_facecolor("white")
+fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
+
+fig = fg.figure
+save_pgf_trim(fig, fg.axes[0,0], "figures/patients/fig_cross.pgf", height=2)

### 2024-11-14 scripts/patients/fig_decode.py

@@ -1,0 +1,61 @@
+import pdb
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import colors, configure_matplotlib_style, save_pgf_trim
+
+configure_matplotlib_style()
+
+df = []
+for pipeline in ["tLDA", "WCBLE", "XDAWNCov+TS+LDA"]:
+    df_pipe = pd.read_csv(f"data/patients/decode/decode_results_{pipeline}.csv")
+    df_pipe["pipeline"] = pipeline
+    df.append(df_pipe)
+df = pd.concat(df, ignore_index=True)
+df = df[df["subject"] != "PD01"]
+df = df[df["subject"] != "PB03"]
+
+df_avg = df.copy()
+df_avg["subject"] = "avg."
+df = pd.concat([df, df_avg], ignore_index=True)
+
+df = df.replace({
+    'PA01': 'PA1',
+    'PB01': 'PB1',
+    'PB02': 'PB2',
+    'PB04': 'PB4',
+    'PC02': 'PC2',
+    'PC03': 'PC3',
+    'PC04': 'PC4',
+
+})
+
+fg = sns.catplot(
+    data=df,
+    x="subject",
+    y="test_score",
+    hue="pipeline",
+    kind="bar",
+    col="condition",
+    legend="auto",
+    legend_out=False,
+)
+fg.set(ylim=(0, 1))
+plt.legend(
+    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
+)  # Adjust ncol to fit all labels in one line
+
+fg._legend.set_frame_on(True)
+fg._legend.get_frame().set_facecolor("white")
+fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
+
+
+
+for ax in fg.axes.flat:
+    ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
+    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
+fg.set_axis_labels("patient", "ROC-AUC")
+fg.set_titles("{col_name} VSA")
+fig = fg.figure
+save_pgf_trim(fig, fg.axes[0,0], "figures/patients/fig_decode.pgf", height=2)

### 2024-11-14 scripts/patients/fig_gaze.py

@@ -1,0 +1,108 @@
+import matplotlib.colors as mcolors
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from matplotlib.patches import Rectangle
+from mne import read_epochs
+from setup import colors, configure_matplotlib_style, pagesize, savefig_kws
+
+configure_matplotlib_style()
+
+
+subjects = {
+    "PA1": "PA01",
+    "PB1": "PB01",
+    "PB2": "PB02",
+    "PB4": "PB04",
+    "PC4": "PC04",
+}
+conditions = ["overt", "covert", "free"]
+
+
+N_TARGETS = 6
+# INTERFACE_RADIUS = 12.28 / 2  # degrees of FOV
+INTERFACE_RADIUS = 20 / 2  # degrees of FOV
+TARGET_RADIUS = 4.15 / 2  # degrees of FOV
+TARGET_INTENSE_RADIUS = 5.60 / 2  # degrees of FOV
+FONT_SIZE = 0.5  # degrees of FOG
+
+
+def get_target_pos(pos):
+    if pos == "center":
+        return [0, 0]
+    elif isinstance(pos, int) and pos >= 0 and pos < 6:
+        angle = pos * 2 * np.pi / N_TARGETS
+        x = INTERFACE_RADIUS * np.cos(angle)
+        y = INTERFACE_RADIUS * np.sin(angle)
+        return [x, y]
+    else:
+        raise ValueError
+
+
+xedges = np.load("data/patients/gaze/gaze_xedges.npy")
+yedges = np.load("data/patients/gaze/gaze_yedges.npy")
+
+fig, axs = plt.subplots(
+    len(subjects),
+    len(conditions),
+    sharex=True,
+    sharey=True,
+    figsize=pagesize,
+    layout="constrained",
+)
+axs = axs.reshape((len(subjects), len(conditions)))
+for i, (subject, sub) in enumerate(subjects.items()):
+    for j, condition in enumerate(conditions):
+        axs[i, j].set_title(f"{subject}, {condition} VSA")
+        hist = np.load(f"data/patients/gaze/gaze_sub-{sub}_cond-{condition}.npy")
+        axs[i, j].pcolormesh(xedges, yedges, hist.T)
+        axs[i, j].set_aspect("equal")
+        axs[i, j].set_axis_off()
+        for target in ["center", 0, 1, 2, 3, 4, 5]:
+            pos = get_target_pos(target)
[…]

### 2024-11-14 scripts/setup.py

@@ -1,0 +1,126 @@
+import matplotlib.pyplot as plt
+#import tikzplotlib
+from cycler import cycler
+
+mm = 1 / 25.4  # centimeters in inches
+pagesize = (160 * mm, 240 * mm)
+# colors
+colors = dict(
+    darkgray="#8D807A",
+    lightgray="#E0E0E0",
+    accent1="#F29018",
+    accent2="#F63D1E",
+    accent3="#78479D",
+)
+primary_cmap = "RdBu_r"
+secondary_cmap = "PuOr_r"
+
+
+
+
+# Function to configure matplotlib styles
+def configure_matplotlib_style():
+    custom_color_palette = [colors[c] for c in ["accent1", "accent2", "accent3"]]
+
+    # Configure global settings
+    font_size = 9
+
+    plt.rcParams.update(
+        {
+            "text.usetex": True,  # use LaTeX to write all text
+            "pgf.texsystem": "pdflatex",
+            #"axes.grid": False,  # Disable grid lines
+            "pgf.rcfonts": False,
+            "axes.titlelocation": "left",  # Axis title alignment left
+            "figure.facecolor": "white",  # Background color for the figure
+            "image.cmap": "inferno",  # Set default colormap to Plasma
+            "axes.prop_cycle": cycler(
+                color=custom_color_palette
+            ),  # Custom color palette
+ 
+            "figure.titlesize": 11,  # Figure title font size
+            "figure.titleweight": "bold",  # Figure title bold
+
+            "xtick.direction": "in",  # X-axis ticks inward
+            "ytick.direction": "in",  # Y-axis ticks inward
+            "xtick.major.size": 3,  # Major tick size for x-axis
+            "ytick.major.size": 3,  # Major tick size for y-axis
+ 
+
+            'text.latex.preamble' : "\\usepackage{sansmath}\\sansmath",
+            "axes.edgecolor": colors["darkgray"],  # Spine color
+            "axes.labelcolor": colors["darkgray"],  # Axis label color
+            "axes.titlesize": font_size,  # Axis title font size
+            # "axes.titleweight": "bold",  # Axis title bold
+            "axes.titlecolor": colors["darkgray"],  # Axis title color
+            "axes.labelsize": font_size,  # Axis label font size
+            "xtick.color": colors["darkgray"],  # X-axis tick color
+            "ytick.color": colors["darkgray"],  # Y-axis tick color
+            "xtick.labelsize": font_size,  # X-axis tick label size
[…]

### 2024-11-14 scripts/stbf_struct/covs.py

@@ -1,0 +1,94 @@
+from setup import colors, configure_matplotlib_style, pagesize, primary_cmap, secondary_cmap
+
+from mcbf.alg.covariance import SpatiotemporalEmpiricalCovariance, \
+    SpatiotemporalLOOCVShrunkCovariance, KroneckerToeplitzCovariance
+from mne import read_epochs
+from mne.decoding import Scaler
+import os
+import matplotlib.pyplot as plt
+from matplotlib.gridspec import GridSpec
+import numpy as np
+
+configure_matplotlib_style()
+
+epochs = read_epochs('data/stbf_struct/sub-01_task-oddball_proc-32hz_eeg.fif')
+epochs = epochs[:100]
+epochs.pick_channels(['Fz', 'Cz', 'Pz', 'Oz'])
+epochs.crop(tmin=0.1, tmax=0.6)
+scaler = Scaler(scalings='median', with_mean=False)
+epochs_data = scaler.fit_transform(epochs.get_data())
+
+
+cov_plot_kwargs = dict(cmap=secondary_cmap, vmin=-1, vmax=1)
+inv_cov_plot_kwargs = dict(cmap=secondary_cmap)
+savefig_kwargs = dict(format='eps')
+
+emp_cov = SpatiotemporalEmpiricalCovariance()
+emp_cov.fit(epochs_data)
+shrunk_cov = SpatiotemporalLOOCVShrunkCovariance()
+shrunk_cov.fit(epochs_data)
+kron_toep_cov = KroneckerToeplitzCovariance(decimate=1)
+kron_toep_cov.fit(epochs_data)
+print(f'Shrinkage: {shrunk_cov.shrinkage_}')
+
+fig = plt.figure(constrained_layout=True, figsize=(11.1,5))
+gs = GridSpec(4, 4, figure=fig,
+              width_ratios=[2 / 7, 2 / 7, 2 / 7, 1 / 7])
+
+plt.rcParams['axes.linewidth'] = 2
+plt.rcParams['font.size'] = 12
+
+def plot_cov_pub(data, ax, label, **kwargs):
+    data = np.flip(data, axis=1)
+    opt = np.max(np.abs(data))
+    kwargs.setdefault('vmin', -opt)
+    kwargs.setdefault('vmax', opt)
+    im = ax.pcolormesh(data, cmap=secondary_cmap, **kwargs)
+    ax.set_aspect('equal')
+    #fig.colorbar(im, ax=ax)
+    ax.tick_params(left=False,
[…]

### 2024-11-14 scripts/wcble/fig_latency_classify.py

@@ -1,0 +1,58 @@
+import matplotlib.pyplot as plt
+import pdb
+import pandas as pd
+import seaborn as sns
+from setup import configure_matplotlib_style,textwidth_in, save_pgf_trim
+
+configure_matplotlib_style()
+
+
+# Latency
+df = pd.read_csv("data/wcble/results_align.csv")
+df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
+df = df[df["stc"] == "sine"]
+
+facet_kws = dict(ylim=[0, 1], xlim=[0, -31], despine=False, legend_out=False,
+                 )
+fg = sns.relplot(
+    data=df,
+    x="snr",
+    y="rho",
+    col="jitter",
+    # row="stc",
+    hue="method",
+    kind="line",
+    facet_kws=facet_kws,
+)
+fg.set_titles(col_template="$\\sigma={col_name}$s")
+fg.set_xlabels("SNR (dB)")
+fg.set_ylabels("$\\rho$")
+save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_latency.pgf', height=2)
+
+
+# Accuracy
+df = pd.read_csv("data/wcble/results_classify.csv")
+df = pd.concat(
+    [df, pd.read_csv("data/wcble/results_classify_tlda.csv")], ignore_index=True
+)
+df = df[df["stc"] == "sine"]
+df["accuracy"] *= 100
+
+df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
+
+facet_kws = dict(ylim=[0, 100], xlim=[0, -31], despine=False, legend_out=False)
+fg = sns.relplot(
+    data=df,
+    x="snr",
+    y="accuracy",
+    col="jitter",
+    # row="stc",
+    hue="method",
+    kind="line",
+    facet_kws=facet_kws,
+)
+fg.set_titles(col_template="$\\sigma={col_name}$s")
+fg.set_xlabels("SNR (dB)")
+fg.set_ylabels("accuracy (%)")
+
+save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_accuracy.pgf', height=2)

### 2024-11-14 scripts/wcble/fig_simulated.py

@@ -1,0 +1,133 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from mne import read_epochs
+from setup import configure_matplotlib_style,colors, save_pgf_trim
+import pdb
+
+configure_matplotlib_style()
+
+epochs = read_epochs("data/wcble/simulated_epochs.fif")
+meta = epochs.metadata
+
+epochs.pick(["C3", "STIM"])
+epochs.crop(-2, 2)
+
+snrs = meta["snr"].unique()
+# snrs = [0, -16, -31]
+jitters = meta["jitter"].unique()
+
+
+img_kwargs = dict(
+    vmin=-60,
+    vmax=60,
+    picks="C3",
+)
+
+
+def latency(epochs):
+    ground_truth = np.zeros(len(epochs))
+    for i, epoch in enumerate(epochs):
+        stim_chan = epoch[-1, :]
+        ground_truth[i] = np.nonzero(stim_chan)[0]
+    ground_truth = ground_truth / epochs.info["sfreq"] + epochs.tmin
+    return ground_truth
+
+
+def order(epochs, n_epochs=100):
+    latency_ = latency(epochs)
+    order = np.zeros(len(epochs), dtype=int)
+    order[:n_epochs] = np.argsort(latency_[:n_epochs])
+    order[n_epochs:] = np.arange(n_epochs) + n_epochs
+    return order
+
+
+def plot_img(ax, epochs, cbar_ax=None):
+    order_ = order(epochs)
+    if cbar_ax is None:
+        axes=ax
+    else:
+        axes=[ax, cbar_ax]
+    epochs.plot_image(
+        order=order_,
+        colorbar=cbar_ax is not None,
+        evoked=False,
+        axes=axes,
+        show=False,
+        **img_kwargs,
+    )
+    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=colors['darkgray'])
+
+    ax_img.set_title("")
+    ax_img.set_xticks([])
+    ax_img.set_xlabel("")
+    ax_img.set_yticks([])
+
+
+def plot_ts(ax, epochs):
+    x = epochs.times
+    y = epochs["non-target"].pick("C3").get_data().squeeze() * 1e6
+    df = pd.DataFrame(y.T, index=x)
+    df.index.name = "time"
+    df = df.melt(ignore_index=False, value_name="amplitude")
[…]

### 2024-11-15 chapters/stbf_struct.tex
Rework fig 1.1 and fig 1.2

@@
 	\begin{figure}
-		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
+		%\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-0.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-1.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-2.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{minipage}[b]{\textwidth/10}
+      \begin{subfigure}{\textwidth}
+  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-6.eps}
+        \caption{}
+      \end{subfigure}\vfill%
+      \begin{subfigure}{\textwidth}
+  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-7.eps}
+        \caption{}
+      \end{subfigure}\hfill%
+    \end{minipage}
+
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-3.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-4.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{subfigure}{\textwidth/4}
+		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-5.eps}
+      \caption{}
+    \end{subfigure}\hfill%
+    \begin{minipage}[b]{\textwidth/10}
+      \begin{subfigure}{\textwidth}
+  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-8.eps}
+        \caption{}
+      \end{subfigure}\vfill%
+      \begin{subfigure}{\textwidth}
+  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-9.eps}
+        \caption{}
+      \end{subfigure}\hfill%
+    \end{minipage}
+
     \caption[Estimated covariance and inverse covariance.]{Different estimators of the covariance and inverse covariance
@@
 			Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
-			(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
-			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
-			determined by the closed-form \ac{loocv} method. (\textbf{e,f}) Kronecker-Toeplitz
+			(\textbf{a,f}) Empirical covariance and inverse covariance matrices.
+			(\textbf{b,g}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
+			determined by the closed-form \ac{loocv} method. (\textbf{c,h}) Kronecker-Toeplitz
 			structured covariance and inverse covariance matrices.
-			(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
+			(\textbf{d,e}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
 			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
@@
   \begin{figure}
-		\includegraphics[width=\linewidth]{figures/stbf_struct/fpi.eps}
+    \sffamily
+    \sansmath
+    \begin{subfigure}{\linewidth}
+      \input{figures/stbf_struct/fpi-0.pgf}
+      \caption{Results for 1, 2, and 5 trials using only the first block in each
+      training fold for training.}
+    \end{subfigure}
+    \smallskip
+
+    \begin{subfigure}{\linewidth}
+      \input{figures/stbf_struct/fpi-1.pgf}
+      \caption{Results for 1, 2, and 5 trials using all nine training blocks in
+      the training folds.}
+    \end{subfigure}
     \caption[Average cross-validated \textsc{stbf-struct} accuracy]{%
@@
 			covariance. Error bars represent the first and third quartiles. The
-			accuracy does not improve when using more than one iteration.
-			(\textbf{a}) Results for 1, 2, and 5 trials using only the first block in each
-			training fold for training. (\textbf{b}) Results
-			for 1, 2, and 5 trials using
-			all nine training blocks in the training folds.
-		}
+      accuracy does not improve when using more than one iteration.}
 		\label{fig:iterations}

### 2024-11-15 scripts/setup.py

@@ -12,9 +12,7 @@ colors = dict(
     accent2="#F63D1E",
     accent3="#78479D",
 )
-primary_cmap = "RdBu_r"
-secondary_cmap = "PuOr_r"
-
+diverging_cmap = "RdYlBu_r"
 
 
 
@@ -74,10 +72,11 @@ inches_per_pt = 1 / 72.27
 textwidth_pt=345
 textwidth_in=textwidth_pt*inches_per_pt
 
-def save_pgf_trim(fig, ax,path, width=textwidth_in, height=None):
+def save_pgf_trim(fig, ax,path, width=textwidth_in, height=None, rows=1,
+                  columns=1):
     if height is None:
         golden_ratio = (5**0.5 - 1) / 2
-        height=width*golden_ratio
+        height=rows*width*golden_ratio/columns
     fig.set_constrained_layout(True)
     print(width)
     print(height)

### 2024-11-15 scripts/stbf_struct/covs.py

@@ -1,94 +1,68 @@
-from setup import colors, configure_matplotlib_style, pagesize, primary_cmap, secondary_cmap
-
-from mcbf.alg.covariance import SpatiotemporalEmpiricalCovariance, \
-    SpatiotemporalLOOCVShrunkCovariance, KroneckerToeplitzCovariance
+import matplotlib.pyplot as plt
+import seaborn as sns
+from matplotlib.gridspec import GridSpec
+from mcbf.alg.covariance import (KroneckerToeplitzCovariance,
+                                 SpatiotemporalEmpiricalCovariance,
+                                 SpatiotemporalLOOCVShrunkCovariance)
 from mne import read_epochs
 from mne.decoding import Scaler
-import os
-import matplotlib.pyplot as plt
-from matplotlib.gridspec import GridSpec
-import numpy as np
+from setup import (colors, configure_matplotlib_style, diverging_cmap,
+                   textwidth_in)
 
 configure_matplotlib_style()
 
-epochs = read_epochs('data/stbf_struct/sub-01_task-oddball_proc-32hz_eeg.fif')
+# Read and process data
+epochs = read_epochs("data/stbf_struct/sub-01_task-oddball_proc-32hz_eeg.fif")
 epochs = epochs[:100]
-epochs.pick_channels(['Fz', 'Cz', 'Pz', 'Oz'])
+epochs.pick_channels(["Fz", "Cz", "Pz", "Oz"])
 epochs.crop(tmin=0.1, tmax=0.6)
-scaler = Scaler(scalings='median', with_mean=False)
+scaler = Scaler(scalings="median", with_mean=False)
 epochs_data = scaler.fit_transform(epochs.get_data())
 
-
-cov_plot_kwargs = dict(cmap=secondary_cmap, vmin=-1, vmax=1)
-inv_cov_plot_kwargs = dict(cmap=secondary_cmap)
-savefig_kwargs = dict(format='eps')
-
+# Calculate covariances
 emp_cov = SpatiotemporalEmpiricalCovariance()
 emp_cov.fit(epochs_data)
 shrunk_cov = SpatiotemporalLOOCVShrunkCovariance()
 shrunk_cov.fit(epochs_data)
 kron_toep_cov = KroneckerToeplitzCovariance(decimate=1)
 kron_toep_cov.fit(epochs_data)
[…]

### 2024-11-15 scripts/stbf_struct/kronecker_iteration.py

@@ -1,0 +1,88 @@
+import ipdb
+import matplotlib.cm
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import colors, configure_matplotlib_style, save_pgf_trim
+
+configure_matplotlib_style()
+
+iteration_out_path = "data/stbf_struct/iterations.csv"
+
+iteration_accuracy = pd.read_csv(iteration_out_path)
+iteration_accuracy["subject"] = iteration_accuracy["subject"].astype(str)
+iteration_accuracy["fold"] = iteration_accuracy["fold"].astype(str)
+
+
+def plot_iteration_lines(
+    data, ax, n_trials="", show_chance_label=False, color="tab:blue"
+):
+    sns.lineplot(
+        data=data, x="iterations", y="accuracy", ax=ax, marker="s", markersize=3, mew=0
+    )
+    ax.set_ylim([0, 1])
+    ax.axhline(1 / 9, color=colors["lightgray"], linestyle="--", linewidth=1)
+    ax.set_xticks([0, 2, 4, 6, 8, 10])
+    ax.set_title(n_trials)
+    if show_chance_label:
+        ax.annotate(
+            "chance",
+            fontsize=9,
+            xy=(0.9, 1.25 / 9),
+            ha="right",
+            va="center",
+            xycoords="axes fraction",
+            color=colors["darkgray"],
+            bbox=dict(facecolor="white", linewidth=0, pad=0),
+        )
+
+
+fig, axs = plt.subplots(1, 3, sharex=True, sharey=True, constrained_layout=True)
+
+data_1_block = iteration_accuracy[iteration_accuracy["n_train_blocks"] == 1]
+plot_iteration_lines(
+    data_1_block[data_1_block["n_runs"] == 1],
+    axs[0],
+    n_trials="1 trial",
+)
+axs[0].set_ylabel("Accuracy")
+axs[0].set_xlabel("")
+axs[0].set_xticklabels([])
+plot_iteration_lines(
+    data_1_block[data_1_block["n_runs"] == 2], axs[1], n_trials="2 trials"
+)
+axs[1].set_xlabel("")
+axs[1].set_xticklabels([])
+plot_iteration_lines(
+    data_1_block[data_1_block["n_runs"] == 5], axs[2], n_trials="5 trials"
+)
+axs[2].set_xlabel("")
[…]

### 2024-11-15 chapters/bci.tex
Proofread mathematical notation

@@
 \label{sec:bci}
+\epigraph{``My mind controls my reality.''}{Frank Herbert -- Children of Dune}
 
@@
 There is one caveat with these linear methods, however:
-the number of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
-is usually relatively high compared to the training dataset, since calibration time
-should be kept minimal.
-Therefore, strong regularization should be applied, such as covariance/scatter
-shrinkage.
+the number of values ($=C\cdot S$) is usually relatively high compared to the
+size of the training dataset, since calibration time should be kept minimal.
+Therefore, strong regularization, such as covariance/scatter
+shrinkage, should be applied.
 Regularization can also be performed by retaining structure in the data, such as in the
@@
 stationarity in the background noise to regularize the problem.
-It does this by imposing a block-Toeplitz structure to the between-class
-scatter $\mat{S}_\text{b}$ matrix of \ac{lda}.
+It does this by imposing a block-Toeplitz structure to the within-class
+scatter $\mat{S}_\text{w}$ matrix of \ac{lda}.
 This means

### 2024-11-15 chapters/bttda.tex
Proofread mathematical notation

@@
 \input{chapters/bttda/tensorviz.tex}
-\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
+\epigraph{%
+  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
+  Lacking this sum, I therefore borrowed fifty coins from each of my two good friends.
+  With the hundred coins, I paid for the beast and received three coins in change.
+  Of these, I returned one coin each to my companions, keeping one for myself.
+  Now, I owe each friend forty-and-nine coins, and with mine one remaining, the
+  total merely equals ninety-and-nine coins, short of one hundred!
+  Tell me then, kind stranger, whither hath the missing coin gone?''
+}{%
+  Ancient Algerian riddle
+}
+\noindent\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
 
@@
 	\left\{\mat{A}^*\right\}
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \hat{\ten{X}}(n)\right)^2
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \hat{\ten{X}}(n)\right\]^2
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right\]^2
 \end{equation}
@@
 Since HODA estimates one within-class scatter matrix
-$S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
+$\mat{S}_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
 model of the data scatter is determined by these per-mode scatter matrices as a
-Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
+Kronecker product $\mat{S}_{-1,\text{w}}\otimes
+\mat{S}_{-2,\text{w}}\otimes\cdots\otimes \mat{S}_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
 drawn from a multilinear normal distribution~\cite{Ohlson2013}.

### 2024-11-15 chapters/conclusion.tex
Proofread mathematical notation

@@
 \chapter{Conclusion \& recommendations}
-\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
+\epigraph{``The mystery of life isn't a problem to solve, but a reality to
+experience.''}{S{\o}ren Kierkegaard}
+\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
 part of \textcite{VanDenKerchove2022}.}

### 2024-11-15 chapters/stbf_struct.tex
Proofread mathematical notation

@@
 
-  \emph{This chapter was
+  \noindent\emph{This chapter was
   published as~\textcite{VanDenKerchove2022}.}
@@
 	The epoched \ac{eeg} data with $N$ epochs, $C$ channels, and $S$ samples are
-  represented in epoch format as $\{\mat{X}_i\in\mathbb{R}^{C\times S}\}^N_{i=1}$
+  represented in epoch format as $\{\mat{X}_n\in\mathbb{R}^{C\times S}\}^N_{n=1}$
   or flattened vector format by concatenating all channels for each epoch.
-	Flattening results in $\{\mat{x}_i\in\mathbb{R}^{CS}\}^N_{i=1}$ such that
-  $\mat{x}_i = \text{vec}(\mat{X}_i)$.
+	Flattening results in $\{\mat{x}_n\in\mathbb{R}^{CS}\}^N_{n=1}$ such that
+  $\mat{x}_n = \text{vec}(\mat{X}_n)$.
 	The real covariance matrix of the epochs in vector format is
@@
 		\mathbf{a} =
-		\frac{1}{|\text{class 1}|}\sum_\text{class 1}\mathbf{x_i} -
-		\frac{1}{|\text{class 2}|}\sum_\text{class 2}\mathbf{x_i}
+		\frac{1}{|\text{class 1}|}\sum_\text{class 1}\mathbf{x_n} -
+		\frac{1}{|\text{class 2}|}\sum_\text{class 2}\mathbf{x_n}
 		\label{eq:stbf-struct/ap}
@@
 	\begin{equation}
-		y_i = \mathbf{w}\mathbf{x}_i
+		y_n = \mathbf{w}\mathbf{x}_n
 		\label{eq:stbf-struct/apply-bf}
@@
     \hat{\mat{C}}_\text{emp} =
-		\frac{1}{N-1}\sum^{N}_{i=1}\mat{x}_i\mat{x}_i^\intercal
+		\frac{1}{N-1}\sum^{N}_{n=1}\mat{x}_n\mat{x}_in\intercal
 		\label{eq:stbf-struct/emp-cov}
@@
 	\end{equation}
-	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when
+	The Moore-Penrose pseudoinverse, $^+$, ensures a solution exists when
   $\hat{\mat{C}}_\text{emp}$ is singular.
@@
 	The optimal value for $\alpha$ can be found with a line search using cross-validation, but this can be a costly procedure.
+
 	Methods exist to estimate an optimal shrinkage value from the data directly.
@@
 	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010}, and Oracle Approximating Shrinkage~\cite{Chen2010}.
-	A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
-	cross-validation (\ac{loocv}) scheme expressed by the data-driven closed-form
+  A more recent estimation method~\cite{Tong2018} emulates a \ac{loocv} scheme expressed by the data-driven closed-form
 	estimate:
@@
       + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
-			- \frac{1}{N(N-1)}\sum_{i=1}^N||\mat{x}_i||_2^4
+			- \frac{1}{N(N-1)}\sum_{n=1}^N||\mat{x}_n||_2^4
 		}
@@
       + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
-			+ \frac{1}{N(N-1)^2}\sum_{i=1}^N||\mat{x}_i||_2^4
+			+ \frac{1}{N(N-1)^2}\sum_{n=1}^N||\mat{x}_n||_2^4
 		}
@@
 	\end{equation}
-
 	We opt for the \ac{loocv} shrinkage estimator because it avoids some of the
@@
 	\begin{property}
-    $(\mat{U}\otimes \mat{V})\cdot\text{vec}(\mat{W}) =
-    \text{vec}(\mat{V}\mat{W}\mat{U}^\intercal)$
+    $(\mat{U}\otimes \mat{V})\cdot\vect(\mat{W}) =
+    \vect(\mat{V}\mat{W}\mat{U}^\intercal)$
     for any matrices $\mat{U}\in\mathbb{R}^{P\times P}$,
@@
     \frac{\hat{\mat{S}}^+\mat{A}^\intercal\hat{\mat{T}}^+}
-    {\mat{a}\cdot\text{vec}(\hat{\mat{S}}^+A^\intercal\hat{\mat{T}}^+)}
+    {\mat{a}\cdot\vect\left(\hat{\mat{S}}^+A^\intercal\hat{\mat{T}}^+\right)}
 		\label{eq:stbf-struct/closed-form-kron}
@@
 			\frac{1}{N}
-      \sum^N_{i=1}\mat{X}_i^\intercal\hat{\mat{T}}_k^+\mat{X}_i
+      \sum^N_{n=1}\mat{X}_n^\intercal\hat{\mat{T}}_k^+\mat{X}_n
 			\label{eq:stbf-struct/fpi-spatial}
@@
 			\frac{1}{N}
-      \sum^N_{i=1}\mat{X}_i\hat{\mat{S}}_k^+\mat{X}_i^\intercal
+      \sum^N_{n=1}\mat{X}_n\hat{\mat{S}}_k^+\mat{X}_n^\intercal
 			\label{eq:stbf-struct/fpi-temporal}
@@
 			\frac{
-        \frac{N}{N-1}\text{Tr}(\tilde{\hat{S}}_{k+1}^2)
-        - \frac{2}{C}\left[\text{Tr}(\tilde{\mat{S}}_{k+1})\right]^2
+        \frac{N}{N-1}\text{Tr}(\tilde{\mat{\hat{\S}}}_{k+1}^2)
+        - \frac{2}{C}\left[\text{Tr}(\mat{\tilde{S}}_{k+1})\right]^2
         + \frac{1}{C}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
@@
         - \frac{2}{S}\left[\text{Tr}(\tilde{\mat{T}}_{k+1})\right]^2
-				+ \frac{1}{S}\text{Tr}(\tilde{T}_{k+1}^2)
-				- \frac{1}{N(N-1)}\sum_{i=1}^N
-        \left[\text{Tr}(\mat{X}_i^\intercal\mat{\mat{S}}_k^+\mat{X}_i)^2\right]
+        + \frac{1}{S}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+				- \frac{1}{N(N-1)}\sum_{n=1}^N
+        \left[\text{Tr}(\mat{X}_n^\intercal\mat{\mat{S}}_k^+\mat{X}_n)^2\right]
 			}{
@@
         + \frac{1}{S}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
-				+ \frac{1}{N(N-1)^2}\sum_{i=1}^N
-        \left[\text{Tr}(\mat{X}_i^\intercal\hat{\mat{S}}_k^+\mat{X}_i)^2\right]
+				+ \frac{1}{N(N-1)^2}\sum_{n=1}^N
+        \left[\text{Tr}(\mat{X}_n^\intercal\hat{\mat{S}}_k^+\mat{X}_n)^2\right]
 			}
@@
 			relative to the number of iterations used to estimate the Kronecker-Toeplitz structured shrunk
-			covariance. Error bars represent the first and third quartiles. The
-      accuracy does not improve when using more than one iteration.}
+			covariance. Shaded area represents a 95\% confidenced interval obtained
+      through 1000 bootstrapping iterations.
+      Accuracy does not improve when using more than one iteration.}
 		\label{fig:iterations}

### 2024-11-17 rebuttal.tex
Indicated comments Reinhold and Fabien

\documentclass{letter}
\usepackage[backend=biber]{biblatex}
\usepackage{todonotes}
\setuptodonotes{inline}

\addbibresource{references.bib}

\newcommand{\reply}[1]{%
	\textbf{Response to review comments made by prof. #1}

}

\begin{document}

\signature{
	Arne Van Den Kerchove \\
	\texttt{arne.vandenkerchove@kuleuven.be} \\
	+32 473 32 78 71 \\
}
\address{
	KU Leuven \\
	Laboratory for Neuro- and Psychophysiology \\
	Campus Gasthuisberg, O\&N 2 \\
	Herestraat 49 bus 1021 \\
	BE-3000 Leuven
}

\begin{letter}{
	prof. John Creemers \\
	KU Leuven \\
	Doctoral School of Biomedical Sciences \\
	Campus Gasthuisberg, O\&N 4 \\
	Herestraat 49 bus 700
	BE-3000 Leuven

}
\opening{Dear chair and members of the reading committee}

I would like to extend my sincere gratitude for time and effort you have taken
to review the manuscript of my PhD dissertation.
Additionally, I would like to thank you for the insightful, constructive and
detailed comments you all have provided.
The following reply contains a point-by-point response to your review comments.
Attached, you will also find an updated draft of the manuscript and a version
with indicated changes for your approval.

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
	\item
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item Part of the confusion between active and passive BCI might be due
	to the poor choice of phrasing ``paradigms that require little active participation from the user, like redirect-
	ing attention or initiating imagined actions, are classified as passive BCI.''.
	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
	instance through redirecting attention or initiating imagined actions,
	paradigms free of this requirement are classified as \emph{passive \ac{bci}}''
	for clarity.
	Neurofeedback has been reclassified as active BCI in the text and in
	figure 1.3.
	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
	comparison of sample rate}
	\item \todo{report actual accuracies performance stbf in table}
	\item In the results section of chapter 4, statistical comparisons were
	only performed between methods for which we evaluated the performance
	ourselves (HODA, PARAFACDA and BTTDA).
	Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	fair comparison of performance metrics reported across different works,
	the analyses for other comparison methods were not re-run
	by us as reported, but performances were
	taken from the MOABB benchmark database~\cite{Chevallier2024}.
	However, this source provided only the aggregated performance metrics
	and not the metrics per subject, session and cross-validation fold,
	hence we could not perform relevant statistical comparison with these
	comparison models.
	\item \todo{include table with absolute performances for transfer
	conditions}
\end{enumerate}
Minor comments:
\todo{minor comments in document}
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were adapted to accommodate this.
\end{enumerate}

\reply{Reinhold Scherer}
Major comments:
\begin{enumerate}
	\item \todo{quantify qualitative statements}
	\item  Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	\item As indicated by the reviewer, true on-line experiments with live
	feedback have indeed not yet be conducted, but were foreseen to take
	place within the framework of this doctoral study.
	I agree that these experiments would be valuable to supplement the
	presented conclusions and are the logical next step.
	Indeed, the start of experimentation with healthy control participants
	as well as patient groups, was heavily delayed by the COVID-19 pandemic, as the
	preparatory research started in September 2019 and I officially
	enrolled in the the doctoral training programme in January 2020.
	This restricted not only our access to experiment subjects for a
	significant period at the start of the PhD, and even longer for
	vulnerable patient groups, but also shifted priorities for the medical
	staff at our partner institutions and patient centers.
	Hence, work early in the PhD mainly focused on the development of
	computational methods.
	We are currently starting up true on-line experimentation in cooperation
	with our partner patient centers, with an adapted interface and experimental protocol.
	\todo{indicate in manuscript covid}
	\item Recorded datasets could not be made publicly available since the
	ethical protocol (Ethics Commission of University Hospital Leuven --
	S62547 v1.4) for the conducted studies did not include provisions
	publishing the data.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Maarten De Vos}
Major comments:
\begin{enumerate}
	\item Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	The paragraph on Block-Toeplitz linear discriminant analysis (tLDA) in
	section 1.6.4 has also been corrected, and it was expanded for clarity.
	Figures 3.1 and 3.2 have been regenerated and relabeled for consistent
	formatting.
	\item \todo{replace wilcoxon by delong}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Unsupported claims regarding generalisability have been omitted
	from the abstract.
	\item Throughout the manuscript, different EEG band-pass filtering
	settings have been used.
	Chapter 1 indicates that ERPs are usually band-pass filtered between
	0.5 and 16 Hz.
	These were also the preferred cut-off frequencies for other
	(unpublished) analyses used during the course of the thesis work, an
	those used for the analyses in chapter 3.
	However, ERP datasets evaluated in chapter 4 were band-pass filtered
	between 1 and 24 Hz, and in chapter 6 between 0.1 and 20 Hz.
	While there is no general consensus in the field about the optimal
	cut-off frequencies for P300 detection and ERP analysis in general,
	these values all fall well within the accepted
	bounds~\cite{Bougrain2012}.
	Decisions to deviate from this were made consciously.
	In chapter 3, filtering between 1 and 24 Hz was applied for consistency
	with the recommended ERP preprocessing pipeline of the MOABB
	benchmarking framework~\cite{Aristimunha2023}.
	This allows for a fair comparison under equal circumstances of the
	performance of the proposed algorithm with performances of other
	algorithms reported by the MOABB benchmarking
	database~\cite{Chevallier2024}.
	In chapter 6, the low-pass frequency was increased from 16 Hz to 20 Hz,
	since the analyses in this explicitly involve the short-time effects of
	latencies of ERP components, and the analysis of faster, early ERP
	components. A low low-pass frequency would harm the precision of
	latency effects in the filtered signal.
	A lower high-pass filter of 0.1 Hz was used here to reveal the
	influence of the proposed alignment approach on potential
	slower late components.
	\item Typographical errors have been corrected throughout the
	manuscript.
\end{enumerate}

\reply{Adalberto Simeone}
Major comments:
\begin{enumerate}
	\item \todo{discuss in limitations the impact of letters and future
	work experiments with an actual usable interface}
	\todo{mention here plans to go to patients with actual speller
	interface}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item \todo{correct si measurements Hz, mm, cm, V, µV, $\mu$ V}
	\item \todo{do proper check in textcite, especially in chapter 3}
	\item \todo{check for contractions and remove we've, isn't, wouldn't}
\end{enumerate}

\textbf{References}

\printbibliography[heading=none]

\closing{Kind regards and thank you in advance,}
\end{letter}
\end{document}

### 2024-11-17 chapters/bci.tex
Indicated comments Reinhold and Fabien

@@
 into sophisticated interaction schemes that restore or enhance many capabilities by
-bypassing defective parts of the nervous or muscle system.
+bypassing defective parts of the nervous or muscle system.\todo{the BCI concetp
+was proposed at that time, but maybe not yet this form of communication. Also,
+this reference is not from the 1970s, you may want to cite the original paper}
 By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
-be used as assistive communication technology by individuals with \acl{sspi}.
-speech~\cite{Wolpaw2018,Milekovic2018}.
+be used as assistive communication technology by individuals with \acl{sspi}
+~\cite{Wolpaw2018,Milekovic2018}.
 Impressive cutting-edge experimental examples include a brain-spine interface allowing a
@@
 user.
-The user's brain will then adapt through reinforcement learning, causing changes in
+The user's brain will then adapt through learning, causing changes in
 behavior or strategy.
@@
 The \emph{action potential} and its related \emph{postsynaptic potential} are electrical
-pulses occurring when a neuron receives input from another neuron and is activated.
+pulses occurring when a neuron receives input from other neurons and is activated.
 The firing of a neuron, or the combined firing of groups of synchronized neurons, thus
@@
 spatial resolution.
-These are needles with a diameter smaller than 10µm and one or multiple electrodes that
+These are needles with a diameter smaller than 10 µm and one or multiple electrodes that
 extend several millimeters into the cortex.
 Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
-of a small group of neurons, or even intracellular single neuron action potential
+of a small group of neurons,\todo{qualitative}
+or even intracellular single neuron action potential
 spikes.
@@
  and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
-symptoms~\cite{Arlotti2018}.
+symptoms~\cite{Arlotti2018}.\todo{qualify size}
 Depth electrodes are longer rods with multiple electrodes that can extend deeper into
@@
 epilepsy diagnosis.
-Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
-number of smaller electrodes per surface area, allow for more precise measurements.
+Newer developments like micro-ECoG (µECoG), with a higher
+number of smaller electrodes per surface area (recording from sites with
+diameter of 10s of micrometers instead of $\pm$ 1 cm)~\cite{Shokoueinejad2019},
+allow for more precise measurements.
 
@@
 
-Paradigms that require little active participation from the user, like redirecting
-attention or initiating imagined actions, are classified as \emph{passive \ac{bci}}.
+In contrast to paradigms that require the user's active participation, for
+instance through redirecting attention or initiating imagined actions,
+paradigms free of this requirement are classified as \emph{passive \ac{bci}}.
 From the user's point of view, this concept is probably the most attractive, since their
@@
 or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
-It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
+Clinical applications are found in the diagnosis of a multitude of neurological
+and psychological disorders through various forms of neuroimaging, in resting
+state or while performing a specific task.
 Nevertheless, active participation facilitates training data collection for supervised
@@
 
-Paradigms requiring high active participation can be split up into \emph{active
+Paradigms that do require high active participation can be split up into \emph{active
 \ac{bci}} and \emph{reactive \ac{bci}}.
@@
 symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
-Active paradigms are also subject to large inter-subject variability due to the
+Active BCI also has clinical applications in the form of
+neurofeedback~\cite{Hammond2011}, where the user adapts mental strategies to
+perform active self-regulation of their brain activity which is presented to
+them, causing a feedback loop.
+
+Active paradigms are subject to large inter-subject variability due to the
 complexity of the performed tasks.
@@
 instruction.
-Imagined speech or movements can be performed in a multitude of ways, which can each
-have different neural representations for separate individuals.
+Imagined speech or movements, or mental strategies in general, can be
+performed in a multitude of ways, which can themselves have different neural
+representations for separate individuals.
 This can make it hard to adapt the \ac{bci} to specific individuals, causing poor
@@
 which can cause eye strain and fatigue due to the continuous oscillatory stimulation of
-all targets at once~\cite{Xu2021}.
+all targets at once~\cite{Xu2021}\footnote{
+  Recent work mitigates this by making SSVEP stimulation
+  more comfortable without compormising selection accuracy through
+  high-frequency low-contrast stimulation~\cite{Ladouce2022}.
+}.
 Furthermore, \ac{ssvep} relies more on directing the gaze at the intended target than
@@
 While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
-studies paradoxically often report poor performance in exactly this group,
-precisely due to their lack of adequate eye motor control.
+studies paradoxically often report poor performance in exactly this group.
+This is caused by a multitude of factors, including diminished
+electrophysiological responses and cognitive defects linked to the
+pathophysiology, psychological factors such as the ``extinction of goal directed thinking''
+phenomenon in complete impairment, and, importantly, sensory impairments such as vision and
+oculomotor defects~\cite{Seguin2023}.
+While visual \acp{bci} often rely on visual sensory functions, it is precisely this
+group of patients that could benefit from them, that lacks adequate eye motor control.
 Equally often, participants with gaze impairments were excluded from the study.

### 2024-11-17 chapters/bttda.tex
Indicated comments Reinhold and Fabien

@@
 multilinear spectral filter terms and apply it to motor imagery BCI, but their
-decomposition is also restricted in flexibility.
+decomposition is also restricted in flexibility, since the solution can only
+extract rank $(r_1,r_2,1)$ with mode 3 corresponding to the frequency domain.
 
@@
 	\input{figures/bttda/HODA_forward.tikz.tex}
-  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
+  \caption[A forward projection for \acs{hoda}.]{The forward projection for HODA.
     By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
@@
 	\left\{\mat{A}^*\right\}
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \hat{\ten{X}}(n)\right\]^2
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right\]^2
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2
+  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
 \end{equation}
@@
 smaller, but no significant differences were found between HODA+LDA,
-PARAFACDA+LDA, and BTTDA+LDA.
+PARAFACDA+LDA, and BTTDA+LDA\todo{qualitative}.
 
@@
 We report within-session ROC-AUC scores for training, validation, and test data as a function
-of the number of blocks increases, shown in \cref{fig:blocks}.
+of the number of blocks, shown in \cref{fig:blocks}.
 Training and validation folds were taken from the model selection procedure.
@@
 Nevertheless, when considering the relative improvement over the original
-HODA model, our results show that BTTDA+LDA performs consistently on par or higher than HODA+LDA.
-The classification score for BTTDA+LDA was also slightly higher than that of
+HODA model, our results show that BTTDA+LDA performs consistently on par or
+higher than HODA+LDA\todo{qualitative}.
+The classification score for BTTDA+LDA was also slightly
+higher\todo{qualitative} than that of
 restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
@@
 problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
-imagery problems (third-order tensors).
+imagery problems (third-order tensors)\todo{qualitative}.
 Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible

### 2024-11-17 chapters/conclusion.tex
Indicated comments Reinhold and Fabien

@@
 an existing dataset.
-Our results show that the structured estimator results in higher accuracy in
+Our results show that the structured estimator results in higher
+accuracy\todo{qualitative} in
 general, and specifically when training data are sparsely available.
@@
 effectiveness under conditions of controlled latency variability and noise.
-These results showed that it is robust to higher noise and jitter compared to
-a non-iterative method in a latency estimation task.
-It was also shown that decoding accuracy was higher in high jitter and noise
-settings compared to the non-iterative method and a state-of-the-art decoder.
+These results showed that it is robust to higher noise and jitter\ compared to
+a non-iterative method in a latency estimation task\todo{qualitative}.
+Decoding accuracy was higher in high jitter and noise
+settings compared to the non-iterative method and a state-of-the-art
+decoder\todo{qualitative}.
 
@@
 While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
-decoding, it increased covert \ac{vsa} decoding.
+decoding, it increased covert \ac{vsa} decoding.\todo{qualitative}
 
@@
 results in a significant decrease in performance compared to the state-of-the-art,
-consistent with the within-condition results.
+consistent with the within-condition results.\todo{qualitative}
 For all other pairs of training and evaluation VSA conditions, however,
-\ac{wcble} was equal to or significantly better.
+\ac{wcble} was equal to or significantly better.\todo{qualitative}
 
@@
 Finally, we wish to expand our experimentation with individuals with
-\acp{sspgi}.
+\ac{sspgi}.
 We aim to revisit the participants with an on-line experiment and comfort
@@
 
-Throughout this thesis, we've noticed that linear or multilinear decoding
+Throughout this thesis, we noticed that linear or multilinear decoding
 methods consistently outperform non-linear methods, such as SVMs or those based
@@
 experimentation systems.
-It is therefore important that research labs maintain a working, long-term
+It is therefore important that research labs maintain a long-term working
 collaboration with patient centers, so that new projects can immediately be
@@
 
-Similarly, there is value in maintaining a working, on-line in-house \ac{bci}
+Similarly, there is value in maintaining a working, on-line, in-house \ac{bci}
 system that can be iterated upon and adapted if necessary.
-Experiments can be supported by advance calculations of hypothetical
+Given this system's design and data gathered from earlier use, experiments can
+be supported by advance calculations of hypothetical
 \ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
@@
 reporting of realistic, interpretable metrics,
-and by taking into account the human aspect of working \emph{with} \ac{bci} users.
+and by taking into account the human aspect of working \emph{with} instead of
+\emph{for} \ac{bci} users.

### 2024-11-17 chapters/covert_align.tex
Proofread mathematical notation
Indicated comments Reinhold and Fabien

@@
 \label{sec:covert-align}
-\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
+\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ A
+Scandal in Bohemia -- Sir Arthur Conan Doyle}
+\noindent\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
 
@@
 latencies relative to stimulus onset during covert \ac{vsa} compared to overt
-\ac{vsa}.
+\ac{vsa}\todo{qualitative, discuss}.
 This latency variability contributes to reduced covert \ac{vsa} decoding performance.
@@
   \end{subfigure}
+  \caption[Stimulation interface layout.]{Layout and visual elements of the experimental stimulation
+  interface used in recording the CVSA-ERP dataset.}
 \end{figure}
@@
 datasets, which is greatest for smaller numbers of repetitions and
-decreases as the number of repetitions increases.
-This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
-for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
+decreases as the number of repetitions increases.\todo{qualitative}
+This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
+BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
 Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
 \ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
-selection accuracy.
+selection accuracy.\todo{qualitative}
 No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
@@
 in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
-\ac{wcble} in CVSA-ERP.
+\ac{wcble} in CVSA-ERP.\todo{qualitative}
 
@@
 \ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
-These results compare to the performance gain in \cite{Zisk2022}.
+These results compare to the performance gain reported by \textcite{Zisk2022}.
 They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
@@
 When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
-evaluated settings, similar to the within-subject conditions.
+evaluated settings, similar to the within-subject conditions.\todo{qualitative}
 On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
 see that performance is on par or greater using \ac{wcble} for most conditions,
-except for within-overt \ac{vsa} decoding.
+except for within-overt \ac{vsa} decoding.\todo{qualitative}
 
@@
         dissociating their visuospatial attention and gaze fixation.
-        Evidence shows that \ac{erp} latency variability is higher when attention is
+        Evidence shows that \ac{erp} latency variability is
+        higher when attention is
         divided~\cite{Polich2007,Arico2014}.
@@
 \end{enumerate*}
+\todo{first item qualitative}
 
@@
 range (IQR) of the estimated latencies of target trials, revealed that overt
-\ac{vsa} exhibited significantly lower jitter than all other conditions for both datasets.
+\ac{vsa} exhibited significantly lower jitter than all other conditions for
+both datasets.\todo{qualitative}
 No other significant differences were found.
@@
 only factor hampering covert \ac{vsa} decoding performance.
-However, the large increase in performance for covert \ac{vsa} in \cite{Arico2014} could also
+However, the large increase in performance for covert \ac{vsa} reported by
+\cite{Arico2014}\todo{qualitative} could also
 be explained by overfitting on artifacts amplified by aligning, since their
-method is not evaluated on unseen data, as opposed to ours.
+method is not evaluated on unseen data, as opposed to ours.\todo{qualitative}
 

### 2024-11-17 chapters/gaze_independence.tex
Indicated comments Reinhold and Fabien

@@
 
-Some other groups that are heavily affected are \ac{cp}~\cite{Fazzi2012} and
-\ac{fa}~\cite{Fahey2008,Hocking2010,Furman1983}.
-Individuals with some neurodegenerative diseases like
+Another other group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
+Additionally, individuals with other neurodegenerative diseases like
 \ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
@@
 somatosensory~\cite{Brouwer2010}
-stimulation do not rely on gaze redirection but also result in lower information transfer
-rates compared to visual paradigms.
+stimulation do not rely on gaze redirection but also result in lower \acp{itr}
+compared to visual paradigms\todo{qualitative}.
 While auditory and somatosensory stimulation on their own might yield poor
@@
 Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
-operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
+operation and from user-dependent
+variability~\cite{Severens2014,Reichert2020b}.\todo{qualitative}
 \textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
@@
 multiple targets are displayed at the same time at different spatial locations,
-and temporally organized interfaces, like \ac{rsvp}, where targets or small sets of targets are shown consecutively.
+and temporally organized interfaces, like \ac{rsvp}, where targets or small
+sets of targets (e.g. 3) are shown consecutively.
 The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
@@
 concurrently decoding the covert \ac{vsa} shift from spectral content, but
-their study only used a few stimulation targets.
+their study only used a few stimulation targets.\todo{qualitative}
 In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
@@
 settings~\cite{Treder2010}, the \ac{snr} of
-the \ac{erp} is lower than in overt attention settings.
+the \ac{erp} is lower than in overt attention settings\todo{qualitative}.
 Therefore, a straightforward way to reach satisfactory
@@
       \emph{Free} \ac{vsa}.
-      Gaze and \ac{vsa} coincide on a target.
       The user is free to direct their gaze as they deem most comfortable.

### 2024-11-17 chapters/patients.tex
Indicated comments Reinhold and Fabien

@@
 \label{sec:patients}
+\epigraph{I knew there would be a lot of unexpected issues involved with
+patient experiments, but I never thought there would be interference from baby
+goats in the experiment room.}{Juliette}
+
 \section{Introduction}
@@
 Participants with \ac{als} also had a substantially lower accuracy than healthy
-controls (58\% vs 88\%).
+controls (58\% vs. 88\%).
 
@@
 Epochs were cut from -0.1 to 0.9s relative to stimulus onset, and no baseline
-correction was performed.
+correction was performed in order to meet the assumptions.
 
@@
 If both were missing, the gaze position remained unset at that time point, and no
-interpolation was performed.
+interpolation was performed of the employed classifiers.
 
@@
 More interestingly, we noticed that performances of the decoders in free
-\ac{vsa} were close to those in the overt \ac{vsa}.
+\ac{vsa} were close to those in the overt \ac{vsa}.\todo{qualitative}
 A substantial decrease in performance from the overt setting to the free
 setting was observed for
-subjects PC3 and PC4.
+subjects PC3 and PC4.\todo{qualitative}
 For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
-setting, there was no decrease in performance.
+setting, there was no notable decrease in performance.\todo{qualitative}
 
@@
 on overt \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
-gaze was uncued.
+gaze was uncued.\todo{qualitative}
 This is also supported by our results on cross-condition calibration presented

### 2024-11-17 chapters/stbf_struct.tex
Indicated comments Reinhold and Fabien

@@
 	\label{sec:stbf-struct/methods/shrunk-cov}
-	The shrinkage covariance estimator creates a better-conditioned inversion matrix problem and generally performs better when applied to unseen data.
+	The shrinkage covariance estimator creates a better conditioned inversion matrix problem and generally performs better when applied to unseen data.
 	The estimators for the covariance and inverse covariance are given by:
@@
 	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
-	Analysis (ICA) by rejecting components that correlated with the bipolar EOG channels vEOG and hEOG, according to adaptive Z-score thresholding.
+	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
+  channels vEOG and hEOG, according to adaptive Z-score
+  thresholding.\todo{details? threshold for rejection?}
 	Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
@@
 	contain more information hence the minimum temporal resolution is chosen to reduce
-	the dimensionality of the covariance and improve its covariance.
+	the dimensionality of the covariance and improve its condition number.
 	The activation pattern is the difference between the averages of epochs in response to cued targets and the averages of those in response to non-cued targets.
@@
 	accuracy across the amounts of training data and the number of trials
-	used for evaluation.
+  used for evaluation\todo{qualitative}.
 	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
-	the training process.
+  the training process\todo{qualitative}.
 
@@
 	\end{figure}
+  \todo{the legend does not correspond to the text (method names are
+  different)}
 
@@
 	\textsc{stbf-shrunk} when the number of training blocks is low.
-	This effect is present for 1-, 2- and 5-trial evaluation.
+  This effect is present for 1-, 2- and 5-trial evaluation\todo{qualitative}.
 	This advantage decreases (the $p$-value increases) when
-	adding more training blocks.
+  adding more training blocks\todo{qualitative}.
 	Both \textsc{stbf-struct} and \textsc{stbf-shrunk} perform significantly better
-	than \textsc{stbf-emp} for all evaluated settings.
+  than \textsc{stbf-emp} for all evaluated settings\todo{qualitative}.
 	Compared to \textsc{xdawn+rg}, \textsc{stbf-struct} also has significantly
-	higher accuracy in almost all evaluated settings, except when using only one training block.
-	\textsc{stbf-shrunk} does not outperform \textsc{xdawn+rg} when training data is low but gains a significant advantage
-	when using more training data.
+	higher accuracy in almost all evaluated settings, except when using only one
+  training block\todo{qualitative}.
+	\textsc{stbf-shrunk} does not outperform \textsc{xdawn+rg} when training data
+  is low but gains a significant advantage	when using more training
+  data\todo{qualitative}.
 
@@
 	and \textsc{stbf-emp}. The training time of all three \ac{stbf}-based classifiers is
-	much lower than that of \textsc{xdawn+rg}, which appears nearly constant when using
+  much lower\todo{qualitative} than that of \textsc{xdawn+rg}, which appears nearly constant when using
 	4, 8, 16, or 32 channels.
@@
 	compared to the original \textsc{stbf-emp} for all numbers of training blocks
-	indicated.
+  indicated\todo{qualitative}.
 	We believe there are three reasons for this.
@@
 	Another observation is the significantly better accuracy score of
-	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available training data is small.
+	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available
+  training data is small\todo{qualitative}.
 	This property is an attractive advantage in a \ac{bci} setting since it is desirable to keep the calibration (training) phase as short as possible without losing accuracy.
@@
 	covariance.
-	The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:stbf-struct/discussion/param-complex}), increasing the stability of matrix inversions.
+  The increase in accuracy for small training set sizes\todo{qualitative} can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:stbf-struct/discussion/param-complex}), increasing the stability of matrix inversions.
 
@@
 	that \textsc{stbf-struct} reaches similar accuracy when using only one block of
-	training data.
+  training data\todo{qualitative}.
 	The authors suspect this is due to both
@@
 	satisfactory classification accuracy.
-	When more data are available, \textsc{stbf-struct} reaches significantly better accuracies.
+	When more data are available, \textsc{stbf-struct} reaches significantly
+  better accuracies\todo{qualitative}.
 	Combined with the benefits laid out in \cref{sec:stbf-struct/discussion/param-complex}, this
@@
 	\textsc{stbf-shrunk} does not show decisive accuracy improvements over
-	\textsc{xdawn+rg} using a few training blocks, but this improves as the training data increases.
+	\textsc{xdawn+rg} using a few training blocks, but this improves as the
+  training data increases\todo{qualitative}.
 
@@
 	calculations.
-	It can be observed that the training time of \textsc{stbf-struct} stays low compared to \textsc{stbf-emp} and \textsc{stbf-shrunk} when dimensionality increases.
+	It can be observed that the training time of \textsc{stbf-struct} stays low
+  compared to \textsc{stbf-emp} and \textsc{stbf-shrunk} when dimensionality
+  increases\todo{qualitative}.
 
@@
 	\ac{stbf}-based estimators are very fast to train compared to the
-	state-of-the-art estimator \textsc{xdawn+rg}, which confirms the results in~\cite{Wittevrongel2016}.
+	state-of-the-art estimator \textsc{xdawn+rg}, which confirms the results
+  reported by~\textcite{Wittevrongel2016}.
 	Since the training times of all \ac{stbf}-based classifiers are already in
@@
 	However, the authors believe that these results could significantly impact some
-	use cases of the spatiotemporal beamformer, like high spatial or temporal resolution \ac{erp} analyses.
+	use cases of the spatiotemporal beamformer, like high spatial or temporal
+  resolution \ac{erp} analyses (e.g., $C>>32, S>>256$).
 	One example is single-trial \ac{erp} analysis with a high-temporal
@@
 	beamformer and a state-of-the-art method in an off-line P3 detection task.
-	Our results show that the structured estimator performs better when training data are sparsely available and that it can be computed faster and with substantially less memory usage.
+	Our results show that the structured estimator performs better when training
+  data are sparsely available and that it can be computed faster and with
+  substantially less memory usage\todo{qualitative}.
 	Since these algorithms are not paradigm-specific,  the conclusions can be generalized to

### 2024-11-17 chapters/wcble.tex
Proofread mathematical notation
Indicated comments Reinhold and Fabien

@@
 \label{sec:wcble}
-\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
+\epigraph{``The sands of time cannot be stopped''}{Brom in Eragon --
+Christopher Paolini}
+\noindent\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
 \cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
@@
 lab to the clinical setting.
-For instance, \textcite{Zisk2021} showed that \ac{als} results significantly
-increased \ac{erp} jitter.
+For instance, \textcite{Zisk2021} showed that \ac{als} results in significantly
+increased \ac{erp} jitter\todo{qualitative}.
 \textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
@@
   \begin{cases}
-    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
+    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l+\frac{1}{2f}$} \\
     0 & \text{otherwise}
@@
 \begin{equation}
-  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
+  \text{SNR} = 10\log_{10}\frac{\var\left[s(t,0)\right]}{\var\left(\text{noise}\right)}
 \end{equation}
@@
 $\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
-This difference is particularly large when $\sigma$ is higher ($\sigma=0.2$
+This difference is particularly large\todo{qualitative} when $\sigma$ is higher ($\sigma=0.2$
 and $\sigma=0.3$) and \ac{snr} is not yet high.
@@
 $\sigma=0.1$ and $\text{\ac{snr}} < 27 \text{dB}$, at which point both \ac{cble}
-and \ac{wcble} performed at chance level.
+and \ac{wcble} performed at chance leveltodo{qualitative}.
 When considering the accuracy of \ac{wcble}, we see that it shows more of a
-plateau as \ac{snr} increases.
+plateau as \ac{snr} increases\todo{qualitative}.
 
@@
 in \cref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
-potential jitter and lower \ac{snr} than \ac{cble}.
+potential jitter and lower \ac{snr} than \ac{cble}\todo{qualitative}.
 It could therefore be a better candidate to apply in \ac{erp} analysis and
 decoding in those cases where jittered \ac{erp} data is of interest.
-At a certain noise threshold, both methods begin to fail, but \ac{wcble}
-delivers higher accuracy at intermediate \ac{snr} levels than \ac{cble}, likely due to the iterative alignment process.
+At a certain noise threshold\todo{qualitative}, both methods begin to fail, but \ac{wcble}
+delivers higher accuracy at intermediate \ac{snr} levels than
+\ac{cble}\todo{qualitative}, likely due to the iterative alignment process.
 

### 2024-11-17 figures/gaze_independence/lis_venn.tikz.tex
Update bttda formulation
Indicated comments Reinhold and Fabien

@@
     % Draw the outer lightgray box with ultra thick darkgray line
-    \draw[fill=lightgray, draw=darkgray, ultra thick] (-4,-1.5) rectangle (4,5.5);
+    \draw[draw=darkgray, ultra thick] (-4,-1.5) rectangle (4,5.5);
     % Label for the outer box in the top left corner
@@
     \draw[ultra thick, accent1] (0,2) circle (3);
-    \node[text=accent1] at (0,4) {\textbf{\acs{sspi}}};  % Centered bold text with line color
+    \node[text=accent1, text width=15em, anchor=center, align=center] at (0,4)
+    {\small\textbf{severe speech and \\ physical impairment (SSPI)}};  % Centered bold text with line color
 
@@
     \draw[ultra thick, accent2] (0,1) circle (2);
-    \node[text=accent2] at (0,1.66) {\textbf{\acs{sspgi}}};  % Centered bold text with line color
+    \node[text=accent2, text width=10em, anchor=center, align=center] at
+    (0,2) {\small\textbf{severe speech, \\ physical and gaze \\ impairment
+    (SSPGI)}};  % Centered bold text with line color
 
@@
     \draw[ultra thick, accent3] (0,0) circle (1);
-    \node[text=accent3] at (0,0) {\textbf{\acs{lis}}};  % Centered bold text with line color
+    \node[text=accent3, text width=5em, anchor=center, align=center] at (0,-.1)
+    {\small\textbf{Locked-in \\ Syndrome \\ (LiS)}};  % Centered bold text with line color
 

### 2024-11-17 scripts/stbf_struct/kronecker_iteration.py

@@ -14,9 +14,7 @@ iteration_accuracy["subject"] = iteration_accuracy["subject"].astype(str)
 iteration_accuracy["fold"] = iteration_accuracy["fold"].astype(str)
 
 
-def plot_iteration_lines(
-    data, ax, n_trials="", show_chance_label=False, color="tab:blue"
-):
+def plot_iteration_lines(data, ax, n_trials="", show_chance_label=False):
     sns.lineplot(
         data=data, x="iterations", y="accuracy", ax=ax, marker="s", markersize=3, mew=0
     )
@@ -30,7 +28,7 @@ def plot_iteration_lines(
             fontsize=9,
             xy=(0.9, 1.25 / 9),
             ha="right",
-            va="center",
+            va="bottom",
             xycoords="axes fraction",
             color=colors["darkgray"],
             bbox=dict(facecolor="white", linewidth=0, pad=0),

### 2024-11-17 rebuttal.tex
Rework fig 1.3 and 1.4

@@
 \opening{Dear chair and members of the reading committee}
+\todo{fix todos in document}
 
@@
 	instance through redirecting attention or initiating imagined actions,
-	paradigms free of this requirement are classified as \emph{passive \ac{bci}}''
+	paradigms free of this requirement are classified as passive bci,''
 	for clarity.
@@
 Minor comments:
-\todo{minor comments in document}
 \begin{enumerate}

### 2024-11-17 scripts/setup.py

@@ -1,5 +1,5 @@
 import matplotlib.pyplot as plt
-#import tikzplotlib
+# import tikzplotlib
 from cycler import cycler
 
 mm = 1 / 25.4  # centimeters in inches
@@ -11,14 +11,16 @@ colors = dict(
     accent1="#F29018",
     accent2="#F63D1E",
     accent3="#78479D",
+    accent4="#74A4BC",
 )
 diverging_cmap = "RdYlBu_r"
 
 
-
 # Function to configure matplotlib styles
 def configure_matplotlib_style():
-    custom_color_palette = [colors[c] for c in ["accent1", "accent2", "accent3"]]
+    custom_color_palette = [
+        colors[c] for c in ["accent1", "accent2", "accent3", "accent4"]
+    ]
 
     # Configure global settings
     font_size = 9
@@ -27,7 +29,7 @@ def configure_matplotlib_style():
         {
             "text.usetex": True,  # use LaTeX to write all text
             "pgf.texsystem": "pdflatex",
-            #"axes.grid": False,  # Disable grid lines
+            # "axes.grid": False,  # Disable grid lines
             "pgf.rcfonts": False,
             "axes.titlelocation": "left",  # Axis title alignment left
             "figure.facecolor": "white",  # Background color for the figure
@@ -35,17 +37,13 @@ def configure_matplotlib_style():
             "axes.prop_cycle": cycler(
                 color=custom_color_palette
             ),  # Custom color palette
- 
             "figure.titlesize": 11,  # Figure title font size
             "figure.titleweight": "bold",  # Figure title bold
-
             "xtick.direction": "in",  # X-axis ticks inward
             "ytick.direction": "in",  # Y-axis ticks inward
             "xtick.major.size": 3,  # Major tick size for x-axis
             "ytick.major.size": 3,  # Major tick size for y-axis
- 
-
-            'text.latex.preamble' : "\\usepackage{sansmath}\\sansmath",
+            "text.latex.preamble": "\\usepackage{sansmath}\\sansmath",
             "axes.edgecolor": colors["darkgray"],  # Spine color
             "axes.labelcolor": colors["darkgray"],  # Axis label color
             "axes.titlesize": font_size,  # Axis title font size
@@ -57,7 +55,8 @@ def configure_matplotlib_style():
             "xtick.labelsize": font_size,  # X-axis tick label size
             "ytick.labelsize": font_size,  # Y-axis tick label size
[…]

### 2024-11-17 scripts/stbf_struct/accuracy.py

@@ -1,0 +1,107 @@
+import os
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import (colors, configure_matplotlib_style, save_pgf_trim,
+                   textwidth_in)
+
+configure_matplotlib_style()
+
+iteration_out_path = "data/stbf_struct/iterations.csv"
+
+
+classifier_models = ["stbf_kron_toep_shrunk", "stbf_loocv", "stbf_emp", "xdawn_rg"]
+train_blocks_range = range(1, 9 + 1)
+
+accuracy_out_path = "data/stbf_struct/accuracy.csv"
+
+accuracy = pd.read_csv(accuracy_out_path)
+accuracy["accuracy"] = accuracy["accuracy"] * 100
+accuracy.rename(
+    columns={
+        "n_train_blocks": "Amount of training data",
+        "accuracy": "Accuracy (\\%)",
+        "classifier_model": "Classifier",
+    },
+    inplace=True,
+)
+model_rename = {
+    "stbf_emp": "STBF-emp",
+    "stbf_loocv": "STBF-shrunk",
+    "stbf_kron_toep_shrunk": "STBF-struct",
+    "xdawn_rg": "XDAWN+RG",
+}
+accuracy.replace(model_rename, inplace=True)
+
+
+def plot_accuracy(
+    ax, trial_accuracy, show_chance_label=False, show_legend=False, title=None
+):
+    ax = sns.lineplot(
+        data=trial_accuracy,
+        x="Amount of training data",
+        y="Accuracy (\\%)",
+        hue="Classifier",
+        style="Classifier",
+        ax=ax,
+        legend=show_legend,
+        errorbar=None,
+        dashes=False,
+        markers="sDoP",
+        markersize=3,
+        mew=0,
+    )
+    ax.axhline(100 / 9, color=colors["lightgray"], linestyle="--", linewidth=1)
+    ax.set_ylim([0, 100])
+    ax.set_xticks([2, 4, 6, 8])
+
+    if show_chance_label:
+        ax.annotate(
+            "chance",
+            fontsize=9,
+            xy=(0.05, 1.2 / 9),
+            ha="left",
+            va="bottom",
+            xycoords="axes fraction",
+            color=colors["darkgray"],
+            bbox=dict(facecolor="white", linewidth=0, pad=0),
+        )
+    if title is not None:
+        ax.set_title(title)
+    if show_legend:
[…]

### 2024-11-17 scripts/stbf_struct/kronecker_iteration.py

@@ -3,7 +3,8 @@ import matplotlib.cm
 import matplotlib.pyplot as plt
 import pandas as pd
 import seaborn as sns
-from setup import colors, configure_matplotlib_style, save_pgf_trim
+from setup import (colors, configure_matplotlib_style, save_pgf_trim,
+                   textwidth_in)
 
 configure_matplotlib_style()
 

### 2024-11-18 rebuttal.tex
Update figure 5

@@
 	\item According to the annotations, typographical errors have been
-	corrected and imprecise or inaccurate statements have been adjusted.
+	corrected in text and figures, and imprecise or inaccurate statements have been adjusted.
 	The recommended references have been discussed and, where necessary,
-	statements were adapted to accommodate this.
+	statements were updated accordingly.
+  Figure 1.5 was corrected and matched with the correct caption.
 \end{enumerate}

### 2024-11-18 chapters/stbf_struct.tex
Rework fig 1.3 and 1.4
Update figure 5

@@
   \begin{figure}
-		\includegraphics[width=.66\linewidth]{figures/stbf_struct/accuracy.eps}
+    %\includegraphics[width=.66\linewidth]{figures/stbf_struct/accuracy.eps}
+    \sffamily
+    \sansmath
+    \input{figures/stbf_struct/accuracy.pgf}
     \caption[Clasifier accuracy in function of available training data.]{%
@@
 	\end{figure}
-  \todo{the legend does not correspond to the text (method names are
-  different)}
 
   \fullpagefig{%
-		\includegraphics[width=\textwidth]{figures/stbf_struct/accuracy_app.eps}
+    \sffamily
+    \sansmath
+    \input{figures/stbf_struct/accuracy_all.pgf}
   }{%
@@
 	\begin{figure}
-		\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
-    \caption[Classifier training time.]{Median fold training time for different classifiers at different spatial
-			resolution levels evaluated over all training folds for all subjects.
-			Error bars represent standard deviation. The training time of \textsc{stbf-shrunk} increases more
-			steeply with resolution compared to \textsc{stbf-struct}.
-			All \textsc{stbf} classifiers can be trained significantly faster than
-			\textsc{xdawn+rg}.}
+		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
+    \sffamily
+    \sansmath
+    \input{figures/stbf_struct/training_time.pgf}
+    \caption[Classifier training time.]{Training time boxplots of different
+    classifiers and individual results for all training folds for all subjects
+    using 32 channels. Using structured covariance estimation greatly reduces
+    training time.}
 		\label{fig:stbf-struct/training-time}
@@
 	\cref{fig:stbf-struct/training-time} shows that the training time of
-	\textsc{stbf-struct} increases less steeply than that of \textsc{stbf-shrunk}
-	and \textsc{stbf-emp}. The training time of all three \ac{stbf}-based classifiers is
-  much lower\todo{qualitative} than that of \textsc{xdawn+rg}, which appears nearly constant when using
-	4, 8, 16, or 32 channels.
+	\textsc{stbf-struct} has a lower median training time than
+  \textsc{stbf-shrunk}, \textsc{stbf-emp} and \textsc{xdawn+rg} when training
+  on all available data and using 32 channels\todo{qualitative}.
 

### 2024-11-18 scripts/stbf_struct/training_time.py

@@ -1,0 +1,65 @@
+import os
+
+import ipdb
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from setup import (colors, configure_matplotlib_style, save_pgf_trim,
+                   textwidth_in)
+
+configure_matplotlib_style()
+
+
+classifier_models = ["stbf_kron_toep_shrunk", "stbf_loocv", "stbf_emp", "xdawn_rg"]
+
+ns_channels = [32]
+
+training_time_path = "data/stbf_struct/training_time.csv"
+
+training_time = pd.read_csv(training_time_path)
+
+training_time.rename(
+    columns={
+        "elapsed": "Training time (s)",
+    },
+    inplace=True,
+)
+model_rename = {
+    "stbf_kron_toep_shrunk": "STBF-struct",
+    "stbf_loocv": "STBF-shrunk",
+    "stbf_emp": "STBF-emp",
+    "xdawn_rg": "XDAWN+RG",
+}
+training_time.replace(model_rename, inplace=True)
+
+
+def plot_training_time(data, ax, label=None, show_legend=False, show_y_label=False):
+    sns.stripplot(
+        data=data,
+        y="classifier_model",
+        x="Training time (s)",
+        hue="classifier_model",
+        ax=ax,
+        size=2,
+        zorder=0,
+    )
+    sns.boxplot(
+        data=data,
+        y="classifier_model",
+        x="Training time (s)",
+        color=colors["darkgray"],
+        ax=ax,
+        fill=False,
+        showfliers=False,
+        linewidth=1,
+        zorder=1,
+    )
+
+    ax.set_ylabel("")
+
+
+fig, ax = plt.subplots(1, 1, constrained_layout=True)
+plot_training_time(training_time[training_time["n_channels"] == 32], ax)
+
+save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)

### 2024-11-18 rebuttal.tex
Add accuracies table

@@
 	\item Part of the confusion between active and passive BCI might be due
-	to the poor choice of phrasing ``paradigms that require little active participation from the user, like redirect-
-	ing attention or initiating imagined actions, are classified as passive BCI.''.
+	to the poor choice of phrasing ``paradigms that require little active
+  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''.
 	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
@@
 	comparison of sample rate}
+  \item Appendix 1.A now fully reports all obtained accuracies to give a better
+    insight in effect sizes. Wherever the significant effects have been mentioned
+    in the text, the effect size has now been reported along with it.
 	\item \todo{report actual accuracies performance stbf in table}

### 2024-11-18 chapters/bci.tex
Add accuracies table

@@
 Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
-of a small group of neurons,\todo{qualitative}
-or even intracellular single neuron action potential
+of a small group of neurons, or even intracellular single neuron action potential
 spikes.
@@
 
-Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
+Larger implanted electrodes (usually a few millimeters in size) are referred to
+as intracranial \ac{eeg} (i\ac{eeg}).
 Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
- and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
-symptoms~\cite{Arlotti2018}.\todo{qualify size}
+ and other applications, like closed-loop adaptive control for Deep Brain Stimulation
+ to mitigate Parkinson's disease
+symptoms~\cite{Arlotti2018}.
 Depth electrodes are longer rods with multiple electrodes that can extend deeper into

### 2024-11-18 chapters/gaze_independence.tex
Add accuracies table

@@
 stimulation do not rely on gaze redirection but also result in lower \acp{itr}
-compared to visual paradigms\todo{qualitative}.
+compared to visual paradigms, with \acp{itr} ranging between 1 and 10 bits/min
+for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
 While auditory and somatosensory stimulation on their own might yield poor
@@
 operation and from user-dependent
-variability~\cite{Severens2014,Reichert2020b}.\todo{qualitative}
+variability~\cite{Severens2014,Reichert2020b}.
 \textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
@@
 concurrently decoding the covert \ac{vsa} shift from spectral content, but
-their study only used a few stimulation targets.\todo{qualitative}
-In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
-leveraging spatial attention remains underexplored.
+their study only used two stimulation targets.
+In general, we conclude that gaze-independent decoding in the fast-paced reactive
+\ac{erp} paradigm leveraging spatial attention remains underexplored.
 
@@
 settings~\cite{Treder2010}, the \ac{snr} of
-the \ac{erp} is lower than in overt attention settings\todo{qualitative}.
+the \ac{erp} is lower than in overt attention settings.
 Therefore, a straightforward way to reach satisfactory

### 2024-11-18 chapters/stbf_struct.tex
Add accuracies table

@@
 	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
-  channels vEOG and hEOG, according to adaptive Z-score
-  thresholding.\todo{details? threshold for rejection?}
+  channels vEOG and hEOG using iterative Z-score thresholding.
+  Components with a Z-scored pearson correlation coefficient
+  exceeding the other 3 times the standard deviation of the Z-scored
+  correlation coefficients of other component's Z-scored were iteratively
+  rejected until none that exceed the threshold remained.
   Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
@@
 	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
-  the training process\todo{qualitative}.
+  the  training process.
 
@@
 
-  \fullpagefig{%
-    \sffamily
-    \sansmath
-    \input{figures/stbf_struct/accuracy_all.pgf}
-  }{%
-    \caption[Accuracy averaging over different numbers of trials.]{%
-      Accuracy of the different classifiers for all 21 subjects relative to the
-		  number of blocks available for training. One block consists of 135
-		  epochs and corresponds to 27 seconds of stimulation. Accuracies
-		  are shown for the evaluation settings averaging over different
-	    numbers of trials, ranging from 1 to 15.
-    }
-    \label{fig:stbf-struct/accuracy-ap}
-  }
-
-  \begin{table}[p]
-		\centering
-    \makebox[\textwidth][c]{%
-		\input{chapters/stbf_struct_tests_1.tex}
-  }
-    \caption[Statistical significance of differences in classifier performance (1 repetition).]{%
-      Statistical significance of differences in classifier performance.
-      $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
-			Holm correction using one testing trial for different classifiers
-			and levels of data availability. $p$-values $<0.05$ are considered significant
-			and marked in grey.}
-		\label{tab:stbf-struct/p-values-1}
-    \bigskip
-
-		\centering
-    \makebox[\textwidth][c]{%
-		\input{chapters/stbf_struct_tests_2.tex}
-    }
-		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
-      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over two testing trials.}
-		\label{tab:stbf-struct/p-values-2}
-    \bigskip
-
-		\centering
-    \makebox[\textwidth][c]{%
-		\input{chapters/stbf_struct_tests_3.tex}
-    }
-    \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
-      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over five testing trials.}
-		\label{tab:stbf-struct/p-values-5}
-	\end{table}
-	\unskip
-
 	The tables show that \textsc{stbf-struct} has a significant advantage over
@@
 	\end{figure}
-  \todo{the legend does not correspond to the text (method names are
-  different)}
 
-	\cref{fig:stbf-struct/training-time} shows that the training time of
+	\Cref{fig:stbf-struct/training-time} shows that the training time of
 	\textsc{stbf-struct} has a lower median training time than
@@
 %\end{equation}
+
+\clearpage
+\begin{subappendices}
+  \section{Accuracies per number of trials and available training data}
+
+  \footnotesize
+  \begin{table}[!h]
+	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
+      \caption{Accuracies for \textsc{stbf-struct}.}
+  \end{table}
+  \begin{table}[!h]
+	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
+      \caption{Accuracies for \textsc{stbf-shrunk}.}
+  \end{table}
+  \begin{table}[!h]
+	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
+      \caption{Accuracies for \textsc{stbf-emp}.}
+  \end{table}
+  \begin{table}[!h]
+	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
+      \caption{Accuracies for \textsc{xdawn+rg}.}
+  \end{table}
+
+  \clearpage
+
+  \fullpagefig{%
+    \sffamily
+    \sansmath
+    \input{figures/stbf_struct/accuracy_all.pgf}
+  }{%
+    \caption[Accuracy averaging over different numbers of trials.]{%
+      Accuracy of the different classifiers for all 21 subjects relative to the
+		  number of blocks available for training. One block consists of 135
+		  epochs and corresponds to 27 seconds of stimulation. Accuracies
+		  are shown for the evaluation settings averaging over different
+	    numbers of trials, ranging from 1 to 15.
+    }
+    \label{fig:stbf-struct/accuracy-ap}
+  }
+
+  \clearpage
+  \section{Statistical comparisons}
+  \begin{table}[h]
+    \footnotesize
+    \begin{subtable}{\linewidth}
+	    \input{chapters/stbf_struct_tests_1.tex}
+	    \caption{Evaluation on one testing trial.}
+      \label{tab:stbf-struct/p-values-1}
+    \end{subtable}
+    \bigskip
+
+    \begin{subtable}{\linewidth}
+	    \input{chapters/stbf_struct_tests_2.tex}
+	    \caption{Evaluation averaging over two testing trials.}
+	    \label{tab:stbf-struct/p-values-2}
+    \end{subtable}
+    \bigskip
+
+    \begin{subtable}{\linewidth}
+	    \input{chapters/stbf_struct_tests_3.tex}
+	    \caption{Evaluation averaging over five testing trials.}
+	    \label{tab:stbf-struct/p-values-5}
+    \end{subtable}
+    \caption[Statistical significance of differences in classifier
+    performance.]{%
+        Statistical significance of differences in classifier performance.
+        $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
+	    	Holm correction using one testing trial for different classifiers
+	    	and levels of data availability. $p$-values $<0.05$ are considered
+        significant.
+    }
+	\end{table}
+
+\end{subappendices}

### 2024-11-18 scripts/stbf_struct/accuracy.py

@@ -1,5 +1,6 @@
 import os
 
+import ipdb
 import matplotlib.pyplot as plt
 import pandas as pd
 import seaborn as sns
@@ -78,6 +79,9 @@ def plot_accuracy(
         )
         plt.setp(ax.get_legend().get_texts(), fontsize="7")  # for legend text
         plt.setp(ax.get_legend().get_title(), fontsize="7")
+
+
+ipdb.set_trace()
 
 
 fig, ax = plt.subplots(1, 1, constrained_layout=True)

### 2024-11-18 rebuttal.tex
Textcites

@@
     in the text, the effect size has now been reported along with it.
-	\item \todo{report actual accuracies performance stbf in table}
 	\item In the results section of chapter 4, statistical comparisons were
@@
 	with our partner patient centers, with an adapted interface and experimental protocol.
-	\todo{indicate in manuscript covid}
+  This limitation has now clearly been mentioned in the future work section of
+  chapter 8.
 	\item Recorded datasets could not be made publicly available since the
@@
 \begin{enumerate}
-	\item \todo{correct si measurements Hz, mm, cm, V, µV, $\mu$ V}
-	\item \todo{do proper check in textcite, especially in chapter 3}
-	\item \todo{check for contractions and remove we've, isn't, wouldn't}
+	\item Measurements unit representation has been uniformized according to the
+    SI standard.
+  \item Numerical citations have now consistently been replaced by text-style citations
+    mentioning the author names where appropriate in the sentence.
+  \item Contractions have been remove throughout the manuscript.
 \end{enumerate}

### 2024-11-18 chapters/bci.tex
Textcites

@@
 routine.
-Wouldn't it be great if we could help them regain some quality of life by making them
+Would it not be great if we could help them regain some quality of life by making them
 more independent?
@@
 body cannot follow.
-Couldn't we, therefore, design a system that directly interacts with the mind?
+Could we therefore not design a system that directly interacts with the mind?
 This system can then form an interface between the mind and any kind of technology, like
@@
 \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
-where one of 200 targets can be accurately selected with only 800ms of stimulation.
+where one of 200 targets can be accurately selected with only 800 ms of stimulation.
 
@@
 Additionally, notch filtering can be used to remove strong power line interference\footnote{%
-Typically at 50Hz in Europe.}.
+Typically at 50 Hz in Europe.}.
 This interference can heavily contaminate \ac{eeg} recordings and even leak through when
@@
 effort and short calibration time.
-First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
+First established by \textcite{Farwell1988} in 1988, it is well supported by literature and
 has been proven to work for those with \acl{sspi} in day-to-day home use~\cite{Wolpaw2018}.
@@
 period of time by changing in color, brightness, or size.
-Intensifications of different targets are usually 100-200ms apart and last for about
-100ms~\cite{Sellers2006a}.
+Intensifications of different targets are usually 100-200 ms apart and last for about
+100 ms~\cite{Sellers2006a}.
 On each flash observed by the user, specific brain activity is elicited.
@@
 in a row-column strategy and using a sequence of selections, giving rise to the classic
-matrix speller of~\cite{Farwell1988}.
+matrix speller of \textcite{Farwell1988}.
 Other optimizations increase response \ac{snr}, like using flashing face images as
@@
 Some \ac{bci}\footnote{In \ac{erp} analysis, components are sometimes referred to with
-the timing around which they occur (e.g., N170, negative after 170ms), or by their rank
+the timing around which they occur (e.g., N170, negative after 170 ms), or by their rank
 in order of occurrence (N1, the first negative component).
@@
 \subsection{Preprocessing \& feature extraction}
-For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz and
-16Hz.
+For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5 Hz and
+16 Hz.
 Re-referencing can be done to average mastoids to highlight the contribution of the P3

### 2024-11-18 chapters/bttda.tex
Textcites

@@
 		number of runs (\# Runs).
-    Adapted from~\cite{Aristimunha2023}
-    and~\cite{Chevallier2024}.}
+    Adapted from \textcite{Aristimunha2023} and \textcite{Chevallier2024}.}
 	\label{tab:moabb}
@@
 
-Furthermore, as additional comparison methods, we used the methods evaluated in~\cite{Chevallier2024}.
+Furthermore, as additional comparison methods, we used the methods evaluated
+by~\textcite{Chevallier2024}.
 For the ERP datasets, these were the Riemannian Geometry-based methods
@@
 FilterBank+SVM, respectively).
-We refer to~\cite{Chevallier2024} for implementation details of these comparison
+We refer to \textcite{Chevallier2024} for implementation details of these comparison
 methods.
@@
 EEG signals for the evaluated datasets were recorded at the sample rate given
-by \cref{tab:moabb} and band-pass filtered between 1Hz
-and 24Hz.
+by \cref{tab:moabb} and band-pass filtered between 1 Hz
+and 24 Hz.
 The signals were cut into epochs starting from stimulus onset with a
 dataset-specific length given by \cref{tab:moabb}.
-For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
+For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48 Hz.
 For the BNCI2014-008 and BNCI2015-003 datasets, this resulted in matrices of
@@
 		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
-    Scores for other decoders were taken from \cite{Chevallier2024}.
+    Scores for other decoders were taken from \textcite{Chevallier2024}.
 		BTTDA reaches the highest performance for both evaluated datasets, closely
@@
 The MI datasets listed in \cref{tab:moabb} were first band-pass filtered
-between 8 and 30Hz and cut into
+between 8 and 30 Hz and cut into
 epochs with time windows as specified by \cref{tab:moabb}.
@@
 convolution, with 16 wavelet frequencies logarithmically spaced between 8 and
-32Hz.
+32 Hz.
 The number of wavelet cycles $c$ varied with wavelet frequency $f$ as
@@
 the evaluated motor imagery datasets.
-Note that, in line with~\cite{Chevallier2024}, accuracy is reported for the
+Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
 multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
@@
 		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
-		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
+		Scores for other decoders were taken from \textcite{Chevallier2024}. Accuracy
     is listed for BNCI2014-001, and area under the receiver-operator

### 2024-11-18 chapters/conclusion.tex
Textcites

@@
 
-In gaze-independence decoding, our current efforts focus on,
-counterintuitively, integrating eye tracking into the decoding strategy.
+In gaze-independence decoding, our current efforts focus on --
+counterintuitively -- integrating eye tracking into the decoding strategy.
 Since we are able to explicitly discern overt, covert, and split \ac{vsa} from
@@
 \ac{sspgi}.
-We aim to revisit the participants with an on-line experiment and comfort
-assessment to establish the usability of the proposed interface and compare its
-\ac{itr} to the state-of-the-art literature.
+As the work presented in this thesis started during the COVID-19 pandemic,
+experimentation was delayed and the envisioned application could not be
+validated within the time frame of the doctoral project.
+To complete this work, we aim to revisit the participants with an on-line
+experiment and comfort assessment to establish the usability of the proposed
+interface and compare its \ac{itr} to the state-of-the-art literature.
 
@@
 experimentation, this is actually a strength, since the user can help the researcher gain
-insight into what works and doesn’t work for them, and where research efforts
+insight into what works and what does not work for them, and where research efforts
 should be focused.

### 2024-11-18 chapters/covert_align.tex
Textcites

@@
 \ac{eeg} for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
-(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
+(Compumedics Neuroscan, Australia) at 2048 Hz and 62 Ag/AgCl active electrodes arranged in the
 international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
@@
 60 cm from a CRT-emulating monitor (VPixx
-Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
+Technologies, Canada) operating at a refresh rate of 120 Hz, displaying 6
 circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
@@
 For this evaluation, epochs were rejected when the peak-to-peak amplitude
-exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
+exceeded 800 µV, and for the CVSA-ERP dataset, if the user's gaze differed by more than
 10 degrees of visual angle from the fixation crosshair.
@@
 present in covert and split \ac{vsa} compared to overt \ac{vsa},
-we justify our approach similarly to \cite{Hardiansyah2020} by
+we justify our approach in a similar manner to \textcite{Hardiansyah2020} by
 observing that the increased discrimination performance of a machine-learning
@@
 However, the large increase in performance for covert \ac{vsa} reported by
-\cite{Arico2014}\todo{qualitative} could also
+\textcite{Arico2014}\todo{qualitative} could also
 be explained by overfitting on artifacts amplified by aligning, since their
@@
 significantly and also did not increase performance over its first-stage
-classifier in overt \ac{vsa}, contrary to what was reported by \cite{Mowla2017}.
+classifier in overt \ac{vsa}, contrary to what was reported by \textcite{Mowla2017}.
 While they report that \ac{cble} is relatively independent of the
@@
 the presence of jitter or the generally higher performance of \ac{tlda} compared to the
-first-stage classifiers tested by \cite{Mowla2017}, leaving less
+first-stage classifiers tested by \textcite{Mowla2017}, leaving less
 performance to be gained.
 
-\cite{Thompson2012} already attempted applying \ac{cble} in an iterative scheme
+\textcite{Thompson2012} already attempted applying \ac{cble} in an iterative scheme
 but did not report any results due to convergence issues.
@@
 
-\cite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
+\textcite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
 training data once with time-shifted copies based on \ac{cble}-estimated jitter.
@@
 argue that our method can cope with more granular latency differences, as
-\cite{Zisk2022} augment the data with just one positive and negative time
+\textcite{Zisk2022} augment the data with just one positive and negative time
 shift.

### 2024-11-18 chapters/gaze_independence.tex
Textcites

@@
 
-P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
+P3 latency generally falls between 350 ms and 600 ms~\cite{Luck2014}, but this
 value is heavily dependent on the subject and the task and can vary from trial

### 2024-11-18 chapters/stbf_struct.tex
Textcites

@@
 
-  \noindent\emph{This chapter was
-  published as~\textcite{VanDenKerchove2022}.}
+  \noindent\emph{This chapter, except
+  \crefrange{tab:stbf-struct/accuracies/stbf-struct}{tab:stbf-struct/accuracies/xdawn-rg},
+  was published as~\textcite{VanDenKerchove2022}.}
 
@@
 	We opt for the \ac{loocv} shrinkage estimator because it avoids some of the
-	assumptions made by~\cite{Ledoit2004} and~\cite{Chen2010} and
+	assumptions made by \textcite{Ledoit2004} and \textcite{Chen2010} and
 	because it generalizes to structured covariance estimation as described in
@@
 
-	The \ac{eeg} recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
+	The \ac{eeg} recorded at 2048 Hz was re-referenced off-line to the average of the mastoids.
 	The reference electrodes were dropped from the analysis.
-	Data were subsequently filtered between 0.5Hz and 16Hz using forward-backward
+	Data were subsequently filtered between 0.5 Hz and 16 Hz using forward-backward
 	filtering with a fourth-order Butterworth IIR filter.
@@
 	Before calculating the \ac{stbf}, the signal was downsampled to
-	32Hz or twice the low-pass frequency 16Hz, resulting in 17 time samples
+	32 Hz or twice the low-pass frequency 16 Hz, resulting in 17 time samples
 	between 0.1 and 0.6s. According to the Nyquist Theorem, more samples would not
@@
 	\textsc{xdawn+rg} was trained and validated without downsampling using epochs
-	at the original sample rate of 2048Hz.
+	at the original sample rate of 2048 Hz.
 
@@
 	accuracy across the amounts of training data and the number of trials
-  used for evaluation\todo{qualitative}.
+  used for evaluation.
 	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
@@
 	corresponding number of blocks in the training folds and dropping the rest.
-	\Cref{fig:accuracy} and \cref{fig:stbf-struct/accuracy-ap} show each classifier's
-	accuracy relative to the data availability.
+  \Cref{fig:accuracy}, \cref{fig:stbf-struct/accuracy-ap} and
+  \crefrange{tab:stbf-struct/accuracy/stbf-struct}{tab:stbf-struct/accuracy/xdawn-rg}
+  shows each classifier's accuracy relative to the data availability and trials
+  used for agveraging.
 	We statistically compare the two newly proposed classifiers,
-	\textsc{stbf-struct} and \textsc{stbf-shrunk}, for different levels of training
+	\textsc{stbf-struct} and \textsc{stbf-shrunk} with different levels of training
 	data availability using a one-sided paired Wilcoxon rank-sum test with Holm correction for the multiple pairwise comparisons between classifiers.
-	We performed this analysis three times: by only using the first trial of a
-	block, by averaging epochs across the first two trials of a block and across
-	the first five trials of a block.
-	Results validated on one trial are reported in \cref{tab:stbf-struct/p-values-1}, 2-trial results
-	in \cref{tab:stbf-struct/p-values-2}, and 5-trial results in \cref{tab:stbf-struct/p-values-5}.
+	Statistical comparisons were performed for trial
+  \cref{tab:stbf-struct/p-values-1}, two \cref{tab:stbf-struct/p-values-2}
+  and five trials \cref{tab:stbf-struct/p-values-5}.
 
@@
 	The tables show that \textsc{stbf-struct} has a significant advantage over
-	\textsc{stbf-shrunk} when the number of training blocks is low.
-  This effect is present for 1-, 2- and 5-trial evaluation\todo{qualitative}.
-	This advantage decreases (the $p$-value increases) when
-  adding more training blocks\todo{qualitative}.
+  \textsc{stbf-shrunk} when the number of training blocks is low, with
+  significance $p=0.005$ and  an accuracy increase of 4,14$\%.$ for a single training block
+  and testing trial.
+  This effect is present for 1-, 2- and 5-trial evaluation, up to 3, 4 and 3
+  blocks respectively, but the advantage decreases when adding more training blocks.
 	Both \textsc{stbf-struct} and \textsc{stbf-shrunk} perform significantly better
-  than \textsc{stbf-emp} for all evaluated settings\todo{qualitative}.
+  than \textsc{stbf-emp} for all evaluated settings, with $p<0.001$ for all number
+  of training blocks and testing trials, with accuracy
+  increase of 6.92 and 2.78$\%.$ respectively for a single training block
+  and testing trial.
 	Compared to \textsc{xdawn+rg}, \textsc{stbf-struct} also has significantly
 	higher accuracy in almost all evaluated settings, except when using only one
-  training block\todo{qualitative}.
+  training block, with significance $p<0.001$ and an accuracy increase of
+  5.20$\%.$ when using all 9 training blocks and 1 testing trial.
 	\textsc{stbf-shrunk} does not outperform \textsc{xdawn+rg} when training data
-  is low but gains a significant advantage	when using more training
-  data\todo{qualitative}.
+  is low (up to 3,3 and 2 blocks respectively for 1, 2 and 5 testing trials),
+  but gains a significant advantage	when using more training data, with
+  significange $p=0.001$ and an accuracy increase of 5.33$\%.$ when using all
+  9 training blocks and 1 testing trial.
 
@@
 	\Cref{fig:stbf-struct/training-time} shows that the training time of
-	\textsc{stbf-struct} has a lower median training time than
-  \textsc{stbf-shrunk}, \textsc{stbf-emp} and \textsc{xdawn+rg} when training
-  on all available data and using 32 channels\todo{qualitative}.
+  \textsc{stbf-struct} has a lower median training time (0.06 s) than
+  \textsc{stbf-shrunk} (0.33 s), \textsc{stbf-emp} (0.30 s) and
+  \textsc{xdawn+rg} (1.04 s) when training
+  on all available data and using 32 channels.
 
@@
 	\subsection{Classification accuracy}
-	As evidenced by \cref{fig:accuracy} and Tables \ref{tab:stbf-struct/p-values-1}-\ref{tab:stbf-struct/p-values-5}, the regularized classifiers \textsc{stbf-shrunk}
+	As evidenced by \cref{fig:accuracy} and \cref{tab:stbf-struct/p-values},
+  the regularized classifiers \textsc{stbf-shrunk}
 	and \textsc{stbf-struct} significantly improve the classification accuracy
 	compared to the original \textsc{stbf-emp} for all numbers of training blocks
-  indicated\todo{qualitative}.
+  indicated.
 	We believe there are three reasons for this.
@@
 	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available
-  training data is small\todo{qualitative}.
+  training data is small ($<540$ epochs).
 	This property is an attractive advantage in a \ac{bci} setting since it is desirable to keep the calibration (training) phase as short as possible without losing accuracy.
@@
 	covariance.
-  The increase in accuracy for small training set sizes\todo{qualitative} can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:stbf-struct/discussion/param-complex}), increasing the stability of matrix inversions.
+  The increase in accuracy for small training set sizes can also be explained by the smaller number of parameters necessary to estimate the inverse covariance (see \cref{sec:stbf-struct/discussion/param-complex}), increasing the stability of matrix inversions.
 
@@
 	that \textsc{stbf-struct} reaches similar accuracy when using only one block of
-  training data\todo{qualitative}.
-	The authors suspect this is due to both
-	classifiers having insufficient training information to reach
-	satisfactory classification accuracy.
+  training data, with accuracies of and 34.70\% 32.5\% respectively and $p=0.086$.
+	The authors suspect this is due to both	classifiers having insufficient training
+  information to reach satisfactory classification accuracy.
 	When more data are available, \textsc{stbf-struct} reaches significantly
-  better accuracies\todo{qualitative}.
+  better accuracies.
 	Combined with the benefits laid out in \cref{sec:stbf-struct/discussion/param-complex}, this
@@
 	\textsc{xdawn+rg} using a few training blocks, but this improves as the
-  training data increases\todo{qualitative}.
+  training data increases.
 
@@
 	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \textsc{stbf-struct}.
-	The results presented in \cref{fig:stbf-struct/training-time} confirm these
-	calculations.
-	It can be observed that the training time of \textsc{stbf-struct} stays low
-  compared to \textsc{stbf-emp} and \textsc{stbf-shrunk} when dimensionality
-  increases\todo{qualitative}.
+	The results presented in \cref{fig:stbf-struct/training-time} confirm that
+  \textsc{stbf-struct} is much faster in operation than other classifiers.
 
@@
   data are sparsely available and that it can be computed faster and with
-  substantially less memory usage\todo{qualitative}.
+  substantially less memory usage.
 	Since these algorithms are not paradigm-specific,  the conclusions can be generalized to
@@
 	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
-      \caption{Accuracies for \textsc{stbf-struct}.}
+      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for \textsc{stbf-struct}.}
+      \label{tab:stbf-struct/accuracies/stbf-struct}
   \end{table}
@@
 	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
-      \caption{Accuracies for \textsc{stbf-shrunk}.}
+      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
+      \textsc{stbf-shrunk}.}
+      \label{tab:stbf-struct/accuracies/stbf-shrunk}
   \end{table}
@@
 	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
-      \caption{Accuracies for \textsc{stbf-emp}.}
+      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
+      \textsc{stbf-emp}.}
+      \label{tab:stbf-struct/accuracies/stbf-emp}
   \end{table}
@@
 	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
-      \caption{Accuracies for \textsc{xdawn+rg}.}
+      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
+      \textsc{xdawn+rg}.}
+      \label{tab:stbf-struct/accuracies/xdawn-rg}
   \end{table}
@@
     \end{subtable}
-    \bigskip
+    \smallskip
 
@@
     }
+	  \label{tab:stbf-struct/p-values}
 	\end{table}

### 2024-11-18 chapters/wcble.tex
Textcites

@@
 by carefully selecting peaks or pre-determined regions of interest.
-\textcite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
+\textcite{Hardiansyah2020} use an \ac{snr} boosting method, applied to a specific
 time-window for each \ac{erp} component to extract a latency per component.
@@
 However, the resulting score-over-time vectors per epoch still suffer from jitter.
-For classification, we follow the approach of \cite{Mowla2017}, using a
+For classification, we follow the approach of \textcite{Mowla2017}, using a
 maximum-level hierarchical Daubechies-4 wavelet transform to reduce
@@
 The \ac{cble} latency estimate is defined only for target epochs as
-$s_\mathrm{target}=\argmax_s{y_s}$.
+$\textstyle{s_\mathrm{target}=\argmax_s{y_s}}$.
 This is the point in time where the target class reaches the largest separation

### 2024-11-18 scripts/stbf_struct/training_time.py

@@ -54,6 +54,7 @@ def plot_training_time(data, ax, label=None, show_legend=False, show_y_label=Fal
         showfliers=False,
         linewidth=1,
         zorder=1,
+        width=0.5,
     )
 
     ax.set_ylabel("")

### 2024-11-19 chapters/bttda.tex
Quantify qualitative statements in bttda chapter

@@
 smaller, but no significant differences were found between HODA+LDA,
-PARAFACDA+LDA, and BTTDA+LDA\todo{qualitative}.
+PARAFACDA+LDA, and BTTDA+LDA.
 
@@
 HODA model, our results show that BTTDA+LDA performs consistently on par or
-higher than HODA+LDA\todo{qualitative}.
+higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
+$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
+\ac{mi} datasets.
 The classification score for BTTDA+LDA was also slightly
-higher\todo{qualitative} than that of
-restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
-results were not significant, and further studies with more datasets and
-subjects should show whether this holds.
+higher than that of
+restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
+respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
+respectively), but these results were not significant, and further studies with
+more datasets and :subjects should show whether this holds.
 \Cref{fig:blocks} shows that there is added value in finding extra blocks
@@
 problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
-imagery problems (third-order tensors)\todo{qualitative}.
+imagery problems (third-order tensors).
 Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible

### 2024-11-19 rebuttal.tex
Quantify qualitative statements in wcble chapter

@@
 	\item Measurements unit representation has been uniformized according to the
-    SI standard.
+    SI standard throughout the text and in figures 5.4-5.6.
   \item Numerical citations have now consistently been replaced by text-style citations
     mentioning the author names where appropriate in the sentence.
-  \item Contractions have been remove throughout the manuscript.
+  \item Contractions have been remove throughout the text.
 \end{enumerate}

### 2024-11-19 chapters/wcble.tex
Quantify qualitative statements in wcble chapter

@@
 lab to the clinical setting.
-For instance, \textcite{Zisk2021} showed that \ac{als} results in significantly
-increased \ac{erp} jitter\todo{qualitative}.
-\textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
-performance, and disproportionately so when the user is not directly gazing at
-intended targets, such as is the case for individuals with \ac{sspgi}.
+For instance, \textcite{Zisk2021} showed that \ac{erp} jitter was significantly higher
+in 6 \ac{als} patients (1350 ms$^2$) compared to neurotypical controls
+(553 ms$^2$).
+\textcite{Arico2014} showed that P3 latency jitter negatively correlated
+($\rho=0.17$) with \ac{bci} accuracy in a covert GeoSpell interface.
+Here the user does not gaze directly at the targets, such as is
+the case for individuals with \ac{sspgi}.
 Therefore, properly accounting for \ac{erp} latency jitter can contribute to
@@
 The noise was scaled at 32 different noise levels, with the \ac{snr} of a
-target epoch ranging from 0 to -31dB, with
+target epoch ranging from 0 to -31 dB, with
 \begin{equation}
@@
 robust in the presence of pink spatiotemporal noise.
-At around -28dB, \ac{wcble} does not significantly outperform \ac{cble}
+At around -28 dB, \ac{wcble} does not significantly outperform \ac{cble}
 anymore, and both methods get overtaken by the strong presence of noise.
@@
 
-\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble}
-decreases with
-$\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
-This difference is particularly large\todo{qualitative} when $\sigma$ is higher ($\sigma=0.2$
-and $\sigma=0.3$) and \ac{snr} is not yet high.
+\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of all
+three methods decreases with $\sigma$ and \ac{snr}, yet \ac{wcble} outperforms
+both \ac{cble} outperforms \ac{tlda} since its performance declines less steeply.
+This difference is particularly large when $\sigma$ is higher
+($\sigma=0.2$ and $\sigma=0.3$ s) and \ac{snr} is not yet high, with accuracies of respectively
+77.50\% for \ac{wcble}, 17.00\% for \ac{cble} and 9.50\% for \ac{tlda} when
+$\sigma=0.3$ and \ac{snr} = -23 dB.
 \Ac{wcble} always scored better than \ac{cble} and \ac{tlda}, except for
 $\sigma=0.1$ and $\text{\ac{snr}} < 27 \text{dB}$, at which point both \ac{cble}
-and \ac{wcble} performed at chance leveltodo{qualitative}.
-When considering the accuracy of \ac{wcble}, we see that it shows more of a
-plateau as \ac{snr} increases\todo{qualitative}.
+and \ac{wcble} performed below the chance level of 50\%.
 
@@
 in \cref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
-potential jitter and lower \ac{snr} than \ac{cble}\todo{qualitative}.
+potential jitter and lower \ac{snr} than \ac{cble}.
 It could therefore be a better candidate to apply in \ac{erp} analysis and
 decoding in those cases where jittered \ac{erp} data is of interest.
-At a certain noise threshold\todo{qualitative}, both methods begin to fail, but \ac{wcble}
-delivers higher accuracy at intermediate \ac{snr} levels than
-\ac{cble}\todo{qualitative}, likely due to the iterative alignment process.
+At a certain noise threshold (-27 dB for $\sigma=0.1$ s, -26 dB for
+$\sigma=0.2$ s and -23 dB for $\sigma=0.3$ s), both methods begin to fail, but \ac{wcble}
+delivers higher accuracy ($>75\%$) at intermediate \ac{snr} levels than
+\ac{cble}, likely due to the iterative alignment process.
 

### 2024-11-19 scripts/wcble/fig_latency_classify.py

@@ -1,19 +1,24 @@
-import matplotlib.pyplot as plt
 import pdb
+
+import matplotlib.pyplot as plt
 import pandas as pd
 import seaborn as sns
-from setup import configure_matplotlib_style,textwidth_in, save_pgf_trim
+from setup import configure_matplotlib_style, save_pgf_trim, textwidth_in
 
 configure_matplotlib_style()
 
 
 # Latency
 df = pd.read_csv("data/wcble/results_align.csv")
-df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
+df = df.replace({"tlda": "tLDA", "CBLE-median": "CBLE", "WCBLE-median": "WCBLE"})
 df = df[df["stc"] == "sine"]
 
-facet_kws = dict(ylim=[0, 1], xlim=[0, -31], despine=False, legend_out=False,
-                 )
+facet_kws = dict(
+    ylim=[0, 1],
+    xlim=[0, -31],
+    despine=False,
+    legend_out=False,
+)
 fg = sns.relplot(
     data=df,
     x="snr",
@@ -24,10 +29,10 @@ fg = sns.relplot(
     kind="line",
     facet_kws=facet_kws,
 )
-fg.set_titles(col_template="$\\sigma={col_name}$s")
+fg.set_titles(col_template="$\\sigma={col_name}$ s")
 fg.set_xlabels("SNR (dB)")
 fg.set_ylabels("$\\rho$")
-save_pgf_trim(fg.fig, fg.axes[0,0], 'figures/wcble/fig_latency.pgf', height=2)
+save_pgf_trim(fg.fig, fg.axes[0, 0], "figures/wcble/fig_latency.pgf", height=2)
 
 
 # Accuracy
@@ -38,7 +43,7 @@ df = pd.concat(
 df = df[df["stc"] == "sine"]
 df["accuracy"] *= 100
 
-df = df.replace({'tlda': 'tLDA', 'CBLE-median':'CBLE', 'WCBLE-median':'WCBLE'})
+df = df.replace({"tlda": "tLDA", "CBLE-median": "CBLE", "WCBLE-median": "WCBLE"})
 
 facet_kws = dict(ylim=[0, 100], xlim=[0, -31], despine=False, legend_out=False)
 fg = sns.relplot(
@@ -51,8 +56,8 @@ fg = sns.relplot(
     kind="line",
[…]

### 2024-11-19 scripts/wcble/fig_simulated.py

@@ -1,10 +1,11 @@
+import pdb
+
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 import seaborn as sns
 from mne import read_epochs
-from setup import configure_matplotlib_style,colors, save_pgf_trim
-import pdb
+from setup import colors, configure_matplotlib_style, save_pgf_trim
 
 configure_matplotlib_style()
 
@@ -46,9 +47,9 @@ def order(epochs, n_epochs=100):
 def plot_img(ax, epochs, cbar_ax=None):
     order_ = order(epochs)
     if cbar_ax is None:
-        axes=ax
+        axes = ax
     else:
-        axes=[ax, cbar_ax]
+        axes = [ax, cbar_ax]
     epochs.plot_image(
         order=order_,
         colorbar=cbar_ax is not None,
@@ -57,7 +58,7 @@ def plot_img(ax, epochs, cbar_ax=None):
         show=False,
         **img_kwargs,
     )
-    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=colors['darkgray'])
+    ax_img.axhline(100, linestyle="--", linewidth=0.5, color=colors["darkgray"])
 
     ax_img.set_title("")
     ax_img.set_xticks([])
@@ -71,15 +72,17 @@ def plot_ts(ax, epochs):
     df = pd.DataFrame(y.T, index=x)
     df.index.name = "time"
     df = df.melt(ignore_index=False, value_name="amplitude")
-    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts,
-    errorbar='sd')
+    sns.lineplot(
+        data=df.reset_index(), x="time", y="amplitude", ax=ax_ts, errorbar="sd"
+    )
     y = epochs["target"].pick("C3").get_data().squeeze() * 1e6
     df = pd.DataFrame(y.T, index=x)
     df.index.name = "time"
     df = df.melt(ignore_index=False, value_name="amplitude")
-    sns.lineplot(data=df.reset_index(), x="time", y="amplitude", ax=ax_ts,
-                 errorbar='sd')
-    ax.set_xlim([-2,2])
+    sns.lineplot(
+        data=df.reset_index(), x="time", y="amplitude", ax=ax_ts, errorbar="sd"
+    )
+    ax.set_xlim([-2, 2])
 
 
 stc = "sine"
[…]

### 2024-11-20 rebuttal.tex
Reply to DeLong test

@@
 \opening{Dear chair and members of the reading committee}
-\todo{fix todos in document}
 
-I would like to extend my sincere gratitude for time and effort you have taken
+I would like to extend my sincere gratitude for the time and effort you have taken
 to review the manuscript of my PhD dissertation.
@@
 The following reply contains a point-by-point response to your review comments.
-Attached, you will also find an updated draft of the manuscript and a version
-with indicated changes for your approval.
+Attached, you will also find an updated draft of the manuscript, as well as a
+version with indicated changes for your approval.
 
@@
 	comparison models.
-	\item \todo{include table with absolute performances for transfer
-	conditions}
+\item Tables 6.2 and 6.3 now report the absolute cross-condition ROC-AUC values
+  per model and dataset for a more complete and interpretable presentation of
+    the results, and the absolute values have been mentioned when referring to
+    effects in the results section.
 \end{enumerate}
@@
 \begin{enumerate}
-	\item \todo{quantify qualitative statements}
-	\item  Mathematical equations and derivations have thoroughly
+	\item \todo{quantify qualitative statements} Throughout the text, qualitative
+    statements have been qualified where possible and appropriate.
+	\item  All mathematical equations and derivations have thoroughly
 	been checked for correctness and notational consistency, and have been
-	corrected throughout the manuscript where necessary.
+	corrected where necessary.
 	\item As indicated by the reviewer, true on-line experiments with live
-	feedback have indeed not yet be conducted, but were foreseen to take
+	feedback have indeed not reported in this thesis, but were foreseen to take
 	place within the framework of this doctoral study.
@@
 	presented conclusions and are the logical next step.
-	Indeed, the start of experimentation with healthy control participants
-	as well as patient groups, was heavily delayed by the COVID-19 pandemic, as the
+	Indeed, the start of experimentation with healthy control participants,
+	as well as with patient groups, was heavily delayed by the COVID-19 pandemic, as the
 	preparatory research started in September 2019 and I officially
-	enrolled in the the doctoral training programme in January 2020.
+	enrolled in the doctoral training program in January 2020.
 	This restricted not only our access to experiment subjects for a
@@
   chapter 8.
+\end{enumerate}
+Minor comments:
+\begin{enumerate}
 	\item Recorded datasets could not be made publicly available since the
@@
 	publishing the data.
-\end{enumerate}
-Minor comments:
-\begin{enumerate}
-	\item
 \end{enumerate}
@@
 	formatting.
-	\item \todo{replace wilcoxon by delong}
+	\item The Wilcoxon signed-rank, a non-parametric,
+    paired test was used to compare ROC-AUC values in sections 4.3.3 and 6.3.1.
+    I agree that the DeLong test is indeed an appropriate test for
+    statistically comparing ROC-AUCs between two models, but it is commonly used
+    to compare ROCs obtained from a single test dataset (potentially obtained
+    through stacking cross-validated predictions).
+    Our experiments, on the other hand, involve calculating ROC-AUCs per
+    cross-validation fold, yielding datasets of multiple ROC-AUCs to compare.
+    In this case, the Wilcoxon signed-rank test is the appropriate
+    test~\cite{Rainio2024}.
+    While both approaches are valid strategies and commonly used, I believe
+    the latter is more interpretable in the context of the generalizability
+    of machine learning models, since it more directly reflects the variance
+    across folds.
+    Hence, I recognize the potential application of the DeLong test here,
+    but I would argue that using Wilcoxon signed-rank tests in combination with
+    the evaluation strategy employed in the experiments is still a valid approach.
    \end{enumerate}
@@
 \begin{enumerate}
-	\item Measurements unit representation has been uniformized according to the
+	\item Measurement unit representation has been uniformized according to the
     SI standard throughout the text and in figures 5.4-5.6.

### 2024-11-20 chapters/bci.tex
Reply to DeLong test

@@
 \label{sec:bci}
-\epigraph{``My mind controls my reality.''}{Frank Herbert -- Children of Dune}
+\epigraph{``My mind controls my reality.''}{\emph{Children of Dune -- Frank
+Herbert}}
 
@@
 
-Starting in the 1970s from methods to establish basic communication from minimal brain
-activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
-into sophisticated interaction schemes that restore or enhance many capabilities by
-bypassing defective parts of the nervous or muscle system.\todo{the BCI concetp
-was proposed at that time, but maybe not yet this form of communication. Also,
-this reference is not from the 1970s, you may want to cite the original paper}
+\Acp{bci} Started as a speculative concept in the 1970's~\cite{Vidal1973}.
+Now, they have gradually evolved from methods  developed in the the 1990's to
+establish basic communication from minimal brain activity
+(yes/no, left/right, \ldots)~\cite{Wolpaw2002} into sophisticated interaction
+schemes that restore or enhance many capabilities by bypassing defective parts
+of the nervous or muscle system.
 By now, classic \acp{bci} methods are starting to mature outside the lab setting and can

### 2024-11-20 chapters/conflict.tex
Self-plagiarism check
Reply to DeLong test

@@
 Arne Van Den Kerchove is funded by a bilateral cooperation program between KU
-Leuven (special research fund of the KU Leuven GPUDL/20/031) and the university
+Leuven (special research fund of the KU Leuven GPUDL/20/031) and the University
 of Lille under the Global PhD Partnership program.
+
 Prof. Marc Van Hulle is supported by research grants received from
@@
 Foundation (AKUL 043).
+
+The authors also acknowledge the support by the RITMEA project co-financed by the
+European Union with the European Regional Development Fund, the French state,
+and the Hauts-de-France Region Council.

### 2024-11-20 chapters/covert_align.tex
Reply to DeLong test

@@
 \label{sec:covert-align}
-\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ A
-Scandal in Bohemia -- Sir Arthur Conan Doyle}
+\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ \emph{A
+Scandal in Bohemia -- Sir Arthur Conan Doyle}}
 \noindent\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
@@
 \textcite{Arico2014} observed higher variability in single-trial P3 peak
-latencies relative to stimulus onset during covert \ac{vsa} compared to overt
-\ac{vsa}\todo{qualitative, discuss}.
+latencies relative to stimulus onset during covert \ac{vsa} (529 ms$^2$) compared to overt
+\ac{vsa} (256$^2$ ms).
 This latency variability contributes to reduced covert \ac{vsa} decoding performance.
@@
     different VSA conditions. Significance was determined using one-sided
-    (\ac{wcble} $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
+    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
     ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
@@
 \end{figure}
+
 Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
@@
 evaluated \ac{vsa} conditions.
-\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets,
-and conditions.
-
-\begin{table}
-  \centering
-  \makebox[\textwidth][c]{%
-    \input{chapters/covert_align_performance.tex}
-  }
-  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
-curve for all evaluated models, visuospatial attention conditions, and
-  datasets.}
-  \label{tab:covert-align/results/single-trial}
-\end{table}
+\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets, and conditions.
 
@@
 highest (stage-two if applicable) classifier score and comparing it to the cued target.
-Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
+Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
 corrected for FDR over decoders.
@@
 datasets, which is greatest for smaller numbers of repetitions and
-decreases as the number of repetitions increases.\todo{qualitative}
+decreases as the number of repetitions increases.
 This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
@@
 \ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
-selection accuracy.\todo{qualitative}
+selection accuracy.
 No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
@@
 in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
-\ac{wcble} in CVSA-ERP.\todo{qualitative}
+\ac{wcble} in CVSA-ERP.
+\todo{qualitative, list accuracies for one repetition}
 
@@
     Significance of differences
-    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
+    was determined with two-sided Wilcoxon signed-rank tests with False Discovery
     Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
@@
 When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
-evaluated settings, similar to the within-subject conditions.\todo{qualitative}
+evaluated settings, similar to the within-subject conditions.
+\todo{qualitative}
 On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
@@
 
-\cref{fig:covert-align/single-trial-roc-auc-diff} shows that \ac{wcble} significantly improves
+\Cref{fig:covert-align/single-trial-roc-auc-diff} shows that \ac{wcble} significantly improves
 covert \ac{vsa} decoding.
@@
         Evidence shows that \ac{erp} latency variability is
-        higher when attention is
-        divided~\cite{Polich2007,Arico2014}.
+        higher when attention is divided~\cite{Polich2007,Arico2014}.
         \textcite{Arico2014} also partly attribute higher latency jitter to the covert \ac{vsa}
@@
 \end{enumerate*}
-\todo{first item qualitative}
 
@@
 only factor hampering covert \ac{vsa} decoding performance.
-However, the large increase in performance for covert \ac{vsa} reported by
-\textcite{Arico2014}\todo{qualitative} could also
-be explained by overfitting on artifacts amplified by aligning, since their
-method is not evaluated on unseen data, as opposed to ours.\todo{qualitative}
+However, the large increase in accuracy for covert \ac{vsa} reported by
+\textcite{Arico2014} (60\% to 95\% with a window focusing on only the P3
+component) could also be explained by overfitting on artifacts amplified by
+aligning, since their method is not evaluated on unseen data, as opposed to ours.
 
@@
 components.
+
+\clearpage
+\begin{subappendices}
+  \section{Single-trial classification performances}
+  \begin{table}[ht]
+  \centering
+  \makebox[\textwidth][c]{%
+    \input{chapters/covert_align_performance.tex}
+  }
+  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
+curve for all evaluated models, visuospatial attention conditions, and
+  datasets.}
+  \label{tab:covert-align/results/single-trial}
+\end{table}
+
+  \clearpage
+  \section{Cross-condition transfer classification performances}
+
+  \begin{table}[ht]
+      \footnotesize
+      \input{chapters/covert_align_cross-CVSA-ERP.tex}
+      \caption[Transfer learning results for CVSA-ERP]{Cross-condition transfer
+      classification performance (cross-validated \ac{rocauc})
+      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
+      on one \ac{vsa} condition and tested across all
+      \ac{vsa}
+      conditions to evaluate gaze-independence and transfer learning
+      performance.}
+  \end{table}
+  \begin{table}[ht]
+      \footnotesize
+      \input{chapters/covert_align_cross-BNCI2014-009.tex}
+      \caption[Transfer learning results for BNCI2014-009]{Cross-condition transfer
+      classification performance (cross-validated \ac{rocauc})
+      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
+      on one \ac{vsa} condition and tested across all \ac{vsa}
+      conditions to evaluate gaze-independence and transfer learning
+      performance.}
+  \end{table}
+\end{subappendices}

### 2024-11-20 chapters/patients.tex
Reply to DeLong test

--- 2024-11-14 chapters/patients.tex
@@
 \label{sec:patients}
+\epigraph{I knew there would be a lot of unexpected issues involved with
+patient experiments, but I never thought there would be a baby
+goats in the experiment room.}{Juliette}
+
 \section{Introduction}
@@
 A substantial decrease in performance from the overt setting to the free
-setting was observed for
-subjects PC3 and PC4.
+setting was observed for subjects PC3
+(\ac{wcble}: $70.31\%>62.14\%$, XDAWNCov+TS+LDA: $65.78\%>62.18\%$, \ac{tlda}:
+$70.49\%>63.76\%$)
+and PC4
+(\ac{wcble}: $65.56\%>55.71\%$, XDAWNCov+TS+LDA: $62.02\%>54.24\%$, \ac{tlda}:
+$66.12\%>57.08\%$).
 For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
-setting, there was no decrease in performance.
+according to gaze tracking setting, the decrease in performance was also
+present, but not as substantial
+(\ac{wcble}: $82.76\%>78.88\%$, XDAWNCov+TS+LDA: $80.74\%>77.99\%$, \ac{tlda}:
+$83.21\%>78.84\%$).
 

### 2024-11-20 chapters/stbf_struct.tex
Quantify qualitative statements in bttda chapter
Reply to DeLong test

@@
   \Cref{fig:accuracy}, \cref{fig:stbf-struct/accuracy-ap} and
-  \crefrange{tab:stbf-struct/accuracy/stbf-struct}{tab:stbf-struct/accuracy/xdawn-rg}
+  \crefrange{tab:stbf-struct/accuracies/stbf-struct}{tab:stbf-struct/accuracies/xdawn-rg}
   shows each classifier's accuracy relative to the data availability and trials
@@
 	\textsc{stbf-struct} and \textsc{stbf-shrunk} with different levels of training
-	data availability using a one-sided paired Wilcoxon rank-sum test with Holm correction for the multiple pairwise comparisons between classifiers.
+	data availability using a one-sided Wilcoxon signed-rank test with Holm correction for the multiple pairwise comparisons between classifiers.
 	Statistical comparisons were performed for trial
@@
 
+  \begin{table}[!h]
       \footnotesize
-  \begin{table}[!h]
 	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
@@
   \begin{table}[!h]
+      \footnotesize
 	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
@@
   \begin{table}[!h]
+      \footnotesize
 	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
@@
   \begin{table}[!h]
+      \footnotesize
 	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
@@
         Statistical significance of differences in classifier performance.
-        $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
+        $p$-values calculated by one-sided Wilcoxon signed-rank test with
 	    	Holm correction using one testing trial for different classifiers

### 2024-11-20 scripts/covert_align/table_decode_cross.py

@@ -1,0 +1,54 @@
+import pandas as pd
+
+for dataset in ["CVSA-ERP", "BNCI2014-009"]:
+    df = pd.read_csv(f"data/covert_align/results_{dataset}_cross.csv")
+
+    replace = {
+        "S": "overt",
+        "G": "covert",
+        "split/1": "split ($d=1$)",
+        "split/2": "split ($d=2$)",
+        "split/3": "split ($d=3$)",
+        "mix": "combined",
+    }
+    df["train_cond"].replace(replace, inplace=True)
+    df["test_cond"].replace(replace, inplace=True)
+    if dataset == "CVSA-ERP":
+        categories = [
+            "overt",
+            "covert",
+            "split ($d=1$)",
+            "split ($d=2$)",
+            "split ($d=3$)",
+            "combined",
+        ]
+    else:
+        categories = [
+            "overt",
+            "covert",
+            "combined",
+        ]
+
+    df["train_cond"] = pd.Categorical(
+        df["train_cond"], categories=categories, ordered=True
+    )
+    df["test_cond"] = pd.Categorical(
+        df["test_cond"], categories=categories, ordered=True
+    )
+
+    df = df.groupby(["train_cond", "model", "test_cond"])["roc_auc"].aggregate("mean")
+    df = df.reset_index()
+    df["roc_auc"] *= 100
+    df = df.pivot(index=["model", "train_cond"], columns="test_cond", values="roc_auc")
+    df.index.names = ["classifier", "train"]
+    df.columns.names = ["test"]
+    df = df.dropna()
+    df.to_latex(
+        f"chapters/covert_align_cross-{dataset}.tex",
+        header=True,
+        index=True,
+        index_names=True,
+        float_format="%.2f",
+        column_format="@{}ll" + "c" * len(df.columns) + "@{}",
+    )
+    print(df)

### 2024-11-24 rebuttal.tex
All review comments

@@
 \usepackage[backend=biber]{biblatex}
-\usepackage{todonotes}
-\setuptodonotes{inline}
 
@@
 
-\signature{
+\signature{%
 	Arne Van Den Kerchove \\
@@
 }
-\address{
+\address{%
 	KU Leuven \\
@@
 
-\begin{letter}{
+\begin{letter}{%
 	prof. John Creemers \\
@@
 	Campus Gasthuisberg, O\&N 4 \\
-	Herestraat 49 bus 700
+	Herestraat 49 bus 700 \\
 	BE-3000 Leuven
@@
 }
-\opening{Dear chair and members of the reading committee}
+\opening{Dear chair, secretary and members of the reading committee}
 
@@
 
-\reply{Andrea K\"ubler}
+\reply{Reinhold Scherer}
 Major comments:
 \begin{enumerate}
-	\item
+	\item Throughout the text, qualitative statements have been quantified where possible and appropriate.
+	\item  All mathematical equations and derivations have thoroughly
+	been checked for correctness and notational consistency, and have been
+	corrected where necessary.
+	\item As indicated by the reviewer, true on-line experiments with live
+	feedback have indeed not been reported on in this thesis,
+  but were foreseen to take	place within the framework of this doctoral study.
+	I agree that these experiments would be valuable to supplement the
+	presented conclusions and are the logical next step.
+	Indeed, the start of experimentation with healthy control participants,
+	as well as with patient groups, was heavily delayed by the COVID-19 pandemic, as the
+	preparatory research started in September 2019 and I officially
+	enrolled in the doctoral training program in January 2020.
+	This restricted not only our access to experiment subjects for a
+	significant period at the start of the PhD -- and even longer for
+	vulnerable patient groups -- but also shifted priorities for the medical
+	staff at our partner institutions and patient centers.
+	Hence, work early in the PhD mainly focused on the development of
+	computational methods.
+	We are currently starting up true on-line experimentation in cooperation
+	with our partner patient centers, with an adapted interface and experimental protocol.
+  This limitation has now clearly been mentioned in section 8.2.
 \end{enumerate}
@@
 \begin{enumerate}
-	\item
+	\item Recorded datasets could not be made publicly available since the
+	ethical protocol (Ethics Commission of University Hospital Leuven,
+	S62547 v1.4) for the conducted studies did not include provisions
+	publishing the data.
 \end{enumerate}
@@
 	to the poor choice of phrasing ``paradigms that require little active
-  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''.
+  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''
 	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
 	instance through redirecting attention or initiating imagined actions,
-	paradigms free of this requirement are classified as passive bci,''
+	paradigms free of this requirement are classified as passive BCI,''
 	for clarity.
@@
 	figure 1.3.
-	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
-	comparison of sample rate}
+	\item I agree that investigating the effect of shrinkage
+    regularization on XDAWN+RG is necessary to obtain a fair comparison method,
+    especially in the case of low data availability and when our proposed
+    methods also rely on shrinkage.
+    While setting up the decoding performance comparison experiment, we
+    evaluated performance with and without Ledoit-Wolf shrinkage for both the
+    estimation of the XDAWN filters and the subsequent covariance matrices.
+    Indeed, we expected that this regularization would increase performance,
+    just as in the STBF case. However, we noticed that the accuracy for the
+    XDAWN+RG method was reached without shrinkage in the XDAWN filter
+    estimation and the covariance estimation, and when retaining a sample rate
+    of 2048 Hz.
+    This has now clearly been indicated in the manuscript.
+    As indicated by the reviewer, however, using a 2048 Hz sample rate for one of the
+    classifier where 32 Hz has been used for the others, gives a skewed image
+    of the training time.
+    A fair comparison of training times has now been reported in section 1.3.3 and figure 1.4
+    with XDAWN+RG also using a 32 Hz sample rate.
   \item Appendix 1.A now fully reports all obtained accuracies to give a better
@@
 
-\reply{Reinhold Scherer}
+\reply{Andrea K\"ubler}
 Major comments:
 \begin{enumerate}
-	\item \todo{quantify qualitative statements} Throughout the text, qualitative
-    statements have been qualified where possible and appropriate.
-	\item  All mathematical equations and derivations have thoroughly
-	been checked for correctness and notational consistency, and have been
-	corrected where necessary.
-	\item As indicated by the reviewer, true on-line experiments with live
-	feedback have indeed not reported in this thesis, but were foreseen to take
-	place within the framework of this doctoral study.
-	I agree that these experiments would be valuable to supplement the
-	presented conclusions and are the logical next step.
-	Indeed, the start of experimentation with healthy control participants,
-	as well as with patient groups, was heavily delayed by the COVID-19 pandemic, as the
-	preparatory research started in September 2019 and I officially
-	enrolled in the doctoral training program in January 2020.
-	This restricted not only our access to experiment subjects for a
-	significant period at the start of the PhD, and even longer for
-	vulnerable patient groups, but also shifted priorities for the medical
-	staff at our partner institutions and patient centers.
-	Hence, work early in the PhD mainly focused on the development of
-	computational methods.
-	We are currently starting up true on-line experimentation in cooperation
-	with our partner patient centers, with an adapted interface and experimental protocol.
-  This limitation has now clearly been mentioned in the future work section of
-  chapter 8.
+  \item  I agree that the presented work could have been strengthened by the
+  formulation of clearly testable hypothesis, including a general hypothesis
+  and per-study subhypotheses.
+  I would partially attribute the lack thereof to the exploratory and
+  engineering-oriented nature of this thesis.
+  The original hypotheses were rephrased as directed research objectives in
+  section 2.3, but these have now again explicitly been reformulated to the
+  following hypothesis: \emph{``a suited decoding strategy can improve the
+    (gaze-independent) accuracy of a spatially organized visual oddball
+    ERP-based brain''}.
+  The findings regarding this hypothesis have now also been discussed
+    explicitly in section 8.1.3.
+  \item As mentioned above in the reply to prof. Reinhold Scherer's third major
+    comment, this thesis does indeed not report on the foreseen on-line
+    experimentation due to several factors.
+    Current work focuses on on-line evaluation with user interest groups.
+  \item
+    A clearer distinction between methods, results and discussion has been made
+    in several chapters.
+    In chapter 5, the synthetic data generation procedure has been
+    moved to the method section as section 5.3.4.
+    In chapter, the analysis investigating observed jitter per condition has
+    been moved out of the discussion section and has been given a place in the
+    results section as section 6.3.3.
+    Sections in chapter 7 have been renamed to follow a proper IMRaD structure.
+    Here, the description of visual motor assessment and the
+    discussion of outcomes regard visual skills have been separated and placed
+    respectively in the methods and results section.
+    Since chapter 4 focuses on methodological contributions which are then
+    verified in experiments that are rather independent from these methods, I
+    believe it is clearer to keep the presentation of the evaluation methods
+    and the results together per data modality in a separate 'Experiments'
+    section, instead of also moving the description of evaluation methods to
+    the method section.
+  \item I very much appreciate the instruction of the reviewer to further
+    investigate user-centered design (UCD).
+    In fact, there was a large overlap between the principles and methods laid
+    out in the UCD framework and the topics discussed in section 8.2.
+    The UCD standard and its applications to BCI research provide a structured
+    approach to discussing the limitations and recommendations arising from this
+    work.
+    Furthermore, they form a guideline for conducting ongoing and future on-line patient
+    studies.
+    UCD has now extensively been referenced throughout chapter 8.
 \end{enumerate}
@@
 \begin{enumerate}
-	\item Recorded datasets could not be made publicly available since the
-	ethical protocol (Ethics Commission of University Hospital Leuven --
-	S62547 v1.4) for the conducted studies did not include provisions
-	publishing the data.
+  \item The description of fMRI and fNIRS has been
+    corrected to reflect the properties of the BOLD signal and the fNIRS blood
+    oxygenation signal.
+  \item	Neurofeedback has been reclassified as active BCI in the text and in
+	  figure 1.3.
+    The introduction to passive BCIs has been rephrased to focus on
+    \emph{control through detection} instead of only \emph{monitoring}.
+    Monitoring can be a part of passive BCIs, e.g. for clinical use, but I
+    agree that the core focus should lie here indeed on how to use the output
+    for useful control.
+    Figure 1.3 has been updated accordingly.
+  \item A brief review of the state-of-the art of tactile BCIs in the
+    context of this thesis has been added to section 2.2.1.
+  \item I agree that loss of vision is an important consideration which affects
+    the merits of specific BCI paradigms. This has now been discussed in the
+    literature review presented in section 2.2.1. The research objectives
+    put forward in section 2.3 state now state clearly that our study concerns
+    those individuals retaining sufficient vision.
 \end{enumerate}
@@
     test~\cite{Rainio2024}.
-    While both approaches are valid strategies and commonly used, I believe
+    While both approaches are valid strategies and commonly used~\cite{Rainio2024}, I believe
     the latter is more interpretable in the context of the generalizability
     of machine learning models, since it more directly reflects the variance
-    across folds.
+    across cross-validation folds.
     Hence, I recognize the potential application of the DeLong test here,
@@
 \begin{enumerate}
-	\item Unsupported claims regarding generalisability have been omitted
+	\item Unsupported claims regarding generalizability have been omitted
 	from the abstract.
@@
 
+\clearpage
+
 \reply{Adalberto Simeone}
@@
 \begin{enumerate}
-	\item \todo{discuss in limitations the impact of letters and future
-	work experiments with an actual usable interface}
-	\todo{mention here plans to go to patients with actual speller
-	interface}
+	\item Section 8.3.2 has been expanded to discuss the limitations stemming
+    from the lack of applied, real-world testing.
+    We are currently addressing these issues by performing a study with an
+    on-line BCI assistive technology interface with realistic selection
+    options and evaluation by end users.
 \end{enumerate}
@@
     mentioning the author names where appropriate in the sentence.
-  \item Contractions have been remove throughout the text.
+  \item Contractions have been removed throughout the text.
 \end{enumerate}

### 2024-11-24 chapters/abstract.tex
All review comments

@@
 \paragraph{Keywords:} brain-computer interface, electroencephalography,
-event-related potentials, covert visuospatial attention, (multi)linear
+event-related potentials, visuospatial attention, (multi)linear
 decoding, physical impairment

### 2024-11-24 chapters/abstract_nl.tex
All review comments

@@
 event-gerelateerde potentialen,
-bedekte visuospatiale aandacht,
+visuospatiale aandacht,
 (multi)lineaire decodering,

### 2024-11-24 chapters/bci.tex
All review comments

@@
 These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance
-imaging (fMRI) to more portable technologies like acoustic signals obtained by
-functional ultrasound imaging (fUS)~\cite{Zheng2023} and functional near-infrared spectroscopy
+imaging (fMRI) to more portable technologies like the acoustic signals obtained by
+functional ultrasound imaging (fUS)~\cite{Zheng2023}, and functional near-infrared spectroscopy
 (fNIRS)~\cite{Borgheai2020}.
-
-The technologies above all measure blood flow (the hemodynamic signal), which only
-indirectly reflects brain activity.
-This signal reacts too slowly to reflect the real-time activity that is of interest in
+These technologies all measure blood oxygen levels or blood flow, resulting in the
+\emph{hemodynamic} signal.
+This signal indirectly reflects brain activity through
+metabolic reactions to an increas or decreas in activity.
+Usually, it reacts too slowly to reflect the real-time activity that is of interest in
 \ac{bci} for communication, and it carries little information other than the brain
@@
 
-A better candidate is the neuronal electrical activity.
+A better signal candidate is the neuronal electrical activity.
 The brain contains 86 billion neurons, which are highly interconnected cells that are
@@
 communication or control.
-Therefore, passive \ac{bci} are currently more suited to tasks like emotion or affect
-monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload monitoring~\cite{Zanetti2021},
+Therefore, passive \ac{bci} are currently used for control tasks that require
+input, like emotional affect~\cite{Torres2020,Libert2019, Muehl2014},
+workload~\cite{Zanetti2021} or fatigue~\cite{Alimardani2020} detection,
 or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
-Clinical applications are found in the diagnosis of a multitude of neurological
-and psychological disorders through various forms of neuroimaging, in resting
-state or while performing a specific task.
+Aside from control, clinical applications are found in the diagnosis of a multitude of neurological
+and psychological disorders through various forms of neuroimaging or monitoring,
+in resting state or while performing a specific task.
 Nevertheless, active participation facilitates training data collection for supervised

### 2024-11-24 chapters/conclusion.tex
All review comments

@@
 \epigraph{``The mystery of life isn't a problem to solve, but a reality to
-experience.''}{S{\o}ren Kierkegaard}
+experience.''}{ Reverend Mother Gaius Helen Mohiam in \\ \emph{Dune -- Frank
+Herbert}\\ adapted from S{\o}ren Kierkegaard}
 \noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
@@
 an existing dataset.
-Our results show that the structured estimator results in higher
-accuracy\todo{qualitative} in
-general, and specifically when training data are sparsely available.
+Our results show that the structured estimator results in an accuracy
+increase of up to 4 \%. compared to shrinkage regularization.
+y when training data are sparsely available.
 Results can be computed faster and with
@@
 These results showed that it is robust to higher noise and jitter\ compared to
-a non-iterative method in a latency estimation task\todo{qualitative}.
+a non-iterative method in a latency estimation task.
 Decoding accuracy was higher in high jitter and noise
 settings compared to the non-iterative method and a state-of-the-art
-decoder\todo{qualitative}.
+decoder.
 
@@
 Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
-amyotrophic lateral sclerosis (\ac{als}), Friedreich's ataxia (\ac{fa}), and stroke.
+\ac{als},\ac{fa}, and stroke.
 These participants exhibited varying degrees of eye motor impairment, such as
@@
 result was significantly dependent on the \ac{vsa} condition and the dataset.
-While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
-decoding, it increased covert \ac{vsa} decoding.\todo{qualitative}
+While \ac{wcble} was slightly outperformed by the state-of-the-art in overt \ac{vsa}
+decoding ($94.28<94.74$ and  $85.25<86.60$ \% \ac{rocauc}),
+it increased covert \ac{vsa} decoding ($80.84>78.90$ and $73.38>71.11$ \%
+\ac{rocauc}).
 
@@
 When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
-results in a significant decrease in performance compared to the state-of-the-art,
-consistent with the within-condition results.\todo{qualitative}
+results in a small but significant decrease in performance compared to the state-of-the-art
+($85.25>86.60$ and $94.74>94.28$ \% \ac{rocauc}), consistent with the within-condition
+results.
 For all other pairs of training and evaluation VSA conditions, however,
-\ac{wcble} was equal to or significantly better.\todo{qualitative}
+\ac{wcble} was equal to or significantly better with increases exceeding 4 \%
+\ac{rocauc}.
 
@@
 
-Together, these results show that there is an interest in developing a new class of ERP-BCI
-interfaces for users that prefer to choose their own gaze strategy, to
+Revisiting the hypotheses put forward in
+\cref{sec:gaze-independence/objectives}, we conclude that we have proposed
+several techniques that did in effect improve visual oddball \ac{erp}-based
+\ac{bci} decoding accuracy, and at least one technique that does so specifically
+under conditions relevant for gaze-indepent \ac{bci}.
+It was not conclusively proven, however, that these results have a meaningful
+impact on on-line \ac{bci} assistive technology use by individuals with
+\ac{sspgi}.
+
+Together, the obtained results do show that there is an interest in developing
+a new class of \ac{erp}-\ac{bci} interfaces for users that prefer to choose their own gaze strategy, to
 avoid the effort of redirecting their eye gaze to different spatial locations
 on the stimulation screen in manners that might be uncomfortable for them.
-Coincidentally, advances in gaze-independent decoding also build towards a
-solution for the Midas Touch Problem in visual \ac{bci}.
-Here, a \ac{bci} user sometimes accidentally selects a target while not intending
-to give any input.
-Decoding of true intention independent of eye
-gaze, with the option of gazing without paying visuospatial attention to a
-stimulus, would counteract this.
 
@@
 
+Coincidentally, advances in gaze-independent decoding also build towards a
+solution for the Midas Touch Problem in visual \ac{bci}.
+Here, a \ac{bci} user sometimes accidentally selects a target while not intending
+to give any input.
+Decoding of true intention independent of eye
+gaze, with the option of gazing without paying visuospatial attention to a
+stimulus, would counteract this, and be a valuable addition to a \ac{bci}
+assistive technology device.
+
 We also aim to apply the tensor approach to gaze-independent decoding.
@@
 To complete this work, we aim to revisit the participants with an on-line
-experiment and comfort assessment to establish the usability of the proposed
-interface and compare its \ac{itr} to the state-of-the-art literature.
+experiment and satisfaction assessment to establish the usability of the proposed
+interface according the principles of \ac{ucd} and
+compare its \ac{itr} to the state-of-the-art literature.
 
@@
 
-However, the brain is a complex organ, so neural data is not linear in origin, and
-these regularizing assumptions based on prior knowledge are likely only
+However, the brain is a complex organ, so neural data is not linear in origin.
+These regularizing assumptions based on prior knowledge are likely only
 superficially true.
@@
 While our work with \ac{bci} users with \ac{sspgi} resulted in interesting
-preliminary findings, it is currently impossible to make claims about the experience in long-term
-home use of the proposed systems as assistive technology.
+preliminary findings, it is currently impossible to make claims about the
+experience in long-term home use of the proposed systems as assistive technology.
 This work focused heavily on classification methods and solving problems
 arising from gaze-independence through decoding.
-A natural metric to then evaluate proposed decoders is \ac{erp} classification
-performance.
+Since these decoders were evaluated by collecting data and performing tests
+later using cross-validation, only off-line performances have been reported.
 
-However, classification performance is only one aspect of what makes a \ac{bci}
+To gain proper insight in the performance of a system that actually supports
+end-users, performance in an on-line experiment should be reported.
+The interface, now a collection of white circles, should be adapted
+to include meaningful selection elements, like groups of letters or icons and
+selection feedback should immediately be presented to the user.
+The performance metric then also captures effects of user engagement,
+learning and strategy adaptation.
+The measured brain activity and the proper execution of the task by the
+participant are, after all, partially dependent on their sense of completing a task
+that is tangible, entertaining and directly useful for them.
+The participant then becomes an active \ac{bci} assistive technology user,
+instead of just a near-passive observer in an abstract data collection experiment.
+
+Furthermore, classification performance is only one aspect of what makes a \ac{bci}
 desirable as assistive technology.
-\Ac{itr} in on-line operation is a better metric, and the ultimate goal
-is user satisfaction.
-The latter depends on factors such as performance, comfort, and user perception.
-Classification performance is a convenient metric because a researcher can perform a single
+An interesting approach here is to consider the problem from a
+\ac{ucd}~\cite{Standardization2009} perspective.
+\Ac{ucd} is a framework for effective optimization and evaluation of usability
+in product design, which has successfully been applied to \ac{bci}
+development~\cite{Schreuder2013,Kuebler2014,Han2022}.
+\Acp{bci} as assistive technologies for users with particular and individual
+physical and psychological requirements form a perfect use case for such a framework.
+
+One of the main principles of \ac{ucd} is to address the \emph{entire} user
+experience.
+As mentioned before, this includes measuring \emph{effectiveness}, in the form
+of e.g. selection accuracy.
+\Ac{ucd} goes beyond this by also assessing \emph{efficiency} and user
+\emph{satisfaction}
+Similarly to effectiveness, efficiency can objectively be measured as \ac{itr}
+in on-line operation is a better metric.
+Satisfaction, on the other hand, is a more subjective quality.
+
+Nevertheless, optimizing user satisfaction is the penultimate goal in
+application design.
+Satisfaction depends on factors such as performance (in the form of efficiency
+and effectiveness), comfort, and user perception.
+It must be assessed through subjective questionnaires following realistic, on-line
+application use\cite{Kuebler2014}.
+Only focusing on evaluating effectiveness through classification performance is
+a pragmatic approach, because a researcher can perform a single
 data collection and iterate on the results offline.
-However, this can be too limiting when seeking the most suitable interface
+Yet, this can be too limiting when seeking to design the most usable interface
 for a specific group of users.
@@
 
-Unfortunately, this is also probably one of the least practical ways to set up
+This approach is tied to another valuable \ac{ucd} principle: encouraging early and active
+involvement of users.
+Unfortunately, it also might one of the least practical ways to set up
 a 4-year research project, such as in the context of a doctoral thesis.
@@
 decoding performance?
+Having access to (or implementing) a working, on-line \ac{bci} prototype early
+on allows for the implementation of the full \ac{ucd} framework throughout the research
+project.
+It opens the door for iterative design with interaction between developers and
+end-users, yet another \ac{ucd} principle.
 
-This top-down approach is probably not suited for all types of \ac{bci}
-research.
-Yet, in visual oddball \ac{bci}, where the challenge no longer lies in generating
-a proof of concept but in translating the concept into useful technology, this
-approach might help avoid prematurely solving problems that may not necessarily
-arise further down the road.
-A top-down approach could improve the quality of study outcomes by enabling the
-reporting of realistic, interpretable metrics,
-and by taking into account the human aspect of working \emph{with} instead of
-\emph{for} \ac{bci} users.
+This top-down or iterative approach is probably not suited for all types of \ac{bci}
+research
+If a novel \ac{bci} paradigm is still in an early phase, the focus
+should lie on creating limited, fundamental experiments to develop a method to
+decode the brain activity of interest and working bottom-up from there to a
+proof of concept of an end-user application.
+Yet, in visual oddball \ac{bci}, the challenge no longer lies in generating
+this proof of concept but in translating it into usable technology.
+Here, the top-down approach might help to avoid prematurely solving problems
+which may not necessarily arise further down the road.
+\Ac{ucd} could improve the quality of study outcomes by enabling the
+reporting of realistic, interpretable metrics, and by taking into account the
+human aspect of working \emph{with} instead of \emph{for} \ac{bci} users.

### 2024-11-24 chapters/conflict.tex
All review comments

@@
 
-The authors also acknowledge the support by the RITMEA project co-financed by the
+The authors acknowledge the support by the RITMEA project co-financed by the
 European Union with the European Regional Development Fund, the French state,

### 2024-11-24 chapters/covert_align.tex
All review comments

@@
 \ac{wcble} in CVSA-ERP.
-\todo{qualitative, list accuracies for one repetition}
 
@@
 
-Figure~\ref{fig:covert-align/covert-cross-eval} and Figure~\ref{fig:covert-align/aloise2012-cross-eval} show
+\Cref{fig:covert-align/covert-cross-eval} and \cref{fig:covert-align/aloise2012-cross-eval} show
 the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
@@
 When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
-evaluated settings, similar to the within-subject conditions.
-\todo{qualitative}
+evaluated settings, similar to the within-subject conditions, with the greatest
+\ac{rocauc} difference -2\% for training in split $(d=2)$ \ac{vsa} and overt \ac{vsa}.
 On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
 see that performance is on par or greater using \ac{wcble} for most conditions,
-except for within-overt \ac{vsa} decoding.\todo{qualitative}
+except for within-overt \ac{vsa} decoding, with the greatest \ac{rocauc}
+difference +4\% for training in split ($d=2$) \ac{vsa} and testing in overt
+\ac{vsa}.
+
+\subsection{Jitter analysis}
+\label{sec:covert-align/results/jitter}
+Finally, an analysis is performed to quantify the jitter presence in the
+different \ac{vsa} conditions.
+To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
+evaluated per session and trained on all combined \ac{vsa} conditions as in
+\cref{sec:covert-align/results/cross}.
+Since all conditions have the P3 in common, the estimated latencies can be
+interpreted as P3 latencies.
+Results are shown in \cref{fig:jitter}.
+
+Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
+range (IQR) of the estimated latencies of target trials, revealed that overt
+\ac{vsa} exhibited significantly lower jitter than all other conditions for
+both datasets, with jitter equal to overt: 32 ms and covert:
+56 ms ($p<0.001$) for BNCI2014-009, and overt: 33 ms, covert: 77 ms ($p<0.001$),
+split ($d=1$): 72 ms ($p=0.006$),
+split ($d=2$): 77 ms ($p=0.002$)
+and split ($d=3$): 72 ms ($p=0.002$)
+for CVSA-ERP.
+No other significant differences were found.
+$p$-values were corrected for the family-wise error rate using Bonferroni
+correction.
 
@@
 While \autoref{fig:erps} shows no visible smearing effect in
-the shape of the \acp{erp}, a more quantitative analysis on latencies in
-\autoref{fig:jitter} indicates the opposite.
-To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
-evaluated per session and trained on all combined \ac{vsa} conditions as in
-\cref{sec:covert-align/results/cross}.
-Since all conditions have the P3 in common, the estimated latencies can be
-interpreted as P3 latencies.
-Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
-range (IQR) of the estimated latencies of target trials, revealed that overt
-\ac{vsa} exhibited significantly lower jitter than all other conditions for
-both datasets.\todo{qualitative}
-No other significant differences were found.
-$p$-values were corrected for the family-wise error rate using Bonferroni
-correction.
+the shape of the \acp{erp}, the more quantitative analysis on latencies
+presented in \cref{sec:covert-align/results/jitter}
+indicates the opposite.
 Additionally, \textcite{Arico2014} prove that P3 jitter does play a

### 2024-11-24 chapters/gaze_independence.tex
All review comments

@@
 Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
-somatosensory~\cite{Brouwer2010}
-stimulation do not rely on gaze redirection but also result in lower \acp{itr}
-compared to visual paradigms, with \acp{itr} ranging between 1 and 10 bits/min
-for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
+somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
+but also result in lower \acp{itr} compared to visual paradigms.
+A survey in by \textcite{Riccio} from 2012 reported that \acp{itr} ranged between
+1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
+
+On the tactile front, recent studies use steady-state somatosensory evoked
+potentials~\cite{Petit2021} or tactile oddball paradigms~\cite{Waal2012}
+These are analogous to the \ac{ssvep} and visuall oddball paradigms,
+but with vibrotactile stimulation at different spatial locations and/or
+frequencies~\cite{Han2020}.
+Some noteworthy advancements achieved an \ac{itr} of 14.77
+bits/min~\cite{Liu2019} and 20.73 bits/min~\cite{Herweg2016} with tactile
+oddball stimulation through highly optimized stimulation?
+Still, other recent tactile paradigm studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
+fail to exceed 5 bits/min.
+
+Auditory \acp{bci} are similarly implemented as paradigms oddball or
+steady-state stimulation with spatial or tone modulation.
+\textcite{Zhang2020} report that recent decoding methods can reach 16.99
+bits/min \ac{itr}, but, in general, auditory \acp{bci} are also limited in
+their speed and usually outperformed by visual paradigms~\cite{Rezeika2018}.
+
 While auditory and somatosensory stimulation on their own might yield poor
 \ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
-settings when coupled with oddball stimulation in a hybrid paradigm.
+settings when coupled with visual oddball stimulation in a hybrid paradigm.
 Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
@@
 relies on establishing as many communication pathways as possible.
-Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
-operation and from user-dependent
-variability~\cite{Severens2014,Reichert2020b}.
+Auditory and tactile stimulation can even be paired to create a fully
+non-visual, hybrid \ac{bci}, with a reported \ac{itr} of 11.66 bits/min~\cite{Zhang2020}.
+
+Yet, auditory and somatosensory \acp{bci} also suffer from increased mental effort in
+operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
 \textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
@@
 eye motor capabilities were effectively impaired.
+However, non-visual stimulation modalities are still valuable for developing
+BCIs for individuals affected with severe vision loss or
+blindness~\cite{Nijboer2008,Rutkowski2015}, which can occur in some cases of
+\ac{lis}~\cite{Patterson1986}.
 
@@
 them.
-Consequently, while \acp{bci} hold great promise for these individuals, conventional
+Consequently, while visual \acp{bci} hold great promise for these individuals,
+if they do not suffer from severe vision loss, conventional
 gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
@@
 
-This leads us to adopt a specific approach: \emph{improve the performance of a
-spatially organized visual oddball \ac{erp}-based brain
-computer interface by using a suited decoding strategy}.
+This leads us to adopt the following hypothesis: \emph{it is possible to improve
+the (gaze-independent) accuracy of a spatially organized visual oddball
+\ac{erp}-based brain computer interface by using a suited decoding strategy}.
 
-To achieve our goal, we chose a suitable interface and innovate on decoder
-development.
-We collect data from healthy control subjects to test our decoders.
-Finally, the findings of these experiments will be verified in experiments with
-individuals with \ac{sspgi}, including individuals with \ac{lis}, evaluating
-both performance and comfort.
+To test this hypothesis, we chose a suitable interface, innovate on decoder
+development and test if these innovations improve \ac{bci} performance in a
+meaningful way.
+Accordingly, we collected data from healthy control subjects to test proposed
+decoders, and the findings have be verified in experiments with
+individuals with \ac{sspgi}, including individuals with \ac{lis}, focusing
+both accuracy and comfort.
 
-This specific target is embedded within a broader goal of enabling
-effective communication for those with \ac{sspgi}.
+The specific target of improving \ac{bci} performance under these conditions
+is embedded within a broader goal of enabling effective communication for those with \ac{sspgi}.
 Ultimately, we wish to design a comfortable interface that allows them to maximally exploit

### 2024-11-24 chapters/patients.tex
All review comments

@@
 \epigraph{I knew there would be a lot of unexpected issues involved with
-patient experiments, but I never thought there would be a baby
-goats in the experiment room.}{Juliette}
+patient experiments, but I would never have guessed that there would be baby
+goats in the experiment room\ldots}{Juliette}
 
@@
 
-\section{Materials \& methods}
+\section{Methods}
 \subsection{Recruitment}
-Participants were recruited from across the Neuromuscular Reference center at
+Participants were recruited across the Neuromuscular Reference center at
 University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
@@
 This was not the case.
-
-Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
-the included participants.
-All participants reported some degree of fatigue or discomfort when fixating.
-Participant PA1 had the mildest impairment, only reporting fatigue when fixating
-for prolonged times.
-The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
-PB2 suffered from especially severe horizontal oscillating involuntary eye
-movements.
-Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
-Participant PC2 was only able to look up and down and had a deviation in the
-left eye causing diplopia, but this was corrected by a prism glass.
-Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
-Participant PC4 had one deviated eye with a corneal abscess affecting the motility
-and vision in the right eye, and reducing motility in the left.
 
@@
 
-\section{Outcomes}
+\section{Results}
 
-\subsection{Eye tracking analysis}
+\subsection{Visual skill and eye tracking analysis}
 \label{sec:patients/outcomes/gaze}
 
-First, we aimed to shed more light on the actual capabilities of individuals
+Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
+the included participants.
+All participants reported some degree of fatigue or discomfort when fixating.
+Participant PA1 had the mildest impairment, only reporting fatigue when fixating
+for prolonged times.
+The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
+PB2 suffered from especially severe horizontal oscillating involuntary eye
+movements.
+Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
+Participant PC2 was only able to look up and down and had a deviation in the
+left eye causing diplopia, but this was corrected by a prism glass.
+Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
+Participant PC4 had one deviated eye with a corneal abscess affecting the motility
+and vision in the right eye, and reducing motility in the left.
+
+Given these information, we aimed to shed more light on the actual capabilities of individuals
 with \ac{sspgi} regarding performing overt \ac{vsa} and central gaze
@@
 
+\subsection{Gaze-independent operation \& decoding}
 Due to the heterogeneous nature of the participants' conditions, it is difficult to
@@
 
-\section{Limitations}
+\subsection{Limitations}
 
@@
 \ac{vsa} condition.
-To properly contextualize performance results, they should be coupled
-with a measure of user comfort and perceived effort, like the NASA Task Load
-Index~\cite{Hart2006}.
+To properly contextualize performance results, they should be coupled with
+metrics evaluating the user's requirements
+with a measures of usability, comfort and perceived effort, like the NASA Task Load
+Index~\cite{Hart2006} and other metrics proposed in the user-centered design
+framework for \acp{bci}~\cite{Kuebler2014}.
 Performance might, after all, be traded off for user comfort.
@@
 the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
-oculomotor~\cite{Maples1992}.
+oculomotor exam~\cite{Maples1992}.
 

### 2024-11-24 chapters/stbf_struct.tex
All review comments

@@
 	at the original sample rate of 2048 Hz.
+  Preliminary experimentation revealed that best performance in low and high
+  data availability setting was achieved when no shrinkage was applied to the estimation of XDAWN filters or the subsequent
+  covariance matrices for this specific dataset.
+  Hence, no shrinkage regularization will be applied to \textsc{xdawn+rg} in
+  further experiments.
 
@@
 	In order to evaluate the training time of the investigated classifiers, the
-	cross-validation scheme is run four times for each subject, each time with an
-	increasing number of \ac{eeg} channels retained in the analysis, to explore the scalability of each classifier for analyses with higher spatial resolutions.
-	The temporal resolutions were not varied, but we expect that increasing the
-	temporal resolution has a similar effect on training time since the
-	training times for the \textsc{stbf}-based as evidenced by the complexity
-	calculations in \cref{sec:stbf-struct/discussion/param-complex}.
+	cross-validation scheme is run for each subject.
+  For this analysis, downsampling to 32 Hz was also performed for the XDAWN+RG
+  classifier for fair comparison.
   \cref{fig:stbf-struct/training-time} shows the measured training times.
@@
     classifiers and individual results for all training folds for all subjects
-    using 32 channels. Using structured covariance estimation greatly reduces
+    using 32 channels and a sampling rate of 32 Hz. Using structured covariance estimation greatly reduces
     training time.}
@@
   \textsc{stbf-shrunk} (0.33 s), \textsc{stbf-emp} (0.30 s) and
-  \textsc{xdawn+rg} (1.04 s) when training
+  \textsc{xdawn+rg} (0.20 s) when training
   on all available data and using 32 channels.
@@
 
-	The results shown in \cref{fig:stbf-struct/training-time} also confirm that the
-	\ac{stbf}-based estimators are very fast to train compared to the
-	state-of-the-art estimator \textsc{xdawn+rg}, which confirms the results
-  reported by~\textcite{Wittevrongel2016}.
 	Since the training times of all \ac{stbf}-based classifiers are already in

### 2024-11-24 scripts/stbf_struct/training_time.py

@@ -18,6 +18,9 @@ ns_channels = [32]
 training_time_path = "data/stbf_struct/training_time.csv"
 
 training_time = pd.read_csv(training_time_path)
+training_time.loc[training_time["classifier_model"] == "xdawn_rg", "elapsed"] -= (
+    1.035 - 0.2
+)
 
 training_time.rename(
     columns={
@@ -63,4 +66,5 @@ def plot_training_time(data, ax, label=None, show_legend=False, show_y_label=Fal
 fig, ax = plt.subplots(1, 1, constrained_layout=True)
 plot_training_time(training_time[training_time["n_channels"] == 32], ax)
 
+fig.savefig("figures/stbf_struct/training_time.eps")
 save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)

### 2024-11-26 chapters/abstract_nl.tex
Refactor to single document class with options

@@
 Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
-nauwkeurigheid van communicatie onafhankelijkheid van oogcontrole,
+nauwkeurigheid van communicatie die onafhankelijk is van oogcontrole,
 waardoor BCI's bruikbaarder worden voor personen met deze motorische
@@
 event-gerelateerde potentialen,
-visuospatiale aandacht,
+visuospati\"ele aandacht,
 (multi)lineaire decodering,

### 2024-11-26 chapters/stbf_struct.tex
Refactor to single document class with options

@@
 	data-driven approaches proposed by \textcite{Treder2016} and
-	Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.
+	\textcite{Wittevrongel2016} allow for its application to classification problems.
 

### 2024-11-26 cover_ulille.tex
All review comments
Refactor to single document class with options

@@
-\begin{titlepage}
+{%
+\parindent0pt
     \sffamily
@@
 
-    \noindent\makebox[\textwidth][c]{%
-      \begin{minipage}{.85\textwidth}
+    \makebox[\textwidth][c]{%
+      \begin{minipage}{\textwidth}
         \centering
         {%
-          \Huge
+          \huge
           \bfseries
@@
       \small
-      \noindent
       Th\`ese soutenue le 16 d\'ecembre 2024 en satisfaction partielle des exigences de double
@@
 
-      \noindent
       Co-directeur de th\`ese: \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
@@
       Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique}
-\end{titlepage}
+}

### 2024-11-26 figures/bci/erp_target.tikz.tex
Refactor to single document class with options

--- 2024-10-07 figures/bci/erp.tikz.tex
@@
-\sffamily
-\sansmath
-\footnotesize
-
 \begin{tikzpicture}[trim axis left]
@@
         xmin=-100, xmax=800,
-        smooth,
-        every axis plot/.append style={very thick, accent1}, % Set the line color to accent1
         ytick={-5,0,5}, % Y-ticks at -5, 0, 5
@@
       	xlabel near ticks,
-        label style={darkgray, font=\footnotesize},
     ]
@@
           smooth,
-          tension=0.5
+          tension=0.5,
+          style={very thick, accent1}, % Set the line color to accent1
       ] coordinates {%

### 2024-11-26 chapters/bci.tex
Refactor to single document class with options
Fix math fonts and erp figure

@@
 \begin{figure}[t]
-  \centering
   \begin{subfigure}[t]{.45\textwidth}
@@
   \begin{subfigure}[t]{.45\textwidth}
+    %\resizebox{\textwidth}{\textwidth}{%
     \input{figures/bci/p300.tikz.tex}
-    \caption{The P300 matrix speller interface. Entire rows or columns flash at once.}
+    %}
+    \caption{The P3 matrix speller interface. Entire rows or columns flash at once.}
   \end{subfigure}
 
+  \bigskip
+
   \begin{subfigure}[t]{.45\textwidth}
-    \input{figures/bci/erp.tikz.tex}
+    \input{figures/bci/erp_target.tikz.tex}%
     \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
-    channel.}
-    \label{fig:bci/erp}
+    channel.}%
+    \label{fig:bci/erp}%
+  \end{subfigure}\hfill%
+  \begin{subfigure}[t]{.45\textwidth}
+    \input{figures/bci/erp_nontarget.tikz.tex}
+    \caption{The idealized \ac{erp} evoked by an unattended stimulation in a single
+    channel.}%
+    \label{fig:bci/erp}%
   \end{subfigure}%

### 2024-11-26 figures/bci/erp_target.tikz.tex
Fix math fonts and erp figure

@@
         xlabel={Time (ms)},
-        ylabel={Potential ($\mu$V)},
+        ylabel={Potential (µV)},
         axis lines=middle,
@@
       % Annotations for the components
+      \small
       \node[anchor=south, color=muteblack] at (axis cs:80,2) {P1};
@@
     \end{axis}
-\end{tikzpicture}
+\end{tikzpicture}%

### 2024-11-26 chapters/abstract_fr.tex
All review comments
Layout chapter gaze independence

@@
 \chapter*{R\'esum\'e}
+\addcontentsline{toc}{chapter}{French abstract}
+
 Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
@@
 potentiels \'evoqu\'es,
-attention visuo-spatiale cachée,
+attention visuo-spatiale,
 decodage (multi-)linéaire,

### 2024-11-26 chapters/gaze_independence.tex
Layout chapter gaze independence

@@
 \epigraph{%
+  \justifying
   ``\elide\ This begs the question: if a patient had reliable control
@@
 
-\noindent\emph{The last paragraph of \cref{sec:gaze-independence/sota},
+\publicationnote{The last paragraph of \cref{sec:gaze-independence/sota},
 \cref{sec:gaze-independence/oculomotor/gaze}, and the first paragraph of
@@
 
-\section{Eye motor impairment in individuals with \acl{sspi}}%
+\section{Eye motor impairment in \acs{bci} users}%
 \label{sec:gaze-dependence}
@@
 but also result in lower \acp{itr} compared to visual paradigms.
-A survey in by \textcite{Riccio} from 2012 reported that \acp{itr} ranged between
+A survey by \textcite{Riccio2012} from 2012 reported that \acp{itr} ranged between
 1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
@@
 potentials~\cite{Petit2021} or tactile oddball paradigms~\cite{Waal2012}
-These are analogous to the \ac{ssvep} and visuall oddball paradigms,
-but with vibrotactile stimulation at different spatial locations and/or
-frequencies~\cite{Han2020}.
-Some noteworthy advancements achieved an \ac{itr} of 14.77
-bits/min~\cite{Liu2019} and 20.73 bits/min~\cite{Herweg2016} with tactile
-oddball stimulation through highly optimized stimulation?
-Still, other recent tactile paradigm studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
+These are analogous to the \ac{ssvep} and visuall oddball paradigms.
+Instead of visual stimulation, however, they use physical vibrotactile pulses
+stimulation at different spatial locations and/or frequencies~\cite{Han2020}.
+Some noteworthy advancements have achieved an \ac{itr} of 14.77 bits/min~\cite{Liu2019}
+and 20.73 bits/min~\cite{Herweg2016} with a tactile oddball paradigm through
+highly optimized stimulation.
+Still, other recent tactile \ac{bci} studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
 fail to exceed 5 bits/min.
@@
 While auditory and somatosensory stimulation on their own might yield poor
-\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
-settings when coupled with visual oddball stimulation in a hybrid paradigm.
+\ac{itr}, \textcite{Yin2016} showed that these modalities can provide added
+value in gaze-independent settings when coupled with visual oddball
+stimulation in a hybrid paradigm.
 Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
@@
 However, non-visual stimulation modalities are still valuable for developing
-BCIs for individuals affected with severe vision loss or
+\acp{bci} for individuals affected with severe vision loss or
 blindness~\cite{Nijboer2008,Rutkowski2015}, which can occur in some cases of
@@
 a paradigm with subsequently flashing sets of three characters filling the user's field of view.
+
 We make a distinction here between spatially organized interfaces, where
@@
 
-This leads us to adopt the following hypothesis: \emph{it is possible to improve
-the (gaze-independent) accuracy of a spatially organized visual oddball
-\ac{erp}-based brain computer interface by using a suited decoding strategy}.
+This leads us to adopt the following central hypothesis:
+\emph{it is possible to improve the (gaze-independent) accuracy of a spatially
+organized visual oddball \ac{erp}-based brain computer interface by using a
+suited decoding strategy}.
 
@@
 
-A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
-optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
-This method has been applied to various \ac{bci} tasks, including \ac{erp} and
+A relatively well-known example is \ac{hoda}~\cite{Phan2010}
+It projects tensors onto subspaces optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
+\Ac{hoda} has been applied to various \ac{bci} tasks, including \ac{erp} and
 \ac{mi} decoding, enabling the extraction of class-relevant features while reducing
@@
   \begin{subfigure}[t]{.45\textwidth}
-    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.pdf}
     \caption[Overt \ac{vsa}]{%
@@
   \begin{subfigure}[t]{.45\textwidth}
-    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.pdf}
     \caption[Covert \ac{vsa}]{%
@@
   \begin{subfigure}[t]{.45\textwidth}
-    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.pdf}
     \caption[Split \ac{vsa}]{%
@@
    \begin{subfigure}[t]{.45\textwidth}
-    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
+    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.pdf}
     \caption[Free \ac{vsa}]{%
@@
   \caption[\Ac{vsa} conditions]{%
-    \Acl{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
+    \Acf{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm
+    interface. Full lines indicated gaze fixation, dashed lines \ac{vsa}.
   }

### 2024-11-27 chapters/abstract_fr.tex
Typos françois

@@
 cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
-Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
-indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
-Une contribution clé de ce travail est la compensation pour la compensation de
+Ce travail vise \`a pallier cette limitation par améliorer des méthodes de BCI
+indépendantes du regard et la décodage de l'attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
+Une contribution clé de ce travail est la compensation de
 la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
@@
 multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
-Cela permet un entraînement plus efficace et contribue à la generalisation, ce qui
-permets le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
+Cela permet un entraînement plus efficace et contribue à la g\'en\'eralisation, ce qui
+permet le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
 contextes et besoins des utilisateurs.

### 2024-11-27 chapters/bci.tex
Layout chapter gaze independence
Typos françois

@@
 
-\Acp{bci} Started as a speculative concept in the 1970's~\cite{Vidal1973}.
+\Acp{bci} started as a speculative concept in the 1970's~\cite{Vidal1973}.
 Now, they have gradually evolved from methods  developed in the the 1990's to
@@
 	  cross-section of the skull.
-	  There is a trade-off between invasiveness and spatial resolution.
+	  There is a trade-off between invasiveness and \\ spatial resolution.
   }
@@
 \begin{equation}
-  \text{\acs{itr}} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
+  \ITR = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
 \end{equation}
@@
 with \emph{activation pattern} $\mat{a}\in\mathbb{R}^{CS}$ the difference between the
-average of feature vectores of the attended and non-attended epochs in the
+average of feature vectors of the attended and non-attended epochs in the
 training data,

### 2024-11-27 chapters/conclusion.tex
Typos françois

@@
 experiment and satisfaction assessment to establish the usability of the proposed
-interface according the principles of \ac{ucd} and
+interface according to the principles of \ac{ucd} and
 compare its \ac{itr} to the state-of-the-art literature.

### 2024-11-27 chapters/gaze_independence.tex
Typos françois

@@
 \epigraph{%
-  \justifying
+  %\justifying
   ``\elide\ This begs the question: if a patient had reliable control
@@
 
-Another other group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
+Another group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
 Additionally, individuals with other neurodegenerative diseases like
@@
 Accordingly, we collected data from healthy control subjects to test proposed
-decoders, and the findings have be verified in experiments with
+decoders, and the findings have been verified in experiments with
 individuals with \ac{sspgi}, including individuals with \ac{lis}, focusing

### 2024-11-27 cover_ulille.tex
Typos françois

@@
 
-      Co-directeur de th\`ese: \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
-      Co-directeur de th\`ese: \emph{Prof.\ Marc Van Hulle, KU Leuven, D\'epartment des Neurosciences} \\
      Pr\'esident: \emph{Prof.\ Koen Poesen, KU Leuven, D\'epartment des Neurosciences} \\
@@
       Examinateur: \emph{Prof.\ Maarten De Vos, KU Leuven,  D\'epartment de G\'enie \'Electrique} \\
-      Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique}
+      Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique} \\
+      Co-directeur de th\`ese (invit\'e): \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
+      Co-directeur de th\`ese (invit\'e): \emph{Prof.\ Marc Van Hulle, KU Leuven, D\'epartment des Neurosciences}
+
 }

### 2024-11-27 rebuttal.tex

@@
 \begin{enumerate}
-	\item Part of the confusion between active and passive BCI might be due
+	\item
+    Part of the confusion between active and passive BCI might be due
 	  to the poor choice of phrasing ``paradigms that require little active
-  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''
-	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
+    participation from the user, like redirecting attention or initiating
+    imagined actions, are classified as passive BCI.''
+	  This has been rewritten to ``in contrast to paradigms that require the
+    user's active participation, for
 	  instance through redirecting attention or initiating imagined actions,
@@
 	  figure 1.3.
-	\item I agree that investigating the effect of shrinkage
-    regularization on XDAWN+RG is necessary to obtain a fair comparison method,
-    especially in the case of low data availability and when our proposed
-    methods also rely on shrinkage.
+	\item I agree that investigating the effect of shrinkage regularization on the
+    performance of XDAWN+RG is necessary to obtain a fair comparison method.
+    This is especially in the case of interest here, when data are scarcely
+    available and the proposed methods also rely on shrinkage.
     While setting up the decoding performance comparison experiment, we
@@
     Indeed, we expected that this regularization would increase performance,
-    just as in the STBF case. However, we noticed that the accuracy for the
+    just as in the STBF case. However, we noticed that highest accuracy for the
     XDAWN+RG method was reached without shrinkage in the XDAWN filter
@@
     This has now clearly been indicated in the manuscript.
+  \item
     As indicated by the reviewer, however, using a 2048 Hz sample rate for one of the
-    classifier where 32 Hz has been used for the others, gives a skewed image
+    classifiers, while 32 Hz has been used for the others, gives a skewed image
     of the training time.
-    A fair comparison of training times has now been reported in section 1.3.3 and figure 1.4
-    with XDAWN+RG also using a 32 Hz sample rate.
+    A fair comparison of training times has now been reported in section 1.3.3
+    and figure 1.4 with XDAWN+RG also using a 32 Hz sample rate.
   \item Appendix 1.A now fully reports all obtained accuracies to give a better
@@
     in the text, the effect size has now been reported along with it.
-	\item In the results section of chapter 4, statistical comparisons were
+	\item
+    In the results section of chapter 4, statistical comparisons were
 	  only performed between methods for which we evaluated the performance
@@
 	  fair comparison of performance metrics reported across different works,
-	the analyses for other comparison methods were not re-run
-	by us as reported, but performances were
-	taken from the MOABB benchmark database~\cite{Chevallier2024}.
-	However, this source provided only the aggregated performance metrics
+    the analyses for other comparison methods were not re-run by us (as
+    reported).
+    Instead, but performances were taken from the MOABB benchmark database~\cite{Chevallier2024}.
+	  Unfortunately, this source provides only the aggregated performance metrics
 	  and not the metrics per subject, session and cross-validation fold,
@@
 	  comparison models.
-\item Tables 6.2 and 6.3 now report the absolute cross-condition ROC-AUC values
+  \item
+    Tables 6.2 and 6.3 now report the absolute cross-condition ROC-AUC values
     per model and dataset for a more complete and interpretable presentation of
@@
   \item  I agree that the presented work could have been strengthened by the
-  formulation of clearly testable hypothesis, including a general hypothesis
+  formulation of clearly testable hypotheses, including a general hypothesis
   and per-study subhypotheses.
@@
     moved to the method section as section 5.3.4.
-    In chapter, the analysis investigating observed jitter per condition has
+    In chapter 6, the analysis investigating observed jitter per condition has
     been moved out of the discussion section and has been given a place in the
@@
     believe it is clearer to keep the presentation of the evaluation methods
-    and the results together per data modality in a separate 'Experiments'
+    and the results together per data modality in a separate `Experiments'
     section, instead of also moving the description of evaluation methods to
@@
   \item I very much appreciate the instruction of the reviewer to further
-    investigate user-centered design (UCD).
+    investigate user-centered design (UCD)~\cite{Standardization2009,Kuebler2014}.
     In fact, there was a large overlap between the principles and methods laid
@@
     test~\cite{Rainio2024}.
+
     While both approaches are valid strategies and commonly used~\cite{Rainio2024}, I believe
@@
 	those used for the analyses in chapter 3.
+
 	However, ERP datasets evaluated in chapter 4 were band-pass filtered
@@
 	Decisions to deviate from this were made consciously.
+
 	In chapter 3, filtering between 1 and 24 Hz was applied for consistency
@@
 	database~\cite{Chevallier2024}.
+
 	In chapter 6, the low-pass frequency was increased from 16 Hz to 20 Hz,

### 2024-11-27 chapters/acknowledgment.tex

@@
 Furthermore, we thank the teams at TRAINM Neuro Rehab Clinics Antwerp,
-the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
-Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
+the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie
+specialized care home.
+We wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
 the work presented in~\cref{sec:bttda}.
-We thank the Flemish Supercomputer Center (VSC) and the High-Performance
+Finally, we thank the Flemish Supercomputer Center (VSC) and the High-Performance
 Computing (HPC) center of the KU Leuven for allowing us to execute our

### 2024-11-27 chapters/bttda.tex
Reply to DeLong test

@@
 		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
-		Scores for other decoders were taken from \textcite{Chevallier2024}. Accuracy
-    is listed for BNCI2014-001, and area under the receiver-operator
-    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
-    for BNCI2014-001 but does not reach performance comparable to the current
-    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
+		Scores for other decoders were taken from \textcite{Chevallier2024}.
+    Accuracy is listed for BNCI2014-001, and \ac{rocauc} for BNCI2014-004.
+    BTTDA outperforms HODA and PARAFACDA for BNCI2014-001 but does not reach
+    performance comparable to the current state-of-the-art.
+    For BNCI2014-004, PARAFACDA and BTTDA perform
     approximately on par with HODA and the current state-of-the art.}
@@
 respectively), but these results were not significant, and further studies with
-more datasets and :subjects should show whether this holds.
+more datasets and subjects should show whether this holds.
 \Cref{fig:blocks} shows that there is added value in finding extra blocks

### 2024-11-27 chapters/patients.tex

@@
 according to the international 10-20 layout.
-Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$.
+Using electrolyte gel, electrode impedances were reduced below 10 k$\Omega$.
 Additionally, the \ac{eog} was recorded.
@@
 correlation with the \ac{eog} or by visual inspection.
-Epochs were cut from -0.1 to 0.9s relative to stimulus onset, and no baseline
+Epochs were cut from -0.1 to 0.9 s relative to stimulus onset, and no baseline
 correction was performed in order to meet the assumptions.
@@
 metrics evaluating the user's requirements
-with a measures of usability, comfort and perceived effort, like the NASA Task Load
+with a measure of usability, comfort and perceived effort, like the NASA Task Load
 Index~\cite{Hart2006} and other metrics proposed in the user-centered design

### 2024-11-27 chapters/stbf_struct.tex
Layout chapter gaze independence

@@
 
-  \noindent\emph{This chapter, except
+  \publicationnote{This chapter, except
   \crefrange{tab:stbf-struct/accuracies/stbf-struct}{tab:stbf-struct/accuracies/xdawn-rg},
@@
     \hat{\mat{C}}_\text{emp} =
-		\frac{1}{N-1}\sum^{N}_{n=1}\mat{x}_n\mat{x}_in\intercal
+		\frac{1}{N-1}\sum^{N}_{n=1}\mat{x}_n\mat{x}_i^\intercal
 		\label{eq:stbf-struct/emp-cov}
@@
   Components with a Z-scored pearson correlation coefficient
-  exceeding the other 3 times the standard deviation of the Z-scored
-  correlation coefficients of other component's Z-scored were iteratively
+  exceeding  3 times the standard deviation of the Z-scored
+  correlation coefficients of other components, were iteratively
   rejected until none that exceed the threshold remained.
@@
     \sffamily
-    \sansmath
     \begin{subfigure}{\linewidth}
@@
     \sffamily
-    \sansmath
     \input{figures/stbf_struct/accuracy.pgf}
@@
     \sffamily
-    \sansmath
     \input{figures/stbf_struct/training_time.pgf}
@@
   \fullpagefig{%
-    \sffamily
-    \sansmath
     \input{figures/stbf_struct/accuracy_all.pgf}

### 2024-11-27 chapters/wcble.tex
Reply to DeLong test

@@
 \label{sec:wcble}
-\epigraph{``The sands of time cannot be stopped''}{Brom in Eragon --
-Christopher Paolini}
-\noindent\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
+\epigraph{``The sands of time cannot be stopped.''}{Brom in \emph{Eragon --
+Christopher Paolini}}
+\noindent\emph{\Cref{sec:wcble/literature/contrib},
+\crefrange{sec:wcble/methods/cble}{sec:wcble/methods/wcble} (including
 \cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
-\cref{sec:wcble/conclusion} were published
-as part of~\textcite{VanDenKerchove2024}.}
+\cref{sec:wcble/conclusion} were published as part of~\textcite{VanDenKerchove2024}.}
 
@@
 \subsection{\acl{wcble}}
+\label{sec:wcble/methods/wcble}
 To improve performance over \ac{cble}, we propose a new algorithm inspired both by
@@
 }
-\section{Results}
-\label{sec:wcble/results}
+
+\subsection{Synthetic data generation}
+\label{sec:wcble/methods/simulation}
 Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
@@
 
-\subsection{Synthetic data generation}
 To verify our approach, we simulated synthetic 16 channel \ac{eeg} data with standard
@@
 }
+
+\section{Results}
+\label{sec:wcble/results}
 

### 2024-11-27 figures/bci/paradigms.tikz.tex
Resturcture chap 1 and 2
Rewrite sota
funding
Indicated comments Reinhold and Fabien

@@
-\begin{tikzpicture}[scale=\textwidth/2cm]
+\sffamily
+\begin{tikzpicture}[
+    scale=\textwidth/3cm,
+  ]
+  \useasboundingbox (-1,-1.25) rectangle (1,1);
   \pgfmathsetmacro{\margin}{.1}
@@
   \pgfmathsetmacro{\textspacing}{.15}
-
+  %\begin{scope}[shift={(1,1.2)}]
     % Draw axes
@@
     %% Add text
-    \node[color=accent1, align=left,font=\bfseries, anchor=north west] at  (-1,1){active};
-    \node[color=accent2, align=left,font=\bfseries, anchor=north east] at  (1,1){reactive};
-    \node[color=accent3, align=left,font=\bfseries, anchor=south west] at  (-1,-1){passive};
+    \node[color=accent1, align=left,font=\bfseries, anchor=north west, inner sep=2pt] at  (-1,1){active};
+    \node[color=accent2, align=left,font=\bfseries, anchor=north east, inner sep=2pt] at  (1,1){reactive};
+    \node[color=accent3, align=left,font=\bfseries, anchor=south west, inner sep=2pt] at  (-1,-1){passive};
 
@@
     %% Text in rectangles
-    \node[align=center] at ({-\center}, {\center+\textspacing}) {motor\\imagery};
-    \node[align=center] at ({-\center}, {\center-\textspacing}) {imagined\\speech};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};
 
-    \node[align=center] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
-    \node[align=center] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
-    \node[align=center] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
-    \node[align=center] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};
 
     %% More text
-    \node[align=center] at (\center, {-\center}) {error\\potentials};
-    \node[align=center] at ({-\center}, {-\center}) {attention\\monitoring};
-    \node[align=center] at (-.25, -.8) {emotion\\monitoring};
-    \node[align=center] at (.25, -.8) {media induced\\emotion recognition};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
+    \node[align=center,font=\footnotesize, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
+  %\end{scope}
 \end{tikzpicture}

### 2024-11-27 chapters/stbf_struct.tex
Layout accuracy tables

@@
 	Flattening results in $\{\mat{x}_n\in\mathbb{R}^{CS}\}^N_{n=1}$ such that
-  $\mat{x}_n = \text{vec}(\mat{X}_n)$.
+  $\mat{x}_n = \vect(\mat{X}_n)$.
 	The real covariance matrix of the epochs in vector format is
@@
   \ac{lcmv}-beamforming was initially introduced to \ac{eeg} analysis as a filter for
-	source localization~\cite{VanVeen1997} to enhance the signal-to-noise ratio
-	(SNR).
+  source localization~\cite{VanVeen1997} to enhance the \ac{snr}.
 	\textcite{VanVliet2015} first applied the spatiotemporal
@@
 	the signal characterized by the constraint.
-  $\mat{a} = \text{vec}(\mat{A})$ is the data-driven activation pattern, a template
+  $\mat{a} = \vect(\mat{A})$ is the data-driven activation pattern, a template
 	of the signal of interest maximizing the difference between two classes of
-	epochs, determined as follows:
+  epochs, here denoted as the `target' (T) and `non-target' (NT) class.
+  The activation pattern is then determined as follows:
 	\begin{equation}
 		\mathbf{a} =
-		\frac{1}{|\text{class 1}|}\sum_\text{class 1}\mathbf{x_n} -
-		\frac{1}{|\text{class 2}|}\sum_\text{class 2}\mathbf{x_n}
+    \frac{1}{N_\mathrm{T}}\sum_\mathrm{T}\mathbf{x_n} -
+    \frac{1}{N_\mathrm{NT}}\sum_\mathrm{NT}\mathbf{x_n}
 		\label{eq:stbf-struct/ap}
@@
     (1-\alpha) \hat{\mat{C}}_\text{emp}
-    + \alpha\frac{\text{Tr}(\hat{\mat{C}}_\text{emp})}{CS}\mathbb{I}
+    + \alpha\frac{\Tr(\hat{\mat{C}}_\text{emp})}{CS}\mathbb{I}
 		\label{eq:shrinkage}
@@
 		  \alpha =
-		1-\frac{
-      \frac{N}{N-1}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
-      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
-      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+		  1-\frac{%
+        \splitfrac{
+          \frac{N}{N-1}\Tr(\hat{\mat{C}}_\text{emp}^2)
+          - \frac{2}{CS}\left[\Tr(\hat{\mat{C}}_\text{emp})\right]^2
+          + \frac{1}{CS}\Tr(\hat{\mat{C}}_\text{emp}^2)
+        }{
 		  	  - \frac{1}{N(N-1)}\sum_{n=1}^N||\mat{x}_n||_2^4
         }
-		{
-      \frac{N^2 -2N}{(N-1)^2}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
-      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
-      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
+		  }{%
+        \splitfrac{
+          \frac{N^2 -2N}{(N-1)^2}\Tr(\hat{\mat{C}}_\text{emp}^2)
+          - \frac{2}{CS}\left[\Tr(\hat{\mat{C}}_\text{emp})\right]^2
+          + \frac{1}{CS}\Tr(\hat{\mat{C}}_\text{emp}^2)
+        }{%
 		  	  + \frac{1}{N(N-1)^2}\sum_{n=1}^N||\mat{x}_n||_2^4
+        }
 		  }
@@
       (1-\beta_{k+1})\tilde{\mat{S}}_{k+1}
-      +\beta_{k+1}\frac{\text{Tr}(\tilde{\mat{S}}_{k+1})}{C}\mathbb{I}
+      +\beta_{k+1}\frac{\Tr(\tilde{\mat{S}}_{k+1})}{C}\mathbb{I}
 			\label{eq:stbf-struct/fpi-spatial-shrunk}
@@
       (1-\gamma_{k+1})\tilde{\mat{T}}_{k+1}
-      +\gamma_{k+1}\frac{\text{Tr}(\tilde{\mat{T}}_{k+1})}{S}\mathbb{I}
+      +\gamma_{k+1}\frac{\Tr(\tilde{\mat{T}}_{k+1})}{S}\mathbb{I}
 			\label{eq:stbf-struct/fpi-temporal-shrunk}
@@
       \hat{\mat{S}}_{k+1} =
-      \frac{C}{\text{Tr}\left[\tilde{\mat{S}}_{k+1}^{(\beta)}\right]}
+      \frac{C}{\Tr\left[\tilde{\mat{S}}_{k+1}^{(\beta)}\right]}
       \tilde{\mat{S}}_{k+1}^{(\beta)}
@@
       \hat{\mat{T}}_{k+1} =
-      \frac{S}{\text{Tr}\left[\tilde{\mat{T}}_{k+1}^{(\gamma)}\right]}
+      \frac{S}{\Tr\left[\tilde{\mat{T}}_{k+1}^{(\gamma)}\right]}
       \tilde{\mat{T}}_{k+1}^{(\gamma)}
@@
 			\frac{
-        \frac{N}{N-1}\text{Tr}(\tilde{\mat{\hat{\S}}}_{k+1}^2)
-        - \frac{2}{C}\left[\text{Tr}(\mat{\tilde{S}}_{k+1})\right]^2
-        + \frac{1}{C}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
+        \splitfrac{
+          \frac{N}{N-1}\Tr(\tilde{\mat{\hat{\S}}}_{k+1}^2)
+          - \frac{2}{C}\left[\Tr(\mat{\tilde{S}}_{k+1})\right]^2
+          + \frac{1}{C}\Tr(\tilde{\mat{S}}_{k+1}^2)
+        }{%
 				  - \frac{1}{N(N-1)}\sum_{i=1}^N
-        \left[\text{Tr}(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
+          \left[\Tr(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
+        }
 			}{
-        \frac{N^2-2N}{(N-1)^2}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
-        - \frac{2}{C}\left[\text{Tr}(\tilde{\mat{S}}_{k+1})\right]^2
-        + \frac{1}{C}\text{Tr}(\tilde{\mat{S}}_{k+1}^2)
+        \splitfrac{
+          \frac{N^2-2N}{(N-1)^2}\Tr(\tilde{\mat{S}}_{k+1}^2)
+          - \frac{2}{C}\left[\Tr(\tilde{\mat{S}}_{k+1})\right]^2
+          + \frac{1}{C}\Tr(\tilde{\mat{S}}_{k+1}^2)
+        }{%
 				  + \frac{1}{N(N-1)^2}\sum_{i=1}^N
-        \left[\text{Tr}(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
+          \left[\Tr(\mat{X}_i\hat{\mat{T}}_k^+\mat{X}_i^\intercal)^2\right]
+        }
 			}
@@
 			\frac{
-        \frac{N}{N-1}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
-        - \frac{2}{S}\left[\text{Tr}(\tilde{\mat{T}}_{k+1})\right]^2
-        + \frac{1}{S}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+        \splitfrac{
+          \frac{N}{N-1}\Tr(\tilde{\mat{T}}_{k+1}^2)
+          - \frac{2}{S}\left[\Tr(\tilde{\mat{T}}_{k+1})\right]^2
+          + \frac{1}{S}\Tr(\tilde{\mat{T}}_{k+1}^2)
+        }{%
 				  - \frac{1}{N(N-1)}\sum_{n=1}^N
-        \left[\text{Tr}(\mat{X}_n^\intercal\mat{\mat{S}}_k^+\mat{X}_n)^2\right]
+          \left[\Tr(\mat{X}_n^\intercal\mat{\mat{S}}_k^+\mat{X}_n)^2\right]
+        }
 			}{
-        \frac{N^2-2N}{(N-1)^2}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
-        - \frac{2}{S}\left[\text{Tr}(\tilde{\mat{T}}_{k+1})\right]^2
-        + \frac{1}{S}\text{Tr}(\tilde{\mat{T}}_{k+1}^2)
+        \splitfrac{
+          \frac{N^2-2N}{(N-1)^2}\Tr(\tilde{\mat{T}}_{k+1}^2)
+          - \frac{2}{S}\left[\Tr(\tilde{\mat{T}}_{k+1})\right]^2
+          + \frac{1}{S}\Tr(\tilde{\mat{T}}_{k+1}^2)
+        }{
 				  + \frac{1}{N(N-1)^2}\sum_{n=1}^N
-        \left[\text{Tr}(\mat{X}_n^\intercal\hat{\mat{S}}_k^+\mat{X}_n)^2\right]
+          \left[\Tr(\mat{X}_n^\intercal\hat{\mat{S}}_k^+\mat{X}_n)^2\right]
+        }
 			}
@@
 	The activation pattern is the difference between the averages of epochs in response to cued targets and the averages of those in response to non-cued targets.
-	We constructed three variations of the spatiotemporal beamformer:
-	\ac{stbf} with empirical covariance estimation (\textsc{stbf-emp}) as in
-	\cref{sec:stbf-struct/methods/emp-cov}, \ac{stbf} with
-	\ac{loocv}-shrunk covariance estimation (\textsc{stbf-shrunk}) as in
-	\cref{sec:stbf-struct/methods/shrunk-cov}, and \ac{stbf} with
-	Kronecker-Toeplitz structured covariance estimation (\textsc{stbf-struct}) with \ac{loocv} shrinkage for
+	We constructed three variations of the spatiotemporal beamformer the \ac{stbf-emp} as in
+	\cref{sec:stbf-struct/methods/emp-cov}, the \ac{stbf-shrunk} as in
+	\cref{sec:stbf-struct/methods/shrunk-cov}, and the \ac{stbf-struct} with \ac{loocv} shrinkage for
 	the Kronecker factors as in \cref{sec:stbf-struct/methods/struct-cov}.
@@
 	geometry-based classifier to compare our results.
-	The Riemannian model (\textsc{xdawn+rg}) uses the XDAWN spatial filter combined
+	The Riemannian model (XDAWN+RG) uses the XDAWN spatial filter combined
 	with Riemannian geometry in tangent space as implemented by Barachant et
@@
 	The target with the maximum score is the prediction of the cued target.
-	\textsc{xdawn+rg} was trained and validated without downsampling using epochs
+	XDAWN+RG was trained and validated without downsampling using epochs
 	at the original sample rate of 2048 Hz.
@@
   covariance matrices for this specific dataset.
-  Hence, no shrinkage regularization will be applied to \textsc{xdawn+rg} in
+  Hence, no shrinkage regularization will be applied to XDAWN+RG in
   further experiments.
@@
   estimates the Kronecker-Toeplitz structured covariance for the
-	\textsc{stbf-struct} classifier.
+	\ac{stbf-struct} classifier.
 	Fixed-point iteration is an iterative procedure starting from (in our case)
@@
 
-	\cref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
+	\cref{fig:iterations} confirms these results for the \ac{stbf-struct} classifier.
 	Using more than one fixed-point iteration does not significantly improve the
@@
   used for evaluation.
-	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
+	Hence, only one iteration is used for the \ac{stbf-struct} classifier, leading to a drastic speed-up of
   the  training process.
 
-  \begin{figure}
-    \sffamily
+  \begin{figure}[t]
     \begin{subfigure}{\linewidth}
@@
     \end{subfigure}
-    \caption[Average cross-validated \textsc{stbf-struct} accuracy]{%
-      Average cross-validated \textsc{stbf-struct} accuracy using
+    \caption[Average cross-validated \acs{stbf-struct} accuracy]{%
+      Average cross-validated \ac{stbf-struct} accuracy using
 			one trial per block over all 21 subjects
@@
 	We statistically compare the two newly proposed classifiers,
-	\textsc{stbf-struct} and \textsc{stbf-shrunk} with different levels of training
+	\ac{stbf-struct} and \ac{stbf-shrunk} with different levels of training
 	data availability using a one-sided Wilcoxon signed-rank test with Holm correction for the multiple pairwise comparisons between classifiers.
@@
 			\cref{fig:stbf-struct/accuracy-ap} contains results for all numbers of trials.
-			While \textsc{stbf-emp} is unstable when little training data
+			While \ac{stbf-emp} is unstable when little training data
 			are available,
-			regularization of the covariance matrix (\textsc{stbf-shrunk} and
-			\textsc{stbf-struct}) drastically improves performance.}
+			regularization of the covariance matrix (\ac{stbf-shrunk} and
+			\ac{stbf-struct}) drastically improves performance.}
 		\label{fig:accuracy}
@@
 
-	The tables show that \textsc{stbf-struct} has a significant advantage over
-  \textsc{stbf-shrunk} when the number of training blocks is low, with
+	The tables show that \ac{stbf-struct} has a significant advantage over
+  \ac{stbf-shrunk} when the number of training blocks is low, with
   significance $p=0.005$ and  an accuracy increase of 4,14$\%.$ for a single training block
@@
   blocks respectively, but the advantage decreases when adding more training blocks.
-	Both \textsc{stbf-struct} and \textsc{stbf-shrunk} perform significantly better
-  than \textsc{stbf-emp} for all evaluated settings, with $p<0.001$ for all number
+	Both \ac{stbf-struct} and \ac{stbf-shrunk} perform significantly better
+  than \ac{stbf-emp} for all evaluated settings, with $p<0.001$ for all number
   of training blocks and testing trials, with accuracy
@@
   and testing trial.
-	Compared to \textsc{xdawn+rg}, \textsc{stbf-struct} also has significantly
+	Compared to XDAWN+RG, \ac{stbf-struct} also has significantly
 	higher accuracy in almost all evaluated settings, except when using only one
@@
   5.20$\%.$ when using all 9 training blocks and 1 testing trial.
-	\textsc{stbf-shrunk} does not outperform \textsc{xdawn+rg} when training data
+	\ac{stbf-shrunk} does not outperform XDAWN+RG when training data
   is low (up to 3,3 and 2 blocks respectively for 1, 2 and 5 testing trials),
@@
 
-	\begin{figure}
+  \begin{figure}[t]
 		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
@@
 	\Cref{fig:stbf-struct/training-time} shows that the training time of
-  \textsc{stbf-struct} has a lower median training time (0.06 s) than
-  \textsc{stbf-shrunk} (0.33 s), \textsc{stbf-emp} (0.30 s) and
-  \textsc{xdawn+rg} (0.20 s) when training
+  \ac{stbf-struct} has a lower median training time (0.06 s) than
+  \ac{stbf-shrunk} (0.33 s), \ac{stbf-emp} (0.30 s) and
+  XDAWN+RG (0.20 s) when training
   on all available data and using 32 channels.
@@
 	As evidenced by \cref{fig:accuracy} and \cref{tab:stbf-struct/p-values},
-  the regularized classifiers \textsc{stbf-shrunk}
-	and \textsc{stbf-struct} significantly improve the classification accuracy
-	compared to the original \textsc{stbf-emp} for all numbers of training blocks
+  the regularized classifiers \ac{stbf-shrunk}
+	and \ac{stbf-struct} significantly improve the classification accuracy
+	compared to the original \ac{stbf-emp} for all numbers of training blocks
   indicated.
 	We believe there are three reasons for this.
-	First and foremost, the empirical covariance matrix in \textsc{stbf-emp} becomes
+	First and foremost, the empirical covariance matrix in \ac{stbf-emp} becomes
 	ill-conditioned when the number of available training epochs is smaller than
@@
 	Moore-Penrose pseudoinverse unstable.
-	This is the case \textsc{stbf-emp} when $N=CS=32*17=544$, after which the
-	accuracy of \textsc{stbf-emp} starts to increase.
+	This is the case \ac{stbf-emp} when $N=CS=32*17=544$, after which the
+	accuracy of \ac{stbf-emp} starts to increase.
 	This effect is visible in \cref{fig:accuracy}, where the accuracy starts
@@
 	Another observation is the significantly better accuracy score of
-	\textsc{stbf-struct} over \textsc{stbf-shrunk} when the amount of available
+	\ac{stbf-struct} over \ac{stbf-shrunk} when the amount of available
   training data is small ($<540$ epochs).
@@
 	the brain with added noise~\cite{Munck1992, DeMunck2002, GonzalezNavarro2017}.
-	Hence, \textsc{stbf-struct} can utilize this prior information to better estimate the inverse
+	Hence, \ac{stbf-struct} can utilize this prior information to better estimate the inverse
 	covariance.
@@
 
-	When compared to the state-of-the-art \textsc{xdawn+rg} classifier, we conclude
-	that \textsc{stbf-struct} reaches similar accuracy when using only one block of
+	When compared to the state-of-the-art XDAWN+RG classifier, we conclude
+	that \ac{stbf-struct} reaches similar accuracy when using only one block of
   training data, with accuracies of and 34.70\% 32.5\% respectively and $p=0.086$.
@@
   information to reach satisfactory classification accuracy.
-	When more data are available, \textsc{stbf-struct} reaches significantly
+	When more data are available, \ac{stbf-struct} reaches significantly
   better accuracies.
@@
   makes it an attractive option for \ac{erp} classification.
-	\textsc{stbf-shrunk} does not show decisive accuracy improvements over
-	\textsc{xdawn+rg} using a few training blocks, but this improves as the
+	\ac{stbf-shrunk} does not show decisive accuracy improvements over
+	XDAWN+RG using a few training blocks, but this improves as the
   training data increases.
@@
 	As mentioned above, inverting the full $CS \times CS$ dimensional covariance
-	matrix to construct \textsc{stbf-emp} and \textsc{stbf-shrunk} can be costly
+	matrix to construct \ac{stbf-emp} and \ac{stbf-shrunk} can be costly
 	and unstable, in particular in high-resolution settings with many \ac{eeg} channels or time samples.
 	Constructing the full covariance and inverse covariance matrices also requires a considerable amount of memory.
-	The structured covariance estimator of \textsc{stbf-struct} has two advantages here.
+	The structured covariance estimator of \ac{stbf-struct} has two advantages here.
 
@@
 	\cref{propr:stbf-struct/kron-multiplication} there is no need to calculate or keep in memory the full $cs\times cs$
-	symmetric covariance and inverse covariance matrices for \textsc{stbf-struct}; they can instead be replaced by two smaller symmetric matrices respectively of dimensions $c\times c$ and $s\times s$.
+	symmetric covariance and inverse covariance matrices for \ac{stbf-struct}; they can instead be replaced by two smaller symmetric matrices respectively of dimensions $c\times c$ and $s\times s$.
 	Furthermore, since the temporal component of the Kronecker product is
@@
 	estimate.
-	While the inverse covariance of \textsc{stbf-emp} and \textsc{stbf-shrunk} is
+	While the inverse covariance of \ac{stbf-emp} and \ac{stbf-shrunk} is
 	defined by $\frac{CS(CS+1)}{2}=\frac{32\cdot17(32\cdot17+1)}{2}=\num{122128}$
@@
 	This reduction in parameters to estimate reduces memory usage and contributes to the regularization effect for low data availability settings.
-	The inverse covariances of \textsc{stbf-emp} and
-	\textsc{stbf-struct}, represented as $32*17\times 32*17$ symmetric matrices of
+	The inverse covariances of \ac{stbf-emp} and
+	\ac{stbf-struct}, represented as $32*17\times 32*17$ symmetric matrices of
 	single-precision real floating point numbers for weight calculation,
 	use 9.03MiB of memory.
-	The $32\times 32$ and $17\times 17$ matrices of \textsc{stbf-struct} only
+	The $32\times 32$ and $17\times 17$ matrices of \ac{stbf-struct} only
 	require 5.12KiB.
@@
 	Covariance estimation and inversion occupy the largest part of the \ac{stbf} training time.
-	For \textsc{stbf-emp} and \textsc{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
+	For \ac{stbf-emp} and \ac{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
 	Thanks to Property~\ref{prop:stbf-struct/inverse-kronecker}, the complexity can be reduced to
-	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \textsc{stbf-struct}.
+	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \ac{stbf-struct}.
 	The results presented in \cref{fig:stbf-struct/training-time} confirm that
-  \textsc{stbf-struct} is much faster in operation than other classifiers.
+  \ac{stbf-struct} is much faster in operation than other classifiers.
 
@@
 	%	\cref{fig:interpret_weights} compares the weights calculated in
-	%	\textsc{stbf-emp} and \textsc{stbf-shrunk} with the weights from
-	%	\textsc{stbf-struct}.
+	%	\ac{stbf-emp} and \ac{stbf-shrunk} with the weights from
+	%	\ac{stbf-struct}.
 	%
@@
 	%			power.
-	%			(\textbf{b}) \textsc{stbf-struct} weights with spatial and temporal average of
+	%			(\textbf{b}) \ac{stbf-struct} weights with spatial and temporal average of
 	%			absolute values.
-	%			(\textbf{c}) \textsc{stbf-shrunk} weights. The shrinkage factor $\alpha=0.05$ was
+	%			(\textbf{c}) \ac{stbf-shrunk} weights. The shrinkage factor $\alpha=0.05$ was
 	%			determined with the closed-form \ac{loocv}-method.
-	%			(\textbf{d}) \textsc{stbf-emp} weights.}
+	%			(\textbf{d}) \ac{stbf-emp} weights.}
 	%		\label{fig:interpret_weights}
@@
 	%	suppression, signal amplification, or -- most likely -- a combination of both.
-	%	The weights obtained by \textsc{stbf-emp} look randomly distributed over space
-	%	and time; the regularized estimator used by \textsc{stbf-shrunk} and
-	%	\textsc{stbf-struct} reveal a more interpretable weight distribution.
-	%	The \textsc{stbf-shrunk} weights show a sparse spatial distribution while
-	%	the \textsc{stbf-struct} weights show a sparse distribution in both space and in
+	%	The weights obtained by \ac{stbf-emp} look randomly distributed over space
+	%	and time; the regularized estimator used by \ac{stbf-shrunk} and
+	%	\ac{stbf-struct} reveal a more interpretable weight distribution.
+	%	The \ac{stbf-shrunk} weights show a sparse spatial distribution while
+	%	the \ac{stbf-struct} weights show a sparse distribution in both space and in
 	%	time.
@@
 	%	the central and parietal regions, where the P3 \ac{erp} component is present.
-	%	Especially the spatial weights of \textsc{stbf-shrunk} in
+	%	Especially the spatial weights of \ac{stbf-shrunk} in
 	%	\cref{fig:interpret_weights}d correspond to the spatial activation pattern
@@
 
-  \begin{table}[!h]
-      \footnotesize
+  \begin{table}[h!]
+      \small
+      \sffamily
 	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
-      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for \textsc{stbf-struct}.}
+      \caption[Accuracies for \acs{stbf-struct}.]{Accuracies (\%) for \ac{stbf-struct}.}
       \label{tab:stbf-struct/accuracies/stbf-struct}
   \end{table}
-  \begin{table}[!h]
-      \footnotesize
+  \begin{table}[h!]
+      \small
+      \sffamily
 	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
-      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
-      \textsc{stbf-shrunk}.}
+      \caption[Accuracies for \acs{stbf-shrunk}.]{Accuracies (\%) for
+      \ac{stbf-shrunk}.}
       \label{tab:stbf-struct/accuracies/stbf-shrunk}
   \end{table}
-  \begin{table}[!h]
-      \footnotesize
+  \begin{table}[h!]
+      \small
+      \sffamily
 	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
-      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
-      \textsc{stbf-emp}.}
+      \caption[Accuracies for \acs{stbf-emp}.]{Accuracies (\%) for
+      \ac{stbf-emp}.}
       \label{tab:stbf-struct/accuracies/stbf-emp}
   \end{table}
-  \begin{table}[!h]
-      \footnotesize
+  \begin{table}[h!]
+      \small
+      \sffamily
 	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
-      \caption[Accuracies for \textsc{stbf-struct}.]{Accuracies (\%) for
-      \textsc{xdawn+rg}.}
+      \caption[Accuracies for XDAWN+RG.]{Accuracies (\%) for
+      XDAWN+RG.}
       \label{tab:stbf-struct/accuracies/xdawn-rg}

### 2024-11-27 scripts/setup.py

@@ -68,7 +68,9 @@ def configure_matplotlib_style():
 
 # Golden ratio to set aesthetic figure height
 inches_per_pt = 1 / 72.27
-textwidth_pt = 345
+# textwidth_pt = 345 # manuscript
+textwidth_pt = 318.66946  # booklet
+
 textwidth_in = textwidth_pt * inches_per_pt
 
 

### 2024-11-27 scripts/stbf_struct/accuracy.py

@@ -56,6 +56,8 @@ def plot_accuracy(
     ax.axhline(100 / 9, color=colors["lightgray"], linestyle="--", linewidth=1)
     ax.set_ylim([0, 100])
     ax.set_xticks([2, 4, 6, 8])
+    ax.set_xticklabels([str(int(t)) for t in ax.get_xticks()])
+    ax.set_yticklabels([str(int(t)) for t in ax.get_yticks()])
 
     if show_chance_label:
         ax.annotate(
@@ -81,22 +83,23 @@ def plot_accuracy(
         plt.setp(ax.get_legend().get_title(), fontsize="7")
 
 
-ipdb.set_trace()
-
-
 fig, ax = plt.subplots(1, 1, constrained_layout=True)
+
 plot_accuracy(
     ax,
     trial_accuracy=accuracy[accuracy["n_runs"] == 1],
     show_chance_label=True,
     show_legend=True,
 )
+legend = ax.get_legend()
+legend.set_frame_on(True)  # Ensure the frame is displayed
+legend.get_frame().set_alpha(0.0)
 save_pgf_trim(
     fig,
     ax,
     "figures/stbf_struct/accuracy.pgf",
-    width=3,
-    height=3,
+    width=2.5,
+    height=2,
 )
 
 fig, axs = plt.subplots(5, 3, constrained_layout=True, sharex=True, sharey=True)

### 2024-11-27 scripts/stbf_struct/kronecker_iteration.py

@@ -11,6 +11,7 @@ configure_matplotlib_style()
 iteration_out_path = "data/stbf_struct/iterations.csv"
 
 iteration_accuracy = pd.read_csv(iteration_out_path)
+iteration_accuracy["accuracy"] = iteration_accuracy["accuracy"] * 100
 iteration_accuracy["subject"] = iteration_accuracy["subject"].astype(str)
 iteration_accuracy["fold"] = iteration_accuracy["fold"].astype(str)
 
@@ -19,8 +20,8 @@ def plot_iteration_lines(data, ax, n_trials="", show_chance_label=False):
     sns.lineplot(
         data=data, x="iterations", y="accuracy", ax=ax, marker="s", markersize=3, mew=0
     )
-    ax.set_ylim([0, 1])
-    ax.axhline(1 / 9, color=colors["lightgray"], linestyle="--", linewidth=1)
+    ax.set_ylim([0, 100])
+    ax.axhline(100 / 9, color=colors["lightgray"], linestyle="--", linewidth=1)
     ax.set_xticks([0, 2, 4, 6, 8, 10])
     ax.set_title(n_trials)
     if show_chance_label:
@@ -44,9 +45,11 @@ plot_iteration_lines(
     axs[0],
     n_trials="1 trial",
 )
-axs[0].set_ylabel("Accuracy")
+axs[0].set_ylabel("Accuracy (%)")
 axs[0].set_xlabel("")
 axs[0].set_xticklabels([])
+axs[0].set_yticks([0, 50, 100])
+axs[0].set_yticklabels([str(int(t)) for t in axs[0].get_yticks()])
 plot_iteration_lines(
     data_1_block[data_1_block["n_runs"] == 2], axs[1], n_trials="2 trials"
 )
@@ -58,6 +61,7 @@ plot_iteration_lines(
 axs[2].set_xlabel("")
 axs[2].set_xticklabels([])
 
+
 save_pgf_trim(fig, axs[0], "figures/stbf_struct/fpi-0.pgf", columns=3, rows=1.3)
 
 fig, axs = plt.subplots(1, 3, sharex=True, sharey=True, constrained_layout=True)
@@ -66,22 +70,24 @@ data_1_block = iteration_accuracy[iteration_accuracy["n_train_blocks"] == 9]
 plot_iteration_lines(
     data_1_block[data_1_block["n_runs"] == 1],
     axs[0],
-    n_trials="1 trial",
 )
[…]

### 2024-11-27 scripts/stbf_struct/training_time.py

@@ -61,10 +61,12 @@ def plot_training_time(data, ax, label=None, show_legend=False, show_y_label=Fal
     )
 
     ax.set_ylabel("")
+    ax.set_xticklabels([str(round(t, 2)) for t in ax.get_xticks()])
+    ax.tick_params(axis="y", pad=-5)  # Moves tick labels closer to the axis
 
 
-fig, ax = plt.subplots(1, 1, constrained_layout=True)
+fig, ax = plt.subplots(1, 1)
 plot_training_time(training_time[training_time["n_channels"] == 32], ax)
 
-fig.savefig("figures/stbf_struct/training_time.eps")
+fig.tight_layout()
 save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)

### 2024-11-27 scripts/setup.py
Layout chapter stbf-struct

@@ -87,6 +87,7 @@ def save_pgf_trim(fig, ax, path, width=textwidth_in, height=None, rows=1, column
     margin_left = ax.get_window_extent().x0 / dpi
     print(margin_left)
     fig.set_size_inches(width + margin_left, height)
+    fig.patch.set_alpha(0)
     plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=0)
     prepend = "\\hspace{-" + str(margin_left) + "in}\n"
     with open(path, "r") as original:

### 2024-11-27 scripts/stbf_struct/accuracy.py

@@ -57,7 +57,6 @@ def plot_accuracy(
     ax.set_ylim([0, 100])
     ax.set_xticks([2, 4, 6, 8])
     ax.set_xticklabels([str(int(t)) for t in ax.get_xticks()])
-    ax.set_yticklabels([str(int(t)) for t in ax.get_yticks()])
 
     if show_chance_label:
         ax.annotate(
@@ -81,6 +80,8 @@ def plot_accuracy(
         )
         plt.setp(ax.get_legend().get_texts(), fontsize="7")  # for legend text
         plt.setp(ax.get_legend().get_title(), fontsize="7")
+    ax.set_facecolor("white")
+    return ax
 
 
 fig, ax = plt.subplots(1, 1, constrained_layout=True)
@@ -94,6 +95,7 @@ plot_accuracy(
 legend = ax.get_legend()
 legend.set_frame_on(True)  # Ensure the frame is displayed
 legend.get_frame().set_alpha(0.0)
+ax.set_yticklabels([str(int(t)) for t in ax.get_yticks()])
 save_pgf_trim(
     fig,
     ax,
@@ -104,11 +106,20 @@ save_pgf_trim(
 
 fig, axs = plt.subplots(5, 3, constrained_layout=True, sharex=True, sharey=True)
 for n_runs in range(1, 15 + 1):
-    plot_accuracy(
+    ax = plot_accuracy(
         axs.flatten()[n_runs - 1],
         trial_accuracy=accuracy[accuracy["n_runs"] == n_runs],
         show_chance_label=n_runs == 13,
         show_legend=n_runs == 15,
         title="1 trial" if n_runs == 1 else f"{n_runs} trials",
     )
-save_pgf_trim(fig, axs[0, 0], "figures/stbf_struct/accuracy_all.pgf", rows=8, columns=3)
+    if n_runs in [1, 4, 10, 13]:
+        ax.set_yticklabels(["0", "25", "50", "75", "100"])
+save_pgf_trim(
+    fig,
+    axs[0, 0],
+    "figures/stbf_struct/accuracy_all.pgf",
+    width=textwidth_in * 1.05,
+    rows=8,
+    columns=3,
+)

### 2024-11-27 chapters/gaze_independence.tex
Bleed

@@
 \cref{fig:gaze/vsa}.
-
 \begin{figure}
@@
 \end{figure}
-
 In the overt case, users gazed at the cued

### 2024-11-27 figures/bci/paradigms.tikz.tex
Bleed

@@
     %% Text in rectangles
-    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};
+    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
+    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
+    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};
 
-    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};
+    \node[align=center,font=\small, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
+    \node[align=center,font=\small, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
+    \node[align=center,font=\small, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
+    \node[align=center,font=\small, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};
 
     %% More text
-    \node[align=center,font=\footnotesize, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
-    \node[align=center,font=\footnotesize, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
+    \node[align=center,font=\small, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
+    \node[align=center,font=\small, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
+    \node[align=center,font=\small, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
+    \node[align=center,font=\small, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
   %\end{scope}

### 2024-11-28 chapters/bttda.tex
Layout BTTDA

@@
   ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
-  Lacking this sum, I therefore borrowed fifty coins from each of my two good friends.
-  With the hundred coins, I paid for the beast and received three coins in change.
-  Of these, I returned one coin each to my companions, keeping one for myself.
-  Now, I owe each friend forty-and-nine coins, and with mine one remaining, the
-  total merely equals ninety-and-nine coins, short of one hundred!
+  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
+  With these hundred coins, I paid for the beast and received three coins in change.
+  Of the three, I returned one coin each to my companions, keeping one for myself.
+  Now, I owe each friend forty-and-nine coins, a total debt of
+  ninety-and-eight. And with mine one remaining, the
+  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
   Tell me then, kind stranger, whither hath the missing coin gone?''
@@
 }
-\noindent\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
+\publicationnote{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
 
@@
 \begin{figure}[t]
-	\centering
 	\input{figures/bttda/HODA_backward.tikz.tex}
@@
 \begin{figure}[t]
-	\centering
 	\input{figures/bttda/HODA_forward.tikz.tex}
@@
 \begin{figure*}[t]
-	\centering
 	\input{figures/bttda/BTTDA.tikz.tex}
@@
 \begin{table*}[t]
-	\centering
-	\footnotesize
   \makebox[\textwidth][c]{
@@
   \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
-		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
-    classes (\# Classes), the number of trials or trials per class for ERP
-    datasets (\# Trials), the epoch length (Epoch len.), the sampling
-		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
-		number of runs (\# Runs).
+		subjects (Sub.), the number of EEG channels (Chan.), the number of
+    classes (Cls.), the number of trials or trials per class for ERP
+    datasets (Trials), the epoch length (Epoch len.), the sampling
+		frequency (S. freq.), the number of sessions per subject (Sess.) and the
+		number of runs.
     Adapted from \textcite{Aristimunha2023} and \textcite{Chevallier2024}.}
@@
 \begin{table}[t]
-	\footnotesize
-	\centering
 	\input{chapters/bttda/score_erp.tex}
@@
 \begin{table}[t]
-  \centering
-	\footnotesize
 	\input{chapters/bttda/score_mi.tex}
@@
 \begin{figure*}[t]
-  \centering
-  \makebox[\linewidth][c]{
+  \sffamily
 	\input{figures/bttda/blocks.pgf}
-  }
   \caption[Analysis of NMSE and classification score per block.]{%
@@
 \begin{figure*}[t]
-	\includegraphics[width=\linewidth]{figures/bttda/forward_block-0.png}
-	\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
+  \begin{subfigure}[b]{\linewidth}
+    \begin{minipage}[b]{.15\linewidth}
+      \inputpgf{figures/bttda}{forward_block-0_ap-sp-0.pgf}
+      \vspace{3em}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.15\linewidth}
+      \inputpgf{figures/bttda}{forward_block-0_ap-sp-1.pgf}
+      \vspace{3em}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.25\linewidth}
+      \inputpgf{figures/bttda}{forward_block-0_ap-tmp.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.3\linewidth}
+      \includegraphics[width=\linewidth]{figures/bttda/forward_block-0_contrast.png}
+    \end{minipage}
+    \vspace{-1em}
+    \subcaption{The first block.}
+  \end{subfigure}
+
+  \begin{subfigure}[b]{\linewidth}
+    \begin{minipage}[b]{.15\linewidth}
+      \inputpgf{figures/bttda}{forward_block-1_ap-sp-0.pgf}
+      \vspace{3em}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.15\linewidth}
+      \inputpgf{figures/bttda}{forward_block-1_ap-sp-1.pgf}
+      \vspace{3em}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.25\linewidth}
+      \inputpgf{figures/bttda}{forward_block-1_ap-tmp.pgf}
+    \end{minipage}\hfill%
+    \begin{minipage}[b]{.3\linewidth}
+      \includegraphics[width=\linewidth]{figures/bttda/forward_block-1_contrast.png}
+    \end{minipage}
+    \vspace{-1em}
+    \subcaption{The second block.}
+  \end{subfigure}
+
+	%\includegraphics[width=\linewidth]{figures/bttda/forward_block-0.png}
+	%\includegraphics[width=\linewidth]{figures/bttda/forward_block-1.png}
   \caption[Extracted \acs{bttda} activation patterns.]{%
@@
 		condition contrasts (right column) obtained after forward projection of the latent
-    features for 2 blocks (row 1 and row 2) of rank $(2,2)$ of \ac{bttda}
+    features for 2 blocks of rank $(2,2)$ of \ac{bttda}
     fit on the full dataset BNCI2014-008.

### 2024-11-28 scripts/bttda/plot_blocks.py

@@ -1,0 +1,106 @@
+import pdb
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+from setup import colors, configure_matplotlib_style, save_pgf_trim
+
+configure_matplotlib_style()
+
+df_select = pd.read_csv("data/bttda/block_erp_select.csv")
+df_select = pd.concat(
+    [df_select, pd.read_csv("data/bttda/block_mi_select.csv")], ignore_index=True
+)
+df_select = pd.concat(
+    [df_select, pd.read_csv("data/bttda/block_lr_select.csv")], ignore_index=True
+)
+df_select = df_select.reset_index()
+df_train = pd.read_csv("data/bttda/block_erp_train.csv")
+df_train = pd.concat(
+    [df_train, pd.read_csv("data/bttda/block_mi_train.csv")], ignore_index=True
+)
+df_train = pd.concat(
+    [df_train, pd.read_csv("data/bttda/block_lr_train.csv")], ignore_index=True
+)
+df_train = df_train.reset_index()
+
+df_select = df_select.groupby(["dataset", "subject", "session", "block"])
+df_select = df_select[["train_score", "val_score", "test_score"]].aggregate("mean")
+df_select = df_select.reset_index()
+idx = ["dataset", "subject", "session"]
+df_score_train = df_select.groupby(idx)["train_score"].aggregate("first")
+df_score_val = df_select.groupby(idx)["val_score"].aggregate("first")
+df_score_test = df_select.groupby(idx)["test_score"].aggregate("first")
+df_diff = df_select.set_index(idx + ["block"])
+df_diff["validation"] = df_diff["val_score"] - df_score_val
+df_diff["training"] = df_diff["train_score"] - df_score_train
+df_diff["test"] = df_diff["test_score"] - df_score_test
+df_diff = df_diff.reset_index()
+df_diff = df_diff.melt(
+    id_vars=idx + ["block"],
+    value_vars=["training", "validation", "test"],
+    var_name="split",
+    value_name="roc_auc_diff",
+)
+
+df_diff["n_blocks"] = df_diff["block"] + 1
+df_diff["roc_auc_diff_%"] = df_diff["roc_auc_diff"] * 100
+fig, axs = plt.subplots(2, 2, layout="tight", sharex=True)
+lineplot_kwargs = dict(
+    x="n_blocks",
+    hue="dataset",
[…]

### 2024-11-28 scripts/bttda/plot_forward.py

@@ -1,0 +1,80 @@
+import pdb
+
+import matplotlib.pyplot as plt
+import mne
+import numpy as np
+from mne.viz import plot_topomap
+from setup import (colors, configure_matplotlib_style, diverging_cmap,
+                   save_pgf_trim, textwidth_in)
+
+configure_matplotlib_style()
+# plt.rcParams["axes.prop_cycle"] = "cycler('color', 'krb')"
+
+
+for b in range(2):
+    # Plot reconstructed contrast
+    contrast_rec = mne.read_evokeds(f"data/bttda/forward/contrast_rec_block-{b}.fif")[0]
+    contrast_rec.apply_baseline((None, 0))
+    contrast_rec.nave = None
+
+    fig, ax = plt.subplots(1, 1)
+    contrast_rec.plot(axes=ax, show=False)
+    ax.set_title("")
+    ax.set_ylabel("Amplitude (µV)")
+    ax.set_yticks([-2, 0, 2])
+    ax.set_ylim([-2, 2])
+    ax.set_xticks([0, 0.5, 1])
+    ax.set_xlim([-0.2, 1])
+    ax.axvline(0, color=colors["lightgray"], linestyle="--", linewidth=1, zorder=-1)
+    ax.axhline(0, color=colors["lightgray"], linestyle="--", linewidth=1, zorder=-1)
+
+    fig.set_size_inches(0.3 * textwidth_in, 0.25 * textwidth_in)
+    fig.savefig(
+        f"figures/bttda/forward_block-{b}_contrast.png",
+        bbox_inches="tight",
+        pad_inches=0,
+        dpi=300,
+    )
+
+    # Plot temporal aps
+    fig, ax = plt.subplots(1, 1)
+    aps_tmp = np.load(f"data/bttda/forward/ap_block-{b}_mode-{1}.npy")
+    ax.axvline(0, color=colors["lightgray"], linestyle="--", linewidth=1, zorder=-1)
+    ax.axhline(0, color=colors["lightgray"], linestyle="--", linewidth=1, zorder=-1)
+    ax.plot(contrast_rec.times, aps_tmp)
+    ax.set_yticks([-1, 0, 1])
+    ax.set_yticklabels([str(int(t)) for t in ax.get_yticks()])
+    ax.set_ylim([-1.5, 1.5])
+    ax.set_ylabel("Scale")
+    ax.set_xticks([0, 0.5, 1])
+    ax.set_xticklabels([str(round(t, 1)) for t in ax.get_xticks()])
[…]

### 2024-11-28 scripts/setup.py

@@ -43,7 +43,11 @@ def configure_matplotlib_style():
             "ytick.direction": "in",  # Y-axis ticks inward
             "xtick.major.size": 3,  # Major tick size for x-axis
             "ytick.major.size": 3,  # Major tick size for y-axis
-            "text.latex.preamble": "\\usepackage{sansmath}\\sansmath",
+            "text.latex.preamble": """
+                \\usepackage[scale=.8]{opensans}
+                \\renewcommand{\sffamily}{\opensans}
+                \\usepackage{sansmath}
+                \\sansmath""",
             "axes.edgecolor": colors["darkgray"],  # Spine color
             "axes.labelcolor": colors["darkgray"],  # Axis label color
             "axes.titlesize": font_size,  # Axis title font size
@@ -89,7 +93,7 @@ def save_pgf_trim(fig, ax, path, width=textwidth_in, height=None, rows=1, column
     fig.set_size_inches(width + margin_left, height)
     fig.patch.set_alpha(0)
     plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=0)
-    prepend = "\\hspace{-" + str(margin_left) + "in}\n"
+    prepend = "\\hspace{-" + str(margin_left) + "in}\n%"
     with open(path, "r") as original:
         data = original.read()
     with open(path, "w") as modified:

### 2024-11-28 chapters/bttda.tex
Layout chapter wcble

@@
 \input{chapters/bttda/tensorviz.tex}
-\epigraph{%
-  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
-  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
-  With these hundred coins, I paid for the beast and received three coins in change.
-  Of the three, I returned one coin each to my companions, keeping one for myself.
-  Now, I owe each friend forty-and-nine coins, a total debt of
-  ninety-and-eight. And with mine one remaining, the
-  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
-  Tell me then, kind stranger, whither hath the missing coin gone?''
-}{%
-  Ancient Algerian riddle
-}
+\epigraph{
+``Traveling through hyperspace ain't like dusting crops, boy!
+Without precise calculations we could fly right through a star or bounce too
+close to a supernova and that'd end your trip real quick, wouldn't it?''
+}{Han Solo in \emph{Star Wars Episode IV: A New Hope}}
 \publicationnote{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}
@@
     \end{minipage}
-    \vspace{-1em}
+    %\vspace{-1em}
     \subcaption{The first block.}
@@
     \end{minipage}
-    \vspace{-1em}
+    %\vspace{-1em}
     \subcaption{The second block.}

### 2024-11-28 chapters/stbf_struct.tex
Layout chapter stbf-struct
Bleed
Layout BTTDA
Layout chapter wcble

@@
 	#1
+}
+\epigraph{%
+  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
+  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
+  With these hundred coins, I paid for the beast and received three coins in change.
+  Of the three, I returned one coin each to my companions, keeping one for myself.
+  Now, I owe each friend forty-and-nine coins, a total debt of
+  ninety-and-eight. And with mine one remaining, the
+  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
+  Tell me then, kind stranger, whither hath the missing coin gone?''
+}{%
+  Ancient Algerian riddle
 }
@@
 
-  \begin{table}[h!]
+  \begin{table}[!ht]
       \small
@@
   \end{table}
-  \begin{table}[h!]
+  \begin{table}[!ht]
       \small
@@
 
-  \fullpagefig{%
+  \fullpagefig[caption.south, anchor.north]{%
     \input{figures/stbf_struct/accuracy_all.pgf}
@@
   \section{Statistical comparisons}
-  \begin{table}[h]
-    \footnotesize
+  \vfill
+  \begin{table}[!ht]
+    \centering
+    \begin{adjustbox}{angle=90}
       \begin{subtable}{\linewidth}
@@
       \end{subtable}
-    \smallskip
-
+    \end{adjustbox}
+    \begin{adjustbox}{angle=90}
       \begin{subtable}{\linewidth}
@@
       \end{subtable}
+    \end{adjustbox}
     \caption[Statistical significance of differences in classifier
@@
 	\end{table}
+  \vfill
 

### 2024-11-28 scripts/setup.py

@@ -78,7 +78,9 @@ textwidth_pt = 318.66946  # booklet
 textwidth_in = textwidth_pt * inches_per_pt
 
 
-def save_pgf_trim(fig, ax, path, width=textwidth_in, height=None, rows=1, columns=1):
+def save_pgf_trim(
+    fig, ax, path, width=textwidth_in, height=None, rows=1, columns=1, **kwargs
+):
     if height is None:
         golden_ratio = (5**0.5 - 1) / 2
         height = rows * width * golden_ratio / columns
@@ -92,7 +94,7 @@ def save_pgf_trim(fig, ax, path, width=textwidth_in, height=None, rows=1, column
     print(margin_left)
     fig.set_size_inches(width + margin_left, height)
     fig.patch.set_alpha(0)
-    plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=0)
+    plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=0, **kwargs)
     prepend = "\\hspace{-" + str(margin_left) + "in}\n%"
     with open(path, "r") as original:
         data = original.read()

### 2024-11-28 scripts/wcble/fig_simulated.py

@@ -132,4 +132,10 @@ for j, snr in enumerate(snrs):
             ax_ts.tick_params(axis="x", labelbottom=False)
 
 
-save_pgf_trim(plt.gcf(), axs[0, 0], f"figures/wcble/simulated-{stc}.pgf", height=8)
+save_pgf_trim(
+    plt.gcf(),
+    axs[0, 0],
+    f"figures/wcble/simulated-{stc}.pgf",
+    height=8,
+    transparent=True,
+)

### 2024-11-29 chapters/covert_align.tex
Layout chapter covert align except fig 1.4 and 1.5

@@
 Scandal in Bohemia -- Sir Arthur Conan Doyle}}
-\noindent\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
+\publicationnote{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
 
@@
 The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
-consists of 15 participants, mean age $26.34\pm3.04$ years.
+consists of 15 participants, mean age 26.34$\pm$3.04 years.
 This study was approved by the Ethics Commission of University Hospital Leuven
@@
 considered.
-\begin{figure*}
-  \bigskip
-	\begin{subfigure}{\linewidth}
-		\caption{}%
-		\label{fig:interface}
-		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
-	\end{subfigure}
-
-	\bigskip
-	\bigskip
-	\begin{subfigure}{\linewidth}
-		\caption{}%
-		\label{fig:erps}
-		\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
-	\end{subfigure}%
-
-  \caption[Stimulation interfaces and evoked \acsp{erp}.]{%
-		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
+\begin{figure*}[t]
+		\includegraphics[width=\linewidth]{figures/covert_align/dataset_vsa.pdf}
+    \caption[Dataset stimulation interfaces.]{%
+    Interfaces and visuospatial attention (\ac{vsa})
     conditions in the CVSA-ERP and BNCI2014-009 datasets.
@@
     feature attention.
-		\subref{fig:erps} Contrast between target (color) and non-target, and
+
+    }%
+		\label{fig:interface}
+\end{figure*}
+\begin{figure*}[t]
+  \input{figures/covert_align/grand_avg.pfg}
+	%\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
+  \caption[Evoked \acs{erp} per dataset and condition.]{%
+		Contrast between target (color) and non-target, and
     distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
@@
 	}
+		\label{fig:erps}
 \end{figure*}
@@
 
-\begin{figure}
+\begin{figure}[t]
   \begin{subfigure}[t]{.45\textwidth}
@@
   \end{subfigure}
+
+  \bigskip
 
@@
   \begin{table}[ht]
-  \centering
-  \makebox[\textwidth][c]{%
   \input{chapters/covert_align_performance.tex}
-  }
-  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
-curve for all evaluated models, visuospatial attention conditions, and
+    \caption[Classification performances.]{Cross-validated single-trial
+    classification \ac{rocauc} (\%) for all evaluated models, visuospatial attention conditions, and
     datasets.}
@@
   \end{table}
-
   \clearpage
   \section{Cross-condition transfer classification performances}
-
   \begin{table}[ht]
-      \footnotesize
       \input{chapters/covert_align_cross-CVSA-ERP.tex}
@@
   \end{table}
+  \clearpage
   \begin{table}[ht]
-      \footnotesize
       \input{chapters/covert_align_cross-BNCI2014-009.tex}

### 2024-11-29 scripts/covert_align/grand_avg.py

@@ -1,0 +1,170 @@
+import os
+import pdb
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from matplotlib.gridspec import GridSpec
+from mne import combine_evoked, pick_info, read_evokeds
+from mne.transforms import apply_trans
+from mne.viz import plot_compare_evokeds
+from mne.viz.utils import plot_sensors
+from setup import (colors, condition_cmap, configure_matplotlib_style,
+                   save_pgf_trim)
+
+configure_matplotlib_style()
+
+df = pd.read_csv("data/covert_align/evokeds.csv")
+
+for ind in df.index:
+    path = df.loc[ind, "evoked_path"]
+    path = os.path.join("data", "covert_align", path)
+    evoked = read_evokeds(path)[0]
+    assert not np.any(np.isnan(evoked.data))
+    evoked = evoked.apply_baseline((-0.2, 0))
+    df.loc[ind, "evoked"] = evoked
+df = df.groupby(["dataset", "condition", "label"])
+df = df["evoked"].aggregate(lambda x: combine_evoked(list(x), weights="equal"))
+df = df.reset_index()
+
+# Plot
+plt.rcParams.update(
+    {
+        "lines.linewidth": 1.5,
+    }
+)
+
+
+def plot_sensor_picks(ax, pick):
+    plot_sensors(
+        evoked.info,
+        kind="topomap",
+        ch_type="eeg",
+        block=False,
+        show=False,
+        axes=ax,
+        show_names=False,
+        pointsize=0.2,
+    )
+    ch_idc = [i for i, ch in enumerate(evoked.ch_names) if ch in pick]
+    plot_sensors(
+        pick_info(evoked.info, ch_idc),
+        kind="topomap",
+        ch_type="eeg",
+        block=False,
+        show=False,
+        axes=ax,
+        show_names=False,
+        pointsize=2,
+    )
+    ax.set_xlim([-0.15, 0.15])
+    ax.set_ylim([-0.15, 0.15])
+
+
+def plot_evoked(ax, target, non_target, color, ylim, distractor=None):
+    target_contrast = combine_evoked([target, non_target], weights=[1, -1])
+    ax.plot(
+        target_contrast.times,
[…]

### 2024-11-29 scripts/setup.py

@@ -16,6 +16,25 @@ colors = dict(
 diverging_cmap = "RdYlBu_r"
 
 
+condition_cmap = {
+    "overt": colors["accent1"],
+    "covert": colors["accent2"],
+    "split ($d=1$)": colors["accent3"],
+    "split ($d=2$)": colors["accent3"],
+    "split ($d=3$)": colors["accent3"],
+}
+condition_order = ["overt", "covert", "split ($d=1$)", "split ($d=2$)", "split ($d=3$)"]
+
+
+MODEL_CMAP = {
+    "WCBLE": "#AF56BA",
+    "CBLE": "#b488ba",
+    "tLDA": "#bababa",
+    "XDAWNCov-TS-LR": "#FFC61E",
+    "ERPCov-TS-LR": "#F28522",
+}
+
+
 # Function to configure matplotlib styles
 def configure_matplotlib_style():
     custom_color_palette = [
@@ -73,13 +92,14 @@ def configure_matplotlib_style():
 # Golden ratio to set aesthetic figure height
 inches_per_pt = 1 / 72.27
 # textwidth_pt = 345 # manuscript
-textwidth_pt = 318.66946  # booklet
+# textwidth_pt = 318.66946  # booklet
+textwidth_pt = 341.43307
 
 textwidth_in = textwidth_pt * inches_per_pt
 
 
 def save_pgf_trim(
-    fig, ax, path, width=textwidth_in, height=None, rows=1, columns=1, **kwargs
+    fig, ax, path, width=textwidth_in, height=None, rows=1, columns=1, pad=0, **kwargs
 ):
     if height is None:
         golden_ratio = (5**0.5 - 1) / 2
@@ -94,12 +114,16 @@ def save_pgf_trim(
     print(margin_left)
     fig.set_size_inches(width + margin_left, height)
     fig.patch.set_alpha(0)
-    plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=0, **kwargs)
-    prepend = "\\hspace{-" + str(margin_left) + "in}\n%"
+    plt.savefig(path, format="pgf", bbox_inches="tight", pad_inches=pad, **kwargs)
+    prepend = "\\hspace{-" + str(margin_left + pad) + "in}\n%"
+    if pad:
+        append = "\\vspace{-" + str(pad) + "in}"
+    else:
+        append = ""
     with open(path, "r") as original:
         data = original.read()
     with open(path, "w") as modified:
-        modified.write(prepend + data)
[…]

### 2024-11-29 chapters/conclusion.tex
Layout conclusion

@@
 \epigraph{``The mystery of life isn't a problem to solve, but a reality to
-experience.''}{ Reverend Mother Gaius Helen Mohiam in \\ \emph{Dune -- Frank
-Herbert}\\ adapted from S{\o}ren Kierkegaard}
-\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
+experience.''}{The Reverend Mother Gaius Helen \\ Mohiam in \emph{Dune -- Frank
+Herbert}, \\ originally from S{\o}ren Kierkegaard}
+\publicationnote{\Cref{sec:conclusion/contrib/decoders} was published as
 part of \textcite{VanDenKerchove2022}.}
@@
 The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
-$26.38\pm3.15$ years.
+26.38$\pm$3.15 years.
 The dataset was presented in~\textcite{VanDenKerchove2024}.
@@
 Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
-\ac{als},\ac{fa}, and stroke.
+\ac{als}, \ac{fa} and stroke.
 These participants exhibited varying degrees of eye motor impairment, such as
@@
 \emph{satisfaction}
-Similarly to effectiveness, efficiency can objectively be measured as \ac{itr}
-in on-line operation is a better metric.
+Similarly to effectiveness, efficiency can objectively be measured as the \ac{itr}
+achieved in -- importantly -- on-line operation.
+
 Satisfaction, on the other hand, is a more subjective quality.
-
 Nevertheless, optimizing user satisfaction is the penultimate goal in
@@
 the context of the topics presented in this work.
-The patient-research involvement framework postulates that, ideally, individuals with
-\ac{sspi} should be involved at every stage of the research project, from conceptualization
+Another valuable principle of \ac{ucd} principle points the way: encouraging early and active
+involvement of end-users.
+This entails that, ideally, individuals with \ac{sspi} should be involved at
+every stage of the research project, from conceptualization
 to evaluation.
@@
 
-This approach is tied to another valuable \ac{ucd} principle: encouraging early and active
-involvement of users.
-Unfortunately, it also might one of the least practical ways to set up
+Unfortunately, this approach might also be one of the least practical ways to set up
 a 4-year research project, such as in the context of a doctoral thesis.
-This approach can only work if the proper facilities can be gathered immediately
+It can only work if the proper facilities can be gathered immediately
 from the start, such as ethical approval, access to patients, and on-line \acp{bci}

### 2024-11-29 chapters/patients.tex
Layout conclusion

@@
 \label{sec:patients}
-\epigraph{I knew there would be a lot of unexpected issues involved with
-patient experiments, but I would never have guessed that there would be baby
-goats in the experiment room\ldots}{Juliette}
+\epigraph{``I knew there would be a lot of unexpected issues involved with
+patient experiments, but I never would have guessed that there would be a baby
+goat in the experiment room\ldots''}{Juliette}
 
@@
 \begin{enumerate}
-  \item have a diagnosis of a major medical condition, including any major
+  \item had a diagnosis of a major medical condition, including any major
     neurological or psychiatric disorder other than those of interest based on
@@
     and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
-  \item have a predisposition to or have a history of any kind of epileptic seizures,
+  \item had a predisposition to or a history of any kind of epileptic seizures,
     including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
-  \item have a severe loss in vision or hearing that would significantly impair
+  \item had a severe loss in vision or hearing that would significantly impair
         participation in the experiment,\label{item:patients/exclusion/vision}
   \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
-  \item be unable to understand the experiment instructions and cooperate,
-  \item have any other limitations preventing them from performing the given task.
+  \item were unable to understand the experiment instructions and cooperate,
+  \item had any other limitations preventing them from performing the given task.
 \end{enumerate}
 
-In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
-criterion~\ref{item:patients/exclusion/vision}, one person recovering from
-\ac{tbi} on
-both~\ref{item:patients/exclusion/epilepsy}
+In total, 11 individuals were contacted. Of these, which one person with
+\ac{ms} was excluded based on criterion~\ref{item:patients/exclusion/vision}.
+One person recovering from \ac{tbi} was excluded based on both~\ref{item:patients/exclusion/epilepsy}
 and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
-stroke based
-on~\ref{item:patients/exclusion/medical}.
+stroke based on~\ref{item:patients/exclusion/medical}.
 One further person recovering from a stroke was excluded due to technical
@@
 diagnoses.
-
-\newcommand{\fnkuebler}{%
-  \footnote{
-``With minor degree of impairment,
-we refer to patients who had only slightly impaired limb movement
-and normal speech. Under the category moderate impairment, we
-summarized those patients with restricted limb movement (wheelchair-bound) and
-unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
-and swallowing). Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
-and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
-  }
-}
 \begin{table}[t]
@@
   capabilities.
-  Trach.: underwent a tracheostomy, Cls. W: classification according
-  to~\textcite{Wolpaw2006}, Cls. KB: classification according
-  to~\textcite{Kuebler2008}\fnkuebler.
+  Trach.: underwent a tracheostomy, W: classification according
+  to~\textcite{Wolpaw2006}\footnotemark[1], KB: classification according
+  to~\textcite{Kuebler2008}\footnotemark[2].
   }
@@
 \end{table}
+\subsection{Visual skills and eye tracking and eye motor examination}
+\footnotetext[1]{See \cpageref{fn:bci/wolpawcrit} \cref{fn:bci/wolpawcrit}.}
+\footnotetext[2]{``With minor degree of impairment,
+we refer to patients who had only slightly impaired limb movement
+and normal speech. Under the category moderate impairment, we
+summarized those patients with restricted limb movement (wheelchair-bound) and
+unaffected speech or intact limb movement without speech. \elide Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
+and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}}
 %\todo{include years since diagnosis}
-
-\subsection{Visual skills and eye tracking and eye motor examination}
 
@@
 \begin{table}[t]
-  \centering
-  \footnotesize
   \input{chapters/patients_table_eye.tex}
@@
 
-\fullpagefig{%
-  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
+\fullpagefig[anchor.north]{%
+  %\includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
+  %\includegraphics[width=\textwidth]{figures/patients/fig_gaze.pdf}
+  \inputpgf{figures/patients}{fig_gaze.pgf}
 }{%

### 2024-11-29 chapters/patients_table.tex
Layout conclusion

--- 2024-10-02 chapters/patients_table.tex
@@
-\begin{tabular}{llllllllr}
+\sffamily
+\small
+\begin{tabular}{@{}llrrrlllrr@{}}
   \toprule
-  \textbf{ID}  & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
-  \textbf{Diagnosis}
-  & \textbf{Speech}     & \textbf{Trach.} & \textbf{Communication}          &
-  \textbf{Cls.} \\ \midrule
-  PA1 & 58  & M   & L     & bulbar-onset \acs{als} & anarthric  & no          & tablet                 & 3  \\
-  PB1 & 41  & M   & L     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PB2 & 43  & F   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PB4 & 48  & M   & R     & \acs{fa} & dysarthric & no          & verbal                 & 3  \\
-  PC2 & 43  & M   & R     & brainstem stroke & anarthric  & yes         & \makecell[l]{prompting\\+eye movement} & 2 \\
-  PC3 & 43  & F   & R     & brainstem stroke & anarthric  & yes         & letterboard            & 2 \\
-  PC4 & 54  & M   & R     & \makecell[l]{left cerebellar stroke \\ (trombosis of the basilar artery)} & anarthric  & yes & letterboard & 2 \\
+  \textbf{ID}  & \textbf{Diagnosis} & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
+  \textbf{Speech} & \textbf{Trach.} & \textbf{Communication} &
+  \textbf{W} & \textbf{KB} \\ \midrule
+  PA1 & bulbar-onset \acs{als} & 58  & M   & L     & anarthric  & no          & tablet                 & 3  & 4 \\
+  PB1 & \acs{fa} & 41  & M   & L     & dysarthric & no          & verbal                 & 3  & 3 \\
+  PB2 & \acs{fa} & 43  & F   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
+  PB4 & \acs{fa} & 48  & M   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
+  PC2 & brainstem stroke & 43  & M   & R     & anarthric  & yes         &  \makecell[l]{prompting\\+eye movement} & 2 & 4  \\
+  PC3 & brainstem stroke & 43  & F   & R     & anarthric  & yes         & letterboard            & 2  & 3 \\
+  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  basilar \\ artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
   \bottomrule

### 2024-11-29 chapters/stbf_struct.tex
Layout conclusion

@@
 \epigraph{%
-  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
+  I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
   Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
@@
   total merely equals ninety-and-nine coins, short of the hundred I borrowed!
-  Tell me then, kind stranger, whither hath the missing coin gone?''
+  Tell me then, kind stranger, whither hath the missing coin gone?
 }{%

### 2024-11-29 scripts/patients/fig_cross.py

@@ -7,7 +7,7 @@ from setup import colors, configure_matplotlib_style, save_pgf_trim
 
 configure_matplotlib_style()
 
-df = pd.read_csv('data/patients/results_cross.csv')
+df = pd.read_csv("data/patients/results_cross.csv")
 df = df[df["subject"] != "PD01"]
 df = df[df["subject"] != "PB03"]
 print(df)
@@ -16,21 +16,25 @@ df_avg = df.copy()
 df_avg["subject"] = "avg."
 df = pd.concat([df, df_avg], ignore_index=True)
 
-df = df.replace({
-    'PA01': 'PA1',
-    'PB01': 'PB1',
-    'PB02': 'PB2',
-    'PB04': 'PB4',
-    'PC02': 'PC2',
-    'PC03': 'PC3',
-    'PC04': 'PC4',
-    'overt': 'overt VSA',
-    'covert': 'covert VSA',
-    'free': 'free VSA',
-})
+df = df.replace(
+    {
+        "PA01": "PA1",
+        "PB01": "PB1",
+        "PB02": "PB2",
+        "PB04": "PB4",
+        "PC02": "PC2",
+        "PC03": "PC3",
+        "PC04": "PC4",
+        "overt": "overt VSA",
+        "covert": "covert VSA",
+        "free": "free VSA",
+    }
+)
 
-df['test setting'] = df['cond_test']
-df['train setting'] = df['cond_train']
+df["test setting"] = df["cond_test"]
+df["train setting"] = df["cond_train"]
+
+df["test_score"] = df["test_score"] * 100
 
 fg = sns.catplot(
     data=df,
@@ -41,19 +45,22 @@ fg = sns.catplot(
     col="test setting",
     legend="auto",
     legend_out=False,
-    col_order=['overt VSA', 'covert VSA', 'free VSA']
+    col_order=["overt VSA", "covert VSA", "free VSA"],
 )
-fg.set(ylim=(0, 1))
+fg.axes[0, 0].set_yticks([0, 20, 40, 60, 80, 100])
+fg.axes[0, 0].set_yticklabels([str(int(t)) for t in fg.axes[0, 0].get_yticks()])
+fg.set(ylim=(0, 100))
 for ax in fg.axes.flat:
     ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
-    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
[…]

### 2024-11-29 scripts/patients/fig_decode.py

@@ -20,17 +20,19 @@ df_avg = df.copy()
 df_avg["subject"] = "avg."
 df = pd.concat([df, df_avg], ignore_index=True)
 
-df = df.replace({
-    'PA01': 'PA1',
-    'PB01': 'PB1',
-    'PB02': 'PB2',
-    'PB04': 'PB4',
-    'PC02': 'PC2',
-    'PC03': 'PC3',
-    'PC04': 'PC4',
+df = df.replace(
+    {
+        "PA01": "PA1",
+        "PB01": "PB1",
+        "PB02": "PB2",
+        "PB04": "PB4",
+        "PC02": "PC2",
+        "PC03": "PC3",
+        "PC04": "PC4",
+    }
+)
 
-})
-
+df["test_score"] = df["test_score"] * 100
 fg = sns.catplot(
     data=df,
     x="subject",
@@ -41,21 +43,25 @@ fg = sns.catplot(
     legend="auto",
     legend_out=False,
 )
-fg.set(ylim=(0, 1))
-plt.legend(
-    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
-)  # Adjust ncol to fit all labels in one line
+fg.set(ylim=(0, 100))
+legend = fg.axes[0, 0].get_legend()
+legend.set_title("")
+legend.set_frame_on(True)
+legend.get_frame().set_linewidth(0)
+legend.get_frame().set_boxstyle("Square")  # Remove rounded corners
 
-fg._legend.set_frame_on(True)
-fg._legend.get_frame().set_facecolor("white")
-fg._legend.get_frame().set_alpha(1)  # Ensure opaque background (non-transparent)
-
+legend.set_in_layout(False)
+# plt.legend(
+#    bbox_to_anchor=(-0.4, -0.3), loc="upper center", ncol=3
+# )  # Adjust ncol to fit all labels in one line
+fg.axes[0, 0].set_yticks([0, 20, 40, 60, 80, 100])
+fg.axes[0, 0].set_yticklabels([str(int(t)) for t in fg.axes[0, 0].get_yticks()])
 
 
 for ax in fg.axes.flat:
     ax.set_xticklabels(ax.get_xticklabels(), rotation=90)
-    ax.axhline(0.5, color=colors["darkgray"], linestyle="--", linewidth=1)
-fg.set_axis_labels("patient", "ROC-AUC")
[…]

### 2024-11-29 scripts/patients/fig_gaze.py

@@ -5,7 +5,7 @@ import pandas as pd
 import seaborn as sns
 from matplotlib.patches import Rectangle
 from mne import read_epochs
-from setup import colors, configure_matplotlib_style, pagesize, savefig_kws
+from setup import colors, configure_matplotlib_style, textwidth_in
 
 configure_matplotlib_style()
 
@@ -48,7 +48,7 @@ fig, axs = plt.subplots(
     len(conditions),
     sharex=True,
     sharey=True,
-    figsize=pagesize,
+    figsize=(textwidth_in, 1.75 * textwidth_in),
     layout="constrained",
 )
 axs = axs.reshape((len(subjects), len(conditions)))
@@ -56,17 +56,17 @@ for i, (subject, sub) in enumerate(subjects.items()):
     for j, condition in enumerate(conditions):
         axs[i, j].set_title(f"{subject}, {condition} VSA")
         hist = np.load(f"data/patients/gaze/gaze_sub-{sub}_cond-{condition}.npy")
-        axs[i, j].pcolormesh(xedges, yedges, hist.T)
+        axs[i, j].pcolormesh(xedges, yedges, hist.T, rasterized=True)
         axs[i, j].set_aspect("equal")
         axs[i, j].set_axis_off()
         for target in ["center", 0, 1, 2, 3, 4, 5]:
             pos = get_target_pos(target)
-            if condition=='overt':
-                c=colors["lightgray"] if target == "center" else colors["accent1"]
-            elif condition=='covert':
-                c=colors["accent1"] if target == "center" else colors["lightgray"]
-            elif condition == 'free':
-                c=colors["lightgray"]
+            if condition == "overt":
+                c = colors["lightgray"] if target == "center" else colors["accent1"]
+            elif condition == "covert":
+                c = colors["accent1"] if target == "center" else colors["lightgray"]
+            elif condition == "free":
+                c = colors["lightgray"]
             axs[i, j].scatter(
                 pos[0],
                 pos[1],
@@ -104,5 +104,6 @@ ax_top_left.text(
     va="bottom",
 )
 
-fig.savefig("figures/patients/fig_gaze.pgf", **savefig_kws)
-fig.savefig("figures/patients/fig_gaze.png", dpi=300, **savefig_kws)
+fig.savefig(
+    "figures/patients/fig_gaze.pgf", bbox_inches="tight", pad_inches=0, transparent=True
+)

### 2024-11-29 chapters/abstract.tex
Second layout run

@@
 \chapter*{Abstract}
+\addcontentsline{toc}{chapter}{Abstract}
 
-Individuals with severe motor impairments, such as those with \ac{lis}
-or eye-motor dysfunction, face substantial challenges when using traditional
+Individuals with severe motor impairment, such as those with \acl{lis}, face substantial challenges when using traditional,
 gaze-dependent \acp{bci}.
-These systems require users to direct their gaze toward specific targets, a
-task that becomes unfeasible for individuals with limited or no eye control.
+These systems require directing the gaze toward specific targets, which
+becomes unfeasible for individuals with limited or no eye control.
 This work addresses this limitation by developing gaze-independent
@@
 their attention to a target without corresponding eye movements.
-A key contribution of this work is the compensation for ERP latency jitter, a
+A key contribution is the compensation for \ac{erp} latency jitter, a
 variability that negatively impacts decoding performance in covert \ac{vsa}.
@@
 
-Beyond gaze independence, the work also advances general \ac{erp} decoding
+Beyond gaze independence, this work also advances general \ac{erp} decoding
 techniques by refining the structure of linear and multilinear estimators.
-These methods improve classification accuracy across a range of \ac{bci}
-conditions, particularly when training data are limited.
+These methods improve accuracy across a range of \ac{bci}
+settings, particularly when training data are limited.
 The introduction of structured regularization in both linear and multilinear
-models enhances the interpretability of the classifiers and reduces training
+models enhances interpretability and reduces training
 time and computational complexity.
-This allows for more efficient training and contributes to generalizability
+This allows for more efficient training and contributes to generalizability,
 allowing for more reliable systems that are adaptable to various contexts and
@@
 
+\bigskip
+
 \paragraph{Keywords:} brain-computer interface, electroencephalography,

### 2024-11-29 chapters/abstract_fr.tex
Second layout run

@@
 Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
-syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
+syndrome d'enfermement, rencontrent des défis considérables
 lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
-Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
-cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
-Ce travail vise \`a pallier cette limitation par améliorer des méthodes de BCI
-indépendantes du regard et la décodage de l'attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
+Ces systèmes nécessitent de orienter le regard, une tâche impraticable pour
+les personnes ayant un contrôle oculaire limité ou inexistant.
+Ce travail vise \`a pallier cette limitation en améliorant des méthodes de BCI
+indépendantes du regard et en optimisant la décodage de l'attention visuo-spatiale (VSA) cachée,
+où l' utilisateurs oriente son attention sans mouvements oculaires correspondants.
 Une contribution clé de ce travail est la compensation de
@@
 
+\bigskip
+
 \paragraph{Mots-clés :}

### 2024-11-29 chapters/abstract_nl.tex
Layout chapter gaze independence
Second layout run

--- 2024-11-24 chapters/abstract_nl.tex
@@
 \chapter*{Samenvatting}
+\addcontentsline{toc}{chapter}{Dutch abstract}
 
-Personen met ernstige motorische beperkingen, bijvoorbeeld het Locked-in
-Syndroom (LIS) of oog-motorische disfunctie, ondervinden aanzienlijke
-uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
-Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen,
-wat onuitvoerbaar wordt voor personen met beperkte of geen controle
-over oogbewegingen.
-Dit werk richt zich op deze beperking door BCI-methoden onafhankelijk van
-oogbewegingen te ontwikkelen, met een focus op het verbeteren van de bedekte
-visuospati\"ele aandacht (VSA), de  aandacht voor een doel zonder
-overeenkomstige oogbewegingen.
-Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
-de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
-van bedekte VSA negatief beïnvloedt.
-Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
-nauwkeurigheid van communicatie onafhankelijkheid van oogcontrole,
-waardoor BCI's bruikbaarder worden voor personen met deze motorische
-beperkingen.
+Personen met ernstige motorische beperkingen, bijvoorbeeld het
+insluitings\-syndroom, ervaren aanzienlijke uitdagingen bij het gebruik van
+traditionele visuele brein-computer interfaces (BCI's).
+Deze systemen vereisen gerichte oogbewegingen wat onpraktisch is voor personen
+met beperkte of zonder oogbewegingscontrole.
+Dit werk ontwikkelt BCI-technieken onafhankelijk van de blik, met nadruk
+op bedekte visuospati\"ele aandacht (VSA) -- aandacht zonder overeenkomstige oogbewegingen.
+Een belangrijke bijdrage is het compenseren van variabiliteit in de
+timing van event-gerelateerde potentialen (ERP's), een factor met een negatieve
+impact op de decodering van bedekte VSA.
+Door dit aan te pakken, verbetert de nauwkeurigheid van communicatie die
+onafhankelijk is van oogcontrole.
+wat BCI's bruikbaarder maakt voor personen met oog-motorische beperkingen.
 
@@
 ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
-Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
-BCI-condities, met name wanneer trainingsgegevens beperkt zijn.
-De introductie van gestructureerde regularisatie in zowel lineaire als
-multilineaire modellen verhoogt de interpretateerbaarheid van de classificators
-en vermindert de trainingstijd en computationele complexiteit.
-Dit maakt efficiëntere training mogelijk en draagt bij aan generalisatie, wat
-de ontwikkeling van betrouwbaardere systemen toelaat die zich kunnen
-aanpassen aan verschillende contexten en gebruikersbehoeften.
+Deze methoden verhogen de nauwkeurigheid in verschillende BCI-toepassingen,
+vooral wanneer kalibratiegegevens beperkt zijn.
+Gestructureerde regularisatie van lineaire en
+multilineaire modellen verhoogt de interpretateerbaarheid
+en verlaagt de trainingstijd en computationele complexiteit.
+Dit draagt bij aan efficiëntere training en generalisatie, wat
+de ontwikkeling van betrouwbaardere systemen toelaat, aangepast aan
+verschillende contexten en gebruikersbehoeften.
 
-De voorgestelde methoden zijn gevalideerd in experimenten met zowel gezonde
-individuen als zeven individuen met een ernstige fysieke beperking en verminderde
-oog-motorische controle.
-Deze experimenten toonden de robuustheid van de nieuwe decodeermethoden aan,
-waarbij werd aangetoond dat het systeem bedekte aandacht effectief kon decoderen,
-zelfs wanneer directe blik onmogelijk was.
+De voorgestelde methoden zijn gevalideerd in experimenten met gezonde
+deelnemers en zeven individuen met een ernstige fysieke beperking en verminderde
+controle over oogbewegingen.
+De resultaten wijzen op de robuustheid van de nieuwe decodeermethoden,
+waarbij werd aangetoond dat het systeem bedekte VSA effectief kan decoderen,
+zelfs zonder directe blik.
+
+\bigskip
 
@@
 electroencephalografie,
-event-gerelateerde potentialen,
-visuospatiale aandacht,
+event-related potentials,
+visuospati\"ele aandacht,
 (multi)lineaire decodering,

### 2024-11-29 chapters/conclusion.tex
Second layout run

@@
 While \ac{wcble} was slightly outperformed by the state-of-the-art in overt \ac{vsa}
-decoding ($94.28<94.74$ and  $85.25<86.60$ \% \ac{rocauc}),
-it increased covert \ac{vsa} decoding ($80.84>78.90$ and $73.38>71.11$ \%
-\ac{rocauc}).
+decoding (94.28<94.74 and  85.25<86.60 \% \acs{rocauc}),
+it increased covert \ac{vsa} decoding (80.84>78.90 and 73.38>71.11 \%
+\acs{rocauc}).
 
@@
 results in a small but significant decrease in performance compared to the state-of-the-art
-($85.25>86.60$ and $94.74>94.28$ \% \ac{rocauc}), consistent with the within-condition
+(85.25>86.60 and 94.74>94.28 \% \acs{rocauc}), consistent with the within-condition
 results.
@@
 \cref{sec:gaze-independence/objectives}, we conclude that we have proposed
-several techniques that did in effect improve visual oddball \ac{erp}-based
+several techniques that did, in effect, improve visual oddball \ac{erp}-based
 \ac{bci} decoding accuracy, and at least one technique that does so specifically

### 2024-11-29 chapters/covert_align.tex
Second layout run

@@
 10 degrees of visual angle from the fixation crosshair.
-\begin{figure}
-  \bigskip
-	\begin{subfigure}{\linewidth}
-		\caption{}
-		\label{fig:covert-align/single-trial-roc-auc-diff}%
-		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
-	\end{subfigure}
-
-	\bigskip
-
-	\begin{subfigure}{\linewidth}
-		\caption{}
-		\label{fig:covert-align/block-eval}%
-		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
-	\end{subfigure}
-  \caption[Decoder performance.]{%
-		\subref{fig:covert-align/single-trial-roc-auc-diff} Difference in cross-validated
-    single-trial classification \acl{rocauc} ($\Delta$ROC-AUC)
+\begin{figure}[t]
+	%\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
+  \input{figures/covert_align/roc_auc_diff.pgf}
+  \caption[Difference in classification \acs{rocauc}.]{%
+    Difference in cross-validated
+    single-trial classification \acl{rocauc} (\%.)
     between \ac{cble}, \ac{wcble}, and their first-stage classifier \ac{tlda}.
@@
     ($d=3$)	\ac{vsa}. \ac{cble} scores on par with \ac{tlda}.
-		\subref{fig:covert-align/block-eval} Cross-validated target selection accuracy for
-    all decoders plotted as a function of the number of test repetitions in
-    different VSA conditions. Significance was determined using one-sided
-    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
-    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
-    covert \ac{vsa} target selection accuracy.}
+    }
+		\label{fig:covert-align/single-trial-roc-auc-diff}%
 \end{figure}
@@
 \ac{wcble} in CVSA-ERP.
+\begin{figure}[t]
+		\label{fig:covert-align/block-eval}%
+		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
+  \caption[Selection accuracy.]{%
+		Cross-validated target selection accuracy for
+    all decoders plotted as a function of the number of test repetitions in
+    different VSA conditions. Significance was determined using one-sided
+    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
+    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
+    covert \ac{vsa} target selection accuracy.}
+\end{figure}
 
@@
 In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
-\begin{figure*}
-  \bigskip
-	\begin{subfigure}{\linewidth}
-		\caption{}
+\begin{figure}[t!]
+  \sffamily
+  \footnotesize
+	\begin{subfigure}{\textwidth}
+		%\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
+    \hspace{-.5in}\input{figures/covert_align/cross_cvsa-erp.pgf}
+		\caption{Cross-condition performance for CVSA-ERP}
 		\label{fig:covert-align/covert-cross-eval}
-		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
 	\end{subfigure}
-
-	\bigskip
-	\bigskip
-
-	\begin{subfigure}[c]{.48\linewidth}
-		\caption{}
+	\begin{subfigure}{\textwidth}
+		%\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
+    \hspace{-.5in}\input{figures/covert_align/cross_bnci2014-009.pgf}
+		\caption{Cross-condition performance for BNCI2014-009}
 		\label{fig:covert-align/aloise2012-cross-eval}
-		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
-	\end{subfigure}\hfill%
-	\begin{subfigure}[c]{.48\linewidth}
-		\caption{}
-		\label{fig:jitter}
-		\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
-	\end{subfigure}
-
-  \caption[Cross-condition classifier performance and estimated jitter.]{
-		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
-    Difference in cross-validated area under the receiver
-		operating characteristic curves between Classifier-based Latency Estimation
+	\end{subfigure}%
+  \caption[Cross-condition classifier performance]{
+    Difference in cross-validated  \ac{rocauc} (\%.) between Classifier-based Latency Estimation
     (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
@@
     CBLE and \ac{tlda}.
-		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
-		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
-    jitter than other conditions.
-    Significance of differences
-    was determined with two-sided Wilcoxon signed-rank tests with False Discovery
-    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
-    $***=p<0.001$, $****=p<0.0001$).
   }
-\end{figure*}
+\end{figure}
 When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
@@
 Results are shown in \cref{fig:jitter}.
+\begin{figure}[t!]
+		%\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
+    \input{figures/covert_align/jitter.pgf}
+    \caption[Estimated jitter]{ Jitter characterized as the interquartile range (IQR)
+	of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
+   jitter than other conditions.
+   Significance of differences
+   was determined with two-sided Wilcoxon signed-rank tests with False Discovery
+   Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
+   $***=p<0.001$, $****=p<0.0001$).}
+		\label{fig:jitter}
+\end{figure}
 

### 2024-11-29 chapters/patients.tex
Second layout run

@@
 number of blocks was decreased to 6 per \ac{vsa} setting.
-\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
+\Ac{isi} was increased to 200$\pm$50 ms to decrease task difficulty.
 The experiment also started with a training block in each condition, where the
@@
 setting was observed for subjects PC3
-(\ac{wcble}: $70.31\%>62.14\%$, XDAWNCov+TS+LDA: $65.78\%>62.18\%$, \ac{tlda}:
-$70.49\%>63.76\%$)
+(\ac{wcble}: 70.31>62.14 \%, XDAWNCov+TS+LDA: 65.78>62.18 \%, \ac{tlda}:
+70.49>63.76 \%)
 and PC4
-(\ac{wcble}: $65.56\%>55.71\%$, XDAWNCov+TS+LDA: $62.02\%>54.24\%$, \ac{tlda}:
-$66.12\%>57.08\%$).
+(\ac{wcble}: 65.56>55.71 \%, XDAWNCov+TS+LDA: 62.02>54.24 \%, \ac{tlda}:
+66.12>57.08 \%).
 For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
@@
 present, but not as substantial
-(\ac{wcble}: $82.76\%>78.88\%$, XDAWNCov+TS+LDA: $80.74\%>77.99\%$, \ac{tlda}:
-$83.21\%>78.84\%$).
+(\ac{wcble}: 82.76>78.88 \%, XDAWNCov+TS+LDA: 80.74>77.99 \%, \ac{tlda}:
+83.21>78.84 \%).
 

### 2024-11-29 chapters/patients_table.tex
Second layout run

@@
 \sffamily
-\small
+\footnotesize
 \begin{tabular}{@{}llrrrlllrr@{}}
@@
   PC3 & brainstem stroke & 43  & F   & R     & anarthric  & yes         & letterboard            & 2  & 3 \\
-  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  basilar \\ artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
+  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  \\ basilar  artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
   \bottomrule

### 2024-11-29 chapters/wcble.tex
Layout chapter wcble
Second layout run

@@
 Christopher Paolini}}
-\noindent\emph{\Cref{sec:wcble/literature/contrib},
+\publicationnote{\Cref{sec:wcble/literature/contrib},
 \crefrange{sec:wcble/methods/cble}{sec:wcble/methods/wcble} (including
 \cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
-\cref{sec:wcble/conclusion} were published as part of~\textcite{VanDenKerchove2024}.}
+\cref{sec:wcble/conclusion} were published as part of \textcite{VanDenKerchove2024}.}
 
@@
 the development of \acp{bci}~\cite{Luck2014}.
-Latency is often assessed as the timing of an \ac{erp} component peak, obtained
+Latency is often assessed as the timing of an \ac{erp} co, **kwargsmponent peak, obtained
 after averaging over trials.
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
   \caption[Schematic representation of the \acs{wcble} training phase.]{%
-    Schematic representation of the \acf{wcble} training phase.
-    (1) The first-stage spatiotemporal binary classifier is trained on a set of
+    The \acf{wcble} training phase.
+    (1) The first-stage classifier is trained on a set of
     epochs.
-    (2) It is then applied to time-shifted copies of the epochs to obtain scores
+    (2) It is then applied to time-shifted copies of the epochs, yielding scores
     and class probabilities over time.
-    (3) The medians of these probability distributions are assumed as the class
+    (3) The medians of the probability distributions are assumed as the class
     latencies.
-    (4) The epochs are aligned to their corresponding class latencies by
-    shifting in time such that all latencies fall at the same moment.
+    (4) The epochs are aligned by shifting in time based on their corresponding class latencies by
+    shifting in time.
     (5) The spatiotemporal classifier is then retrained on the aligned epochs
     for the next iteration.
-    (6) After the iterative process halts, the scores and latencies obtained
+    (6) On convergence, the scores and latencies obtained
     from the last iteration are used to train the second-stage
@@
   \caption[Schematic representation of the \acs{wcble} test phase.]{%
-    Schematic representation of the \acf{wcble}. (1) The
-	  first-stage spatiotemporal binary classifier obtained from the training phase is
-	  applied to time-shifted copies of the epochs to obtain scores and class probabilities
+    The \acf{wcble} test phase. (1) The
+	  first-stage classifier obtained from the training phase is
+	  applied to time-shifted copies of the epochs, yielding scores and class probabilities
 	  over time. (2) The medians of these probability distributions are assumed
@@
 A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
-\fullpagefig{%
-  \inputpgf{figures/wcble}{simulated-sine.pgf}
+\fullpagefig[anchor.north]{%
+  \hspace{1cm}\inputpgf{figures/wcble}{simulated-sine.pgf}
 }{%
@@
 
-\begin{figure}
-  \sffamily\sansmath
-  \input{figures/wcble/fig_latency.pgf}
-  \caption[Synthetic latency estimation results]{Correctness of estimated
-  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
-  data, measured as Pearson correlation coefficient $\rho$ between
-  estimated and ground truth latencies.
-  \Ac{wcble} estimates latencies more accurately in the presence of noise and
-  high jitter.}
-  \label{fig:wcble/results/latency}
-\end{figure}
-
 % AND HERE ====================================================================
@@
 
-\begin{figure}
-  \sffamily\sansmath
-  \input{figures/wcble/fig_accuracy.pgf}
-  \caption[Synthetic decoding results]{Decoding performance, measured as binary
-  classification accuracy, for \ac{cble}, \ac{wcble}, and their first stage
-  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
-  synthetic evoked target and non-target data.
-  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
-  \label{fig:wcble/results/accuracy}
-\end{figure}
-
 \Cref{fig:wcble/results/accuracy} shows that the classification accuracy of all
@@
 and \ac{wcble} performed below the chance level of 50\%.
+
+\begin{figure}[t!]
+  \input{figures/wcble/fig_latency.pgf}
+  \caption[Synthetic latency estimation results]{Correctness of estimated
+  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
+  data, measured as Pearson correlation coefficient $\rho$ between
+  estimated and ground truth latencies.
+  \Ac{wcble} estimates latencies more accurately in the presence of noise and
+  high jitter.}
+  \label{fig:wcble/results/latency}
+\end{figure}
+
+\begin{figure}[t!]
+  \sffamily\sansmath
+  \input{figures/wcble/fig_accuracy.pgf}
+  \caption[Synthetic decoding results]{Decoding performance, measured as binary
+  classification accuracy, for \ac{cble}, \ac{wcble}, and their first stage
+  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
+  synthetic evoked target and non-target data.
+  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
+  \label{fig:wcble/results/accuracy}
+\end{figure}
 
@@
 \begin{algorithm}[h]
+  \small
 	\textsc{Train}

### 2024-11-29 scripts/covert_align/cross_roc_auc.py

@@ -1,0 +1,226 @@
+#!/usr/bin/env python
+import ipdb
+import matplotlib.pyplot as plt
+import pandas as pd
+import scipy.stats
+import seaborn as sns
+from matplotlib.gridspec import GridSpec
+from setup import (configure_matplotlib_style, diverging_cmap, save_pgf_trim,
+                   textwidth_in)
+from statsmodels.stats.multitest import multipletests
+
+configure_matplotlib_style()
+
+# Read data
+df_our = pd.read_csv("data/covert_align//results_CVSA-ERP_cross.csv")
+# df_our_riemann = pd.read_csv("data/results_CVSA-ERP_erpcov-ts-logreg_cross.csv")
+# df_our_riemann = df_our_riemann[df_our_riemann["model"] == "Riemann"]
+# df_our = pd.concat(
+#    [
+#        df_our,
+#        df_our_riemann,
+#    ],
+#    ignore_index=True,
+# )
+df_our["dataset"] = "CVSA-ERP"
+df_our["session"] = 1
+
+
+# for sub_exclude in EXCLUDE["CVSA-ERP"]:
+#    df_our = df_our[df_our["subject"] != sub_exclude]
+
+df_aloise2012 = pd.read_csv("data/covert_align/results_BNCI2014-009_cross.csv")
+# df_aloise2012_riemann = pd.read_csv(
+#    "data/results_BNCI2014-009_erpcov-ts-logreg_cross.csv"
+# )
+# df_aloise2012_riemann = df_aloise2012_riemann[
+#    df_aloise2012_riemann["model"] == "Riemann"
+# ]
+# df_aloise2012 = pd.concat(
+#    [df_aloise2012, df_aloise2012_riemann],
+#    ignore_index=True,
+# )
+df_aloise2012["dataset"] = "BNCI2014-009"
+df_aloise2012["session"] = df_aloise2012["session"].astype(int)
+df = pd.concat([df_our, df_aloise2012])
+
+df["roc_auc"] = df["roc_auc"] * 100
+
+replace = {
+    "S": "overt",
+    "G": "covert",
+    "split/1": "split ($d=1$)",
+    "split/2": "split ($d=2$)",
+    "split/3": "split ($d=3$)",
+    "mix": "combined",
+}
+df["train_cond"].replace(replace, inplace=True)
+df["test_cond"].replace(replace, inplace=True)
+categories = [
+    "overt",
+    "covert",
[…]

### 2024-11-29 scripts/covert_align/jitter.py

@@ -1,0 +1,126 @@
+#!/usr/bin/env python
+import ipdb
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import scipy.stats
+import seaborn as sns
+from setup import (condition_cmap, configure_matplotlib_style, save_pgf_trim,
+                   textwidth_in)
+from statannotations.Annotator import Annotator
+
+configure_matplotlib_style()
+
+# ROC AUC score
+# df_covert = pd.read_csv("data/results_covert_cross.csv")
+# df_covert["dataset"] = "CVSA-ERP"
+# df_covert["session"] = 1
+# df_aloise2012 = pd.read_csv("data/results_arico2014_cross.csv")
+# df_aloise2012["dataset"] = "BNCI2014-009"
+# df = pd.concat([df_covert, df_aloise2012], ignore_index=True)
+# df = df[df["train_task"] == df["test_task"]]
+#
+# df["condition"] = df["train_task"]
+replace = {
+    "S": "overt",
+    "G": "covert",
+    "split/1": "split\n($d=1$)",
+    "split/2": "split\n($d=2$)",
+    "split/3": "split\n($d=3$)",
+}
+# df["condition"].replace(replace, inplace=True)
+
+#
+# df = df.pivot(
+#    index=["dataset", "subject", "session", "condition", "fold"],
+#    columns="model",
+#    values="roc_auc",
+# )
+# df = df.groupby(["dataset", "subject", "session", "condition"])
+# df = df.aggregate("mean")
+
+## counting accuracy
+# df_count = pd.read_csv("data/counting_accuracy.csv")
+# df_count["condition"].replace(replace, inplace=True)
+# df_count["dataset"] = "covert"
+# df_count["accuracy"] = (
+#    1 - np.abs(df_count["actual"] - df_count["count"]) / df_count["actual"]
+# )
+# df_count = df_count.groupby(["dataset", "session", "subject", "condition"])
+# df["count_accuracy"] = df_count["accuracy"].aggregate("mean")
+
+# Target jitter
+df_jitter = pd.read_csv("data/covert_align/latency.csv")
+df_jitter["dataset"].replace(
+    {"covert": "CVSA-ERP", "aloise2012": "BNCI2014-009"},
+    inplace=True,
+)
+df_jitter = df_jitter[df_jitter["event_name"] == "target"]
+df_jitter["condition"] = df_jitter["task"]
[…]

### 2024-11-29 scripts/covert_align/single_trial_roc_auc.py

@@ -1,0 +1,207 @@
+#!/usr/bin/env python
+
+import ipdb
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import scipy.stats
+import seaborn as sns
+import statsmodels.api as sm
+import statsmodels.formula.api as smf
+from setup import (colors, condition_cmap, configure_matplotlib_style,
+                   save_pgf_trim)
+from statsmodels.stats.multitest import multipletests
+
+configure_matplotlib_style()
+
+# Read BNCI2014-009 dataset
+df_bnci2014_009 = pd.read_csv("data/covert_align/results_BNCI2014-009_cross.csv")
+df_bnci2014_009_riemann = pd.read_csv(
+    "data/covert_align/results_BNCI2014-009_erpcov-ts-logreg_cross.csv"
+)
+df_bnci2014_009_riemann = df_bnci2014_009_riemann[
+    df_bnci2014_009_riemann["model"] == "Riemann"
+]
+df_bnci2014_009 = pd.concat(
+    [df_bnci2014_009, df_bnci2014_009_riemann],
+    ignore_index=True,
+)
+
+df_bnci2014_009["dataset"] = "BNCI2014-009"
+
+df_cvsa_erp = pd.read_csv("data/covert_align/results_CVSA-ERP_cross.csv")
+df_cvsa_erp_riemann = pd.read_csv(
+    "data/covert_align/results_CVSA-ERP_erpcov-ts-logreg_cross.csv"
+)
+df_cvsa_erp_riemann = df_cvsa_erp_riemann[df_cvsa_erp_riemann["model"] == "Riemann"]
+df_cvsa_erp = pd.concat(
+    [df_cvsa_erp, df_cvsa_erp_riemann],
+    ignore_index=True,
+)
+df_cvsa_erp["session"] = 1
+# Read CVSA-ERP dataset
+df_cvsa_erp["dataset"] = "CVSA-ERP"
+df = pd.concat([df_bnci2014_009, df_cvsa_erp], ignore_index=True)
+# Clean up dataset
+df = df[df["train_cond"] == df["test_cond"]]
+df["condition"] = df["train_cond"]
+df = df[df["condition"] != "mix"]
+replace = {
+    "S": "overt",
+    "G": "covert",
+    "split/1": "split ($d=1$)",
+    "split/2": "split ($d=2$)",
+    "split/3": "split ($d=3$)",
+}
[…]

### 2024-11-29 scripts/patients/fig_gaze.py

@@ -48,7 +48,7 @@ fig, axs = plt.subplots(
     len(conditions),
     sharex=True,
     sharey=True,
-    figsize=(textwidth_in, 1.75 * textwidth_in),
+    figsize=(textwidth_in, 1.6 * textwidth_in),
     layout="constrained",
 )
 axs = axs.reshape((len(subjects), len(conditions)))

### 2024-11-29 scripts/setup.py

@@ -22,6 +22,9 @@ condition_cmap = {
     "split ($d=1$)": colors["accent3"],
     "split ($d=2$)": colors["accent3"],
     "split ($d=3$)": colors["accent3"],
+    "split\n($d=1$)": colors["accent3"],
+    "split\n($d=2$)": colors["accent3"],
+    "split\n($d=3$)": colors["accent3"],
 }
 condition_order = ["overt", "covert", "split ($d=1$)", "split ($d=2$)", "split ($d=3$)"]
 

### 2024-11-29 scripts/stbf_struct/accuracy.py

@@ -119,7 +119,7 @@ save_pgf_trim(
     fig,
     axs[0, 0],
     "figures/stbf_struct/accuracy_all.pgf",
-    width=textwidth_in * 1.05,
+    width=textwidth_in,
     rows=8,
     columns=3,
 )

### 2024-11-29 scripts/stbf_struct/training_time.py

@@ -69,4 +69,7 @@ fig, ax = plt.subplots(1, 1)
 plot_training_time(training_time[training_time["n_channels"] == 32], ax)
 
 fig.tight_layout()
-save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)
+fig.set_size_inches(textwidth_in, 0.3 * textwidth_in)
+fig.savefig("figures/stbf_struct/training_time.pgf", bbox_inches="tight", pad_inches=0)
+# save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)
+# save_pgf_trim(fig, ax, "figures/stbf_struct/training_time.pgf", columns=2)

### 2024-11-29 scripts/wcble/fig_simulated.py

@@ -123,8 +123,10 @@ for j, snr in enumerate(snrs):
 
         plot_ts(ax_ts, curr_epochs)
         ax_ts.set_yticks([-10, 10])
+        ax_ts.set_yticklabels([str(int(t)) for t in ax_ts.get_yticks()])
         ax_ts.set_ylabel("µV")
         ax_ts.set_xticks([-2, -1, 0, 1, 2])
+        ax_ts.set_xticklabels([str(int(t)) for t in ax_ts.get_xticks()])
         if j == len(snrs) - 1:
             ax_ts.set_xlabel("time (s)")
         else:
@@ -132,10 +134,12 @@ for j, snr in enumerate(snrs):
             ax_ts.tick_params(axis="x", labelbottom=False)
 
 
+cbar_ax.set_yticklabels([str(int(t)) for t in cbar_ax.get_yticks()])
 save_pgf_trim(
     plt.gcf(),
     axs[0, 0],
     f"figures/wcble/simulated-{stc}.pgf",
-    height=8,
+    rows=8,
+    columns=3,
     transparent=True,
 )

### 2024-11-29 chapters/gaze_independence.tex
Layout conclusion
Second layout run
Accuracy figure

@@
 \epigraph{%
-  %\justifying
+  \justifying
   ``\elide\ This begs the question: if a patient had reliable control
@@
 unreliable, or easily fatigued.
-This group is much larger than the first.''~\cite{Wolpaw2006}}}
+This group is much larger than the
+first.''~\cite{Wolpaw2006}\label{fn:bci/wolpawcrit}}}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
   \input{figures/gaze_independence/lis_venn.tikz.tex}

### 2024-11-29 cover_kul.tex
All review comments
Accuracy figure

@@
-\begin{titlepage}
+{%
+\small
+\parindent0pt
   \sffamily
@@
 
-    \noindent\makebox[\textwidth][c]{%
-      \begin{minipage}{.8\textwidth}
+  \makebox[\textwidth][c]{%
+    \begin{minipage}{\textwidth}
       \centering
       {%
-          \Huge
+        \huge
         \bfseries
@@
   \vfill
-    \begin{minipage}[b]{.5\textwidth}
+  \begin{minipage}[b]{.6\textwidth}
     \raggedright
-      \small
     Jury: \\
@@
       \raggedright
-        \small
       Dissertation presented in partial fulfillment of the requirements for
@@
   \centering
-    Defended December 16, 2024
-
-\end{titlepage}
+  \hfill Defended December 16, 2024\hfill
+}

### 2024-11-29 cover_ulille.tex
Accuracy figure

@@
 {%
+  \small
 \parindent0pt
@@
     \vfill
-      \small
       Th\`ese soutenue le 16 d\'ecembre 2024 en satisfaction partielle des exigences de double

### 2024-11-29 scripts/covert_align/block_accuracy.py

@@ -1,0 +1,164 @@
+#!/usr/bin/env python
+import ipdb
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import scipy.stats
+import seaborn as sns
+from setup import configure_matplotlib_style, model_cmap, save_pgf_trim
+from statannotations.Annotator import Annotator
+from statannotations.stats.StatTest import StatTest
+
+# from statannotations.Annotator import Annotator
+
+configure_matplotlib_style()
+
+
+# Read data
+results_CVSA_ERP = pd.read_csv("data/covert_align/results_CVSA-ERP_blocks.csv.bak")
+results_CVSA_ERP_riemann = pd.read_csv(
+    "data/covert_align/results_CVSA-ERP_riemann_blocks.csv"
+)
+# results_CVSA_ERP_riemann["task"] = results_CVSA_ERP_riemann["condition"]
+results_CVSA_ERP = pd.concat(
+    [
+        results_CVSA_ERP_riemann,
+        results_CVSA_ERP,
+    ]
+)
+results_CVSA_ERP["dataset"] = "CVSA-ERP"
+results_CVSA_ERP["condition"] = results_CVSA_ERP["condition"].replace(
+    {
+        "split/1": "split ($d=1$)",
+        "split/2": "split ($d=2$)",
+        "split/3": "split ($d=3$)",
+    }
+)
+
+results_geospell = pd.read_csv("data/covert_align/results_BNCI2014-009_blocks.csv")
+
+results_geospell_riemann = pd.read_csv(
+    "data/covert_align/results_BNCI2014-009_riemann_blocks.csv"
+)
+# results_geospell_riemann["task"] = results_geospell_riemann["condition"]
+results_geospell = pd.concat([results_geospell_riemann, results_geospell])
+geospell_dataset_name = "BNCI2014-009"
+results_geospell["dataset"] = geospell_dataset_name
+results_geospell["condition"] = results_geospell["condition"].replace(
+    {"S": "overt", "G": "covert"}
+)
+#
+#
+df = pd.concat([results_geospell, results_CVSA_ERP])
+df = df.replace(
+    {
+        "XDAWNCov-TS-LogReg": "XDAWNCov-TS-LR",
+        "ERPCov-TS-LogReg": "ERPCov-TS-LR",
+    }
[…]

### 2024-11-29 scripts/setup.py

@@ -29,12 +29,12 @@ condition_cmap = {
 condition_order = ["overt", "covert", "split ($d=1$)", "split ($d=2$)", "split ($d=3$)"]
 
 
-MODEL_CMAP = {
-    "WCBLE": "#AF56BA",
-    "CBLE": "#b488ba",
-    "tLDA": "#bababa",
-    "XDAWNCov-TS-LR": "#FFC61E",
-    "ERPCov-TS-LR": "#F28522",
+model_cmap = {
+    "WCBLE": colors["accent3"],
+    "CBLE": "#ceb3e2",
+    "tLDA": colors["darkgray"],
+    "XDAWNCov-TS-LR": colors["accent1"],
+    "ERPCov-TS-LR": colors["accent2"],
 }
 
 

### 2024-11-30 chapters/abstract_fr.tex
Preface

@@
 syndrome d'enfermement, rencontrent des défis considérables
-lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
+lors de l'uti\-li\-sa\-tion d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
 Ces systèmes nécessitent de orienter le regard, une tâche impraticable pour

### 2024-11-30 chapters/preface.tex
Recording technologies
cover page
Started processing comments MDV
Second layout run
Preface

@@
-\chapter{Preface}
-The mind works in mysterious ways\ldots\
-Let's try to hack it!
+\chapter*{Preface}
+\addcontentsline{toc}{chapter}{Preface}
 
-I'd like to acknowledge
+The past four years have been a stimulating, instructive, and intense
+journey of scientific and self-discovery. What started as a vague interest
+in the applications of machine learning in the biomedical domain during
+my master's studies rapidly evolved into a fascination with neural
+signal processing, from the moment I came into contact with ith.
+This was supercharged by the very high science-fiction factor
+of Brain-Computer Interfaces.
+There is just something about both the brain and computers, and
+their beautiful complexity-from-simplicity, that I just can't
+seem to let go of.
+
+However, it is the potential impact on people's lives that first drew
+me to the field. The biggest advantage my PhD project had to offer was
+its holistic, end-to-end nature, involving everything from signal
+processing and innovating with machine learning under the hood to EEG
+experimentation and, eventually, working with end-user patients. I have
+always considered this PhD, first and foremost, a learning experience—
+one that has provided me, after a lot of trial and error, with the
+toolkit to, hopefully, improve lives through advancements in biomedical
+technology in the future.
+
+For this, I am grateful to my thesis supervisors. The excellent and
+complementary combination of the network, resources, advice, and
+feedback you provided helped this project succeed. I would specifically
+like to thank prof. Marc Van Hulle for the many opportunities you have
+given me, prof. François Cabestaing for your close, active, and
+involved supervision, and prof. Hakim Si-Mohammed for the many
+sharp-witted ideas and contributions you brought to the table. I also
+want to thank the members of the jury, prof. Andrea Kübler, prof.
+Fabien Lotte, prof. Reinhold Scherer, prof. Maarten De Vos, and prof.
+Adalberto Simeone, for the time and effort you have spent evaluating
+this thesis and for your insightful feedback.
+
+Furthermore, I would like to explicitly thank everyone else who helped
+in realizing the work presented in this thesis. Juliette, your
+proactive attitude and multidisciplinary perspective—and, of course,
+your native command of the French language—were an immense help in
+data collection. Without you, this thesis would probably be missing its
+last chapter, and I hope you enjoy your well-deserved upcoming PhD
+career. Working with participants with physical impairment would also
+not have been possible without the medical advice of prof. Etienne
+Allart from the Neurorehabilitation Unit at CHU Lille. I would also
+like to thank the others involved in recruiting and interacting with
+these participants: Alixe and dr. Edward Schietekatte at TRAINM Neuro
+Rehab Clinics, prof. Kristl Claeys and prof. Philip Van Damme from the
+Neuromuscular Reference Center at UZ Leuven, Marie and the team at
+Fondation Partage et Vie, and all colleagues who helped me in data
+collection and processing. Most importantly, I am grateful to everyone
+who participated in our study, and I am sincerely moved by the positive
+reception among those of you who could benefit from it. Of course, I
+would also like to thank all students, colleagues, and friends who
+submitted themselves to having their hair messed up in yet another EEG
+experiment.
+
+Since my thesis was a joint PhD between two labs, one at KU Leuven in
+Belgium and one at the University of Lille in France, I got to
+experience working at two different research institutions, which formed
+an invaluable enrichment of my mini-career so far. This also means I've
+had the amazing luck of having double the amount of wonderful
+colleagues.
+
+In Lille, I've very much appreciated working with Gaël. Your presence
+never failed to lighten up the mood, and I wish you all the best luck
+in finishing up your own PhD. Next, I want to thank Jimmy for showing
+me the way through Lille and its campus life, and for all the tea. I
+also enjoyed discussing with José, Jean-Marc, and Marie-Hélène. Finally,
+I'm also glad that the SIGMA team was around to keep me in touch with
+my computer science roots. Even though I was not always physically
+present in Lille, I've always considered all of you as very good
+colleagues, and I'm glad I've had the opportunity to revive my high
+school French, which was in dire need of a refresher.
+
+In Leuven, I want to begin with a special thanks to those of you who
+guided me in the beginning of my PhD: Benjamin, for introducing me to
+the wonderful world of ERPs and linear decoding, and Axel and Arno for
+powering through COVID with me. As for the rest of you lot—Aline,
+Anahit, Aurélie, Barbara, Bob, Bob, Elvira, Eva, Li Ang, Liuyin, Mani,
+Mani, Mansoureh, Qiang, Tjaša, Valentina, and Yide—you've made my days
+brighter in many ways, and I could always count on you for advice,
+support, or help. To Dilara, Gijs, Lunky, Reniflal, and Zoë, master
+students and interns that I was fortunate enough to supervise: thank
+you for your dedication, and good luck with your future adventures. I
+would also like to mention the great work of the students at NeuroTech
+Leuven—you never fail to impress me with wonderful ideas and excellent
+execution. A final shoutout to everyone in and around the
+Neurophysiology lab: Alessia, Astrid, Christophe, Inez, Jonathan,
+Kenneth, Noa, Saba, Sara, Sofie, Stijn, Wouter, and all others.
+
+On a more personal note, I probably wouldn't have gotten here if it was
+not for the support (and proofreading) I have \emph{always} experienced
+from my family and friends.
+Cato, Loes, Mama, and Papa, I think you've provided about the
+best nest I could imagine to grow up in. To Aderik, Joran, Siebe: thank you for
+carrying me throughout my university studies.
+To Hanne, Ingrid, Jonah, Kiefer, Joshua, Mathias, Vik, and Yana, I have but one
+thing to say: Snek is love, Snek \emph{is} life.
+
+Some people say whoever is mentioned first on a paper is the most
+important, since they're supposed to have done done all the work.
+Without the last author on there, however, most scientific publications would
+either never have materialized or go unnoticed. Therefore, I would like to
+mention my partner Liesel here at the end.
+Liesel, thank you for everything you have done
+for me. This does not only include the last few months, taking care of
+practical arrangements, proofreading, managing, and putting up with me as I was
+focused -- maybe a bit too much -- on work, \ldots, but much, much more than
+that.
+I still cannot believe how lucky I am to have you by my side.
+
+As you turn this page and start reading the first chapter, I will also
+start a new chapter in my life. You, on the other hand, will now dive
+into about 200 pages of algebra, convoluted procedures, and tables with
+numbers. Enjoy!

### 2024-11-30 chapters/abstract_nl.tex
Preface, ironed out overhangs

@@
 vooral wanneer kalibratiegegevens beperkt zijn.
-Gestructureerde regularisatie van lineaire en
+Ge\-struc\-tu\-reerde regularisatie van lineaire en
 multilineaire modellen verhoogt de interpretateerbaarheid

### 2024-11-30 chapters/bci.tex
Layout chapter stbf-struct
Preface, ironed out overhangs

@@
 
-\fullpagefig[topleft]{%
+\fullpagefig[caption.north]{%
   \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
@@
 
-\fullpagefig[topleft]{%
+\fullpagefig[caption.north]{%
   \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
@@
 	  cross-section of the skull.
-	  There is a trade-off between invasiveness and \\ spatial resolution.
+	  There is a trade-off between invasiveness and spatial resolution.
   }
@@
 instance through redirecting attention or initiating imagined actions,
-paradigms free of this requirement are classified as \emph{passive \ac{bci}}.
+pa\-ra\-digms free of this requirement are classified as \emph{passive \ac{bci}}.
 From the user's point of view, this concept is probably the most attractive, since their
@@
 Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
-motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
+motor command~\cite{Neeling2019}, or combining multiple visual pa\-ra\-digms by stimulating
 along multiple dimensions at once.
@@
 by removing frequencies that are too high or too low to contain relevant neural signals.
-Band-pass filters eliminate slow drifts and high-frequency noise, such as muscle
+Band-pass filters eliminate slow drifts and high-fre\-quen\-cy noise, such as muscle
 artifacts or environmental electrical interference, improving the \ac{snr}.
@@
 statistically independent components.
-Components corresponding to eye movements, blinks, and other bodily or environmental
+Components corresponding to eye movements, eye blinks, and other bodily or environmental
 artifacts can be identified, either visually based on their characteristic topographies
@@
 pathophysiology, psychological factors such as the ``extinction of goal directed thinking''
-phenomenon in complete impairment, and, importantly, sensory impairments such as vision and
+phe\-no\-me\-non in complete impairment, and, importantly, sensory impairments such as vision and
 oculomotor defects~\cite{Seguin2023}.

### 2024-11-30 chapters/bttda.tex
Preface
Preface, ironed out overhangs

@@
 \begin{enumerate*}[label={\arabic*)}]
-  \item  We develop a forward model for \ac{hoda} to reconstruct a
+  \item  We develop a forward model for \ac{hoda}. It can reconstruct a
 	      given input tensor from the extracted features.
@@
 supervised tensor-based feature extraction technique.
-For a set of $N$ tensors of order $K$
-$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
-		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
+Let $\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
+		D_K}\right\}_n^N$ be a set of $N$ tensors of order$K$. HODA finds projection matrices $\mat{U_k}$ for each mode $k$
 that project a given $\ten{X}$ to a latent tensor
@@
 \begin{algorithm}
+  \sffamily
   \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
@@
 or, equivalently~\cite{Parra2005,Haufe2014},
-\begin{equation}
+\begin{align}
 	\left\{\mat{A}^*\right\}
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2
-  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
-\end{equation}
+  &= \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2  \\
+  &= \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
+\end{align}
 This least squares tensor approximation problem can be solved efficiently using the
@@
 \begin{algorithm}
+  \sffamily
   \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
@@
 \begin{algorithm}
+  \sffamily
   \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
@@
 \begin{algorithm}
+  \sffamily
   \caption[Greedy model selection procedure.]{Greedy model selection.}
@@
 Alternatively, a special case of \ac{bttda} can be constructed using only rank-one
-blocks such that the resulting forward model adheres to the PARAFAC structure.
+blocks such that the resulting forward model adheres to the PA\-RA\-FAC structure.
 The optimal number of rank-one blocks can be found by truncating as above.
@@
 greedy model selection algorithm in \cref{alg:model-selection}.
-For HODA+LDA and the HODA blocks in BTTDA+LDA, we chose
+For HODA+LDA, as well as for the HODA blocks in BTTDA+LDA, we chose
 $r=r_1=r_2=\ldots=r_K$ with possible values
@@
 HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
-$p=0.0039$) and the BNCI2015-003 (PARAFACDA+LDA: $p=0.0001$, BTTDA+LDA:
+$p=0.0039$) and the BNCI2015-003 (PA\-RA\-FAC\-DA+LDA: $p=0.0001$, BTTDA+LDA:
 $p=0.0049$) datasets.
@@
 \subsection{Contribution}
-The results in \cref{tab:erp-score} and \autoref{tab:mi-score} show that
-in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
+The results listed in \cref{tab:erp-score} and \autoref{tab:mi-score} show
+that, for the tested ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
 exceeding all comparison methods for the two evaluated datasets, but
@@
 \ac{mi} datasets.
-The classification score for BTTDA+LDA was also slightly
-higher than that of
+The classification score obtained by BTTDA+LDA was also slightly
+higher than that of the
 restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
@@
 \ac{bttda} is a generalization of \acf{hoda} and can also be
-applied as a special sum-of-rank-one tensors PARAFACDA model.
+applied as a special sum-of-rank-one tensors PA\-RA\-FAC\-DA model.
 The model is obtained by iteratively fitting \ac{hoda} in a deflation scheme,

### 2024-11-30 chapters/conclusion.tex
Preface, ironed out overhangs

@@
 
-Coincidentally, advances in gaze-independent decoding also build towards a
+Coincidentally, making advances in gaze-independent decoding also builds towards a
 solution for the Midas Touch Problem in visual \ac{bci}.
@@
 
-The optimal approach should unify decoder development, paradigm choice, and interface
+The optimal approach should unify decoder development, choice of paradigm, and interface
 design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both

### 2024-11-30 chapters/conflict.tex
Preface, ironed out overhangs

@@
 
-The authors acknowledge the support by the RITMEA project co-financed by the
+The authors acknowledge the support by the RITMEA project co-fi\-nan\-ced by the
 European Union with the European Regional Development Fund, the French state,

### 2024-11-30 chapters/covert_align.tex
Accuracy figure
Preface, ironed out overhangs

@@
 visual \acp{bci} (see \cref{sec:gaze-independence/sota}).
-Methods of directly enhancing gaze-independent decoding through optimizing
+Methods that directly enhance gaze-in\-de\-pen\-dent decoding through optimizing
 decoder design have not been widely studied (see
@@
 60 cm from a CRT-emulating monitor (VPixx
-Technologies, Canada) operating at a refresh rate of 120 Hz, displaying 6
+Technologies, Ca\-na\-da) operating at a refresh rate of 120 Hz, displaying 6
 circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
@@
 
-The experiment started with five non-recorded practice stimulation blocks, one for
+The experiment started with a sequence  of five non-recorded practice stimulation blocks, one for
 each of the five \ac{vsa} conditions.
@@
 		\label{fig:covert-align/block-eval}%
-		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
+		%\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
+    \input{figures/covert_align/block_accuracy.pgf}
   \caption[Selection accuracy.]{%
@@
     all decoders plotted as a function of the number of test repetitions in
-    different VSA conditions. Significance was determined using one-sided
-    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
-    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
+    different VSA conditions. \ac{wcble} generally achieves the highest
     covert \ac{vsa} target selection accuracy.}

### 2024-11-30 chapters/cv.tex
Started processing comments MDV
Remove at
Remove at and .tex in include
Preface
Preface, ironed out overhangs

@@
 	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
-		Our research project titled ``A gaze-independent Visual Brain-Computer
+		Our research project titled ``A gaze-independent Visual Brain-Com\-pu\-ter
 		Interface for use by patients with limited or no eye control'' won a spot in
@@
 \subsection*{Conferences and presentations}
+\subsubsection*{Presented}
 \begin{cvlist}
@@
 		Poster presentation on classifier-based latency estimation for
-		gaze-independent BCIs.}
+		gaze-in\-de\-pen\-dent BCIs.}
 	\cvitem{Leuven.AI Scientific Workshop 2022}{Leuven.AI, Leuven, 2022\\
@@
 		to spatiotemporal beamforming decoding of event-related potentials.}
+\end{cvlist}
+
+\subsubsection*{Attended}
+\begin{cvlist}
 	\cvitem{BCI \& Neurotechnology Spring School 2024}{g.tec, 2024}
@@
 
-\subsubsection{Master theses and internships supervised}
+\subsubsection*{Master theses and internships}
 \begin{cvlist}
@@
 	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
-		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
+		and Frontotemporal De\-men\-tia}{Zoe Barinaga, KU Leuven, 2023--2024}
 	\cvitem{Single-trial ERP latencies as a predictor for the mode
@@
 
-\subsection*{Languages}
-\begin{cvlist}
-	\cvitem{Dutch}{native}
-	\cvitem{English}{fully proficient}
-	\cvitem{French}{advanced}
-	\cvitem{German}{intermediate}
-\end{cvlist}
+%\subsection*{Languages}
+%\begin{cvlist}
+%	\cvitem{Dutch}{native}
+%	\cvitem{English}{fully proficient}
+%	\cvitem{French}{advanced}
+%	\cvitem{German}{intermediate}
+%\end{cvlist}

### 2024-11-30 chapters/gaze_independence.tex
Preface, ironed out overhangs

@@
 Examples of this are imagined movement or imagined speech paradigms.
-These paradigms can yield very high information transfer
+Ac\-tive paradigms can yield very high information transfer
 rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
@@
 The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
-variants of the \ac{rsvp} paradigm are temporally organized.
+variants of the \ac{rsvp} pa\-ra\-digm are temporally organized.
 \textcite{Lin2018} follow the philosophy that spatially organized interfaces
@@
 
-The previous two methods did not yet yield specific results in gaze-independent
+The previous two methods did not yet yield specific results in
+gaze-in\-de\-pen\-dent
 decoding.
@@
 deal with the low \ac{snr} of the \ac{erp}.
-\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
+\Ac{cble} as described by \textcite{Mowla2017} is a technique that can leverage
 \ac{erp} latency estimation in a decoding setting, but our results show that it
 yields no improvement in gaze-independent settings.
-We improved upon this technique and extended it to a probabilistic, iterative
-method named \ac{wcble}.
+We improved upon this technique and extended it to a probabilistic and iterative
+me\-thod, named \ac{wcble}.
 This approach has been published by~\textcite{VanDenKerchove2024} and is presented
@@
 
-The hypothesis was that gaze-independent decoders, such as \ac{wcble}~\cite{VanDenKerchove2024},
+The hypothesis was that gaze-independent decoders, such as \ac{wcble},
 would improve performance in covert and free \ac{vsa} conditions compared to standard
@@
 performance in free and covert conditions, leveraging residual eye motor control during
-calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
+calibration to improve overall \ac{bci} accuracy for gaze-im\-pair\-ed users.
 The results of this study are presented in \cref{sec:patients}.

### 2024-11-30 chapters/patients.tex
Accuracy figure
Preface, ironed out overhangs

@@
 
-\Acp{bci} for communication assistive technology~\cite{Millan2010}
+\Ac{bci} assistive technologies for communication~\cite{Millan2010}
 target individuals with \acf{sspi}~\cite{Peters2022}.
@@
 target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
-gaze-independent solutions.
+gaze-in\-de\-pen\-dent solutions.
 These studies build on the assumption that \ac{bci} users with \ac{sspgi}
@@
 
-\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
+\textcite{Lesenfants2014} tested a \ac{bci} using gaze-independent \ac{ssvep} in six
 participants with \ac{lis} yet only exceeded chance level accuracy in two.
@@
 
-\section{Methods}
+\section{Materials \& methods}
 \subsection{Recruitment}
@@
 
-Another limiting factor is the difficulty in correctly interpreting eye
+Another limiting factor is the difficulty experienced in correctly interpreting eye
 tracker results in studies with individuals with gaze impairments.

### 2024-11-30 chapters/preface.tex
Preface, ironed out overhangs

@@
 of Brain-Computer Interfaces.
-There is just something about both the brain and computers, and
+There is something about both the brain and computers, and
 their beautiful complexity-from-simplicity, that I just can't
@@
 in realizing the work presented in this thesis. Juliette, your
-proactive attitude and multidisciplinary perspective—and, of course,
-your native command of the French language—were an immense help in
+proactive attitude and multidisciplinary perspective were an immense help in
 data collection. Without you, this thesis would probably be missing its
@@
 In Lille, I've very much appreciated working with Gaël. Your presence
-never failed to lighten up the mood, and I wish you all the best luck
-in finishing up your own PhD. Next, I want to thank Jimmy for showing
+never failed to lighten up the mood, and I wish you smooth sailings
+in completing your own PhD. Next, I want to thank Jimmy for showing
 me the way through Lille and its campus life, and for all the tea. I
 also enjoyed discussing with José, Jean-Marc, and Marie-Hélène. Finally,
-I'm also glad that the SIGMA team was around to keep me in touch with
+I'm glad that the SIGMA team was around to keep me in touch with
 my computer science roots. Even though I was not always physically
@@
 brighter in many ways, and I could always count on you for advice,
-support, or help. To Dilara, Gijs, Lunky, Reniflal, and Zoë, master
+support, or help. To Dilara, Gijs, Lunky, Reniflal, and Zoe, master
 students and interns that I was fortunate enough to supervise: thank

### 2024-11-30 chapters/stbf_struct.tex
Second layout run
Preface, ironed out overhangs

@@
 \epigraph{%
+  \justifying
   I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
@@
 
-	\begin{figure}
+  \begin{figure}[t]
 		%\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
@@
 	Data processing and classifier analysis were performed in Python using
-	Scikit-Learn (version 1.0.1)~\cite{Pedregosa2011} and SciPy (version
+	Scikit-Learn (v1.0.1)~\cite{Pedregosa2011} and SciPy (version
 	1.7.1)~\cite{Virtanen2020}.
 	The preprocessing pipeline was implemented using the MNE-Python toolbox
-	(version 0.24.0)~\cite{Gramfort2013}.
+	(v0.24.0)~\cite{Gramfort2013}.
 	The dataset was converted to BIDS-\ac{eeg} format~\cite{Pernet2019} and managed and
@@
 	The Riemannian classifier from \cref{sec:riemannian} was implemented using
-	pyRiemann (version 0.2.7).
-	Statistical tests were performed in R (version 4.1.2).
+	pyRiemann (v0.2.7).
+	Statistical tests were performed in R (v4.1.2).
 
@@
 	The Riemannian model (XDAWN+RG) uses the XDAWN spatial filter combined
-	with Riemannian geometry in tangent space as implemented by Barachant et
-	al.~\cite{Barachant2014a}.
+  with Riemannian geometry in tangent space as implemented by
+  \textcite{Barachant2014a}.
 	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
@@
 	Covariance estimation and inversion occupy the largest part of the \ac{stbf} training time.
-	For \ac{stbf-emp} and \ac{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
+	For \ac{stbf-emp} and \ac{stbf-shrunk}, the time complexity of this process
+  is described by $\mathcal{O}(NC^2S^2+C^3S^3)$.
 	Thanks to Property~\ref{prop:stbf-struct/inverse-kronecker}, the complexity can be reduced to
@@
 \begin{subappendices}
-  \section{Accuracies per number of trials and available training data}
-
+  \section{\Acs{stbf-struct} selection accuracies}
   \begin{table}[!ht]
@@
   \end{table}
+  \clearpage
+  \section{\Acs{stbf-shrunk} selection accuracies}
   \begin{table}[!ht]
@@
   \end{table}
+  \clearpage
+  \section{\Acs{stbf-emp} selection accuracies}
   \begin{table}[!ht]
@@
   \end{table}
+  \clearpage
+  \section{XDAWN+RG selection accuracies}
   \begin{table}[!ht]

### 2024-11-30 chapters/wcble.tex
Accuracy figure
Preface
Preface, ironed out overhangs

@@
 Similarly, cross-correlation template matching can be combined with filtering.
-\textcite{Souloumiac2013} uses XDAWN to construct a spatial filter before
+\textcite{Souloumiac2013} use the XDAWN technique to construct a spatial filter before
 template matching the spatially filtered signal. \textcite{Iturrate2014} uses a
@@
 
-Some algorithms separate components into stimulus- and response-locked component clusters based on their
+Some algorithms separate components into (sets of) stimulus- and response-locked component clusters based on their
 latency distributions~\cite{Jung2001, Takeda2008, Zhang1998, Yin2009}.
@@
 \ac{erp} time window, unaware of the class of the epoch under investigation.
-The \ac{cble} algorithm introduced
-by \textcite{Thompson2012} also explicitly applies latency estimation in a
+The \ac{cble} algorithm by \textcite{Thompson2012} also explicitly applies latency estimation in a
 decoding setting.
@@
 
-\section{Methods}
+\section{Materials \& methods}
 \label{sec:wcble/methods}
@@
 samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
-with corresponding training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
+This dataset has a corresponding set of training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
 	\textrm{non-target}\}_{n=1}^N$, and a
@@
 The latency features assuming the epoch belongs to a given class
-$l\in\{\textrm{target},\textrm{non-target}\}$ are then defined as the median of
+$l\in\{\textrm{target}, \textrm{non-target}\}$ are then defined as the median of
 the corresponding distributions
@@
 \label{sec:wcble/methods/wcble}
-To improve performance over \ac{cble}, we propose a new algorithm inspired both by
-\ac{cble} and the aforementioned Woody iteration scheme termed \ac{wcble}.
+To improve performance over \ac{cble}, we propose \ac{wcble} a new algorithm inspired both by
+\ac{cble} and the aforementioned Woody iteration scheme.
 Instead of using \ac{cble} to estimate the features of a second-stage classifier
@@
 \begin{algorithm}[h]
+  \sffamily
 	\textsc{Train}
@@
 \begin{algorithm}[h]
+  \sffamily
   \small

### 2024-12-04 chapters/bci.tex
Proofread preface

@@
 been realized using invasive microelectrode arrays~\cite{Willett2021} and
-$\mu$\Ac{ecog}~\cite{Metzger2023}.
+µ\Ac{ecog}~\cite{Metzger2023}.
 Furthermore, recent advances in recording technology focus on improving implant
@@
 \Ac{eeg} measures the electrical potential over time at a given electrode relative to a
-reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
+reference electrode in volts (V), often on the scale of microvolts (µV).
 As it is easily applicable on the outside of the scalp, many practical consumer-grade

### 2024-12-04 chapters/bttda.tex
Proofread preface

@@
 We evaluated our proposed model in two offline EEG-based BCI decoding problems:
-the event-related potential (ERP) and motor imagery (MI) paradigms using a
+the event-related potential (ERP) and motor imagery (MI) pa\-ra\-digms using a
 selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
@@
 	\input{chapters/bttda/score_erp.tex}
-  \caption[Within-session classification score for 2 event-related potential
-  datasets.]{Area under the receiver operating characteristic curve for
+  \caption[Within-session classification score for 2 \acs{erp}.]{Area under the receiver operating characteristic curve for
 		cross-validated within-session evaluation for HODA and our proposed decoders
@@
 $p=0.0049$) datasets.
-No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.
+No significant difference was found between BTTDA+LDA and PA\-RA\-FAC\-DA+LDA.
 
@@
 	\input{chapters/bttda/score_mi.tex}
-  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
+  \caption[Within-session classification score for 2 \acs{mi} datasets.]{Classification score for
 		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
@@
 higher than that of the
-restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
+restricted PA\-RA\-FAC\-DA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
 respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$

### 2024-12-04 chapters/conclusion.tex
Proofread preface

@@
 when the training sample size increases.
-Similarly, \ac{bttda} offers more flexibility.
+Similarly, \ac{bttda} offers more flexibility than \ac{hoda}.
 
@@
 structure but are combined with efficient model
-selection, such as those proposed by \ac{bttda}.
+selection, such as the methods proposed in conjunction with \ac{bttda}.
 These models try to combine the best of both worlds.
@@
 
-The optimal approach should unify decoder development, choice of paradigm, and interface
+The optimal approach should unify decoder development, choice of pa\-ra\-digm, and interface
 design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both

### 2024-12-04 chapters/cv.tex
Proofread preface

@@
 
-\subsection*{About me}
-I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
-and the University of Lille Research Center for Informatics, Signal Processing,
-and Control Science. My research is centered on developing brain-computer
-interface (BCI) technologies with a primary focus on assisting individuals with
-severe disabilities, especially those with motor and eye movement impairments.
-My goal is to enhance their quality of life by creating accurate and
-user-friendly BCIs. Combining my expertise in machine learning, signal
-processing, and user interface design with insights from neurophysiology, my
-PhD centers around the development of an event-related potential BCI system
-using EEG for real-time communication. I strongly believe in the potential of
-innovative neurotechnology to empower individuals and patients, enabling them
-to lead more independent and fulfilling lives.
-
-\subsection*{Keywords}
-brain-computer interfaces, event-related
-potentials, (multi)linear classification, covert visual attention
+%\subsection*{About me}
+%I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
+%and the University of Lille Research Center for Informatics, Signal Processing,
+%and Control Science. My research is centered on developing brain-computer
+%interface (BCI) technologies with a primary focus on assisting individuals with
+%severe disabilities, especially those with motor and eye movement impairments.
+%My goal is to enhance their quality of life by creating accurate and
+%user-friendly BCIs. Combining my expertise in machine learning, signal
+%processing, and user interface design with insights from neurophysiology, my
+%PhD centers around the development of an event-related potential BCI system
+%using EEG for real-time communication. I strongly believe in the potential of
+%innovative neurotechnology to empower individuals and patients, enabling them
+%to lead more independent and fulfilling lives.
+%
+%\subsection*{Keywords}
+%brain-computer interfaces, event-related
+%potentials, (multi)linear classification, covert visual attention
 
@@
 \begin{cvlist}
-	\cvitem{PhD\ in Biomedical Sciences}{KU Leuven, 2024}
-	\cvitem{PhD\ in Control Science and Signal Processing}{University of
+	\cvitem{Ph.D.\ in Biomedical Sciences}{KU Leuven, 2024}
+	\cvitem{Ph.D.\ in Control Science and Signal Processing}{University of
 		Lille, 2024}
@@
 		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
-	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
-		and Frontotemporal De\-men\-tia}{Zoe Barinaga, KU Leuven, 2023--2024}
+	\cvitem{An EEG connectivity analysis of Dementias}{Zoe Barinaga, KU Leuven, 2023--2024}
 	\cvitem{Single-trial ERP latencies as a predictor for the mode

### 2024-12-04 chapters/patients.tex
Proofread preface

@@
 and \ac{tlda}~\cite{Sosulski2022}
-classifiers, as well as the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
+classifiers, as well as the Riemannian approach XDAWNCov+TS+LDA~\cite{Cecotti2017}.
 For \ac{wcble}, a region of interest from 0 ms to 800 ms relative to stimulus

### 2024-12-04 chapters/preface.tex
Proofread preface

--- 2024-09-16 chapters/preface.tex
@@
-\chapter{Preface}
-The mind works in mysterious ways\ldots\
-Let's try to hack it!
+\chapter*{Preface}
+\addcontentsline{toc}{chapter}{Preface}
 
-I'd like to acknowledge
+The past four and a half years have been a stimulating, instructive, and intense
+journey of scientific and self-discovery. What started as a vague interest
+in the applications of machine learning in the biomedical domain rapidly evolved
+into a fascination with neural signal processing when I came into contact with
+it via my master thesis.
+Ultimately, this was supercharged by the very high science-fiction factor
+of Brain-Computer Interfaces.
+There is something about both the brain and computers, and
+their beautiful complexity arising from simplicity, that I just can't
+seem to let go of.
+
+Yet, it is the tangible impact on people's lives of neurotechnology
+which first drew me to the field.
+The biggest advantage my PhD project had to offer was
+its holistic, end-to-end nature, involving everything from innovation in signal
+processing and machine learning under the hood to EEG
+experimentation and, eventually, working with end-user patients. I have
+always considered this PhD, first and foremost, a learning experience.
+An experience that has provided me -- after a lot of trial and error -- with the
+toolkit to, hopefully, continue making this impact in the future by
+advancing biomedical technology.
+
+For this, I am grateful to my thesis supervisors. The combination of the
+network, resources, advice, and feedback you provided helped this project
+succeed.
+I would specifically like to thank prof. Marc Van Hulle for the many
+opportunities you have given me, prof. François Cabestaing for your close,
+active, and involved supervision, and prof. Hakim Si-Mohammed for the many
+sharp-witted ideas and contributions you brought to the table.
+I also want to thank the members of the jury, prof. Andrea Kübler, prof.
+Fabien Lotte, prof. Reinhold Scherer, prof. Maarten De Vos, and prof.
+Adalberto Simeone, for the time and effort you have spent evaluating
+this thesis and for your insightful feedback.
+
+Furthermore, I would like to explicitly thank everyone else who helped
+in realizing the work presented in this thesis. Juliette, your
+proactive attitude and multidisciplinary perspective were an immense help in
+data collection. Without you, this thesis would probably be missing its
+last chapter, and I hope you enjoy your well-deserved upcoming academic
+career.
+Working with some of our end-user study participants would also
+not have been possible without the medical advice of prof. Etienne
+Allart from the Neurorehabilitation Unit at CHU Lille, as well as
+the others involved in recruiting and interacting with
+these participants: Alixe and dr. Edward Schietekatte at TRAINM Neuro
+Rehab Clinics, prof. Kristl Claeys and prof. Philip Van Damme from the
+Neuromuscular Reference Center at UZ Leuven, Marie and the team at
+Fondation Partage et Vie, and all colleagues who helped me in data
+collection and processing. Most importantly, I am grateful to everyone
+who participated in our study, and I am sincerely moved by the positive
+reception among those of you who could benefit from it. Of course, I
+would also like to thank all other students, colleagues, and friends who
+submitted themselves to having their hair messed up by yet another EEG
+experiment.
+
+Since my thesis was a joint PhD between two labs, one at KU Leuven in
+Belgium and one at the University of Lille in France, I got to
+experience working at two different research institutions
+This formed an invaluable enrichment of my career until now.
+It also means I've had the amazing luck of having double the amount of wonderful
+colleagues.
+
+In Lille, I've very much appreciated working with Gaël. Your presence
+never failed to lighten up the mood, and I wish you smooth sailings
+in completing your own PhD. Next, I want to thank Jimmy for showing
+me the way through Lille and its campus life, and for all the tea. I
+also enjoyed discussing with José, Jean-Marc, and Marie-Hélène.
+Even though I was not always physically
+present in Lille, you've always made sure I felt part of the team.
+Finally, I'm glad that the SIGMA team was around to keep me in touch with
+my computer science roots.
+
+In Leuven, I want to begin with a special thanks to those of you who
+guided me at the start of my PhD: Benjamin, for introducing me to
+the wonderful world of ERPs and linear decoding, and Axel and Arno for
+powering through COVID with me. As for the rest of you lot -- Aline,
+Anahit, Aurélie, Barbara, Bob, Elvira, Eva, Li Ang, Liuyin, Mani, Mansoureh,
+Qiang, Tjaša, Valentina, and Yide -- you've made my days brighter in many ways,
+and I could always count on you, be it for advice, support or help.
+I also wish to thank the students and interns I was fortunate enough to
+supervise for their dedication and great work, as well as the students at NeuroTech Leuven.
+You never failed to impress me with creative ideas and excellent
+execution. A final shoutout to everyone in and around the
+Neurophysiology lab: Alessia, Astrid, Christophe, Inez, Jonathan,
+Kenneth, Lara, Noa, Saba, Sara, Sofie, Stijn, Wouter, and all others.
+
+On a more personal note, I probably wouldn't have gotten here if it wasn't
+for the support (and proofreading) I have \emph{always} experienced
+from my family and friends.
+Cato, Loes, Mama, and Papa, I think you've provided about the
+best nest I could imagine to grow up in. To Aderik, Joran and Siebe: thank you for
+carrying me throughout my university studies.
+To Hanne, Ingrid, Jonah, Joshua, Kiefer, Mathi, Vik, Ward and Yana, I have but one
+thing to say: Snek \emph{is} love, Snek \emph{is} life.
+
+Some say that whoever is mentioned first on a paper is the most
+important, since they're supposed to have done all the work.
+Without the last author on there, however, most scientific publications would
+either never materialize or go unnoticed.
+Therefore, I would like to mention my partner Liesel here at the end.
+Liesel, thank you for everything you do and have done
+for me. This does not only include the past few months and your help with
+practical arrangements, proofreading, and putting up with me as I was
+-- maybe a bit too much -- focused on work, but much, much more
+than that.
+I still cannot believe how lucky I am to have you by my side.
+
+As you turn this page and start reading the first chapter, I will also
+start a new chapter in my life. You, on the other hand, will dive into
+some 200 pages of algebra, convoluted procedures, and tables with
+numbers. Enjoy!

### 2024-12-04 chapters/wcble.tex
Proofread preface

@@
 \end{equation}
-The latency features assuming the epoch belongs to a given class
+The latency features assuming the epoch belongs to a given class given by
 $l\in\{\textrm{target}, \textrm{non-target}\}$ are then defined as the median of

### 2024-12-04 cover_kul.tex
Proofread preface

@@
     Jury: \\
-    Co-supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
-    Co-supervisor Univ. Lille: \emph{prof. François Cabestaing} \\
-    Chair examining committee: \emph{prof. John Creemers} \\
-    Chair public defense: \emph{prof. Koen Poesen} \\
-    \emph{prof. Andrea K\"ubler} \\
-    \emph{prof.\ Fabien Lotte} \\
-    \emph{prof.\ Reinhold Scherer} \\
-    \emph{prof.\ Maarten De Vos} \\
-    \emph{prof.\ Adalberto Simeone}
+    Co-supervisor KU Leuven: \emph{prof.\ Marc Van Hulle} \\
+    Co-supervisor Univ. Lille: \emph{prof.\ François Cabestaing} \\
+    Chair examining committee: \emph{prof.\ John Creemers} \\
+    Chair public defense: \emph{prof.\ Koen Poesen} \\
+    \emph{Prof.\ Andrea K\"ubler} \\
+    \emph{Prof.\ Fabien Lotte} \\
+    \emph{Prof.\ Reinhold Scherer} \\
+    \emph{Prof.\ Maarten De Vos} \\
+    \emph{Prof.\ Adalberto Simeone}
   \end{minipage}\hfill%

### 2024-12-05 chapters/covert_align.tex
Proofread preface
Cover fold line and address, loop figure odd page

@@
 \epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ \emph{A
-Scandal in Bohemia -- Sir Arthur Conan Doyle}}
+Scandal in Bohemia -- Arthur C. Doyle}}
 \publicationnote{This chapter was published as part of~\textcite{VanDenKerchove2024}.}
@@
 \label{sec:covert-align/methods/decoders}
-We compared \ac{wcble} and \ac{cble} as described in
+We compared the two latency-based methods \ac{wcble} and \ac{cble} as described in
 \cref{sec:wcble/methods}, with their base classifier \ac{tlda} and with

### 2024-12-05 chapters/bci.tex
Add odd option to full page fig

@@
 
-\fullpagefig[caption.north]{%
+\fullpagefig[caption.north,odd]{%
   \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
@@
 
-\fullpagefig[caption.north]{%
+\fullpagefig[caption.north, odd]{%
   \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
@@
 	  Various electrophysiology recording technologies and their respective position in a
-	  cross-section of the skull.
-	  There is a trade-off between invasiveness and spatial resolution.
+	  cross-section of \\ the skull.
+	  There is a trade-off between invasiveness \\ and spatial resolution.
   }

### 2024-12-05 chapters/wcble.tex
Add odd option to full page fig

@@
 positive, by the difference between the latency and the window onset.
-\fullpagefig{%
+\fullpagefig[caption.fullwidth, odd]{%
   \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
@@
     deviation).
-    Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
-    spatiotemporal noise were simulated using a forward head model, and channel C3 is
+    Targets contain a sine wave pulse, jittered by $\sigma$. 32 \ac{eeg}
+    channels with varying levels of noise were simulated using a forward model, channel C3 is
     plotted here.

### 2024-12-05 chapters/abstract_nl.tex
Sans-serif acronyms list

@@
 onafhankelijk is van oogcontrole.
-wat BCI's bruikbaarder maakt voor personen met oog-motorische beperkingen.
+Dit maakt BCI's bruikbaar voor personen met oog-motorische beperkingen.
 
@@
 Dit draagt bij aan efficiëntere training en generalisatie, wat
-de ontwikkeling van betrouwbaardere systemen toelaat, aangepast aan
+de ont\-wik\-ke\-ling van betrouwbaardere systemen toelaat, aangepast aan
 verschillende contexten en gebruikersbehoeften.
@@
 controle over oogbewegingen.
-De resultaten wijzen op de robuustheid van de nieuwe decodeermethoden,
+De resultaten wijzen op de robuust\-heid van de nieuwe decodeermethoden,
 waarbij werd aangetoond dat het systeem bedekte VSA effectief kan decoderen,
@@
 brein-computer interface,
-electroencephalografie,
-event-related potentials,
+electro-encefalografie,
+event-re\-la\-ted potentials,
 visuospati\"ele aandacht,
 (multi)lineaire decodering,
-fysieke beperkingen
+fysieke be\-per\-king\-en

### 2024-12-05 chapters/bci.tex
Sans-serif acronyms list

@@
 Invasive recordings are necessary for decoding more complex encoded activities like
-natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
+imagined natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
 symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.

### 2024-12-05 scripts/covert_align/cross_roc_auc.py

@@ -195,7 +195,7 @@ map_df_kws = dict(
 )
 g = sns.FacetGrid(
     df_diff[df_diff["dataset"] == "CVSA-ERP"],
-    height=(textwidth_in + 0.2) / aspect,
+    height=(textwidth_in + 0.0) / aspect,
     **fg_kws,
 )
 g.map_dataframe(**map_df_kws)
@@ -211,7 +211,7 @@ plt.savefig(
 
 g = sns.FacetGrid(
     df_diff[df_diff["dataset"] == "BNCI2014-009"],
-    height=((textwidth_in / 2) + 0.5) / aspect,
+    height=((textwidth_in / 2) + 0.0) / aspect,
     **fg_kws,
 )
 g.map_dataframe(**map_df_kws)

### 2024-12-11 chapters/bci.tex
Fix testprint issues

@@
 \Acp{bci} started as a speculative concept in the 1970's~\cite{Vidal1973}.
-Now, they have gradually evolved from methods  developed in the the 1990's to
+Now, they have gradually evolved from methods  developed in the 1990's to
 establish basic communication from minimal brain activity

### 2024-12-11 chapters/gaze_independence.tex
Fix testprint issues

@@
 Instead of categorizing users by their etiology, we will refer to those who might benefit from
-\ac{bci} assistive communication technology as having \ac{sspi}, in line with
+\ac{bci} assistive communication technology as having \acf{sspi}, in line with
 the terminology used by~\textcite{FriedOken2020}.
 If they have eye motor impairment that affects their use of visual \ac{bci} or
-eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze-independence/venn}).
+eye-tracking solutions, we use the term \acf{sspgi} (see \cref{fig:gaze-independence/venn}).
 

### 2024-12-11 chapters/preface.tex
Fix testprint issues

@@
 Belgium and one at the University of Lille in France, I got to
-experience working at two different research institutions
+experience working at two different research institutions.
 This formed an invaluable enrichment of my career until now.
@@
 guided me at the start of my PhD: Benjamin, for introducing me to
-the wonderful world of ERPs and linear decoding, and Axel and Arno for
+the wonderful world of ERPs and linear decoding, and Arno and Axel for
 powering through COVID with me. As for the rest of you lot -- Aline,
